### 2022-05-10 gmea.Rmd
initial script

---
title: "GSE74548: Gene Methylation Enrichment Analysis"
author: "hcy_meth meta-analysis group"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
---

##  Introduction

Previously, Mandhri and I analysed the B-PROOF 450K data, trying to understand
whether vitamin supplementation caused changes in gene methylation.
We used Limma and some basic analyses, which showed no specific probes with FDR<0.05,
nor any DMRs.

In this analysis we will use the principle of Gene Set Enrichment Analysis, applying it
to many probes belonging to genes.
If the probes are trending in concert, then we can make some judgement about the
enrichment of those probes.
The statistical test used is the Wilcox test, which can be applied to competitive 
or self contained test types.

```{r,libs}

library("parallel")
library("dplyr")
library("kableExtra")
library("eulerr")
library("mitch")

```

# DM1 - post trial supplement versus placebo

## Whole gene analysis

```{r,dma1a}

dm <- read.table("dma1a.tsv")
head(dm,50) %>%
  kbl(caption = "Top significant genes with limma") %>%
  kable_paper("hover", full_width = F)

# histogram of t values
hist(dm$t,breaks=seq(from=-6,to=6,by=1))

# set cores to used for parallel execution
CORES= detectCores()

# make a vector of unique gene names
gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
gnl <- strsplit( dm$UCSC_RefGene_Name ,";")

# run the test (self contained)
gme_res <- mclapply( 1:length(gn), function(i) {
  g <- gn[i]
  myrows <- which(lapply(gnl,function(x) {
    length(which(x==g))
  } ) > 0 )
  tstats <- dm[myrows,"t"]
  tstats <- tstats[order(tstats)]
  myn <- length(tstats)
  mymean <- mean(tstats)
  mymedian <- median(tstats)
  wtselfcont <- wilcox.test(tstats)

  res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,"median"=mymedian,
           "p-value(selfcontained)"=wtselfcont$p.value)
} , mc.cores=CORES)

# format the results
gme_res_df <- do.call(rbind, gme_res)
rownames(gme_res_df) <- gme_res_df[,1]
gme_res_df <- gme_res_df[,-1]
tmp <- apply(gme_res_df,2,as.numeric)
rownames(tmp) <- rownames(gme_res_df)
gme_res_df <- as.data.frame(tmp)

# results are sorted by the sum of -log10 of the p-values
gme_res_df$sig <- -log10(gme_res_df[,4])

gme_res_df <- gme_res_df[order(-gme_res_df$sig),]

gme_res_df$`fdr(selfcontained)` <- p.adjust(gme_res_df$`p-value(selfcontained)`)

gme_res_df_wholegene <- gme_res_df

write.table(gme_res_df_wholegene,file="gmea_wholegene.tsv")

```

## Table of top results.

```{r, toptable}

head(gme_res_df,50) %>%
  kbl(caption = "Top significant genes with GMEA") %>%
  kable_paper("hover", full_width = F)

```

## Volcano plots

```{r,volcano1}

# volcano selfcont
sig <- subset(gme_res_df,`fdr(selfcontained)` < 0.05)
plot(gme_res_df$median , -log10(gme_res_df$`p-value(selfcontained)`) ,
  xlab="effect size (mean t-stat)", ylab="-log10(p-value)",
  pch=19, cex=0.5, col="gray",main="self contained test")
grid()
points(sig$median , -log10(sig$`p-value(selfcontained)`) ,
  pch=19, cex=0.5, col="red")

```

## Boxplots

Boxplots smallest pvalue.

```{r,box1}

par(mfrow=c(1,2))
n=50
# self contained
gs <- head(rownames(gme_res_df[order(gme_res_df$`p-value(selfcontained)`),]),n)
tstats <- mclapply(gs, function(g) {
  myrows <- which(lapply(gnl,function(x) {
    length(which(x==g))
  } ) > 0 )
  dm[myrows,"t"]
},mc.cores=CORES)
names(tstats) <- gs
tstats <- tstats[order(unlist(lapply(tstats,median)))]
boxplot(tstats,horizontal=TRUE,las=1,
  main="smallest p-val(selfcont)",cex.axis=0.6,
  xlab="t-statistic")
grid()

n=50
# effect size median
sig <- subset(gme_res_df,`fdr(selfcontained)` < 0.05)
gs <- head(rownames(sig[order(-abs(sig$median)),]),n)
tstats <- mclapply(gs, function(g) {
  myrows <- which(lapply(gnl,function(x) {
    length(which(x==g))
  } ) > 0 )
  dm[myrows,"t"]
},mc.cores=CORES)
names(tstats) <- gs
tstats <- tstats[order(unlist(lapply(tstats,median)))]
boxplot(tstats,horizontal=TRUE,las=1,
  main="biggest effect size(median)",cex.axis=0.6,
  xlab="t-statistic")
grid()

```

## Mitch

```{r,dm1_mitch}

dmscore <- data.frame( gme_res_df$median * gme_res_df$sig)

rownames(dmscore) <- rownames(gme_res_df)

colnames(dmscore) <- "metric"

#download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", destfile="ReactomePathw>
#unzip("ReactomePathways.gmt.zip")
genesets <- gmt_import("ReactomePathways.gmt")

res <- mitch_calc(x=dmscore, genesets=genesets, priority="effect")

head(res$enrichment_result,20) %>%
  kbl(caption = "Top enriched gene sets with GMEA-Mitch") %>%
  kable_paper("hover", full_width = F)

mitch_report(res,outfile="dma1a_mitch.html",overwrite=TRUE)

```

## Promoter analysis only

```{r,promo}

dm <- read.table("dma1a.tsv")
dm <- subset(dm,Regulatory_Feature_Group == "Promoter_Associated")
dm$adj.P.Val <- p.adjust(dm$P.Value)

head(dm,50) %>%
  kbl(caption = "Top significant genes with limma") %>%
  kable_paper("hover", full_width = F)

# histogram of t values
hist(dm$t,breaks=seq(from=-6,to=6,by=1))

# set cores to used for parallel execution
CORES= detectCores()

# make a vector of unique gene names
gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
gnl <- strsplit( dm$UCSC_RefGene_Name ,";")

# run the test
gme_res <- mclapply( 1:length(gn), function(i) {
  g <- gn[i]
  myrows <- which(lapply(gnl,function(x) {
    length(which(x==g))
  } ) > 0 )
  tstats <- dm[myrows,"t"]
  tstats <- tstats[order(tstats)]
  myn <- length(tstats)
  mymean <- mean(tstats)
  mymedian <- median(tstats)
  wtselfcont <- wilcox.test(tstats)

  res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,"median"=mymedian,
           "p-value(selfcontained)"=wtselfcont$p.value)
} , mc.cores=CORES)

# format the results
gme_res_df <- do.call(rbind, gme_res)
rownames(gme_res_df) <- gme_res_df[,1]
gme_res_df <- gme_res_df[,-1]
tmp <- apply(gme_res_df,2,as.numeric)
rownames(tmp) <- rownames(gme_res_df)
gme_res_df <- as.data.frame(tmp)

# results are sorted by the sum of -log10 of the p-values
gme_res_df$sig <- -log10(gme_res_df[,4])

gme_res_df <- gme_res_df[order(-gme_res_df$sig),]

gme_res_df$`fdr(selfcontained)` <- p.adjust(gme_res_df$`p-value(selfcontained)`)

gme_res_df_promoter <- gme_res_df

write.table(gme_res_df_promoter ,file="gmea_promo.tsv")

```

## Table of top results.

```{r, toptable_promoter}

head(gme_res_df,50) %>%
  kbl(caption = "Top significant genes with GMEA") %>%
  kable_paper("hover", full_width = F)

```

## Volcano plots

```{r,volcano1_promoter}

# volcano selfcont
sig <- subset(gme_res_df,`fdr(selfcontained)` < 0.05)
plot(gme_res_df$median , -log10(gme_res_df$`p-value(selfcontained)`) ,
  xlab="effect size (mean t-stat)", ylab="-log10(p-value)",
  pch=19, cex=0.5, col="gray",main="self contained test")
grid()
points(sig$median , -log10(sig$`p-value(selfcontained)`) ,
  pch=19, cex=0.5, col="red")

```

## Boxplots

Boxplots smallest pvalue.

```{r,box1_promoter}

par(mfrow=c(1,2))
n=50
# self contained
gs <- head(rownames(gme_res_df[order(gme_res_df$`p-value(selfcontained)`),]),n)
tstats <- mclapply(gs, function(g) {
  myrows <- which(lapply(gnl,function(x) {
    length(which(x==g))
  } ) > 0 )
  dm[myrows,"t"]
},mc.cores=CORES)
names(tstats) <- gs
tstats <- tstats[order(unlist(lapply(tstats,median)))]
boxplot(tstats,horizontal=TRUE,las=1,
  main="smallest p-val(selfcont)",cex.axis=0.6,
  xlab="t-statistic")
grid()

n=50
# effect size median
sig <- subset(gme_res_df,`fdr(selfcontained)` < 0.05)
gs <- head(rownames(sig[order(-abs(sig$median)),]),n)
tstats <- mclapply(gs, function(g) {
  myrows <- which(lapply(gnl,function(x) {
    length(which(x==g))
  } ) > 0 )
  dm[myrows,"t"]
},mc.cores=CORES)
names(tstats) <- gs
tstats <- tstats[order(unlist(lapply(tstats,median)))]
boxplot(tstats,horizontal=TRUE,las=1,
  main="biggest effect size(median)",cex.axis=0.6,
  xlab="t-statistic")
grid()

```

## Mitch

```{r,dm1_mitch_promo}

dmscore <- data.frame( gme_res_df$median * gme_res_df$sig)

rownames(dmscore) <- rownames(gme_res_df)

colnames(dmscore) <- "metric"

#download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", destfile="ReactomePathways.gmt.zip")
#unzip("ReactomePathways.gmt.zip")
genesets <- gmt_import("ReactomePathways.gmt")

res <- mitch_calc(x=dmscore, genesets=genesets,priority="effect")

head(res$enrichment_result,20) %>%
  kbl(caption = "Top enriched gene sets with GMEA-Mitch (promoter only)") %>%
  kable_paper("hover", full_width = F)

mitch_report(res,outfile="dma1a_mitch_promo.html",overwrite=TRUE)

```

## Multi-mitch

```{r,dm1_mitch_promoter}

dmscore_wholegene <- data.frame( gme_res_df_wholegene$median * gme_res_df_wholegene$sig )
rownames(dmscore_wholegene) <- rownames(gme_res_df_wholegene)
colnames(dmscore_wholegene) <- "wholegene"

dmscore_promoter <- data.frame( gme_res_df_promoter$median * gme_res_df_promoter$sig )
rownames(dmscore_promoter) <- rownames(gme_res_df_promoter)
colnames(dmscore_promoter) <- "promoter"

mm <- merge(dmscore_wholegene,dmscore_promoter,by=0)
rownames(mm) <- mm[,1]
mm[,1] = NULL

#download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", destfile="ReactomePathw>
#unzip("ReactomePathways.gmt.zip")
genesets <- gmt_import("ReactomePathways.gmt")

mmres <- mitch_calc(x=mm, genesets=genesets,priority="effect")

head(mmres$enrichment_result,20) %>%
  kbl(caption = "Top enriched gene sets with GMEA-Mitch (wholegene and promoter)") %>%
  kable_paper("hover", full_width = F)

mitch_report(mmres,outfile="mitch_multi.html",overwrite=TRUE)

```

## Session Information

For reproducibility.

```{r,sessioninfo}

sessionInfo()

```

### 2022-05-11 gmea.Rmd
adjustments for dma3 (hcy)

@@
 
-# DM1 - post trial supplement versus placebo
+# DM3 - post trial supplement versus placebo
 
@@
 
-```{r,dma1a}
+```{r,dma3a}
 
-dm <- read.table("dma1a.tsv")
+dm <- read.table("dma3a.tsv")
+
+# testing
+dm <- head(dm,40000)
+
 head(dm,50) %>%
@@
 
-mitch_report(res,outfile="dma1a_mitch.html",overwrite=TRUE)
+mitch_report(res,outfile="dma3a_mitch.html",overwrite=TRUE)
 
@@
 
-dm <- read.table("dma1a.tsv")
+dm <- read.table("dma3a.tsv.gz")
 dm <- subset(dm,Regulatory_Feature_Group == "Promoter_Associated")
+
+# testing
+dm <- head(dm,20000)
+
 dm$adj.P.Val <- p.adjust(dm$P.Value)
@@
 
-mitch_report(res,outfile="dma1a_mitch_promo.html",overwrite=TRUE)
+mitch_report(res,outfile="dma3a_mitch_promo.html",overwrite=TRUE)
 

### 2022-05-11 gmea.Rmd
small fixes

@@
 
-dm <- read.table("dma3a.tsv")
+dm <- read.table("dma3a.tsv.gz")
 
@@
 
-#download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", destfile="ReactomePathw>
-#unzip("ReactomePathways.gmt.zip")
+if ( ! file.exists("ReactomePathways.gmt") ) {
+  download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip",
+    destfile="ReactomePathways.gmt.zip")
+  unzip("ReactomePathways.gmt.zip")
+}
+file.info("ReactomePathways.gmt")
 genesets <- gmt_import("ReactomePathways.gmt")
+length(genesets)
 

### 2022-05-11 gmea.Rmd
when there's no FDR sig results

@@
 gs <- head(rownames(sig[order(-abs(sig$median)),]),n)
+
+if ( length(gs) >2 ) {
   tstats <- mclapply(gs, function(g) {
@@
   grid()
+} else {
+  plot(1)
+  mtext("too few significant genes found")
+}
 

### 2022-05-12 gmea.Rmd
fixed a bug with promoter analysis

@@
 
-# DM3 - post trial supplement versus placebo
+# DM3 - association of Hcy with gene methylation
+
+It is thought that high plasma homocysteine can inhibit DNA methylation.
+Let's see whether that is the case, and which genes are most affected.
+This analysis is conducted at the whole gene level as well as on the level of promoters.
 
@@
 # testing
-dm <- head(dm,40000)
+#dm <- head(dm,40000)
 
@@
 dm <- read.table("dma3a.tsv.gz")
-dm <- subset(dm,Regulatory_Feature_Group == "Promoter_Associated")
+dm <- dm[grep("Promoter_Associated",dm$Regulatory_Feature_Group),]
 
 # testing
-dm <- head(dm,20000)
+#dm <- head(dm,20000)
 
@@
 
-#download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", destfile="ReactomePathways.gmt.zip")
-#unzip("ReactomePathways.gmt.zip")
 genesets <- gmt_import("ReactomePathways.gmt")
@@
 
-#download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", destfile="ReactomePathw>
-#unzip("ReactomePathways.gmt.zip")
 genesets <- gmt_import("ReactomePathways.gmt")
@@
 
-mitch_report(mmres,outfile="mitch_multi.html",overwrite=TRUE)
+#mitch_report(mmres,outfile="mitch_multi.html",overwrite=TRUE)
 

### 2022-05-14 gmea.Rmd
optimise speed

@@
 library("mitch")
+library("IlluminaHumanMethylation450kmanifest")
+library("IlluminaHumanMethylation450kanno.ilmn12.hg19")
+library("tictoc")
+
+```
+
+## Get array annotation data
+
+```{r,anno}
+
+ann450k <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
+myann <- data.frame(ann450k[,c("UCSC_RefGene_Name","Regulatory_Feature_Group")])
+promoters <- grep("Prom",myann$Regulatory_Feature_Group)
 
@@
 
-## Whole gene analysis
+## Read in and inspect the data
 
@@
 dm <- read.table("dma3a.tsv.gz")
+dm <- dm[,4:9]
+dm <- merge(myann,dm,by=0)
+rownames(dm) <- dm[,1]
+dm[,1] = NULL
 
@@
 
+# trim down the input dataset
+dm <- dm[,c("UCSC_RefGene_Name","t")]
+
 # histogram of t values
@@
 
+```
+
+## Whole gene analysis
+
+```{r,wg}
 # set cores to used for parallel execution
@@
 
-# make a vector of unique gene names
+calc_sc <- function(dm) {
   gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
   gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
-
-# run the test (self contained)
+  gnl <- mclapply(gnl,unique,mc.cores=CORES)
+  dm$UCSC_RefGene_Name <- gnl
+  l <- mclapply(1:nrow(dm), function(i) {
+    a <- dm[i,]
+    len <- length(a[[1]][[1]])
+    tvals <- as.numeric(rep(a[2],len))
+    genes <- a[[1]][[1]]
+    data.frame(genes,tvals)
+  },mc.cores=CORES)
+  df <- do.call(rbind,l)
   gme_res <- mclapply( 1:length(gn), function(i) {
     g <- gn[i]
-  myrows <- which(lapply(gnl,function(x) {
-    length(which(x==g))
-  } ) > 0 )
-  tstats <- dm[myrows,"t"]
-  tstats <- tstats[order(tstats)]
+    tstats <- df[which(df$genes==g),"tvals"] 
     myn <- length(tstats)
@@
     wtselfcont <- wilcox.test(tstats)
-
     res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,"median"=mymedian,
-           "p-value(selfcontained)"=wtselfcont$p.value)
-} , mc.cores=CORES)
-
-# format the results
+      "p-value(sc)"=wtselfcont$p.value)
+  } , mc.cores=CORES )
   gme_res_df <- do.call(rbind, gme_res)
@@
   gme_res_df <- as.data.frame(tmp)
+  gme_res_df$sig <- -log10(gme_res_df[,4])
+  gme_res_df <- gme_res_df[order(-gme_res_df$sig),]
+  gme_res_df$`fdr(sc)` <- p.adjust(gme_res_df$`p-value(sc)`)
+  out <- list("df"=df,"gme_res_df"=gme_res_df)
+  return(out)
+}
 
-# results are sorted by the sum of -log10 of the p-values
-gme_res_df$sig <- -log10(gme_res_df[,4])
-
-gme_res_df <- gme_res_df[order(-gme_res_df$sig),]
-
-gme_res_df$`fdr(selfcontained)` <- p.adjust(gme_res_df$`p-value(selfcontained)`)
-
-gme_res_df_wholegene <- gme_res_df
-
-write.table(gme_res_df_wholegene,file="gmea_wholegene.tsv")
+tic()
+gme_res_wholegene <- calc_sc(dm)
+time2 <- toc() #38 44 41 40 44 41 40
+time2
+df <- gme_res_wholegene[[1]]
+res <- gme_res_wholegene[[2]]
+write.table(res,file="gmea_wholegene.tsv")
 
@@
 
-head(gme_res_df,50) %>%
+head(res,50) %>%
   kbl(caption = "Top significant genes with GMEA") %>%
@@
 # volcano selfcont
-sig <- subset(gme_res_df,`fdr(selfcontained)` < 0.05)
-plot(gme_res_df$median , -log10(gme_res_df$`p-value(selfcontained)`) ,
+sig <- subset(res,`fdr(sc)` < 0.05)
+plot(res$median , -log10(res$`p-value(sc)`) ,
   xlab="effect size (mean t-stat)", ylab="-log10(p-value)",
@@
 grid()
-points(sig$median , -log10(sig$`p-value(selfcontained)`) ,
+points(sig$median , -log10(sig$`p-value(sc)`) ,
   pch=19, cex=0.5, col="red")
@@
 # self contained
-gs <- head(rownames(gme_res_df[order(gme_res_df$`p-value(selfcontained)`),]),n)
-tstats <- mclapply(gs, function(g) {
-  myrows <- which(lapply(gnl,function(x) {
-    length(which(x==g))
-  } ) > 0 )
-  dm[myrows,"t"]
-},mc.cores=CORES)
+
+gs <- head(rownames(res),50)
+tstats <- lapply(gs, function(g) {
+  df[which(df$genes==g),"tvals"]
+})
 names(tstats) <- gs
+
 tstats <- tstats[order(unlist(lapply(tstats,median)))]
@@
 # effect size median
-sig <- subset(gme_res_df,`fdr(selfcontained)` < 0.05)
+sig <- subset(res,`fdr(sc)` < 0.05)
 gs <- head(rownames(sig[order(-abs(sig$median)),]),n)
@@
 if ( length(gs) >2 ) {
-  tstats <- mclapply(gs, function(g) {
-    myrows <- which(lapply(gnl,function(x) {
-      length(which(x==g))
-    } ) > 0 )
-    dm[myrows,"t"]
-  },mc.cores=CORES)
+  tstats <- lapply(gs, function(g) {
+    df[which(df$genes==g),"tvals"]
+  })
   names(tstats) <- gs
@@
 
-dmscore <- data.frame( gme_res_df$median * gme_res_df$sig)
+dmscore <- data.frame( res$median * res$sig)
 
-rownames(dmscore) <- rownames(gme_res_df)
+rownames(dmscore) <- rownames(res)
 
@@
 
-res <- mitch_calc(x=dmscore, genesets=genesets, priority="effect")
+mres <- mitch_calc(x=dmscore, genesets=genesets, priority="effect")
 
-head(res$enrichment_result,20) %>%
+head(mres$enrichment_result,20) %>%
   kbl(caption = "Top enriched gene sets with GMEA-Mitch") %>%
@@
 
-mitch_report(res,outfile="dma3a_mitch.html",overwrite=TRUE)
+mitch_report(mres,outfile="dma3a_mitch.html",overwrite=TRUE)
 
@@
 dm <- read.table("dma3a.tsv.gz")
+dm <- dm[,4:9]
+dm <- merge(myann,dm,by=0)
+rownames(dm) <- dm[,1]
+dm[,1] = NULL
 dm <- dm[grep("Promoter_Associated",dm$Regulatory_Feature_Group),]
-
-# testing
-#dm <- head(dm,20000)
-
-dm$adj.P.Val <- p.adjust(dm$P.Value)
 
@@
 
+```
+
+```{r,promo2}
+
 # set cores to used for parallel execution
-CORES= detectCores()
 
-# make a vector of unique gene names
-gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
-gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
+tic()
+gme_res_promoter <- calc_sc(dm)
+time2 <- toc() #15.7 20.2 20.3 17.5 17.9 17.7 20.6 
+time2
 
-# run the test
-gme_res <- mclapply( 1:length(gn), function(i) {
-  g <- gn[i]
-  myrows <- which(lapply(gnl,function(x) {
-    length(which(x==g))
-  } ) > 0 )
-  tstats <- dm[myrows,"t"]
-  tstats <- tstats[order(tstats)]
-  myn <- length(tstats)
-  mymean <- mean(tstats)
-  mymedian <- median(tstats)
-  wtselfcont <- wilcox.test(tstats)
-
-  res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,"median"=mymedian,
-           "p-value(selfcontained)"=wtselfcont$p.value)
-} , mc.cores=CORES)
-
-# format the results
-gme_res_df <- do.call(rbind, gme_res)
-rownames(gme_res_df) <- gme_res_df[,1]
-gme_res_df <- gme_res_df[,-1]
-tmp <- apply(gme_res_df,2,as.numeric)
-rownames(tmp) <- rownames(gme_res_df)
-gme_res_df <- as.data.frame(tmp)
-
-# results are sorted by the sum of -log10 of the p-values
-gme_res_df$sig <- -log10(gme_res_df[,4])
-
-gme_res_df <- gme_res_df[order(-gme_res_df$sig),]
-
-gme_res_df$`fdr(selfcontained)` <- p.adjust(gme_res_df$`p-value(selfcontained)`)
-
-gme_res_df_promoter <- gme_res_df
-
-write.table(gme_res_df_promoter ,file="gmea_promo.tsv")
+df <- gme_res_promoter[[1]]
+res <- gme_res_promoter[[2]] 
+write.table(res ,file="gmea_promo.tsv")
 
@@
 
-res <- mitch_calc(x=dmscore, genesets=genesets,priority="effect")
+mres <- mitch_calc(x=dmscore, genesets=genesets,priority="effect")
 
-head(res$enrichment_result,20) %>%
+head(mres$enrichment_result,20) %>%
   kbl(caption = "Top enriched gene sets with GMEA-Mitch (promoter only)") %>%
@@
 
-mitch_report(res,outfile="dma3a_mitch_promo.html",overwrite=TRUE)
-
-```
-
-## Multi-mitch
-
-```{r,dm1_mitch_promoter}
-
-dmscore_wholegene <- data.frame( gme_res_df_wholegene$median * gme_res_df_wholegene$sig )
-rownames(dmscore_wholegene) <- rownames(gme_res_df_wholegene)
-colnames(dmscore_wholegene) <- "wholegene"
-
-dmscore_promoter <- data.frame( gme_res_df_promoter$median * gme_res_df_promoter$sig )
-rownames(dmscore_promoter) <- rownames(gme_res_df_promoter)
-colnames(dmscore_promoter) <- "promoter"
-
-mm <- merge(dmscore_wholegene,dmscore_promoter,by=0)
-rownames(mm) <- mm[,1]
-mm[,1] = NULL
-
-genesets <- gmt_import("ReactomePathways.gmt")
-
-mmres <- mitch_calc(x=mm, genesets=genesets,priority="effect")
-
-head(mmres$enrichment_result,20) %>%
-  kbl(caption = "Top enriched gene sets with GMEA-Mitch (wholegene and promoter)") %>%
-  kable_paper("hover", full_width = F)
-
-#mitch_report(mmres,outfile="mitch_multi.html",overwrite=TRUE)
+mitch_report(mres,outfile="dma3a_mitch_promo.html",overwrite=TRUE)
 

### 2022-05-17 GSE158433/GSE158422.Rmd
added GSE158422 to enable comparison of gene expression and methylation in cancer

---
title: "GSE158422: DNA methylation in Lung squamous cell carcinoma"
author: "The GMEA team"
date: "`r Sys.Date()`"
output:
  html_document:
  toc: true
theme: cosmo
---

## Introduction

The purpose of this analysis is to apply different enrichment analysis approaches,
and see which approach is most parsimonious with gene expressoin differences.

We will specifically be testing the GMEA approach versus existing approaches,
namely GOseq and mitch.

This study (GSE158433) is a super-series contains matched RNA-seq (GSE158420) and 
Infinium EPIC array (GSE158422) data for matched control and cancer tissues.

In this script, we will be processing the EPIC DNA methylation data and performing the
statistical analysis in limma and saving the results for downstream enrichment analysis.

```{r,packages}

suppressPackageStartupMessages({
  library("plyr")
  library("R.utils")
  library("missMethyl")
  library("limma")
  library("DMRcate")
  library("DMRcatedata")
  library("topconfects")
  library("minfi")
  library("IlluminaHumanMethylation450kmanifest")
  library("RColorBrewer")
  library("IlluminaHumanMethylation450kanno.ilmn12.hg19")
  library("GEOquery")
  library("eulerr")
  library("plyr")
  library("gplots")
  library("reshape2")
  library("forestplot")
  library("beeswarm")
  library("RCircos")
  library("qqman")
  library("ENmix")
})

```

## Obtaining array annotations

```{r,annotation}

anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
myann <- data.frame(anno[,c("UCSC_RefGene_Name","Regulatory_Feature_Group","Islands_Name","Relation_to_Island")])
promoters <- grep("Prom",myann$Regulatory_Feature_Group)

```

## Obtain array data

GSE158422 is the accession number for the array data.

```{r,arraydata}

ARRAY_DATA="GSE158422_RAW.tar"
if(!file.exists(ARRAY_DATA)){
  download.file("ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE158nnn/GSE158422/suppl/GSE158422_RAW.tar",
    destfile=DEST) 
  untar(tarfile = ARRAY_DATA)
}
  SERIES_MATRIX="GSE158422_series_matrix.txt.gz"
download.file("ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE158nnn/GSE158422/matrix/GSE158422_series_matrix.txt.gz", 
  destfile=SERIES_MATRIX)
gse <- getGEO(filename=SERIES_MATRIX)
baseDir <- "."
sample_metadata <- pData(phenoData(gse))
targets <- sample_metadata
mybase <- unique(gsub("_Red.idat.gz" ,"",  gsub("_Grn.idat.gz", "" ,list.files(".",pattern = "GSM",recursive = TRUE))))
targets$Basename <- mybase
head(targets)
rgSet <- read.metharray.exp(targets = targets)
rgSet

```

## Filtering and normalisation

```{r,norm,fig.cap="Figure 1. Normalisation of bead-array data with SWAN."}

mSet <- preprocessRaw(rgSet)
mSetSw <- SWAN(mSet,verbose=FALSE)
par(mfrow=c(1,2), cex=0.8)
densityByProbeType(mSet[,1], main = "Raw")
densityByProbeType(mSetSw[,1], main = "SWAN")

```

```{r,filterprobes}

# include sex chromosomes
detP <- detectionP(rgSet)
keep <- rowSums(detP < 0.01) == ncol(rgSet)
mSetSw <- mSetSw[keep,]

# exclude SNP probes
mSetSw <- mapToGenome(mSetSw)
mSetSw_nosnp <- dropLociWithSnps(mSetSw)

# exclude sex chromosomes
xyprobes <- anno$Name[anno$chr %in% c("chrX","chrY")]
mSetFlt <- mSetSw[which(!rownames(mSetSw) %in% xyprobes),]

```

```{r,beta_m_vals}

# include sex chromosomes
meth <- getMeth(mSetSw)
unmeth <- getUnmeth(mSetSw)
Mval <- log2((meth + 100)/(unmeth + 100))
beta <- getBeta(mSetSw)
dim(Mval)

# exclude sex chromosomes
meth <- getMeth(mSetFlt)
unmeth <- getUnmeth(mSetFlt)
Mval_flt <- log2((meth + 100)/(unmeth + 100))
beta_flt <- getBeta(mSetFlt)
dim(Mval_flt)

```

# MDS analysis

```{r,scree,fig.cap="Figure 2. Screeplot shows contribution of top principal components to the overall variation in the experiment."}

# scree plot shows the amount of variation in a dataset that is accounted
# for by the first N principal components
myscree <- function(mx,n=10,main="") {
  pc <- princomp(mx)$sdev
  pcp <- pc/sum(pc)*100
  pcp <- pcp[1:10]
  barplot(pcp,cex.names = 1,las=2,ylim=c(0,60),
      ylab="percent (%) variance explained", main=main)
  text((0.5:length(pcp)*1.2),pcp,label=signif(pcp,3),pos=3,cex=0.8)
}

par(mfrow=c(2,1))  
myscree(Mval,main="incl sex chr")
myscree(Mval_flt,main="excl sex chr")

```

#MDS plot 1

```{r,mds1,fig.width = 8 ,fig.height = 8,fig.cap="Figure 3. MDS plot coloured by ART classification."}

colnames(targets) <- gsub(" ","_",colnames(targets)) 
colnames(targets) <- gsub(":","_",colnames(targets))
targets$sex<-factor(targets$`gender_ch1`)
targets$art<-factor(targets$`art_subtype`)
sample_groups <- factor(targets$art)
colour_palette=brewer.pal(n = length(levels(sample_groups)), name = "Paired")
colours <- colour_palette[as.integer(factor(targets$art))]
plot(1,axes = FALSE,xlab="",ylab="",main="MDS by ART type")
legend("center",legend=levels(sample_groups),pch=16,cex=1.2,col=colour_palette)
mydist <- plotMDS(Mval, labels=targets$Sample_Name,col=colours,main="sex chromosomes included")
mydist_flt <- plotMDS(Mval_flt, labels=targets$Sample_Name,col=colours,main="sex chromosomes excluded")

```

## save data object

```{r,save}

save.image("GSE158422.Rdata")

```

### 2022-05-18 GSE158433/GSE158422.Rmd
GSE progress

@@
 
-ARRAY_DATA="GSE158422_RAW.tar"
+ARRAY_DATA="GSE158433_RAW.tar"
 if(!file.exists(ARRAY_DATA)){
   download.file("ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE158nnn/GSE158422/suppl/GSE158422_RAW.tar",
-    destfile=DEST) 
+    destfile=ARRAY_DATA) 
   untar(tarfile = ARRAY_DATA)
 }
+
 SERIES_MATRIX="GSE158422_series_matrix.txt.gz"
@@
 
+# Infer sex
+
+```{r,infersex}
+
+xprobes <- anno$Name[anno$chr %in% "chrX"]
+yprobes <- anno$Name[anno$chr %in% "chrY"]
+
+xdat <- Mval[which( rownames(Mval) %in% xprobes ),]
+ydat <- Mval[which( rownames(Mval) %in% yprobes ),]
+gsm <- sapply( strsplit(colnames(Mval),"_" ) , "[[" , 1)
+plot(colMeans(xdat) , colMeans(ydat), cex=1)
+text(colMeans(xdat) , colMeans(ydat),labels=gsm )
+grid()
+abline(v=-0.1,col="red",lty=2)
+
+which( colMeans(xdat)  > -0.1 )
+
+sex_female <- as.numeric(colMeans(xdat)  > -0.1)
+
+```
+
 # MDS analysis
@@
 colnames(targets) <- gsub(":","_",colnames(targets))
-targets$sex<-factor(targets$`gender_ch1`)
-targets$art<-factor(targets$`art_subtype`)
-sample_groups <- factor(targets$art)
+targets$sex<-factor(sex_female)
+targets$case<-factor(targets$`characteristics_ch1`)
+sample_groups <- factor(targets$case)
 colour_palette=brewer.pal(n = length(levels(sample_groups)), name = "Paired")
-colours <- colour_palette[as.integer(factor(targets$art))]
-plot(1,axes = FALSE,xlab="",ylab="",main="MDS by ART type")
+colours <- colour_palette[as.integer(factor(targets$case))]
+plot(1,axes = FALSE,xlab="",ylab="",main="MDS by case/ctrl",cex=0)
 legend("center",legend=levels(sample_groups),pch=16,cex=1.2,col=colour_palette)

### 2022-05-18 meth_functions.R
meth_functions.R

@@ -1,0 +1,445 @@
+suppressPackageStartupMessages({
+  library("plyr")
+  library("R.utils")
+  library("missMethyl")
+  library("limma")
+  library("DMRcate")
+  library("DMRcatedata")
+  library("topconfects")
+  library("minfi")
+  library("IlluminaHumanMethylation450kmanifest")
+  library("RColorBrewer")
+  library("IlluminaHumanMethylation450kanno.ilmn12.hg19")
+  library("GEOquery")
+  library("eulerr")
+  library("plyr")
+  library("gplots")
+  library("reshape2")
+  library("forestplot")
+  library("beeswarm")
+  library("RCircos")
+  library("qqman")
+  library("ENmix")
+})
+
+
+# scree plot shows the amount of variation in a dataset that is accounted
+# for by the first N principal components
+myscree <- function(mx,n=10,main="") {
+  pc <- princomp(mx)$sdev
+  pcp <- pc/sum(pc)*100
+  pcp <- pcp[1:10]
+  barplot(pcp,cex.names = 1,las=2,ylim=c(0,60),
+      ylab="percent (%) variance explained", main=main)
+  text((0.5:length(pcp)*1.2),pcp,label=signif(pcp,3),pos=3,cex=0.8)
+}
+
+
+# Here is a function to make a volcano plot
+make_volcano <- function(dm,name,mx) {
+    sig <- subset(dm,adj.P.Val<0.05)
+    N_SIG=nrow(sig)
+    N_UP=nrow(subset(sig,logFC>0))
+    N_DN=nrow(subset(sig,logFC<0))
+    HEADER=paste(N_SIG,"@5%FDR,", N_UP, "up", N_DN, "dn")
+    plot(dm$logFC,-log10(dm$P.Val),cex=0.5,pch=19,col="darkgray",
+        main=name, xlab="log FC", ylab="-log10 pval")
+    mtext(HEADER)
+    grid()
+    points(sig$logFC,-log10(sig$P.Val),cex=0.5,pch=19,col="red")
+}
+
+# Metal volcano
+volcano_metal <- function(metal,name) {
+  sig <- subset(metal,FDR<0.05)
+  N_SIG=nrow(sig)
+  N_UP=nrow(subset(sig,Zscore>0))
+  N_DN=nrow(subset(sig,Zscore<0))
+  HEADER=paste(N_SIG,"@5%FDR,", N_UP, "up", N_DN, "dn")
[…]

### 2022-05-18 GSE158433/GSE158422.Rmd
progress on GSE158422

@@
 })
+
+source("../meth_functions.R")
 
@@
 legend("center",legend=levels(sample_groups),pch=16,cex=1.2,col=colour_palette)
-mydist <- plotMDS(Mval, labels=targets$Sample_Name,col=colours,main="sex chromosomes included")
-mydist_flt <- plotMDS(Mval_flt, labels=targets$Sample_Name,col=colours,main="sex chromosomes excluded")
+mydist <- plotMDS(Mval, labels=gsm,col=colours,main="sex chromosomes included")
+mydist_flt <- plotMDS(Mval_flt, labels=gsm,col=colours,main="sex chromosomes excluded")
+
+```
+
+## Differential analysis
+
+Paired normal versus cancer tissue.
+
+```{r,dm}
+
+samplesheet<-targets
+groups <- factor(samplesheet$case,levels=c("tumor: normal","tumor: tumor"))
+sex <- factor(samplesheet$sex,levels=c("0","1"))
+Mvals<-Mval
+betas<-beta
+dim(Mvals)
+dim(betas)
+
+GSE158422 <- dm_analysis(samplesheet=samplesheet,
+  sex=sex,groups=groups,mx=Mvals,name="GSE158422",
+  myann=myann ,beta= betas)
+
+head(GSE158422$dma,10)
+head(GSE158422$dmr,10)
 

### 2022-05-18 GSE158433/GSE158422.Rmd
progress on GSE158422.Rmd

@@
 SERIES_MATRIX="GSE158422_series_matrix.txt.gz"
+if(!file.exists(SERIES_MATRIX)){
 download.file("ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE158nnn/GSE158422/matrix/GSE158422_series_matrix.txt.gz", 
   destfile=SERIES_MATRIX)
+}
 gse <- getGEO(filename=SERIES_MATRIX)
@@
 
+saveRDS(GSE158422,file="GSE158422.rds")
+write.table(GSE158422,file="GSE158422_limma.tsv",sep="\t",quote=FALSE)
+
 ```

### 2022-05-18 GSE158433/GSE158433_gmea.Rmd
gmea for GSE158433

---
title: "GSE158433: GMEA of DNA methylation and RNA-seq differences in lung squamous cell carcinoma"
author: "The GMEA team"
date: "`r Sys.Date()`"
output:
  html_document:
  toc: true
theme: cosmo
---

## Introduction

The purpose of this analysis is to apply different enrichment analysis approaches,
and see which approach is most parsimonious with gene expressoin differences.

We will specifically be testing the GMEA approach versus existing approaches,
namely GOseq and mitch.

This study (GSE158433) is a super-series contains matched RNA-seq (GSE158420) and 
Infinium EPIC array (GSE158422) data for matched control and cancer tissues.

In this script, we will be processing the EPIC DNA methylation data and performing the
statistical analysis in limma and saving the results for downstream enrichment analysis.

```{r,packages}

suppressPackageStartupMessages({
  library("plyr")
  library("R.utils")
  library("missMethyl")
  library("limma")
  library("DMRcate")
  library("DMRcatedata")
  library("topconfects")
  library("minfi")
  library("IlluminaHumanMethylation450kmanifest")
  library("RColorBrewer")
  library("IlluminaHumanMethylation450kanno.ilmn12.hg19")
  library("GEOquery")
  library("eulerr")
  library("plyr")
  library("gplots")
  library("reshape2")
  library("forestplot")
  library("beeswarm")
  library("RCircos")
  library("qqman")
  library("ENmix")
  library("tictoc")
  library("mitch")
})

source("../meth_functions.R")
CORES=detectCores()/2

```

## Obtaining array annotations

```{r,annotation}

anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
myann <- data.frame(anno[,c("UCSC_RefGene_Name","Regulatory_Feature_Group","Islands_Name","Relation_to_Island")])
promoters <- grep("Prom",myann$Regulatory_Feature_Group)

```

## Load array result

GSE158422 is the accession number for the array data.

```{r,arraydataload}

#GSE158422 <- readRDS("GSE158422.rds")
dm <- read.table("GSE158422_limma.tsv",header=TRUE,sep="\t")

head(dm,50) %>%
  kbl(caption = "Top significant probes with limma") %>%
  kable_paper("hover", full_width = F)

rownames(dm) <- dm[,1]
dm <- dm[,c("UCSC_RefGene_Name","t")]

```

## GMEA whole gene

```{r,gmea}

hist(dm$t)
tic() ; res <- calc_sc(dm) ; toc() #32 cores 95.5
res2 <- res[[2]]

head(res2,20) %>%
  kbl(caption = "Top significant genes with GMEA") %>%
  kable_paper("hover", full_width = F)

sig <- subset(res2,`fdr(sc)`<0.05)

es <- sig[order(-abs(sig$median)),]

head(es,20) %>%
  kbl(caption = "Top effect size probes with GMEA") %>%
  kable_paper("hover", full_width = F)

gmea_volc(res2)
gmea_barplot(res2)
res2_wg <- res2

```

## Mitch whole gene

```{r,mitch_wg}

genesets <- gmt_import("../ReactomePathways.gmt")

metric <- res2$sig * res2$median
metric <- as.data.frame(metric)
rownames(metric) <- rownames(res2)

mres <- mitch_calc(x=metric, genesets=genesets)

head(mres$enrichment_result,20) %>%
  kbl(caption = "Top differential pathways with GMEA+mitch") %>%
  kable_paper("hover", full_width = F)

```

## GMEA promoter

```{r,gmea}

dm <- read.table("GSE158422_limma.tsv",header=TRUE,sep="\t")
dm <- dm[grep("Promot",dm$Regulatory_Feature_Group),]
dm <- dm[,c("UCSC_RefGene_Name","t")]

hist(dm$t)
tic() ; res <- calc_sc(dm) ; toc() #32 cores 17.7
res2 <- res[[2]]

head(res2,20) %>%
  kbl(caption = "Top significant promoters with GMEA") %>%
  kable_paper("hover", full_width = F)

sig <- subset(res2,`fdr(sc)`<0.05)

es <- sig[order(-abs(sig$median)),]

head(es,20) %>%
  kbl(caption = "Top effect size probes with GMEA") %>%
  kable_paper("hover", full_width = F)

gmea_volc(res2)
gmea_barplot(res2)
res2_p <- res2

```

## Mitch promoter

```{r,mitchpromoter}

genesets <- gmt_import("../ReactomePathways.gmt")

metric <- res2$sig * res2$median
metric <- as.data.frame(metric)
rownames(metric) <- rownames(res2)

mres <- mitch_calc(x=metric, genesets=genesets)

head(mres$enrichment_result,20) %>%
  kbl(caption = "Top differential pathways with GMEA+mitch") %>%
  kable_paper("hover", full_width = F)

```

## save data object

```{r,save}

save.image("GSE158422.Rdata")

```

### 2022-05-18 GSE158433/GSE158433_gmea.Rmd
bugfix

@@
   library("mitch")
+  library("kableExtra")
 })
@@
 
-```{r,gmea}
+```{r,gmea_wg}
 
@@
 
-```{r,gmea}
+```{r,gmea_promoter}
 

### 2022-05-18 GSE158433/GSE158433_gmea.Rmd
figure size

@@
   html_document:
-  toc: true
+    toc: yes
+    toc_float: true
+    code_folding: hide
+    fig_width: 7
+    fig_height: 7
 theme: cosmo

### 2022-05-24 GSE158433/GSE158420.Rmd
integrative analysis begun

---
title: "GSE158420: RNA expression in Lung squamous cell carcinoma"
author: "The GMEA team"
date: "`r Sys.Date()`"
output:
  html_document:
  toc: true
theme: cosmo
---

## Introduction

The purpose of this analysis is to apply different enrichment analysis approaches,
and see which approach is most parsimonious with gene expression differences.

This study (GSE158433) is a super-series contains matched RNA-seq (GSE158420) and 
Infinium EPIC array (GSE158422) data for matched control and cancer tissues.

In this script, we will be analysing the gene expression profiles, followed by mitch
enrichment analysis.

```{r,packages}

suppressPackageStartupMessages({
  library("DESeq2")
  library("mitch")
  library("gplots")
  library("kableExtra")
})

```

## Download expression data

The data is available from the following address.
There are matched cancer and normal tissues.

https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE158420

The count matrix contains excel errors which need to be fixed manually or with sed.

```{r,datadl1}

URL="https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE158420&format=file&file=GSE158420%5Fcounts%2Etxt%2Egz"
FILE="GSE158420_counts.txt.gz"
FILE2="GSE158420_counts.txt"
download.file(URL,destfile=FILE)

x <- read.table(FILE2)

head(x)

```

## Samplesheet

The sample names will be converted to samplesheet.

```{r,samplesheet1}

ss <- data.frame(colnames(x))
names(ss) <- "label"
ss$tumor <- factor(grepl("T",ss$label))
ss$patient <- factor(gsub("T","",gsub("N","",ss$label)))
rownames(ss) <- ss$label

```

## Filtering and normalisation

```{r,filt}

dim(x)
xx <- x[which(rowMeans(x)>=10),]
dim(xx)

rpm <- xx/colSums(xx) * 1000000

```

## MDS plot

```{r,mds}

colours <- gsub("FALSE","lightblue",gsub("TRUE","pink",as.character(ss$tumor)))

labels <- gsub("N","",gsub("T","",gsub("X","",colnames(xx))))

plot(cmdscale(dist(t(xx))), xlab="Coordinate 1", ylab="Coordinate 2", type = "p",pch=19,col=colours,cex=2)
text(cmdscale(dist(t(xx))), labels=labels )

```

## DESEQ2

```{r,deseq2}

dds <- DESeqDataSetFromMatrix(countData = xx , colData = ss, design = ~ patient + tumor )
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds, blind=FALSE)
zz<-cbind(as.data.frame(z),assay(vsd))
dge<-as.data.frame(zz[order(zz$pvalue),])
dge[1:10,1:6]

sig <- subset(dge,padj<0.05)
dge_up <- rownames(subset(sig,log2FoldChange>0))
dge_dn <- rownames(subset(sig,log2FoldChange<0))
length(dge_up)
length(dge_dn)

write.table(dge,file="GSE158420_dge.tsv",sep="\t",quote=FALSE)

```

## Mitch

```{r,mitch}

download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", destfile="ReactomePathways.gmt.zip")
unzip("ReactomePathways.gmt.zip")
genesets <- gmt_import("ReactomePathways.gmt")

m <- mitch_import(dge,DEtype="deseq2")

res <- mitch_calc(m, genesets, priority="effect")

head(res$enrichment_result,20) %>% 
  kbl(caption = "top differentially regulated pathways") %>%
  kable_paper("hover", full_width = F)

```

## Session information

```{r,sessioninfo}

sessionInfo()

```

### 2022-05-24 GSE158433/GSE158433_gmea.Rmd
integrative analysis begun

@@
 gmea_volc(res2)
-gmea_barplot(res2)
+gmea_boxplot(res)
 res2_wg <- res2
@@
 
+Methylation only.
+
 ```{r,mitch_wg}
@@
 
-metric <- res2$sig * res2$median
-metric <- as.data.frame(metric)
-rownames(metric) <- rownames(res2)
+meth <- res2$sig * res2$median
+meth <- as.data.frame(meth)
+rownames(meth) <- rownames(res2)
 
-mres <- mitch_calc(x=metric, genesets=genesets)
+mres <- mitch_calc(x=meth, genesets=genesets)
 
@@
   kable_paper("hover", full_width = F)
+
+```
+
+Methylation and gene expression.
+
+```{r,integrate_wg}
+
+dge <- read.table("GSE158420_dge.tsv")
+dgem <- mitch_import(dge,DEtype="deseq2")
+
+int <- merge(meth,dgem,by=0)
+rownames(int) <- int$Row.names
+int[,1]=NULL
+names(int) <- c("meth","rna")
+
+int_res <- mitch_calc(int,genesets=genesets)
+
+int_res_tbl <- int_res$enrichment_result
+
+head(int_res_tbl, 20) %>%
+  kbl(caption = "Integrative results with mitch") %>%
+  kable_paper("hover", full_width = F)
+
+plot(int_res_tbl$s.meth,int_res_tbl$s.rna,ylab="RNA",xlab="meth")
+
+abline(h=0,lty=2,lwd=2,col="red")
+abline(v=0,lty=2,lwd=2,col="red")
+
+uu <- length(which(int_res_tbl$s.meth>0 & int_res_tbl$s.rna>0))
+ud <- length(which(int_res_tbl$s.meth>0 & int_res_tbl$s.rna<0))
+dd <- length(which(int_res_tbl$s.meth<0 & int_res_tbl$s.rna<0))
+du <- length(which(int_res_tbl$s.meth<0 & int_res_tbl$s.rna>0))
+
+uu
+ud
+dd
+du
+
+uu + dd
+ud + du
+
+mitch_report(int_res,outfile="int_wholegene.html",overwrite=TRUE)
 
@@
 gmea_volc(res2)
-gmea_barplot(res2)
+gmea_boxplot(res)
 res2_p <- res2
@@
 
-```{r,mitchpromoter}
+Methylation only
+
+```{r,mitchpromoter1}
 
@@
 
-metric <- res2$sig * res2$median
-metric <- as.data.frame(metric)
-rownames(metric) <- rownames(res2)
+pmeth <- res2$sig * res2$median
+pmeth <- as.data.frame(pmeth)
+rownames(pmeth) <- rownames(res2)
 
-mres <- mitch_calc(x=metric, genesets=genesets)
+mres <- mitch_calc(x=pmeth, genesets=genesets)
 
@@
 
-## save data object
+Methylation and gene expression.
 
-```{r,save}
+```{r,integrate_promoter}
 
-save.image("GSE158422.Rdata")
+#dge <- read.table("GSE158420_dge.tsv")
+#dgem <- mitch_import(dge,DEtype="deseq2")
+
+int <- merge(pmeth,dgem,by=0)
+rownames(int) <- int$Row.names
+int[,1]=NULL
+names(int) <- c("meth","rna")
+
+int_res <- mitch_calc(int,genesets=genesets)
+int_res_tbl <- int_res$enrichment_result
+
+head(int_res_tbl,20) %>%
+  kbl(caption = "Integrative results with mitch") %>%
+  kable_paper("hover", full_width = F)
+
+int_res_tbl <- int_res_tbl[order(-abs(int_res_tbl$s.dist)),]
+
+head(int_res_tbl,20) %>%
+  kbl(caption = "Integrative results with mitch") %>%
+  kable_paper("hover", full_width = F)
+
+plot(int_res_tbl$s.meth,int_res_tbl$s.rna,ylab="RNA",xlab="meth")
+
+abline(h=0,lty=2,lwd=2,col="red")
+abline(v=0,lty=2,lwd=2,col="red")
+
+uu <- length(which(int_res_tbl$s.meth>0 & int_res_tbl$s.rna>0))
+ud <- length(which(int_res_tbl$s.meth>0 & int_res_tbl$s.rna<0))
+dd <- length(which(int_res_tbl$s.meth<0 & int_res_tbl$s.rna<0))
+du <- length(which(int_res_tbl$s.meth<0 & int_res_tbl$s.rna>0))
+
+uu
+ud
+dd
+du
+
+uu + dd
+ud + du
+
+mitch_report(int_res,outfile="int_promoter.html",overwrite=TRUE)
 
 ```
+
+## Session Information
+
+```{r,session}
+
+sessionInfo()
+
+```

### 2022-05-24 meth_functions.R

@@ -443,3 +443,84 @@ run_mitch_1d <- function(dma,name) {
     type = c("output", "message"), split = FALSE)
   return(res$enrichment_result)
 }
+
+
+calc_sc <- function(dm) {
+  gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
+  gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
+  gnl <- mclapply(gnl,unique,mc.cores=CORES)
+  dm$UCSC_RefGene_Name <- gnl
+  l <- mclapply(1:nrow(dm), function(i) {
+    a <- dm[i,]
+    len <- length(a[[1]][[1]])
+    tvals <- as.numeric(rep(a[2],len))
+    genes <- a[[1]][[1]]
+    data.frame(genes,tvals)
+  },mc.cores=CORES)
+  df <- do.call(rbind,l)
+  gme_res <- mclapply( 1:length(gn), function(i) {
+    g <- gn[i]
+    tstats <- df[which(df$genes==g),"tvals"]
+    myn <- length(tstats)
+    mymean <- mean(tstats)
+    mymedian <- median(tstats)
+    wtselfcont <- wilcox.test(tstats)
+    res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,"median"=mymedian,
+      "p-value(sc)"=wtselfcont$p.value)
+  } , mc.cores=CORES )
+  gme_res_df <- do.call(rbind, gme_res)
+  rownames(gme_res_df) <- gme_res_df[,1]
+  gme_res_df <- gme_res_df[,-1]
+  tmp <- apply(gme_res_df,2,as.numeric)
+  rownames(tmp) <- rownames(gme_res_df)
+  gme_res_df <- as.data.frame(tmp)
+  gme_res_df$sig <- -log10(gme_res_df[,4])
+  gme_res_df <- gme_res_df[order(-gme_res_df$sig),]
+  gme_res_df$`fdr(sc)` <- p.adjust(gme_res_df$`p-value(sc)`)
+  out <- list("df"=df,"gme_res_df"=gme_res_df)
+  return(out)
+}
+
+gmea_volc <- function(res) {
+  sig <- subset(res,`fdr(sc)` < 0.05)
+  plot(res$median , -log10(res$`p-value(sc)`) ,
+    xlab="effect size (mean t-stat)", ylab="-log10(p-value)",
+    pch=19, cex=0.5, col="gray",main="self contained test")
+  grid()
[…]

### 2022-05-24 GSE158433/GSE158433_gmea.Rmd
add integrative analysis
priority

@@
 
-mres <- mitch_calc(x=meth, genesets=genesets)
+mres <- mitch_calc(x=meth, genesets=genesets,priority="effect")
 
@@
 
-int_res <- mitch_calc(int,genesets=genesets)
+plot(int,xlim=c(-100,100))
+abline(h=0,lty=2,lwd=2,col="red")
+abline(v=0,lty=2,lwd=2,col="red")
+
+uu <- length(which(int$meth>0 & int$rna>0))
+ud <- length(which(int$meth>0 & int$rna<0))
+dd <- length(which(int$meth<0 & int$rna<0))
+du <- length(which(int$meth<0 & int$rna>0))
+
+uu
+ud
+dd
+du
+
+uu + dd
+ud + du
+
+int_res <- mitch_calc(int,genesets=genesets,priority="effect")
 
@@
 
-mres <- mitch_calc(x=pmeth, genesets=genesets)
+mres <- mitch_calc(x=pmeth, genesets=genesets,priority="effect")
 
@@
 
-int_res <- mitch_calc(int,genesets=genesets)
+plot(int)
+abline(h=0,lty=2,lwd=2,col="red")
+abline(v=0,lty=2,lwd=2,col="red")
+
+uu <- length(which(int$meth>0 & int$rna>0))
+ud <- length(which(int$meth>0 & int$rna<0))
+dd <- length(which(int$meth<0 & int$rna<0))
+du <- length(which(int$meth<0 & int$rna>0))
+
+uu
+ud
+dd
+du
+
+uu + dd
+ud + du
+
+int_res <- mitch_calc(int,genesets=genesets,priority="effect")
+
 int_res_tbl <- int_res$enrichment_result

### 2022-10-06 GSE158433/GSE158422_gmea.Rmd
reorganising

---
title: "GSE158433: GMEA of DNA methylation differences in lung squamous cell carcinoma"
author: "The GMEA team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
---

## Introduction

The purpose of this analysis is to explore different enrichment approaches to the
cancer methylation data: Infinium EPIC array (GSE158422) data for matched control and cancer tissues.

We will specifically be testing the GMEA approach versus existing approaches:

* 1-way Wilcox text

* 1-way t-test

* FGSEA test

* roast test

After each of these tests, downstream GSEA can be conducted.

```{r,packages}

suppressPackageStartupMessages({
  library("plyr")
  library("R.utils")
  library("missMethyl")
  library("limma")
  library("DMRcate")
  library("DMRcatedata")
  library("topconfects")
  library("minfi")
  library("IlluminaHumanMethylation450kmanifest")
  library("RColorBrewer")
  library("IlluminaHumanMethylation450kanno.ilmn12.hg19")
  library("GEOquery")
  library("eulerr")
  library("plyr")
  library("gplots")
  library("reshape2")
  library("forestplot")
  library("beeswarm")
  library("RCircos")
  library("qqman")
  library("ENmix")
  library("tictoc")
  library("mitch")
  library("kableExtra")
  library("fgsea")
})

source("../meth_functions.R")
CORES=8

```

## Obtaining array annotations

```{r,annotation}

anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
myann <- data.frame(anno[,c("UCSC_RefGene_Name",
  "Regulatory_Feature_Group","Islands_Name","Relation_to_Island")])
promoters <- grep("Prom",myann$Regulatory_Feature_Group)

```

Gene probe sets

```{r,sets}

gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
names(gp2) <- rownames(gp)
sets <- split(rep(names(gp2), lengths(gp2)), unlist(gp2))

hist(unlist(lapply(sets,length)),xlim=c(0,200),breaks=500,xlab="set size",main="probes per gene")

```

## Load array result

GSE158422 is the accession number for the array data.

```{r,arraydataload}

#GSE158422 <- readRDS("GSE158422.rds")
dm <- read.table("GSE158422_limma.tsv",header=TRUE,sep="\t")

head(dm,50) %>%
  kbl(caption = "Top significant probes with limma") %>%
  kable_paper("hover", full_width = F)

rownames(dm) <- dm[,1]
dm <- dm[,c("UCSC_RefGene_Name","t")]

```

## GMEA whole gene - wilcox test

```{r,gmea_wg}

hist(dm$t)
tic() ; res <- calc_sc(dm) ; toc() #32 cores 95.5
res2 <- res[[2]]

head(res2,20) %>%
  kbl(caption = "Top significant genes with GMEA") %>%
  kable_paper("hover", full_width = F)

sig <- subset(res2,`fdr(sc)`<0.05)

es <- sig[order(-abs(sig$median)),]

head(es,20) %>%
  kbl(caption = "Top effect size probes with GMEA") %>%
  kable_paper("hover", full_width = F)

gmea_volc(res2)
gmea_boxplot(res)
res2_wg <- res2

```

Investigate how probe number influences significance.

```{r,probebias}

plot(res2$nprobes,res2$sig,log="x",ylim=c(0,50),pch=19,cex=0.6)
points(sig$nprobes,sig$sig,col="red",pch=19,cex=0.62)
MIN = min(sig$nprobes)
LEFT = nrow(subset(res2,nprobes<MIN))
RIGHT = nrow(subset(res2,nprobes>MIN))
SIG = nrow(sig)
TOT = nrow(res2)
HEADER <- paste(TOT, "genes in total.", SIG, "with FDR<0.05.",
  RIGHT, "well covered and", LEFT, "poorly covered")
mtext(HEADER)
abline(v=MIN,lty=2)

```

## GMEA whole gene - t.test

```{r,gmea_t}

dm <- read.table("GSE158422_limma.tsv",header=TRUE,sep="\t")
rownames(dm) <- dm[,1]
dm <- dm[,c("UCSC_RefGene_Name","t")]

hist(dm$t)
tic() ; res <- calc_sc2(dm) ; toc() #
res2 <- res[[2]]

head(res2,20) %>%
  kbl(caption = "Top significant genes with GMEA-t") %>%
  kable_paper("hover", full_width = F)

sig <- subset(res2,`fdr(sc)`<0.05)

es <- sig[order(-abs(sig$median)),]

head(es,20) %>%
  kbl(caption = "Top effect size probes with GMEA") %>%
  kable_paper("hover", full_width = F)

gmea_volc(res2)
gmea_boxplot(res)
res2_t <- res2

plot(res2$nprobes,res2$sig,log="x",ylim=c(0,50),pch=19,cex=0.6)
points(sig$nprobes,sig$sig,col="red",pch=19,cex=0.62)
MIN = min(sig$nprobes)
LEFT = nrow(subset(res2,nprobes<MIN))
RIGHT = nrow(subset(res2,nprobes>MIN))
SIG = nrow(sig)
TOT = nrow(res2)
HEADER <- paste(TOT, "genes in total.", SIG, "with FDR<0.05.",
  RIGHT, "well covered and", LEFT, "poorly covered")
mtext(HEADER)
abline(v=MIN,lty=2)

```

## GMEA using FGSEA test

```{r,gmea_fgsea}

hist(dm$t)
head(dm)
tstats <- dm$t
names(tstats) <- rownames(dm)

tic()
fgseaRes <- fgsea(pathways = sets,
                  stats    = tstats,
                  minSize  = 5,
                  nproc = 16)
toc()

fgseaRes <- as.data.frame(fgseaRes)[,1:7]
fgseaRes <- fgseaRes[order(fgseaRes$pval),]

head(fgseaRes,20) %>%
  kbl(caption = "Top significant genes with GMEA-FGSEA") %>%
  kable_paper("hover", full_width = F)

sig <- subset(fgseaRes,padj<0.05)
SIG = nrow(sig)

plot(fgseaRes$ES,-log10(fgseaRes$pval))
points(sig$ES,-log10(sig$pval),col="red")

plot(fgseaRes$size,-log10(fgseaRes$pval),log="x")
points(sig$size,-log10(sig$pval),col="red")

gmea_volc(res2)
gmea_boxplot(res)
res2_wg <- res2

```

## GMEA using ROAST test

TODO

## Mitch whole gene

Methylation only.

```{r,mitch_wg}

genesets <- gmt_import("../ReactomePathways.gmt")

meth <- res2$sig * res2$median
meth <- as.data.frame(meth)
rownames(meth) <- rownames(res2)

mres <- mitch_calc(x=meth, genesets=genesets,priority="effect")

head(mres$enrichment_result,20) %>%
  kbl(caption = "Top differential pathways with GMEA+mitch") %>%
  kable_paper("hover", full_width = F)

```

## Session Information

```{r,session}

sessionInfo()

```

### 2022-10-06 meth_functions.R
func update

@@ -453,7 +453,7 @@ calc_sc <- function(dm) {
   l <- mclapply(1:nrow(dm), function(i) {
     a <- dm[i,]
     len <- length(a[[1]][[1]])
-    tvals <- as.numeric(rep(a[2],len))
+    tvals <- as.numeric(rep(a["t"],len))
     genes <- a[[1]][[1]]
     data.frame(genes,tvals)
   },mc.cores=CORES)
@@ -480,6 +480,53 @@ calc_sc <- function(dm) {
   out <- list("df"=df,"gme_res_df"=gme_res_df)
   return(out)
 }
+
+
+# use a t-test for calculating significance
+calc_sc2 <- function(dm) {
+  gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
+  gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
+  gnl <- mclapply(gnl,unique,mc.cores=CORES)
+  dm$UCSC_RefGene_Name <- gnl
+  l <- mclapply(1:nrow(dm), function(i) {
+    a <- dm[i,]
+    len <- length(a[[1]][[1]])
+    tvals <- as.numeric(rep(a["t"],len))
+    genes <- a[[1]][[1]]
+    data.frame(genes,tvals)
+  },mc.cores=CORES)
+  df <- do.call(rbind,l)
+  keep <- names(which(table(df$genes)>1))
+  df <- df[df$genes %in% keep,]
+  gn <- unique(df$genes)
+
+  gme_res <- lapply( 1:length(gn), function(i) {
+    g <- gn[i]
+    tstats <- df[which(df$genes==g),"tvals"]
+    myn <- length(tstats)
+    mymean <- mean(tstats)
+    mymedian <- median(tstats)
+    tselfcont <- t.test(tstats)
+    res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,"median"=mymedian,
+      "p-value(sc)"=tselfcont$p.value)
+  } )
+
+  gme_res_df <- do.call(rbind, gme_res)
+  rownames(gme_res_df) <- gme_res_df[,1]
+  gme_res_df <- gme_res_df[,-1]
+  tmp <- apply(gme_res_df,2,as.numeric)
+  rownames(tmp) <- rownames(gme_res_df)
+  gme_res_df <- as.data.frame(tmp)
[…]

### 2022-10-10 GSE158433/GSE158422.Rmd
added mroast and fry

--- 2022-05-18 GSE158433/GSE158422.Rmd
@@
 
-ARRAY_DATA="GSE158433_RAW.tar"
+ARRAY_DATA="GSE158422_RAW.tar"
 if(!file.exists(ARRAY_DATA)){
-  download.file("ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE158nnn/GSE158422/suppl/GSE158422_RAW.tar",
+  download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE158422&format=file",
     destfile=ARRAY_DATA)
@@
 SERIES_MATRIX="GSE158422_series_matrix.txt.gz"
-download.file("ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE158nnn/GSE158422/matrix/GSE158422_series_matrix.txt.gz", 
+if(!file.exists(SERIES_MATRIX)){
+download.file("https://ftp.ncbi.nlm.nih.gov/geo/series/GSE158nnn/GSE158422/matrix/GSE158422_series_matrix.txt.gz",
   destfile=SERIES_MATRIX)
+}
 gse <- getGEO(filename=SERIES_MATRIX)
@@
 
+saveRDS(GSE158422,file="GSE158422.rds")
+write.table(GSE158422$dma,file="GSE158422_limma.tsv",sep="\t",quote=FALSE)
+
 ```

### 2022-10-10 GSE158433/GSE158422_fry.Rmd
added mroast and fry

---
title: "GSE158422: DNA methylation in Lung squamous cell carcinoma - GMEA using fry()"
author: "The GMEA team"
date: "`r Sys.Date()`"
output:
  html_document:
  toc: true
theme: cosmo
---

## Introduction

Here I will use fry to perform GMEA analysis.

```{r,packages}

suppressPackageStartupMessages({
  library("plyr")
  library("R.utils")
  library("missMethyl")
  library("limma")
  library("DMRcate")
  library("DMRcatedata")
  library("topconfects")
  library("minfi")
  library("IlluminaHumanMethylation450kmanifest")
  library("RColorBrewer")
  library("IlluminaHumanMethylation450kanno.ilmn12.hg19")
  library("GEOquery")
  library("eulerr")
  library("plyr")
  library("gplots")
  library("reshape2")
  library("forestplot")
  library("beeswarm")
  library("RCircos")
  library("qqman")
  library("ENmix")

  library("tictoc")
  library("kableExtra")

})

source("../meth_functions.R")

```

## Load data

```{r,annotation}

load("GSE158422.Rdata")

anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
myann <- data.frame(anno[,c("UCSC_RefGene_Name","Regulatory_Feature_Group","Islands_Name","Relation_to_Island")])
promoters <- grep("Prom",myann$Regulatory_Feature_Group)
promoters <- myann[promoters,]

```

Curate probe-sets.

Firstly for whole gene.

```{r,probesets_wholegene}

gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
names(gp2) <- rownames(gp)
sets <- split(rep(names(gp2), lengths(gp2)), unlist(gp2))

summary(unlist(lapply(sets,length)))

```

Now for promoters only.

```{r,probesets_promoters}

gp <- promoters[,"UCSC_RefGene_Name",drop=FALSE]
gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
names(gp2) <- rownames(gp)
psets <- split(rep(names(gp2), lengths(gp2)), unlist(gp2))

summary(unlist(lapply(psets,length)))

```

## Differential analysis

Paired normal versus cancer tissue.

```{r,dm}

samplesheet<-targets
groups <- factor(samplesheet$case,levels=c("tumor: normal","tumor: tumor"))
sex <- factor(samplesheet$sex,levels=c("0","1"))
Mvals<-Mval
betas<-beta
dim(Mvals)
dim(betas)

GSE158422 <- dm_analysis(samplesheet=samplesheet,
  sex=sex,groups=groups,mx=Mvals,name="GSE158422",
  myann=myann ,beta= betas)

mx=Mvals
name="GSE158422"
beta=betas
design <- model.matrix(~ sex + groups)
mxs <- mx[,which( colnames(mx) %in% samplesheet$Basename )]
fit.reduced <- lmFit(mxs,design)
fit.reduced <- eBayes(fit.reduced)
summary(decideTests(fit.reduced))

```

## set up mroast

Roast is a self contained test involving sample permulations.

This is beneficial for two reasons.

1. Sample permutations will prevent against false positives caused by strong
probe-probe correlation structures.

2. Competitive tests are not suited to experiments where there are unequal up and down
regulaiton, which is common for epigenetics studies.

Parallel analysis of GMEA with whole gene and promoters only.

mroast with 1000 whole gene probe sets, 2000 rotations: 42 s

mroast with 1000 gene promoter probe sets, 2000 rotations: 41 s

```{r,mroast1}

tic()
mro_res1 <- mroast(y=mxs, index = head(sets,1000), design = design, contrast = ncol(design), nrot = 2000 )
toc()

head(mro_res1,20) %>%
  kbl(caption="mroast test with 1000 whole gene probe sets") %>%
  kable_styling(full_width=FALSE)

tic()
mro_res2 <- mroast(y=mxs, index = head(psets,1000), design = design, contrast = ncol(design), nrot = 2000 )
toc()

head(mro_res2,20) %>%
  kbl(caption="mroast test with 1000 promoter only probe sets") %>%
  kable_styling(full_width=FALSE)

```

## Set up fry test

Fry is good because the p-values are analytically determined, and will have high precision.

Small test whole gene: 28 secs.
Small test gene promoter: 27 secs.

```{r,smallfry1}

tic()
fry_res1 <- fry(y=mxs, index = head(sets,1000), design = design, contrast = ncol(design) )
toc()

head(fry_res1,20) %>%
  kbl(caption="fry test with 1000 whole gene probe sets") %>%
  kable_styling(full_width=FALSE)

tic()
fry_res2 <- fry(y=mxs, index = head(psets,1000), design = design, contrast = ncol(design) )
toc()

head(fry_res2,20) %>%
  kbl(caption="fry test with 1000 gene promoter probe sets") %>%
  kable_styling(full_width=FALSE)

```

Now we can run the test with all probe sets.

Whole gene: 639 sec (10.5 mins)

Promoter: 261 sec (4.35 mins)

```{r,bigfry1}

tic()
fry_res1 <- fry(y=mxs, index = sets, design = design, contrast = ncol(design) )
toc()

head(fry_res1,20) %>%
  kbl(caption="fry test with all whole gene probe sets") %>%
  kable_styling(full_width=FALSE)

tic()
fry_res2 <- fry(y=mxs, index = psets, design = design, contrast = ncol(design) )
toc()

head(fry_res2,20) %>%
  kbl(caption="fry test with all gene promoter probe sets") %>%
  kable_styling(full_width=FALSE)

```

## save data object

```{r,save}

save.image("GSE158422_fry.Rdata")

```

### 2022-10-11 GSE158433/GSE158422.Rmd
getting fry to work

@@
 gse <- getGEO(filename=SERIES_MATRIX)
-baseDir <- "."
+baseDir <- "IDAT"
 sample_metadata <- pData(phenoData(gse))
@@
 rgSet
+
+targets$Basename <- gsub("IDAT/","",mybase)
+head(targets)
 
@@
 myscree(Mval_flt,main="excl sex chr")
+par(mfrow=c(1,1))
 
@@
 sex <- factor(samplesheet$sex,levels=c("0","1"))
-Mvals<-Mval
-betas<-beta
-dim(Mvals)
-dim(betas)
+dim(Mval)
+dim(beta)
 
 GSE158422 <- dm_analysis(samplesheet=samplesheet,
-  sex=sex,groups=groups,mx=Mvals,name="GSE158422",
-  myann=myann ,beta= betas)
+  sex=sex,groups=groups,mx=Mval,name="GSE158422",
+  myann=myann ,beta= beta)
 

### 2022-10-11 GSE158433/GSE158422_fry.Rmd
getting fry to work

@@
   library("RColorBrewer")
-  library("IlluminaHumanMethylation450kanno.ilmn12.hg19")
+  library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
   library("GEOquery")
@@
 
-## Differential analysis
+## Design matrix
 
-Paired normal versus cancer tissue.
-
-```{r,dm}
+```{r,design}
 
@@
 sex <- factor(samplesheet$sex,levels=c("0","1"))
-Mvals<-Mval
-betas<-beta
-dim(Mvals)
-dim(betas)
-
-GSE158422 <- dm_analysis(samplesheet=samplesheet,
-  sex=sex,groups=groups,mx=Mvals,name="GSE158422",
-  myann=myann ,beta= betas)
-
-mx=Mvals
-name="GSE158422"
-beta=betas
 design <- model.matrix(~ sex + groups)
-mxs <- mx[,which( colnames(mx) %in% samplesheet$Basename )]
-fit.reduced <- lmFit(mxs,design)
-fit.reduced <- eBayes(fit.reduced)
-summary(decideTests(fit.reduced))
+mxs <- Mval
 

### 2022-11-14 GSE158433/GSE158422.Rmd
reorganising analysis: gene and geneset Rmd reports

@@
 
-The purpose of this analysis is to apply different enrichment analysis approaches,
-and see which approach is most parsimonious with gene expressoin differences.
+This work is to conduct the methylation data processing and differential analysis of GSE158422, which is matched
+lung squamous cell carcinoma (LUSC) and control tissue.
 
-We will specifically be testing the GMEA approach versus existing approaches,
-namely GOseq and mitch.
-
-This study (GSE158433) is a super-series contains matched RNA-seq (GSE158420) and 
-Infinium EPIC array (GSE158422) data for matched control and cancer tissues.
-
-In this script, we will be processing the EPIC DNA methylation data and performing the
-statistical analysis in limma and saving the results for downstream enrichment analysis.
+The data is going to be saved as an Rdata file and then used for enrichment analysis in a subsequent Rmd script.
 

### 2022-11-14 GSE158433/GSE158422_genesetlevel.Rmd
reorganising analysis: gene and geneset Rmd reports

--- 2022-10-11 GSE158433/GSE158422_fry.Rmd
@@
 
-Here I will use fry to perform GMEA analysis.
+There are many ways to perform enrichment analysis of Infinium methylation data.
+The purpose of this work is to determine which one of these methods perform "best".
+Best performance is a function of specificity, sensitivity and biological consistency.
+
+* Specificity can be defined as the ability to control false discovery rate appropriately.
+This will be evaluated by analyzing a randomised data set.
+
+* Sensitivity is defined as the ability to identify genesets undergoing differentially methylation.
+This can be evaluated if we begin with a large study with many samples, analyse it and then repeat
+the process with fewer samples.
+The more sensitive methods will be able to identify differentially methylated pathways with
+a smaller sample size.
+
+* As methylation is known to be a characteristic of silenced genes, we can use this feature to examine
+whether the methylation pathway enrichment analysis is congruent with gene expression data, as this
+dataset has matched methylation and gene expression profiles for all samples.
+
+In addition I will be measuring at the execution time for these approaches.
+
+The approaches that I will be examining:
+
+* GOmeth: an approach to filter differential methylated probes followed by over-representation testing.
+
+* ebGSEA: global test to rank genes followed by Wilcoxon Test or Known Population Median Test
+
+* methylGSA: mRRA robust rank aggregation, mGLM
+
+* Fry test on probes belonging to a gene, followed by Camera PR based on significance/effectsize?
+
+* Median of probe intensity for each gene in each sample followed by Fry test on pathways.
 
@@
   library("kableExtra")
+  library("mitch")
+  library("fgsea")
 
@@
 
+## Pathway analysis
+
+Simple GSEA using FGSEA (gene permutation) and mitch (rank-ANOVA).
+
+```{r,fgsea}
+
+#download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", destfile="ReactomePathways.gmt.zip")
+#unzip("ReactomePathways.gmt.zip")
+genesets <- gmt_import("ReactomePathways.gmt")
+
+stat1 <- sign(as.numeric(factor(fry_res1$Direction))-1.5)  * -log10(fry_res1$PValue)
+stat1[is.na(stat1)] <- 0
+names(stat1) <- rownames(fry_res1)
+fres1 <- fgseaMultilevel(pathways=genesets,stats=stat1)
+fres1 <- fres1[order(fres1$pval),]
+
+head(fres1,20) %>%
+  kbl(caption="WG fry + fgsea Reactome results") %>%
+  kable_styling(full_width=FALSE)
+
+stat2 <- sign(as.numeric(factor(fry_res2$Direction))-1.5)  * -log10(fry_res2$PValue)
+stat2[is.na(stat2)] <- 0
+names(stat2) <- rownames(fry_res2)
+fres2 <- fgseaMultilevel(pathways=genesets,stats=stat2)
+fres2 <- fres2[order(fres2$pval),]
+head(fres2,20)  %>%
+  kbl(caption="Promoter fry + fgsea Reactome results") %>%
+  kable_styling(full_width=FALSE)
+
+```
+
+```{r,mitch}
+
+stat1 <- data.frame(stat1)
+mres1 <- mitch_calc(stat1, genesets, priority="effect")
+head(mres1$enrichment_result,20) %>%
+  kbl(caption="WG fry + mitch Reactome results") %>%
+  kable_styling(full_width=FALSE)
+
+stat2 <- data.frame(stat2)
+mres2 <- mitch_calc(stat2, genesets, priority="effect")
+head(mres2$enrichment_result,20) %>%
+  kbl(caption="Promoter fry + mitch Reactome results") %>%
+  kable_styling(full_width=FALSE)
+
+```
+
 ## save data object

### 2022-11-24 gmea2.Rmd
some work on the b-proof project

---
title: "GSE74548: Gene Methylation Enrichment Analysis"
author: "hcy_meth meta-analysis group"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
---

##  Introduction

Previously, Mandhri and I analysed the B-PROOF 450K data, trying to understand
whether vitamin supplementation caused changes in gene methylation.
We used Limma and some basic analyses, which showed no specific probes with FDR<0.05,
nor any DMRs.

Here we will attempt another approach.
First aggregate median signal for probes belonging to a gene, followed by limma and enrichment analysis.

To validate this approach it is possible to cut the dataset in half and see what % of results are consistent.

```{r,libs}

suppressPackageStartupMessages({
  library("limma")
  library("parallel")
  library("dplyr")
  library("kableExtra")
  library("eulerr")
  library("mitch")
  library("IlluminaHumanMethylation450kmanifest")
  library("IlluminaHumanMethylation450kanno.ilmn12.hg19")
  library("tictoc")
})

source("meth_functions.R")

```

## Get array annotation data

```{r,anno}

ann450k <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
myann <- data.frame(ann450k[,c("UCSC_RefGene_Name","Regulatory_Feature_Group")])
gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
names(gp2) <- rownames(gp)
sets <- split(rep(names(gp2), lengths(gp2)), unlist(gp2))
promoters <- grep("Prom",myann$Regulatory_Feature_Group)

```

# DM3 - association of Hcy with gene methylation

It is thought that high plasma homocysteine can inhibit DNA methylation.
Let's see whether that is the case, and which genes are most affected.
This analysis is conducted at the whole gene level as well as on the level of promoters.

This approach involves performing limma, and then for each gene the median t-score is
calculated.
This value is then used by mitch.

```{r,dma3_mitch}

design <- readRDS(file="hcy_design.rds")
mval <- readRDS(file="hcy_mval.rds")

tic()
res <- main(mval=mval,design=design,sets=sets,cores=4)
res_df <- res[[1]] # enrichment results
res_top <- res[[2]] # limma results
toc() ## 502-522 sec elapsed on 8 cores (8.3-8.7 mins)

hist(res_top$t)

hist(res_df$median)

head(res_df,30) %>%
  kbl(caption="Top GMEA genes") %>%
  kable_styling("hover", full_width = F)

genesets <- gmt_import("ReactomePathways.gmt")

y <- res_df[,"median",drop=FALSE]

colnames(y) <- "meth"

res <- mitch_calc(y, genesets, priority="effect")

head(res$enrichment_result,30) %>%
  kbl(caption="Top GMEA pathways", row.names = F) %>%
  kable_styling("hover", full_width = F)

```

A slight variation is to use cameraPR instead of mitch to better control for correlation.

```{r,dma3_cameraPR}

stat <- res_df$median

names(stat) <- rownames(res_df)
stat[is.na(stat)] <- 0

tic()
cres <- cameraPR(statistic=stat, index=genesets, use.ranks = FALSE, inter.gene.cor=0.01, sort = TRUE)
cres <- subset(cres,NGenes>4)
cres$FDR <- p.adjust(cres$PValue,method="fdr")
mymedians <- sapply(1:nrow(cres),function(i) {
  myset <- genesets[[which(names(genesets) == rownames(cres)[i])]]
  mystats <- stat[names(stat) %in% myset]
  median(mystats)
})
cres$median <- mymedians
toc() # 1.0 sec

sig <- subset(cres,FDR<0.05)

head(sig[order(-abs(sig$median)),],20) %>%
  kbl(caption = "Top GMEA pathways found with CAMERA after 5% FDR filtering") %>%
  kable_styling("hover", full_width = F)

```

Now try another approach by first aggregating and then performing limma, then pathway enrichment.

Aggregate mval to genes and examine pathway enrichment in blood.

```{r,gsmea_hcy}

gsmea <- function(mval,design,probesets,genesets,cores=2) {
  ag <- mclapply(probesets,function(ps) {
    mval <- mval[rownames(mval) %in% ps,,drop=FALSE]
    if (nrow(mval>1)) {
      o <- apply(mval,2,median)
    } else {
      o <- mval
    }
   o
  },mc.cores=cores)
  ag <- do.call(rbind,ag)
  res <- main(mval=ag,design=design,sets=genesets,cores=cores )
}

tic()
hcy_gsmea <- gsmea(mval=mval,design=design,probesets=sets,genesets=genesets,cores=8)
toc()

hcy_gsmea$res <- subset(hcy_gsmea$res,nprobes>=5)
hcy_gsmea$res$FDR <- p.adjust(hcy_gsmea$res$PValue,method="fdr")

head(hcy_gsmea$res,20) %>%
  kbl(caption = "Pathways found using the agg,lim,fry method") %>%
  kable_styling("hover", full_width = F)

```

## Whole gene analysis

```{r,wg}
# set cores to used for parallel execution
CORES= detectCores()

dm <- read.table("dma3a.tsv.gz")
dm <- dm[,4:9]
dm <- merge(myann,dm,by=0)
rownames(dm) <- dm[,1]
dm[,1] = NULL

calc_sc <- function(dm) {
  gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
  gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
  gnl <- mclapply(gnl,unique,mc.cores=CORES)
  dm$UCSC_RefGene_Name <- gnl
  l <- mclapply(1:nrow(dm), function(i) {
    a <- dm[i,]
    len <- length(a[[1]][[1]])
    tvals <- as.numeric(rep(a["t"],len))
    genes <- a[[1]][[1]]
    data.frame(genes,tvals)
  },mc.cores=CORES)
  df <- do.call(rbind,l)
  gme_res <- mclapply( 1:length(gn), function(i) {
    g <- gn[i]
    tstats <- df[which(df$genes==g),"tvals"]
    myn <- length(tstats)
    mymean <- mean(tstats)
    mymedian <- median(tstats)
    wtselfcont <- wilcox.test(tstats)
    res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,"median"=mymedian,
      "p-value(sc)"=wtselfcont$p.value)
  } , mc.cores=CORES )
  gme_res_df <- do.call(rbind, gme_res)
  rownames(gme_res_df) <- gme_res_df[,1]
  gme_res_df <- gme_res_df[,-1]
  tmp <- apply(gme_res_df,2,as.numeric)
  rownames(tmp) <- rownames(gme_res_df)
  gme_res_df <- as.data.frame(tmp)
  gme_res_df$sig <- -log10(gme_res_df[,4])
  gme_res_df <- gme_res_df[order(-gme_res_df$sig),]
  gme_res_df$`fdr(sc)` <- p.adjust(gme_res_df$`p-value(sc)`)
  out <- list("df"=df,"gme_res_df"=gme_res_df)
  return(out)
}

tic()
gme_res_wholegene <- calc_sc(dm)
time2 <- toc() #38 44 41 40 44 41 40
time2
df <- gme_res_wholegene[[1]]
res <- gme_res_wholegene[[2]]
write.table(res,file="gmea_wholegene.tsv")

```

## Table of top results.

```{r, toptable}

head(res,50) %>%
  kbl(caption = "Top significant genes with GMEA") %>%
  kable_styling("hover", full_width = F)

```

## Volcano plots

```{r,volcano1}

# volcano selfcont
sig <- subset(res,`fdr(sc)` < 0.05)
plot(res$median , -log10(res$`p-value(sc)`) ,
  xlab="effect size (mean t-stat)", ylab="-log10(p-value)",
  pch=19, cex=0.5, col="gray",main="self contained test")
grid()
points(sig$median , -log10(sig$`p-value(sc)`) ,
  pch=19, cex=0.5, col="red")

```

## Probe bias

```{r,probe_bias1}

plot(res$nprobes,res$sig,log="x",ylim=c(0,50),pch=19,cex=0.6)
points(sig$nprobes,sig$sig,col="red",pch=19,cex=0.62)
MIN = min(sig$nprobes)
LEFT = nrow(subset(res,nprobes<MIN))
RIGHT = nrow(subset(res,nprobes>MIN))
SIG = nrow(sig)
TOT = nrow(res)
HEADER <- paste(TOT, "genes in total.", SIG, "with FDR<0.05.",
  RIGHT, "well covered and", LEFT, "poorly covered")
mtext(HEADER)
abline(v=MIN,lty=2)

```

## Boxplots

Boxplots smallest pvalue.

```{r,box1}

par(mfrow=c(1,2))
n=50
# self contained

gs <- head(rownames(res),50)
tstats <- lapply(gs, function(g) {
  df[which(df$genes==g),"tvals"]
})
names(tstats) <- gs

tstats <- tstats[order(unlist(lapply(tstats,median)))]
boxplot(tstats,horizontal=TRUE,las=1,
  main="smallest p-val(selfcont)",cex.axis=0.6,
  xlab="t-statistic")
grid()

n=50
# effect size median
sig <- subset(res,`fdr(sc)` < 0.05)
gs <- head(rownames(sig[order(-abs(sig$median)),]),n)

if ( length(gs) >2 ) {
  tstats <- lapply(gs, function(g) {
    df[which(df$genes==g),"tvals"]
  })
  names(tstats) <- gs
  tstats <- tstats[order(unlist(lapply(tstats,median)))]
  boxplot(tstats,horizontal=TRUE,las=1,
    main="biggest effect size(median)",cex.axis=0.6,
    xlab="t-statistic")
  grid()
} else {
  plot(1)
  mtext("too few significant genes found")
}

```

## Mitch

```{r,dm1_mitch}

dmscore <- data.frame( res$median * res$sig)

rownames(dmscore) <- rownames(res)

colnames(dmscore) <- "metric"

if ( ! file.exists("ReactomePathways.gmt") ) {
  download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip",
    destfile="ReactomePathways.gmt.zip")
  unzip("ReactomePathways.gmt.zip")
}
file.info("ReactomePathways.gmt")
genesets <- gmt_import("ReactomePathways.gmt")
length(genesets)

mres <- mitch_calc(x=dmscore, genesets=genesets, priority="effect")

head(mres$enrichment_result,20) %>%
  kbl(caption = "Top enriched gene sets with GMEA-Mitch", row.names = FALSE) %>%
  kable_styling("hover", full_width = F)

mitch_report(mres,outfile="dma3a_mitch.html",overwrite=TRUE)

```

## Session Information

For reproducibility.

```{r,sessioninfo}

sessionInfo()

```

### 2023-09-14 figs/gmea.Rmd
refreshing hcy work

--- 2022-11-24 gmea2.Rmd
@@
 
-Here we will attempt another approach.
-First aggregate median signal for probes belonging to a gene, followed by limma and enrichment analysis.
+Here we will attempt functional enrichment analysis on this data, firstly using GOMeth/gsameth, and secondly
+with a new method called GMEA (Gene Methylation Enrichment Analysis).
+GMEA involves first aggregating median signal for probes belonging to a gene, followed by limma and subsequentenrichment analysis.
 
@@
   library("IlluminaHumanMethylation450kanno.ilmn12.hg19")
+  library("missMethyl")
   library("tictoc")

### 2023-09-14 figs/step_01a_GSE74548_prep.Rmd
refreshing hcy work

---
title: "GSE74548: Preparation"
author: "hcy_meth meta-analysis group"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
---

##  Introduction

In this report, I will take you through a re-analysis methylation data first described by
[Kok et al (2015)](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4644301/).

In their study, they analysed the DNA methylation patterns of 87 participants of aged 65-75 years with midly elevated Homocysteine levels. THese individuals were randomly asigned to take 400 μg folic acid and 500 μg vitamin B12 per day or a placebo during an intervention period of 2 years.

The platform used in the study is the Illumina Infinium HumanMethylation450k BeadChip assay.
The authors used a pipeline based on DMRs
([Peters et al, 2017](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4429355/)), together with Benjamini-Hochberg(BH) procedure ([Benjamini et al, 1995](https://rss.onlinelibrary.wiley.com/doi/abs/10.1111/j.2517-6161.1995.tb02031.x)).

The methylation data have been deposited to NCBI GEO repository accession number
[GSE74548](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE74548)

The main conclusions from the original study were:

* Long-term supplementation with folic acid and vitamin B12 resulted in DNA methylation changes in leukocytes of older persons.

* A change in DNA methylation was observed to be different between the participants receiving folic acid and vitamin B12 versus placebo.

* DNA methylation levels of several genomic loci were found to correlate to serum levels of either folate, vitamin B12, or plasma homocysteine.

* Most prominent DNA methylation patterns associated with supplemental intake or status of B-vitamins are related to developmental processes as well as carcinogenesis.

The aim of this work is to;

1. Asses the relationship between Homocysteine to DNA methylation patterns

2. develop the analytical pipelines required for efficient re-analysis of 450K array data,

3. to confirm that we are able to obtain differential methylation results that are similar
to those obtained in the original study, and

4. to critically evaluate the conclusions made in the original study.

## Loading packages

These packackes will help us to perform vital steps such as normalisation, filtering, 
differential analysis, etc, and provide information about the array probe annotaions.

```{r,packages}

Sys.setenv("VROOM_CONNECTION_SIZE" = 524288 * 2)

knitr::opts_chunk$set(dev = "png")

suppressPackageStartupMessages({
  library("missMethyl")
  library("GEOquery")
  library("limma")
  library("topconfects")
  library("minfi")
  library("IlluminaHumanMethylation450kmanifest")
  library("IlluminaHumanMethylation450kanno.ilmn12.hg19")
  library("DMRcate")
  library("mitch")
  library("kableExtra")
  library("forestplot")
  library("RColorBrewer")
  library("plyr")
  library("R.utils")
  library("eulerr")
  library("gplots")
  library("reshape2")
  library("beeswarm")
  library("RCircos")
})
# Annotation
ann450k <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
myann <- data.frame(ann450k[,c("UCSC_RefGene_Name","Regulatory_Feature_Group")])
promoters <- grep("Prom",myann$Regulatory_Feature_Group)

```

## Loading functions

These functions provide shortcuts to help with charts and other analysis.

```{r,load_functions}

source("meth_functions.R")
myranks<-function(x) {
  x$score <- sign(x$logFC)/log10(x$adj.P.Val)
  y <- x[,"score",drop=FALSE]
  y$rn <- x$Row.names
  return(y)
}

# heatmap for continuous data
heatmap_c<-function(dm,name,mx,n, groups) {
  my_palette <- colorRampPalette(c("blue", "white", "red"))(25)
  topgenes <-  rownames(head(dm[order(dm$P.Value),],n))
  ss <- mx[which(rownames(mx) %in% topgenes),]
  mycols <- colorRampPalette(c("white","yellow", "orange", "red","darkred"))(n = length(groups))
  colCols <- mycols[order(groups)]
  heatmap.2(ss,scale="row",margin=c(10, 10),cexRow=0.6,trace="none",cexCol=0.4,
  ColSideColors=colCols ,  col=my_palette, main=name)
}

# heatmap for continuous data - topconfects
make_heatmap2_c <- function(confects,name,mx,n, groups) {
  topgenes <-  head(confects$table$name,n)
  my_palette <- colorRampPalette(c("blue", "white", "red"))(25)
  ss <- mx[which(rownames(mx) %in% topgenes),]
  mycols <- colorRampPalette(c("white","yellow", "orange", "red","darkred"))(n = length(groups))
  colCols <- mycols[order(groups)]
  heatmap.2(ss,scale="row",margin=c(10, 10),cexRow=0.6,trace="none",cexCol=0.4,
  ColSideColors=colCols ,  col=my_palette, main=name)
}

```

## Data import

Data will be imported from GEO data base under the accession number-GSE74548
https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE74548

```{r,data}

if ( !dir.exists("GSE74548") ) { dir.create("GSE74548") }

ARRAY_SAMPLESHEET="GSE74548/GSE74548_Sample_description_BProof.txt"
# only download it if it is not present on the system
if ( !file.exists(ARRAY_SAMPLESHEET ) ) {
    DLFILE=paste(ARRAY_SAMPLESHEET,".gz",sep="")
    download.file("https://ftp.ncbi.nlm.nih.gov/geo/series/GSE74nnn/GSE74548/suppl/GSE74548_Sample_description_BProof.txt.gz",
        destfile = DLFILE)
    gunzip(DLFILE)
}

ARRAY_DATA="GSE74548/GSE74548_RAW.tar"
# only download it if it is not present on the system
if ( !dir.exists("GSE74548/IDAT") ) {
  dir.create("GSE74548/IDAT")
  download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE74548&format=file",
    destfile = ARRAY_DATA)
    untar(exdir = "GSE74548/IDAT", tarfile = ARRAY_DATA)
}

baseDir <- "GSE74548"
R.utils::gunzip("GSE74548/IDAT/GPL13534_HumanMethylation450_15017482_v.1.1.csv.gz",overwrite=TRUE, remove=FALSE)
targets <- read.metharray.sheet(baseDir,pattern="csv$")

if (! file.exists("GSE74548_series_matrix.txt.gz") ) {
  URL="https://ftp.ncbi.nlm.nih.gov/geo/series/GSE74nnn/GSE74548/matrix/GSE74548_series_matrix.txt.gz"
  download.file(URL,destfile = "GSE74548_series_matrix.txt.gz")
}
gse<- getGEO(filename = "GSE74548_series_matrix.txt.gz")

targets <- pData(phenoData(gse))
targets <- targets[order(rownames(targets)),]
mybase <- unique(gsub("_Red.idat.gz" ,"", gsub("_Grn.idat.gz", "" ,list.files("./GSE74548",pattern = "GSM",recursive = TRUE))))
mybase <- paste("GSE74548/", mybase, sep = "")
# sample number discrepancy. 311 IDAT FILES but GEO states 174 subjects.
gsm <-sapply(strsplit(mybase,"_"),"[[",1)
gsm <-gsub("GSE74548/IDAT/","",gsm)
#only using sampels described in meta data
targets$Basename<- mybase[which(gsm %in% rownames(targets))]
rgSet <- read.metharray.exp(targets = targets)
targets[1:6,1:5]

```

## Normalisation

```{r,norm,fig.cap="Figure 1. Normalisation of bead-array data with SWAN."}

mSet <- preprocessRaw(rgSet)
mSetSw <- SWAN(mSet,verbose=FALSE)
par(mfrow=c(1,2), cex=0.8)
densityByProbeType(mSet[,1], main = "Raw")
densityByProbeType(mSetSw[,1], main = "SWAN")

```

## Filter probes

Here we are running parallel analyses, both including and excluding sex chromosomes.

```{r,filterprobes}

# include sex chromosomes
detP <- detectionP(rgSet)
keep <- rowSums(detP < 0.01) == ncol(rgSet)
mSetSw <- mSetSw[keep,]

# exclude SNP probes
mSetSw <- mapToGenome(mSetSw)
mSetSw_nosnp <- dropLociWithSnps(mSetSw)
dim(mSetSw)
dim(mSetSw_nosnp)
mSetSw <- mSetSw_nosnp

# exclude sex chromosomes
keep <- !(featureNames(mSetSw) %in% ann450k$Name[ann450k$chr %in% c("chrX","chrY")])
mSetFlt <- mSetSw[keep,]
mSetFlt[1:6,1:5]
dim(mSetFlt)

```

## Extracting Beta and M-values

```{r,beta_m_vals}

# include sex chromosomes
meth <- getMeth(mSetSw)
unmeth <- getUnmeth(mSetSw)
Mval <- log2((meth + 100)/(unmeth + 100))
beta <- getBeta(mSetSw)

# exclude sex chromosomes
meth <- getMeth(mSetFlt)
unmeth <- getUnmeth(mSetFlt)
Mval_flt <- log2((meth + 100)/(unmeth + 100))
beta_flt <- getBeta(mSetFlt)

```

## sex chromosome diagnostic

```{r,sex chromosome diagnostic,fig.height=8,fig.width=8}

colnames(Mval) <- sapply(strsplit(colnames(Mval),"_"),"[[",1)
colnames(Mval_flt) <- sapply(strsplit(colnames(Mval_flt),"_"),"[[",1)

cgx <- rownames(Locations[which(Locations$chr %in% "chrX"),])
cgy <- rownames(Locations[which(Locations$chr %in% "chrY"),])

mvx <- Mval[which(rownames(Mval) %in% cgx),]
mvy <- Mval[which(rownames(Mval) %in% cgy),]

targets_m <- rownames(subset(targets,`gender:ch1`=="male"))
str(targets_m)
targets_f <- rownames(subset(targets,`gender:ch1`=="female"))

Mvalm <- Mval[,colnames(Mval) %in% targets_m]
Mvalf <- Mval[,colnames(Mval) %in% targets_f]

mvxm <- Mvalm[which(rownames(Mvalm) %in% cgx),]
mvym <- Mvalm[which(rownames(Mvalm) %in% cgy),]

mvxf <- Mvalf[which(rownames(Mvalf) %in% cgx),]
mvyf <- Mvalf[which(rownames(Mvalf) %in% cgy),]

plot(colMeans(mvx),colMeans(mvy),col="gray")
points(colMeans(mvxm),colMeans(mvym),col="blue")
points(colMeans(mvxf),colMeans(mvyf),col="red")

plot(colMeans(mvx),colMeans(mvy),col="gray")
points(colMeans(mvxm),colMeans(mvym),col="lightblue",pch=19,cex=1.5)
points(colMeans(mvxf),colMeans(mvyf),col="pink",pch=19,cex=1.5)
text(colMeans(mvx),colMeans(mvy),labels = colnames(mvx),cex=0.75)

```

## MDS analysis

[Multidimensional scaling(https://en.wikipedia.org/wiki/Multidimensional_scaling)  plot is a method used
to identify the major sources of variation in a dataset. In the MDS plots below, I will be plotting the 
first two dimensions (principal components [PCs]), with each sample label coloured either by Hcy 
classification, sample group,age, folate levels and vitb12 levels.

We will begin with MDS analysis including the sex chromosomes and then exclude them.

First, let's quantify the contribution of the major principal components. with a scree plot, so we can see 
whether most of the variation is captured in the first two PCs or whether it is spread over more PCs.
As we can see in Figure 2, the main source of variation is what is shown in PC1, and a much lesser
extent on the other dimensions. Interestingly, excluding sex chromosomes does not seem to change the 
relative contributions of PCs very much.

```{r,mds1,fig.width = 8 ,fig.height = 8,fig.cap="Figure 3. MDS plot coloured by targets groups"}

targets[1:6,1:5]
dim(targets)
targets$group <- factor(targets$source_name_ch1)
sample_group<-factor(targets$group)
targets$sex <- factor(targets$`gender:ch1`)
dim(sample_group)

colour_palette=brewer.pal(n = length(levels(targets$group)), name = "Paired")
colours <- colour_palette[as.integer(factor(targets$source_name_ch1))]
plot(1,axes = FALSE,xlab="",ylab="",main="treatment groups")
legend("center",legend=levels(targets$group),pch=16,cex=1.2,col=colour_palette)
mydist <- plotMDS(Mval, labels=targets$geo_accession,col=colours,main="sex chromosomes included")
mydist_flt <- plotMDS(Mval_flt, labels=targets$geo_accession,col=colours,main="sex chromosomes excluded")

```

```{r,hist1}

hcy<- targets$characteristics_ch1.10
hcy<-strsplit(as.character(targets$characteristics_ch1.10), " ")
hcy<-sapply(hcy, "[",5)
hcy<-as.numeric(hcy)
hist(hcy,breaks = 20,xlab = "hcy levels")
hcy_groups<-cut(hcy,breaks = c(0,10,16,30),labels = c("low","medium","high"))
table(hcy_groups)

```

```{r,mds2,fig.width = 8 ,fig.height = 8,fig.cap="Figure 3. MDS plot coloured by homocysteine levels"}

targets[1:6,1:5]
targets$sex <- factor(targets$`gender:ch1`)
targets$hcy_groups <- hcy_groups
sample_group<-hcy_groups

colour_palette=brewer.pal(n = length(levels(sample_group)), name = "Paired")
colours <- colour_palette[as.integer(factor(hcy_groups))]
plot(1,axes = FALSE,xlab="",ylab="",main="hcy levels")
legend("center",legend=levels(hcy_groups),pch=16,cex=1.2,col=colour_palette)
plotMDS(mydist, labels=targets$geo_accession,col=colours,main="sex chromosomes included")
plotMDS(mydist_flt, labels=targets$geo_accession,col=colours,main="sex chromosomes excluded")

```

```{r,hist2}

age<- targets$characteristics_ch1.5
age<-strsplit(as.character(targets$characteristics_ch1.5), " ")
age<-sapply(age , "[",4)
age<-as.numeric(age)
hist(age,breaks = 20,xlab = "age levels")
age_groups<-cut(age,breaks = c(0,65,70,75),labels = c("old","older","oldest"))
table(age_groups)

```

```{r,mds3,fig.width = 8 ,fig.height = 8,fig.cap="Figure 3. MDS plot coloured by age"}

targets$age <- factor(age_groups)
sample_group <-factor(age_groups)
colour_palette=brewer.pal(n = length(levels(sample_group)), name = "Paired")
colours <- colour_palette[as.integer(factor(age_groups))]
plot(1,axes = FALSE,xlab="",ylab="",main="age")
legend("center",legend=levels(age_groups),pch=16,cex=1.2,col=colour_palette)
plotMDS(mydist, labels=targets$geo_accession,col=colours,main="sex chromosomes included")
plotMDS(mydist_flt,labels=targets$geo_accession,col=colours,main="sex chromosomes excluded")

```

```{r,hist3}

folate_levels<- targets$characteristics_ch1.8
folate_levels<-strsplit(as.character(targets$characteristics_ch1.8), " ")
folate_levels<-sapply(folate_levels , "[",5)
folate_levels<-as.numeric(folate_levels)
hist(folate_levels,breaks = 20,xlab = "folate levels")
folate_levels<-cut(folate_levels,breaks = c(0,20,35,50,70,90),labels = c("lowest","low","medium","high","highest"))
table(folate_levels)
folate_levels

```

```{r,mds4,fig.width = 8 ,fig.height = 8,fig.cap="Figure 3. MDS plot coloured by folate levels"}

targets$folate_levels <- folate_levels
sample_group<-folate_levels

colour_palette=brewer.pal(n = length(levels(sample_group)), name = "Paired")
colours <- colour_palette[as.integer(factor(targets$folate_levels))]
plot(1,axes = FALSE,xlab="",ylab="",main="Folate levels")
legend("center",legend=levels(targets$folate_levels),pch=16,cex=1.2,col=colour_palette)
plotMDS(mydist, labels=targets$geo_accession,col=colours,main="sex chromosomes included")
plotMDS(mydist_flt, labels=targets$geo_accession,col=colours,main="sex chromosomes excluded")

```

```{r,hist4}

vitb12_levels<- targets$characteristics_ch1.9
vitb12_levels<-strsplit(as.character(targets$characteristics_ch1.9), " ")
vitb12_levels<-sapply(vitb12_levels , "[",6)
vitb12_levels<-as.numeric(vitb12_levels)
hist(vitb12_levels,breaks = 20,xlab = "vitb12 levels")
vitb12_levels<-cut(vitb12_levels,breaks = c(100,300,500,700,900,1115),labels = c("lowest","low","medium","high","highest"))
table(vitb12_levels)
vitb12_levels

```

```{r,mds5,fig.width = 8 ,fig.height = 8,fig.cap="Figure 3. MDS plot coloured by vitaminb12 levels"}

targets$vitb12_levels <- vitb12_levels
sample_group<-vitb12_levels

colour_palette=brewer.pal(n = length(levels(sample_group)), name = "Paired")
colours <- colour_palette[as.integer(factor(targets$vitb12_levels))]
plot(1,axes = FALSE,xlab="",ylab="",main="vitb12 levels")
legend("center",legend=levels(targets$vitb12_levels),pch=16,cex=1.2,col=colour_palette)
plotMDS(mydist, labels=targets$geo_accession,col=colours,main="sex chromosomes included")
plotMDS(mydist_flt, labels=targets$geo_accession,col=colours,main="sex chromosomes excluded")

```

# Analysis of covariates

```{r,folate vs hcy vs age}

samplesheet <- targets
folate_levels <- samplesheet$characteristics_ch1.8
folate_levels <-strsplit(as.character(samplesheet$characteristics_ch1.8), " ")
folate_levels <-sapply(folate_levels , "[",5)
folate_levels <-as.numeric(folate_levels)
folate_levels

hcy<- samplesheet$characteristics_ch1.10
hcy<-strsplit(as.character(samplesheet$characteristics_ch1.10), " ")
hcy<-sapply(hcy, "[",5)
hcy<- as.numeric(hcy)

age<- samplesheet$characteristics_ch1.5
age<-strsplit(as.character(samplesheet$characteristics_ch1.5), " ")
age<-sapply(age , "[",4)
age<-as.numeric(age)

pairs(data.frame(folate_levels, hcy, age))

```

# Differential analysis

There are several differential contrasts that would be of interest to us in this study:

1. placebo(43) vs supplement(44) follow-up

2. placebo(43) vs supplement(44) Baseline

3. hcy levels(172)

4. folate levels(172)

5. vitb12 levels(172)

```{r,dm1A,fig.height=8,fig.width=8}

samplesheet<-targets[grep("follow-up",targets$source_name_ch1),]
sex <- factor(samplesheet$`gender:ch1`)
age<-samplesheet$`age at baseline:ch1`
age<-as.numeric(age)

groups<-factor(samplesheet$source_name_ch1,levels = c("Buffy coat, placebo, follow-up","Buffy coat, FA/vB12, follow-up"))
mx <-Mval_flt
name="placebo_vs_supplement_follow-up"
design <- model.matrix(~age+ sex +groups)
mxs <- mx[,which(colnames(mx) %in% rownames(samplesheet) )]
fit.reduced <- lmFit(mxs,design)
fit.reduced <- eBayes(fit.reduced)
summary(decideTests(fit.reduced))
dm <- topTable(fit.reduced,coef=4, number = Inf)
dma <- merge(myann,dm,by=0)
dma1a <- dma[order(dma$P.Value),]

write.table(dma1a,file="dma1a.tsv")

head(dma1a, 50) %>% kbl() %>% kable_paper("hover", full_width = F)

dma1a_d<-nrow(subset(dm,adj.P.Val<0.05,logFC<0))
dma1a_u<-nrow(subset(dm,adj.P.Val<0.05,logFC>0))

confects <- limma_confects(fit.reduced, coef=3, fdr=0.05)
head(confects$table, 50) %>% kbl() %>% kable_paper("hover", full_width = F)

ann450k[which(rownames(ann450k )%in% confects$table[1:10,6]),c(1:3,24)]  %>% kbl() %>% kable_paper("hover", full_width = F)
colCols <- as.numeric(as.factor(groups))

make_volcano(dma1a,name = "placebo_vs_supplement_follow-up(sex,age)",mx=Mval_flt)
rownames(dma1a)<-dma1a[,1]
make_beeswarms(dm=dma1a ,name="placebo_vs_supplement_follow-up(sex,age)" , mx=beta_flt , groups=groups , n= 15)
make_beeswarms_confects(confects = confects ,name="placebo_vs_supplement_follow-up(sex,age)" , mx=beta_flt , groups=groups , n= 15)
make_heatmap(dm=dma1a , name="placebo_vs_supplement_baseline(sex,age)" , mx=mxs ,n = 50, groups=groups)
make_heatmap2(confects = confects , name="placebo_vs_supplement_baseline(sex,age)",mx=mxs ,n = 50, groups=groups)

# gene methylation analysis

genes <- unique(unlist(strsplit( dma1a$UCSC_RefGene_Name ,";")))

```

```{r,dm2A}

samplesheet<-targets[grep("baseline",targets$source_name_ch1),]
sex <- factor(samplesheet$`gender:ch1`)
age<-samplesheet$`age at baseline:ch1`
age<-as.numeric(age)
age
groups<-factor(samplesheet$source_name_ch1,levels = c("Buffy coat, placebo, baseline","Buffy coat, FA/vB12, baseline"))
mx <-Mval_flt
name="placebo_vs_supplement_baseline"
design <- model.matrix(~age+ sex +groups)
mxs <- mx[,which(colnames(mx) %in% rownames(samplesheet) )]
fit.reduced <- lmFit(mxs,design)
fit.reduced <- eBayes(fit.reduced)
summary(decideTests(fit.reduced))
dm <- topTable(fit.reduced,coef=4, number = Inf)
dma <- merge(myann,dm,by=0)
dma2a <- dma[order(dma$P.Value),]

write.table(dma2a,file="dma2a.tsv")

head(dma2a, 50) %>% kbl() %>% kable_paper("hover", full_width = F)

dma2a_d<-nrow(subset(dm,adj.P.Val<0.05,logFC<0))
dma2a_u<-nrow(subset(dm,adj.P.Val<0.05,logFC>0))

confects <- limma_confects(fit.reduced, coef=3, fdr=0.05)
head(confects$table, 50) %>% kbl() %>% kable_paper("hover", full_width = F)
ann450k[which(rownames(ann450k )%in% confects$table[1:10,6]),c(1:3,24)]  %>% kbl() %>% kable_paper("hover", full_width = F)
make_volcano(dma2a,name = "placebo_vs_supplement_baseline(sex,age)",mx=Mval_flt)
rownames(dma2a)<-dma2a[,1]
make_beeswarms(dm=dma2a ,name="placebo_vs_supplement_baseline(sex,age)" , mx=beta_flt , groups=groups , n= 15)
make_beeswarms_confects(confects = confects ,name="placebo_vs_supplement_baseline(sex,age)" , mx=beta_flt , groups=groups , n= 15)
make_heatmap(dm=dma2a , name="plb_supl_baseline" , mx=mxs ,n = 50, groups=groups)
make_heatmap2(confects = confects , name="placebo_vs_supplement_baseline(sex,age)",mx=mxs ,n = 50, groups=groups)

```

```{r,dm3A,fig.height=8,fig.width=8}

samplesheet<-targets
hcy<- samplesheet$characteristics_ch1.10
hcy<-strsplit(as.character(samplesheet$characteristics_ch1.10), " ")

hcy<-sapply(hcy, "[",5)

hcy<- as.numeric(hcy)
samplesheet<-samplesheet[which(!is.na(hcy)),]
hcy<-hcy[which(!is.na(hcy))]

groups<-cut(hcy,breaks = c(0,10,16,30),labels = c("low","medium","high"))
groups<-as.integer(groups)

sex <- factor(samplesheet$`gender:ch1`)
class(sex)

age<-samplesheet$`age at baseline:ch1`
age<-as.numeric(age)
mx <-Mval_flt
name="hcy_levels "

design <- model.matrix(~ age+ sex + groups)
mxs <- mx[,which( colnames(mx) %in% rownames(samplesheet) )]
saveRDS(mxs,"dm3_mx.Rds")
saveRDS(design,"dm3_design.Rds")

fit.reduced <- lmFit(mxs,design)
fit.reduced <- eBayes(fit.reduced)
summary(decideTests(fit.reduced))
dm <- topTable(fit.reduced,coef=4, number = Inf)
dma <- merge(myann,dm,by=0)
dma3a <- dma[order(dma$P.Value),]

write.table(dma3a,file="dma3a.tsv")

head(dma3a, 50) %>% kbl() %>% kable_paper("hover", full_width = F)

dma3a_d<-nrow(subset(dm,adj.P.Val<0.05,logFC<0))
dma3a_u<-nrow(subset(dm,adj.P.Val<0.05,logFC>0))

confects <- limma_confects(fit.reduced, coef=3, fdr=0.05)
head(confects$table, 50) %>% kbl() %>% kable_paper("hover", full_width = F)
ann450k[which(rownames(ann450k )%in% confects$table[1:10,6]),c(1:3,24)]  %>% kbl() %>% kable_paper("hover", full_width = F)
make_volcano(dma3a,name = "hcy_levels(sex,age)",mx=Mval_flt)
rownames(dma3a)<-dma3a[,1]
make_beeswarms(dm=dma3a ,name="hcy_levels(sex,age)" , mx=beta_flt , groups=groups , n= 15)
make_beeswarms_confects(confects = confects ,name="hcy_levels(sex,age)" , mx=beta_flt , groups=groups , n= 15)

heatmap_c(dm=dma3a , name="hcy_levels(sex,age)" , mx=mxs ,n = 50, groups=groups)
make_heatmap2_c(confects = confects , name="hcy_levels(sex,age)",mx=mxs ,n = 50, groups=groups)

```

```{r,dm4A}

samplesheet<-targets
folate_levels<- samplesheet$characteristics_ch1.8

folate_levels<-strsplit(as.character(samplesheet$characteristics_ch1.8), " ")
folate_levels<-sapply(folate_levels , "[",5)
folate_levels<-as.numeric(folate_levels)

sex <- factor(samplesheet$`gender:ch1`)
age<-samplesheet$`age at baseline:ch1`
age<-as.numeric(age)
groups<-factor(samplesheet$source_name_ch1,levels = c("Buffy coat, placebo, baseline","Buffy coat, FA/vB12, baseline"))
mx <-Mval_flt
name="folate_levels"
age<-samplesheet$`age at baseline:ch1`
age<-as.numeric(age)

design <- model.matrix(~ age+ sex + folate_levels)
mxs <- mx[,which( colnames(mx) %in% rownames(samplesheet) )]
fit.reduced <- lmFit(mxs,design)
fit.reduced <- eBayes(fit.reduced)
summary(decideTests(fit.reduced))
dm <- topTable(fit.reduced,coef=4, number = Inf)
dma <- merge(myann,dm,by=0)
dma4a <- dma[order(dma$P.Value),]

write.table(dma4a,file="dma4a.tsv")

head(dma4a, 50) %>% kbl() %>% kable_paper("hover", full_width = F)

dma4a_d<-nrow(subset(dm,adj.P.Val<0.05,logFC<0))
dma4a_u<-nrow(subset(dm,adj.P.Val<0.05,logFC>0))

confects <- limma_confects(fit.reduced, coef=3, fdr=0.05)
head(confects$table, 50) %>% kbl() %>% kable_paper("hover", full_width = F)
ann450k[which(rownames(ann450k )%in% confects$table[1:10,6]),c(1:3,24)]  %>% kbl() %>% kable_paper("hover", full_width = F)
make_volcano(dma4a,name = "folate_levels(sex,age)",mx=Mval_flt)
rownames(dma4a)<-dma4a[,1]
make_beeswarms(dm=dma4a ,name="folate_levels(sex,age)" , mx=beta_flt , groups=groups , n= 15)
make_beeswarms_confects(confects = confects ,name="folate_levels(sex,age)" , mx=beta_flt , groups=groups , n= 15)
make_heatmap(dm=dma4a , name="folate_levels(sex,age)" , mx=mxs ,n = 50, groups=groups)
make_heatmap2(confects = confects , name="folate_levels(sex,age)",mx=mxs ,n = 50, groups=groups)

```

```{r,dm5A}

samplesheet<-targets
vitb12_levels<- targets$characteristics_ch1.9
vitb12_levels<-strsplit(as.character(samplesheet$characteristics_ch1.9), " ")
vitb12_levels<-sapply(vitb12_levels , "[",6)
vitb12_levels<-as.numeric(vitb12_levels)

sex <- factor(samplesheet$`gender:ch1`)

age<-samplesheet$`age at baseline:ch1`
age<-as.numeric(age)

mx <-Mval_flt
name="vitmin_b12_levels"

groups<-factor(samplesheet$source_name_ch1,levels = c("Buffy coat, placebo, baseline","Buffy coat, FA/vB12, baseline"))
design <- model.matrix(~ age+ sex + vitb12_levels)
mxs <- mx[,which( colnames(mx) %in% rownames(samplesheet) )]
fit.reduced <- lmFit(mxs,design)
fit.reduced <- eBayes(fit.reduced)
summary(decideTests(fit.reduced))
dm <- topTable(fit.reduced,coef=4, number = Inf)
dma <- merge(myann,dm,by=0)
dma5a <- dma[order(dma$P.Value),]

write.table(dma5a,file="dma5a.tsv")

head(dma5a, 50) %>% kbl() %>% kable_paper("hover", full_width = F)

dma5a_d<- nrow(subset(dm,adj.P.Val<0.05,logFC<0))
dma5a_u<-nrow(subset(dm,adj.P.Val<0.05,logFC>0))

confects <- limma_confects(fit.reduced, coef=3, fdr=0.05)
head(confects$table, 50) %>% kbl() %>% kable_paper("hover", full_width = F)
ann450k[which(rownames(ann450k )%in% confects$table[1:10,6]),c(1:3,24)]  %>% kbl() %>% kable_paper("hover", full_width = F)
make_volcano(dma5a,name = "vitmin_b12_levels(sex,age)",mx=Mval_flt)
rownames(dma5a)<-dma5a[,1]
make_beeswarms(dm=dma5a ,name="vitmin_b12_levels" , mx=beta_flt , groups=groups , n= 15)
make_beeswarms_confects(confects = confects ,name="vitmin_b12_levels" , mx=beta_flt , groups=groups , n= 15)
make_heatmap(dm=dma5a , name="vitmin_b12_levels(sex,age)" , mx=mxs ,n = 50, groups=groups)
make_heatmap2(confects = confects , name="vitmin_b12_levels(sex,age)",mx=mxs ,n = 50, groups=groups)

```

```{r,summarised dms}

downs<-c(dma1a_d,dma2a_d,dma3a_d, dma4a_d,dma5a_d)
ups<-c(dma1a_u,dma2a_u, dma3a_u,dma4a_u,dma5a_u)
my_summary<-data.frame(downs,ups)
row.names(my_summary)<-c("placebo vs supplement follow-up(sex,age)",
                         "placebo vs supplement Baseline(sex,age)",
                         "hcy levels(sex,age)",
                         "folate levels(sex,age)",
                         "vitb12 levels(sex,age)")
library(knitr)
kable(my_summary)

```

## Venn diagrams of the differential methylated probes for each contrast

First, we look at the similarity of dmps altered by fresh and frozen procedures.
The overlap is very large.
This indicates a high degree of similarity between the profiles.

```{r,venn1,fig.cap="Comparison of probes in related to placebo vs supplement in follow-up(sex,age)", fig.width = 8 ,fig.height = 8}

v1 <- list("plb v sup fl(up)" =dma1a_u , 
           "hcy(up)" =dma3a_d ,
           "plb v sup fl(dwn)" =dma1a_d ,
           "hcy(dwn)" =dma3a_d)
head(v1)
plot(euler(v1, shape = "ellipse"), quantities = TRUE)

```

```{r,venn2,fig.cap="Comparison of probes in related to hcy(sex) vs folate level(sex,age)", fig.width = 8 ,fig.height = 8}

v2 <- list("fol(up)" =dma4a_u, 
           "hcy(up)" =dma3a_u,
           "fol(dwn)" =dma4a_d,
           "hcy(dwn)" =dma3a_d)
head(v2)
str(v2)
plot(euler(v2, shape = "ellipse"), quantities = TRUE)

```

```{r,venn3,fig.cap="Comparison of probes in related to hcy(age) vs folate level(sex,age)", fig.width = 8 ,fig.height = 8}

v3 <- list("plb v sup fl(up)" =dma1a_u, 
           "fol(up)" = dma4a_u ,
           "plb v sup fl(dwn)" =dma1a_d ,
           "fol(dwn)" =dma4a_d)
head(v3)
plot(euler(v3, shape = "ellipse"), quantities = TRUE)

```

```{r,venn4,fig.cap="Comparison of probes in related to hcy vs folate level(sex,age)", fig.width = 8 ,fig.height = 8}

v4 <- list("hcy(up)" =dma3a_u, 
           "fol(up)" = dma4a_u ,
           "hcy(dwn)" =dma3a_d ,
           "fol(dwn)" =dma4a_d)
head(v4)
plot(euler(v4, shape = "ellipse"), quantities = TRUE)

```

```{r,venn5,fig.cap="Comparison of probes in related to hcy vs vitb12 level(sex,age)", fig.width = 8 ,fig.height = 8}

v5<- list("hcy(up)" =dma3a_u, 
           "vitb12(up)" = dma5a_u ,
           "hcy(dwn)" =dma3a_d ,
           "vitb12(dwn)" =dma5a_d)
head(v5)
plot(euler(v5, shape = "ellipse"), quantities = TRUE)

```

## Spearman correlations of each contrast

While the above Venn diagrams are suggestive of similarity between contrasts, the best 
way to assess this is with correlation analysis. I have chosen Spearman as based on
the directional p-value ranking metrics

```{r,spearman, fig.width = 8 ,fig.height = 8}

mycontrasts <- list("placebo_vs_supplement-follow-up(sex,age)"=dma1a,"placebo vs supplement Baseline(sex,age)"=dma2a,"hcy levels(sex,age)"=dma3a,"folate levels(sex,age)"=dma4a,"vitb12 levels(sex,age)"=dma5a)
lapply(mycontrasts, head)
myrnks<-lapply(X = mycontrasts, FUN = myranks)
str(myrnks)
df <- join_all(myrnks,by="rn")

rownames(df) <- df$rn
df$rn=NULL
colnames(df) <- names(mycontrasts)
head(df)
mycors <- cor(df,method = "spearman")
my_palette <- colorRampPalette(c("darkred","red", "orange", "yellow","white"))(n = 25)
heatmap.2(mycors,scale="none",margin=c(10, 10),cexRow=0.8,trace="none",cexCol=0.8,
    col=my_palette,main="Spearman correlations")

```

## Enrichment analysis

Will be done in another Rmd file to reduce the size of the final HTML.

```{r,save}

save.image("GSE74548.Rdata")

```

## Session information

```{r,sessioninfo}

sessionInfo()

```

### 2023-09-14 figs/step_01b_test_gometh.Rmd
refreshing hcy work

---
title: "GOmeth test"
author: "hcy_meth meta-analysis group"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
---

## Introduction

Here we will run test enrichment analysis approaches for undderstanding the effect of elevated
homocysteine on blood methylomes.
Homocysteine is thought to be an inhibitor of methyltransferases and hyperhomocysteine is
associated with overall reduced genomic methylation.
Previously we analysed the B-PROOF study data which enabled us to identify probes whose methylation correlatesor anti-correlated with homocysteine level.

The two enrichment analysis approaches we will use are:

* `gsameth`: an ORA approach that addressed the biases inherent with infinium array analysis.

* `gmea`: an experimental application of functional class scoring for infinium analysis.

```{r,libs}

suppressPackageStartupMessages({
  library("missMethyl")
  library("kableExtra")
  library("mitch")
  library("eulerr")
  library("IlluminaHumanMethylation450kmanifest")
  library("IlluminaHumanMethylation450kanno.ilmn12.hg19")
  library("tictoc")
})

```

Load the data.

```{r,load}

dm <- read.table("dma3a.tsv")
rownames(dm) <- dm$Row.names
dm$Row.names=NULL

head(dm, 10) %>% kbl() %>% kable_paper("hover", full_width = F)

```

Reactome pathways were downloaded on the 14th Sept 2023 from MsigDB.

```{r,gsdl}

gs_entrez <- gmt_import("c2.cp.reactome.v2023.1.Hs.entrez.gmt")

gs_symbols <- gmt_import("c2.cp.reactome.v2023.1.Hs.symbols.gmt")

```

## ORA with GSAMETH - parth of the MissMethyl package

Now perform ORA analysis using gsameth with the top 500 probes.

```{r,gsameth1}

sigup <- rownames(head(subset(dm,logFC>0),1000))
gsaup <- gsameth(sig.cpg=head(sigup,500), all.cpg=rownames(dm), collection=gs_entrez)
gsaup <- gsaup[order(gsaup$P.DE),]
head(gsaup,10) %>% kbl(caption="Top hypermethylated pathways") %>% kable_paper("hover", full_width = F)

sigdn <- rownames(head(subset(dm,logFC<0),1000))
gsadn <- gsameth(sig.cpg=head(sigdn,500), all.cpg=rownames(dm), collection=gs_entrez)
gsadn <- gsadn[order(gsadn$P.DE),]
head(gsadn,10) %>% kbl(caption="Top hypomethylated pathways") %>% kable_paper("hover", full_width = F)

```

## GMEA

Now with GMEA.

Start with gathering the annotation set.

```{r,anno}

ann450k <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
myann <- data.frame(ann450k[,c("UCSC_RefGene_Name","Regulatory_Feature_Group")])
promoters <- grep("Prom",myann$Regulatory_Feature_Group)

```

Histogram of limma t-statistics.

```{r,hist}

dm <- dm[,c("UCSC_RefGene_Name","t")]
hist(dm$t,breaks=seq(from=-6,to=6,by=1)) ; grid()

```

Aggregate probe t-scores to gene level scores.

The object generated has two parts:

1. `df`: a simple table of the gene name and the median tvalue

2. `gme_res_df`: results of 1-sample Wilcox test.

```{r,agg_wt}

CORES= detectCores()

agg <- function(dm) {
  gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
  gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
  gnl <- mclapply(gnl,unique,mc.cores=CORES)
  dm$UCSC_RefGene_Name <- gnl
  l <- mclapply(1:nrow(dm), function(i) {
    a <- dm[i,]
    len <- length(a[[1]][[1]])
    tvals <- as.numeric(rep(a[2],len))
    genes <- a[[1]][[1]]
    data.frame(genes,tvals)
  },mc.cores=CORES)
  df <- do.call(rbind,l)
  gme_res <- mclapply( 1:length(gn), function(i) {
    g <- gn[i]
    tstats <- df[which(df$genes==g),"tvals"]
    myn <- length(tstats)
    mymean <- mean(tstats)
    mymedian <- median(tstats)
    wtselfcont <- wilcox.test(tstats)
    res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,"median"=mymedian,
      "p-value(sc)"=wtselfcont$p.value)
  } , mc.cores=CORES )
  gme_res_df <- do.call(rbind, gme_res)
  rownames(gme_res_df) <- gme_res_df[,1]
  gme_res_df <- gme_res_df[,-1]
  tmp <- apply(gme_res_df,2,as.numeric)
  rownames(tmp) <- rownames(gme_res_df)
  gme_res_df <- as.data.frame(tmp)
  gme_res_df$sig <- -log10(gme_res_df[,4])
  gme_res_df <- gme_res_df[order(-gme_res_df$sig),]
  gme_res_df$`fdr(sc)` <- p.adjust(gme_res_df$`p-value(sc)`)
  out <- list("df"=df,"gme_res_df"=gme_res_df)
  return(out)
}

tic()
dmagg <- agg(dm)
toc()
head(dmagg$gme_res_df) %>% kbl(caption="top gmea genes (Wilcox)") %>% kable_paper("hover", full_width = F)

```

Does 1-sample t-test work better?

```{r,agg_tt}

agg2 <- function(dm) {
  gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
  gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
  gnl <- mclapply(gnl,unique,mc.cores=CORES)
  dm$UCSC_RefGene_Name <- gnl
  l <- mclapply(1:nrow(dm), function(i) {
    a <- dm[i,]
    len <- length(a[[1]][[1]])
    tvals <- as.numeric(rep(a["t"],len))
    genes <- a[[1]][[1]]
    data.frame(genes,tvals)
  },mc.cores=CORES)
  df <- do.call(rbind,l)
  keep <- names(which(table(df$genes)>1))
  df <- df[df$genes %in% keep,]
  gn <- unique(df$genes)
  gme_res <- lapply( 1:length(gn), function(i) {
    g <- gn[i]
    tstats <- df[which(df$genes==g),"tvals"]
    myn <- length(tstats)
    mymean <- mean(tstats)
    mymedian <- median(tstats)
    tselfcont <- t.test(tstats)
    res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,"median"=mymedian,
      "p-value(sc)"=tselfcont$p.value)
  } )
  gme_res_df <- do.call(rbind, gme_res)
  rownames(gme_res_df) <- gme_res_df[,1]
  gme_res_df <- gme_res_df[,-1]
  tmp <- apply(gme_res_df,2,as.numeric)
  rownames(tmp) <- rownames(gme_res_df)
  gme_res_df <- as.data.frame(tmp)
  gme_res_df$sig <- -log10(gme_res_df[,4])
  gme_res_df <- gme_res_df[order(-gme_res_df$sig),]
  gme_res_df$`fdr(sc)` <- p.adjust(gme_res_df$`p-value(sc)`)
  out <- list("df"=df,"gme_res_df"=gme_res_df)
  return(out)
}

tic()
dmagg2 <- agg2(dm)
toc()
head(dmagg2$gme_res_df) %>% kbl(caption="top gmea genes (t-test)") %>% kable_paper("hover", full_width = F)

```

This results shows that wilcox test executes faster but 1-sample t-test gives more significant findings.

Check that the distribution is even.

```{r,tmeddist}

hist(dmagg2$gme_res_df$median,breaks=30) ; grid()

```

If the distribution is even, then established enrichment analysis can be performed.
Options include FGSEA and mitch.

Here I will try downstream analysis with mitch.

```{r,mitch1}

m <- dmagg2$gme_res_df

m <- m[,"median",drop=FALSE]

mres <- mitch_calc(x=m, genesets=gs_symbols, priority="effect")

head(mres$enrichment_result,20) %>% kbl(caption="top mitch pathways") %>% kable_paper("hover", full_width = F)
hist(mres$enrichment_result$s.dist) ; grid()

tbl <- mres$enrichment_result

sig <- subset(tbl,p.adjustANOVA<0.05)
nrow(sig)

plot(tbl$s.dist,-log10(tbl$pANOVA),pch=19,col="gray")
grid()
points(sig$s.dist,-log10(sig$pANOVA),pch=19,col="red")

msig <- sig

```

Some interesting pathways found.

There was no overlap with `gsameth()`.

Interestingly many gene sets exhibit increased methylation.

There could be a directional bias, so I will also conduct a 1-sample t-test.

```{r,t_enrichment_test1}

ttenrich <- function(m,genesets) {
  res <- mclapply( 1:length(genesets), function(i) {
    gs <- genesets[i]
    name <- names(gs)
    n_members <- length(which(rownames(m) %in% gs[[1]]))
    if ( n_members > 4 ) {
      tstats <- m[which(rownames(m) %in% gs[[1]]),]
      myn <- length(tstats)
      mymean <- mean(tstats)
      mymedian <- median(tstats)
      wt <- t.test(tstats)
      res <- c(name,myn,mymean,mymedian,wt$p.value)
    }
  } , mc.cores = CORES)
  res_df <- do.call(rbind, res)
  rownames(res_df) <- res_df[,1]
  res_df <- res_df[,-1]
  colnames(res_df) <- c("n_genes","t_mean","t_median","p.value")
  tmp <- apply(res_df,2,as.numeric)
  rownames(tmp) <- rownames(res_df)
  res_df <- tmp
  res_df <- as.data.frame(res_df)
  res_df <- res_df[order(res_df$p.value),]
  res_df$logp <- -log10(res_df$p.value )
  res_df$fdr <- p.adjust(res_df$p.value,method="fdr")
  res_df[order(abs(res_df$p.value)),]
  return(res_df)
}

gse_res_df <- ttenrich(m=m,genesets=gs_symbols)

head(gse_res_df,30) %>%
  kbl(caption = "Top significant genesets") %>%
  kable_paper("hover", full_width = F)

head(gse_res_df[order(-abs(gse_res_df$t_median)),],30) %>%
  kbl(caption = "Top effect size genesets") %>%
  kable_paper("hover", full_width = F)

sig <- subset(gse_res_df,fdr<0.05)
nrow(sig)

tsig <- sig

```

## Session information

For reproducibility.

```{r,session}

sessionInfo()

```

### 2023-09-14 old/GSE158433/GSE158422_gmea.Rmd
refreshing hcy work

--- 2022-10-06 GSE158433/GSE158422_gmea.Rmd
@@
 
+hist(dm$t)
+
 ```
@@
 
-hist(dm$t)
-tic() ; res <- calc_sc(dm) ; toc() #32 cores 95.5
-res2 <- res[[2]]
+tic()
+resw <- calc_sc(dm)
+toc() #16 cores 129 sec
+res2 <- resw[[2]]
 
@@
 
-hist(dm$t)
-tic() ; res <- calc_sc2(dm) ; toc() #
-res2 <- res[[2]]
+tic()
+rest <- calc_sc2(dm)
+toc() #16 cores 252 sec
+
+res2 <- rest[[2]]
 
@@
 
-hist(dm$t)
 head(dm)
@@
 SIG = nrow(sig)
+UP = nrow(subset(sig,ES>0))
+DN = nrow(subset(sig,ES<0))
 
@@
 points(sig$ES,-log10(sig$pval),col="red")
+mtext(paste(SIG,"signifiant genes,",UP,"up,",DN,"down"))
 
@@
 points(sig$size,-log10(sig$pval),col="red")
+mtext(paste(SIG,"signifiant genes,",UP,"up,",DN,"down"))
 
-gmea_volc(res2)
-gmea_boxplot(res)
-res2_wg <- res2
+```
+
+## Using CAMERA
+
+Need to add volcano and nprobe charts here.
+
+```{r,cam_func}
+
+stat <- dm$t
+names(stat) <- rownames(dm)
+
+# parallel approach
+CORES=detectCores()
+#sets <- head(sets,100000)
+sets_split <- split(sets,1:CORES)
+
+tic()
+
+cres <- mclapply(X=sets_split, function(mysets) {
+  cameraPR(statistic=stat, index=mysets, use.ranks = FALSE, inter.gene.cor=0.01, sort = TRUE)
+}, mc.cores=CORES)
+
+cres <- do.call(rbind,cres)
+rownames(cres) <- sub("\\.","@", rownames(cres))
+rownames(cres) <- sapply(strsplit(rownames(cres),"@"),"[[",2)
+cres <- cres[order(cres$PValue),]
+g <- rownames(cres)[1]
+
+m <- mclapply(rownames(cres) , function(g) {
+  probes <- sets[[g]]
+  scores <- stat[which(names(stat) %in% probes)]
+  mymedian <- median(scores)
+  mymean <- mean(scores)
+  c(mymean,mymedian)
+},mc.cores=CORES)
+
+mdf <- do.call(rbind,m)
+colnames(mdf) <- c("mean","median")
+cres <- cbind(mdf,cres)
+cres$FDR <- p.adjust(cres$PValue,method="fdr")
+toc()
+
+head(subset(cres,Direction=="Up"),25)
+head(subset(cres,Direction=="Down"),25)
+
+head(cres,50) %>%
+  kbl(caption = "Top significant genes with CAMERA") %>%
+  kable_paper("hover", full_width = F)
+
+sig <- subset(cres,FDR<0.05)
+SIG <- nrow(sig)
+
+plot(cres$median,-log10(cres$PValue))
+points(sig$median,-log10(sig$PValue),col="red")
+
+plot(cres$NGenes,-log10(cres$PValue),log="x")
+points(sig$NGenes,-log10(sig$PValue),col="red")
 
@@
 
-TODO
+ROAST is self contained test using rotations.
+
+## TODO
+
+The goal is to use a self contained test that can handle extensive correlation.
+
+Globaltest, ANCOVA and SAM-GS
+
+GSVA
+
+Q2 (Tian et al. [14])
+14.
+Tian L, Greenberg SA, Kong SW, et al.
+Discovering statistically significant pathways in expression profiling studies.
+Proc Natl Acad Sci USA.
+2005;102(38):13544–9.
+[PMC free article] [PubMed] [Google Scholar]
+
+FCS.SC—self-contained version of FCS (Pavlidis et al. [24])
+24.
+Pavlidis P, Qin J, Arango V, et al.
+Using the gene ontology for microarray data mining: a comparison of methods and application to age effect in human prefrontal cortex.
+Neurochem Res.
+2004;29(6):1213–22.
+[PubMed] [Google Scholar]
 

### 2023-09-14 old/GSE158433/GSE158433_integration.Rmd
refreshing hcy work

--- 2022-05-24 GSE158433/GSE158433_gmea.Rmd
@@
   library("kableExtra")
+  library("fgsea")
 })
@@
 source("../meth_functions.R")
-CORES=detectCores()/2
+CORES=8
 
@@
 anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
-myann <- data.frame(anno[,c("UCSC_RefGene_Name","Regulatory_Feature_Group","Islands_Name","Relation_to_Island")])
+myann <- data.frame(anno[,c("UCSC_RefGene_Name",
+  "Regulatory_Feature_Group","Islands_Name","Relation_to_Island")])
 promoters <- grep("Prom",myann$Regulatory_Feature_Group)
+
+```
+
+Gene probe sets
+
+```{r,sets}
+
+gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
+gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
+names(gp2) <- rownames(gp)
+sets <- split(rep(names(gp2), lengths(gp2)), unlist(gp2))
+
+hist(unlist(lapply(sets,length)),xlim=c(0,200),breaks=500,xlab="set size",main="probes per gene")
 
@@
 
-## GMEA whole gene
+## GMEA whole gene - wilcox test
 
@@
 res2_wg <- res2
+
+```
+
+Investigate how probe number influences significance.
+
+```{r,probebias}
+
+plot(res2$nprobes,res2$sig,log="x",ylim=c(0,50),pch=19,cex=0.6)
+points(sig$nprobes,sig$sig,col="red",pch=19,cex=0.62)
+MIN = min(sig$nprobes)
+LEFT = nrow(subset(res2,nprobes<MIN))
+RIGHT = nrow(subset(res2,nprobes>MIN))
+SIG = nrow(sig)
+TOT = nrow(res2)
+HEADER <- paste(TOT, "genes in total.", SIG, "with FDR<0.05.",
+  RIGHT, "well covered and", LEFT, "poorly covered")
+mtext(HEADER)
+abline(v=MIN,lty=2)
 

### 2023-09-14 old/gmea_camera.Rmd
refreshing hcy work

--- 2022-05-14 gmea.Rmd
@@
 ---
-title: "GSE74548: Gene Methylation Enrichment Analysis"
+title: "GSE74548: Gene Methylation Enrichment Analysis - accounting for correlation"
 author: "hcy_meth meta-analysis group"
@@
 enrichment of those probes.
-The statistical test used is the Wilcox test, which can be applied to competitive 
-or self contained test types.
+The statistical test used is the CAMERA test, which is a competitive test that attempts to account for
+correlation between genes in a set, or in this case probes belonging to a gene.
 
@@
 library("mitch")
+library("limma")
 library("IlluminaHumanMethylation450kmanifest")
@@
 
+CORES=detectCores()
+
 ann450k <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
@@
 promoters <- grep("Prom",myann$Regulatory_Feature_Group)
+
+# make a list of probes that belong to a gene
+gn <- unique(unlist(strsplit( myann$UCSC_RefGene_Name ,";")))
+gnl <- strsplit( myann$UCSC_RefGene_Name ,";")
+gnl <- mclapply(gnl,unique,mc.cores=CORES)
+myann$UCSC_RefGene_Name <- gnl
+
+gnl <- gnl[which(lapply(gnl,length)>0)]
+
+if ( ! file.exists("sets.Rds") ) {
+
+  l <- mclapply(1:nrow(myann), function(i) {
+    a <- myann[i,]
+    len <- length(a[[1]][[1]])
+    probe <- rep(rownames(a),len)
+    genes <- a[[1]][[1]]
+    data.frame(genes,probe)
+  },mc.cores=CORES)
+
+  df <- do.call(rbind,l)
+
+  sets <- mclapply(X=gn, FUN=function(g) {
+    df[which(df$genes == g),2]
+  } , mc.cores=CORES/2)
+
+  names(sets) <- gn
+
+  saveRDS(object=sets,file="sets.Rds")
+
+} else {
+
+  sets <- readRDS("sets.Rds")
+
+}
 
@@
 
+## Using CAMERA
+
+```{r,cam_func}
+
+stat <- dm$t
+names(stat) <- rownames(dm)
+
+# slow
+sets2 <- head(sets,1000)
+tic()
+cres <- cameraPR(statistic=stat, index=sets2, use.ranks = FALSE, inter.gene.cor=0.01, sort = TRUE)
+toc()
+
+# fast
+CORES=detectCores()
+#sets <- head(sets,100000)
+sets_split <- split(sets,1:CORES)
+
+tic()
+cres <- mclapply(X=sets_split, function(mysets) {
+  cameraPR(statistic=stat, index=mysets, use.ranks = FALSE, inter.gene.cor=0.01, sort = TRUE)
+}, mc.cores=CORES)
+cres <- do.call(rbind,cres)
+rownames(cres) <- sapply(strsplit(rownames(cres),"\\."),"[[",2)
+cres <- cres[order(cres$PValue),]
+g <- rownames(cres)[1]
+m <- mclapply(rownames(cres) , function(g) {
+  probes <- sets[[g]]
+  scores <- stat[which(names(stat) %in% probes)]
+  mymedian <- median(scores)
+  mymean <- mean(scores)
+  c(mymean,mymedian)
+},mc.cores=CORES)
+mdf <- do.call(rbind,m)
+colnames(mdf) <- c("mean","median")
+cres <- cbind(mdf,cres)
+toc()
+
+head(subset(cres,Direction=="Up"),25)
+head(subset(cres,Direction=="Down"),25)
+
+head(cres,50) %>%
+  kbl(caption = "Top significant genes with CAMERA") %>%
+  kable_paper("hover", full_width = F)
+
+```
+
+## CAMERA + pathway analysis
+
+There's no ES and the p-value is biased, so not sure about this
+
 ## Promoter analysis only

### 2023-09-14 figs/meth_functions.R

@@ -1,0 +1,573 @@
+suppressPackageStartupMessages({
+  library("plyr")
+  library("R.utils")
+  library("missMethyl")
+  library("limma")
+  library("DMRcate")
+  library("DMRcatedata")
+  library("topconfects")
+  library("minfi")
+  library("IlluminaHumanMethylation450kmanifest")
+  library("RColorBrewer")
+  library("IlluminaHumanMethylation450kanno.ilmn12.hg19")
+  library("GEOquery")
+  library("eulerr")
+  library("plyr")
+  library("gplots")
+  library("reshape2")
+  library("forestplot")
+  library("beeswarm")
+  library("RCircos")
+  library("qqman")
+  library("ENmix")
+})
+
+
+# scree plot shows the amount of variation in a dataset that is accounted
+# for by the first N principal components
+myscree <- function(mx,n=10,main="") {
+  pc <- princomp(mx)$sdev
+  pcp <- pc/sum(pc)*100
+  pcp <- pcp[1:10]
+  barplot(pcp,cex.names = 1,las=2,ylim=c(0,60),
+      ylab="percent (%) variance explained", main=main)
+  text((0.5:length(pcp)*1.2),pcp,label=signif(pcp,3),pos=3,cex=0.8)
+}
+
+
+# Here is a function to make a volcano plot
+make_volcano <- function(dm,name,mx) {
+    sig <- subset(dm,adj.P.Val<0.05)
+    N_SIG=nrow(sig)
+    N_UP=nrow(subset(sig,logFC>0))
+    N_DN=nrow(subset(sig,logFC<0))
+    HEADER=paste(N_SIG,"@5%FDR,", N_UP, "up", N_DN, "dn")
+    plot(dm$logFC,-log10(dm$P.Val),cex=0.5,pch=19,col="darkgray",
+        main=name, xlab="log FC", ylab="-log10 pval")
+    mtext(HEADER)
+    grid()
+    points(sig$logFC,-log10(sig$P.Val),cex=0.5,pch=19,col="red")
+}
+
+# Metal volcano
+volcano_metal <- function(metal,name) {
+  sig <- subset(metal,FDR<0.05)
+  N_SIG=nrow(sig)
+  N_UP=nrow(subset(sig,Zscore>0))
+  N_DN=nrow(subset(sig,Zscore<0))
+  HEADER=paste(N_SIG,"@5%FDR,", N_UP, "up", N_DN, "dn")
[…]

### 2023-09-14 meth_functions.R
remove ti

@@ -1,573 +1,0 @@
-suppressPackageStartupMessages({
-  library("plyr")
-  library("R.utils")
-  library("missMethyl")
-  library("limma")
-  library("DMRcate")
-  library("DMRcatedata")
-  library("topconfects")
-  library("minfi")
-  library("IlluminaHumanMethylation450kmanifest")
-  library("RColorBrewer")
-  library("IlluminaHumanMethylation450kanno.ilmn12.hg19")
-  library("GEOquery")
-  library("eulerr")
-  library("plyr")
-  library("gplots")
-  library("reshape2")
-  library("forestplot")
-  library("beeswarm")
-  library("RCircos")
-  library("qqman")
-  library("ENmix")
-})
-
-
-# scree plot shows the amount of variation in a dataset that is accounted
-# for by the first N principal components
-myscree <- function(mx,n=10,main="") {
-  pc <- princomp(mx)$sdev
-  pcp <- pc/sum(pc)*100
-  pcp <- pcp[1:10]
-  barplot(pcp,cex.names = 1,las=2,ylim=c(0,60),
-      ylab="percent (%) variance explained", main=main)
-  text((0.5:length(pcp)*1.2),pcp,label=signif(pcp,3),pos=3,cex=0.8)
-}
-
-
-# Here is a function to make a volcano plot
-make_volcano <- function(dm,name,mx) {
-    sig <- subset(dm,adj.P.Val<0.05)
-    N_SIG=nrow(sig)
-    N_UP=nrow(subset(sig,logFC>0))
-    N_DN=nrow(subset(sig,logFC<0))
-    HEADER=paste(N_SIG,"@5%FDR,", N_UP, "up", N_DN, "dn")
-    plot(dm$logFC,-log10(dm$P.Val),cex=0.5,pch=19,col="darkgray",
-        main=name, xlab="log FC", ylab="-log10 pval")
-    mtext(HEADER)
-    grid()
-    points(sig$logFC,-log10(sig$P.Val),cex=0.5,pch=19,col="red")
-}
-
-# Metal volcano
-volcano_metal <- function(metal,name) {
-  sig <- subset(metal,FDR<0.05)
-  N_SIG=nrow(sig)
-  N_UP=nrow(subset(sig,Zscore>0))
-  N_DN=nrow(subset(sig,Zscore<0))
-  HEADER=paste(N_SIG,"@5%FDR,", N_UP, "up", N_DN, "dn")
[…]

### 2023-09-14 figs/step_01b_test_gometh.Rmd
additions

@@
 sigup <- rownames(head(subset(dm,logFC>0),1000))
+sigdn <- rownames(head(subset(dm,logFC<0),1000))
+
+tic()
 gsaup <- gsameth(sig.cpg=head(sigup,500), all.cpg=rownames(dm), collection=gs_entrez)
+gsadn <- gsameth(sig.cpg=head(sigdn,500), all.cpg=rownames(dm), collection=gs_entrez)
+toc()
+
 gsaup <- gsaup[order(gsaup$P.DE),]
@@
 
-sigdn <- rownames(head(subset(dm,logFC<0),1000))
-gsadn <- gsameth(sig.cpg=head(sigdn,500), all.cpg=rownames(dm), collection=gs_entrez)
 gsadn <- gsadn[order(gsadn$P.DE),]
@@
 
+tic()
 mres <- mitch_calc(x=m, genesets=gs_symbols, priority="effect")
+toc()
 
@@
     }
-  } , mc.cores = CORES)
+  } , mc.cores = CORES/2)
   res_df <- do.call(rbind, res)
@@
 
-gse_res_df <- ttenrich(m=m,genesets=gs_symbols)
+tic()
+tres <- ttenrich(m=m,genesets=gs_symbols)
+toc()
 
-head(gse_res_df,30) %>%
+head(tres,30) %>%
   kbl(caption = "Top significant genesets") %>%
@@
 
-head(gse_res_df[order(-abs(gse_res_df$t_median)),],30) %>%
+head(tres[order(-abs(tres$t_median)),],30) %>%
   kbl(caption = "Top effect size genesets") %>%
@@
 
-sig <- subset(gse_res_df,fdr<0.05)
+sig <- subset(tres,fdr<0.05)
 nrow(sig)
+
+plot(tres$t_median,-log10(tres$p.value),pch=19,col="gray")
+grid()
+points(sig$t_median,-log10(sig$p.value),pch=19,col="red")
 
@@
 ```
+
+## TODO
+
+Find the overlap between mitch (competitive test) and ttest (self contained).
+
+Split data set into 2 halves to show consistency.
 

### 2023-09-14 figs/step_01b_test_gometh.Rmd
overlap

@@
 
+head(sig[order(-abs(sig$t_median)),],30) %>%
+  kbl(caption = "Top effect size genesets after FDR filtering") %>%
+  kable_paper("hover", full_width = F)
+
 plot(tres$t_median,-log10(tres$p.value),pch=19,col="gray")
@@
 
+## Overlap
+
+Use Euler diagrams to find the similarity.
+As medians are normally distributed about the zero, one would expect equal numbers of up and down-
+methylated gene sets.
+Therefore the t-test based approach could be considered better.
+
+```{r,ol1}
+
+msig_up <- subset(msig,s.dist>0)$set
+msig_dn <- subset(msig,s.dist<0)$set
+
+tsig_up <- rownames(subset(tsig,t_median>0))
+tsig_dn <- rownames(subset(tsig,t_median<0))
+
+v1 <- list("mitch up"=msig_up, "mitch dn"=msig_dn,
+  "ttest up"=tsig_up,"ttest dn"=tsig_dn)
+
+plot(euler(v1),quantities = TRUE)
+
+```
+
 ## TODO
-
-Find the overlap between mitch (competitive test) and ttest (self contained).
 

### 2023-09-14 figs/step_01c_splithcy.Rmd
working on split/analyse

--- 2023-09-14 figs/step_01b_test_gometh.Rmd
@@
 ---
-title: "GOmeth test"
+title: "Validation of GMEA"
 author: "hcy_meth meta-analysis group"
@@
 
-Here we will run test enrichment analysis approaches for undderstanding the effect of elevated
-homocysteine on blood methylomes.
-Homocysteine is thought to be an inhibitor of methyltransferases and hyperhomocysteine is
-associated with overall reduced genomic methylation.
-Previously we analysed the B-PROOF study data which enabled us to identify probes whose methylation correlatesor anti-correlated with homocysteine level.
-
-The two enrichment analysis approaches we will use are:
-
-* `gsameth`: an ORA approach that addressed the biases inherent with infinium array analysis.
-
-* `gmea`: an experimental application of functional class scoring for infinium analysis.
+Previously we demonstrated the ability of GMEA to detect differentially methylated pathways.
+There remains a question about how reliable these observations are.
+To resolve it, we will use the Hcy dataset which consists of 172 samples.
+We will split the data into two groups and analyse these separately.
+If the pathway results are similar, then we can be confident about the accuracy.
 
@@
 
-dm <- read.table("dma3a.tsv")
-rownames(dm) <- dm$Row.names
-dm$Row.names=NULL
+design <- readRDS("dm3_design.Rds")
+dim(design)
 
-head(dm, 10) %>% kbl() %>% kable_paper("hover", full_width = F)
+dm <- readRDS("dm3_mx.Rds")
+dim(dm)
+
+```
+
+Split the data.
+
+```{r,split1}
+
+g1 <- which(as.data.frame(design)$groups==1)
+g2 <- which(as.data.frame(design)$groups==2)
+g3 <- which(as.data.frame(design)$groups==3)
+
+seed=100
+
+set.seed(seed); g1s <- sample(g1,floor(length(g1)/2))
+g1b <- setdiff(g1,g1s)
+
+set.seed(seed); g2s <- sample(g2,floor(length(g2)/2))
+g2b <- setdiff(g2,g2s)
+
+set.seed(seed); g3s <- sample(g3,floor(length(g3)/2))
+g3b <- setdiff(g3,g3s)
+
+gxs <- c(g1s,g2s,g3s)
+gxs <- gxs[order(gxs)]
+
+gxb <- c(g1b,g2b,g3b)
+gxb <- gxb[order(gxb)]
+
+dess <- design[c(gxs),]
+dms <- dm[,c(gxs)]
+
+desb <- design[c(gxb),]
+dmb <- dm[,c(gxb)]
 
@@
 gs_symbols <- gmt_import("c2.cp.reactome.v2023.1.Hs.symbols.gmt")
-
-```
-
-## ORA with GSAMETH - parth of the MissMethyl package
-
-Now perform ORA analysis using gsameth with the top 500 probes.
-
-```{r,gsameth1}
-
-sigup <- rownames(head(subset(dm,logFC>0),1000))
-sigdn <- rownames(head(subset(dm,logFC<0),1000))
-
-tic()
-gsaup <- gsameth(sig.cpg=head(sigup,500), all.cpg=rownames(dm), collection=gs_entrez)
-gsadn <- gsameth(sig.cpg=head(sigdn,500), all.cpg=rownames(dm), collection=gs_entrez)
-toc()
-
-gsaup <- gsaup[order(gsaup$P.DE),]
-head(gsaup,10) %>% kbl(caption="Top hypermethylated pathways") %>% kable_paper("hover", full_width = F)
-
-gsadn <- gsadn[order(gsadn$P.DE),]
-head(gsadn,10) %>% kbl(caption="Top hypomethylated pathways") %>% kable_paper("hover", full_width = F)
 
@@
 hist(dm$t,breaks=seq(from=-6,to=6,by=1)) ; grid()
-
-```
-
-Aggregate probe t-scores to gene level scores.
-
-The object generated has two parts:
-
-1. `df`: a simple table of the gene name and the median tvalue
-
-2. `gme_res_df`: results of 1-sample Wilcox test.
-
-```{r,agg_wt}
-
-CORES= detectCores()
-
-agg <- function(dm) {
-  gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
-  gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
-  gnl <- mclapply(gnl,unique,mc.cores=CORES)
-  dm$UCSC_RefGene_Name <- gnl
-  l <- mclapply(1:nrow(dm), function(i) {
-    a <- dm[i,]
-    len <- length(a[[1]][[1]])
-    tvals <- as.numeric(rep(a[2],len))
-    genes <- a[[1]][[1]]
-    data.frame(genes,tvals)
-  },mc.cores=CORES)
-  df <- do.call(rbind,l)
-  gme_res <- mclapply( 1:length(gn), function(i) {
-    g <- gn[i]
-    tstats <- df[which(df$genes==g),"tvals"]
-    myn <- length(tstats)
-    mymean <- mean(tstats)
-    mymedian <- median(tstats)
-    wtselfcont <- wilcox.test(tstats)
-    res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,"median"=mymedian,
-      "p-value(sc)"=wtselfcont$p.value)
-  } , mc.cores=CORES )
-  gme_res_df <- do.call(rbind, gme_res)
-  rownames(gme_res_df) <- gme_res_df[,1]
-  gme_res_df <- gme_res_df[,-1]
-  tmp <- apply(gme_res_df,2,as.numeric)
-  rownames(tmp) <- rownames(gme_res_df)
-  gme_res_df <- as.data.frame(tmp)
-  gme_res_df$sig <- -log10(gme_res_df[,4])
-  gme_res_df <- gme_res_df[order(-gme_res_df$sig),]
-  gme_res_df$`fdr(sc)` <- p.adjust(gme_res_df$`p-value(sc)`)
-  out <- list("df"=df,"gme_res_df"=gme_res_df)
-  return(out)
-}
-
-tic()
-dmagg <- agg(dm)
-toc()
-head(dmagg$gme_res_df) %>% kbl(caption="top gmea genes (Wilcox)") %>% kable_paper("hover", full_width = F)
 
@@
 
-```
-
-This results shows that wilcox test executes faster but 1-sample t-test gives more significant findings.
-
-Check that the distribution is even.
-
-```{r,tmeddist}
-
 hist(dmagg2$gme_res_df$median,breaks=30) ; grid()
@@
 
-If the distribution is even, then established enrichment analysis can be performed.
-Options include FGSEA and mitch.
-
-Here I will try downstream analysis with mitch.
-
-```{r,mitch1}
-
-m <- dmagg2$gme_res_df
-
-m <- m[,"median",drop=FALSE]
-
-tic()
-mres <- mitch_calc(x=m, genesets=gs_symbols, priority="effect")
-toc()
-
-head(mres$enrichment_result,20) %>% kbl(caption="top mitch pathways") %>% kable_paper("hover", full_width = F)
-hist(mres$enrichment_result$s.dist) ; grid()
-
-tbl <- mres$enrichment_result
-
-sig <- subset(tbl,p.adjustANOVA<0.05)
-nrow(sig)
-
-plot(tbl$s.dist,-log10(tbl$pANOVA),pch=19,col="gray")
-grid()
-points(sig$s.dist,-log10(sig$pANOVA),pch=19,col="red")
-
-msig <- sig
-
-```
-
-Some interesting pathways found.
-
-There was no overlap with `gsameth()`.
-
-Interestingly many gene sets exhibit increased methylation.
-
-There could be a directional bias, so I will also conduct a 1-sample t-test.
+Conduct a 1-sample t-test.
 

### 2023-09-15 figs/step_01c_splithcy.Rmd
split+analyse

@@
 suppressPackageStartupMessages({
-  library("missMethyl")
+  library("mitch")
+  library("limma")
   library("kableExtra")
-  library("mitch")
   library("eulerr")
@@
 
-Load the data.
-
-```{r,load}
-
-design <- readRDS("dm3_design.Rds")
-dim(design)
-
-dm <- readRDS("dm3_mx.Rds")
-dim(dm)
-
-```
-
-Split the data.
-
-```{r,split1}
-
-g1 <- which(as.data.frame(design)$groups==1)
-g2 <- which(as.data.frame(design)$groups==2)
-g3 <- which(as.data.frame(design)$groups==3)
-
-seed=100
-
-set.seed(seed); g1s <- sample(g1,floor(length(g1)/2))
-g1b <- setdiff(g1,g1s)
-
-set.seed(seed); g2s <- sample(g2,floor(length(g2)/2))
-g2b <- setdiff(g2,g2s)
-
-set.seed(seed); g3s <- sample(g3,floor(length(g3)/2))
-g3b <- setdiff(g3,g3s)
-
-gxs <- c(g1s,g2s,g3s)
-gxs <- gxs[order(gxs)]
-
-gxb <- c(g1b,g2b,g3b)
-gxb <- gxb[order(gxb)]
-
-dess <- design[c(gxs),]
-dms <- dm[,c(gxs)]
-
-desb <- design[c(gxb),]
-dmb <- dm[,c(gxb)]
-
-```
+## Load gene sets
 
@@
 
-## GMEA
-
-Now with GMEA.
-
-Start with gathering the annotation set.
+## Load annotation data
 
@@
 
-Histogram of limma t-statistics.
+## Load functions for GMEA
 
-```{r,hist}
+```{r,gmeafunc}
 
-dm <- dm[,c("UCSC_RefGene_Name","t")]
-hist(dm$t,breaks=seq(from=-6,to=6,by=1)) ; grid()
-
-```
-
-Does 1-sample t-test work better?
-
-```{r,agg_tt}
-
-agg2 <- function(dm) {
+agg <- function(dm,cores=1) {
   gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
   gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
-  gnl <- mclapply(gnl,unique,mc.cores=CORES)
+  gnl <- mclapply(gnl,unique,mc.cores=cores)
   dm$UCSC_RefGene_Name <- gnl
@@
     data.frame(genes,tvals)
-  },mc.cores=CORES)
+  },mc.cores=cores)
   df <- do.call(rbind,l)
@@
 
-tic()
-dmagg2 <- agg2(dm)
-toc()
-head(dmagg2$gme_res_df) %>% kbl(caption="top gmea genes (t-test)") %>% kable_paper("hover", full_width = F)
-
-hist(dmagg2$gme_res_df$median,breaks=30) ; grid()
-
-```
-
-Conduct a 1-sample t-test.
-
-```{r,t_enrichment_test1}
-
-ttenrich <- function(m,genesets) {
+ttenrich <- function(m,genesets,cores=1) {
   res <- mclapply( 1:length(genesets), function(i) {
@@
     }
-  } , mc.cores = CORES/2)
+  } , mc.cores = cores)
   res_df <- do.call(rbind, res)
@@
 
-tic()
-tres <- ttenrich(m=m,genesets=gs_symbols)
-toc()
+```
 
-head(tres,30) %>%
-  kbl(caption = "Top significant genesets") %>%
-  kable_paper("hover", full_width = F)
+## Load methylation data
 
-head(tres[order(-abs(tres$t_median)),],30) %>%
-  kbl(caption = "Top effect size genesets") %>%
-  kable_paper("hover", full_width = F)
+Load the data.
 
+```{r,load}
+
+design <- as.data.frame(readRDS("dm3_design.Rds"))
+dim(design)
+
+mx <- readRDS("dm3_mx.Rds")
+dim(mx)
+
+```
+
+## Split data
+
+Split the data into two smaller groups (s and b).
+
+```{r,split1}
+
+g1 <- which(as.data.frame(design)$groups==1)
+g2 <- which(as.data.frame(design)$groups==2)
+g3 <- which(as.data.frame(design)$groups==3)
+
+seed=100
+
+set.seed(seed); g1s <- sample(g1,floor(length(g1)/2))
+g1b <- setdiff(g1,g1s)
+
+set.seed(seed); g2s <- sample(g2,floor(length(g2)/2))
+g2b <- setdiff(g2,g2s)
+
+set.seed(seed); g3s <- sample(g3,floor(length(g3)/2))
+g3b <- setdiff(g3,g3s)
+
+gxs <- c(g1s,g2s,g3s)
+gxs <- gxs[order(gxs)]
+
+gxb <- c(g1b,g2b,g3b)
+gxb <- gxb[order(gxb)]
+
+dess <- design[c(gxs),]
+mxs <- mx[,c(gxs)]
+
+desb <- design[c(gxb),]
+mxb <- mx[,c(gxb)]
+
+```
+
+## Limma + GMEA
+
+Using the S dataset, apply limma to conduct differential methylation analysis,
+then GMEA.
+GMEA has two steps:
+
+1. Aggregate probe t-statistics to genes.
+
+2. Perform enrichment test using 1 sample t-test.
+
+```{r,limmagmea1}
+
+d <- model.matrix(~dess$age+ dess$sex + dess$groups)
+
+fit.reduced <- lmFit(mxs,d)
+fit.reduced <- eBayes(fit.reduced)
+dm <- topTable(fit.reduced,coef=4, number = Inf)
+dma <- merge(myann,dm,by=0)
+dma <- dma[order(dma$P.Value),]
+dma <- dma[,c("UCSC_RefGene_Name","t")]
+
+dmagg <- agg(dma,cores=8)
+m <- dmagg$gme_res_df
+m <- m[,"median",drop=FALSE]
+tres <- ttenrich(m=m,genesets=gs_symbols,cores=8)
 sig <- subset(tres,fdr<0.05)
-nrow(sig)
+ssig <- sig
 
-head(sig[order(-abs(sig$t_median)),],30) %>%
-  kbl(caption = "Top effect size genesets after FDR filtering") %>%
-  kable_paper("hover", full_width = F)
+```
 
-plot(tres$t_median,-log10(tres$p.value),pch=19,col="gray")
-grid()
-points(sig$t_median,-log10(sig$p.value),pch=19,col="red")
+Now repeat with the b set.
 
-tsig <- sig
+```{r,limmagmea2}
+
+d <- model.matrix(~desb$age+ desb$sex + desb$groups)
+
+fit.reduced <- lmFit(mxb,d)
+fit.reduced <- eBayes(fit.reduced)
+dm <- topTable(fit.reduced,coef=4, number = Inf)
+dma <- merge(myann,dm,by=0)
+dma <- dma[order(dma$P.Value),]
+dma <- dma[,c("UCSC_RefGene_Name","t")]
+
+dmagg <- agg(dma,cores=8)
+m <- dmagg$gme_res_df
+m <- m[,"median",drop=FALSE]
+tres <- ttenrich(m=m,genesets=gs_symbols,cores=8)
+sig <- subset(tres,fdr<0.05)
+bsig <- sig
 
@@
 
-msig_up <- subset(msig,s.dist>0)$set
-msig_dn <- subset(msig,s.dist<0)$set
+ssig_up <- rownames(subset(ssig,t_median>0))
+ssig_dn <- rownames(subset(ssig,t_median<0))
 
-tsig_up <- rownames(subset(tsig,t_median>0))
-tsig_dn <- rownames(subset(tsig,t_median<0))
+bsig_up <- rownames(subset(bsig,t_median>0))
+bsig_dn <- rownames(subset(bsig,t_median<0))
 
-v1 <- list("mitch up"=msig_up, "mitch dn"=msig_dn,
-  "ttest up"=tsig_up,"ttest dn"=tsig_dn)
+v1 <- list("s up"=ssig_up, "s dn"=ssig_dn,
+  "b up"=bsig_up,"b dn"=bsig_dn)
 
@@
 
-Split data set into 2 halves to show consistency.
+* Make sure it is all working properly.
+
+* Repeat for other seed values and aggregate the results.
 

### 2023-09-15 figs/step_01a_GSE74548_prep.Rmd
rownames for design matrix

@@
 saveRDS(mxs,"dm3_mx.Rds")
+rownames(design) <- rownames(samplesheet)
+head(design)
 saveRDS(design,"dm3_design.Rds")

### 2023-09-15 figs/step_01c_splithcy.Rmd
split and analyse

@@
     mymedian <- median(tstats)
-    tselfcont <- t.test(tstats)
+    tselfcont <- wilcox.test(tstats)
     res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,"median"=mymedian,
@@
       mymedian <- median(tstats)
-      wt <- t.test(tstats)
+      wt <- wilcox.test(tstats)
       res <- c(name,myn,mymean,mymedian,wt$p.value)
@@
 
-## Split data
+## Split, limma, GMEA
 
@@
 
-```{r,split1}
-
-g1 <- which(as.data.frame(design)$groups==1)
-g2 <- which(as.data.frame(design)$groups==2)
-g3 <- which(as.data.frame(design)$groups==3)
-
-seed=100
-
-set.seed(seed); g1s <- sample(g1,floor(length(g1)/2))
-g1b <- setdiff(g1,g1s)
-
-set.seed(seed); g2s <- sample(g2,floor(length(g2)/2))
-g2b <- setdiff(g2,g2s)
-
-set.seed(seed); g3s <- sample(g3,floor(length(g3)/2))
-g3b <- setdiff(g3,g3s)
-
-gxs <- c(g1s,g2s,g3s)
-gxs <- gxs[order(gxs)]
-
-gxb <- c(g1b,g2b,g3b)
-gxb <- gxb[order(gxb)]
-
-dess <- design[c(gxs),]
-mxs <- mx[,c(gxs)]
-
-desb <- design[c(gxb),]
-mxb <- mx[,c(gxb)]
-
-```
-
-## Limma + GMEA
-
-Using the S dataset, apply limma to conduct differential methylation analysis,
+Using the S and B datasets separately, apply limma to conduct differential methylation analysis,
 then GMEA.
@@
 
-```{r,limmagmea1}
+```{r,splitanalysis}
 
+g1 <- which(as.data.frame(design)$groups==1)
+g2 <- which(as.data.frame(design)$groups==2)
+g3 <- which(as.data.frame(design)$groups==3)
+
+split_analyse <- function(seed){
+  set.seed(seed); g1s <- sample(g1,floor(length(g1)/2))
+  g1b <- setdiff(g1,g1s)
+  set.seed(seed); g2s <- sample(g2,floor(length(g2)/2))
+  g2b <- setdiff(g2,g2s)
+  set.seed(seed); g3s <- sample(g3,floor(length(g3)/2))
+  g3b <- setdiff(g3,g3s)
+  gxs <- c(g1s,g2s,g3s)
+  gxs <- gxs[order(gxs)]
+  gxb <- c(g1b,g2b,g3b)
+  gxb <- gxb[order(gxb)]
+  dess <- design[gxs,]
+  mxs <- mx[,gxs]
+  desb <- design[gxb,]
+  mxb <- mx[,gxb]
+
+  #s dataset
   d <- model.matrix(~dess$age+ dess$sex + dess$groups)
-
   fit.reduced <- lmFit(mxs,d)
@@
   dma <- dma[,c("UCSC_RefGene_Name","t")]
-
   dmagg <- agg(dma,cores=8)
@@
   tres <- ttenrich(m=m,genesets=gs_symbols,cores=8)
-sig <- subset(tres,fdr<0.05)
+  sig <- subset(tres,fdr<0.01)
   ssig <- sig
 
-```
-
-Now repeat with the b set.
-
-```{r,limmagmea2}
-
+  #b dataset
   d <- model.matrix(~desb$age+ desb$sex + desb$groups)
-
   fit.reduced <- lmFit(mxb,d)
@@
   tres <- ttenrich(m=m,genesets=gs_symbols,cores=8)
-sig <- subset(tres,fdr<0.05)
+  sig <- subset(tres,fdr<0.01)
   bsig <- sig
+
+  return(list("ssig"=ssig,"bsig"=bsig))
+}
 
@@
 
+ol <- function(res){
+  ssig <- res$ssig
+  bsig <- res$bsig
   ssig_up <- rownames(subset(ssig,t_median>0))
   ssig_dn <- rownames(subset(ssig,t_median<0))
-
   bsig_up <- rownames(subset(bsig,t_median>0))
   bsig_dn <- rownames(subset(bsig,t_median<0))
-
   v1 <- list("s up"=ssig_up, "s dn"=ssig_dn,
     "b up"=bsig_up,"b dn"=bsig_dn)
+  plot(euler(v1),quantities = TRUE)
+}
 
-plot(euler(v1),quantities = TRUE)
+res <- split_analyse(seed=200)
+ol(res=res)
 

### 2023-09-20 figs/step_01x_gmea2.Rmd
different approaches for fcs

---
title: "Test different aggregation approaches"
author: "hcy_meth meta-analysis group"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
---

## Introduction

In this analysis I will try out an approach that preserved per-sample information.
It aggregates methylaiton values for all CpGs into a single median value.

```{r,libs}

suppressPackageStartupMessages({
  library("missMethyl")
  library("kableExtra")
  library("mitch")
  library("eulerr")
  library("IlluminaHumanMethylation450kmanifest")
  library("IlluminaHumanMethylation450kanno.ilmn12.hg19")
  library("tictoc")
  library("limma")
})

CORES=12

```

## Annotation data

Start with gathering the annotation set.

```{r,anno}

ann450k <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
myann <- data.frame(ann450k[,c("UCSC_RefGene_Name","Regulatory_Feature_Group")])
promoters <- grep("Prom",myann$Regulatory_Feature_Group)

```

## Load methylation data

```{r,load}

mx <- readRDS("dm3_mx.Rds")
design <- readRDS("dm3_design.Rds")

```

## Load pathways

Reactome pathways were downloaded on the 14th Sept 2023 from MsigDB.

```{r,gsdl}

gs_entrez <- gmt_import("c2.cp.reactome.v2023.1.Hs.entrez.gmt")

gs_symbols <- gmt_import("c2.cp.reactome.v2023.1.Hs.symbols.gmt")

```

## Join

Determine the median of probe values

```{r,med1}

gn <- unique(unlist(strsplit( myann$UCSC_RefGene_Name ,";")))
gnl <- strsplit( myann$UCSC_RefGene_Name ,";")
gnl <- mclapply(gnl,unique,mc.cores=CORES)
myann$gnl <- gnl

mymed <- function(g) {
  probes <- rownames(myann[grep(g,myann$gnl),])
  rows <- which(rownames(mx) %in% probes)
  if ( length(rows) > 1 ) {
    b <- mx[rows,]
    med <- apply(b,2,median)
    med <- matrix(med,nrow=1)
    colnames(med) <- colnames(b)
    rownames(med) <- g
    return(med)
  }
}

med <- mclapply(gn,mymed,mc.cores=CORES)
med <- med[lapply(med,length)>0]
medf <- do.call(rbind,med)

```

Now aggregate by median for gene sets.

```{r,med2}

i=1
mymed2 <- function(mx,gsets,i) {
  nm <- names(gsets)[i]
  gs <- gsets[[i]]
  med <- matrix(apply(mx[which(rownames(mx) %in% gs),],2,median),nrow=1)
  colnames(med) <- colnames(mx)
  rownames(med) <- nm
  return(med)
}

med2 <- mclapply(1:length(gs_symbols), function(i) {
  mymed2(mx=medf,gsets=gs_symbols,i=i)
},mc.cores=CORES)

medf2 <- do.call(rbind,med2)

```

## Limma differential methylation

Then do a limma test.
See if it is more robust than other approaches.

```{r,limma1}

fit.reduced <- lmFit(medf,design)
fit.reduced <- eBayes(fit.reduced)
summary(decideTests(fit.reduced))
dm <- topTable(fit.reduced,coef=4, number = Inf)

#write.table(dma3a,file="dma3a.tsv")

head(dm, 30) %>% kbl() %>% kable_paper("hover", full_width = F)

```

## Mitch

```{r,mitchtest}

y <- mitch_import(dm, DEtype="limma")

res <- mitch_calc(y, genesets=gs_symbols, priority="effect",cores=CORES)

sig <- subset(res$enrichment_result,p.adjustANOVA<0.05)

nrow(subset(sig,s.dist>0))

nrow(subset(sig,s.dist<0))

head(subset(sig,s.dist>0),20) %>%
  kbl(caption="hypermethylated pathways") %>%
  kable_paper("hover", full_width = F)

head(subset(sig,s.dist<0),20) %>%
  kbl(caption="hypomethylated pathways") %>%
  kable_paper("hover", full_width = F)

```

## 1-sample t-test

```{r,ttest}

ttenrich <- function(m,genesets,cores=1) {
  res <- mclapply( 1:length(genesets), function(i) {
    gs <- genesets[i]
    name <- names(gs)
    n_members <- length(which(rownames(m) %in% gs[[1]]))
    if ( n_members > 4 ) {
      tstats <- m[which(rownames(m) %in% gs[[1]]),]
      myn <- length(tstats)
      mymean <- mean(tstats)
      mymedian <- median(tstats)
      wt <- t.test(tstats)
      res <- c(name,myn,mymean,mymedian,wt$p.value)
    }
  } , mc.cores = cores)
  res_df <- do.call(rbind, res)
  rownames(res_df) <- res_df[,1]
  res_df <- res_df[,-1]
  colnames(res_df) <- c("n_genes","t_mean","t_median","p.value")
  tmp <- apply(res_df,2,as.numeric)
  rownames(tmp) <- rownames(res_df)
  res_df <- tmp
  res_df <- as.data.frame(res_df)
  res_df <- res_df[order(res_df$p.value),]
  res_df$logp <- -log10(res_df$p.value )
  res_df$fdr <- p.adjust(res_df$p.value,method="fdr")
  res_df[order(abs(res_df$p.value)),]
  return(res_df)
}

m <- as.data.frame(dm$t)
rownames(m) <- rownames(dm)
colnames(m) <- "t"

tres <- ttenrich(m=m,genesets=gs_symbols,cores=CORES)

sig <- subset(tres,fdr<0.05)

nrow(sig)

nrow(subset(sig,t_median>0))

head(sig[order(-sig$t_median),],20) %>%
  kbl(caption="hypermethylated pathways") %>%
  kable_paper("hover", full_width = F)

nrow(subset(sig,t_median<0))

head(sig[order(sig$t_median),],20) %>%
  kbl(caption="hypomethylated pathways") %>%
  kable_paper("hover", full_width = F)

```

## Todo

Compare mitch and t-test approaches.

## Session information

For reproducibility.

```{r,session}

sessionInfo()

```

### 2023-09-25 figs/step_01b_test_gometh.Rmd
limma-agg-ttest approach

--- 2023-09-14 figs/step_01b_test_gometh.Rmd
@@
 
-Here we will run test enrichment analysis approaches for undderstanding the effect of elevated
+Here we will run test enrichment analysis approaches for understanding the effect of elevated
 homocysteine on blood methylomes.
@@
   library("mitch")
+  library("limma")
   library("eulerr")
@@
 })
+
+CORES=12
 
@@
 
+gsig_up <-  subset(gsaup,FDR<0.1)
+gsig_dn <-  subset(gsadn,FDR<0.1)
+
 ```
@@
 Now with GMEA.
+There are two approaches:
+
+1. limma-agg-1stt (la1).
+In this approach, the limma test is conducted on probes, followed by aggregation of the differential
+methylation t-scores. Each gene gets a median t-stat value which is evaluated downstream by a 1 sample
+t-test.
+
+2. agg-limma-1stt (al1).
+Here, the probe beta values for each gene are aggregated to a median and then limma is run on these
+values.
+To determine gene set enrichment, a 1-sample t-test is conducted.
 
@@
 
-Histogram of limma t-statistics.
+Check histogram of limma t-statistics.
 
@@
 
-Aggregate probe t-scores to gene level scores.
+## LA: limma-aggregate approach
+
+Aggregate probe t-scores to gene level scores and conduct a 1-sample t-test on the gene level.
 
@@
 
-```{r,agg_wt}
+```{r,agg_tt}
 
-CORES= detectCores()
-
-agg <- function(dm) {
+agg2 <- function(dm,cores=1) {
   gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
   gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
-  gnl <- mclapply(gnl,unique,mc.cores=CORES)
-  dm$UCSC_RefGene_Name <- gnl
-  l <- mclapply(1:nrow(dm), function(i) {
-    a <- dm[i,]
-    len <- length(a[[1]][[1]])
-    tvals <- as.numeric(rep(a[2],len))
-    genes <- a[[1]][[1]]
-    data.frame(genes,tvals)
-  },mc.cores=CORES)
-  df <- do.call(rbind,l)
-  gme_res <- mclapply( 1:length(gn), function(i) {
-    g <- gn[i]
-    tstats <- df[which(df$genes==g),"tvals"]
-    myn <- length(tstats)
-    mymean <- mean(tstats)
-    mymedian <- median(tstats)
-    wtselfcont <- wilcox.test(tstats)
-    res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,"median"=mymedian,
-      "p-value(sc)"=wtselfcont$p.value)
-  } , mc.cores=CORES )
-  gme_res_df <- do.call(rbind, gme_res)
-  rownames(gme_res_df) <- gme_res_df[,1]
-  gme_res_df <- gme_res_df[,-1]
-  tmp <- apply(gme_res_df,2,as.numeric)
-  rownames(tmp) <- rownames(gme_res_df)
-  gme_res_df <- as.data.frame(tmp)
-  gme_res_df$sig <- -log10(gme_res_df[,4])
-  gme_res_df <- gme_res_df[order(-gme_res_df$sig),]
-  gme_res_df$`fdr(sc)` <- p.adjust(gme_res_df$`p-value(sc)`)
-  out <- list("df"=df,"gme_res_df"=gme_res_df)
-  return(out)
-}
-
-tic()
-dmagg <- agg(dm)
-toc()
-head(dmagg$gme_res_df) %>% kbl(caption="top gmea genes (Wilcox)") %>% kable_paper("hover", full_width = F)
-
-```
-
-Does 1-sample t-test work better?
-
-```{r,agg_tt}
-
-agg2 <- function(dm) {
-  gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
-  gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
-  gnl <- mclapply(gnl,unique,mc.cores=CORES)
+  gnl <- mclapply(gnl,unique,mc.cores=cores)
   dm$UCSC_RefGene_Name <- gnl
@@
 tic()
-dmagg2 <- agg2(dm)
+dmagg2 <- agg2(dm,cores=CORES)
 toc()
-head(dmagg2$gme_res_df) %>% kbl(caption="top gmea genes (t-test)") %>% kable_paper("hover", full_width = F)
+
+head(dmagg2$gme_res_df) %>%
+  kbl(caption="top gmea genes (t-test)") %>%
+  kable_paper("hover", full_width = F)
 
 ```
-
-This results shows that wilcox test executes faster but 1-sample t-test gives more significant findings.
 
@@
 
-If the distribution is even, then established enrichment analysis can be performed.
-Options include FGSEA and mitch.
-
-Here I will try downstream analysis with mitch.
-
-```{r,mitch1}
-
-m <- dmagg2$gme_res_df
-
-m <- m[,"median",drop=FALSE]
-
-tic()
-mres <- mitch_calc(x=m, genesets=gs_symbols, priority="effect")
-toc()
-
-head(mres$enrichment_result,20) %>% kbl(caption="top mitch pathways") %>% kable_paper("hover", full_width = F)
-hist(mres$enrichment_result$s.dist) ; grid()
-
-tbl <- mres$enrichment_result
-
-sig <- subset(tbl,p.adjustANOVA<0.05)
-nrow(sig)
-
-plot(tbl$s.dist,-log10(tbl$pANOVA),pch=19,col="gray")
-grid()
-points(sig$s.dist,-log10(sig$pANOVA),pch=19,col="red")
-
-msig <- sig
-
-```
-
-Some interesting pathways found.
-
-There was no overlap with `gsameth()`.
-
-Interestingly many gene sets exhibit increased methylation.
-
-There could be a directional bias, so I will also conduct a 1-sample t-test.
+1-sample t-test for gene set enrichment.
 
@@
 
-ttenrich <- function(m,genesets) {
+ttenrich <- function(m,genesets,cores=1) {
   res <- mclapply( 1:length(genesets), function(i) {
@@
 
+# need to get the median column before analysis
+m <- dmagg2$gme_res_df[,"median",drop=FALSE]
+
 tic()
-tres <- ttenrich(m=m,genesets=gs_symbols)
+lares <- ttenrich(m=m,genesets=gs_symbols)
 toc()
 
-head(tres,30) %>%
-  kbl(caption = "Top significant genesets") %>%
+head(lares,30) %>%
+  kbl(caption = "Top significant genesets using LA approach") %>%
   kable_paper("hover", full_width = F)
 
-head(tres[order(-abs(tres$t_median)),],30) %>%
-  kbl(caption = "Top effect size genesets") %>%
+head(lares[order(-abs(lares$t_median)),],30) %>%
+  kbl(caption = "Top effect size genesets using LA approach") %>%
   kable_paper("hover", full_width = F)
 
-sig <- subset(tres,fdr<0.05)
-nrow(sig)
+lasig <- subset(lares,fdr<0.05)
+nrow(lasig)
 
-plot(tres$t_median,-log10(tres$p.value),pch=19,col="gray")
+head(lasig[order(-abs(lasig$t_median)),],30) %>%
+  kbl(caption = "Top effect size genesets after FDR filtering using LA approach") %>%
+  kable_paper("hover", full_width = F)
+
+plot(lares$t_median,-log10(lares$p.value),pch=19,col="gray")
 grid()
-points(sig$t_median,-log10(sig$p.value),pch=19,col="red")
-
-tsig <- sig
+points(lasig$t_median,-log10(lasig$p.value),pch=19,col="red")
 
@@
 
-## TODO
+## AL: Aggregate limma t-test
 
-Find the overlap between mitch (competitive test) and ttest (self contained).
+Determine the median of probe values
 
-Split data set into 2 halves to show consistency.
+```{r,med1}
+
+gn <- unique(unlist(strsplit( myann$UCSC_RefGene_Name ,";")))
+gnl <- strsplit( myann$UCSC_RefGene_Name ,";")
+gnl <- mclapply(gnl,unique,mc.cores=CORES)
+myann$gnl <- gnl
+
+mymed <- function(g) {
+  probes <- rownames(myann[grep(g,myann$gnl),])
+  rows <- which(rownames(mx) %in% probes)
+  if ( length(rows) > 1 ) {
+    b <- mx[rows,]
+    med <- apply(b,2,median)
+    med <- matrix(med,nrow=1)
+    colnames(med) <- colnames(b)
+    rownames(med) <- g
+    return(med)
+  }
+}
+
+mx <- readRDS("dm3_mx.Rds")
+design <- readRDS("dm3_design.Rds")
+
+med <- mclapply(gn,mymed,mc.cores=CORES)
+med <- med[lapply(med,length)>0]
+medf <- do.call(rbind,med)
+
+```
+
+Then do a limma test on median gene methylation.
+See if it is more robust than other approaches.
+
+```{r,limma1}
+
+fit.reduced <- lmFit(medf,design)
+fit.reduced <- eBayes(fit.reduced)
+summary(decideTests(fit.reduced))
+dm <- topTable(fit.reduced,coef=4, number = Inf)
+
+#write.table(dma3a,file="dma3a.tsv")
+
+head(dm, 30) %>%
+  kbl(caption="limma gene methylation results") %>%
+  kable_paper("hover", full_width = F)
+
+```
+
+## 1-sample t-test
+
+```{r,ttest}
+
+ttenrich <- function(m,genesets,cores=1) {
+  res <- mclapply( 1:length(genesets), function(i) {
+    gs <- genesets[i]
+    name <- names(gs)
+    n_members <- length(which(rownames(m) %in% gs[[1]]))
+    if ( n_members > 4 ) {
+      tstats <- m[which(rownames(m) %in% gs[[1]]),]
+      myn <- length(tstats)
+      mymean <- mean(tstats)
+      mymedian <- median(tstats)
+      wt <- t.test(tstats)
+      res <- c(name,myn,mymean,mymedian,wt$p.value)
+    }
+  } , mc.cores = cores)
+  res_df <- do.call(rbind, res)
+  rownames(res_df) <- res_df[,1]
+  res_df <- res_df[,-1]
+  colnames(res_df) <- c("n_genes","t_mean","t_median","p.value")
+  tmp <- apply(res_df,2,as.numeric)
+  rownames(tmp) <- rownames(res_df)
+  res_df <- tmp
+  res_df <- as.data.frame(res_df)
+  res_df <- res_df[order(res_df$p.value),]
+  res_df$logp <- -log10(res_df$p.value )
+  res_df$fdr <- p.adjust(res_df$p.value,method="fdr")
+  res_df[order(abs(res_df$p.value)),]
+  return(res_df)
+}
+
+m <- as.data.frame(dm$t)
+rownames(m) <- rownames(dm)
+colnames(m) <- "t"
+
+alres <- ttenrich(m=m,genesets=gs_symbols,cores=CORES)
+
+alsig <- subset(alres,fdr<0.05)
+
+nrow(alsig)
+
+nrow(subset(alsig,t_median>0))
+
+head(alsig[order(-alsig$t_median),],20) %>%
+  kbl(caption="hypermethylated pathways") %>%
+  kable_paper("hover", full_width = F)
+
+nrow(subset(alsig,t_median<0))
+
+head(alsig[order(alsig$t_median),],20) %>%
+  kbl(caption="hypomethylated pathways") %>%
+  kable_paper("hover", full_width = F)
+
+```
+
+## Overlap
+
+Use Euler diagrams to find the similarity.
+As medians are normally distributed about the zero, one would expect equal numbers of up and down-
+methylated gene sets.
+Therefore the t-test based approach could be considered better.
+
+```{r,ol1}
+
+lasig_up <- rownames(subset(lasig,t_median>0))
+lasig_dn <- rownames(subset(lasig,t_median<0))
+
+alsig_up <- rownames(subset(alsig,t_median>0))
+alsig_dn <- rownames(subset(alsig,t_median<0))
+
+v1 <- list("LA up"=lasig_up, "LA dn"=lasig_dn,
+  "AL up"=alsig_up,"AL dn"=alsig_dn,
+  "GSAmeth up"=gsig_up,"GSAmeth dn"=gsig_dn)
+
+plot(euler(v1),quantities = TRUE)
+
+```
 

### 2023-09-25 figs/step_01x_gmea2.Rmd
limma-agg-ttest approach

@@
 
-sig <- subset(res$enrichment_result,p.adjustANOVA<0.05)
+msig <- subset(res$enrichment_result,p.adjustANOVA<0.05)
 
-nrow(subset(sig,s.dist>0))
+nrow(subset(msig,s.dist>0))
 
-nrow(subset(sig,s.dist<0))
+nrow(subset(msig,s.dist<0))
 
-head(subset(sig,s.dist>0),20) %>%
+head(subset(msig,s.dist>0),20) %>%
   kbl(caption="hypermethylated pathways") %>%
@@
 
-head(subset(sig,s.dist<0),20) %>%
+head(subset(msig,s.dist<0),20) %>%
   kbl(caption="hypomethylated pathways") %>%
@@
 
-sig <- subset(tres,fdr<0.05)
+tsig <- subset(tres,fdr<0.05)
 
-nrow(sig)
+nrow(tsig)
 
-nrow(subset(sig,t_median>0))
+nrow(subset(tsig,t_median>0))
 
-head(sig[order(-sig$t_median),],20) %>%
+head(sig[order(-tsig$t_median),],20) %>%
   kbl(caption="hypermethylated pathways") %>%
@@
 
-nrow(subset(sig,t_median<0))
+nrow(subset(tsig,t_median<0))
 
-head(sig[order(sig$t_median),],20) %>%
+head(sig[order(tsig$t_median),],20) %>%
   kbl(caption="hypomethylated pathways") %>%
@@
 
-## Todo
+## Compare mitch and t-test approaches.
 
-Compare mitch and t-test approaches.
+```{r,compare1}
+
+save.image("step_01x_gmea2.Rdata")
+
+tt <- tres[,c("t_median","fdr"),drop=FALSE]
+
+mres <- res$enrichment_result
+mm <- mres[,c("s.dist","p.adjustANOVA"),drop=FALSE]
+rownames(mm) <- mres$set
+
+zz <- merge(tt,mm,by=0)
+rownames(zz) <- zz$Row.names
+zz$Row.names=NULL
+
+# extract rows with t sig
+ts <- subset(zz,fdr<0.05)
+
+# extract rows with m sig
+ms <- subset(zz,p.adjustANOVA<0.05)
+
+# extract rows where both are significant
+ss <- subset(zz,p.adjustANOVA<0.05 & fdr<0.05)
+
+plot(zz$t_median,zz$s.dist,pch=19,col="gray",
+  xlab="median",ylab="s.dist") ; grid()
+points(ms$t_median,ms$s.dist,pch=19,cex=1.1,col="red")
+points(ts$t_median,ts$s.dist,pch=19,cex=1.1,col="blue")
+points(ss$t_median,ss$s.dist,pch=19,cex=1.2,col="purple")
+mtext("red:sig t-test, blue:sig mitch, purple:sig both",cex=1.2)
+
+```
+
+Euler diagram.
+
+```{r,euler1}
+
+msig_up <- subset(msig,s.dist>0)$set
+msig_dn <- subset(msig,s.dist<0)$set
+
+tsig_up <- rownames(subset(tsig,t_median>0))
+tsig_dn <- rownames(subset(tsig,t_median<0))
+
+ll <- list("m up"=msig_up,"m dn"=msig_dn,"t up"=tsig_up,"t dn"=tsig_dn)
+
+plot(euler(ll),quantities = TRUE)
+
+```
 
@@
 
+save.image("step_01x_gmea2.Rdata")
+
 sessionInfo()

### 2023-09-25 figs/GSE158422_gmea.Rmd
save for later

---
title: "GSE158422: DNA methylation in Lung squamous cell carcinoma - gene methylation enrichment analysis"
author: "The GMEA team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
---

## Introduction

Functional class sorting is widely used for pathway analysis using tools like GSEA,
yet there is no consensus in how this can be conducted for Illumina Infinium
methylation array data.

Here we propose a simple approach which involves the following:

1. Limma test on probes.

2. For each gene, calculate the median t-statistic from step 1.

3. Use this median t-stat in Camera pre-ranked test for gene sets.

In this example, I'm using matched infinium EPIC 850k data from (n=37) normal and lung cancer samples (GSE158422).
The data was previously preprocessed and normalised using the minfi package (see the folder called "misc").

Here the gene sets are obtained from Reactome.

## Requirements

This analysis was run on a 8C/16T computer with 64GB RAM running at 3.8 GHz.
This workflow used 34 GB RAM and took
RAM usage can be moderated by reducing the parallel core count.

```{r,packages}

suppressPackageStartupMessages({
  library("limma")
  library("eulerr")
  library("IlluminaHumanMethylation450kmanifest")
  library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
  library("tictoc")
  library("mitch")
  library("kableExtra")
  library("beeswarm")
})

CORES=12

```

## Load data

* annotations

* probe sets

* gene sets

* design matrix

* mval matrix

```{r,annotation}

anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
myann <- data.frame(anno[,c("UCSC_RefGene_Name","Regulatory_Feature_Group","Islands_Name","Relation_to_Island")])

gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
names(gp2) <- rownames(gp)
sets <- split(rep(names(gp2), lengths(gp2)), unlist(gp2))

summary(unlist(lapply(sets,length)))

genesets <- gmt_import("https://ziemann-lab.net/public/gmea_prototype/ReactomePathways.gmt")

if (!file.exists("GSE158422_design.rds")) {
  download.file("https://ziemann-lab.net/public/gmea_prototype/GSE158422_design.rds", "GSE158422_design.rds")
}
design <- readRDS("GSE158422_design.rds")

if (!file.exists("GSE158422_design.rds")) {
 download.file("https://ziemann-lab.net/public/gmea_prototype/GSE158422_mx.rds","GSE158422_mx.rds")
}
mval <- readRDS("GSE158422_mx.rds")

```

Reactome pathways were downloaded on the 14th Sept 2023 from MsigDB.

```{r,gsdl}

gs_entrez <- gmt_import("c2.cp.reactome.v2023.1.Hs.entrez.gmt")

gs_symbols <- gmt_import("c2.cp.reactome.v2023.1.Hs.symbols.gmt")

```

## GMEA

There are two obvious ways to perform GMEA:

1. limma-agg-1stt (la1).
In this approach, the limma test is conducted on probes, followed by aggregation of the differential
methylation t-scores. Each gene gets a median t-stat value which is evaluated downstream by a 1 sample
t-test.

2. agg-limma-1stt (al1).
Here, the probe beta values for each gene are aggregated to a median and then limma is run on these
values.
To determine gene set enrichment, a 1-sample t-test is conducted.

## LA: limma aggregate 1-sample t-test

```{r,probelimma1}

sex <- as.data.frame(design)$sex
tumor <- as.data.frame(design)$tumor
patient <- as.character(unlist(lapply(1:ncol(mval),function(i) {c(i,i)})))
patient <- head(patient,ncol(mval))
design <- model.matrix(~ patient + tumor )

fit.reduced <- lmFit(mval,design)
fit.reduced <- eBayes(fit.reduced)
summary(decideTests(fit.reduced))
dm <- topTable(fit.reduced,coef=4, number = Inf)
dm <- merge(myann,dm,by=0)
dm <- dm[order(dm$P.Value),]
rownames(dm) <- dm$Row.names
dm$Row.names=NULL

head(dma) %>%
  kbl(caption="Top DM probes") %>%
  kable_styling("hover",full_width=FALSE)

agg <- function(dm,cores=1) {
  gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
  gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
  gnl <- mclapply(gnl,unique,mc.cores=cores)
  dm$UCSC_RefGene_Name <- gnl
  l <- mclapply(1:nrow(dm), function(i) {
    a <- dm[i,]
    len <- length(a[[1]][[1]])
    tvals <- as.numeric(rep(a["t"],len))
    genes <- a[[1]][[1]]
    data.frame(genes,tvals)
  },mc.cores=cores)
  df <- do.call(rbind,l)
  keep <- names(which(table(df$genes)>1))
  df <- df[df$genes %in% keep,]
  gn <- unique(df$genes)
  gme_res <- lapply( 1:length(gn), function(i) {
    g <- gn[i]
    tstats <- df[which(df$genes==g),"tvals"]
    myn <- length(tstats)
    mymean <- mean(tstats)
    mymedian <- median(tstats)
    if ( length(tstats) > 2 ) {
      ttest <- t.test(tstats)
      pval <- ttest$p.value
    } else {
      pval = 1
    }
    res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,
      "median"=mymedian, pval=pval)
  } )
  gme_res_df <- do.call(rbind, gme_res)
  rownames(gme_res_df) <- gme_res_df[,1]
  gme_res_df <- gme_res_df[,-1]
  tmp <- apply(gme_res_df,2,as.numeric)
  rownames(tmp) <- rownames(gme_res_df)
  gme_res_df <- as.data.frame(tmp)
  gme_res_df$sig <- -log10(gme_res_df[,4])
  gme_res_df <- gme_res_df[order(-gme_res_df$sig),]
  gme_res_df$fdr <- p.adjust(gme_res_df$pval)
  out <- list("df"=df,"gme_res_df"=gme_res_df)
  return(out)
}

tic()
dmagg <- agg(dm,cores=CORES)
toc()

head(dmagg$gme_res_df,20) %>%
  kbl(caption="top gmea genes (t-test)") %>%
  kable_paper("hover", full_width = F)

```

Check whether the distribution is even - it's not.

```{r,tmeddist}

hist(dmagg$gme_res_df$median,breaks=30) ; grid()

```

1-sample t-test for gene set enrichment.

```{r,t_enrichment_test1}

ttenrich <- function(m,genesets,cores=1) {
  res <- mclapply( 1:length(genesets), function(i) {
    gs <- genesets[i]
    name <- names(gs)
    n_members <- length(which(rownames(m) %in% gs[[1]]))
    if ( n_members > 4 ) {
      tstats <- m[which(rownames(m) %in% gs[[1]]),]
      myn <- length(tstats)
      mymean <- mean(tstats)
      mymedian <- median(tstats)
      wt <- t.test(tstats)
      res <- c(name,myn,mymean,mymedian,wt$p.value)
    }
  } , mc.cores = cores)
  res_df <- do.call(rbind, res)
  rownames(res_df) <- res_df[,1]
  res_df <- res_df[,-1]
  colnames(res_df) <- c("n_genes","t_mean","t_median","pval")
  tmp <- apply(res_df,2,as.numeric)
  rownames(tmp) <- rownames(res_df)
  res_df <- tmp
  res_df <- as.data.frame(res_df)
  res_df <- res_df[order(res_df$pval),]
  res_df$logp <- -log10(res_df$pval )
  res_df$fdr <- p.adjust(res_df$pval,method="fdr")
  res_df[order(abs(res_df$pval)),]
  return(res_df)
}

# need to get the median column before analysis
m <- dmagg$gme_res_df[,"median",drop=FALSE]

tic()
lares <- ttenrich(m=m,genesets=gs_symbols,cores=CORES)
toc()

head(lares,30) %>%
  kbl(caption = "Top significant genesets using LA approach") %>%
  kable_paper("hover", full_width = F)

head(lares[order(-abs(lares$t_median)),],30) %>%
  kbl(caption = "Top effect size genesets using LA approach") %>%
  kable_paper("hover", full_width = F)

lasig <- subset(lares,fdr<0.05)
nrow(lasig)

head(lasig[order(-abs(lasig$t_median)),],30) %>%
  kbl(caption = "Top effect size genesets after FDR filtering using LA approach") %>%
  kable_paper("hover", full_width = F)

plot(lares$t_median,-log10(lares$pval),pch=19,col="gray")
grid()
points(lasig$t_median,-log10(lasig$pval),pch=19,col="red")

```

## AL: Aggregate limma t-test

Determine the median of probe values

```{r,med1}

mx <- mval

gn <- unique(unlist(strsplit( myann$UCSC_RefGene_Name ,";")))
gnl <- strsplit( myann$UCSC_RefGene_Name ,";")
gnl <- mclapply(gnl,unique,mc.cores=CORES)
myann$gnl <- gnl

mymed <- function(g) {
  probes <- rownames(myann[grep(g,myann$gnl),])
  rows <- which(rownames(mx) %in% probes)
  if ( length(rows) > 1 ) {
    b <- mx[rows,]
    med <- apply(b,2,median)
    med <- matrix(med,nrow=1)
    colnames(med) <- colnames(b)
    rownames(med) <- g
    return(med)
  }
}

med <- mclapply(gn,mymed,mc.cores=CORES)
med <- med[lapply(med,length)>0]
medf <- do.call(rbind,med)

```

Then do a limma test on median gene methylation.
See if it is more robust than other approaches.

```{r,limma1}

fit.reduced <- lmFit(medf,design)
fit.reduced <- eBayes(fit.reduced)
summary(decideTests(fit.reduced))
magg <- topTable(fit.reduced,coef=ncol(design), number = Inf)

#write.table(dma3a,file="dma3a.tsv")

head(magg, 30) %>%
  kbl(caption="limma gene methylation results") %>%
  kable_paper("hover", full_width = F)

```

1-sample t-test for enrichment

```{r,ttest}

ttenrich <- function(m,genesets,cores=1) {
  res <- mclapply( 1:length(genesets), function(i) {
    gs <- genesets[i]
    name <- names(gs)
    n_members <- length(which(rownames(m) %in% gs[[1]]))
    if ( n_members > 4 ) {
      tstats <- m[which(rownames(m) %in% gs[[1]]),]
      myn <- length(tstats)
      mymean <- mean(tstats)
      mymedian <- median(tstats)
      wt <- t.test(tstats)
      pval = wt$p.value
      res <- c(name,myn,mymean,mymedian,pval)
    }
  } , mc.cores = cores)
  res_df <- do.call(rbind, res)
  rownames(res_df) <- res_df[,1]
  res_df <- res_df[,-1]
  colnames(res_df) <- c("n_genes","t_mean","t_median","pval")
  tmp <- apply(res_df,2,as.numeric)
  rownames(tmp) <- rownames(res_df)
  res_df <- tmp
  res_df <- as.data.frame(res_df)
  res_df <- res_df[order(res_df$pval),]
  res_df$logp <- -log10(res_df$pval )
  res_df$fdr <- p.adjust(res_df$pval,method="fdr")
  res_df[order(abs(res_df$pval)),]
  return(res_df)
}

m <- as.data.frame(magg$t)
rownames(m) <- rownames(magg)
colnames(m) <- "t"

alres <- ttenrich(m=m,genesets=gs_symbols,cores=CORES)

alsig <- subset(alres,fdr<0.05)

nrow(alsig)

nrow(subset(alsig,t_median>0))

head(alsig[order(-alsig$t_median),],20) %>%
  kbl(caption="hypermethylated pathways") %>%
  kable_paper("hover", full_width = F)

nrow(subset(alsig,t_median<0))

head(alsig[order(alsig$t_median),],20) %>%
  kbl(caption="hypomethylated pathways") %>%
  kable_paper("hover", full_width = F)

```

## Overlap

Use Euler diagrams to find the similarity.
As medians are normally distributed about the zero, one would expect equal numbers of up and down-
methylated gene sets.
Therefore the t-test based approach could be considered better.

```{r,ol1}

lasig_up <- rownames(subset(lasig,t_median>0))
lasig_dn <- rownames(subset(lasig,t_median<0))

alsig_up <- rownames(subset(alsig,t_median>0))
alsig_dn <- rownames(subset(alsig,t_median<0))

v1 <- list("LA up"=lasig_up, "LA dn"=lasig_dn,
  "AL up"=alsig_up,"AL dn"=alsig_dn)

plot(euler(v1),quantities = TRUE)

```

## Session information

```{r,save}

save.image("GSE158422_gmea.Rdata")

sessionInfo()

```

### 2023-09-25 figs/GSE158422_rna_int.Rmd
more work on RNA integration

---
title: "GSE158422: DNA methylation in Lung squamous cell carcinoma - RNA expression and integration"
author: "The GMEA team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
---

## Introduction

```{r,packages}

suppressPackageStartupMessages({
  library("DESeq2")
  library("eulerr")
  library("IlluminaHumanMethylation450kmanifest")
  library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
  library("tictoc")
  library("mitch")
  library("kableExtra")
  library("beeswarm")
  library("HGNChelper")
  library("gplots")
})

CORES=12

```

## Load annotations

```{r,annotation}

load("GSE158422_gmea.Rdata")

```

Reactome pathways were downloaded on the 14th Sept 2023 from MsigDB.

```{r,gsdl}

gs_entrez <- gmt_import("c2.cp.reactome.v2023.1.Hs.entrez.gmt")

gs_symbols <- gmt_import("c2.cp.reactome.v2023.1.Hs.symbols.gmt")

```

## Load RNA expression

From GEO https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE158420
GSE158420_counts.txt.gz

## DESEQ2

```{r,deseq1}

x <- read.table("GSE158420_counts.txt.gz",row.names=NULL)
dim(x)
x <- aggregate(. ~ row.names,x,sum)
dim(x)

#RUFKM
head(x)

```

There are some gene name errors, so I will use HGNChelper to repair them.

```{r,gene_name_fix}

human=x$row.names
checkGeneSymbols(human)
fix <- checkGeneSymbols(human)
rownames(x) <- fix$Suggested.Symbol
x$gene <- fix$Suggested.Symbol
x <- aggregate(. ~ gene,x,sum)
dim(x)
row.names(x)=x$gene
x$gene=NULL
head(x)

```

Now do DESeq2

```{r,deseq2}

patient <- head(as.character(unlist(lapply(1:ncol(x),function(i) {c(i,i)}))),ncol(x))
tumor <- rep(c(0,1),ncol(x)/2)
ss <- data.frame(patient,tumor)

dds <- DESeqDataSetFromMatrix(countData = x , colData=ss, design = ~ patient + tumor)
res <- DESeq(dds)
z <- results(res)
vsd <- vst(dds, blind=FALSE)
zz <-cbind(as.data.frame(z),assay(vsd))
dge <- as.data.frame(zz[order(zz$pvalue),])

sig <- subset(dge,padj<0.05)
SIG = nrow(sig)
DN = nrow(subset(sig,log2FoldChange<0))
UP = nrow(subset(sig,log2FoldChange>0))
HEADER = paste("mRNA", SIG , "DEGs,", UP ,"upregulated,", DN, "downregulated")
# smear
plot(log2(dge$baseMean),dge$log2FoldChange,cex=0.6,cex.axis=1.2,cex.lab=1.3,
  xlab="log2 base mean",
  ylab="log2 fold change" ,pch=19,col="#838383")
points(log2(sig$baseMean),sig$log2FoldChange,cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
top <- head(sig,20)
# volcano
plot(dge$log2FoldChange, -log2(dge$pvalue) ,cex=0.6, cex.lab=1.3,cex.axis=1.2,
 ,xlab="log2 fold change", ylab="-log2 p-value" ,pch=19,col="#838383")
points(sig$log2FoldChange, -log2(sig$pvalue),cex=0.6,pch=19,col="red"
mtext((HEADER),cex=1.2)
# top N gene heatmap
colCols <- rep(c(0,1),ncol(x)/2)
colCols <- gsub("0","gray",colCols)
colCols <- gsub("1","black",colCols)

colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2(  as.matrix(dge[1:50,c(7:ncol(dge))]), col=colfunc(25),scale="row",
  ColSideColors=colCols ,
  trace="none",margins = c(6,10), cexRow=.6, cexCol=.5, main="Top 50 genes")

```

Pathway enrichment with mitch.

```{r,mitch}

y <- mitch_import(dge, DEtype="deseq2")

res <- mitch_calc(y, genesets=gs_symbols, priority="effect",cores=CORES)

mres <- res$enrichment_result
msig <- subset(mres,p.adjustANOVA<0.05)

head(subset(mres,p.adjustANOVA<0.05 & s.dist>0),20) %>%
  kbl(caption="mRNA upregulated pathways") %>%
  kable_styling("hover",full_width=FALSE)

head(subset(mres,p.adjustANOVA<0.05 & s.dist<0),20) %>%
  kbl(caption="mRNA downregulated pathways") %>%
  kable_styling("hover",full_width=FALSE)

```

## Load methylation enrichment data and calculate overlap

Use Euler diagrams to find the similarity.
As medians are normally distributed about the zero, one would expect equal numbers of up and down-
methylated gene sets.

```{r,overlap1}

load("GSE158422_gmea.Rdata")

head(mres)

head(alres)

head(lsres)

msig_up <- subset(msig,s.dist>0)$set
msig_dn <- subset(msig,s.dist<0)$set

lasig_up <- rownames(subset(lasig,t_median>0))
lasig_dn <- rownames(subset(lasig,t_median<0))

alsig_up <- rownames(subset(alsig,t_median>0))
alsig_dn <- rownames(subset(alsig,t_median<0))

v1 <- list("LA up"=lasig_up, "LA dn"=lasig_dn,
  "RNA up"=msig_up,"RNA dn"=msig_dn)
plot(euler(v1),quantities = TRUE)

v1 <- list("AL up"=alsig_up,"AL dn"=alsig_dn,
  "RNA up"=msig_up,"RNA dn"=msig_dn)
plot(euler(v1),quantities = TRUE)

```

## Session information

```{r,save}

sessionInfo()

```

### 2023-09-25 figs/step_01b_test_gometh.Rmd
more work on RNA integration

@@
     mymedian <- median(tstats)
-    tselfcont <- t.test(tstats)
+    if ( length(tstats) ) > 2 {
+      ttest <- t.test(tstats)
+      pval <- ttest$p.value
+    } else {
+      pval = 1
+    }
     res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,"median"=mymedian,
-      "p-value(sc)"=tselfcont$p.value)
+      "pval"=ttest$p.value)
   } )
@@
 summary(decideTests(fit.reduced))
-dm <- topTable(fit.reduced,coef=4, number = Inf)
+magg <- topTable(fit.reduced,coef=4, number = Inf)
 
-#write.table(dma3a,file="dma3a.tsv")
-
-head(dm, 30) %>%
+head(magg, 30) %>%
   kbl(caption="limma gene methylation results") %>%
@@
 
-m <- as.data.frame(dm$t)
-rownames(m) <- rownames(dm)
+m <- as.data.frame(magg$t)
+rownames(m) <- rownames(magg)
 colnames(m) <- "t"

### 2023-09-27 figs/GSE158422_gmea.Rmd
more work on integrating methylation and RNA expression

@@
 
+boxplot(list("tumor"=matrix(colMeans(mval),ncol=2)[,1],"normal"=matrix(colMeans(mval),ncol=2)[,2]),
+  main="mean probe methylation mval")
+
 ```
@@
 
-## GMEA
+# GMEA: all probes
 
@@
 
-## LA: limma aggregate 1-sample t-test
+## LA: limma aggregate 1-sample t-test all probes
 
@@
 
-head(dma) %>%
+hist(dm$t,breaks=30,main="all probes")
+hist(subset(dm,Relation_to_Island=="Island")$t,main="CpG island")
+hist(subset(dm,Relation_to_Island=="N_Shore")$t,main="CpG shore")
+hist(subset(dm,Relation_to_Island=="N_Shelf")$t,main="CpG shelf")
+
+hist(subset(dm,Regulatory_Feature_Group=="",main="No genomic feature")
+hist(subset(dm,Regulatory_Feature_Group=="Promoter_Associated")$t,main="promoter")
+hist(subset(dm,Regulatory_Feature_Group=="Gene_Associated")$t,main="gene")
+hist(subset(dm,Regulatory_Feature_Group=="Unclassified")$t,main="unclassified")
+
+head(dm) %>%
   kbl(caption="Top DM probes") %>%
@@
 
+hist(dmagg$gme_res_df$mean,breaks=30) ; grid()
+
 ```
@@
 
+# try mean and median
+m <- dmagg$gme_res_df[,"mean",drop=FALSE]
+lares <- ttenrich(m=m,genesets=gs_symbols,cores=CORES)
+hist(lares$t_mean)
+
 # need to get the median column before analysis
 m <- dmagg$gme_res_df[,"median",drop=FALSE]
-
-tic()
 lares <- ttenrich(m=m,genesets=gs_symbols,cores=CORES)
-toc()
+hist(lares$t_median)
 
@@
 
-## AL: Aggregate limma t-test
+## AL: Aggregate limma t-test all probes
 
@@
 
+# compare median and mean based aggregation
 mymed <- function(g) {
@@
 
+boxplot(list("normal"=matrix(colMeans(mval),ncol=2)[,2],"tumor"=matrix(colMeans(mval),ncol=2)[,1]),
+  main="mean mval for control and tumor samples (all probes)")
+
+boxplot(list("normal"=matrix(colMeans(medf),ncol=2)[,2],"tumor"=matrix(colMeans(medf),ncol=2)[,1]),
+  main="mean mval for control and tumor samples (gene associated)")
+
 ```
@@
 
-#write.table(dma3a,file="dma3a.tsv")
+hist(magg$t)
 
@@
 ```
+
+Very curious that the distribution is negatively skewed now.
+
+Let's do another test to see whether this is correct or not.
+
+```{r,test}
+
+mytest <- lapply(1:nrow(medf),function(i) {
+  ctl <- matrix(medf[i,],ncol=2)[,2]
+  tum <- matrix(medf[i,],ncol=2)[,1]
+  tt <- t.test(tum,ctl)
+  t=tt$statistic ; p=tt$p.value ; est=rev(tt$estimate)
+  names(est) = c("ctl","tum")
+  c(t,p,est)
+} )
+
+mytest <- as.data.frame(do.call(rbind,mytest))
+colnames(mytest) = c("t","p","ctl","tum")
+rownames(mytest) <- rownames(medf)
+hist(mytest[,"t"])
+
+mytest <- mytest[order(mytest$p),]
+head(mytest,20) %>%
+  kbl(caption="Top significant differential gene methylation results using dumb test") %>%
+  kable_paper("hover", full_width = F)
+
+nrow(subset(mytest,p<0.05 & t>0))
+nrow(subset(mytest,p<0.05 & t<0))
+
+```
+
+Hypomethylated genes still outnumber hypermethylated genes.
+
+It does skew downwards after aggregation.
 
@@
 
+hist(alres$t_mean)
+
 alsig <- subset(alres,fdr<0.05)
@@
 
+# GMEA: promoter probes only
+
+## Limma-aggregate-1-sample t-test promoter only
+
+```{r,ls_promoter}
+
+dmp <- dm[grep("Promo",dm$Regulatory_Feature_Group),]
+
+tic()
+dmpagg <- agg(dmp,cores=CORES)
+toc()
+
+head(dmpagg$gme_res_df,20) %>%
+  kbl(caption="top gmea genes (t-test)") %>%
+  kable_paper("hover", full_width = F)
+
+```
+
+Check whether the distribution is even - it's not.
+
+```{r,laphist}
+
+hist(dmpagg$gme_res_df$median,breaks=30) ; grid()
+
+```
+
+1-sample t-test for gene set enrichment.
+
+```{r,t_enrichment_test1_promoter}
+
+# need to get the median column before analysis
+m <- dmpagg$gme_res_df[,"median",drop=FALSE]
+
+tic()
+lapres <- ttenrich(m=m,genesets=gs_symbols,cores=CORES)
+toc()
+
+head(lapres,30) %>%
+  kbl(caption = "Top significant genesets using LA approach for promoters") %>%
+  kable_paper("hover", full_width = F)
+
+head(lapres[order(-abs(lapres$t_median)),],30) %>%
+  kbl(caption = "Top effect size genesets using LA approach") %>%
+  kable_paper("hover", full_width = F)
+
+lapsig <- subset(lapres,fdr<0.05)
+nrow(lapsig)
+
+head(lapsig[order(-abs(lapsig$t_median)),],30) %>%
+  kbl(caption = "Top effect size genesets after FDR filtering using LA approach for promoters") %>%
+  kable_paper("hover", full_width = F)
+
+plot(lapres$t_median,-log10(lapres$pval),pch=19,col="gray")
+grid()
+points(lapsig$t_median,-log10(lapsig$pval),pch=19,col="red")
+
+```
+
+## AL: Aggregate limma t-test Promoter only
+
+Determine the median of probe values.
+
+```{r,med1p}
+
+gn <- unique(unlist(strsplit( myann$UCSC_RefGene_Name ,";")))
+gnl <- strsplit( myann$UCSC_RefGene_Name ,";")
+gnl <- mclapply(gnl,unique,mc.cores=CORES)
+myann$gnl <- gnl
+myannp <- myann[grep("Promot",myann$Regulatory_Feature_Group),]
+
+mxp <- mval[rownames(mval) %in% rownames(myannp),]
+
+mymed <- function(g) {
+  probes <- rownames(myannp[grep(g,myannp$gnl),])
+  rows <- which(rownames(mxp) %in% probes)
+  if ( length(rows) > 1 ) {
+    b <- mxp[rows,]
+    med <- apply(b,2,median)
+    med <- matrix(med,nrow=1)
+    colnames(med) <- colnames(b)
+    rownames(med) <- g
+    return(med)
+  }
+}
+
+medp <- mclapply(gn,mymed,mc.cores=CORES)
+medp <- medp[lapply(medp,length)>0]
+medfp <- do.call(rbind,medp)
+
+```
+
+Then do a limma test on median gene methylation.
+See if it is more robust than other approaches.
+
+```{r,limma1p}
+
+fit.reduced <- lmFit(medfp,design)
+fit.reduced <- eBayes(fit.reduced)
+summary(decideTests(fit.reduced))
+maggp <- topTable(fit.reduced,coef=ncol(design), number = Inf)
+
+head(maggp, 30) %>%
+  kbl(caption="limma gene methylation results") %>%
+  kable_paper("hover", full_width = F)
+
+```
+
+1-sample t-test for enrichment promoter.
+
+```{r,ttestp}
+
+m <- as.data.frame(maggp$t)
+rownames(m) <- rownames(maggp)
+colnames(m) <- "t"
+
+alpres <- ttenrich(m=m,genesets=gs_symbols,cores=CORES)
+
+alpsig <- subset(alpres,fdr<0.05)
+
+nrow(alpsig)
+
+nrow(subset(alpsig,t_median>0))
+
+head(alpsig[order(-alpsig$t_median),],20) %>%
+  kbl(caption="hypermethylated pathways") %>%
+  kable_paper("hover", full_width = F)
+
+nrow(subset(alpsig,t_median<0))
+
+head(alpsig[order(alpsig$t_median),],20) %>%
+  kbl(caption="hypomethylated pathways") %>%
+  kable_paper("hover", full_width = F)
+
+```
+
+## Promoter Overlap
+
+Use Euler diagrams to find the similarity.
+As medians are normally distributed about the zero, one would expect equal numbers of up and down-
+methylated gene sets.
+Therefore the t-test based approach could be considered better.
+
+```{r,ol1p}
+
+lapsig_up <- rownames(subset(lapsig,t_median>0))
+lapsig_dn <- rownames(subset(lapsig,t_median<0))
+
+alpsig_up <- rownames(subset(alpsig,t_median>0))
+alpsig_dn <- rownames(subset(alpsig,t_median<0))
+
+v1 <- list("LA up"=lapsig_up, "LA dn"=lapsig_dn,
+  "AL up"=alpsig_up,"AL dn"=alpsig_dn)
+
+plot(euler(v1),quantities = TRUE)
+
+```
+
 ## Session information

### 2023-09-27 figs/GSE158422_rna_int.Rmd
more work on integrating methylation and RNA expression

@@
 human=x$row.names
-checkGeneSymbols(human)
 fix <- checkGeneSymbols(human)
-rownames(x) <- fix$Suggested.Symbol
 x$gene <- fix$Suggested.Symbol
+x$row.names=NULL
 x <- aggregate(. ~ gene,x,sum)
@@
  ,xlab="log2 fold change", ylab="-log2 p-value" ,pch=19,col="#838383")
-points(sig$log2FoldChange, -log2(sig$pvalue),cex=0.6,pch=19,col="red"
+points(sig$log2FoldChange, -log2(sig$pvalue),cex=0.6,pch=19,col="red")
 mtext((HEADER),cex=1.2)
@@
 
-head(lsres)
+head(lares)
 
@@
 
+## Load promoter methylation enrichment data and calculate overlap
+
+```{r,overlap2}
+
+head(alpres)
+
+head(lapres)
+
+lapsig_up <- rownames(subset(lapsig,t_median>0))
+lapsig_dn <- rownames(subset(lapsig,t_median<0))
+
+alpsig_up <- rownames(subset(alpsig,t_median>0))
+alpsig_dn <- rownames(subset(alpsig,t_median<0))
+
+v1 <- list("LAP up"=lapsig_up, "LAP dn"=lapsig_dn,
+  "RNA up"=msig_up,"RNA dn"=msig_dn)
+plot(euler(v1),quantities = TRUE)
+
+v1 <- list("ALP up"=alpsig_up,"ALP dn"=alpsig_dn,
+  "RNA up"=msig_up,"RNA dn"=msig_dn)
+plot(euler(v1),quantities = TRUE)
+
+```
+
+## Conclusion
+
+Although there were some differentially methylated pathways analysed with different methods,
+they were not concordant with differential expression.
+
 ## Session information

### 2023-09-28 figs/GSE158422_split.Rmd
split + analyse working LFG

---
title: "GSE158422: DNA methylation in Lung squamous cell carcinoma - gene methylation enrichment analysis"
author: "The GMEA team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
---

## Introduction

Here we are splitting the dataset and testing concordance between LA and AL approaches.

```{r,packages}

suppressPackageStartupMessages({
  library("limma")
  library("eulerr")
  library("IlluminaHumanMethylation450kmanifest")
  library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
  library("tictoc")
  library("mitch")
  library("kableExtra")
  library("beeswarm")
})

CORES=12

```

## Load data

* annotations

* probe sets

* gene sets

* design matrix

* mval matrix

```{r,annotation}

anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
myann <- data.frame(anno[,c("UCSC_RefGene_Name","Regulatory_Feature_Group","Islands_Name","Relation_to_Island")])

gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
names(gp2) <- rownames(gp)
sets <- split(rep(names(gp2), lengths(gp2)), unlist(gp2))

summary(unlist(lapply(sets,length)))

genesets <- gmt_import("https://ziemann-lab.net/public/gmea_prototype/ReactomePathways.gmt")

if (!file.exists("GSE158422_design.rds")) {
  download.file("https://ziemann-lab.net/public/gmea_prototype/GSE158422_design.rds", "GSE158422_design.rds")
}
design <- readRDS("GSE158422_design.rds")

if (!file.exists("GSE158422_design.rds")) {
 download.file("https://ziemann-lab.net/public/gmea_prototype/GSE158422_mx.rds","GSE158422_mx.rds")
}
mval <- readRDS("GSE158422_mx.rds")

boxplot(list("normal"=matrix(colMeans(mval),ncol=2)[,2],"tumor"=matrix(colMeans(mval),ncol=2)[,1]),
  main="mean probe methylation mval")

```

Reactome pathways were downloaded on the 14th Sept 2023 from MsigDB.

```{r,gsdl}

gs_entrez <- gmt_import("c2.cp.reactome.v2023.1.Hs.entrez.gmt")

gs_symbols <- gmt_import("c2.cp.reactome.v2023.1.Hs.symbols.gmt")

```

# GMEA FUNCTIONS

## LA

This process runs limma first and then aggregates the results before doing an enrichment test.

```{r,lafunc}

# limma
runlimma <- function(mval,design,myann) {
  fit.reduced <- lmFit(mval,design)
  fit.reduced <- eBayes(fit.reduced)
  summary(decideTests(fit.reduced))
  dm <- topTable(fit.reduced,coef=4, number = Inf)
  dm <- merge(myann,dm,by=0)
  dm <- dm[order(dm$P.Value),]
  rownames(dm) <- dm$Row.names
  dm$Row.names=NULL
  return(dm)
}

# aggregate
agg <- function(dm,cores=1) {
  gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
  gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
  gnl <- mclapply(gnl,unique,mc.cores=cores)
  dm$UCSC_RefGene_Name <- gnl
  l <- mclapply(1:nrow(dm), function(i) {
    a <- dm[i,]
    len <- length(a[[1]][[1]])
    tvals <- as.numeric(rep(a["t"],len))
    genes <- a[[1]][[1]]
    data.frame(genes,tvals)
  },mc.cores=cores)
  df <- do.call(rbind,l)
  keep <- names(which(table(df$genes)>1))
  df <- df[df$genes %in% keep,]
  gn <- unique(df$genes)
  gme_res <- lapply( 1:length(gn), function(i) {
    g <- gn[i]
    tstats <- df[which(df$genes==g),"tvals"]
    myn <- length(tstats)
    mymean <- mean(tstats)
    mymedian <- median(tstats)
    if ( length(tstats) > 2 ) {
      ttest <- t.test(tstats)
      pval <- ttest$p.value
    } else {
      pval = 1
    }
    res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,
      "median"=mymedian, pval=pval)
  } )
  gme_res_df <- do.call(rbind, gme_res)
  rownames(gme_res_df) <- gme_res_df[,1]
  gme_res_df <- gme_res_df[,-1]
  tmp <- apply(gme_res_df,2,as.numeric)
  rownames(tmp) <- rownames(gme_res_df)
  gme_res_df <- as.data.frame(tmp)
  gme_res_df$sig <- -log10(gme_res_df[,4])
  gme_res_df <- gme_res_df[order(-gme_res_df$sig),]
  gme_res_df$fdr <- p.adjust(gme_res_df$pval)
  out <- list("df"=df,"gme_res_df"=gme_res_df)
  return(out)
}

# enrich
ttenrich <- function(m,genesets,cores=1) {
  res <- mclapply( 1:length(genesets), function(i) {
    gs <- genesets[i]
    name <- names(gs)
    n_members <- length(which(rownames(m) %in% gs[[1]]))
    if ( n_members > 4 ) {
      tstats <- m[which(rownames(m) %in% gs[[1]]),]
      myn <- length(tstats)
      mymean <- mean(tstats)
      mymedian <- median(tstats)
      wt <- t.test(tstats)
      res <- c(name,myn,mymean,mymedian,wt$p.value)
    }
  } , mc.cores = cores)
  res_df <- do.call(rbind, res)
  rownames(res_df) <- res_df[,1]
  res_df <- res_df[,-1]
  colnames(res_df) <- c("n_genes","t_mean","t_median","pval")
  tmp <- apply(res_df,2,as.numeric)
  rownames(tmp) <- rownames(res_df)
  res_df <- tmp
  res_df <- as.data.frame(res_df)
  res_df <- res_df[order(res_df$pval),]
  res_df$logp <- -log10(res_df$pval )
  res_df$fdr <- p.adjust(res_df$pval,method="fdr")
  res_df[order(abs(res_df$pval)),]
  return(res_df)
}

```

## AL: Aggregate limma

Functions for aggregate-limma-enrich approach.

```{r,alfunc}

# chromosome by chromosome will be much faster
magg <- function(mval,myann,cores=1){
  gn <- unique(unlist(strsplit( myann$UCSC_RefGene_Name ,";")))
  gnl <- strsplit( myann$UCSC_RefGene_Name ,";")
  gnl <- mclapply(gnl,unique,mc.cores=cores)
  myann$gnl <- gnl
  keep <- rownames(subset(myann,UCSC_RefGene_Name!=""))
  mx <- mval[rownames(mval) %in% keep,]
  mymed <- function(g) {
    probes <- rownames(myann[grep(g,myann$gnl),])
    rows <- which(rownames(mx) %in% probes)
    if ( length(rows) > 1 ) {
      b <- mx[rows,]
      med <- apply(b,2,median)
      med <- matrix(med,nrow=1)
      colnames(med) <- colnames(b)
      rownames(med) <- g
      return(med)
    }
  }
  med <- mclapply(gn,mymed,mc.cores=cores)
  med <- med[lapply(med,length)>0]
  medf <- do.call(rbind,med)
  return(medf)
}

chragg <- function(mval,myann,cores=1){
  keep <- rownames(subset(myann,UCSC_RefGene_Name!=""))
  mx <- mval[rownames(mval) %in% keep,]
  chrs <- unique(anno$chr)
  myorder <- unlist(lapply(chrs,function(mychr) { nrow(subset(annodf,chr==mychr)) } ))
  chrs <- chrs[order(-myorder)]
  leadercores <- floor(sqrt(cores))
  workercores <- ceiling(sqrt(cores))
  chrmedf <- mclapply(chrs,function(chr) {
    annodf <- as.data.frame(anno)
    chrfrag <- annodf[annodf$chr==chr,]
    chrprobes <-rownames(chrfrag)
    chrmx <- mx[rownames(mx) %in% chrprobes,]
    chranno <- myann[rownames(myann) %in% chrprobes,]
    chrmedf <- magg(mval=chrmx,myann=chranno,cores=workercores)
    return(chrmedf)
  },mc.cores=leadercores)
  medf <- do.call(rbind, chrmedf)
  return(medf)
}

agglimma <- function(medf,design) {
  fit.reduced <- lmFit(medf,design)
  fit.reduced <- eBayes(fit.reduced)
  dmagg <- topTable(fit.reduced,coef=ncol(design), number = Inf)
  nondup <- !duplicated(dmagg$ID)
  dmagg <- dmagg[nondup,]
  rownames(dmagg) <- dmagg$ID
  dmagg$ID = NULL
  return(dmagg)
}

#ttenrich is same

```

## Prep

```{r,prep}

# prep
sex <- as.data.frame(design)$sex
tumor <- as.data.frame(design)$tumor
patient <- as.character(unlist(lapply(1:ncol(mval),function(i) {c(i,i)})))
patient <- head(patient,ncol(mval))
design <- model.matrix(~ patient + tumor )
rownames(design) <- colnames(mval)

```

## LA workflow split and analyse

split 74 samples 37 patients split into 18 + 18
Conduct LA workflow analysis.
Then make overlap.

```{r,lafunc}

splitanalyze_la <- function(seed){
  set.seed(seed)
  a <- sample(1:37,18)
  a <- a[order(a)]
  nona <- setdiff(1:37,a)
  set.seed(seed)
  b <- sample(nonctl,18)
  b <- b[order(b)]
  a <- c( (a*2)-1 ,a*2 )
  a <- a[order(a)]
  b <- c( (b*2)-1 ,b*2 )
  b <- b[order(b)]

  design_a <- design[a,]
  design_a <- design_a[,colSums(design_a)>0]
  mval_a <- mval[,a]
  design_b <- design[b,]
  design_b <- design_b[,colSums(design_b)>0]
  mval_b <- mval[,b]

  # la1
  dm1 <- runlimma(mval_a,design_a,myann)
  dmagg1 <- agg(dm1,cores=CORES)
  m1 <- dmagg1$gme_res_df[,"mean",drop=FALSE]
  lares1 <- ttenrich(m=m1,genesets=gs_symbols,cores=CORES)
  lasig1 <- subset(lares1,fdr<0.01)

  # la2
  dm2 <- runlimma(mval_b,design_b,myann)
  dmagg2 <- agg(dm2,cores=CORES)
  m2 <- dmagg2$gme_res_df[,"mean",drop=FALSE]
  lares2 <- ttenrich(m=m2,genesets=gs_symbols,cores=CORES)
  lasig2 <- subset(lares2,fdr<0.01)

  lasig1_up <- rownames(subset(lasig1,t_median>0))
  lasig1_dn <- rownames(subset(lasig1,t_median<0))
  lasig2_up <- rownames(subset(lasig2,t_median>0))
  lasig2_dn <- rownames(subset(lasig2,t_median<0))

  ##v1 <- list("LA1 up"=lasig1_up, "LA1 dn"=lasig1_dn,
  ##  "LA2 up"=lasig2_up,"LA2 dn"=lasig2_dn)
  ## plot(euler(v1),quantities = TRUE)

  total <- length(unique(c(lasig1_up,lasig2_up,lasig1_dn,lasig2_dn)))
  common <- length(intersect(lasig1_up, lasig2_up)) + length(intersect(lasig1_dn, lasig2_dn))
  discordant <- length(intersect(lasig1_up, lasig2_dn)) + length(intersect(lasig1_dn, lasig2_up))
  uncommon = total - common - discordant
  result <- c("total"=total,"common"=common,"discordant"=discordant,"uncommon"=uncommon)
}

lares <- lapply( seq(100,1000,100) ,splitanalyze_la)

```

## AL workflow split and analyse

```{alfunc}

splitanalyze_al <- function(seed){
  set.seed(seed)
  a <- sample(1:37,18)
  a <- a[order(a)]
  nona <- setdiff(1:37,a)
  set.seed(seed)
  b <- sample(nonctl,18)
  b <- b[order(b)]
  a <- c( (a*2)-1 ,a*2 )
  a <- a[order(a)]
  b <- c( (b*2)-1 ,b*2 )
  b <- b[order(b)]

  design_a <- design[a,]
  design_a <- design_a[,colSums(design_a)>0]
  mval_a <- mval[,a]
  design_b <- design[b,]
  design_b <- design_b[,colSums(design_b)>0]
  mval_b <- mval[,b]

  #al
  medf1 <- chragg(mval_a,myann,cores=CORES)
  magg1 <- agglimma(medf1,design_a)
  m1 <- as.data.frame(magg1$t)
  rownames(m1) <- rownames(magg1)
  colnames(m1) <- "t"
  alres1 <- ttenrich(m=m1,genesets=gs_symbols,cores=CORES)

  #al
  medf2 <- chragg(mval_b,myann,cores=CORES)
  magg2 <- agglimma(medf2,design_b)
  m2 <- as.data.frame(magg2$t)
  rownames(m2) <- rownames(magg2)
  colnames(m2) <- "t"
  alres2 <- ttenrich(m=m2,genesets=gs_symbols,cores=CORES)

  alsig1_up <- rownames(subset(alsig1,t_median>0))
  alsig1_dn <- rownames(subset(alsig1,t_median<0))
  alsig2_up <- rownames(subset(alsig2,t_median>0))
  alsig2_dn <- rownames(subset(alsig2,t_median<0))

  ##v1 <- list("LA1 up"=lasig1_up, "LA1 dn"=lasig1_dn,
  ##  "LA2 up"=lasig2_up,"LA2 dn"=lasig2_dn)
  ## plot(euler(v1),quantities = TRUE)

  total <- length(unique(c(alsig1_up,alsig2_up,alsig1_dn,alsig2_dn)))
  common <- length(intersect(alsig1_up, alsig2_up)) + length(intersect(alsig1_dn, alsig2_dn))
  discordant <- length(intersect(alsig1_up, alsig2_dn)) + length(intersect(alsig1_dn, alsig2_up))
  uncommon = total - common - discordant
  result <- c("total"=total,"common"=common,"discordant"=discordant,"uncommon"=uncommon)
}

alres <- lapply( seq(100,1000,100) ,splitanalyze_al)

## Session information

```{r,save}

save.image("GSE158422_gmea.Rdata")

sessionInfo()

```

### 2023-09-29 figs/GSE158422_split.Rmd
split and analyse

@@
 chragg <- function(mval,myann,cores=1){
+  annodf <- as.data.frame(anno)
   keep <- rownames(subset(myann,UCSC_RefGene_Name!=""))
@@
   chrs <- unique(anno$chr)
-  myorder <- unlist(lapply(chrs,function(mychr) { nrow(subset(annodf,chr==mychr)) } ))
+  myorder <- unlist(lapply(chrs,function(mychr) { nrow( annodf[annodf$chr==mychr,] ) } ))
   chrs <- chrs[order(-myorder)]
@@
   chrmedf <- mclapply(chrs,function(chr) {
-    annodf <- as.data.frame(anno)
     chrfrag <- annodf[annodf$chr==chr,]
@@
 
-```{r,lafunc}
+```{r,laflow}
 
@@
   set.seed(seed)
-  b <- sample(nonctl,18)
+  b <- sample(nona,18)
   b <- b[order(b)]
@@
 
-```{alfunc}
+```{r,alflow}
 
@@
 
+```
+
 ## Session information

### 2023-09-30 figs/GSE158422_split.Rmd
compare

@@
   lares1 <- ttenrich(m=m1,genesets=gs_symbols,cores=CORES)
-  lasig1 <- subset(lares1,fdr<0.01)
+  lasig1 <- subset(lares1,fdr<0.05)
 
@@
   lares2 <- ttenrich(m=m2,genesets=gs_symbols,cores=CORES)
-  lasig2 <- subset(lares2,fdr<0.01)
+  lasig2 <- subset(lares2,fdr<0.05)
 
@@
 lares <- lapply( seq(100,1000,100) ,splitanalyze_la)
+
+lares
 
@@
   alres1 <- ttenrich(m=m1,genesets=gs_symbols,cores=CORES)
+  alsig1 <- subset(alres1,fdr<0.05)
 
@@
   alres2 <- ttenrich(m=m2,genesets=gs_symbols,cores=CORES)
+  alsig2 <- subset(alres2,fdr<0.05)
 
@@
 
+alres
+
+```
+
+## Compare
+
+```{r,compare}
+
+lares <- as.data.frame(do.call(rbind,lares))
+lares$p_comm <-lares$common / lares$total
+lares$p_disc <-lares$discordant / lares$total
+lares$p_uncom <-lares$uncommon / lares$total
+lares
+
+alres <- as.data.frame(do.call(rbind,alres))
+alres$p_comm <-alres$common / alres$total
+alres$p_disc <-alres$discordant / alres$total
+alres$p_uncom <-alres$uncommon / alres$total
+alres
+
 ```

### 2023-09-30 figs/GSE158422_split.Rmd
collect results

--- 2023-09-29 figs/GSE158422_split.Rmd
@@
   lares1 <- ttenrich(m=m1,genesets=gs_symbols,cores=CORES)
-  lasig1 <- subset(lares1,fdr<0.01)
 
@@
   lares2 <- ttenrich(m=m2,genesets=gs_symbols,cores=CORES)
-  lasig2 <- subset(lares2,fdr<0.01)
-
-  lasig1_up <- rownames(subset(lasig1,t_median>0))
-  lasig1_dn <- rownames(subset(lasig1,t_median<0))
-  lasig2_up <- rownames(subset(lasig2,t_median>0))
-  lasig2_dn <- rownames(subset(lasig2,t_median<0))
-
-  ##v1 <- list("LA1 up"=lasig1_up, "LA1 dn"=lasig1_dn,
-  ##  "LA2 up"=lasig2_up,"LA2 dn"=lasig2_dn)
-  ## plot(euler(v1),quantities = TRUE)
-
-  total <- length(unique(c(lasig1_up,lasig2_up,lasig1_dn,lasig2_dn)))
-  common <- length(intersect(lasig1_up, lasig2_up)) + length(intersect(lasig1_dn, lasig2_dn))
-  discordant <- length(intersect(lasig1_up, lasig2_dn)) + length(intersect(lasig1_dn, lasig2_up))
-  uncommon = total - common - discordant
-  result <- c("total"=total,"common"=common,"discordant"=discordant,"uncommon"=uncommon)
 }
@@
 
+  list("alres1"=alres1,"alres2"=alres2)
+}
+
+alres <- lapply( seq(100,1000,100) ,splitanalyze_al)
+
+alres
+
+```
+
+## Compare
+
+```{r,compare}
+
+lacompare <- function(lares){
+  lares1 <- lares[[1]]
+  lares2 <- lares[[2]]
+  lasig1 <- subset(lares1,fdr<0.05)
+  lasig2 <- subset(lares2,fdr<0.05)
+  lasig1_up <- rownames(subset(lasig1,t_median>0))
+  lasig1_dn <- rownames(subset(lasig1,t_median<0))
+  lasig2_up <- rownames(subset(lasig2,t_median>0))
+  lasig2_dn <- rownames(subset(lasig2,t_median<0))
+
+  total <- length(unique(c(lasig1_up,lasig2_up,lasig1_dn,lasig2_dn)))
+  common <- length(intersect(lasig1_up, lasig2_up)) + length(intersect(lasig1_dn, lasig2_dn))
+  discordant <- length(intersect(lasig1_up, lasig2_dn)) + length(intersect(lasig1_dn, lasig2_up))
+  uncommon = total - common - discordant
+
+  lares <- as.data.frame(do.call(rbind,lares))
+  lares$p_comm <-lares$common / lares$total
+  lares$p_disc <-lares$discordant / lares$total
+  lares$p_uncom <-lares$uncommon / lares$total
+  lares
+}
+
+alcompare <-function(alres){
+  alres1 <- alres[[1]]
+  alres2 <- alres[[2]]
+  alsig1 <- subset(alres1,fdr<0.05)
+  alsig2 <- subset(alres2,fdr<0.05)
+
   alsig1_up <- rownames(subset(alsig1,t_median>0))
@@
   alsig2_dn <- rownames(subset(alsig2,t_median<0))
-
-  ##v1 <- list("LA1 up"=lasig1_up, "LA1 dn"=lasig1_dn,
-  ##  "LA2 up"=lasig2_up,"LA2 dn"=lasig2_dn)
-  ## plot(euler(v1),quantities = TRUE)
 
@@
   result <- c("total"=total,"common"=common,"discordant"=discordant,"uncommon"=uncommon)
+
+  alres <- as.data.frame(do.call(rbind,alres))
+  alres$p_comm <-alres$common / alres$total
+  alres$p_disc <-alres$discordant / alres$total
+  alres$p_uncom <-alres$uncommon / alres$total
+  alres
 }
-
-alres <- lapply( seq(100,1000,100) ,splitanalyze_al)
 

### 2023-09-30 figs/GSE158422_split.Rmd
fix outputs

@@
 
-lares <- lapply( seq(100,1000,100) ,splitanalyze_la)
-
-lares
+lares <- lapply( seq(100,300,100) ,splitanalyze_la)
 
@@
 
-alres <- lapply( seq(100,1000,100) ,splitanalyze_al)
-
-alres
+alres <- lapply( seq(100,300,100) ,splitanalyze_al)
 
@@
 
-  lares <- as.data.frame(do.call(rbind,lares))
-  lares$p_comm <-lares$common / lares$total
-  lares$p_disc <-lares$discordant / lares$total
-  lares$p_uncom <-lares$uncommon / lares$total
-  lares
+  laout <- data.frame(total,common,discordant,uncommon)
+  laout$p_comm <-laout$common / laout$total
+  laout$p_disc <-laout$discordant / laout$total
+  laout$p_uncom <-laout$uncommon / laout$total
+  laout
 }
+
+laout <- do.call(rbind,lapply(lares,lacompare))
 
@@
 
-  alres <- as.data.frame(do.call(rbind,alres))
-  alres$p_comm <-alres$common / alres$total
-  alres$p_disc <-alres$discordant / alres$total
-  alres$p_uncom <-alres$uncommon / alres$total
-  alres
+  alout <- data.frame(total,common,discordant,uncommon)
+  alout$p_comm <-alout$common / alout$total
+  alout$p_disc <-alout$discordant / alout$total
+  alout$p_uncom <-alout$uncommon / alout$total
+  alout
 }
+
+alout <- do.call(rbind,lapply(alres,alcompare))
 

### 2023-10-01 figs/GSE158422_split.Rmd
tidy up results

@@
   lares2 <- ttenrich(m=m2,genesets=gs_symbols,cores=CORES)
+
+  list("lares1"=lares1,"lares2"=lares2)
+
 }
@@
 
+laout <- lapply(lares,lacompare)
 laout <- do.call(rbind,lapply(lares,lacompare))

### 2023-10-01 figs/GSE158422_split.Rmd
aggagg approach

@@
 # GMEA FUNCTIONS
+
+There are three approaches.
+First, limma can be done first, followed by aggregation by gene and then enrichment
+test (called LA).
+Secondly, probe values can be aggregated by gene, followed by limma, followed by
+enrichment test (AL).
+Third, probe values can be aggregated to gene level, and then aggregated to gene set
+level followed by limma test (AA).
 
@@
 
-## AL: Aggregate limma
+## AL: Aggregate limma functions
 
@@
 
-#ttenrich is same
+```
+
+## AA Aggregate-aggregate-limma functions
+
+```{r,aafunc}
+
+gsagg <- function(x,genesets,cores=1) {
+  meds <- mclapply(1:length(genesets), function(i) {
+    gs = genesets[[i]]
+    xx <- x[rownames(x) %in% gs,]
+    med <- apply(xx,2,median)
+  },mc.cores=cores)
+  mymed <- do.call(rbind,meds)
+  rownames(mymed) <- names(genesets)
+  as.data.frame(mymed)
+}
+
+aalimma <- function(agag,design) {
+  fit.reduced <- lmFit(agag,design)
+  fit.reduced <- eBayes(fit.reduced)
+  dmagg <- topTable(fit.reduced,coef=ncol(design), number = Inf)
+  return(dmagg)
+}
+
+aal <- function(mval,myann,genesets,design,cores=1) {
+  medf <- chragg(mval,myann,cores=CORES)
+  agag <- gsagg(x=medf,genesets=gs_symbols,cores=CORES)
+  aal <- aalimma(agag=agag,design=design)
+  return(aal)
+}
+
+aalres <- aal(mval=mval,myann=myann,genesets=gs_symbols,design=design,cores=CORES)
 
@@
 
+laout
+
+alout
+
 ```

### 2023-10-01 figs/GSE158422_split.Rmd
aal

@@
 
+## AA workflow split and analyse
+
+```{r,aaflow}
+
+splitanalyze_aa <- function(seed){
+  set.seed(seed)
+  a <- sample(1:37,18)
+  a <- a[order(a)]
+  nona <- setdiff(1:37,a)
+  set.seed(seed)
+  b <- sample(nona,18)
+  b <- b[order(b)]
+  a <- c( (a*2)-1 ,a*2 )
+  a <- a[order(a)]
+  b <- c( (b*2)-1 ,b*2 )
+  b <- b[order(b)]
+
+  design_a <- design[a,]
+  design_a <- design_a[,colSums(design_a)>0]
+  mval_a <- mval[,a]
+  design_b <- design[b,]
+  design_b <- design_b[,colSums(design_b)>0]
+  mval_b <- mval[,b]
+
+  #aa1
+  aares1 <- aal(mval=mval_a,myann=myann,genesets=gs_symbols,
+    design=design_a,cores=CORES)
+
+  #aa2
+  aares2 <- aal(mval=mval_b,myann=myann,genesets=gs_symbols,
+    design=design_b,cores=CORES)
+
+  list("aares1"=aares1,"aares2"=aares2)
+}
+
+aares <- lapply( seq(100,300,100) ,splitanalyze_aa)
+
+```
+
 ## Compare
 
-```{r,compare}
+```{r,summarise1}
 
@@
 }
-
-laout <- lapply(lares,lacompare)
-laout <- do.call(rbind,lapply(lares,lacompare))
 
@@
 
+aacompare <-function(aares){
+  aares1 <- aares[[1]]
+  aares2 <- aares[[2]]
+  aasig1 <- subset(aares1,adj.P.Val<0.05)
+  aasig2 <- subset(aares2,adj.P.Val<0.05)
+
+  aasig1_up <- rownames(subset(aasig1,logFC>0))
+  aasig1_dn <- rownames(subset(aasig1,logFC<0))
+  aasig2_up <- rownames(subset(aasig2,logFC>0))
+  aasig2_dn <- rownames(subset(aasig2,logFC<0))
+
+  total <- length(unique(c(aasig1_up,aasig2_up,aasig1_dn,aasig2_dn)))
+  common <- length(intersect(aasig1_up, aasig2_up)) + length(intersect(aasig1_dn, aasig2_dn))
+  discordant <- length(intersect(aasig1_up, aasig2_dn)) + length(intersect(aasig1_dn, aasig2_up))
+  uncommon = total - common - discordant
+  result <- c("total"=total,"common"=common,"discordant"=discordant,"uncommon"=uncommon)
+
+  aaout <- data.frame(total,common,discordant,uncommon)
+  aaout$p_comm <- aaout$common / aaout$total
+  aaout$p_disc <- aaout$discordant / aaout$total
+  aaout$p_uncom <-aaout$uncommon / aaout$total
+  aaout
+}
+
+```
+
+```{r,compare1}
+
+laout <- do.call(rbind,lapply(lares,lacompare))
+
 alout <- do.call(rbind,lapply(alres,alcompare))
+
+aaout <- do.call(rbind,lapply(aares,aacompare))
 
@@
 alout
+
+aaout
 

### 2023-10-02 figs/GSE158422_split.Rmd
working nicely

@@
 
-CORES=12
+CORES=28
 
@@
 
-lares <- lapply( seq(100,300,100) ,splitanalyze_la)
+lares <- lapply( seq(100,5000,100) ,splitanalyze_la)
 
@@
 
-alres <- lapply( seq(100,300,100) ,splitanalyze_al)
+alres <- lapply( seq(100,5000,100) ,splitanalyze_al)
 
@@
 
-aares <- lapply( seq(100,300,100) ,splitanalyze_aa)
+aares <- lapply( seq(100,5000,100) ,splitanalyze_aa)
 
@@
 
+res <- do.call(rbind,lapply(list(laout,alout,aaout),colMeans))
+
+rownames(res) <- c("la","al","aa")
+
+res <- as.data.frame(res)
+res %>%
+  kbl(caption="overall summary") %>%
+  kable_paper("hover", full_width = F)
+
+```
+
+Plot some data.
+
+```{r,plots1,fig.height=5,fig.width=9}
+
+par(mfrow=c(1,3))
+barplot(res$p_comm,main="proportion common",names.arg=c("la","al","aa"))
+barplot(res$p_uncom,main="proportion uncommon",names.arg=c("la","al","aa"))
+barplot(res$p_disc,main="proportion discordant",names.arg=c("la","al","aa"))
+
 ```

### 2023-10-02 figs/GSE158422_split.Rmd
output plots

@@
     mymean <- mean(tstats)
-    mymedian <- median(tstats)
+    mymedian <- mean(tstats)
     if ( length(tstats) > 2 ) {
@@
       mymean <- mean(tstats)
-      mymedian <- median(tstats)
+      mymedian <- mean(tstats)
       wt <- t.test(tstats)
@@
       b <- mx[rows,]
-      med <- apply(b,2,median)
+      med <- apply(b,2,mean)
       med <- matrix(med,nrow=1)
@@
     xx <- x[rownames(x) %in% gs,]
-    med <- apply(xx,2,median)
+    med <- apply(xx,2,mean)
   },mc.cores=cores)
@@
 res <- as.data.frame(res)
+
 res %>%
@@
   kable_paper("hover", full_width = F)
+
+apply(laout,2,median)
+apply(alout,2,median)
+apply(aaout,2,median)
 
@@
 
+Raw counts and proportions.
+
 ```{r,plots1,fig.height=5,fig.width=9}
@@
 par(mfrow=c(1,3))
+
+barplot(res$common,main="no. common",names.arg=c("la","al","aa"))
+barplot(res$uncommon,main="no. uncommon",names.arg=c("la","al","aa"))
+barplot(res$discordant,main="no. discordant",names.arg=c("la","al","aa"))
+
+boxplot(list("la"=laout$common,"al"=alout$common,"aa"=aaout$common),
+  main="no. common",col="white",ylim=c(0,1550))
+beeswarm(list("la"=laout$common,"al"=alout$common,"aa"=aaout$common), add=TRUE,cex=1.2,pch=19)
+
+boxplot(list("la"=laout$uncommon,"al"=alout$uncommon,"aa"=aaout$uncommon),
+  main="no. uncommon",col="white",ylim=c(0,1550))
+beeswarm(list("la"=laout$uncommon,"al"=alout$uncommon,"aa"=aaout$uncommon), add=TRUE,cex=1.2,pch=19)
+
+boxplot(list("la"=laout$discordant,"al"=alout$discordant,"aa"=aaout$discordant),
+  main="no. discordant",col="white",ylim=c(0,1550))
+beeswarm(list("la"=laout$discordant,"al"=alout$discordant,"aa"=aaout$discordant), add=TRUE,cex=1.2,pch=19)
+
 barplot(res$p_comm,main="proportion common",names.arg=c("la","al","aa"))
@@
 barplot(res$p_disc,main="proportion discordant",names.arg=c("la","al","aa"))
+
+boxplot(list("la"=laout$p_comm,"al"=alout$p_comm,"aa"=aaout$p_comm),
+  main="proportion common",col="white",ylim=c(0,1))
+beeswarm(list("la"=laout$p_comm,"al"=alout$p_comm,"aa"=aaout$p_comm), add=TRUE,cex=1.2,pch=19)
+
+boxplot(list("la"=laout$p_uncom,"al"=alout$p_uncom,"aa"=aaout$p_uncom),
+  main="proportion uncommon",col="white",ylim=c(0,1))
+beeswarm(list("la"=laout$p_uncom,"al"=alout$p_uncom,"aa"=aaout$p_uncom), add=TRUE,cex=1.2,pch=19)
+
+boxplot(list("la"=laout$p_disc,"al"=alout$p_disc,"aa"=aaout$p_disc),
+  main="proportion discordant",col="white",ylim=c(0,1))
+beeswarm(list("la"=laout$p_disc,"al"=alout$p_disc,"aa"=aaout$p_disc), add=TRUE,cex=1.2,pch=19)
 
@@
 
-save.image("GSE158422_gmea.Rdata")
+save.image("GSE158422_split.Rdata")
 

### 2023-10-03 figs/GSE158422_split.Rmd
progress

@@
 
-CORES=28
+CORES=8
 
@@
   agag <- gsagg(x=medf,genesets=gs_symbols,cores=CORES)
-  aal <- aalimma(agag=agag,design=design)
-  return(aal)
+  aalres <- aalimma(agag=agag,design=design)
+  return(aalres)
 }
 
-aalres <- aal(mval=mval,myann=myann,genesets=gs_symbols,design=design,cores=CORES)
+#aalres <- aal(mval=mval,myann=myann,genesets=gs_symbols,design=design,cores=CORES)
 
@@
   dm1 <- runlimma(mval_a,design_a,myann)
-  dmagg1 <- agg(dm1,cores=CORES)
+  dmagg1 <- agg(dm1,cores=floor(CORES/3))
   m1 <- dmagg1$gme_res_df[,"mean",drop=FALSE]
@@
   dm2 <- runlimma(mval_b,design_b,myann)
-  dmagg2 <- agg(dm2,cores=CORES)
+  dmagg2 <- agg(dm2,cores=floor(CORES/3))
   m2 <- dmagg2$gme_res_df[,"mean",drop=FALSE]
@@
 
-lares <- lapply( seq(100,5000,100) ,splitanalyze_la)
+lares <- mclapply( seq(100,5000,100) ,splitanalyze_la,mc.cores=CORES)
 
@@
   #al
-  medf1 <- chragg(mval_a,myann,cores=CORES)
+  medf1 <- chragg(mval_a,myann,cores=floor(CORES/3))
   magg1 <- agglimma(medf1,design_a)
@@
   #al
-  medf2 <- chragg(mval_b,myann,cores=CORES)
+  medf2 <- chragg(mval_b,myann,cores=floor(CORES/3))
   magg2 <- agglimma(medf2,design_b)
@@
 
-alres <- lapply( seq(100,5000,100) ,splitanalyze_al)
+alres <- mclapply( seq(100,5000,100) ,splitanalyze_al,mc.cores=CORES)
 
@@
   aares1 <- aal(mval=mval_a,myann=myann,genesets=gs_symbols,
-    design=design_a,cores=CORES)
+    design=design_a,cores=floor(CORES/3))
 
@@
   aares2 <- aal(mval=mval_b,myann=myann,genesets=gs_symbols,
-    design=design_b,cores=CORES)
+    design=design_b,cores=floor(CORES/3))
 
@@
 
-aares <- lapply( seq(100,5000,100) ,splitanalyze_aa)
+aares <- mclapply( seq(100,5000,100) ,splitanalyze_aa,mc.cores=CORES)
 
@@
 
-aacompare <-function(aares){
-  aares1 <- aares[[1]]
-  aares2 <- aares[[2]]
-  aasig1 <- subset(aares1,adj.P.Val<0.05)
-  aasig2 <- subset(aares2,adj.P.Val<0.05)
+aacompare <-function(aa){
+  aa1 <- aa[[1]]
+  aa2 <- aa[[2]]
+  aasig1 <- subset(aa1,adj.P.Val<0.05)
+  aasig2 <- subset(aa2,adj.P.Val<0.05)
 
@@
 
-barplot(res$common,main="no. common",names.arg=c("la","al","aa"))
-barplot(res$uncommon,main="no. uncommon",names.arg=c("la","al","aa"))
-barplot(res$discordant,main="no. discordant",names.arg=c("la","al","aa"))
+barplot(res$common,main="no. common",names.arg=c("LA","AL","AA"))
+barplot(res$uncommon,main="no. uncommon",names.arg=c("LA","AL","AA"))
+barplot(res$discordant,main="no. discordant",names.arg=c("LA","AL","AA"))
 
-boxplot(list("la"=laout$common,"al"=alout$common,"aa"=aaout$common),
+boxplot(list("LA"=laout$common,"AL"=alout$common,"AA"=aaout$common),
   main="no. common",col="white",ylim=c(0,1550))
-beeswarm(list("la"=laout$common,"al"=alout$common,"aa"=aaout$common), add=TRUE,cex=1.2,pch=19)
+beeswarm(list("LA"=laout$common,"AL"=alout$common,"AA"=aaout$common), add=TRUE,cex=1.2,pch=19)
 
-boxplot(list("la"=laout$uncommon,"al"=alout$uncommon,"aa"=aaout$uncommon),
+boxplot(list("LA"=laout$uncommon,"AL"=alout$uncommon,"AA"=aaout$uncommon),
   main="no. uncommon",col="white",ylim=c(0,1550))
-beeswarm(list("la"=laout$uncommon,"al"=alout$uncommon,"aa"=aaout$uncommon), add=TRUE,cex=1.2,pch=19)
+beeswarm(list("LA"=laout$uncommon,"AL"=alout$uncommon,"AA"=aaout$uncommon), add=TRUE,cex=1.2,pch=19)
 
-boxplot(list("la"=laout$discordant,"al"=alout$discordant,"aa"=aaout$discordant),
+boxplot(list("LA"=laout$discordant,"AL"=alout$discordant,"AA"=aaout$discordant),
   main="no. discordant",col="white",ylim=c(0,1550))
-beeswarm(list("la"=laout$discordant,"al"=alout$discordant,"aa"=aaout$discordant), add=TRUE,cex=1.2,pch=19)
+beeswarm(list("LA"=laout$discordant,"AL"=alout$discordant,"AA"=aaout$discordant), add=TRUE,cex=1.2,pch=19)
 
-barplot(res$p_comm,main="proportion common",names.arg=c("la","al","aa"))
-barplot(res$p_uncom,main="proportion uncommon",names.arg=c("la","al","aa"))
-barplot(res$p_disc,main="proportion discordant",names.arg=c("la","al","aa"))
+barplot(res$p_comm,main="proportion common",names.arg=c("LA","AL","AA"))
+barplot(res$p_uncom,main="proportion uncommon",names.arg=c("LA","AL","AA"))
+barplot(res$p_disc,main="proportion discordant",names.arg=c("LA","AL","AA"))
 
-boxplot(list("la"=laout$p_comm,"al"=alout$p_comm,"aa"=aaout$p_comm),
+boxplot(list("LA"=laout$p_comm,"AL"=alout$p_comm,"AA"=aaout$p_comm),
   main="proportion common",col="white",ylim=c(0,1))
-beeswarm(list("la"=laout$p_comm,"al"=alout$p_comm,"aa"=aaout$p_comm), add=TRUE,cex=1.2,pch=19)
+beeswarm(list("LA"=laout$p_comm,"AL"=alout$p_comm,"AA"=aaout$p_comm), add=TRUE,cex=1.2,pch=19)
 
-boxplot(list("la"=laout$p_uncom,"al"=alout$p_uncom,"aa"=aaout$p_uncom),
+boxplot(list("LA"=laout$p_uncom,"AL"=alout$p_uncom,"AA"=aaout$p_uncom),
   main="proportion uncommon",col="white",ylim=c(0,1))
-beeswarm(list("la"=laout$p_uncom,"al"=alout$p_uncom,"aa"=aaout$p_uncom), add=TRUE,cex=1.2,pch=19)
+beeswarm(list("LA"=laout$p_uncom,"AL"=alout$p_uncom,"AA"=aaout$p_uncom), add=TRUE,cex=1.2,pch=19)
 
-boxplot(list("la"=laout$p_disc,"al"=alout$p_disc,"aa"=aaout$p_disc),
+boxplot(list("LA"=laout$p_disc,"AL"=alout$p_disc,"AA"=aaout$p_disc),
   main="proportion discordant",col="white",ylim=c(0,1))
-beeswarm(list("la"=laout$p_disc,"al"=alout$p_disc,"aa"=aaout$p_disc), add=TRUE,cex=1.2,pch=19)
+beeswarm(list("LA"=laout$p_disc,"AL"=alout$p_disc,"AA"=aaout$p_disc), add=TRUE,cex=1.2,pch=19)
 

### 2023-10-03 manuscript/manuscript.Rmd
progress

---
title: "Functional class scoring-based enrichment tests of Infinium DNA methylation data"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
bibliography: gmea.bib
csl: plos-computational-biology.csl
---

<strong>
Mark Ziemann<sup>1,2*</sup>,
Mandhri Abeysooriya<sup>2,3*</sup>,
Anusuiya Bora<sup>2</sup>,
Severin Lamon<sup>3</sup>,
Yen Wong<sup>4</sup>,
Jeffrey Craig<sup>4</sup>
</strong>

1. Burnet Institute, Melbourne, Australia

2. Deakin University, School of Life and Environmental Sciences, Geelong, Australia.

3. Deakin University, School of Exercise and Nutrition, Burwood, Australia.

4. Deakin University, School of Medicine, Geelong, Australia.

(*) Correspondence: m.ziemann@deakin.edu.au

| Author | ORCID |
| --- | --- |
| Mark Ziemann | 0000-0002-7688-6974 |
| Anusuiya Bora | 0009-0006-2908-1352 |

Source: https://github.com/markziemann/gmea

<br>

## Abstract

* Infinium methylation beadarrays remain one of the most popular techniques for studying genomic
methylation, but tools for downstream pathways analysis remain limited.

* Functional class scoring (FCS) is a group of pathway enrichment techniques that involve the ranking of
genes and evaluation of their collective regulation in biological systems.

* Here we propose three different approaches to apply FCS to Infinium data, which vary upon when the
differential methylation analysis is conducted, at the level of probes, genes or gene sets.

* Using real EPIC array data, we demonstrate that aggregation of probe methylation to gene level before
differential methylation analysis yields a high degree of true positives and low number of contradictory
results.

* Using matched RNA-seq data we evaluate the relationship of methylation differences at promoters and
gene bodies with RNA expression at the level of pathways.

* We apply this approach to two further contexts, where existing methods failed to identify any
pathway-level effects in 450K and EPIC datasets: (i) the effect of assisted reproductive technology on
DNA methylation, and (2) the association of elevated plasma homocysteine with methylation patterns.

* The methods described here have been incorporated into a package called gmea, short for gene
methylation enrichment analysis.

**Keywords:** Pathway analysis, functional enrichment analysis, Infinium Array, DNA methylation,
epigenetics.

<br>

## Introduction

* DNA methylation is the most widely studied epigenetic mark, for its role in development and
disease.

* Many hundreds of epigenome-wide association studies (EWASs) are conducted each year to understand DNA
methylation patterns in disease and other contexts.

* Infinium arrays remain the preferred platform for genomic methylation profiling due to relatively
low cost and analytical simplicity as compared to high throughput methylation sequencing.

* Infinium arrays have multiple probes per gene in different locations including promoters, CpG islands,
gene bodies and enhancers, which complicates downstream functional enrichment analysis, otherwise known
as pathway analysis or gene set enrichment analysis.

* Enrichment analysis comes in two popular types; over-representation analysis (ORA) and functional
class scoring (FCS).
In ORA, genes with measurements that meet an arbitrary significance threshold are selected and compared
to a background list consisting of genes measured in the assay.
The test seeks to identify sets of genes (eg: ontologies, pathways) that are over-represented in the
gene list of interest relative to the background.
Functional class scoring takes a different approach by ranking all detected genes by a differential
regulation score (eg: fold change, confident effect size) followed by a test to assess whether each
set of genes has distribution of scores that is different to the null.
Currently there are no specialised approaches for the application of FCS to Infinium array data.

* Here, we evaluate different approaches to apply FCS to Infinium array.
These methods vary in the sequence of data aggregation and differential testing steps which are
required to go from probe-level methylation measurements to evidence of pathway-level differential
methylation.

## Methods

## Results

## Discussion

## Bibliography

### 2023-10-03 manuscript/manuscript.Rmd
writing starts

@@
 
+### Implementation overview
+
+Functional enrichment analysis is a process of data summarisation from genes to gene sets (pathways).
+This is made more complicated for Infinium array data due to the presence of multiple probes per gene,
+meaning the data needs to be summarised from probes to genes to gene sets.
+
+We propose three different approaches for applying enrichment analysis depending on when the differential
+methylation step is conducted (**Figure 1**).
+
+1. <strong>Differential methylation analysis is conducted on probes and these results are aggregated
+to the gene and gene set level (called "LA").<.strong>
+In this approach differential analysis was conduced on probe values using limma.
+Differential probe methylation results (t-statistics) were then aggregated to the gene level by median.
+These median t-statistics were subject to downstream enrichment analysis using a one-sample two-sided
+t-test (a self-contained test) to estimate whether the t-statistic distribution of the gene set varies
+from zero.
+The use of a self-contained test is warranted as global hyper- or hypo-methylation is common between
+conditions, where competitive tests may give misleading results.
+The use of parametric tests is also warranted, as probe t-statistics typically exhibit a normal
+distribution, even in the presence of global hyper- or hypo-methylation.
+
+2. <strong>Differential methylation analysis conducted after probe-level methylation is aggregated to
+gene-level (called "AL").</strong>
+As each gene may have many probes, these individual data points were summarised into a single
+value for each individual.
+We used the mean of M-values for each probe belonging to a gene, which generated a new gene-level
+M-value matrix.
+This gene-level methylation matrix underwent differential analysis using limma, followed by an
+enrichment test as above (one-sample two-tailed t-test).
+
+3. <strong>Methylation values are aggregated to the gene set level for differential testing
+(called "AA").</strong>
+M-values for each probe were aggregated to the gene level by mean, and then aggregated again to the gene
+set level by mean or median.
+The resulting matrix of aggregated M-values underwent differential analysis with limma and the results
+including log fold changes and significance values were used to infer pathway regulation.
+This approach has a caveat in that it may partly violate the assumption of independent sampling, as
+it is known that some genes are represented in multiple gene sets.
+
+### Method validation
+
+As there are no established methods for benchmarking Infinium array based pathway regulation using
+known ground truth synthetic data, we used a split-and-analyse approach with a real dataset to
+evaluate the three approaches outlined above.
+We downloaded raw intensity EPIC IDAT files from GEO for study GSE158422 which consists of
+lung tumour and normal adjacent tissues from 37 patients.
+Probe annotations were obtained from the "IlluminaHumanMethylationEPICanno.ilm10b4.hg19" Bioconductor
+package.
+Reactome gene sets were downloaded from the MSigDB website.
+From this set of samples, 18 patients were randomly selected into analysis A, where control and tumour
+samples were compared.
+From the remaining 19 samples, 18 were randomly selected into analysis B, where control and tumour
+samples were compared as above to act as an independent validation of analysis A.
+Analyses A and B were conducted using one of the three enrichment approaches outlined above, and the
+gene sets meeting a 5% FDR threshold in the hyper- and hypo-methylated directions were saved.
+Gene sets identified to be common in A and B analyses (FDR<0.05 and direction) were considered validated,
+gene sets identified in only one method were not considered validated, and gene sets that gave discordant
+direction (hyper- vs hypo-methylation) between A and B were classed as contradictory.
+This process was repeated 50 times to rigourously evaluate method performance.
+
+### Sensitivity analysis
+
+The same dataset including all 37 patients underwent analysis using the LA method, gsameth and ebGSEA.
+The effect size (enrichment score) of significantly differentially methylated pathways (FDR<0.05) was
+recorded.
+Then, a random subset of n patients was selected followed by pathway enrichment analysis with the
+respective method.
+The subset results were compared to the full n=37 results to test the sensitivity of these enrichment
+methods at differenent effect size levels.
+This process was repeated 20 times.
+
+### Association of methylation with gene expression
+
+As methylation at promoters and gene bodies is considered to have differing association with gene
+expression, we conducted LA analysis at gene promoters, gene bodies and whole gene separately and
+examined the relationship with RNA-seq based pathway regulation.
+RNA-seq counts were downloaded from the GEO database (Accession: GSE158420).
+HGNChelper was used to repair transmogrified gene names, followed by DESeq2 based differential analysis.
+Enrichment analysis of RNA-seq data was conducted using the mitch package with REACTOME gene sets.
+
+### Pathway-level differential methylation in assisted reproductive technology
+(i) the effect of assisted reproductive technology DNA methylation
+
+### Pathway-level differential methylation in hyperhomocysteinemia
+(2) the association of elevated plasma homocysteine with methylation patterns.
+
 ## Results
+
+### Evaluation of FCS methods for Infinium methylation array data
+
+We applied the three distinct approaches ("LA", "AL" and "AA") for pathway level differential
+methylation analysis to a dataset consisting of lung tumour and adjacent normal tissue from 37 patients.
+The dataset was split randomly into two groups of 18 patients and the overlap between subset was used to
+assess the consistency and validity of findings with 50 simulations.
+From the 1550 Reactome gene sets in the dataset, the AL method recorded the highest number of
+"valid" pathways with an average of 919.8 being identified in both subsets, which was much higher than
+LA and AA methods with 295.3 and 276.7 gene sets respectively (Figure 1A).
+The AL method also recorded the lowest number of uncommon sets (382.0) which was better than the LA and
+AA methods with 634.8 and 908.3 respectively (Figure 1B).
+We also noticed that the LA method sometimes yielded contradictory results, so we quantified the number
+of discordant observations, that is statistically significant but in the opposite direction.
+True to this observation, the LA method gave 461.6 discordant pathways which indicates a high rate of
+false-positives.
+Meanwhile the AL and AA methods gave a mean of 0.1 and 0.02 discordant pathways respectively.
+Taken together, these results indicate that the AL method gives the highest rate of common results
+suggesting good accuracy in terms of sensitivity and specificity in this context.
+
+### Comparision of AL method to existing pathway enrichment approaches
 

### 2023-10-03 figs/GSE158422_split.Rmd
finalised

@@
     mymean <- mean(tstats)
-    mymedian <- mean(tstats)
+    mymedian <- median(tstats)
     if ( length(tstats) > 2 ) {
@@
       mymean <- mean(tstats)
-      mymedian <- mean(tstats)
+      mymedian <- median(tstats)
       wt <- t.test(tstats)
@@
 
+Median value works well here.
+
 ```{r,aafunc}
@@
     xx <- x[rownames(x) %in% gs,]
-    med <- apply(xx,2,mean)
+    med <- apply(xx,2,median)
   },mc.cores=cores)
@@
 
-apply(laout,2,median)
-apply(alout,2,median)
-apply(aaout,2,median)
+apply(laout,2,mean)
+apply(alout,2,mean)
+apply(aaout,2,mean)
 

### 2023-10-03 figs/GSE158422_split.Rmd
multicore management

@@
 
+# optimised for 128 GB sever with 32 threads
 CORES=8
@@
 aal <- function(mval,myann,genesets,design,cores=1) {
-  medf <- chragg(mval,myann,cores=CORES)
-  agag <- gsagg(x=medf,genesets=gs_symbols,cores=CORES)
+  medf <- chragg(mval,myann,cores=cores)
+  agag <- gsagg(x=medf,genesets=gs_symbols,cores=cores)
   aalres <- aalimma(agag=agag,design=design)
@@
   m1 <- dmagg1$gme_res_df[,"mean",drop=FALSE]
-  lares1 <- ttenrich(m=m1,genesets=gs_symbols,cores=CORES)
+  lares1 <- ttenrich(m=m1,genesets=gs_symbols,cores=CORES/3)
 
@@
   m2 <- dmagg2$gme_res_df[,"mean",drop=FALSE]
-  lares2 <- ttenrich(m=m2,genesets=gs_symbols,cores=CORES)
+  lares2 <- ttenrich(m=m2,genesets=gs_symbols,cores=CORES/3)
 
@@
   colnames(m1) <- "t"
-  alres1 <- ttenrich(m=m1,genesets=gs_symbols,cores=CORES)
+  alres1 <- ttenrich(m=m1,genesets=gs_symbols,cores=CORES/3)
 
@@
   colnames(m2) <- "t"
-  alres2 <- ttenrich(m=m2,genesets=gs_symbols,cores=CORES)
+  alres2 <- ttenrich(m=m2,genesets=gs_symbols,cores=CORES/3)
 
@@
 
-aares <- mclapply( seq(100,5000,100) ,splitanalyze_aa,mc.cores=CORES)
+aares <- mclapply( seq(100,300,100) ,splitanalyze_aa,mc.cores=CORES)
 

### 2023-10-04 figs/GSE158422_sensitivity.Rmd
moving on to sensitivity analysis

--- 2023-09-25 figs/GSE158422_gmea.Rmd
@@
 ---
-title: "GSE158422: DNA methylation in Lung squamous cell carcinoma - gene methylation enrichment analysis"
+title: "GSE158422: DNA methylation in Lung squamous cell carcinoma - gene methylation enrichment analysis - sensitivity analysis"
 author: "The GMEA team"
@@
 
-Functional class sorting is widely used for pathway analysis using tools like GSEA,
-yet there is no consensus in how this can be conducted for Illumina Infinium
-methylation array data.
+In the previous analysis, it was shown that the "AL" (aggregation-limma-enrichment) method
+yielded good performance with a cohort of cancer-normal tissue pairs.
 
-Here we propose a simple approach which involves the following:
+In this piece of work, we want to ascertain the sensitivity of this approach as compared to
+existing methods such as gsameth and ebGSEA.
 
-1. Limma test on probes.
-
-2. For each gene, calculate the median t-statistic from step 1.
-
-3. Use this median t-stat in Camera pre-ranked test for gene sets.
-
-In this example, I'm using matched infinium EPIC 850k data from (n=37) normal and lung cancer samples (GSE158422).
-The data was previously preprocessed and normalised using the minfi package (see the folder called "misc").
-
-Here the gene sets are obtained from Reactome.
+To achieve this we will analyse with the full 37 patient sample pairs, and repeatedly downsample
+the dataset down to n=3 patients.
+What we expect is that the "AL" method could have a higher degree of recall as compared to these
+other approaches.
+We expect that "AL" could identify gene sets with larger effect sizes (enrichment scores) using
+fewer samples.
+This means that for a given set of datasets, this approach could yield more findings, or
+alternatively if pathway analysis is the primary outcome of a study, it need not be as large which
+can save money and difficulty with establishing large sample collections.
 
@@
 
-This analysis was run on a 8C/16T computer with 64GB RAM running at 3.8 GHz.
-This workflow used 34 GB RAM and took
+This analysis was run on a 16C/32T computer with 128GB RAM running at 4.2 GHz.
 RAM usage can be moderated by reducing the parallel core count.
@@
 
-## GMEA
+# AL GMEA: all probes
 
-There are two obvious ways to perform GMEA:
+Start with curating the sample info.
 
-1. limma-agg-1stt (la1).
-In this approach, the limma test is conducted on probes, followed by aggregation of the differential
-methylation t-scores. Each gene gets a median t-stat value which is evaluated downstream by a 1 sample
-t-test.
-
-2. agg-limma-1stt (al1).
-Here, the probe beta values for each gene are aggregated to a median and then limma is run on these
-values.
-To determine gene set enrichment, a 1-sample t-test is conducted.
-
-## LA: limma aggregate 1-sample t-test
-
-```{r,probelimma1}
+```{r,samplesheet}
 
@@
 
-fit.reduced <- lmFit(mval,design)
-fit.reduced <- eBayes(fit.reduced)
-summary(decideTests(fit.reduced))
-dm <- topTable(fit.reduced,coef=4, number = Inf)
-dm <- merge(myann,dm,by=0)
-dm <- dm[order(dm$P.Value),]
-rownames(dm) <- dm$Row.names
-dm$Row.names=NULL
+```
 
-head(dma) %>%
-  kbl(caption="Top DM probes") %>%
-  kable_styling("hover",full_width=FALSE)
+## Functions
 
-agg <- function(dm,cores=1) {
-  gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
-  gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
+```{r,functions}
+
+# magg is a general function that aggregates probe methylation measurements by gene
+# in this variation it aggregates by MEAN, which performs better than median
+magg <- function(mval,myann,cores=1){
+  gn <- unique(unlist(strsplit( myann$UCSC_RefGene_Name ,";")))
+  gnl <- strsplit( myann$UCSC_RefGene_Name ,";")
   gnl <- mclapply(gnl,unique,mc.cores=cores)
-  dm$UCSC_RefGene_Name <- gnl
-  l <- mclapply(1:nrow(dm), function(i) {
-    a <- dm[i,]
-    len <- length(a[[1]][[1]])
-    tvals <- as.numeric(rep(a["t"],len))
-    genes <- a[[1]][[1]]
-    data.frame(genes,tvals)
-  },mc.cores=cores)
-  df <- do.call(rbind,l)
-  keep <- names(which(table(df$genes)>1))
-  df <- df[df$genes %in% keep,]
-  gn <- unique(df$genes)
-  gme_res <- lapply( 1:length(gn), function(i) {
-    g <- gn[i]
-    tstats <- df[which(df$genes==g),"tvals"]
-    myn <- length(tstats)
-    mymean <- mean(tstats)
-    mymedian <- median(tstats)
-    if ( length(tstats) > 2 ) {
-      ttest <- t.test(tstats)
-      pval <- ttest$p.value
-    } else {
-      pval = 1
+  myann$gnl <- gnl
+  keep <- rownames(subset(myann,UCSC_RefGene_Name!=""))
+  mx <- mval[rownames(mval) %in% keep,]
+  mymed <- function(g) {
+    probes <- rownames(myann[grep(g,myann$gnl),])
+    rows <- which(rownames(mx) %in% probes)
+    if ( length(rows) > 1 ) {
+      b <- mx[rows,]
+      med <- apply(b,2,mean)
+      med <- matrix(med,nrow=1)
+      colnames(med) <- colnames(b)
+      rownames(med) <- g
+      return(med)
     }
-    res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,
-      "median"=mymedian, pval=pval)
-  } )
-  gme_res_df <- do.call(rbind, gme_res)
-  rownames(gme_res_df) <- gme_res_df[,1]
-  gme_res_df <- gme_res_df[,-1]
-  tmp <- apply(gme_res_df,2,as.numeric)
-  rownames(tmp) <- rownames(gme_res_df)
-  gme_res_df <- as.data.frame(tmp)
-  gme_res_df$sig <- -log10(gme_res_df[,4])
-  gme_res_df <- gme_res_df[order(-gme_res_df$sig),]
-  gme_res_df$fdr <- p.adjust(gme_res_df$pval)
-  out <- list("df"=df,"gme_res_df"=gme_res_df)
-  return(out)
+  }
+  med <- mclapply(gn,mymed,mc.cores=cores)
+  med <- med[lapply(med,length)>0]
+  medf <- do.call(rbind,med)
+  return(medf)
 }
 
-tic()
-dmagg <- agg(dm,cores=CORES)
-toc()
+# chragg is a specialised function that performs the magg routine on chromosomes in parallel
+chragg <- function(mval,myann,cores=1){
+  annodf <- as.data.frame(anno)
+  keep <- rownames(subset(myann,UCSC_RefGene_Name!=""))
+  mx <- mval[rownames(mval) %in% keep,]
+  chrs <- unique(anno$chr)
+  myorder <- unlist(lapply(chrs,function(mychr) { nrow( annodf[annodf$chr==mychr,] ) } ))
+  chrs <- chrs[order(-myorder)]
+  leadercores <- floor(sqrt(cores))
+  workercores <- ceiling(sqrt(cores))
+  chrmedf <- mclapply(chrs,function(chr) {
+    chrfrag <- annodf[annodf$chr==chr,]
+    chrprobes <-rownames(chrfrag)
+    chrmx <- mx[rownames(mx) %in% chrprobes,]
+    chranno <- myann[rownames(myann) %in% chrprobes,]
+    chrmedf <- magg(mval=chrmx,myann=chranno,cores=workercores)
+    return(chrmedf)
+  },mc.cores=leadercores)
+  medf <- do.call(rbind, chrmedf)
+  return(medf)
+}
 
-head(dmagg$gme_res_df,20) %>%
-  kbl(caption="top gmea genes (t-test)") %>%
-  kable_paper("hover", full_width = F)
+# agglimma is a helper function that runs limma analysis on the aggregated methylation data
+# and fixes any duplicated gene names - most significant one is retained
+agglimma <- function(medf,design) {
+  fit.reduced <- lmFit(medf,design)
+  fit.reduced <- eBayes(fit.reduced)
+  dmagg <- topTable(fit.reduced,coef=ncol(design), number = Inf)
+  nondup <- !duplicated(dmagg$ID)
+  dmagg <- dmagg[nondup,]
+  rownames(dmagg) <- dmagg$ID
+  dmagg$ID = NULL
+  return(dmagg)
+}
 
-```
-
-Check whether the distribution is even - it's not.
-
-```{r,tmeddist}
-
-hist(dmagg$gme_res_df$median,breaks=30) ; grid()
-
-```
-
-1-sample t-test for gene set enrichment.
-
-```{r,t_enrichment_test1}
-
+# ttenrich is a function designed to conduct analysis of differential methylation.
+# It takes the t-statistics from limma for a geneset and determines whether this gene
+# set has a distribution different to the null (zero) using a one-sample two-sided t-test.
 ttenrich <- function(m,genesets,cores=1) {
@@
 
-# need to get the median column before analysis
-m <- dmagg$gme_res_df[,"median",drop=FALSE]
-
-tic()
-lares <- ttenrich(m=m,genesets=gs_symbols,cores=CORES)
-toc()
-
-head(lares,30) %>%
-  kbl(caption = "Top significant genesets using LA approach") %>%
-  kable_paper("hover", full_width = F)
-
-head(lares[order(-abs(lares$t_median)),],30) %>%
-  kbl(caption = "Top effect size genesets using LA approach") %>%
-  kable_paper("hover", full_width = F)
-
-lasig <- subset(lares,fdr<0.05)
-nrow(lasig)
-
-head(lasig[order(-abs(lasig$t_median)),],30) %>%
-  kbl(caption = "Top effect size genesets after FDR filtering using LA approach") %>%
-  kable_paper("hover", full_width = F)
-
-plot(lares$t_median,-log10(lares$pval),pch=19,col="gray")
-grid()
-points(lasig$t_median,-log10(lasig$pval),pch=19,col="red")
-
 ```
 
-## AL: Aggregate limma t-test
+Run AL method on methylation data
 
-Determine the median of probe values
+```{}
 
-```{r,med1}
+# aggregation step
+medf <- chragg(mval,myann,cores=CORES)
 
-mx <- mval
+# differential analysis
+dmagg <- agglimma(medf,design)
+hist(dmagg$t)
 
-gn <- unique(unlist(strsplit( myann$UCSC_RefGene_Name ,";")))
-gnl <- strsplit( myann$UCSC_RefGene_Name ,";")
-gnl <- mclapply(gnl,unique,mc.cores=CORES)
-myann$gnl <- gnl
-
-mymed <- function(g) {
-  probes <- rownames(myann[grep(g,myann$gnl),])
-  rows <- which(rownames(mx) %in% probes)
-  if ( length(rows) > 1 ) {
-    b <- mx[rows,]
-    med <- apply(b,2,median)
-    med <- matrix(med,nrow=1)
-    colnames(med) <- colnames(b)
-    rownames(med) <- g
-    return(med)
-  }
-}
-
-med <- mclapply(gn,mymed,mc.cores=CORES)
-med <- med[lapply(med,length)>0]
-medf <- do.call(rbind,med)
-
-```
-
-Then do a limma test on median gene methylation.
-See if it is more robust than other approaches.
-
-```{r,limma1}
-
-fit.reduced <- lmFit(medf,design)
-fit.reduced <- eBayes(fit.reduced)
-summary(decideTests(fit.reduced))
-magg <- topTable(fit.reduced,coef=ncol(design), number = Inf)
-
-#write.table(dma3a,file="dma3a.tsv")
-
-head(magg, 30) %>%
+head(dmagg, 30) %>%
   kbl(caption="limma gene methylation results") %>%
   kable_paper("hover", full_width = F)
+# get t-stats
+m <- as.data.frame(dmagg$t)
+rownames(m1) <- rownames(magg1)
+colnames(m1) <- "t"
 
-```
-
-1-sample t-test for enrichment
-
-```{r,ttest}
-
-ttenrich <- function(m,genesets,cores=1) {
-  res <- mclapply( 1:length(genesets), function(i) {
-    gs <- genesets[i]
-    name <- names(gs)
-    n_members <- length(which(rownames(m) %in% gs[[1]]))
-    if ( n_members > 4 ) {
-      tstats <- m[which(rownames(m) %in% gs[[1]]),]
-      myn <- length(tstats)
-      mymean <- mean(tstats)
-      mymedian <- median(tstats)
-      wt <- t.test(tstats)
-      pval = wt$p.value
-      res <- c(name,myn,mymean,mymedian,pval)
-    }
-  } , mc.cores = cores)
-  res_df <- do.call(rbind, res)
-  rownames(res_df) <- res_df[,1]
-  res_df <- res_df[,-1]
-  colnames(res_df) <- c("n_genes","t_mean","t_median","pval")
-  tmp <- apply(res_df,2,as.numeric)
-  rownames(tmp) <- rownames(res_df)
-  res_df <- tmp
-  res_df <- as.data.frame(res_df)
-  res_df <- res_df[order(res_df$pval),]
-  res_df$logp <- -log10(res_df$pval )
-  res_df$fdr <- p.adjust(res_df$pval,method="fdr")
-  res_df[order(abs(res_df$pval)),]
-  return(res_df)
-}
-
-m <- as.data.frame(magg$t)
-rownames(m) <- rownames(magg)
-colnames(m) <- "t"
-
+# enrichment test
 alres <- ttenrich(m=m,genesets=gs_symbols,cores=CORES)
+hist(alres$t_mean)
 
@@
 
+# GMEA: promoter probes only
+
+## Limma-aggregate-1-sample t-test promoter only
+
+```{r,ls_promoter}
+
+dmp <- dm[grep("Promo",dm$Regulatory_Feature_Group),]
+
+tic()
+dmpagg <- agg(dmp,cores=CORES)
+toc()
+
+head(dmpagg$gme_res_df,20) %>%
+  kbl(caption="top gmea genes (t-test)") %>%
+  kable_paper("hover", full_width = F)
+
+```
+
 ## Session information

### 2023-10-04 figs/GSE158422_sensitivity.Rmd
working on sensitivity

@@
   library("beeswarm")
+  library("missMethyl")
 })
 
-CORES=12
+CORES=24
 
@@
 
-Start with curating the sample info.
+Start with curating the sample info and design matrix.
 
@@
 
-Run AL method on methylation data
+Run AL method on methylation data.
+Record timings.
 
-```{}
+```{r,rungmea}
 
 # aggregation step
+tic()
 medf <- chragg(mval,myann,cores=CORES)
+toc() #51.5 sec
 
 # differential analysis
+tic()
 dmagg <- agglimma(medf,design)
+toc() #0.46 sec
+
 hist(dmagg$t)
@@
   kable_paper("hover", full_width = F)
+
 # get t-stats
 m <- as.data.frame(dmagg$t)
-rownames(m1) <- rownames(magg1)
-colnames(m1) <- "t"
+rownames(m) <- rownames(dmagg)
+colnames(m) <- "t"
 
 # enrichment test
+tic()
 alres <- ttenrich(m=m,genesets=gs_symbols,cores=CORES)
+toc() #1.1 sec
+
 hist(alres$t_mean)
@@
 
+# GSAMETH
+
+Individual probes have a distribution that leans positive, but after aggregation to genes it is actually
+negative.
+
+```{r,rungsameth1}
+
+# limma
+runlimma <- function(mval,design,myann) {
+  fit.reduced <- lmFit(mval,design)
+  fit.reduced <- eBayes(fit.reduced)
+  summary(decideTests(fit.reduced))
+  dm <- topTable(fit.reduced,coef=4, number = Inf)
+  dm <- merge(myann,dm,by=0)
+  dm <- dm[order(dm$P.Value),]
+  rownames(dm) <- dm$Row.names
+  dm$Row.names=NULL
+  return(dm)
+}
+
+dm <- runlimma(mval,design,myann)
+
+hist(dm$t,breaks=100,xlim=c(-10,10)) ; grid()
+
+head(dm, 10) %>% kbl() %>% kable_paper("hover", full_width = F)
+
+sigup <- rownames(subset(dm,logFC > 0 & adj.P.Val < 0.05))
+sigdn <- rownames(subset(dm,logFC < 0 & adj.P.Val < 0.05))
+
+length(sigup)
+length(sigdn)
+
+tic()
+gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez)
+gsadn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez)
+toc() # 12.9 sec
+
+gsaup <- gsaup[order(gsaup$P.DE),]
+head(gsaup,10) %>% kbl(caption="Top hypermethylated pathways") %>% kable_paper("hover", full_width = F)
+
+gsadn <- gsadn[order(gsadn$P.DE),]
+head(gsadn,10) %>% kbl(caption="Top hypomethylated pathways") %>% kable_paper("hover", full_width = F)
+
+gsig_up <-  rownames(subset(gsaup,FDR<0.05))
+gsig_dn <-  rownames(subset(gsadn,FDR<0.05))
+
+```
+
 ## Overlap
@@
 
-lasig_up <- rownames(subset(lasig,t_median>0))
-lasig_dn <- rownames(subset(lasig,t_median<0))
-
 alsig_up <- rownames(subset(alsig,t_median>0))
@@
 
-v1 <- list("LA up"=lasig_up, "LA dn"=lasig_dn,
-  "AL up"=alsig_up,"AL dn"=alsig_dn)
+gsig_up <-  rownames(subset(gsaup,FDR<0.05))
+gsig_dn <-  rownames(subset(gsadn,FDR<0.05))
+
+v1 <- list("AL up"=alsig_up,"AL dn"=alsig_dn,
+  "GSAmeth up"=gsig_up,"GSAmeth dn"=gsig_dn)
 
@@
 
-tic()
-dmpagg <- agg(dmp,cores=CORES)
-toc()
+#tic()
+#dmpagg <- agg(dmp,cores=CORES)
+#toc()
 
-head(dmpagg$gme_res_df,20) %>%
-  kbl(caption="top gmea genes (t-test)") %>%
-  kable_paper("hover", full_width = F)
+#head(dmpagg$gme_res_df,20) %>%
+#  kbl(caption="top gmea genes (t-test)") %>%
+#  kable_paper("hover", full_width = F)
 
@@
 
-save.image("GSE158422_gmea.Rdata")
+save.image("GSE158422_sensitivity.Rdata")
 

### 2023-10-09 figs/GSE158422_sensitivity.Rmd
sensitivity good

@@
 
+Downsample
+
+```{r,al_down1,fig.height=10}
+
+#medf already exists
+
+SAMPLESIZES=c(2,3,6,10,16,22,30)
+
+ssres <- lapply(SAMPLESIZES,function(n) {
+  SEEDS <- seq(100,5000,100)
+  dsres <- mclapply(SEEDS, function(SEED) {
+    set.seed(SEED)
+    mysample <- sample(1:37,n,replace=FALSE)*2
+    mysample <- c(mysample,(mysample-1))
+    mysample <- mysample[order(mysample)]
+    design2 <- design[mysample,]
+    design2 <- design2[,colSums(design2)>0]
+    medf2 <- medf[,mysample]
+    dmagg2 <- agglimma(medf2,design2)
+    m2 <- as.data.frame(dmagg2$t)
+    rownames(m2) <- rownames(dmagg2)
+    colnames(m2) <- "t"
+    alres2 <- ttenrich(m=m2,genesets=gs_symbols,cores=CORES)
+    head(alres2)
+    alsig2 <- subset(alres2,fdr<0.05)
+    TOT=nrow(alsig)
+    TP=length(which(rownames(alsig2) %in% rownames(alsig)))
+    FP=length(which(!rownames(alsig2) %in% rownames(alsig)))
+    TPES=mean(abs(alsig[which(rownames(alsig) %in% rownames(alsig2)),"t_mean"]))
+    RES=c("TOT"=TOT,"TP"=TP,"FP"=FP,"TPES"=TPES)
+    RES
+  },mc.cores=2)
+  do.call(rbind,dsres)
+})
+
+par(mfrow=c(3,1))
+
+dots <- lapply(ssres,function(x) { x[,"TP"] })
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. true positives",xlab="sample size",cex=0)
+beeswarm(dots,add=TRUE,pch=1,cex=0.6)
+abline(h= ssres[[1]][1,1])
+
+dots <- lapply(ssres,function(x) { x[,"FP"] })
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. false positives",xlab="sample size",cex=0)
+beeswarm(dots,add=TRUE,pch=1,cex=0.6)
+
+dots <- lapply(ssres,function(x) { x[,"TP"] })
+TPdots <- do.call(rbind,dots)
+dots <- lapply(ssres,function(x) { x[,"FP"] })
+FPdots <- do.call(rbind,dots)
+ratio <- t(TPdots/FPdots)
+ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
+
+boxplot(ratio,names=SAMPLESIZES,col="white",ylab="TP per FP",xlab="sample size",cex=0)
+beeswarm(ratio,add=TRUE,pch=1,cex=0.6)
+
+```
+
 # GSAMETH

### 2023-10-11 figs/GSE158422_sensitivity.Rmd
gsa sensitivity

@@
 dots <- lapply(ssres,function(x) { x[,"TP"] })
-boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. true positives",xlab="sample size",cex=0)
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. true positives",xlab="sample size",cex=0,main="GMEA") ; grid()
 beeswarm(dots,add=TRUE,pch=1,cex=0.6)
@@
 dots <- lapply(ssres,function(x) { x[,"FP"] })
-boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. false positives",xlab="sample size",cex=0)
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. false positives",xlab="sample size",cex=0) ; grid()
 beeswarm(dots,add=TRUE,pch=1,cex=0.6)
@@
 FPdots <- do.call(rbind,dots)
-ratio <- t(TPdots/FPdots)
+ratio <- t(FPdots / (TPdots + FPdots))
 ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
-
-boxplot(ratio,names=SAMPLESIZES,col="white",ylab="TP per FP",xlab="sample size",cex=0)
+boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion FP",xlab="sample size",cex=0) ; grid()
 beeswarm(ratio,add=TRUE,pch=1,cex=0.6)
@@
 
-Individual probes have a distribution that leans positive, but after aggregation to genes it is actually
-negative.
+Run GSA meth on the full dataset.
 
@@
 gsaup <- gsaup[order(gsaup$P.DE),]
-head(gsaup,10) %>% kbl(caption="Top hypermethylated pathways") %>% kable_paper("hover", full_width = F)
+gsadn <- gsadn[order(gsadn$P.DE),]
 
-gsadn <- gsadn[order(gsadn$P.DE),]
-head(gsadn,10) %>% kbl(caption="Top hypomethylated pathways") %>% kable_paper("hover", full_width = F)
+gsig_up <-  subset(gsaup,FDR<0.05)
+gsig_up %>% kbl(caption="Hypermethylated pathways") %>% kable_paper("hover", full_width = F)
+gsig_dn <-  subset(gsadn,FDR<0.05)
+gsig_dn %>% kbl(caption="Hypomethylated pathways") %>% kable_paper("hover", full_width = F)
+gsig <- rbind(gsig_up,gsig_dn)
 
-gsig_up <-  rownames(subset(gsaup,FDR<0.05))
-gsig_dn <-  rownames(subset(gsadn,FDR<0.05))
+```
+
+Only 4 hyper and 1 hypomethylated gene set were identified.
+This is surprising as there were 1299 hypermethylated probes corresponding to 808 genes, and
+1016 hypomethylated probes corresponding to 657 genes.
+
+## GSAmeth Downsample
+
+```{r,gsamethdownsample}
+
+SAMPLESIZES=c(2,3,6,10,16,22,30)
+
+gsres <- lapply(SAMPLESIZES,function(n) {
+  SEEDS <- seq(100,5000,100)
+  dsres <- mclapply(SEEDS, function(SEED) {
+    set.seed(SEED)
+    mysample <- sample(1:37,n,replace=FALSE)*2
+    mysample <- c(mysample,(mysample-1))
+    mysample <- mysample[order(mysample)]
+    design2 <- design[mysample,]
+    design2 <- design2[,colSums(design2)>0]
+    #mval downsample
+    mval2 <- mval[,mysample]
+    dm2 <- runlimma(mval2,design2,myann)
+    sigup2 <- rownames(subset(dm2,adj.P.Val<0.05 & logFC>0))
+    sigdn2 <- rownames(subset(dm2,adj.P.Val<0.05 & logFC<0))
+    #gsamet
+    gsaup2 <- gsameth(sig.cpg=sigup2, all.cpg=rownames(dm2), collection=gs_entrez)
+    gsadn2 <- gsameth(sig.cpg=sigdn2, all.cpg=rownames(dm2), collection=gs_entrez)
+    gsig_up2 <- subset(gsaup2,FDR<0.05)
+    gsig_dn2 <- subset(gsadn2,FDR<0.05)
+    gsig2 <- rbind(gsig_up2,gsig_dn2)
+    TOT=nrow(gsig2)
+    TP=length(which(rownames(gsig2) %in% rownames(gsig)))
+    FP=length(which(!rownames(gsig2) %in% rownames(gsig)))
+    TPES=0
+    RES=c("TOT"=TOT,"TP"=TP,"FP"=FP,"TPES"=TPES)
+    RES
+  },mc.cores=8)
+  do.call(rbind,dsres)
+})
+
+mx <- matrix(rep(0,200),ncol=4)
+colnames(mx) <- c("TOT","TP","FP","TPES")
+
+# fix empty matrix
+gsres <- lapply(gsres,function(x) {
+  if ( typeof(x) == "character" ) {
+    mx
+  } else {
+    x
+  }
+} )
+
+par(mfrow=c(3,1))
+
+dots <- lapply(gsres,function(x) { x[,"TP"] })
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. true positives",xlab="sample size",cex=0,main="GSAmeth")
+beeswarm(dots,add=TRUE,pch=1,cex=0.6)
+abline(h= ssres[[1]][1,1])
+
+dots <- lapply(gsres,function(x) { x[,"FP"] })
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. false positives",xlab="sample size",cex=0)
+beeswarm(dots,add=TRUE,pch=1,cex=0.6)
+
+dots <- lapply(gsres,function(x) { x[,"TP"] })
+TPdots <- do.call(rbind,dots)
+dots <- lapply(gsres,function(x) { x[,"FP"] })
+FPdots <- do.call(rbind,dots)
+ratio <- t(FPdots / (TPdots + FPdots))
+ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
+boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion FP",xlab="sample size",cex=0) ; grid()
+beeswarm(ratio,add=TRUE,pch=1,cex=0.6)
 
@@
 
-# GMEA: promoter probes only
-
-## Limma-aggregate-1-sample t-test promoter only
-
-```{r,ls_promoter}
-
-dmp <- dm[grep("Promo",dm$Regulatory_Feature_Group),]
-
-#tic()
-#dmpagg <- agg(dmp,cores=CORES)
-#toc()
-
-#head(dmpagg$gme_res_df,20) %>%
-#  kbl(caption="top gmea genes (t-test)") %>%
-#  kable_paper("hover", full_width = F)
-
-```
-
 ## Session information

### 2023-10-12 figs/GSE158422_sensitivity.Rmd
maybe a competitive test is better?

@@
 
-boxplot(list("tumor"=matrix(colMeans(mval),ncol=2)[,1],"normal"=matrix(colMeans(mval),ncol=2)[,2]),
-  main="mean probe methylation mval")
+mval_means <- list("tumor"=matrix(colMeans(mval),ncol=2)[,1],"normal"=matrix(colMeans(mval),ncol=2)[,2])
+boxplot(mval_means, main="mean probe methylation mval",cex=0,col="white")
+beeswarm(mval_means,add=TRUE,pch=1,cex=1.5)
+lapply(mval_means,mean)
 
 ```
+
+We will assume mean probe methylation is a proxy measure for global methylation.
+The boxplot appears to show higher values in tumour samples, based on median and quartiles.
+However the mean values are actually lower in tumour samples compared to normal.
 
@@
 
+medf_means <- list("tumor"=matrix(colMeans(medf),ncol=2)[,1],"normal"=matrix(colMeans(medf),ncol=2)[,2])
+boxplot(medf_means, main="mean gene methylation mval",cex=0,col="white")
+beeswarm(medf_means,add=TRUE,pch=1,cex=1.5)
+
+lapply(medf_means,mean)
+
 # differential analysis
@@
 
-Downsample
+Try a competitive test.
+
+```{r,compet}
+
+mres <- mitch_calc(x=m,minsetsize=5,genesets=gs_symbols,cores=32,priority="effect")
+
+# hypermethylated
+mres_up <- subset(mres$enrichment_result,p.adjustANOVA<0.05 & s.dist > 0)
+nrow(mres_up)
+head(mres_up,20) %>%
+  kbl(caption="hypermethylated pathways with competitive test") %>%
+  kable_paper("hover", full_width = F)
+
+# hypomethylated
+mres_dn <- subset(mres$enrichment_result,p.adjustANOVA<0.05 & s.dist < 0)
+nrow(mres_dn)
+head(mres_dn,20) %>%
+  kbl(caption="hypomethylated pathways with competitive test") %>%
+  kable_paper("hover", full_width = F)
+
+```
+
+These results look more in line with cancer pathogenesis.
+
+Downsample AL test
 
@@
 
-```{r,gsamethdownsample}
+```{r,gsamethdownsample,fig.height=10}
 
@@
 
+par(mfrow=c(1,1))
+
 ```
@@
 
+msig_up <- subset(mres$enrichment_result,p.adjustANOVA<0.05 & s.dist > 0)$set
+msig_dn <- subset(mres$enrichment_result,p.adjustANOVA<0.05 & s.dist < 0)$set
+
 gsig_up <-  rownames(subset(gsaup,FDR<0.05))
@@
 v1 <- list("AL up"=alsig_up,"AL dn"=alsig_dn,
+  "mitch up"=msig_up,"mitch dn"=msig_dn,
   "GSAmeth up"=gsig_up,"GSAmeth dn"=gsig_dn)

### 2023-10-12 figs/GSE158422_split.Rmd
moving on to sensitivity analysis
add mitch competitive test

@@
 # optimised for 128 GB sever with 32 threads
-CORES=8
+CORES=6
 
@@
 
+## Agg-limma-mitch function
+
+This approach uses the aggregated mvals, limma and instead of a 1-sample t-test it uses mitch
+which is a competitive test and could give more interpretable results.
+
+```{r,alm}
+
+runmitch <- function(m,genesets,cores=1) {
+  mres <- mitch_calc(m,genesets,cores=cores)
+  mres <- mres$enrichment_result
+  rownames(mres) <- mres$set
+  mres$set=NULL
+  return(mres)
+}
+
+```
+
 ## Prep
@@
 
-aares <- mclapply( seq(100,300,100) ,splitanalyze_aa,mc.cores=CORES)
+aares <- mclapply( seq(100,5000,100) ,splitanalyze_aa,mc.cores=CORES)
+
+```
+
+## ALM competitive test
+
+```{r,almflow}
+
+splitanalyze_alm <- function(seed){
+  set.seed(seed)
+  a <- sample(1:37,18)
+  a <- a[order(a)]
+  nona <- setdiff(1:37,a)
+  set.seed(seed)
+  b <- sample(nona,18)
+  b <- b[order(b)]
+  a <- c( (a*2)-1 ,a*2 )
+  a <- a[order(a)]
+  b <- c( (b*2)-1 ,b*2 )
+  b <- b[order(b)]
+
+  design_a <- design[a,]
+  design_a <- design_a[,colSums(design_a)>0]
+  mval_a <- mval[,a]
+  design_b <- design[b,]
+  design_b <- design_b[,colSums(design_b)>0]
+  mval_b <- mval[,b]
+
+  #al
+  medf1 <- chragg(mval_a,myann,cores=floor(CORES/3))
+  magg1 <- agglimma(medf1,design_a)
+  m1 <- as.data.frame(magg1$t)
+  rownames(m1) <- rownames(magg1)
+  colnames(m1) <- "t"
+  almres1 <- runmitch(m=m1,genesets=gs_symbols,cores=CORES)
+
+  #al
+  medf2 <- chragg(mval_b,myann,cores=floor(CORES/3))
+  magg2 <- agglimma(medf2,design_b)
+  m2 <- as.data.frame(magg2$t)
+  rownames(m2) <- rownames(magg2)
+  colnames(m2) <- "t"
+  almres2 <- runmitch(m=m2,genesets=gs_symbols,cores=CORES)
+
+  list("almres1"=almres1,"almres2"=almres2)
+
+}
+
+almres <- mclapply( seq(100,5000,100) ,splitanalyze_alm,mc.cores=CORES)
 
@@
 
+almcompare <-function(almres){
+  almres1 <- almres[[1]]
+  almres2 <- almres[[2]]
+  almsig1 <- subset(almres1,p.adjustANOVA<0.05)
+  almsig2 <- subset(almres2,p.adjustANOVA<0.05)
+
+  almsig1_up <- rownames(subset(almsig1,s.dist>0))
+  almsig1_dn <- rownames(subset(almsig1,s.dist<0))
+  almsig2_up <- rownames(subset(almsig2,s.dist>0))
+  almsig2_dn <- rownames(subset(almsig2,s.dist<0))
+
+  total <- length(unique(c(almsig1_up,almsig2_up,almsig1_dn,almsig2_dn)))
+  common <- length(intersect(almsig1_up, almsig2_up)) + length(intersect(almsig1_dn, almsig2_dn))
+  discordant <- length(intersect(almsig1_up, almsig2_dn)) + length(intersect(almsig1_dn, almsig2_up))
+  uncommon = total - common - discordant
+  result <- c("total"=total,"common"=common,"discordant"=discordant,"uncommon"=uncommon)
+
+  almout <- data.frame(total,common,discordant,uncommon)
+  almout$p_comm <-almout$common / almout$total
+  almout$p_disc <-almout$discordant / almout$total
+  almout$p_uncom <-almout$uncommon / almout$total
+  almout
+}
+
 ```
@@
 
+almout <- do.call(rbind,lapply(almres,alcompare))
+
 laout
@@
 
-res <- do.call(rbind,lapply(list(laout,alout,aaout),colMeans))
+almout
 
-rownames(res) <- c("la","al","aa")
+res <- do.call(rbind,lapply(list(laout,alout,aaout,almout),colMeans))
+
+rownames(res) <- c("la","al","aa","alm")
 
@@
 apply(aaout,2,mean)
+apply(almout,2,mean)
 
@@
 
-barplot(res$common,main="no. common",names.arg=c("LA","AL","AA"))
-barplot(res$uncommon,main="no. uncommon",names.arg=c("LA","AL","AA"))
-barplot(res$discordant,main="no. discordant",names.arg=c("LA","AL","AA"))
+barplot(res$common,main="no. common",names.arg=c("LA","AL","AA","ALM"))
+barplot(res$uncommon,main="no. uncommon",names.arg=c("LA","AL","AA","ALM"))
+barplot(res$discordant,main="no. discordant",names.arg=c("LA","AL","AA","ALM"))
 
-boxplot(list("LA"=laout$common,"AL"=alout$common,"AA"=aaout$common),
+boxplot(list("LA"=laout$common,"AL"=alout$common,"AA"=aaout$common,"ALM"=almout$common),
   main="no. common",col="white",ylim=c(0,1550))
-beeswarm(list("LA"=laout$common,"AL"=alout$common,"AA"=aaout$common), add=TRUE,cex=1.2,pch=19)
+beeswarm(list("LA"=laout$common,"AL"=alout$common,"AA"=aaout$common,"ALM"=almout$common), add=TRUE,cex=1.2,pch=19)
 
-boxplot(list("LA"=laout$uncommon,"AL"=alout$uncommon,"AA"=aaout$uncommon),
+boxplot(list("LA"=laout$uncommon,"AL"=alout$uncommon,"AA"=aaout$uncommon,"ALM"=almout$uncommon),
   main="no. uncommon",col="white",ylim=c(0,1550))
-beeswarm(list("LA"=laout$uncommon,"AL"=alout$uncommon,"AA"=aaout$uncommon), add=TRUE,cex=1.2,pch=19)
+beeswarm(list("LA"=laout$uncommon,"AL"=alout$uncommon,"AA"=aaout$uncommon,"ALM"=almout$uncommon), add=TRUE,cex=1.2,pch=19)
 
-boxplot(list("LA"=laout$discordant,"AL"=alout$discordant,"AA"=aaout$discordant),
+boxplot(list("LA"=laout$discordant,"AL"=alout$discordant,"AA"=aaout$discordant,"ALM"=almout$discordant),
   main="no. discordant",col="white",ylim=c(0,1550))
-beeswarm(list("LA"=laout$discordant,"AL"=alout$discordant,"AA"=aaout$discordant), add=TRUE,cex=1.2,pch=19)
+beeswarm(list("LA"=laout$discordant,"AL"=alout$discordant,"AA"=aaout$discordant,"ALM"=almout$discordant), add=TRUE,cex=1.2,pch=19)
 
-barplot(res$p_comm,main="proportion common",names.arg=c("LA","AL","AA"))
-barplot(res$p_uncom,main="proportion uncommon",names.arg=c("LA","AL","AA"))
-barplot(res$p_disc,main="proportion discordant",names.arg=c("LA","AL","AA"))
+barplot(res$p_comm,main="proportion common",names.arg=c("LA","AL","AA","ALM"))
+barplot(res$p_uncom,main="proportion uncommon",names.arg=c("LA","AL","AA","ALM"))
+barplot(res$p_disc,main="proportion discordant",names.arg=c("LA","AL","AA","ALM"))
 
-boxplot(list("LA"=laout$p_comm,"AL"=alout$p_comm,"AA"=aaout$p_comm),
+boxplot(list("LA"=laout$p_comm,"AL"=alout$p_comm,"AA"=aaout$p_comm,"ALM"=almout$p_comm),
   main="proportion common",col="white",ylim=c(0,1))
-beeswarm(list("LA"=laout$p_comm,"AL"=alout$p_comm,"AA"=aaout$p_comm), add=TRUE,cex=1.2,pch=19)
+beeswarm(list("LA"=laout$p_comm,"AL"=alout$p_comm,"AA"=aaout$p_comm,"ALM"=almout$p_comm), add=TRUE,cex=1.2,pch=19)
 
-boxplot(list("LA"=laout$p_uncom,"AL"=alout$p_uncom,"AA"=aaout$p_uncom),
+boxplot(list("LA"=laout$p_uncom,"AL"=alout$p_uncom,"AA"=aaout$p_uncom,"ALM"=almout$p_uncom),
   main="proportion uncommon",col="white",ylim=c(0,1))
-beeswarm(list("LA"=laout$p_uncom,"AL"=alout$p_uncom,"AA"=aaout$p_uncom), add=TRUE,cex=1.2,pch=19)
+beeswarm(list("LA"=laout$p_uncom,"AL"=alout$p_uncom,"AA"=aaout$p_uncom,"ALM"=almout$p_uncom), add=TRUE,cex=1.2,pch=19)
 
-boxplot(list("LA"=laout$p_disc,"AL"=alout$p_disc,"AA"=aaout$p_disc),
+boxplot(list("LA"=laout$p_disc,"AL"=alout$p_disc,"AA"=aaout$p_disc,"ALM"=almout$p_disc),
   main="proportion discordant",col="white",ylim=c(0,1))
-beeswarm(list("LA"=laout$p_disc,"AL"=alout$p_disc,"AA"=aaout$p_disc), add=TRUE,cex=1.2,pch=19)
+beeswarm(list("LA"=laout$p_disc,"AL"=alout$p_disc,"AA"=aaout$p_disc,"ALM"=almout$p_disc), add=TRUE,cex=1.2,pch=19)
 

### 2023-10-12 figs/GSE158422_sensitivity.Rmd
script updates with competitive test

@@
 
+# runmitch is a helper function that runs mitch on a 1 column dataframe and just returns
+# the differential table
+runmitch <- function(m,genesets,cores=1) {
+  mres <- mitch_calc(m,genesets,minsetsize=5,cores=cores)
+  mres <- mres$enrichment_result
+  rownames(mres) <- mres$set
+  mres$set=NULL
+  return(mres)
+}
+
 ```
@@
 
-Try a competitive test.
-
-```{r,compet}
-
-mres <- mitch_calc(x=m,minsetsize=5,genesets=gs_symbols,cores=32,priority="effect")
-
-# hypermethylated
-mres_up <- subset(mres$enrichment_result,p.adjustANOVA<0.05 & s.dist > 0)
-nrow(mres_up)
-head(mres_up,20) %>%
-  kbl(caption="hypermethylated pathways with competitive test") %>%
-  kable_paper("hover", full_width = F)
-
-# hypomethylated
-mres_dn <- subset(mres$enrichment_result,p.adjustANOVA<0.05 & s.dist < 0)
-nrow(mres_dn)
-head(mres_dn,20) %>%
-  kbl(caption="hypomethylated pathways with competitive test") %>%
-  kable_paper("hover", full_width = F)
-
-```
-
-These results look more in line with cancer pathogenesis.
-
-Downsample AL test
+## Downsample AL test
 
@@
 dots <- lapply(ssres,function(x) { x[,"FP"] })
+FPdots <- do.call(rbind,dots)
+ratio <- t(FPdots / (TPdots + FPdots))
+ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
+boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion FP",xlab="sample size",cex=0) ; grid()
+beeswarm(ratio,add=TRUE,pch=1,cex=0.6)
+
+```
+
+## ALM aggregate limma mitch
+
+Try a competitive test.
+
+```{r,compet}
+
+mres <- mitch_calc(x=m,minsetsize=5,genesets=gs_symbols,cores=32,priority="effect")
+almsig <- subset(mres$enrichment_result,p.adjustANOVA<0.05)
+rownames(almsig) <- almsig$set
+alm$set=NULL
+
+# hypermethylated
+mres_up <- subset(mres$enrichment_result,p.adjustANOVA<0.05 & s.dist > 0)
+nrow(mres_up)
+head(mres_up,20) %>%
+  kbl(caption="hypermethylated pathways with competitive test") %>%
+  kable_paper("hover", full_width = F)
+
+# hypomethylated
+mres_dn <- subset(mres$enrichment_result,p.adjustANOVA<0.05 & s.dist < 0)
+nrow(mres_dn)
+head(mres_dn,20) %>%
+  kbl(caption="hypomethylated pathways with competitive test") %>%
+  kable_paper("hover", full_width = F)
+
+```
+
+These results look more in line with cancer pathogenesis.
+
+Downsample alm method
+
+```{r,almdown}
+
+SAMPLESIZES=c(2,3,6,10,16,22,30)
+
+mmres <- lapply(SAMPLESIZES,function(n) {
+  SEEDS <- seq(100,5000,100)
+  dsres <- mclapply(SEEDS, function(SEED) {
+    set.seed(SEED)
+    mysample <- sample(1:37,n,replace=FALSE)*2
+    mysample <- c(mysample,(mysample-1))
+    mysample <- mysample[order(mysample)]
+    design2 <- design[mysample,]
+    design2 <- design2[,colSums(design2)>0]
+    medf2 <- medf[,mysample]
+    dmagg2 <- agglimma(medf2,design2)
+    m2 <- as.data.frame(dmagg2$t)
+    rownames(m2) <- rownames(dmagg2)
+    colnames(m2) <- "t"
+    almres2 <- runmitch(m=m2,genesets=gs_symbols,cores=CORES)
+    head(almres2)
+    almsig2 <- subset(almres2,p.adjustANOVA<0.05)
+    TOT=nrow(almsig)
+    TP=length(which(rownames(almsig2) %in% rownames(almsig)))
+    FP=length(which(!rownames(almsig2) %in% rownames(almsig)))
+    TPES=1
+    RES=c("TOT"=TOT,"TP"=TP,"FP"=FP,"TPES"=TPES)
+    RES
+  },mc.cores=2)
+  do.call(rbind,dsres)
+})
+
+par(mfrow=c(3,1))
+
+dots <- lapply(mmres,function(x) { x[,"TP"] })
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. true positives",xlab="sample size",cex=0,main="ALM") ; grid()
+beeswarm(dots,add=TRUE,pch=1,cex=0.6)
+abline(h=mmres[[1]][1,1])
+
+dots <- lapply(mmres,function(x) { x[,"FP"] })
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. false positives",xlab="sample size",cex=0) ; grid()
+beeswarm(dots,add=TRUE,pch=1,cex=0.6)
+
+dots <- lapply(mmres,function(x) { x[,"TP"] })
+TPdots <- do.call(rbind,dots)
+dots <- lapply(mmres,function(x) { x[,"FP"] })
 FPdots <- do.call(rbind,dots)
@@
 beeswarm(dots,add=TRUE,pch=1,cex=0.6)
-abline(h= ssres[[1]][1,1])
+abline(h=gsres[[1]][1,1])
 

### 2023-10-12 figs/GSE158422_split.Rmd
script updates with competitive test

@@
 
-genesets <- gmt_import("https://ziemann-lab.net/public/gmea_prototype/ReactomePathways.gmt")
+#genesets <- gmt_import("https://ziemann-lab.net/public/gmea_prototype/ReactomePathways.gmt")
 
@@
   medf <- chragg(mval,myann,cores=cores)
-  agag <- gsagg(x=medf,genesets=gs_symbols,cores=cores)
+  agag <- gsagg(x=medf,genesets=genesets,cores=cores)
   aalres <- aalimma(agag=agag,design=design)
@@
 runmitch <- function(m,genesets,cores=1) {
-  mres <- mitch_calc(m,genesets,cores=cores)
+  mres <- mitch_calc(m,genesets,minsetsize=5,cores=cores)
   mres <- mres$enrichment_result
@@
 
-  # la1
   dm1 <- runlimma(mval_a,design_a,myann)
@@
 
-  # la2
   dm2 <- runlimma(mval_b,design_b,myann)
@@
 
-  #al
   medf1 <- chragg(mval_a,myann,cores=floor(CORES/3))
@@
 
-  #al
   medf2 <- chragg(mval_b,myann,cores=floor(CORES/3))
@@
 
-  #aa1
   aares1 <- aal(mval=mval_a,myann=myann,genesets=gs_symbols,
@@
 
-  #aa2
   aares2 <- aal(mval=mval_b,myann=myann,genesets=gs_symbols,
@@
 
-almout <- do.call(rbind,lapply(almres,alcompare))
+almout <- do.call(rbind,lapply(almres,almcompare))
 

### 2023-10-14 figs/GSE158422_simulate.Rmd
started work on simulation

--- 2023-10-12 figs/GSE158422_split.Rmd
@@
 ---
-title: "GSE158422: DNA methylation in Lung squamous cell carcinoma - gene methylation enrichment analysis"
+title: "Testing gene methylation enrichment analysis approaches using simulated data"
 author: "The GMEA team"
@@
 
-Here we are splitting the dataset and testing concordance between LA and AL approaches.
+In this report we establish a new method for generating simulated data with known ground truth.
+This will be used to test different gene methylation enrichment approaches systematically.
+
+The general steps are:
+
+1. Import GSE158422 data corresponding to control (non-tumour tissue).
+
+2. From the 37 samples, create two groups of 18 samples.
+One of these will be considered "control" and the other "case".
+
+3. Some gene sets will be selected to be differentially methylated.
+A requirement is that there will be no overlap between them.
+Half of these will be hypermethylated and the others will be hypomethylated.
+
+4. The changes will be incorporated into the "case" samples.
+
+5. The enrichment analysis will be conducted.
+
+6. The accuracy will be calculated.
 
@@
 
+# Reactome
+
 Reactome pathways were downloaded on the 14th Sept 2023 from MsigDB.
@@
 gs_symbols <- gmt_import("c2.cp.reactome.v2023.1.Hs.symbols.gmt")
+
+```
+
+Select nonoverlapping gene sets.
+
+```{r,gset_selection}
+
+gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
+
+gset_selection <- function(genesetdatabase,seed,gene_catalog) {
+  set.seed(seed)
+  gsets <- gs_symbols[sample(1:length(genesetdatabase))]
+  genelist=NULL
+  genesets=NULL
+  for ( i in 1:length(gsets) ) {
+    gs <- gsets[i]
+    inx <- length(intersect(unlist(gs),genelist))
+    num <- length(intersect(unlist(gs),gene_catalog))
+    if ( inx == 0 & num >= 10 ) {
+      genesets <- c(genesets,gs)
+      genelist <- c(unname(unlist(gs)),genelist)
+    }
+  }
+  return(genesets)
+}
+
+gset_mod <- gset_selection(genesetdatabase=gs_symbols,seed=100,gene_catalog=gene_catalog)
+
+```
+
+# Incorporate changes
+
+TODO: to incorporate changes to case samples.
+
+Need to figure out what magnitude to change.
+Will refer to the cancer/normal comparison.
+
+Select genes and probes to alter.
+
+```{r,incorp}
+
+seed=100
+
+# select gene sets to alter
+gset_mod <- gset_selection(genesetdatabase=gs_symbols,seed=seed,gene_catalog=gene_catalog)
+
+# select normal samples only
+normal <- mval[,(1:(ncol(mval)/2)*2)]
+
+# divide sets between hyper and hypomethylated
+nset <- floor(length(gset_mod)/2)
+set.seed(seed) ; gtup <-sample(gset_mod,48)
+set.seed(seed) ; gtdn <- sample(setdiff(gset_mod,gtup),nset)
+gup <- unname(unlist(gtup))
+gdn <- unname(unlist(gtdn))
+
+# divide samples between ctrl and case
+set.seed(seed) ; ctrl <- sample(1:ncol(normal),18)
+set.seed(seed) ; case <- sample(setdiff(1:37,ctrl),18)
+mval_ctrl <- mval[,ctrl]
+mval_case <- mval[,case]
+
+# select probes
 

### 2023-10-15 figs/GSE158422_simulate.Rmd
simwork

@@
 
+frac_genes=0.5
+frac_probes=0.5
+
 # select gene sets to alter
@@
 
-# divide sets between hyper and hypomethylated
+# divide gene sets between hyper and hypomethylated
 nset <- floor(length(gset_mod)/2)
@@
 gdn <- unname(unlist(gtdn))
+
+# make probe-gene vector
+probe2gene <- strsplit(myann$UCSC_RefGene_Name,";")
+names(probe2gene) <- rownames(myann)
+probe2gene <- unlist(probe2gene)
+
+# select probes hypermethylated
+set.seed(seed) ; gup2 <- sample(gup,floor(length(gup)/2))
+pup <- names(probe2gene[which(probe2gene %in% gup2)])
+set.seed(seed) ; pup2 <- sample(pup,floor(length(pup)/2))
+
+# select probes hypomethylated
+set.seed(seed) ; gdn2 <- sample(gdn,floor(length(gdn)/2))
+pdn <- names(probe2gene[which(probe2gene %in% gdn2)])
+set.seed(seed) ; pdn2 <- sample(pdn,floor(length(pup)/2))
 
@@
 
-# select probes
+# incorporate altered signals - change by +1 or -1
+mval_case[rownames(mval_case) %in% pup2,] <-  mval_case[rownames(mval_case) %in% pup2,] +1
+mval_case[rownames(mval_case) %in% pdn2,] <-  mval_case[rownames(mval_case) %in% pdn2,] -1
+mval2 <- cbind(mval_ctrl,mval_case)
+
+ss <- data.frame(colnames(mval2))
+colnames(ss) <- "sample"
+ss$case <- c(rep(0,18),rep(1,18))
+
+d <- model.matrix(~ ss$case)
+fit.reduced <- lmFit(mval2,d)
+fit.reduced <- eBayes(fit.reduced)
+
+summary(decideTests(fit.reduced))
+dm <- topTable(fit.reduced,coef=ncol(d), number = Inf)
 

### 2023-10-15 figs/GSE158422_simulate.Rmd
sim

--- 2023-10-14 figs/GSE158422_simulate.Rmd
@@
 seed=100
+frac_genes=0.5
+frac_probes=0.5
+normal_mval <- mval[,(1:(ncol(mval)/2)*2)]
 
+incorp_dm <- function(genesetdatabase,myann,gene_catalog,mval,seed,frac_genes,frac_probes) {
   # select gene sets to alter
-gset_mod <- gset_selection(genesetdatabase=gs_symbols,seed=seed,gene_catalog=gene_catalog)
-
-# select normal samples only
-normal <- mval[,(1:(ncol(mval)/2)*2)]
-
-# divide sets between hyper and hypomethylated
+  gset_mod <- gset_selection(genesetdatabase=genesetdatabase,seed=seed,gene_catalog=gene_catalog)
+  # divide gene sets between hyper and hypomethylated
   nset <- floor(length(gset_mod)/2)
-set.seed(seed) ; gtup <-sample(gset_mod,48)
+  set.seed(seed) ; gtup <-sample(gset_mod,nset)
   set.seed(seed) ; gtdn <- sample(setdiff(gset_mod,gtup),nset)
@@
   gdn <- unname(unlist(gtdn))
-
+  # make probe-gene vector
+  probe2gene <- strsplit(myann$UCSC_RefGene_Name,";")
+  names(probe2gene) <- rownames(myann)
+  probe2gene <- unlist(probe2gene)
+  # select probes hypermethylated
+  set.seed(seed) ; gup2 <- sample(gup,floor(length(gup)*frac_genes))
+  pup <- names(probe2gene[which(probe2gene %in% gup2)])
+  set.seed(seed) ; pup2 <- sample(pup,floor(length(pup)*frac_probes))
+  # select probes hypomethylated
+  set.seed(seed) ; gdn2 <- sample(gdn,floor(length(gdn)*frac_genes))
+  pdn <- names(probe2gene[which(probe2gene %in% gdn2)])
+  set.seed(seed) ; pdn2 <- sample(pdn,floor(length(pup)*frac_probes))
   # divide samples between ctrl and case
-set.seed(seed) ; ctrl <- sample(1:ncol(normal),18)
-set.seed(seed) ; case <- sample(setdiff(1:37,ctrl),18)
+  groupsize=floor(ncol(mval)/2)
+  ncols <- ncol(mval)
+  set.seed(seed) ; ctrl <- sample(1:ncols,groupsize)
+  set.seed(seed) ; case <- sample(setdiff(1:ncols,ctrl),groupsize)
   mval_ctrl <- mval[,ctrl]
   mval_case <- mval[,case]
+  # incorporate altered signals - change by +1 or -1
+  mval_case[rownames(mval_case) %in% pup2,] <-  mval_case[rownames(mval_case) %in% pup2,] +1
+  mval_case[rownames(mval_case) %in% pdn2,] <-  mval_case[rownames(mval_case) %in% pdn2,] -1
+  mval2 <- cbind(mval_ctrl,mval_case)
+  result <- list("mval"=mval2,"probes up"=pup2,"probes down"=pdn2,
+    "genes up"=gup2,"genes down"=gdn2,
+    "genesets up"=gtup,"genesets down"=gtdn)
+  return(result)
+}
 
-# select probes
+```
+
+This is how to use the function
+
+```{r,run1sim}
+
+sim <- incorp_dm(genesetdatabase=gs_symbols,myann=myann,gene_catalog=gene_catalog,
+  mval=normal_mval,seed=100,frac_genes=0.5,frac_probes=0.5)
+
+mval2 <- sim$mval
+ncols <- ncol(mval2)
+groupsize <- ncols/2
+ss <- data.frame(colnames(mval2))
+colnames(ss) <- "sample"
+ss$case <- c(rep(0,groupsize),rep(1,groupsize))
+
+d <- model.matrix(~ ss$case)
+fit.reduced <- lmFit(mval2,d)
+fit.reduced <- eBayes(fit.reduced)
+
+summary(decideTests(fit.reduced))
+dm <- topTable(fit.reduced,coef=ncol(d), number = Inf)
+head(dm)
 

### 2023-10-15 figs/GSE158422_simulate.Rmd
gsa sim

--- 2023-10-14 figs/GSE158422_simulate.Rmd
@@
 seed=100
+frac_genes=0.5
+frac_probes=0.5
+nsamples=10
+normal_mval <- mval[,(1:(ncol(mval)/2)*2)]
 
+incorp_dm <- function(genesetdatabase,myann,gene_catalog,mval,seed,frac_genes,frac_probes,groupsize) {
   # select gene sets to alter
-gset_mod <- gset_selection(genesetdatabase=gs_symbols,seed=seed,gene_catalog=gene_catalog)
-
-# select normal samples only
-normal <- mval[,(1:(ncol(mval)/2)*2)]
-
-# divide sets between hyper and hypomethylated
+  gset_mod <- gset_selection(genesetdatabase=genesetdatabase,seed=seed,gene_catalog=gene_catalog)
+  # divide gene sets between hyper and hypomethylated
   nset <- floor(length(gset_mod)/2)
-set.seed(seed) ; gtup <-sample(gset_mod,48)
+  set.seed(seed) ; gtup <-sample(gset_mod,nset)
   set.seed(seed) ; gtdn <- sample(setdiff(gset_mod,gtup),nset)
@@
   gdn <- unname(unlist(gtdn))
-
+  # make probe-gene vector
+  probe2gene <- strsplit(myann$UCSC_RefGene_Name,";")
+  names(probe2gene) <- rownames(myann)
+  probe2gene <- unlist(probe2gene)
+  # select probes hypermethylated
+  set.seed(seed) ; gup2 <- sample(gup,floor(length(gup)*frac_genes))
+  pup <- names(probe2gene[which(probe2gene %in% gup2)])
+  set.seed(seed) ; pup2 <- sample(pup,floor(length(pup)*frac_probes))
+  # select probes hypomethylated
+  set.seed(seed) ; gdn2 <- sample(gdn,floor(length(gdn)*frac_genes))
+  pdn <- names(probe2gene[which(probe2gene %in% gdn2)])
+  set.seed(seed) ; pdn2 <- sample(pdn,floor(length(pup)*frac_probes))
   # divide samples between ctrl and case
-set.seed(seed) ; ctrl <- sample(1:ncol(normal),18)
-set.seed(seed) ; case <- sample(setdiff(1:37,ctrl),18)
+  ncols <- ncol(mval)
+  maxgroupsize=floor(ncols/2)
+  if ( groupsize > maxgroupsize ) { stop("groupsize cannot be larger than half the ncols of mval") }
+  set.seed(seed) ; ctrl <- sample(1:ncols,groupsize)
+  set.seed(seed) ; case <- sample(setdiff(1:ncols,ctrl),groupsize)
   mval_ctrl <- mval[,ctrl]
   mval_case <- mval[,case]
+  # incorporate altered signals - change by +1 or -1
+  mval_case[rownames(mval_case) %in% pup2,] <-  mval_case[rownames(mval_case) %in% pup2,] +1
+  mval_case[rownames(mval_case) %in% pdn2,] <-  mval_case[rownames(mval_case) %in% pdn2,] -1
+  mval2 <- cbind(mval_ctrl,mval_case)
+  result <- list("mval"=mval2,"probes up"=pup2,"probes down"=pdn2,
+    "genes up"=gup2,"genes down"=gdn2,
+    "genesets up"=gtup,"genesets down"=gtdn)
+  return(result)
+}
 
-# select probes
+```
+
+## GSAMETH function
+
+```{r,gsameth_func}
+
+# limma
+runlimma <- function(mval,design,myann) {
+  fit.reduced <- lmFit(mval,design)
+  fit.reduced <- eBayes(fit.reduced)
+  dm <- topTable(fit.reduced,coef=ncol(design), number = Inf)
+  dm <- merge(myann,dm,by=0)
+  dm <- dm[order(dm$P.Value),]
+  rownames(dm) <- dm$Row.names
+  dm$Row.names=NULL
+  return(dm)
+}
+
+```
+
+This is how to use the function
+
+```{r,rungsameth}
+
+simgsa <- function(genesetdatabase,myann,gene_catalog,
+  mval,seed,frac_genes,frac_probes,groupsize) {
+
+  sim <- incorp_dm(genesetdatabase=gs_symbols,myann=myann,gene_catalog=gene_catalog,
+    mval=mval,seed=seed,frac_genes=0.5,frac_probes=0.5,groupsize=groupsize)
+  mval2 <- sim$mval
+  ncols <- ncol(mval2)
+  groupsize <- ncols/2
+  ss <- data.frame(colnames(mval2))
+  colnames(ss) <- "sample"
+  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
+  d <- model.matrix(~ ss$case )
+  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
+  pup3 <- rownames(subset(dm3,adj.P.Val<0.05 & logFC>0))
+  pdn3 <- rownames(subset(dm3,adj.P.Val<0.05 & logFC<0))
+  gsaup3 <- gsameth(sig.cpg=pup3, all.cpg=rownames(dm3), collection=gs_entrez)
+  gsadn3 <- gsameth(sig.cpg=pdn3, all.cpg=rownames(dm3), collection=gs_entrez)
+  gsig_up3 <- rownames(subset(gsaup3,FDR<0.05))
+  gsig_dn3 <- rownames(subset(gsadn3,FDR<0.05))
+  gtup <- names(sim[[6]])
+  gtdn <- names(sim[[7]])
+  UPTP=length(intersect(gsig_up3 ,gtup))
+  UPFP=length(setdiff(gsig_up3 ,gtup))
+  UPFN=length(setdiff(gtup,gsig_up3))
+  DNTP=length(intersect(gsig_dn3 ,gtdn))
+  DNFP=length(setdiff(gsig_dn3 ,gtdn))
+  DNFN=length(setdiff(gtdn,gsig_dn3))
+  TP=UPTP+DNTP
+  FP=UPFP+DNFP
+  FN=UPFN+DNFN
+  TN=nrow(gsadn3)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
+  PREC=TP/(TP+FP)
+  REC=TP/(TP+FN)
+  F1=TP/(TP+(0.5*(FP+FN)))
+  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
+  return(result)
+}
+
+groupsize=10
+seed=100
+
+gres <- mclapply(1:50,function(i) {
+  simgsa(genesetdatabase=gs_symbols,myann=myann,gene_catalog=gene_catalog,
+    mval=normal_mval,seed=i*100,frac_genes=0.5,frac_probes=0.5,groupsize=groupsize)
+},mc.cores=10)
 

### 2023-10-16 figs/GSE158422_simulate.Rmd
plumbing

--- 2023-10-15 figs/GSE158422_simulate.Rmd
@@
 suppressPackageStartupMessages({
+  library("stringi")
   library("limma")
-  library("eulerr")
+  library("missMethyl")
   library("IlluminaHumanMethylation450kmanifest")
   library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
+  library('org.Hs.eg.db')
   library("tictoc")
@@
   library("beeswarm")
+  library("eulerr")
 })
@@
 
-# Reactome
+# Gene set database
 
-Reactome pathways were downloaded on the 14th Sept 2023 from MsigDB.
+We could use Reactome pathways, however these have a lot of overlapping sets, which could cause inflated
+false positives.
+A better solution could be to select random gene sets with sizes similar to Reactomes.
+Reactomes were downloaded on the 14th Sept 2023 from MsigDB.
 
@@
 
+Get a vector of genes, and the length of gene sets from Reactome.
+
+Then generate random gene sets.
+
+```{r,gset_prep}
+
+gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
+
+lengths <- unname(unlist(lapply(gs_symbols,length)))
+
+randomGeneSets <- function(gene_catalog, lengths, seed){
+  num_gsets <- length(lengths)
+  set.seed(seed) ; seeds <- sample(1:1e6, num_gsets)
+  gsets <- lapply(1:num_gsets,function(i) {
+    set.seed(seeds[i]) ; gs <- sample(gene_catalog,lengths[i])
+    return(gs)
+  } )
+  names(gsets)<-stri_rand_strings(length(gsets), 15, pattern = "[A-Za-z]")
+  return(gsets)
+}
+
+gsets <- randomGeneSets(gene_catalog,lengths,seed=100)
+
+```
+
 Select nonoverlapping gene sets.
@@
 
-gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
-
 gset_selection <- function(genesetdatabase,seed,gene_catalog) {
   set.seed(seed)
-  gsets <- gs_symbols[sample(1:length(genesetdatabase))]
+  gsets <- genesetdatabase[sample(1:length(genesetdatabase))]
   genelist=NULL
@@
 
-gset_mod <- gset_selection(genesetdatabase=gs_symbols,seed=100,gene_catalog=gene_catalog)
+gset_mod <- gset_selection(genesetdatabase=gsets,seed=100,gene_catalog=gene_catalog)
 
@@
 seed=100
-
 frac_genes=0.5
 frac_probes=0.5
+delta=1
+nsamples=10
+normal_mval <- mval[,(1:(ncol(mval)/2)*2)]
 
-# select gene sets to alter
-gset_mod <- gset_selection(genesetdatabase=gs_symbols,seed=seed,gene_catalog=gene_catalog)
-
-# select normal samples only
-normal <- mval[,(1:(ncol(mval)/2)*2)]
+incorp_dm <- function(genesets,myann,mval,seed,
+  frac_genes,frac_probes,groupsize,delta=1) {
 
   # divide gene sets between hyper and hypomethylated
-nset <- floor(length(gset_mod)/2)
-set.seed(seed) ; gtup <-sample(gset_mod,48)
-set.seed(seed) ; gtdn <- sample(setdiff(gset_mod,gtup),nset)
+  nset <- floor(length(genesets)/2)
+  set.seed(seed) ; gtup <-sample(genesets,nset)
+  set.seed(seed) ; gtdn <- sample(setdiff(genesets,gtup),nset)
   gup <- unname(unlist(gtup))
   gdn <- unname(unlist(gtdn))
-
   # make probe-gene vector
@@
   probe2gene <- unlist(probe2gene)
-
   # select probes hypermethylated
-set.seed(seed) ; gup2 <- sample(gup,floor(length(gup)/2))
+  set.seed(seed) ; gup2 <- sample(gup,floor(length(gup)*frac_genes))
   pup <- names(probe2gene[which(probe2gene %in% gup2)])
-set.seed(seed) ; pup2 <- sample(pup,floor(length(pup)/2))
-
+  set.seed(seed) ; pup2 <- sample(pup,floor(length(pup)*frac_probes))
   # select probes hypomethylated
-set.seed(seed) ; gdn2 <- sample(gdn,floor(length(gdn)/2))
+  set.seed(seed) ; gdn2 <- sample(gdn,floor(length(gdn)*frac_genes))
   pdn <- names(probe2gene[which(probe2gene %in% gdn2)])
-set.seed(seed) ; pdn2 <- sample(pdn,floor(length(pup)/2))
-
+  set.seed(seed) ; pdn2 <- sample(pdn,floor(length(pup)*frac_probes))
   # divide samples between ctrl and case
-set.seed(seed) ; ctrl <- sample(1:ncol(normal),18)
-set.seed(seed) ; case <- sample(setdiff(1:37,ctrl),18)
+  ncols <- ncol(mval)
+  maxgroupsize=floor(ncols/2)
+  if ( groupsize > maxgroupsize ) { stop("groupsize cannot be larger than half the ncols of mval") }
+  set.seed(seed) ; ctrl <- sample(1:ncols,groupsize)
+  set.seed(seed) ; case <- sample(setdiff(1:ncols,ctrl),groupsize)
   mval_ctrl <- mval[,ctrl]
   mval_case <- mval[,case]
-
   # incorporate altered signals - change by +1 or -1
-mval_case[rownames(mval_case) %in% pup2,] <-  mval_case[rownames(mval_case) %in% pup2,] +1
-mval_case[rownames(mval_case) %in% pdn2,] <-  mval_case[rownames(mval_case) %in% pdn2,] -1
+  mval_case[rownames(mval_case) %in% pup2,] <-  mval_case[rownames(mval_case) %in% pup2,] + delta
+  mval_case[rownames(mval_case) %in% pdn2,] <-  mval_case[rownames(mval_case) %in% pdn2,] - delta
   mval2 <- cbind(mval_ctrl,mval_case)
-
-ss <- data.frame(colnames(mval2))
-colnames(ss) <- "sample"
-ss$case <- c(rep(0,18),rep(1,18))
-
-d <- model.matrix(~ ss$case)
-fit.reduced <- lmFit(mval2,d)
-fit.reduced <- eBayes(fit.reduced)
-
-summary(decideTests(fit.reduced))
-dm <- topTable(fit.reduced,coef=ncol(d), number = Inf)
+  result <- list("mval"=mval2,"probes up"=pup2,"probes down"=pdn2,
+    "genes up"=gup2,"genes down"=gdn2,
+    "genesets up"=gtup,"genesets down"=gtdn)
+  return(result)
+}
 
@@
 
-# GMEA FUNCTIONS
+## GSAMETH function
 
-There are three approaches.
-First, limma can be done first, followed by aggregation by gene and then enrichment
-test (called LA).
-Secondly, probe values can be aggregated by gene, followed by limma, followed by
-enrichment test (AL).
-Third, probe values can be aggregated to gene level, and then aggregated to gene set
-level followed by limma test (AA).
-
-## LA
-
-This process runs limma first and then aggregates the results before doing an enrichment test.
-
-```{r,lafunc}
+```{r,gsameth_func}
 
@@
   fit.reduced <- eBayes(fit.reduced)
-  summary(decideTests(fit.reduced))
-  dm <- topTable(fit.reduced,coef=4, number = Inf)
+  dm <- topTable(fit.reduced,coef=ncol(design), number = Inf)
   dm <- merge(myann,dm,by=0)
@@
 }
+
+```
+
+This is how to use the function
+
+This could be complicated as it requires translation of symbols to entrez IDs, but
+could be simplified if the entrez translation is done at the gsameth step.
+
+```{r,rungsameth}
+
+simgsa <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
+  # generate gene sets
+  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
+  lengths <- unname(unlist(lapply(genesetdatabase,length)))
+  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
+  # select gene sets to alter
+  gset_mod <- gset_selection(genesetdatabase=gsets,seed=seed,gene_catalog=gene_catalog)
+  # cut down length to num_dm_sets
+  set.seed(seed) ; gset_mod <- sample(gset_mod,num_dm_sets)
+  # incorporate select changes
+  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
+    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
+  # set up limma
+  mval2 <- sim$mval
+  ncols <- ncol(mval2)
+  groupsize <- ncols/2
+  ss <- data.frame(colnames(mval2))
+  colnames(ss) <- "sample"
+  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
+  d <- model.matrix(~ ss$case )
+  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
+  pup3 <- rownames(subset(dm3,adj.P.Val<0.05 & logFC>0))
+  pdn3 <- rownames(subset(dm3,adj.P.Val<0.05 & logFC<0))
+  # convert gene sets to entrez
+  gene2entrez <- mapIds(org.Hs.eg.db, gene_catalog, 'ENTREZID', 'SYMBOL')
+  gsets_entrez <- lapply(gsets,function(gs) {
+    gs2 <- unique(gene2entrez[names(gene2entrez) %in% gs])
+    gs2 <- gs2[!is.na(gs2)]
+    return(gs2)
+  })
+  gsaup3 <- gsameth(sig.cpg=pup3, all.cpg=rownames(dm3), collection=gsets_entrez)
+  gsadn3 <- gsameth(sig.cpg=pdn3, all.cpg=rownames(dm3), collection=gsets_entrez)
+  gsig_up3 <- rownames(subset(gsaup3,FDR<0.05))
+  gsig_dn3 <- rownames(subset(gsadn3,FDR<0.05))
+  gtup <- names(sim[[6]])
+  gtdn <- names(sim[[7]])
+  UPTP=length(intersect(gsig_up3 ,gtup))
+  UPFP=length(setdiff(gsig_up3 ,gtup))
+  UPFN=length(setdiff(gtup,gsig_up3))
+  DNTP=length(intersect(gsig_dn3 ,gtdn))
+  DNFP=length(setdiff(gsig_dn3 ,gtdn))
+  DNFN=length(setdiff(gtdn,gsig_dn3))
+  TP=UPTP+DNTP
+  FP=UPFP+DNFP
+  FN=UPFN+DNFN
+  TN=nrow(gsadn3)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
+  PREC=TP/(TP+FP)
+  REC=TP/(TP+FN)
+  F1=TP/(TP+(0.5*(FP+FN)))
+  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
+  return(result)
+}
+
+```
+
+## LA function
+
+This process runs limma first and then aggregates the results before doing an enrichment test.
+
+```{r,lafunc}
 
@@
 
+simla <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
+  # generate gene sets
+  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
+  lengths <- unname(unlist(lapply(genesetdatabase,length)))
+  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
+  # select gene sets to alter
+  gset_mod <- gset_selection(genesetdatabase=gsets,seed=seed,gene_catalog=gene_catalog)
+  # cut down length to num_dm_sets
+  set.seed(seed) ; gset_mod <- sample(gset_mod,num_dm_sets)
+  # incorporate select changes
+  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
+    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
+  # set up limma
+  mval2 <- sim$mval
+  ncols <- ncol(mval2)
+  groupsize <- ncols/2
+  ss <- data.frame(colnames(mval2))
+  colnames(ss) <- "sample"
+  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
+  d <- model.matrix(~ ss$case )
+  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
+  dmagg1 <- agg(dm3,cores=4)
+  m1 <- dmagg1$gme_res_df[,"mean",drop=FALSE]
+  lares1 <- ttenrich(m=m1,genesets=gsets,cores=10)
+  gsig_up3 <- rownames(subset(lares1, fdr < 0.05 & t_mean > 0))
+  gsig_dn3 <- rownames(subset(lares1, fdr < 0.05 & t_mean < 0))
+  gtup <- names(sim[[6]])
+  gtdn <- names(sim[[7]])
+  UPTP=length(intersect(gsig_up3 ,gtup))
+  UPFP=length(setdiff(gsig_up3 ,gtup))
+  UPFN=length(setdiff(gtup,gsig_up3))
+  DNTP=length(intersect(gsig_dn3 ,gtdn))
+  DNFP=length(setdiff(gsig_dn3 ,gtdn))
+  DNFN=length(setdiff(gtdn,gsig_dn3))
+  TP=UPTP+DNTP
+  FP=UPFP+DNFP
+  FN=UPFN+DNFN
+  TN=nrow(lares1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
+  PREC=TP/(TP+FP)
+  REC=TP/(TP+FN)
+  F1=TP/(TP+(0.5*(FP+FN)))
+  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
+  return(result)
+}
+
 ```
@@
 
+simal <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
+  # generate gene sets
+  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
+  lengths <- unname(unlist(lapply(genesetdatabase,length)))
+  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
+  # select gene sets to alter
+  gset_mod <- gset_selection(genesetdatabase=gsets,seed=seed,gene_catalog=gene_catalog)
+  # cut down length to num_dm_sets
+  set.seed(seed) ; gset_mod <- sample(gset_mod,num_dm_sets)
+  # incorporate select changes
+  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
+    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
+  # set up limma
+  mval2 <- sim$mval
+  ncols <- ncol(mval2)
+  groupsize <- ncols/2
+  ss <- data.frame(colnames(mval2))
+  colnames(ss) <- "sample"
+  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
+  d <- model.matrix(~ ss$case )
+  # al pipeline
+  medf1 <- chragg(mval=mval2,myann=myann,cores=4)
+  magg1 <- agglimma(medf1,d)
+  m1 <- as.data.frame(magg1$t)
+  rownames(m1) <- rownames(magg1)
+  colnames(m1) <- "t"
+  alres1 <- ttenrich(m=m1,genesets=gsets,cores=10)
+  # summarise results
+  gsig_up3 <- rownames(subset(alres1, fdr < 0.05 & t_mean > 0))
+  gsig_dn3 <- rownames(subset(alres1, fdr < 0.05 & t_mean < 0))
+  gtup <- names(sim[[6]])
+  gtdn <- names(sim[[7]])
+  UPTP=length(intersect(gsig_up3 ,gtup))
+  UPFP=length(setdiff(gsig_up3 ,gtup))
+  UPFN=length(setdiff(gtup,gsig_up3))
+  DNTP=length(intersect(gsig_dn3 ,gtdn))
+  DNFP=length(setdiff(gsig_dn3 ,gtdn))
+  DNFN=length(setdiff(gtdn,gsig_dn3))
+  TP=UPTP+DNTP
+  FP=UPFP+DNFP
+  FN=UPFN+DNFN
+  TN=nrow(alres1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
+  PREC=TP/(TP+FP)
+  REC=TP/(TP+FN)
+  F1=TP/(TP+(0.5*(FP+FN)))
+  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
+  return(result)
+}
+
 ```
@@
 
-#aalres <- aal(mval=mval,myann=myann,genesets=gs_symbols,design=design,cores=CORES)
+simaa <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
+  # generate gene sets
+  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
+  lengths <- unname(unlist(lapply(genesetdatabase,length)))
+  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
+  # select gene sets to alter
+  gset_mod <- gset_selection(genesetdatabase=gsets,seed=seed,gene_catalog=gene_catalog)
+  # cut down length to num_dm_sets
+  set.seed(seed) ; gset_mod <- sample(gset_mod,num_dm_sets)
+  # incorporate select changes
+  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
+    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
+  # set up limma
+  mval2 <- sim$mval
+  ncols <- ncol(mval2)
+  groupsize <- ncols/2
+  ss <- data.frame(colnames(mval2))
+  colnames(ss) <- "sample"
+  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
+  d <- model.matrix(~ ss$case )
+  # aa pipeline
+  aares1 <- aal(mval=mval2, myann=myann, genesets=gsets, design=d, cores=4)
+  # summarise results
+  gsig_up3 <- rownames(subset(aares1, adj.P.Val < 0.05 & logFC > 0))
+  gsig_dn3 <- rownames(subset(aares1, adj.P.Val < 0.05 & logFC < 0))
+  gtup <- names(sim[[6]])
+  gtdn <- names(sim[[7]])
+  UPTP=length(intersect(gsig_up3 ,gtup))
+  UPFP=length(setdiff(gsig_up3 ,gtup))
+  UPFN=length(setdiff(gtup,gsig_up3))
+  DNTP=length(intersect(gsig_dn3 ,gtdn))
+  DNFP=length(setdiff(gsig_dn3 ,gtdn))
+  DNFN=length(setdiff(gtdn,gsig_dn3))
+  TP=UPTP+DNTP
+  FP=UPFP+DNFP
+  FN=UPFN+DNFN
+  TN=nrow(aares1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
+  PREC=TP/(TP+FP)
+  REC=TP/(TP+FN)
+  F1=TP/(TP+(0.5*(FP+FN)))
+  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
+  return(result)
+}
+
+genesetdatabase=gs_symbols
+groupsize=10
+delta=1
+num_dm_sets=50
+
+aares <- mclapply(1:6,function(i) {
+  simaa(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
+    groupsize=10, delta=delta, num_dm_sets=num_dm_sets)
+},mc.cores=6)
+
+aares <- do.call(rbind,aares)
+aares
 
@@
 
-```
+simalm <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
+  # generate gene sets
+  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
+  lengths <- unname(unlist(lapply(genesetdatabase,length)))
+  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
+  # select gene sets to alter
+  gset_mod <- gset_selection(genesetdatabase=gsets,seed=seed,gene_catalog=gene_catalog)
+  # cut down length to num_dm_sets
+  set.seed(seed) ; gset_mod <- sample(gset_mod,num_dm_sets)
+  # incorporate select changes
+  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
+    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
+  # set up limma
+  mval2 <- sim$mval
+  ncols <- ncol(mval2)
+  groupsize <- ncols/2
+  ss <- data.frame(colnames(mval2))
+  colnames(ss) <- "sample"
+  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
+  d <- model.matrix(~ ss$case )
+  # alm pipeline
+  medf1 <- chragg(mval2,myann,cores=4)
+  magg1 <- agglimma(medf1,d)
+  m1 <- as.data.frame(magg1$t)
+  rownames(m1) <- rownames(magg1)
+  colnames(m1) <- "t"
+  almres1 <- runmitch(m=m1,genesets=gsets,cores=4)
+  # summarise results
+  gsig_up3 <- rownames( subset( almres1, p.adjustANOVA < 0.05 & s.dist > 0 ) )
+  gsig_dn3 <- rownames( subset( almres1, p.adjustANOVA < 0.05 & s.dist < 0 ) )
+  gtup <- names(sim[[6]])
+  gtdn <- names(sim[[7]])
+  UPTP=length(intersect(gsig_up3 ,gtup))
+  UPFP=length(setdiff(gsig_up3 ,gtup))
+  UPFN=length(setdiff(gtup,gsig_up3))
+  DNTP=length(intersect(gsig_dn3 ,gtdn))
+  DNFP=length(setdiff(gsig_dn3 ,gtdn))
+  DNFN=length(setdiff(gtdn,gsig_dn3))
+  TP=UPTP+DNTP
+  FP=UPFP+DNFP
+  FN=UPFN+DNFN
+  TN=nrow(almres1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
+  PREC=TP/(TP+FP)
+  REC=TP/(TP+FN)
+  F1=TP/(TP+(0.5*(FP+FN)))
+  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
+  return(result)
+}
 
-## Prep
+genesetdatabase=gs_symbols
+groupsize=10
+delta=1
+num_dm_sets=50
 
-```{r,prep}
+almres <- mclapply(1:6,function(i) {
+  simalm(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
+    groupsize=10, delta=delta, num_dm_sets=num_dm_sets)
+},mc.cores=6)
 
-# prep
-sex <- as.data.frame(design)$sex
-tumor <- as.data.frame(design)$tumor
-patient <- as.character(unlist(lapply(1:ncol(mval),function(i) {c(i,i)})))
-patient <- head(patient,ncol(mval))
-design <- model.matrix(~ patient + tumor )
-rownames(design) <- colnames(mval)
+almres <- do.call(rbind,almres)
+almres
 
@@
 
-## LA workflow split and analyse
+# Run analyses
 
-split 74 samples 37 patients split into 18 + 18
-Conduct LA workflow analysis.
-Then make overlap.
+## GSA meth sim
 
-```{r,laflow}
+```{r,rungsa}
 
-splitanalyze_la <- function(seed){
-  set.seed(seed)
-  a <- sample(1:37,18)
-  a <- a[order(a)]
-  nona <- setdiff(1:37,a)
-  set.seed(seed)
-  b <- sample(nona,18)
-  b <- b[order(b)]
-  a <- c( (a*2)-1 ,a*2 )
-  a <- a[order(a)]
-  b <- c( (b*2)-1 ,b*2 )
-  b <- b[order(b)]
+groupsize=10
+delta=1
+num_dm_sets=50
 
-  design_a <- design[a,]
-  design_a <- design_a[,colSums(design_a)>0]
-  mval_a <- mval[,a]
-  design_b <- design[b,]
-  design_b <- design_b[,colSums(design_b)>0]
-  mval_b <- mval[,b]
+gres <- lapply(1:6,function(i) {
+  simgsa(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
+    groupsize=10, delta=delta, num_dm_sets=num_dm_sets)
+})
 
-  dm1 <- runlimma(mval_a,design_a,myann)
-  dmagg1 <- agg(dm1,cores=floor(CORES/3))
-  m1 <- dmagg1$gme_res_df[,"mean",drop=FALSE]
-  lares1 <- ttenrich(m=m1,genesets=gs_symbols,cores=CORES/3)
-
-  dm2 <- runlimma(mval_b,design_b,myann)
-  dmagg2 <- agg(dm2,cores=floor(CORES/3))
-  m2 <- dmagg2$gme_res_df[,"mean",drop=FALSE]
-  lares2 <- ttenrich(m=m2,genesets=gs_symbols,cores=CORES/3)
-
-  list("lares1"=lares1,"lares2"=lares2)
-
-}
-
-lares <- mclapply( seq(100,5000,100) ,splitanalyze_la,mc.cores=CORES)
+gres <- do.call(rbind,gres)
+gres
 
@@
 
-## AL workflow split and analyse
+## LA sim
 
-```{r,alflow}
+```{r,runla}
 
-splitanalyze_al <- function(seed){
-  set.seed(seed)
-  a <- sample(1:37,18)
-  a <- a[order(a)]
-  nona <- setdiff(1:37,a)
-  set.seed(seed)
-  b <- sample(nona,18)
-  b <- b[order(b)]
-  a <- c( (a*2)-1 ,a*2 )
-  a <- a[order(a)]
-  b <- c( (b*2)-1 ,b*2 )
-  b <- b[order(b)]
+lares <- mclapply(1:6,function(i) {
+  simla(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
+    groupsize=10, delta=delta, num_dm_sets=num_dm_sets)
+},mc.cores=6)
 
-  design_a <- design[a,]
-  design_a <- design_a[,colSums(design_a)>0]
-  mval_a <- mval[,a]
-  design_b <- design[b,]
-  design_b <- design_b[,colSums(design_b)>0]
-  mval_b <- mval[,b]
-
-  medf1 <- chragg(mval_a,myann,cores=floor(CORES/3))
-  magg1 <- agglimma(medf1,design_a)
-  m1 <- as.data.frame(magg1$t)
-  rownames(m1) <- rownames(magg1)
-  colnames(m1) <- "t"
-  alres1 <- ttenrich(m=m1,genesets=gs_symbols,cores=CORES/3)
-
-  medf2 <- chragg(mval_b,myann,cores=floor(CORES/3))
-  magg2 <- agglimma(medf2,design_b)
-  m2 <- as.data.frame(magg2$t)
-  rownames(m2) <- rownames(magg2)
-  colnames(m2) <- "t"
-  alres2 <- ttenrich(m=m2,genesets=gs_symbols,cores=CORES/3)
-
-  list("alres1"=alres1,"alres2"=alres2)
-}
-
-alres <- mclapply( seq(100,5000,100) ,splitanalyze_al,mc.cores=CORES)
+lares <- do.call(rbind,lares)
+lares
 
@@
 
-## AA workflow split and analyse
+## AL sim
 
-```{r,aaflow}
+```{r,runal}
 
-splitanalyze_aa <- function(seed){
-  set.seed(seed)
-  a <- sample(1:37,18)
-  a <- a[order(a)]
-  nona <- setdiff(1:37,a)
-  set.seed(seed)
-  b <- sample(nona,18)
-  b <- b[order(b)]
-  a <- c( (a*2)-1 ,a*2 )
-  a <- a[order(a)]
-  b <- c( (b*2)-1 ,b*2 )
-  b <- b[order(b)]
+alres <- mclapply(1:6,function(i) {
+  simal(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
+    groupsize=10, delta=delta, num_dm_sets=num_dm_sets)
+},mc.cores=6)
 
-  design_a <- design[a,]
-  design_a <- design_a[,colSums(design_a)>0]
-  mval_a <- mval[,a]
-  design_b <- design[b,]
-  design_b <- design_b[,colSums(design_b)>0]
-  mval_b <- mval[,b]
+alres <- do.call(rbind,alres)
+alres
 
-  aares1 <- aal(mval=mval_a,myann=myann,genesets=gs_symbols,
-    design=design_a,cores=floor(CORES/3))
+```
 
-  aares2 <- aal(mval=mval_b,myann=myann,genesets=gs_symbols,
-    design=design_b,cores=floor(CORES/3))
+## AA sim
 
-  list("aares1"=aares1,"aares2"=aares2)
-}
+```{r,runaa}
 
-aares <- mclapply( seq(100,5000,100) ,splitanalyze_aa,mc.cores=CORES)
+aares <- mclapply(1:6,function(i) {
+  simaa(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
+    groupsize=10, delta=delta, num_dm_sets=num_dm_sets)
+},mc.cores=6)
+
+aares <- do.call(rbind,aares)
+aares
 
@@
 
-```{r,almflow}
+```{r,runalm}
 
-splitanalyze_alm <- function(seed){
-  set.seed(seed)
-  a <- sample(1:37,18)
-  a <- a[order(a)]
-  nona <- setdiff(1:37,a)
-  set.seed(seed)
-  b <- sample(nona,18)
-  b <- b[order(b)]
-  a <- c( (a*2)-1 ,a*2 )
-  a <- a[order(a)]
-  b <- c( (b*2)-1 ,b*2 )
-  b <- b[order(b)]
+almres <- mclapply(1:6,function(i) {
+  simalm(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
+    groupsize=10, delta=delta, num_dm_sets=num_dm_sets)
+},mc.cores=6)
 
-  design_a <- design[a,]
-  design_a <- design_a[,colSums(design_a)>0]
-  mval_a <- mval[,a]
-  design_b <- design[b,]
-  design_b <- design_b[,colSums(design_b)>0]
-  mval_b <- mval[,b]
-
-  medf1 <- chragg(mval_a,myann,cores=floor(CORES/3))
-  magg1 <- agglimma(medf1,design_a)
-  m1 <- as.data.frame(magg1$t)
-  rownames(m1) <- rownames(magg1)
-  colnames(m1) <- "t"
-  almres1 <- runmitch(m=m1,genesets=gs_symbols,cores=CORES)
-
-  medf2 <- chragg(mval_b,myann,cores=floor(CORES/3))
-  magg2 <- agglimma(medf2,design_b)
-  m2 <- as.data.frame(magg2$t)
-  rownames(m2) <- rownames(magg2)
-  colnames(m2) <- "t"
-  almres2 <- runmitch(m=m2,genesets=gs_symbols,cores=CORES)
-
-  list("almres1"=almres1,"almres2"=almres2)
-
-}
-
-almres <- mclapply( seq(100,5000,100) ,splitanalyze_alm,mc.cores=CORES)
-
-```
-
-## Compare
-
-```{r,summarise1}
-
-lacompare <- function(lares){
-  lares1 <- lares[[1]]
-  lares2 <- lares[[2]]
-  lasig1 <- subset(lares1,fdr<0.05)
-  lasig2 <- subset(lares2,fdr<0.05)
-  lasig1_up <- rownames(subset(lasig1,t_median>0))
-  lasig1_dn <- rownames(subset(lasig1,t_median<0))
-  lasig2_up <- rownames(subset(lasig2,t_median>0))
-  lasig2_dn <- rownames(subset(lasig2,t_median<0))
-
-  total <- length(unique(c(lasig1_up,lasig2_up,lasig1_dn,lasig2_dn)))
-  common <- length(intersect(lasig1_up, lasig2_up)) + length(intersect(lasig1_dn, lasig2_dn))
-  discordant <- length(intersect(lasig1_up, lasig2_dn)) + length(intersect(lasig1_dn, lasig2_up))
-  uncommon = total - common - discordant
-
-  laout <- data.frame(total,common,discordant,uncommon)
-  laout$p_comm <-laout$common / laout$total
-  laout$p_disc <-laout$discordant / laout$total
-  laout$p_uncom <-laout$uncommon / laout$total
-  laout
-}
-
-alcompare <-function(alres){
-  alres1 <- alres[[1]]
-  alres2 <- alres[[2]]
-  alsig1 <- subset(alres1,fdr<0.05)
-  alsig2 <- subset(alres2,fdr<0.05)
-
-  alsig1_up <- rownames(subset(alsig1,t_median>0))
-  alsig1_dn <- rownames(subset(alsig1,t_median<0))
-  alsig2_up <- rownames(subset(alsig2,t_median>0))
-  alsig2_dn <- rownames(subset(alsig2,t_median<0))
-
-  total <- length(unique(c(alsig1_up,alsig2_up,alsig1_dn,alsig2_dn)))
-  common <- length(intersect(alsig1_up, alsig2_up)) + length(intersect(alsig1_dn, alsig2_dn))
-  discordant <- length(intersect(alsig1_up, alsig2_dn)) + length(intersect(alsig1_dn, alsig2_up))
-  uncommon = total - common - discordant
-  result <- c("total"=total,"common"=common,"discordant"=discordant,"uncommon"=uncommon)
-
-  alout <- data.frame(total,common,discordant,uncommon)
-  alout$p_comm <-alout$common / alout$total
-  alout$p_disc <-alout$discordant / alout$total
-  alout$p_uncom <-alout$uncommon / alout$total
-  alout
-}
-
-aacompare <-function(aa){
-  aa1 <- aa[[1]]
-  aa2 <- aa[[2]]
-  aasig1 <- subset(aa1,adj.P.Val<0.05)
-  aasig2 <- subset(aa2,adj.P.Val<0.05)
-
-  aasig1_up <- rownames(subset(aasig1,logFC>0))
-  aasig1_dn <- rownames(subset(aasig1,logFC<0))
-  aasig2_up <- rownames(subset(aasig2,logFC>0))
-  aasig2_dn <- rownames(subset(aasig2,logFC<0))
-
-  total <- length(unique(c(aasig1_up,aasig2_up,aasig1_dn,aasig2_dn)))
-  common <- length(intersect(aasig1_up, aasig2_up)) + length(intersect(aasig1_dn, aasig2_dn))
-  discordant <- length(intersect(aasig1_up, aasig2_dn)) + length(intersect(aasig1_dn, aasig2_up))
-  uncommon = total - common - discordant
-  result <- c("total"=total,"common"=common,"discordant"=discordant,"uncommon"=uncommon)
-
-  aaout <- data.frame(total,common,discordant,uncommon)
-  aaout$p_comm <- aaout$common / aaout$total
-  aaout$p_disc <- aaout$discordant / aaout$total
-  aaout$p_uncom <-aaout$uncommon / aaout$total
-  aaout
-}
-
-almcompare <-function(almres){
-  almres1 <- almres[[1]]
-  almres2 <- almres[[2]]
-  almsig1 <- subset(almres1,p.adjustANOVA<0.05)
-  almsig2 <- subset(almres2,p.adjustANOVA<0.05)
-
-  almsig1_up <- rownames(subset(almsig1,s.dist>0))
-  almsig1_dn <- rownames(subset(almsig1,s.dist<0))
-  almsig2_up <- rownames(subset(almsig2,s.dist>0))
-  almsig2_dn <- rownames(subset(almsig2,s.dist<0))
-
-  total <- length(unique(c(almsig1_up,almsig2_up,almsig1_dn,almsig2_dn)))
-  common <- length(intersect(almsig1_up, almsig2_up)) + length(intersect(almsig1_dn, almsig2_dn))
-  discordant <- length(intersect(almsig1_up, almsig2_dn)) + length(intersect(almsig1_dn, almsig2_up))
-  uncommon = total - common - discordant
-  result <- c("total"=total,"common"=common,"discordant"=discordant,"uncommon"=uncommon)
-
-  almout <- data.frame(total,common,discordant,uncommon)
-  almout$p_comm <-almout$common / almout$total
-  almout$p_disc <-almout$discordant / almout$total
-  almout$p_uncom <-almout$uncommon / almout$total
-  almout
-}
-
-```
-
-```{r,compare1}
-
-laout <- do.call(rbind,lapply(lares,lacompare))
-
-alout <- do.call(rbind,lapply(alres,alcompare))
-
-aaout <- do.call(rbind,lapply(aares,aacompare))
-
-almout <- do.call(rbind,lapply(almres,almcompare))
-
-laout
-
-alout
-
-aaout
-
-almout
-
-res <- do.call(rbind,lapply(list(laout,alout,aaout,almout),colMeans))
-
-rownames(res) <- c("la","al","aa","alm")
-
-res <- as.data.frame(res)
-
-res %>%
-  kbl(caption="overall summary") %>%
-  kable_paper("hover", full_width = F)
-
-apply(laout,2,mean)
-apply(alout,2,mean)
-apply(aaout,2,mean)
-apply(almout,2,mean)
-
-```
-
-Plot some data.
-
-Raw counts and proportions.
-
-```{r,plots1,fig.height=5,fig.width=9}
-
-par(mfrow=c(1,3))
-
-barplot(res$common,main="no. common",names.arg=c("LA","AL","AA","ALM"))
-barplot(res$uncommon,main="no. uncommon",names.arg=c("LA","AL","AA","ALM"))
-barplot(res$discordant,main="no. discordant",names.arg=c("LA","AL","AA","ALM"))
-
-boxplot(list("LA"=laout$common,"AL"=alout$common,"AA"=aaout$common,"ALM"=almout$common),
-  main="no. common",col="white",ylim=c(0,1550))
-beeswarm(list("LA"=laout$common,"AL"=alout$common,"AA"=aaout$common,"ALM"=almout$common), add=TRUE,cex=1.2,pch=19)
-
-boxplot(list("LA"=laout$uncommon,"AL"=alout$uncommon,"AA"=aaout$uncommon,"ALM"=almout$uncommon),
-  main="no. uncommon",col="white",ylim=c(0,1550))
-beeswarm(list("LA"=laout$uncommon,"AL"=alout$uncommon,"AA"=aaout$uncommon,"ALM"=almout$uncommon), add=TRUE,cex=1.2,pch=19)
-
-boxplot(list("LA"=laout$discordant,"AL"=alout$discordant,"AA"=aaout$discordant,"ALM"=almout$discordant),
-  main="no. discordant",col="white",ylim=c(0,1550))
-beeswarm(list("LA"=laout$discordant,"AL"=alout$discordant,"AA"=aaout$discordant,"ALM"=almout$discordant), add=TRUE,cex=1.2,pch=19)
-
-barplot(res$p_comm,main="proportion common",names.arg=c("LA","AL","AA","ALM"))
-barplot(res$p_uncom,main="proportion uncommon",names.arg=c("LA","AL","AA","ALM"))
-barplot(res$p_disc,main="proportion discordant",names.arg=c("LA","AL","AA","ALM"))
-
-boxplot(list("LA"=laout$p_comm,"AL"=alout$p_comm,"AA"=aaout$p_comm,"ALM"=almout$p_comm),
-  main="proportion common",col="white",ylim=c(0,1))
-beeswarm(list("LA"=laout$p_comm,"AL"=alout$p_comm,"AA"=aaout$p_comm,"ALM"=almout$p_comm), add=TRUE,cex=1.2,pch=19)
-
-boxplot(list("LA"=laout$p_uncom,"AL"=alout$p_uncom,"AA"=aaout$p_uncom,"ALM"=almout$p_uncom),
-  main="proportion uncommon",col="white",ylim=c(0,1))
-beeswarm(list("LA"=laout$p_uncom,"AL"=alout$p_uncom,"AA"=aaout$p_uncom,"ALM"=almout$p_uncom), add=TRUE,cex=1.2,pch=19)
-
-boxplot(list("LA"=laout$p_disc,"AL"=alout$p_disc,"AA"=aaout$p_disc,"ALM"=almout$p_disc),
-  main="proportion discordant",col="white",ylim=c(0,1))
-beeswarm(list("LA"=laout$p_disc,"AL"=alout$p_disc,"AA"=aaout$p_disc,"ALM"=almout$p_disc), add=TRUE,cex=1.2,pch=19)
+almres <- do.call(rbind,almres)
+almres
 
@@
 
-save.image("GSE158422_split.Rdata")
+save.image("GSE158422_simulate.Rdata")
 

### 2023-10-16 figs/GSE158422_simulate.Rmd
simulation

@@
 
+3. Create random gene sets that have similar sized to Reactome pathways.
+
 3. Some gene sets will be selected to be differentially methylated.
-A requirement is that there will be no overlap between them.
 Half of these will be hypermethylated and the others will be hypomethylated.
@@
 Select nonoverlapping gene sets.
+
+Not in use now. Might delete later.
 
@@
   # select gene sets to alter
-  gset_mod <- gset_selection(genesetdatabase=gsets,seed=seed,gene_catalog=gene_catalog)
-  # cut down length to num_dm_sets
-  set.seed(seed) ; gset_mod <- sample(gset_mod,num_dm_sets)
+  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
   # incorporate select changes
@@
   # convert gene sets to entrez
-  gene2entrez <- mapIds(org.Hs.eg.db, gene_catalog, 'ENTREZID', 'SYMBOL')
+  suppressWarnings(suppressMessages({ gene2entrez <- mapIds(org.Hs.eg.db, gene_catalog, 'ENTREZID', 'SYMBOL') }))
   gsets_entrez <- lapply(gsets,function(gs) {
@@
   })
+  suppressWarnings(suppressMessages({
     gsaup3 <- gsameth(sig.cpg=pup3, all.cpg=rownames(dm3), collection=gsets_entrez)
     gsadn3 <- gsameth(sig.cpg=pdn3, all.cpg=rownames(dm3), collection=gsets_entrez)
+  }))
   gsig_up3 <- rownames(subset(gsaup3,FDR<0.05))
@@
 
-Median value works well here.
+Use mean value works well here.
 
@@
 
-genesetdatabase=gs_symbols
-groupsize=10
-delta=1
-num_dm_sets=50
-
-aares <- mclapply(1:6,function(i) {
-  simaa(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
-    groupsize=10, delta=delta, num_dm_sets=num_dm_sets)
-},mc.cores=6)
-
-aares <- do.call(rbind,aares)
-aares
-
 ```
@@
 runmitch <- function(m,genesets,cores=1) {
-  mres <- mitch_calc(m,genesets,minsetsize=5,cores=cores)
+  suppressMessages({ mres <- mitch_calc(m,genesets,minsetsize=5,cores=cores) })
   mres <- mres$enrichment_result
@@
 
-genesetdatabase=gs_symbols
-groupsize=10
-delta=1
-num_dm_sets=50
-
-almres <- mclapply(1:6,function(i) {
-  simalm(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
-    groupsize=10, delta=delta, num_dm_sets=num_dm_sets)
-},mc.cores=6)
-
-almres <- do.call(rbind,almres)
-almres
-
 ```
@@
 
+Set assumptions.
+
+```{r,set}
+
+groupsize=5
+delta=1
+num_dm_sets=50
+
+```
+
 ## GSA meth sim
+
+Cannot be run in multicore due to fragility of AnnotationDbi SQLite objects.
 
@@
 
-groupsize=10
-delta=1
-num_dm_sets=50
-
 gres <- lapply(1:6,function(i) {
   simgsa(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
-    groupsize=10, delta=delta, num_dm_sets=num_dm_sets)
+    groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
 })
@@
   simla(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
-    groupsize=10, delta=delta, num_dm_sets=num_dm_sets)
+    groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
 },mc.cores=6)
@@
   simal(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
-    groupsize=10, delta=delta, num_dm_sets=num_dm_sets)
+    groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
 },mc.cores=6)
@@
   simaa(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
-    groupsize=10, delta=delta, num_dm_sets=num_dm_sets)
+    groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
 },mc.cores=6)
@@
   simalm(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
-    groupsize=10, delta=delta, num_dm_sets=num_dm_sets)
+    groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
 },mc.cores=6)

### 2023-10-16 figs/GSE158422_simulate.Rmd
bugfix

@@
   pdn <- names(probe2gene[which(probe2gene %in% gdn2)])
-  set.seed(seed) ; pdn2 <- sample(pdn,floor(length(pup)*frac_probes))
+  set.seed(seed) ; pdn2 <- sample(pdn,floor(length(pdn)*frac_probes))
   # divide samples between ctrl and case
@@
   pdn3 <- rownames(subset(dm3,adj.P.Val<0.05 & logFC<0))
+  if ( length(pup3) < 250 ) { pup3 <- head(rownames(subset(dm3, logFC > 0)), 250) }
+  if ( length(pdn3) < 250 ) { pdn3 <- head(rownames(subset(dm3, logFC < 0)), 250) }
   # convert gene sets to entrez
@@
 
-groupsize=5
-delta=1
+groupsize=3
+delta=0.6
 num_dm_sets=50
@@
 
-gres <- lapply(1:6,function(i) {
+gres <- lapply(1:100,function(i) {
   simgsa(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
@@
 
-lares <- mclapply(1:6,function(i) {
+lares <- mclapply(1:100,function(i) {
   simla(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
@@
 
-alres <- mclapply(1:6,function(i) {
+alres <- mclapply(1:100,function(i) {
   simal(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
@@
 
-aares <- mclapply(1:6,function(i) {
+aares <- mclapply(1:100,function(i) {
   simaa(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
@@
 
-almres <- mclapply(1:6,function(i) {
+almres <- mclapply(1:100,function(i) {
   simalm(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,

### 2023-10-17 figs/GSE158422_simulate.Rmd
looks like it is working

@@
 false positives.
-A better solution could be to select random gene sets with sizes similar to Reactomes.
-Reactomes were downloaded on the 14th Sept 2023 from MsigDB.
-
-```{r,gsdl}
-
-gs_entrez <- gmt_import("c2.cp.reactome.v2023.1.Hs.entrez.gmt")
-
-gs_symbols <- gmt_import("c2.cp.reactome.v2023.1.Hs.symbols.gmt")
-
-```
-
-Get a vector of genes, and the length of gene sets from Reactome.
-
-Then generate random gene sets.
+A better solution could be to select random gene sets with size range between 10 and 100 genes.
 
@@
 
-lengths <- unname(unlist(lapply(gs_symbols,length)))
+lengths <- rep(1:10,100)*10
 
@@
   m1 <- dmagg1$gme_res_df[,"mean",drop=FALSE]
-  lares1 <- ttenrich(m=m1,genesets=gsets,cores=10)
+  lares1 <- ttenrich(m=m1,genesets=gsets,cores=4)
   gsig_up3 <- rownames(subset(lares1, fdr < 0.05 & t_mean > 0))
@@
   colnames(m1) <- "t"
-  alres1 <- ttenrich(m=m1,genesets=gsets,cores=10)
+  alres1 <- ttenrich(m=m1,genesets=gsets,cores=4)
   # summarise results
@@
 
-groupsize=3
-delta=0.6
 num_dm_sets=50
+sims=4
+
+groupsizes=c(3,6,12,18)
+deltas=c(0.1,0.3,0.5,0.7,1.0)
+params <- expand.grid("groupsizes"=groupsizes,"deltas"=deltas)
+params
 
@@
 
-gres <- lapply(1:100,function(i) {
-  simgsa(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
-    groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
+gres <- lapply(1:nrow(params) , function(j) {
+  groupsize = params[j,1]
+  delta = params[j,2]
+  res <- lapply(1:sims,function(i) {
+    simgsa(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5,
+      frac_probes=0.5, groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
+  })
+  res <- do.call(rbind,res)
+  return(res)
 })
 
-gres <- do.call(rbind,gres)
 gres
@@
 
-lares <- mclapply(1:100,function(i) {
+lares <- lapply(1:nrow(params) , function(j) {
+  groupsize = params[j,1]
+  delta = params[j,2]
+  res <- mclapply(1:sims,function(i) {
     simla(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
       groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
-},mc.cores=6)
+  },mc.cores=4)
+  res <- do.call(rbind,res)
+  return(res)
+})
 
-lares <- do.call(rbind,lares)
 lares
@@
 
-alres <- mclapply(1:100,function(i) {
+alres <- lapply(1:nrow(params) , function(j) {
+  groupsize = params[j,1]
+  delta = params[j,2]
+  res <- mclapply(1:50,function(i) {
     simal(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
       groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
-},mc.cores=6)
+  },mc.cores=4)
+  res <- do.call(rbind,alres)
+  return(res)
+})
 
-alres <- do.call(rbind,alres)
 alres
@@
 
-aares <- mclapply(1:100,function(i) {
+aares <- lapply(1:nrow(params) , function(j) {
+  groupsize = params[j,1]
+  delta = params[j,2]
+  res <- mclapply(1:50,function(i) {
     simaa(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
       groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
-},mc.cores=6)
+  },mc.cores=4)
+  res <- do.call(rbind,res)
+  return(res)
+})
 
-aares <- do.call(rbind,aares)
 aares
@@
 
-almres <- mclapply(1:100,function(i) {
+almres <- lapply(1:nrow(params) , function(j) {
+  groupsize = params[j,1]
+  delta = params[j,2]
+  res <- mclapply(1:sims,function(i) {
     simalm(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
       groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
-},mc.cores=6)
+  },mc.cores=4)
+  res <- do.call(rbind,res)
+  return(res)
+})
 
-almres <- do.call(rbind,almres)
 almres

### 2023-10-18 figs/GSE158422_simulate.Rmd
summarising simulation data

@@
 
-# enrich
-ttenrich <- function(m,genesets,cores=1) {
+# enrich parametric
+ttenrich <- function(m,genesets,cores=1,testtype="selfcontained") {
   res <- mclapply( 1:length(genesets), function(i) {
+    scores <- m[,1]
     gs <- genesets[i]
@@
       mymedian <- median(tstats)
-      wt <- t.test(tstats)
+      if ( testtype == "selfcontained" ) { wt <- t.test(tstats) }
+      if ( testtype == "competitive" ) { wt <- t.test(tstats,scores) }
       res <- c(name,myn,mymean,mymedian,wt$p.value)
@@
 
+# enrich non-parametric
+wtenrich <- function(m,genesets,cores=1,testtype="selfcontained") {
+  res <- mclapply( 1:length(genesets), function(i) {
+    scores <- m[,1]
+    gs <- genesets[i]
+    name <- names(gs)
+    n_members <- length(which(rownames(m) %in% gs[[1]]))
+    if ( n_members > 4 ) {
+      tstats <- m[which(rownames(m) %in% gs[[1]]),]
+      myn <- length(tstats)
+      mymean <- mean(tstats)
+      mymedian <- median(tstats)
+      if ( testtype == "selfcontained" ) { wt <- wilcox.test(tstats) }
+      if ( testtype == "competitive" ) { wt <- wilcox.test(tstats,scores) }
+      res <- c(name,myn,mymean,mymedian,wt$p.value)
+    }
+  } , mc.cores = cores)
+  res_df <- do.call(rbind, res)
+  rownames(res_df) <- res_df[,1]
+  res_df <- res_df[,-1]
+  colnames(res_df) <- c("n_genes","t_mean","t_median","pval")
+  tmp <- apply(res_df,2,as.numeric)
+  rownames(tmp) <- rownames(res_df)
+  res_df <- tmp
+  res_df <- as.data.frame(res_df)
+  res_df <- res_df[order(res_df$pval),]
+  res_df$logp <- -log10(res_df$pval )
+  res_df$fdr <- p.adjust(res_df$pval,method="fdr")
+  res_df[order(abs(res_df$pval)),]
+  return(res_df)
+}
+
+# parametric self-contained
 simla <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
@@
   m1 <- dmagg1$gme_res_df[,"mean",drop=FALSE]
-  lares1 <- ttenrich(m=m1,genesets=gsets,cores=4)
+  lares1 <- ttenrich(m=m1,genesets=gsets,cores=4,testtype="selfcontained")
   gsig_up3 <- rownames(subset(lares1, fdr < 0.05 & t_mean > 0))
@@
 
+# parametric competitive
+simlac <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
+  # generate gene sets
+  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
+  lengths <- unname(unlist(lapply(genesetdatabase,length)))
+  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
+  # select gene sets to alter
+  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
+  # incorporate select changes
+  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
+    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
+  # set up limma
+  mval2 <- sim$mval
+  ncols <- ncol(mval2)
+  groupsize <- ncols/2
+  ss <- data.frame(colnames(mval2))
+  colnames(ss) <- "sample"
+  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
+  d <- model.matrix(~ ss$case )
+  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
+  dmagg1 <- agg(dm3,cores=4)
+  m1 <- dmagg1$gme_res_df[,"mean",drop=FALSE]
+  lares1 <- ttenrich(m=m1,genesets=gsets,cores=4,testtype="competitive")
+  gsig_up3 <- rownames(subset(lares1, fdr < 0.05 & t_mean > 0))
+  gsig_dn3 <- rownames(subset(lares1, fdr < 0.05 & t_mean < 0))
+  gtup <- names(sim[[6]])
+  gtdn <- names(sim[[7]])
+  UPTP=length(intersect(gsig_up3 ,gtup))
+  UPFP=length(setdiff(gsig_up3 ,gtup))
+  UPFN=length(setdiff(gtup,gsig_up3))
+  DNTP=length(intersect(gsig_dn3 ,gtdn))
+  DNFP=length(setdiff(gsig_dn3 ,gtdn))
+  DNFN=length(setdiff(gtdn,gsig_dn3))
+  TP=UPTP+DNTP
+  FP=UPFP+DNFP
+  FN=UPFN+DNFN
+  TN=nrow(lares1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
+  PREC=TP/(TP+FP)
+  REC=TP/(TP+FN)
+  F1=TP/(TP+(0.5*(FP+FN)))
+  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
+  return(result)
+}
+
+# nonparametric self-contained
+simnla <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
+  # generate gene sets
+  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
+  lengths <- unname(unlist(lapply(genesetdatabase,length)))
+  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
+  # select gene sets to alter
+  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
+  # incorporate select changes
+  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
+    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
+  # set up limma
+  mval2 <- sim$mval
+  ncols <- ncol(mval2)
+  groupsize <- ncols/2
+  ss <- data.frame(colnames(mval2))
+  colnames(ss) <- "sample"
+  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
+  d <- model.matrix(~ ss$case )
+  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
+  dmagg1 <- agg(dm3,cores=4)
+  m1 <- dmagg1$gme_res_df[,"mean",drop=FALSE]
+  lares1 <- wtenrich(m=m1,genesets=gsets,cores=4,testtype="selfcontained")
+  gsig_up3 <- rownames(subset(lares1, fdr < 0.05 & t_mean > 0))
+  gsig_dn3 <- rownames(subset(lares1, fdr < 0.05 & t_mean < 0))
+  gtup <- names(sim[[6]])
+  gtdn <- names(sim[[7]])
+  UPTP=length(intersect(gsig_up3 ,gtup))
+  UPFP=length(setdiff(gsig_up3 ,gtup))
+  UPFN=length(setdiff(gtup,gsig_up3))
+  DNTP=length(intersect(gsig_dn3 ,gtdn))
+  DNFP=length(setdiff(gsig_dn3 ,gtdn))
+  DNFN=length(setdiff(gtdn,gsig_dn3))
+  TP=UPTP+DNTP
+  FP=UPFP+DNFP
+  FN=UPFN+DNFN
+  TN=nrow(lares1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
+  PREC=TP/(TP+FP)
+  REC=TP/(TP+FN)
+  F1=TP/(TP+(0.5*(FP+FN)))
+  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
+  return(result)
+}
+
+# nonparametric competitive
+simnlac <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
+  # generate gene sets
+  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
+  lengths <- unname(unlist(lapply(genesetdatabase,length)))
+  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
+  # select gene sets to alter
+  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
+  # incorporate select changes
+  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
+    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
+  # set up limma
+  mval2 <- sim$mval
+  ncols <- ncol(mval2)
+  groupsize <- ncols/2
+  ss <- data.frame(colnames(mval2))
+  colnames(ss) <- "sample"
+  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
+  d <- model.matrix(~ ss$case )
+  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
+  dmagg1 <- agg(dm3,cores=4)
+  m1 <- dmagg1$gme_res_df[,"mean",drop=FALSE]
+  lares1 <- wtenrich(m=m1,genesets=gsets,cores=4,testtype="competitive")
+  gsig_up3 <- rownames(subset(lares1, fdr < 0.05 & t_mean > 0))
+  gsig_dn3 <- rownames(subset(lares1, fdr < 0.05 & t_mean < 0))
+  gtup <- names(sim[[6]])
+  gtdn <- names(sim[[7]])
+  UPTP=length(intersect(gsig_up3 ,gtup))
+  UPFP=length(setdiff(gsig_up3 ,gtup))
+  UPFN=length(setdiff(gtup,gsig_up3))
+  DNTP=length(intersect(gsig_dn3 ,gtdn))
+  DNFP=length(setdiff(gsig_dn3 ,gtdn))
+  DNFN=length(setdiff(gtdn,gsig_dn3))
+  TP=UPTP+DNTP
+  FP=UPFP+DNFP
+  FN=UPFN+DNFN
+  TN=nrow(lares1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
+  PREC=TP/(TP+FP)
+  REC=TP/(TP+FN)
+  F1=TP/(TP+(0.5*(FP+FN)))
+  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
+  return(result)
+}
+
 ```
@@
 
+# AL approach parametric self-contained
 simal <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
@@
   colnames(m1) <- "t"
-  alres1 <- ttenrich(m=m1,genesets=gsets,cores=4)
+  alres1 <- ttenrich(m=m1,genesets=gsets,cores=4,testtype="selfcontained")
+  # summarise results
+  gsig_up3 <- rownames(subset(alres1, fdr < 0.05 & t_mean > 0))
+  gsig_dn3 <- rownames(subset(alres1, fdr < 0.05 & t_mean < 0))
+  gtup <- names(sim[[6]])
+  gtdn <- names(sim[[7]])
+  UPTP=length(intersect(gsig_up3 ,gtup))
+  UPFP=length(setdiff(gsig_up3 ,gtup))
+  UPFN=length(setdiff(gtup,gsig_up3))
+  DNTP=length(intersect(gsig_dn3 ,gtdn))
+  DNFP=length(setdiff(gsig_dn3 ,gtdn))
+  DNFN=length(setdiff(gtdn,gsig_dn3))
+  TP=UPTP+DNTP
+  FP=UPFP+DNFP
+  FN=UPFN+DNFN
+  TN=nrow(alres1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
+  PREC=TP/(TP+FP)
+  REC=TP/(TP+FN)
+  F1=TP/(TP+(0.5*(FP+FN)))
+  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
+  return(result)
+}
+
+# AL approach parametric competitive
+simalc <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
+  # generate gene sets
+  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
+  lengths <- unname(unlist(lapply(genesetdatabase,length)))
+  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
+  # select gene sets to alter
+  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
+  # incorporate select changes
+  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
+    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
+  # set up limma
+  mval2 <- sim$mval
+  ncols <- ncol(mval2)
+  groupsize <- ncols/2
+  ss <- data.frame(colnames(mval2))
+  colnames(ss) <- "sample"
+  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
+  d <- model.matrix(~ ss$case )
+  # al pipeline
+  medf1 <- chragg(mval=mval2,myann=myann,cores=4)
+  magg1 <- agglimma(medf1,d)
+  m1 <- as.data.frame(magg1$t)
+  rownames(m1) <- rownames(magg1)
+  colnames(m1) <- "t"
+  alres1 <- ttenrich(m=m1,genesets=gsets,cores=4,testtype="competitive")
+  # summarise results
+  gsig_up3 <- rownames(subset(alres1, fdr < 0.05 & t_mean > 0))
+  gsig_dn3 <- rownames(subset(alres1, fdr < 0.05 & t_mean < 0))
+  gtup <- names(sim[[6]])
+  gtdn <- names(sim[[7]])
+  UPTP=length(intersect(gsig_up3 ,gtup))
+  UPFP=length(setdiff(gsig_up3 ,gtup))
+  UPFN=length(setdiff(gtup,gsig_up3))
+  DNTP=length(intersect(gsig_dn3 ,gtdn))
+  DNFP=length(setdiff(gsig_dn3 ,gtdn))
+  DNFN=length(setdiff(gtdn,gsig_dn3))
+  TP=UPTP+DNTP
+  FP=UPFP+DNFP
+  FN=UPFN+DNFN
+  TN=nrow(alres1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
+  PREC=TP/(TP+FP)
+  REC=TP/(TP+FN)
+  F1=TP/(TP+(0.5*(FP+FN)))
+  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
+  return(result)
+}
+
+# AL approach nonparametric self-contained
+simnal <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
+  # generate gene sets
+  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
+  lengths <- unname(unlist(lapply(genesetdatabase,length)))
+  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
+  # select gene sets to alter
+  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
+  # incorporate select changes
+  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
+    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
+  # set up limma
+  mval2 <- sim$mval
+  ncols <- ncol(mval2)
+  groupsize <- ncols/2
+  ss <- data.frame(colnames(mval2))
+  colnames(ss) <- "sample"
+  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
+  d <- model.matrix(~ ss$case )
+  # al pipeline
+  medf1 <- chragg(mval=mval2,myann=myann,cores=4)
+  magg1 <- agglimma(medf1,d)
+  m1 <- as.data.frame(magg1$t)
+  rownames(m1) <- rownames(magg1)
+  colnames(m1) <- "t"
+  alres1 <- wtenrich(m=m1,genesets=gsets,cores=4,testtype="selfcontained")
+  # summarise results
+  gsig_up3 <- rownames(subset(alres1, fdr < 0.05 & t_mean > 0))
+  gsig_dn3 <- rownames(subset(alres1, fdr < 0.05 & t_mean < 0))
+  gtup <- names(sim[[6]])
+  gtdn <- names(sim[[7]])
+  UPTP=length(intersect(gsig_up3 ,gtup))
+  UPFP=length(setdiff(gsig_up3 ,gtup))
+  UPFN=length(setdiff(gtup,gsig_up3))
+  DNTP=length(intersect(gsig_dn3 ,gtdn))
+  DNFP=length(setdiff(gsig_dn3 ,gtdn))
+  DNFN=length(setdiff(gtdn,gsig_dn3))
+  TP=UPTP+DNTP
+  FP=UPFP+DNFP
+  FN=UPFN+DNFN
+  TN=nrow(alres1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
+  PREC=TP/(TP+FP)
+  REC=TP/(TP+FN)
+  F1=TP/(TP+(0.5*(FP+FN)))
+  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
+  return(result)
+}
+
+# AL approach nonparametric competitive
+simnalc <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
+  # generate gene sets
+  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
+  lengths <- unname(unlist(lapply(genesetdatabase,length)))
+  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
+  # select gene sets to alter
+  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
+  # incorporate select changes
+  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
+    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
+  # set up limma
+  mval2 <- sim$mval
+  ncols <- ncol(mval2)
+  groupsize <- ncols/2
+  ss <- data.frame(colnames(mval2))
+  colnames(ss) <- "sample"
+  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
+  d <- model.matrix(~ ss$case )
+  # al pipeline
+  medf1 <- chragg(mval=mval2,myann=myann,cores=4)
+  magg1 <- agglimma(medf1,d)
+  m1 <- as.data.frame(magg1$t)
+  rownames(m1) <- rownames(magg1)
+  colnames(m1) <- "t"
+  alres1 <- wtenrich(m=m1,genesets=gsets,cores=4,testtype="competitive")
   # summarise results
@@
   res <- lapply(1:sims,function(i) {
-    simgsa(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5,
+    simgsa(genesetdatabase=gsets, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5,
       frac_probes=0.5, groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
@@
 
+gres2 <- do.call(rbind,lapply(gres,colMeans))
+
+gres2
+
 ```
@@
 ## LA sim
+
+### parametric self-contained
 
@@
   res <- mclapply(1:sims,function(i) {
-    simla(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
+    simla(genesetdatabase=gsets, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
       groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
@@
 
-lares
-
-```
-
-## AL sim
-
-```{r,runal}
-
-alres <- lapply(1:nrow(params) , function(j) {
+lares2 <- do.call(rbind,lapply(lares,colMeans))
+
+lares2
+
+```
+
+### parametric competitive
+
+```{r,runlac}
+
+lacres <- lapply(1:nrow(params) , function(j) {
   groupsize = params[j,1]
   delta = params[j,2]
-  res <- mclapply(1:50,function(i) {
-    simal(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
-      groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
-  },mc.cores=4)
-  res <- do.call(rbind,alres)
-  return(res)
-})
-
-alres
-
-```
-
-## AA sim
-
-```{r,runaa}
-
-aares <- lapply(1:nrow(params) , function(j) {
-  groupsize = params[j,1]
-  delta = params[j,2]
-  res <- mclapply(1:50,function(i) {
-    simaa(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
+  res <- mclapply(1:sims,function(i) {
+    simlac(genesetdatabase=gsets, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
       groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
@@
 
+lacres
+
+lacres2 <- do.call(rbind,lapply(lacres,colMeans))
+
+lacres2
+
+```
+
+### nonparametric self-contained
+
+```{r,runnla}
+
+nlares <- lapply(1:nrow(params) , function(j) {
+  groupsize = params[j,1]
+  delta = params[j,2]
+  res <- mclapply(1:sims,function(i) {
+    simnla(genesetdatabase=gsets, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
+      groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
+  },mc.cores=4)
+  res <- do.call(rbind,res)
+  return(res)
+})
+
+nlares
+
+nlares2 <- do.call(rbind,lapply(nlares,colMeans))
+
+nlares2
+
+```
+
+### nonparametric competitive
+
+```{r,runnlac}
+
+nlacres <- lapply(1:nrow(params) , function(j) {
+  groupsize = params[j,1]
+  delta = params[j,2]
+  res <- mclapply(1:sims,function(i) {
+    simnlac(genesetdatabase=gsets, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
+      groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
+  },mc.cores=4)
+  res <- do.call(rbind,res)
+  return(res)
+})
+
+nlacres
+
+nlacres2 <- do.call(rbind,lapply(nlacres,colMeans))
+
+nlacres2
+
+```
+
+## AL sim
+
+### parametric self-contained
+
+```{r,runal}
+
+alres <- lapply(1:nrow(params) , function(j) {
+  groupsize = params[j,1]
+  delta = params[j,2]
+  res <- mclapply(1:sims,function(i) {
+    simal(genesetdatabase=gsets, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
+      groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
+  },mc.cores=4)
+  res <- do.call(rbind,res)
+  return(res)
+})
+
+alres
+
+alres2 <- do.call(rbind,lapply(alres,colMeans))
+
+alres2
+
+```
+
+### parametric competitive
+
+```{r,runalc}
+
+alcres <- lapply(1:nrow(params) , function(j) {
+  groupsize = params[j,1]
+  delta = params[j,2]
+  res <- mclapply(1:sims,function(i) {
+    simalc(genesetdatabase=gsets, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
+      groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
+  },mc.cores=4)
+  res <- do.call(rbind,res)
+  return(res)
+})
+
+alcres
+
+alcres2 <- do.call(rbind,lapply(alcres,colMeans))
+
+alcres2
+
+```
+
+### nonparametric self-contained
+
+```{r,runnal}
+
+nalres <- lapply(1:nrow(params) , function(j) {
+  groupsize = params[j,1]
+  delta = params[j,2]
+  res <- mclapply(1:sims,function(i) {
+    simnal(genesetdatabase=gsets, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
+      groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
+  },mc.cores=4)
+  res <- do.call(rbind,res)
+  return(res)
+})
+
+nalres
+
+nalres2 <- do.call(rbind,lapply(nalres,colMeans))
+
+nalres2
+
+```
+
+### nonparametric competitive
+
+```{r,runnalc}
+
+nalcres <- lapply(1:nrow(params) , function(j) {
+  groupsize = params[j,1]
+  delta = params[j,2]
+  res <- mclapply(1:sims,function(i) {
+    simnalc(genesetdatabase=gsets, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
+      groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
+  },mc.cores=4)
+  res <- do.call(rbind,res)
+  return(res)
+})
+
+nalcres
+
+nalcres2 <- do.call(rbind,lapply(nalcres,colMeans))
+
+nalcres2
+
+```
+
+## AA sim
+
+```{r,runaa}
+
+aares <- lapply(1:nrow(params) , function(j) {
+  groupsize = params[j,1]
+  delta = params[j,2]
+  res <- mclapply(1:sims,function(i) {
+    simaa(genesetdatabase=gsets, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
+      groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
+  },mc.cores=4)
+  res <- do.call(rbind,res)
+  return(res)
+})
+
 aares
+
+aares2 <- do.call(rbind,lapply(aares,colMeans))
+
+aares2
 
@@
   res <- mclapply(1:sims,function(i) {
-    simalm(genesetdatabase=gs_symbols, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
+    simalm(genesetdatabase=gsets, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
       groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
@@
 
+almres2 <- do.call(rbind,lapply(almres,colMeans))
+
+almres2
+
 ```
@@
 ```
+
+## Notes
+
+LA
+
+* simla: parametric self-contained
+
+* simlac: parametric competitive
+
+* simnla: nonparametric self-contained
+
+* simnlac: nonparametric competitive
+
+AL
+
+* simal: parametric self-contained
+
+* simalc: parametric competitive
+
+* simnal: nonparametric self-contained
+
+* simnalc: nonparametric competitive

### 2023-10-18 figs/GSE158422_simulate.Rmd
progress on simulation

@@
   library('org.Hs.eg.db')
-  library("tictoc")
+  library("psych")
   library("mitch")
   library("kableExtra")
-  library("beeswarm")
-  library("eulerr")
 })
@@
 
+F1 <- function(x,y) {
+  ( 2 * x * y ) / ( x + y )
+}
+
 ```
@@
 num_dm_sets=50
-sims=4
-
-groupsizes=c(3,6,12,18)
-deltas=c(0.1,0.3,0.5,0.7,1.0)
+sims=10
+
+groupsizes=c(3,6,12)
+deltas=c(0.1,0.2,0.3,0.5)
 params <- expand.grid("groupsizes"=groupsizes,"deltas"=deltas)
@@
 
+gres3p <- do.call(rbind,lapply(groupsizes, function (g) { gres2[params$groupsizes==g,"PREC"] }))
+colnames(gres3p) <- deltas
+rownames(gres3p) <- groupsizes
+gres3p
+
+gres3r <- do.call(rbind,lapply(groupsizes, function (g) { gres2[params$groupsizes==g,"REC"] }))
+colnames(gres3r) <- deltas
+rownames(gres3r) <- groupsizes
+gres3r
+
+F1(gres3p,gres3r)
+
 ```
@@
 
+lares3p <- do.call(rbind,lapply(groupsizes, function (g) { lares2[params$groupsizes==g,"PREC"] }))
+colnames(lares3p) <- deltas
+rownames(lares3p) <- groupsizes
+lares3p
+
+lares3r <- do.call(rbind,lapply(groupsizes, function (g) { lares2[params$groupsizes==g,"REC"] }))
+colnames(lares3r) <- deltas
+rownames(lares3r) <- groupsizes
+lares3r
+
+F1(lares3p,lares3r)
+
 ```
@@
 
+lacres3p <- do.call(rbind,lapply(groupsizes, function (g) { lacres2[params$groupsizes==g,"PREC"] }))
+colnames(lacres3p) <- deltas
+rownames(lacres3p) <- groupsizes
+lacres3p
+
+lacres3r <- do.call(rbind,lapply(groupsizes, function (g) { lacres2[params$groupsizes==g,"REC"] }))
+colnames(lacres3r) <- deltas
+rownames(lacres3r) <- groupsizes
+lacres3r
+
+F1(lacres3p,lacres3r)
+
 ```
@@
 
+nlares3p <- do.call(rbind,lapply(groupsizes, function (g) { nlares2[params$groupsizes==g,"PREC"] }))
+colnames(nlares3p) <- deltas
+rownames(nlares3p) <- groupsizes
+nlares3p
+
+nlares3r <- do.call(rbind,lapply(groupsizes, function (g) { nlares2[params$groupsizes==g,"REC"] }))
+colnames(nlares3r) <- deltas
+rownames(nlares3r) <- groupsizes
+nlares3r
+
+F1(nlares3p,nlares3r)
+
 ```
@@
 
+nlacres3p <- do.call(rbind,lapply(groupsizes, function (g) { nlacres2[params$groupsizes==g,"PREC"] }))
+colnames(nlacres3p) <- deltas
+rownames(nlacres3p) <- groupsizes
+nlacres3p
+
+nlacres3r <- do.call(rbind,lapply(groupsizes, function (g) { nlacres2[params$groupsizes==g,"REC"] }))
+colnames(nlacres3r) <- deltas
+rownames(nlacres3r) <- groupsizes
+nlacres3r
+
+F1(nlacres3p,nlacres3r)
+
 ```
@@
 
+alres3p <- do.call(rbind,lapply(groupsizes, function (g) { alres2[params$groupsizes==g,"PREC"] }))
+colnames(alres3p) <- deltas
+rownames(alres3p) <- groupsizes
+alres3p
+
+alres3r <- do.call(rbind,lapply(groupsizes, function (g) { alres2[params$groupsizes==g,"REC"] }))
+colnames(alres3r) <- deltas
+rownames(alres3r) <- groupsizes
+alres3r
+
+F1(alres3p,alres3r)
+
 ```
@@
 
+alcres3p <- do.call(rbind,lapply(groupsizes, function (g) { alcres2[params$groupsizes==g,"PREC"] }))
+colnames(alcres3p) <- deltas
+rownames(alcres3p) <- groupsizes
+alcres3p
+
+alcres3r <- do.call(rbind,lapply(groupsizes, function (g) { alcres2[params$groupsizes==g,"REC"] }))
+colnames(alcres3r) <- deltas
+rownames(alcres3r) <- groupsizes
+alcres3r
+
+F1(alcres3p,alcres3r)
+
 ```
@@
 
+nalres3p <- do.call(rbind,lapply(groupsizes, function (g) { nalres2[params$groupsizes==g,"PREC"] }))
+colnames(nalres3p) <- deltas
+rownames(nalres3p) <- groupsizes
+nalres3p
+
+nalres3r <- do.call(rbind,lapply(groupsizes, function (g) { nalres2[params$groupsizes==g,"REC"] }))
+colnames(nalres3r) <- deltas
+rownames(nalres3r) <- groupsizes
+nalres3r
+
+F1(nalres3p,nalres3r)
+
 ```
@@
 
+nalcres3p <- do.call(rbind,lapply(groupsizes, function (g) { nalcres2[params$groupsizes==g,"PREC"] }))
+colnames(nalcres3p) <- deltas
+rownames(nalcres3p) <- groupsizes
+nalcres3p
+
+nalcres3r <- do.call(rbind,lapply(groupsizes, function (g) { nalcres2[params$groupsizes==g,"REC"] }))
+colnames(nalcres3r) <- deltas
+rownames(nalcres3r) <- groupsizes
+nalcres3r
+
+F1(nalcres3p,nalcres3r)
+
 ```
@@
 
+aares3p <- do.call(rbind,lapply(groupsizes, function (g) { aares2[params$groupsizes==g,"PREC"] }))
+colnames(aares3p) <- deltas
+rownames(aares3p) <- groupsizes
+aares3p
+
+aares3r <- do.call(rbind,lapply(groupsizes, function (g) { aares2[params$groupsizes==g,"REC"] }))
+colnames(aares3r) <- deltas
+rownames(aares3r) <- groupsizes
+aares3r
+
+F1(aares3p,aares3r)
+
 ```
@@
 
+almres3p <- do.call(rbind,lapply(groupsizes, function (g) { almres2[params$groupsizes==g,"PREC"] }))
+colnames(almres3p) <- deltas
+rownames(almres3p) <- groupsizes
+almres3p
+
+almres3r <- do.call(rbind,lapply(groupsizes, function (g) { almres2[params$groupsizes==g,"REC"] }))
+colnames(almres3r) <- deltas
+rownames(almres3r) <- groupsizes
+almres3r
+
+F1(almres3p,almres3r)
+
 ```

### 2023-10-19 figs/GSE158422_simulate.Rmd
simulation results

--- 2023-10-18 figs/GSE158422_simulate.Rmd
@@
 
-# parametric self-contained
-simla <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
-  # generate gene sets
-  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
-  lengths <- unname(unlist(lapply(genesetdatabase,length)))
-  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
-  # select gene sets to alter
-  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
-  # incorporate select changes
-  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
-    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
-  # set up limma
-  mval2 <- sim$mval
-  ncols <- ncol(mval2)
-  groupsize <- ncols/2
-  ss <- data.frame(colnames(mval2))
-  colnames(ss) <- "sample"
-  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
-  d <- model.matrix(~ ss$case )
-  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
-  dmagg1 <- agg(dm3,cores=4)
-  m1 <- dmagg1$gme_res_df[,"mean",drop=FALSE]
-  lares1 <- ttenrich(m=m1,genesets=gsets,cores=4,testtype="selfcontained")
-  gsig_up3 <- rownames(subset(lares1, fdr < 0.05 & t_mean > 0))
-  gsig_dn3 <- rownames(subset(lares1, fdr < 0.05 & t_mean < 0))
-  gtup <- names(sim[[6]])
-  gtdn <- names(sim[[7]])
-  UPTP=length(intersect(gsig_up3 ,gtup))
-  UPFP=length(setdiff(gsig_up3 ,gtup))
-  UPFN=length(setdiff(gtup,gsig_up3))
-  DNTP=length(intersect(gsig_dn3 ,gtdn))
-  DNFP=length(setdiff(gsig_dn3 ,gtdn))
-  DNFN=length(setdiff(gtdn,gsig_dn3))
-  TP=UPTP+DNTP
-  FP=UPFP+DNFP
-  FN=UPFN+DNFN
-  TN=nrow(lares1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
-  PREC=TP/(TP+FP)
-  REC=TP/(TP+FN)
-  F1=TP/(TP+(0.5*(FP+FN)))
-  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
-  return(result)
-}
-
 # parametric competitive
@@
 
-# nonparametric self-contained
-simnla <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
-  # generate gene sets
-  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
-  lengths <- unname(unlist(lapply(genesetdatabase,length)))
-  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
-  # select gene sets to alter
-  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
-  # incorporate select changes
-  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
-    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
-  # set up limma
-  mval2 <- sim$mval
-  ncols <- ncol(mval2)
-  groupsize <- ncols/2
-  ss <- data.frame(colnames(mval2))
-  colnames(ss) <- "sample"
-  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
-  d <- model.matrix(~ ss$case )
-  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
-  dmagg1 <- agg(dm3,cores=4)
-  m1 <- dmagg1$gme_res_df[,"mean",drop=FALSE]
-  lares1 <- wtenrich(m=m1,genesets=gsets,cores=4,testtype="selfcontained")
-  gsig_up3 <- rownames(subset(lares1, fdr < 0.05 & t_mean > 0))
-  gsig_dn3 <- rownames(subset(lares1, fdr < 0.05 & t_mean < 0))
-  gtup <- names(sim[[6]])
-  gtdn <- names(sim[[7]])
-  UPTP=length(intersect(gsig_up3 ,gtup))
-  UPFP=length(setdiff(gsig_up3 ,gtup))
-  UPFN=length(setdiff(gtup,gsig_up3))
-  DNTP=length(intersect(gsig_dn3 ,gtdn))
-  DNFP=length(setdiff(gsig_dn3 ,gtdn))
-  DNFN=length(setdiff(gtdn,gsig_dn3))
-  TP=UPTP+DNTP
-  FP=UPFP+DNFP
-  FN=UPFN+DNFN
-  TN=nrow(lares1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
-  PREC=TP/(TP+FP)
-  REC=TP/(TP+FN)
-  F1=TP/(TP+(0.5*(FP+FN)))
-  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
-  return(result)
-}
-
 # nonparametric competitive
@@
   return(dmagg)
-}
-
-# AL approach parametric self-contained
-simal <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
-  # generate gene sets
-  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
-  lengths <- unname(unlist(lapply(genesetdatabase,length)))
-  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
-  # select gene sets to alter
-  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
-  # incorporate select changes
-  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
-    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
-  # set up limma
-  mval2 <- sim$mval
-  ncols <- ncol(mval2)
-  groupsize <- ncols/2
-  ss <- data.frame(colnames(mval2))
-  colnames(ss) <- "sample"
-  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
-  d <- model.matrix(~ ss$case )
-  # al pipeline
-  medf1 <- chragg(mval=mval2,myann=myann,cores=4)
-  magg1 <- agglimma(medf1,d)
-  m1 <- as.data.frame(magg1$t)
-  rownames(m1) <- rownames(magg1)
-  colnames(m1) <- "t"
-  alres1 <- ttenrich(m=m1,genesets=gsets,cores=4,testtype="selfcontained")
-  # summarise results
-  gsig_up3 <- rownames(subset(alres1, fdr < 0.05 & t_mean > 0))
-  gsig_dn3 <- rownames(subset(alres1, fdr < 0.05 & t_mean < 0))
-  gtup <- names(sim[[6]])
-  gtdn <- names(sim[[7]])
-  UPTP=length(intersect(gsig_up3 ,gtup))
-  UPFP=length(setdiff(gsig_up3 ,gtup))
-  UPFN=length(setdiff(gtup,gsig_up3))
-  DNTP=length(intersect(gsig_dn3 ,gtdn))
-  DNFP=length(setdiff(gsig_dn3 ,gtdn))
-  DNFN=length(setdiff(gtdn,gsig_dn3))
-  TP=UPTP+DNTP
-  FP=UPFP+DNFP
-  FN=UPFN+DNFN
-  TN=nrow(alres1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
-  PREC=TP/(TP+FP)
-  REC=TP/(TP+FN)
-  F1=TP/(TP+(0.5*(FP+FN)))
-  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
-  return(result)
 }
@@
 
-# AL approach nonparametric self-contained
-simnal <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
-  # generate gene sets
-  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
-  lengths <- unname(unlist(lapply(genesetdatabase,length)))
-  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
-  # select gene sets to alter
-  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
-  # incorporate select changes
-  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
-    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
-  # set up limma
-  mval2 <- sim$mval
-  ncols <- ncol(mval2)
-  groupsize <- ncols/2
-  ss <- data.frame(colnames(mval2))
-  colnames(ss) <- "sample"
-  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
-  d <- model.matrix(~ ss$case )
-  # al pipeline
-  medf1 <- chragg(mval=mval2,myann=myann,cores=4)
-  magg1 <- agglimma(medf1,d)
-  m1 <- as.data.frame(magg1$t)
-  rownames(m1) <- rownames(magg1)
-  colnames(m1) <- "t"
-  alres1 <- wtenrich(m=m1,genesets=gsets,cores=4,testtype="selfcontained")
-  # summarise results
-  gsig_up3 <- rownames(subset(alres1, fdr < 0.05 & t_mean > 0))
-  gsig_dn3 <- rownames(subset(alres1, fdr < 0.05 & t_mean < 0))
-  gtup <- names(sim[[6]])
-  gtdn <- names(sim[[7]])
-  UPTP=length(intersect(gsig_up3 ,gtup))
-  UPFP=length(setdiff(gsig_up3 ,gtup))
-  UPFN=length(setdiff(gtup,gsig_up3))
-  DNTP=length(intersect(gsig_dn3 ,gtdn))
-  DNFP=length(setdiff(gsig_dn3 ,gtdn))
-  DNFN=length(setdiff(gtdn,gsig_dn3))
-  TP=UPTP+DNTP
-  FP=UPFP+DNFP
-  FN=UPFN+DNFN
-  TN=nrow(alres1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
-  PREC=TP/(TP+FP)
-  REC=TP/(TP+FN)
-  F1=TP/(TP+(0.5*(FP+FN)))
-  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
-  return(result)
-}
-
 # AL approach nonparametric competitive
@@
 
-## AA Aggregate-aggregate-limma functions
-
-Use mean value works well here.
-
-```{r,aafunc}
-
-gsagg <- function(x,genesets,cores=1) {
-  meds <- mclapply(1:length(genesets), function(i) {
-    gs = genesets[[i]]
-    xx <- x[rownames(x) %in% gs,]
-    med <- apply(xx,2,mean)
-  },mc.cores=cores)
-  mymed <- do.call(rbind,meds)
-  rownames(mymed) <- names(genesets)
-  as.data.frame(mymed)
-}
-
-aalimma <- function(agag,design) {
-  fit.reduced <- lmFit(agag,design)
-  fit.reduced <- eBayes(fit.reduced)
-  dmagg <- topTable(fit.reduced,coef=ncol(design), number = Inf)
-  return(dmagg)
-}
-
-aal <- function(mval,myann,genesets,design,cores=1) {
-  medf <- chragg(mval,myann,cores=cores)
-  agag <- gsagg(x=medf,genesets=genesets,cores=cores)
-  aalres <- aalimma(agag=agag,design=design)
-  return(aalres)
-}
-
-simaa <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
-  # generate gene sets
-  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
-  lengths <- unname(unlist(lapply(genesetdatabase,length)))
-  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
-  # select gene sets to alter
-  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
-  # incorporate select changes
-  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
-    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
-  # set up limma
-  mval2 <- sim$mval
-  ncols <- ncol(mval2)
-  groupsize <- ncols/2
-  ss <- data.frame(colnames(mval2))
-  colnames(ss) <- "sample"
-  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
-  d <- model.matrix(~ ss$case )
-  # aa pipeline
-  aares1 <- aal(mval=mval2, myann=myann, genesets=gsets, design=d, cores=4)
-  # summarise results
-  gsig_up3 <- rownames(subset(aares1, adj.P.Val < 0.05 & logFC > 0))
-  gsig_dn3 <- rownames(subset(aares1, adj.P.Val < 0.05 & logFC < 0))
-  gtup <- names(sim[[6]])
-  gtdn <- names(sim[[7]])
-  UPTP=length(intersect(gsig_up3 ,gtup))
-  UPFP=length(setdiff(gsig_up3 ,gtup))
-  UPFN=length(setdiff(gtup,gsig_up3))
-  DNTP=length(intersect(gsig_dn3 ,gtdn))
-  DNFP=length(setdiff(gsig_dn3 ,gtdn))
-  DNFN=length(setdiff(gtdn,gsig_dn3))
-  TP=UPTP+DNTP
-  FP=UPFP+DNFP
-  FN=UPFN+DNFN
-  TN=nrow(aares1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
-  PREC=TP/(TP+FP)
-  REC=TP/(TP+FN)
-  F1=TP/(TP+(0.5*(FP+FN)))
-  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
-  return(result)
-}
-
-```
-
 ## Agg-limma-mitch function
@@
 num_dm_sets=50
-sims=4
-
-groupsizes=c(3,6,12,18)
-deltas=c(0.1,0.3,0.5,0.7,1.0)
+sims=10
+
+groupsizes=c(3,5,8,12)
+deltas=c(0.1,0.2,0.3,0.4,0.5)
 params <- expand.grid("groupsizes"=groupsizes,"deltas"=deltas)
-params
+
+params %>% kbl(caption="Parameters to test") %>% kable_paper("hover", full_width = F)
 
@@
 
-gres
-
 gres2 <- do.call(rbind,lapply(gres,colMeans))
 
-gres2
+gres2 %>% kbl(caption="GSAmeth results") %>% kable_paper("hover", full_width = F)
+
+gres3p <- do.call(rbind,lapply(groupsizes, function (g) { gres2[params$groupsizes==g,"PREC"] }))
+colnames(gres3p) <- deltas
+rownames(gres3p) <- groupsizes
+gres3p %>% kbl(caption="GSAmeth precision") %>% kable_paper("hover", full_width = F)
+
+gres3r <- do.call(rbind,lapply(groupsizes, function (g) { gres2[params$groupsizes==g,"REC"] }))
+colnames(gres3r) <- deltas
+rownames(gres3r) <- groupsizes
+gres3r %>% kbl(caption="GSAmeth recall") %>% kable_paper("hover", full_width = F)
+
+F1(gres3p,gres3r) %>% kbl(caption="GSAmeth F1") %>% kable_paper("hover", full_width = F)
 
@@
 ## LA sim
-
-### parametric self-contained
-
-```{r,runla}
-
-lares <- lapply(1:nrow(params) , function(j) {
-  groupsize = params[j,1]
-  delta = params[j,2]
-  res <- mclapply(1:sims,function(i) {
-    simla(genesetdatabase=gsets, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
-      groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
-  },mc.cores=4)
-  res <- do.call(rbind,res)
-  return(res)
-})
-
-lares2 <- do.call(rbind,lapply(lares,colMeans))
-
-lares2
-
-```
 
@@
 
-lacres
-
 lacres2 <- do.call(rbind,lapply(lacres,colMeans))
 
-lacres2
-
-```
-
-### nonparametric self-contained
-
-```{r,runnla}
-
-nlares <- lapply(1:nrow(params) , function(j) {
-  groupsize = params[j,1]
-  delta = params[j,2]
-  res <- mclapply(1:sims,function(i) {
-    simnla(genesetdatabase=gsets, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
-      groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
-  },mc.cores=4)
-  res <- do.call(rbind,res)
-  return(res)
-})
-
-nlares
-
-nlares2 <- do.call(rbind,lapply(nlares,colMeans))
-
-nlares2
+lacres2 %>% kbl(caption="LA parametric results") %>% kable_paper("hover", full_width = F)
+
+lacres3p <- do.call(rbind,lapply(groupsizes, function (g) { lacres2[params$groupsizes==g,"PREC"] }))
+colnames(lacres3p) <- deltas
+rownames(lacres3p) <- groupsizes
+lacres3p %>% kbl(caption="LA parametric precision") %>% kable_paper("hover", full_width = F)
+
+lacres3r <- do.call(rbind,lapply(groupsizes, function (g) { lacres2[params$groupsizes==g,"REC"] }))
+colnames(lacres3r) <- deltas
+rownames(lacres3r) <- groupsizes
+lacres3r %>% kbl(caption="LA parametric recall") %>% kable_paper("hover", full_width = F)
+
+F1(lacres3p,lacres3r) %>% kbl(caption="LA parametric F1") %>% kable_paper("hover", full_width = F)
 
@@
 
-nlacres
-
 nlacres2 <- do.call(rbind,lapply(nlacres,colMeans))
 
-nlacres2
+nlacres2 %>% kbl(caption="LA nonparametric results") %>% kable_paper("hover", full_width = F)
+
+nlacres3p <- do.call(rbind,lapply(groupsizes, function (g) { nlacres2[params$groupsizes==g,"PREC"] }))
+colnames(nlacres3p) <- deltas
+rownames(nlacres3p) <- groupsizes
+nlacres3p %>% kbl(caption="LA nonparametric precision") %>% kable_paper("hover", full_width = F)
+
+nlacres3r <- do.call(rbind,lapply(groupsizes, function (g) { nlacres2[params$groupsizes==g,"REC"] }))
+colnames(nlacres3r) <- deltas
+rownames(nlacres3r) <- groupsizes
+nlacres3r %>% kbl(caption="LA nonparametric recall") %>% kable_paper("hover", full_width = F)
+
+F1(nlacres3p,nlacres3r) %>% kbl(caption="LA nonparametric F1") %>% kable_paper("hover", full_width = F)
 
@@
 ## AL sim
-
-### parametric self-contained
-
-```{r,runal}
-
-alres <- lapply(1:nrow(params) , function(j) {
-  groupsize = params[j,1]
-  delta = params[j,2]
-  res <- mclapply(1:sims,function(i) {
-    simal(genesetdatabase=gsets, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
-      groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
-  },mc.cores=4)
-  res <- do.call(rbind,res)
-  return(res)
-})
-
-alres
-
-alres2 <- do.call(rbind,lapply(alres,colMeans))
-
-alres2
-
-```
 
@@
 
-alcres
-
 alcres2 <- do.call(rbind,lapply(alcres,colMeans))
 
-alcres2
-
-```
-
-### nonparametric self-contained
-
-```{r,runnal}
-
-nalres <- lapply(1:nrow(params) , function(j) {
-  groupsize = params[j,1]
-  delta = params[j,2]
-  res <- mclapply(1:sims,function(i) {
-    simnal(genesetdatabase=gsets, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
-      groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
-  },mc.cores=4)
-  res <- do.call(rbind,res)
-  return(res)
-})
-
-nalres
-
-nalres2 <- do.call(rbind,lapply(nalres,colMeans))
-
-nalres2
+alcres2 %>% kbl(caption="AL parametric results") %>% kable_paper("hover", full_width = F)
+
+alcres3p <- do.call(rbind,lapply(groupsizes, function (g) { alcres2[params$groupsizes==g,"PREC"] }))
+colnames(alcres3p) <- deltas
+rownames(alcres3p) <- groupsizes
+alcres3p %>% kbl(caption="AL parametric precision") %>% kable_paper("hover", full_width = F)
+
+alcres3r <- do.call(rbind,lapply(groupsizes, function (g) { alcres2[params$groupsizes==g,"REC"] }))
+colnames(alcres3r) <- deltas
+rownames(alcres3r) <- groupsizes
+alcres3r %>% kbl(caption="AL parametric recall") %>% kable_paper("hover", full_width = F)
+
+F1(alcres3p,alcres3r) %>% kbl(caption="AL parametric F1") %>% kable_paper("hover", full_width = F)
 
@@
 
-nalcres
-
 nalcres2 <- do.call(rbind,lapply(nalcres,colMeans))
 
-nalcres2
-
-```
-
-## AA sim
-
-```{r,runaa}
-
-aares <- lapply(1:nrow(params) , function(j) {
-  groupsize = params[j,1]
-  delta = params[j,2]
-  res <- mclapply(1:sims,function(i) {
-    simaa(genesetdatabase=gsets, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
-      groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
-  },mc.cores=4)
-  res <- do.call(rbind,res)
-  return(res)
-})
-
-aares
-
-aares2 <- do.call(rbind,lapply(aares,colMeans))
-
-aares2
+nalcres2 %>% kbl(caption="AL nonparametric results") %>% kable_paper("hover", full_width = F)
+
+nalcres3p <- do.call(rbind,lapply(groupsizes, function (g) { nalcres2[params$groupsizes==g,"PREC"] }))
+colnames(nalcres3p) <- deltas
+rownames(nalcres3p) <- groupsizes
+nalcres3p %>% kbl(caption="AL nonparametric precision") %>% kable_paper("hover", full_width = F)
+
+nalcres3r <- do.call(rbind,lapply(groupsizes, function (g) { nalcres2[params$groupsizes==g,"REC"] }))
+colnames(nalcres3r) <- deltas
+rownames(nalcres3r) <- groupsizes
+nalcres3r %>% kbl(caption="AL nonparametric recall") %>% kable_paper("hover", full_width = F)
+
+F1(nalcres3p,nalcres3r) %>% kbl(caption="AL nonparametric F1") %>% kable_paper("hover", full_width = F)
 
@@
 
-almres
-
 almres2 <- do.call(rbind,lapply(almres,colMeans))
 
-almres2
+almres2 %>% kbl(caption="ALM results") %>% kable_paper("hover", full_width = F)
+
+almres3p <- do.call(rbind,lapply(groupsizes, function (g) { almres2[params$groupsizes==g,"PREC"] }))
+colnames(almres3p) <- deltas
+rownames(almres3p) <- groupsizes
+almres3p %>% kbl(caption="ALM precision") %>% kable_paper("hover", full_width = F)
+
+almres3r <- do.call(rbind,lapply(groupsizes, function (g) { almres2[params$groupsizes==g,"REC"] }))
+colnames(almres3r) <- deltas
+rownames(almres3r) <- groupsizes
+almres3r %>% kbl(caption="ALM recall") %>% kable_paper("hover", full_width = F)
+
+F1(almres3p,almres3r) %>% kbl(caption="ALM F1") %>% kable_paper("hover", full_width = F)
 

### 2023-10-23 figs/GSE158422_simulate.Rmd
aggregate types: top, geometric mean

@@
 if (!file.exists("GSE158422_design.rds")) {
+  options(timeout=10000000000000)
   download.file("https://ziemann-lab.net/public/gmea_prototype/GSE158422_mx.rds","GSE158422_mx.rds")
@@
 
+# geometric mean might come in handy
+geomean <- function(x) { exp(mean(log(x))) }
+
 # aggregate
@@
     mymedian <- median(tstats)
+    top <- tstats[order(-abs(tstats))][1]
     if ( length(tstats) > 2 ) {
@@
     res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,
-      "median"=mymedian, pval=pval)
+      "median"=mymedian, "top"=top, "pval"=pval)
   } )
@@
   gme_res_df <- as.data.frame(tmp)
-  gme_res_df$sig <- -log10(gme_res_df[,4])
+  gme_res_df$sig <- -log10(gme_res_df[,"pval"])
+  gme_res_df <- gme_res_df[order(-gme_res_df$sig),]
+  gme_res_df$fdr <- p.adjust(gme_res_df$pval)
+  out <- list("df"=df,"gme_res_df"=gme_res_df)
+  return(out)
+}
+
+# geometric mean aggregate
+geoagg <- function(dm,cores=1) {
+  gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
+  gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
+  gnl <- mclapply(gnl,unique,mc.cores=cores)
+  dm$UCSC_RefGene_Name <- gnl
+  l <- mclapply(1:nrow(dm), function(i) {
+    a <- dm[i,]
+    len <- length(a[[1]][[1]])
+    tvals <- as.numeric(rep(a["t"],len))
+    genes <- a[[1]][[1]]
+    data.frame(genes,tvals)
+  },mc.cores=cores)
+  df <- do.call(rbind,l)
+  keep <- names(which(table(df$genes)>1))
+  df <- df[df$genes %in% keep,]
+  gn <- unique(df$genes)
+  gme_res <- lapply( 1:length(gn), function(i) {
+    g <- gn[i]
+    tstats <- df[which(df$genes==g),"tvals"]
+    myn <- length(tstats)
+    mymean <- geomean(tstats)
+    mymedian <- median(tstats)
+    top <- tstats[order(-abs(tstats))][1]
+    if ( length(tstats) > 2 ) {
+      ttest <- t.test(tstats)
+      pval <- ttest$p.value
+    } else {
+      pval = 1
+    }
+    res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,
+      "median"=mymedian, "top"=top, "pval"=pval)
+  } )
+  gme_res_df <- do.call(rbind, gme_res)
+  rownames(gme_res_df) <- gme_res_df[,1]
+  gme_res_df <- gme_res_df[,-1]
+  tmp <- apply(gme_res_df,2,as.numeric)
+  rownames(tmp) <- rownames(gme_res_df)
+  gme_res_df <- as.data.frame(tmp)
+  gme_res_df$sig <- -log10(gme_res_df[,"pval"])
   gme_res_df <- gme_res_df[order(-gme_res_df$sig),]
@@
   dmagg1 <- agg(dm3,cores=4)
+  m1 <- dmagg1$gme_res_df[,"mean",drop=FALSE]
+  lares1 <- ttenrich(m=m1,genesets=gsets,cores=4,testtype="competitive")
+  gsig_up3 <- rownames(subset(lares1, fdr < 0.05 & t_mean > 0))
+  gsig_dn3 <- rownames(subset(lares1, fdr < 0.05 & t_mean < 0))
+  gtup <- names(sim[[6]])
+  gtdn <- names(sim[[7]])
+  UPTP=length(intersect(gsig_up3 ,gtup))
+  UPFP=length(setdiff(gsig_up3 ,gtup))
+  UPFN=length(setdiff(gtup,gsig_up3))
+  DNTP=length(intersect(gsig_dn3 ,gtdn))
+  DNFP=length(setdiff(gsig_dn3 ,gtdn))
+  DNFN=length(setdiff(gtdn,gsig_dn3))
+  TP=UPTP+DNTP
+  FP=UPFP+DNFP
+  FN=UPFN+DNFN
+  TN=nrow(lares1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
+  PREC=TP/(TP+FP)
+  REC=TP/(TP+FN)
+  F1=TP/(TP+(0.5*(FP+FN)))
+  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
+  return(result)
+}
+
+# parametric competitive top
+simlactop <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
+  # generate gene sets
+  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
+  lengths <- unname(unlist(lapply(genesetdatabase,length)))
+  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
+  # select gene sets to alter
+  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
+  # incorporate select changes
+  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
+    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
+  # set up limma
+  mval2 <- sim$mval
+  ncols <- ncol(mval2)
+  groupsize <- ncols/2
+  ss <- data.frame(colnames(mval2))
+  colnames(ss) <- "sample"
+  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
+  d <- model.matrix(~ ss$case )
+  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
+  dmagg1 <- agg(dm3,cores=4)
+  m1 <- dmagg1$gme_res_df[,"top",drop=FALSE]
+  lares1 <- ttenrich(m=m1,genesets=gsets,cores=4,testtype="competitive")
+  gsig_up3 <- rownames(subset(lares1, fdr < 0.05 & t_mean > 0))
+  gsig_dn3 <- rownames(subset(lares1, fdr < 0.05 & t_mean < 0))
+  gtup <- names(sim[[6]])
+  gtdn <- names(sim[[7]])
+  UPTP=length(intersect(gsig_up3 ,gtup))
+  UPFP=length(setdiff(gsig_up3 ,gtup))
+  UPFN=length(setdiff(gtup,gsig_up3))
+  DNTP=length(intersect(gsig_dn3 ,gtdn))
+  DNFP=length(setdiff(gsig_dn3 ,gtdn))
+  DNFN=length(setdiff(gtdn,gsig_dn3))
+  TP=UPTP+DNTP
+  FP=UPFP+DNFP
+  FN=UPFN+DNFN
+  TN=nrow(lares1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
+  PREC=TP/(TP+FP)
+  REC=TP/(TP+FN)
+  F1=TP/(TP+(0.5*(FP+FN)))
+  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
+  return(result)
+}
+
+# parametric competitive geometric
+simlacgeo <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
+  # generate gene sets
+  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
+  lengths <- unname(unlist(lapply(genesetdatabase,length)))
+  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
+  # select gene sets to alter
+  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
+  # incorporate select changes
+  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
+    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
+  # set up limma
+  mval2 <- sim$mval
+  ncols <- ncol(mval2)
+  groupsize <- ncols/2
+  ss <- data.frame(colnames(mval2))
+  colnames(ss) <- "sample"
+  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
+  d <- model.matrix(~ ss$case )
+  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
+  dmagg1 <- geoagg(dm3,cores=4)
   m1 <- dmagg1$gme_res_df[,"mean",drop=FALSE]
@@
       groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
-  },mc.cores=4)
+  },mc.cores=2)
   res <- do.call(rbind,res)

### 2023-10-23 figs/GSE158422_simulate.Rmd
geometric sim

@@
 
+### parametric competitive with "top" aggregation
+
+```{r,runlac}
+
+lactopres <- lapply(1:nrow(params) , function(j) {
+  groupsize = params[j,1]
+  delta = params[j,2]
+  res <- mclapply(1:sims,function(i) {
+    simlactop(genesetdatabase=gsets, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
+      groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
+  },mc.cores=2)
+  res <- do.call(rbind,res)
+  return(res)
+})
+
+lactopres2 <- do.call(rbind,lapply(lactopres,colMeans))
+
+lactopres2 %>% kbl(caption="LA parametric results (top agg)") %>% kable_paper("hover", full_width = F)
+
+lactopres3p <- do.call(rbind,lapply(groupsizes, function (g) { lactopres2[params$groupsizes==g,"PREC"] }))
+colnames(lactopres3p) <- deltas
+rownames(lactopres3p) <- groupsizes
+lactopres3p %>% kbl(caption="LA parametric precision (top agg)") %>% kable_paper("hover", full_width = F)
+
+lactopres3r <- do.call(rbind,lapply(groupsizes, function (g) { lactopres2[params$groupsizes==g,"REC"] }))
+colnames(lactopres3r) <- deltas
+rownames(lactopres3r) <- groupsizes
+lactopres3r %>% kbl(caption="LA parametric recall (top agg)") %>% kable_paper("hover", full_width = F)
+
+F1(lactopres3p,lactopres3r) %>% kbl(caption="LA parametric F1 (top agg)") %>% kable_paper("hover", full_width = F)
+
+```
+
+### parametric competitive aggregation with geometric mean
+
+```{r,runlac}
+
+lacgeores <- lapply(1:nrow(params) , function(j) {
+  groupsize = params[j,1]
+  delta = params[j,2]
+  res <- mclapply(1:sims,function(i) {
+    simlacgeo(genesetdatabase=gsets, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
+      groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
+  },mc.cores=2)
+  res <- do.call(rbind,res)
+  return(res)
+})
+
+lacgeores2 <- do.call(rbind,lapply(lacgeores,colMeans))
+
+lacgeores2 %>% kbl(caption="LA parametric results with geometric mean") %>% kable_paper("hover", full_width = F)
+
+lacgeores3p <- do.call(rbind,lapply(groupsizes, function (g) { lacgeores2[params$groupsizes==g,"PREC"] }))
+colnames(lacgeores3p) <- deltas
+rownames(lacgeores3p) <- groupsizes
+lacgeores3p %>% kbl(caption="LA parametric precision with geometric mean") %>% kable_paper("hover", full_width = F)
+
+lacgeores3r <- do.call(rbind,lapply(groupsizes, function (g) { lacgeores2[params$groupsizes==g,"REC"] }))
+colnames(lacgeores3r) <- deltas
+rownames(lacgeores3r) <- groupsizes
+lacgeores3r %>% kbl(caption="LA parametric recall with geometric mean") %>% kable_paper("hover", full_width = F)
+
+F1(lacgeores3p,lacgeores3r) %>% kbl(caption="LA parametric F1 with geometric mean") %>% kable_paper("hover", full_width = F)
+
+```
+
 ### nonparametric competitive

### 2023-10-23 figs/GSE158422_simulate.Rmd
geomean

@@
 suppressPackageStartupMessages({
+  library("splicejam")
   library("stringi")
@@
 
-# geometric mean might come in handy
-geomean <- function(x) { exp(mean(log(x))) }
-
 # aggregate
@@
     myn <- length(tstats)
-    mymean <- geomean(tstats)
+    mymean <- splicejam::jamGeomean(tstats)
     mymedian <- median(tstats)
@@
 
-groupsizes=c(3,5,8,12)
-deltas=c(0.1,0.2,0.3,0.4,0.5)
+#groupsizes=c(3,5,8,12)
+#deltas=c(0.1,0.2,0.3,0.4,0.5)
+
+groupsizes=5
+deltas=0.4
+
 params <- expand.grid("groupsizes"=groupsizes,"deltas"=deltas)
@@
 
+Too many false positives.
+
 ```{r,runlac}
@@
 
+Precision is same but recall was very low.
+
 ```{r,runlac}

### 2023-10-26 figs/GSE158422_simulate.Rmd
slight mods

--- 2023-10-23 figs/GSE158422_simulate.Rmd
@@
     myn <- length(tstats)
-    mymean <- geomean(tstats)
+    mymean <- jamGeomean(tstats)
     mymedian <- median(tstats)
@@
       groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
-  },mc.cores=2)
+  },mc.cores=4)
   res <- do.call(rbind,res)
@@
 
-```{r,runlac}
+Too many false positives.
+
+```{r,runlactop}
 
@@
 
-```{r,runlac}
+Precision is same but recall was very low.
+
+```{r,runlacgeomean}
+
+jamGeomean <- function(x, na.rm=TRUE) {
+   x2 <- mean(log2(1 + abs(x)) * sign(x),
+      na.rm=na.rm);
+   sign(x2) * (2 ^ abs(x2) - 1);
+}
 

### 2023-10-27 figs/GSE158422_simulate050.Rmd
update with LA mitch

--- 2023-10-26 figs/GSE158422_simulate.Rmd
@@
 false positives.
-A better solution could be to select random gene sets with size range between 10 and 100 genes.
+After some testing, I found that the gene set size strongly impacts the accuracy.
+
+For reference, Reactome sets have a mean of 48 and median of 15, while MSigDB has a median of 47 and mean of 116.
+
+Therefore, a reasonable analysis would include small (20), medium (50) and large (100) sets.
 
@@
 
-lengths <- rep(1:10,100)*10
+lengths <- rep(50,1000)
 
@@
 
-# parametric competitive
+# LA parametric competitive
 simlac <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
@@
 
-# parametric competitive top
+# LA parametric competitive top
 simlactop <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
@@
 
-# parametric competitive geometric
+# LA parametric competitive geometric
 simlacgeo <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
@@
 
-# nonparametric competitive
+# LA nonparametric competitive
 simnlac <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
@@
   gsig_dn3 <- rownames(subset(lares1, fdr < 0.05 & t_mean < 0))
+  gtup <- names(sim[[6]])
+  gtdn <- names(sim[[7]])
+  UPTP=length(intersect(gsig_up3 ,gtup))
+  UPFP=length(setdiff(gsig_up3 ,gtup))
+  UPFN=length(setdiff(gtup,gsig_up3))
+  DNTP=length(intersect(gsig_dn3 ,gtdn))
+  DNFP=length(setdiff(gsig_dn3 ,gtdn))
+  DNFN=length(setdiff(gtdn,gsig_dn3))
+  TP=UPTP+DNTP
+  FP=UPFP+DNFP
+  FN=UPFN+DNFN
+  TN=nrow(lares1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
+  PREC=TP/(TP+FP)
+  REC=TP/(TP+FN)
+  F1=TP/(TP+(0.5*(FP+FN)))
+  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
+  return(result)
+}
+
+# LA competitive mitch
+simlacm <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
+  # generate gene sets
+  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
+  lengths <- unname(unlist(lapply(genesetdatabase,length)))
+  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
+  # select gene sets to alter
+  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
+  # incorporate select changes
+  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
+    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
+  # set up limma
+  mval2 <- sim$mval
+  ncols <- ncol(mval2)
+  groupsize <- ncols/2
+  ss <- data.frame(colnames(mval2))
+  colnames(ss) <- "sample"
+  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
+  d <- model.matrix(~ ss$case )
+  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
+  dmagg1 <- agg(dm3,cores=4)
+  m1 <- dmagg1$gme_res_df[,"mean",drop=FALSE]
+  lamres1 <- runmitch(m=m1,genesets=gsets,cores=4)
+  gsig_up3 <- rownames( subset( lamres1, p.adjustANOVA < 0.05 & s.dist > 0 ) )
+  gsig_dn3 <- rownames( subset( lamres1, p.adjustANOVA < 0.05 & s.dist < 0 ) )
   gtup <- names(sim[[6]])
@@
 num_dm_sets=50
-sims=10
+sims=4
 
@@
 #deltas=c(0.1,0.2,0.3,0.4,0.5)
-
 groupsizes=5
@@
 
-```{r,runlacgeomean}
-
-jamGeomean <- function(x, na.rm=TRUE) {
-   x2 <- mean(log2(1 + abs(x)) * sign(x),
-      na.rm=na.rm);
-   sign(x2) * (2 ^ abs(x2) - 1);
-}
-
-lacgeores <- lapply(1:nrow(params) , function(j) {
-  groupsize = params[j,1]
-  delta = params[j,2]
-  res <- mclapply(1:sims,function(i) {
-    simlacgeo(genesetdatabase=gsets, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
-      groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
-  },mc.cores=4)
-  res <- do.call(rbind,res)
-  return(res)
-})
-
-lacgeores2 <- do.call(rbind,lapply(lacgeores,colMeans))
-
-lacgeores2 %>% kbl(caption="LA parametric results with geometric mean") %>% kable_paper("hover", full_width = F)
-
-lacgeores3p <- do.call(rbind,lapply(groupsizes, function (g) { lacgeores2[params$groupsizes==g,"PREC"] }))
-colnames(lacgeores3p) <- deltas
-rownames(lacgeores3p) <- groupsizes
-lacgeores3p %>% kbl(caption="LA parametric precision with geometric mean") %>% kable_paper("hover", full_width = F)
-
-lacgeores3r <- do.call(rbind,lapply(groupsizes, function (g) { lacgeores2[params$groupsizes==g,"REC"] }))
-colnames(lacgeores3r) <- deltas
-rownames(lacgeores3r) <- groupsizes
-lacgeores3r %>% kbl(caption="LA parametric recall with geometric mean") %>% kable_paper("hover", full_width = F)
-
-F1(lacgeores3p,lacgeores3r) %>% kbl(caption="LA parametric F1 with geometric mean") %>% kable_paper("hover", full_width = F)
-
-```
+Not included.
 
@@
 
+### LA competitive mitch
+
+```{r,runlacm}
+
+simlacm
+
+lacmres <- lapply(1:nrow(params) , function(j) {
+  groupsize = params[j,1]
+  delta = params[j,2]
+  res <- mclapply(1:sims,function(i) {
+    simlacm(genesetdatabase=gsets, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
+      groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
+  },mc.cores=4)
+  res <- do.call(rbind,res)
+  return(res)
+})
+
+lacmres2 <- do.call(rbind,lapply(lacmres,colMeans))
+
+lacmres2 %>% kbl(caption="LA mitch results") %>% kable_paper("hover", full_width = F)
+
+lacmres3p <- do.call(rbind,lapply(groupsizes, function (g) { lacmres2[params$groupsizes==g,"PREC"] }))
+colnames(lacmres3p) <- deltas
+rownames(lacmres3p) <- groupsizes
+lacmres3p %>% kbl(caption="LA mitch precision") %>% kable_paper("hover", full_width = F)
+
+lacmres3r <- do.call(rbind,lapply(groupsizes, function (g) { lacmres2[params$groupsizes==g,"REC"] }))
+colnames(lacmres3r) <- deltas
+rownames(lacmres3r) <- groupsizes
+lacmres3r %>% kbl(caption="LA mitch recall") %>% kable_paper("hover", full_width = F)
+
+F1(lacmres3p,lacmres3r) %>% kbl(caption="LA mitch F1") %>% kable_paper("hover", full_width = F)
+
+```
+
 ## AL sim

### 2023-10-27 figs/GSE158422_simulate020.Rmd
progress on simulation results

--- 2023-10-27 figs/GSE158422_simulate050.Rmd
@@
 
-lengths <- rep(50,1000)
+lengths <- rep(20,1000)
 
@@
   FN=UPFN+DNFN
-  TN=nrow(lares1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
+  TN=nrow(lamres1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
   PREC=TP/(TP+FP)
@@
 
+```
+
+## AA Aggregate-aggregate-limma functions
+
+Use mean value works well here.
+
+```{r,aafunc}
+
+gsagg <- function(x,genesets,cores=1) {
+  meds <- mclapply(1:length(genesets), function(i) {
+    gs = genesets[[i]]
+    xx <- x[rownames(x) %in% gs,]
+    med <- apply(xx,2,mean)
+  },mc.cores=cores)
+  mymed <- do.call(rbind,meds)
+  rownames(mymed) <- names(genesets)
+  as.data.frame(mymed)
+}
+
+aalimma <- function(agag,design) {
+  fit.reduced <- lmFit(agag,design)
+  fit.reduced <- eBayes(fit.reduced)
+  dmagg <- topTable(fit.reduced,coef=ncol(design), number = Inf)
+  return(dmagg)
+}
+
+aal <- function(mval,myann,genesets,design,cores=1) {
+  medf <- chragg(mval,myann,cores=cores)
+  agag <- gsagg(x=medf,genesets=genesets,cores=cores)
+  aalres <- aalimma(agag=agag,design=design)
+  return(aalres)
+}
+
+simaa <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
+  # generate gene sets
+  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
+  lengths <- unname(unlist(lapply(genesetdatabase,length)))
+  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
+  # select gene sets to alter
+  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
+  # incorporate select changes
+  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
+    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
+  # set up limma
+  mval2 <- sim$mval
+  ncols <- ncol(mval2)
+  groupsize <- ncols/2
+  ss <- data.frame(colnames(mval2))
+  colnames(ss) <- "sample"
+  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
+  d <- model.matrix(~ ss$case )
+  # aa pipeline
+  aares1 <- aal(mval=mval2, myann=myann, genesets=gsets, design=d, cores=4)
+  # summarise results
+  gsig_up3 <- rownames(subset(aares1, adj.P.Val < 0.05 & logFC > 0))
+  gsig_dn3 <- rownames(subset(aares1, adj.P.Val < 0.05 & logFC < 0))
+  gtup <- names(sim[[6]])
+  gtdn <- names(sim[[7]])
+  UPTP=length(intersect(gsig_up3 ,gtup))
+  UPFP=length(setdiff(gsig_up3 ,gtup))
+  UPFN=length(setdiff(gtup,gsig_up3))
+  DNTP=length(intersect(gsig_dn3 ,gtdn))
+  DNFP=length(setdiff(gsig_dn3 ,gtdn))
+  DNFN=length(setdiff(gtdn,gsig_dn3))
+  TP=UPTP+DNTP
+  FP=UPFP+DNFP
+  FN=UPFN+DNFN
+  TN=nrow(aares1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
+  PREC=TP/(TP+FP)
+  REC=TP/(TP+FN)
+  F1=TP/(TP+(0.5*(FP+FN)))
+  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
+  return(result)
+}
+
 F1 <- function(x,y) {
@@
 num_dm_sets=50
-sims=4
-
-#groupsizes=c(3,5,8,12)
-#deltas=c(0.1,0.2,0.3,0.4,0.5)
-groupsizes=5
-deltas=0.4
+sims=10
+
+groupsizes=c(3,5,8,12)
+deltas=c(0.1,0.2,0.3,0.4,0.5)
+#groupsizes=5
+#deltas=0.4
 
@@
 
-simlacm
-
 lacmres <- lapply(1:nrow(params) , function(j) {
@@
 
+## AA sim
+
+```{r,runaa}
+
+aares <- lapply(1:nrow(params) , function(j) {
+  groupsize = params[j,1]
+  delta = params[j,2]
+  res <- mclapply(1:sims,function(i) {
+    simaa(genesetdatabase=gsets, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
+      groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
+  },mc.cores=4)
+  res <- do.call(rbind,res)
+  return(res)
+})
+
+aares
+
+aares2 <- do.call(rbind,lapply(aares,colMeans))
+
+aares2
+
+aares2 %>% kbl(caption="AA results") %>% kable_paper("hover", full_width = F)
+
+aares3p <- do.call(rbind,lapply(groupsizes, function (g) { aares2[params$groupsizes==g,"PREC"] }))
+colnames(aares3p) <- deltas
+rownames(aares3p) <- groupsizes
+aares3p %>% kbl(caption="AA precision") %>% kable_paper("hover", full_width = F)
+
+aares3r <- do.call(rbind,lapply(groupsizes, function (g) { aares2[params$groupsizes==g,"REC"] }))
+colnames(aares3r) <- deltas
+rownames(aares3r) <- groupsizes
+aares3r %>% kbl(caption="AA recall") %>% kable_paper("hover", full_width = F)
+
+F1(aares3p,almres3r) %>% kbl(caption="AA F1") %>% kable_paper("hover", full_width = F)
+
+```
+
 ## Session information

### 2023-10-29 manuscript/manuscript.Rmd
writing

@@
 ---
-title: "Functional class scoring-based enrichment tests of Infinium DNA methylation data"
+title: "Finally, functional class scoring enrichment analysis of Infinium DNA methylation data"
 date: "`r Sys.Date()`"
@@
 
-* Infinium arrays remain the preferred platform for genomic methylation profiling due to relatively
-low cost and analytical simplicity as compared to high throughput methylation sequencing.
+* Infinium arrays remain the preferred platform for EWASs due to low cost and analytical simplicity
+as compared to high throughput methylation sequencing.
 
-* Infinium arrays have multiple probes per gene in different locations including promoters, CpG islands,
-gene bodies and enhancers, which complicates downstream functional enrichment analysis, otherwise known
-as pathway analysis or gene set enrichment analysis.
+* Infinium arrays typically have multiple probes per gene in different locations including promoters,
+CpG islands, gene bodies and enhancers, which complicates downstream functional enrichment analysis,
+otherwise known as pathway analysis or gene set enrichment analysis.
 
@@
 class scoring (FCS).
-In ORA, genes with measurements that meet an arbitrary significance threshold are selected and compared
+
+In ORA, genes with probes that meet an arbitrary significance threshold are selected and compared
 to a background list consisting of genes measured in the assay.
@@
 gene list of interest relative to the background.
+
 Functional class scoring takes a different approach by ranking all detected genes by a differential
@@
 
-* Here, we evaluate different approaches to apply FCS to Infinium array.
-These methods vary in the sequence of data aggregation and differential testing steps which are
-required to go from probe-level methylation measurements to evidence of pathway-level differential
-methylation.
+* Here, we systematically evaluate different approaches to apply FCS to Infinium array.
+
+* Due to the fact that genes have many probes, there are many concevable ways that FCS could be applied
+to Infinium Methylation data.
+
+* Method group 1: Limma is performed on the probe level and the probes beloning to each gene are scored
+and these scores are aggregated to the gene level.
+The gene scores undergo an enrichment test.
+
+* Method group 2: The probe methylation values beloning to each gene are aggregated before differential
+methylation analysis with limma.
+Then, differential methylation results for each gene are scored, followed by an enrichment test.
+
+With the above, there are options for how the aggregation steps are done, whether parametric or
+non-parametric tests are used and whether the enrichment tests are self-contained or competitive.
+
+To determine the best approach to applying FCS to Infinium methylation data, in this study we use
+simulated data.
+The proposed FCS methods are also compared against existing ORA methods.
 

### 2023-10-30 manuscript/manuscript.Rmd
adding simulation to the ms

@@
 This is made more complicated for Infinium array data due to the presence of multiple probes per gene,
-meaning the data needs to be summarised from probes to genes to gene sets.
+meaning the data needs to be summarised from probes to genes and then to gene sets.
+Here, we outline seven potential approaches for FCS of Infinium methylation data:
 
-We propose three different approaches for applying enrichment analysis depending on when the differential
-methylation step is conducted (**Figure 1**).
+1. Limma Average t-test (LAT).
+Differential methylation analysis is conducted at the probe level with limma, and the limma
+t-statistics for each gene are summarised (mean).
+The mean t-statistics are used in a downstream two-sample two-way t-test of gene set enrichment.
 
-1. <strong>Differential methylation analysis is conducted on probes and these results are aggregated
-to the gene and gene set level (called "LA").<.strong>
-In this approach differential analysis was conduced on probe values using limma.
-Differential probe methylation results (t-statistics) were then aggregated to the gene level by median.
-These median t-statistics were subject to downstream enrichment analysis using a one-sample two-sided
-t-test (a self-contained test) to estimate whether the t-statistic distribution of the gene set varies
-from zero.
-The use of a self-contained test is warranted as global hyper- or hypo-methylation is common between
-conditions, where competitive tests may give misleading results.
-The use of parametric tests is also warranted, as probe t-statistics typically exhibit a normal
-distribution, even in the presence of global hyper- or hypo-methylation.
+2. Limma Top t-test (LTT).
+As above, except instead of calculating the mean t-statistic, the probe with the largest magnitude is selected
+to represent the gene.
 
-2. <strong>Differential methylation analysis conducted after probe-level methylation is aggregated to
-gene-level (called "AL").</strong>
-As each gene may have many probes, these individual data points were summarised into a single
-value for each individual.
-We used the mean of M-values for each probe belonging to a gene, which generated a new gene-level
-M-value matrix.
-This gene-level methylation matrix underwent differential analysis using limma, followed by an
-enrichment test as above (one-sample two-tailed t-test).
+3. Limma Average Wilcox (LAW).
+Similar to LAT, except instead of the two-sided t-test of gene set enrichment, a non-parametric alternative,
+the Wilcoxon signed-rank test.
 
-3. <strong>Methylation values are aggregated to the gene set level for differential testing
-(called "AA").</strong>
-M-values for each probe were aggregated to the gene level by mean, and then aggregated again to the gene
-set level by mean or median.
-The resulting matrix of aggregated M-values underwent differential analysis with limma and the results
-including log fold changes and significance values were used to infer pathway regulation.
-This approach has a caveat in that it may partly violate the assumption of independent sampling, as
-it is known that some genes are represented in multiple gene sets.
+4. Limma Average Mitch (LAM).
+Similar to LAW, however instead of the Wilcoxon test, the Mitch package is used to execute a two-way,
+two-sample ANOVA-on-ranks test (Kruskal Wallis).
+
+5. Aggregate Limma t-test (ALT).
+In this method, all probe measurements beloning to each gene are averaged before conducting differential
+methylation andlysis with limma.
+The limma t-statistics are then used downstream for enrichment analysis with the two-sample two-way t-test.
+
+6. Aggregate Limma Wilcox (ALW).
+Similar to ALT, however it uses the Wilcoxon non-parametric test for gene set enrichment.
+
+7. Aggregate Limma Mitch (ALM).
+Similar to ALW, however uses the Mitch package to execute a two-way, two-sample ANOVA-on-ranks test
+(Kruskal Wallis).
+
+The above were compared to a standard ORA based approach (GSA), which involves limma on probes, selection of
+statistically significant probes (FDR<0.05) for ORA with the `GSAmeth()` function of missMethyl which conducts a
+modified hypergeometric test that accounts for multiple probe biases.
+In the case that fewer than 250 significant probes are identified, then 250 probes with the smallest p-values
+were selected.
 
 ### Method validation
+
+To assess these methods, we adopted a simulated data approach.
+EPIC Infinium array data derived from non-cancerous lung tissue of 37 cancer patients was used for our
+evaluations.
+We randomly sampled datasets to serve as control and cases (no replacement), with group sizes varying between
+3 and 12.
+One thousand random gene sets were created with sizes varying between 20 and 100 members.
+Randow gene sets were used to avoid problems caused by the large overlap among real gene sets.
+Throughout our evaluations, 50 gene sets were selected to be differentially methylated, 25 with increased and 25
+with decreased methylation.
+From these gene sets, half of the members were selected.
+For those selected genes, half of the annotated probes were selected.
+We adjusted the m-values in the case group by a specified amount, which we call the "delta", which we varied
+between 0.1 and 0.4.
+Following the incorporation of methylation changes to selected probes in the case group, the data underwent
+enrichment analysis using `GSAmeth` approach, or one of the seven FCS methods outlined above.
+Enrichment results were then filtered for FDR<0.05 and correct direction of methylation change.
+Differentially methylated gene sets observed were compared to gene sets selected to be the ground truth, so that
+the number of true positives (TP), false positives (FP) and false negatives (FN) could be determined.
+At each setting for group size, delta and gene set size, 20 replications were conducted each with a set seed
+to ensure reproducibility.
+Mean TP, FP and FN values were determined, and then precision, recall and F1 scores were calculated.
 

### 2023-11-01 figs/GSE158422_simulate050.Rmd
bugfix
simulation results good

@@
 
+## AA sim
+
+```{r,runaa}
+
+aares <- lapply(1:nrow(params) , function(j) {
+  groupsize = params[j,1]
+  delta = params[j,2]
+  res <- mclapply(1:sims,function(i) {
+    simaa(genesetdatabase=gsets, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
+      groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
+  },mc.cores=4)
+  res <- do.call(rbind,res)
+  return(res)
+})
+
+aares2 <- do.call(rbind,lapply(aares,colMeans))
+
+aares2
+
+aares2 %>% kbl(caption="AA results") %>% kable_paper("hover", full_width = F)
+
+aares3p <- do.call(rbind,lapply(groupsizes, function (g) { aares2[params$groupsizes==g,"PREC"] }))
+colnames(aares3p) <- deltas
+rownames(aares3p) <- groupsizes
+aares3p %>% kbl(caption="AA precision") %>% kable_paper("hover", full_width = F)
+
+aares3r <- do.call(rbind,lapply(groupsizes, function (g) { aares2[params$groupsizes==g,"REC"] }))
+colnames(aares3r) <- deltas
+rownames(aares3r) <- groupsizes
+aares3r %>% kbl(caption="AA recall") %>% kable_paper("hover", full_width = F)
+
+F1(aares3p,almres3r) %>% kbl(caption="AA F1") %>% kable_paper("hover", full_width = F)
+
+```
+
 ## Session information
@@
 
-save.image("GSE158422_simulate.Rdata")
+save.image("GSE158422_simulate050.Rdata")
 

### 2023-11-01 manuscript/manuscript.Rmd
ms

@@
 
+Simulations were conducted using the seven FCS methods (LAT, LTT, LAW, LAM, ALT, ALW, ALM) and GSAmeth,
+with group sizes of 3-12 and delta values between 0.1 and 0.5, and random gene sets with sizes 20, 50
+and 100.
+Precision and recall were calculated for these tests (Figure 1).
+As expected, precision and recall values were better when group sizes and delta values were larger.
+
+These data showed a strong relationship between the size of the gene sets and the precision and recall,
+regardless of the enrichment test used.
+On the whole, precision and recall were better
+
 We applied the three distinct approaches ("LA", "AL" and "AA") for pathway level differential

### 2023-11-02 figs/GSE158422_simulate020.Rmd
simulation results good
updates on data presentation

@@
 gres2 <- do.call(rbind,lapply(gres,colMeans))
+gres2[,"PREC"] <- gres2[,"TP"] / ( gres2[,"TP"] + gres2[,"FP"] )
 
@@
 lacres2 <- do.call(rbind,lapply(lacres,colMeans))
+lacres2[,"PREC"] <- lacres2[,"TP"] / ( lacres2[,"TP"] + lacres2[,"FP"] )
 
@@
 lactopres2 <- do.call(rbind,lapply(lactopres,colMeans))
+lactopres2[,"PREC"] <- lactopres2[,"TP"] / ( lactopres2[,"TP"] + lactopres2[,"FP"] )
 
@@
 nlacres2 <- do.call(rbind,lapply(nlacres,colMeans))
+nlacres2[,"PREC"] <- nlacres2[,"TP"] / ( nlacres2[,"TP"] + nlacres2[,"FP"] )
 
@@
 lacmres2 <- do.call(rbind,lapply(lacmres,colMeans))
+lacmres2[,"PREC"] <- lacmres2[,"TP"] / ( lacmres2[,"TP"] + lacmres2[,"FP"] )
 
@@
 alcres2 <- do.call(rbind,lapply(alcres,colMeans))
+alcres2[,"PREC"] <- alcres2[,"TP"] / ( alcres2[,"TP"] + alcres2[,"FP"] )
 
@@
 nalcres2 <- do.call(rbind,lapply(nalcres,colMeans))
+nalcres2[,"PREC"] <- nalcres2[,"TP"] / ( nalcres2[,"TP"] + nalcres2[,"FP"] )
 
@@
 almres2 <- do.call(rbind,lapply(almres,colMeans))
+almres2[,"PREC"] <- almres2[,"TP"] / ( almres2[,"TP"] + almres2[,"FP"] )
 
@@
 
-aares
-
 aares2 <- do.call(rbind,lapply(aares,colMeans))
-
-aares2
+aares2[,"PREC"] <- aares2[,"TP"] / ( aares2[,"TP"] + aares2[,"FP"] )
 
@@
 
-save.image("GSE158422_simulate.Rdata")
+save.image("GSE158422_simulate020.Rdata")
 

### 2023-11-02 figs/GSE158422_simulate050.Rmd
updates on data presentation

@@
 aares2 <- do.call(rbind,lapply(aares,colMeans))
-
-aares2
+aares2[,"PREC"] <- aares2[,"TP"] / ( aares2[,"TP"] + aares2[,"FP"] )
 

### 2023-11-02 figs/GSE158422_simulate100.Rmd
progress on simulation results
simulation results good
updates on data presentation

--- 2023-10-27 figs/GSE158422_simulate020.Rmd
@@
 
-lengths <- rep(20,1000)
+lengths <- rep(100,1000)
 
@@
 
-aares
-
 aares2 <- do.call(rbind,lapply(aares,colMeans))
@@
 
-save.image("GSE158422_simulate.Rdata")
+save.image("GSE158422_simulate100.Rdata")
 

### 2023-11-02 figs/GSE158422_simulate_summary.Rmd
summary of simulation results good

---
title: "Testing gene methylation enrichment analysis approaches using simulated data - summarise data"
author: "The GMEA team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
---

## Introduction

In this report we establish a new method for generating simulated data with known ground truth.
This will be used to test different gene methylation enrichment approaches systematically.

The general steps are:

1. Import GSE158422 data corresponding to control (non-tumour tissue).

2. From the 37 samples, create two groups of 18 samples.
One of these will be considered "control" and the other "case".

3. Create random gene sets that have similar sized to Reactome pathways.

3. Some gene sets will be selected to be differentially methylated.
Half of these will be hypermethylated and the others will be hypomethylated.

4. The changes will be incorporated into the "case" samples.

5. The enrichment analysis will be conducted.

6. The accuracy will be calculated.

```{r,packages}

suppressPackageStartupMessages({
  library("stringi")
  library("limma")
  library("missMethyl")
  library("IlluminaHumanMethylation450kmanifest")
  library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
  library('org.Hs.eg.db')
  library("psych")
  library("mitch")
  library("kableExtra")
})

```

## Functions

```{r,functions}

F1 <- function(x,y) {
  ( 2 * x * y ) / ( x + y )
}

```

# Geneset size = 100

```{r,gsa}

load("GSE158422_simulate100.Rdata")

par(mfrow=c(2,3))

gres2[,"PREC"] <- gres2[,"TP"] / ( gres2[,"TP"] + gres2[,"FP"] )
gres3p <- do.call(rbind,lapply(groupsizes, function (g) { gres2[params$groupsizes==g,"PREC"] }))
colnames(gres3p) <- deltas
rownames(gres3p) <- groupsizes
gres3p %>% kbl(caption="GSAmeth precision") %>% kable_paper("hover", full_width = F)

gres3r <- do.call(rbind,lapply(groupsizes, function (g) { gres2[params$groupsizes==g,"REC"] }))
colnames(gres3r) <- deltas
rownames(gres3r) <- groupsizes
gres3r %>% kbl(caption="GSAmeth recall") %>% kable_paper("hover", full_width = F)

gres3f <- F1(gres3p,gres3r)
gres3f %>% kbl(caption="GSAmeth F1") %>% kable_paper("hover", full_width = F)

plot(rownames(gres3p),gres3p[,"0.5"],type="b",las=1,ylab="precision",xlab="",
  main="",ylim=c(0,1))
lines(rownames(gres3p),gres3p[,"0.4"],type="b",col="red")
lines(rownames(gres3p),gres3p[,"0.3"],type="b",col="blue")
lines(rownames(gres3p),gres3p[,"0.2"],type="b",col="darkgreen")
grid()

legend("bottomright", legend=c("0.5", "0.4", "0.3", "0.2"),title="delta",
  lty=1, lwd=2, col=c("black", "red", "blue", "darkgreen"), cex=1)

plot(rownames(gres3r),gres3r[,"0.5"],type="b",las=1,ylab="recall",xlab="group size",
  main="GSA",ylim=c(0,1))
lines(rownames(gres3r),gres3r[,"0.4"],type="b",col="red")
lines(rownames(gres3r),gres3r[,"0.3"],type="b",col="blue")
lines(rownames(gres3r),gres3r[,"0.2"],type="b",col="darkgreen")
grid()

plot(rownames(gres3r),gres3f[,"0.5"],type="b",las=1,ylab="F1",xlab="",
  main="",ylim=c(0,1))
lines(rownames(gres3r),gres3f[,"0.4"],type="b",col="red")
lines(rownames(gres3r),gres3f[,"0.3"],type="b",col="blue")
lines(rownames(gres3r),gres3f[,"0.2"],type="b",col="darkgreen")
grid()

lacmres2[,"PREC"] <- lacmres2[,"TP"] / ( lacmres2[,"TP"] + lacmres2[,"FP"] )
lacmres3p <- do.call(rbind,lapply(groupsizes, function (g) { lacmres2[params$groupsizes==g,"PREC"] }))
colnames(lacmres3p) <- deltas
rownames(lacmres3p) <- groupsizes
lacmres3p %>% kbl(caption="LAM precision") %>% kable_paper("hover", full_width = F)

lacmres3r <- do.call(rbind,lapply(groupsizes, function (g) { lacmres2[params$groupsizes==g,"REC"] }))
colnames(lacmres3r) <- deltas
rownames(lacmres3r) <- groupsizes
lacmres3r %>% kbl(caption="LAM recall") %>% kable_paper("hover", full_width = F)

lacmres3f <- F1(lacmres3p,lacmres3r)
lacmres3f %>% kbl(caption="LAM F1") %>% kable_paper("hover", full_width = F)

plot(rownames(lacmres3p),lacmres3p[,"0.5"],type="b",las=1,ylab="precision",xlab="",
  main="",ylim=c(0,1))
lines(rownames(lacmres3p),lacmres3p[,"0.4"],type="b",col="red")
lines(rownames(lacmres3p),lacmres3p[,"0.3"],type="b",col="blue")
lines(rownames(lacmres3p),lacmres3p[,"0.2"],type="b",col="darkgreen")
grid()

legend("bottomright", legend=c("0.5", "0.4", "0.3", "0.2"),title="delta",
  lty=1, lwd=2, col=c("black", "red", "blue", "darkgreen"), cex=1)

plot(rownames(lacmres3r),lacmres3r[,"0.5"],type="b",las=1,ylab="recall",xlab="group size",
  main="LAM",ylim=c(0,1))
lines(rownames(lacmres3r),lacmres3r[,"0.4"],type="b",col="red")
lines(rownames(lacmres3r),lacmres3r[,"0.3"],type="b",col="blue")
lines(rownames(lacmres3r),lacmres3r[,"0.2"],type="b",,col="darkgreen")
grid()

plot(rownames(lacmres3r),lacmres3f[,"0.5"],type="b",las=1,ylab="F1",xlab="",
  main="",ylim=c(0,1))
lines(rownames(lacmres3r),lacmres3f[,"0.4"],type="b",col="red")
lines(rownames(lacmres3r),lacmres3f[,"0.3"],type="b",col="blue")
lines(rownames(lacmres3r),lacmres3f[,"0.2"],type="b",col="darkgreen")
grid()

dev.off()

```

## Session information

```{r,save}

sessionInfo()

```

## Notes

LA

* simla: parametric self-contained

* simlac: parametric competitive

* simnla: nonparametric self-contained

* simnlac: nonparametric competitive

AL

* simal: parametric self-contained

* simalc: parametric competitive

* simnal: nonparametric self-contained

* simnalc: nonparametric competitive

### 2023-11-03 figs/GSE158422_simulate_summary.Rmd
fig1

@@
 
-In this report we establish a new method for generating simulated data with known ground truth.
-This will be used to test different gene methylation enrichment approaches systematically.
+Here is an explanation of the different objects:
 
-The general steps are:
+0. GSA: gres
 
-1. Import GSE158422 data corresponding to control (non-tumour tissue).
+1. LAT: lacres
 
-2. From the 37 samples, create two groups of 18 samples.
-One of these will be considered "control" and the other "case".
+2. LTT: lactopres
 
-3. Create random gene sets that have similar sized to Reactome pathways.
+3. LAW: nlacres
 
-3. Some gene sets will be selected to be differentially methylated.
-Half of these will be hypermethylated and the others will be hypomethylated.
+4. LAM: lacmres
 
-4. The changes will be incorporated into the "case" samples.
+5. ALT: alcres
 
-5. The enrichment analysis will be conducted.
+6. ALW: nalcres
 
-6. The accuracy will be calculated.
+7. ALM: almres
 
@@
 
+summarise <- function(x) {
+  PREC <- sum(x[,"TP"]) / ( sum(x[,"TP"]) + sum(x[,"FP"]) )
+  REC <- sum(x[,"TP"]) / ( sum(x[,"TP"]) + sum(x[,"FN"]) )
+  F <- F1(PREC,REC)
+  return(c("PRECISION"=PREC,"RECALL"=REC,"F1"=F))
+}
+
 ```
 
-# Geneset size = 100
+## Overall summary
 
-```{r,gsa}
+Make a set of three barplots which summarise method performance over the range of group size and
+delta values, for gene set sizes of 20, 50 and 200.
+
+```{r,summary,fig.width=10,fig.height=4}
+
+par(mfrow=c(1,3))
+
+load("GSE158422_simulate020.Rdata")
+
+l <- list("GSA"=gres2,"LAT"=lacres2,"LTT"=lactopres2,"LAW"=nlacres2,
+  "LAM"=lacmres2,"ALT"=alcres2,"ALW"=nalcres2,"ALM"=almres2)
+
+x <- do.call(rbind,lapply(l,summarise))
+
+barplot(t(x),beside=TRUE,ylim=c(0,1.2),ylab="index", legend = colnames(x) ,main="20 genes per set")
+grid()
+
+load("GSE158422_simulate050.Rdata")
+
+l <- list("GSA"=gres2,"LAT"=lacres2,"LTT"=lactopres2,"LAW"=nlacres2,
+  "LAM"=lacmres2,"ALT"=alcres2,"ALW"=nalcres2,"ALM"=almres2)
+
+x <- do.call(rbind,lapply(l,summarise))
+
+par(mfrow=c(1,1))
+
+barplot(t(x),beside=TRUE,ylim=c(0,1.2),ylab="index", legend = colnames(x) ,main="50 genes per set")
+grid()
 
 load("GSE158422_simulate100.Rdata")
+
+l <- list("GSA"=gres2,"LAT"=lacres2,"LTT"=lactopres2,"LAW"=nlacres2,
+  "LAM"=lacmres2,"ALT"=alcres2,"ALW"=nalcres2,"ALM"=almres2)
+
+x <- do.call(rbind,lapply(l,summarise))
+
+par(mfrow=c(1,1))
+
+barplot(t(x),beside=TRUE,ylim=c(0,1.2),ylab="index", legend = colnames(x) ,main="100 genes per set")
+grid()
+
+```
+
+## In depth analysis of GSA vs LAM
+
+As GSA appears to be the best-performing existing method for enrichment analysis, it is a good idea
+to compare it to LAM which is the best one out of the new methods evaluated here.
+The following charts show the performance of GSA and LAM including precision and recall over the
+range of group sizes and delta values when the sets contain 20, 50 and 100 genes each.
+
+```{r,020_gsa_vs_lam}
+
+# 20
+load("GSE158422_simulate020.Rdata")
 
@@
 
-dev.off()
+# 50
+load("GSE158422_simulate050.Rdata")
+
+gres2[,"PREC"] <- gres2[,"TP"] / ( gres2[,"TP"] + gres2[,"FP"] )
+gres3p <- do.call(rbind,lapply(groupsizes, function (g) { gres2[params$groupsizes==g,"PREC"] }))
+colnames(gres3p) <- deltas
+rownames(gres3p) <- groupsizes
+gres3p %>% kbl(caption="GSAmeth precision") %>% kable_paper("hover", full_width = F)
+
+gres3r <- do.call(rbind,lapply(groupsizes, function (g) { gres2[params$groupsizes==g,"REC"] }))
+colnames(gres3r) <- deltas
+rownames(gres3r) <- groupsizes
+gres3r %>% kbl(caption="GSAmeth recall") %>% kable_paper("hover", full_width = F)
+
+gres3f <- F1(gres3p,gres3r)
+gres3f %>% kbl(caption="GSAmeth F1") %>% kable_paper("hover", full_width = F)
+
+plot(rownames(gres3p),gres3p[,"0.5"],type="b",las=1,ylab="precision",xlab="",
+  main="",ylim=c(0,1))
+lines(rownames(gres3p),gres3p[,"0.4"],type="b",col="red")
+lines(rownames(gres3p),gres3p[,"0.3"],type="b",col="blue")
+lines(rownames(gres3p),gres3p[,"0.2"],type="b",col="darkgreen")
+grid()
+
+legend("bottomright", legend=c("0.5", "0.4", "0.3", "0.2"),title="delta",
+  lty=1, lwd=2, col=c("black", "red", "blue", "darkgreen"), cex=1)
+
+plot(rownames(gres3r),gres3r[,"0.5"],type="b",las=1,ylab="recall",xlab="group size",
+  main="GSA",ylim=c(0,1))
+lines(rownames(gres3r),gres3r[,"0.4"],type="b",col="red")
+lines(rownames(gres3r),gres3r[,"0.3"],type="b",col="blue")
+lines(rownames(gres3r),gres3r[,"0.2"],type="b",col="darkgreen")
+grid()
+
+plot(rownames(gres3r),gres3f[,"0.5"],type="b",las=1,ylab="F1",xlab="",
+  main="",ylim=c(0,1))
+lines(rownames(gres3r),gres3f[,"0.4"],type="b",col="red")
+lines(rownames(gres3r),gres3f[,"0.3"],type="b",col="blue")
+lines(rownames(gres3r),gres3f[,"0.2"],type="b",col="darkgreen")
+grid()
+
+lacmres2[,"PREC"] <- lacmres2[,"TP"] / ( lacmres2[,"TP"] + lacmres2[,"FP"] )
+lacmres3p <- do.call(rbind,lapply(groupsizes, function (g) { lacmres2[params$groupsizes==g,"PREC"] }))
+colnames(lacmres3p) <- deltas
+rownames(lacmres3p) <- groupsizes
+lacmres3p %>% kbl(caption="LAM precision") %>% kable_paper("hover", full_width = F)
+
+lacmres3r <- do.call(rbind,lapply(groupsizes, function (g) { lacmres2[params$groupsizes==g,"REC"] }))
+colnames(lacmres3r) <- deltas
+rownames(lacmres3r) <- groupsizes
+lacmres3r %>% kbl(caption="LAM recall") %>% kable_paper("hover", full_width = F)
+
+lacmres3f <- F1(lacmres3p,lacmres3r)
+lacmres3f %>% kbl(caption="LAM F1") %>% kable_paper("hover", full_width = F)
+
+plot(rownames(lacmres3p),lacmres3p[,"0.5"],type="b",las=1,ylab="precision",xlab="",
+  main="",ylim=c(0,1))
+lines(rownames(lacmres3p),lacmres3p[,"0.4"],type="b",col="red")
+lines(rownames(lacmres3p),lacmres3p[,"0.3"],type="b",col="blue")
+lines(rownames(lacmres3p),lacmres3p[,"0.2"],type="b",col="darkgreen")
+grid()
+
+legend("bottomright", legend=c("0.5", "0.4", "0.3", "0.2"),title="delta",
+  lty=1, lwd=2, col=c("black", "red", "blue", "darkgreen"), cex=1)
+
+plot(rownames(lacmres3r),lacmres3r[,"0.5"],type="b",las=1,ylab="recall",xlab="group size",
+  main="LAM",ylim=c(0,1))
+lines(rownames(lacmres3r),lacmres3r[,"0.4"],type="b",col="red")
+lines(rownames(lacmres3r),lacmres3r[,"0.3"],type="b",col="blue")
+lines(rownames(lacmres3r),lacmres3r[,"0.2"],type="b",,col="darkgreen")
+grid()
+
+plot(rownames(lacmres3r),lacmres3f[,"0.5"],type="b",las=1,ylab="F1",xlab="",
+  main="",ylim=c(0,1))
+lines(rownames(lacmres3r),lacmres3f[,"0.4"],type="b",col="red")
+lines(rownames(lacmres3r),lacmres3f[,"0.3"],type="b",col="blue")
+lines(rownames(lacmres3r),lacmres3f[,"0.2"],type="b",col="darkgreen")
+grid()
+
+# 100
+load("GSE158422_simulate100.Rdata")
+
+gres2[,"PREC"] <- gres2[,"TP"] / ( gres2[,"TP"] + gres2[,"FP"] )
+gres3p <- do.call(rbind,lapply(groupsizes, function (g) { gres2[params$groupsizes==g,"PREC"] }))
+colnames(gres3p) <- deltas
+rownames(gres3p) <- groupsizes
+gres3p %>% kbl(caption="GSAmeth precision") %>% kable_paper("hover", full_width = F)
+
+gres3r <- do.call(rbind,lapply(groupsizes, function (g) { gres2[params$groupsizes==g,"REC"] }))
+colnames(gres3r) <- deltas
+rownames(gres3r) <- groupsizes
+gres3r %>% kbl(caption="GSAmeth recall") %>% kable_paper("hover", full_width = F)
+
+gres3f <- F1(gres3p,gres3r)
+gres3f %>% kbl(caption="GSAmeth F1") %>% kable_paper("hover", full_width = F)
+
+plot(rownames(gres3p),gres3p[,"0.5"],type="b",las=1,ylab="precision",xlab="",
+  main="",ylim=c(0,1))
+lines(rownames(gres3p),gres3p[,"0.4"],type="b",col="red")
+lines(rownames(gres3p),gres3p[,"0.3"],type="b",col="blue")
+lines(rownames(gres3p),gres3p[,"0.2"],type="b",col="darkgreen")
+grid()
+
+legend("bottomright", legend=c("0.5", "0.4", "0.3", "0.2"),title="delta",
+  lty=1, lwd=2, col=c("black", "red", "blue", "darkgreen"), cex=1)
+
+plot(rownames(gres3r),gres3r[,"0.5"],type="b",las=1,ylab="recall",xlab="group size",
+  main="GSA",ylim=c(0,1))
+lines(rownames(gres3r),gres3r[,"0.4"],type="b",col="red")
+lines(rownames(gres3r),gres3r[,"0.3"],type="b",col="blue")
+lines(rownames(gres3r),gres3r[,"0.2"],type="b",col="darkgreen")
+grid()
+
+plot(rownames(gres3r),gres3f[,"0.5"],type="b",las=1,ylab="F1",xlab="",
+  main="",ylim=c(0,1))
+lines(rownames(gres3r),gres3f[,"0.4"],type="b",col="red")
+lines(rownames(gres3r),gres3f[,"0.3"],type="b",col="blue")
+lines(rownames(gres3r),gres3f[,"0.2"],type="b",col="darkgreen")
+grid()
+
+lacmres2[,"PREC"] <- lacmres2[,"TP"] / ( lacmres2[,"TP"] + lacmres2[,"FP"] )
+lacmres3p <- do.call(rbind,lapply(groupsizes, function (g) { lacmres2[params$groupsizes==g,"PREC"] }))
+colnames(lacmres3p) <- deltas
+rownames(lacmres3p) <- groupsizes
+lacmres3p %>% kbl(caption="LAM precision") %>% kable_paper("hover", full_width = F)
+
+lacmres3r <- do.call(rbind,lapply(groupsizes, function (g) { lacmres2[params$groupsizes==g,"REC"] }))
+colnames(lacmres3r) <- deltas
+rownames(lacmres3r) <- groupsizes
+lacmres3r %>% kbl(caption="LAM recall") %>% kable_paper("hover", full_width = F)
+
+lacmres3f <- F1(lacmres3p,lacmres3r)
+lacmres3f %>% kbl(caption="LAM F1") %>% kable_paper("hover", full_width = F)
+
+plot(rownames(lacmres3p),lacmres3p[,"0.5"],type="b",las=1,ylab="precision",xlab="",
+  main="",ylim=c(0,1))
+lines(rownames(lacmres3p),lacmres3p[,"0.4"],type="b",col="red")
+lines(rownames(lacmres3p),lacmres3p[,"0.3"],type="b",col="blue")
+lines(rownames(lacmres3p),lacmres3p[,"0.2"],type="b",col="darkgreen")
+grid()
+
+legend("bottomright", legend=c("0.5", "0.4", "0.3", "0.2"),title="delta",
+  lty=1, lwd=2, col=c("black", "red", "blue", "darkgreen"), cex=1)
+
+plot(rownames(lacmres3r),lacmres3r[,"0.5"],type="b",las=1,ylab="recall",xlab="group size",
+  main="LAM",ylim=c(0,1))
+lines(rownames(lacmres3r),lacmres3r[,"0.4"],type="b",col="red")
+lines(rownames(lacmres3r),lacmres3r[,"0.3"],type="b",col="blue")
+lines(rownames(lacmres3r),lacmres3r[,"0.2"],type="b",,col="darkgreen")
+grid()
+
+plot(rownames(lacmres3r),lacmres3f[,"0.5"],type="b",las=1,ylab="F1",xlab="",
+  main="",ylim=c(0,1))
+lines(rownames(lacmres3r),lacmres3f[,"0.4"],type="b",col="red")
+lines(rownames(lacmres3r),lacmres3f[,"0.3"],type="b",col="blue")
+lines(rownames(lacmres3r),lacmres3f[,"0.2"],type="b",col="darkgreen")
+grid()
+
+par(mfrow=c(1,1))
 

### 2023-11-03 figs/GSE158422_simulate_summary.Rmd
summary

@@
 
-```{r,summary,fig.width=10,fig.height=4}
+```{r,summary,fig.width=12,fig.height=3}
 
 par(mfrow=c(1,3))
+
+mar = c(5.1, 4.1, 5.1, 2.1)
 
@@
 
-barplot(t(x),beside=TRUE,ylim=c(0,1.2),ylab="index", legend = colnames(x) ,main="20 genes per set")
+cols <- c("white","gray","black")
+
+barplot(t(x),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, main="20 genes per set")
 grid()
+
+legend("topleft", inset=.1, c("Precision","Recall","F1"), fill=cols, cex=0.8)
 
@@
 
-par(mfrow=c(1,1))
-
-barplot(t(x),beside=TRUE,ylim=c(0,1.2),ylab="index", legend = colnames(x) ,main="50 genes per set")
+barplot(t(x),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, main="50 genes per set")
 grid()
@@
 
-par(mfrow=c(1,1))
-
-barplot(t(x),beside=TRUE,ylim=c(0,1.2),ylab="index", legend = colnames(x) ,main="100 genes per set")
+barplot(t(x),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, main="100 genes per set")
 grid()

### 2023-11-03 figs/GSE158422_simulate_summary.Rmd
results

@@
 
-```{r,summary,fig.width=12,fig.height=3}
-
-par(mfrow=c(1,3))
+```{r,summary,fig.width=5,fig.height=4}
 
@@
 
+x %>% kbl(caption="sim 20 res") %>% kable_paper("hover", full_width = F)
+
 cols <- c("white","gray","black")
 
-barplot(t(x),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, main="20 genes per set")
+barplot(t(x),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, cex.names=0.7, cex.axis=0.7, main="20 genes per set")
 grid()
 
-legend("topleft", inset=.1, c("Precision","Recall","F1"), fill=cols, cex=0.8)
+legend("topleft", inset=.1, c("Precision","Recall","F1"), fill=cols, cex=0.7)
 
@@
 
-barplot(t(x),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, main="50 genes per set")
+x %>% kbl(caption="sim 50 res") %>% kable_paper("hover", full_width = F)
+
+barplot(t(x),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, cex.names=0.7, cex.axis=0.7, main="50 genes per set")
 grid()
@@
 
-barplot(t(x),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, main="100 genes per set")
+x %>% kbl(caption="sim 100 res") %>% kable_paper("hover", full_width = F)
+
+barplot(t(x),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, cex.names=0.7, cex.axis=0.7, main="100 genes per set")
 grid()

### 2023-11-03 manuscript/manuscript.Rmd
results

@@
 One thousand random gene sets were created with sizes varying between 20 and 100 members.
-Randow gene sets were used to avoid problems caused by the large overlap among real gene sets.
+Random gene sets were used to avoid problems caused by the large overlap among real gene sets.
 Throughout our evaluations, 50 gene sets were selected to be differentially methylated, 25 with increased and 25
@@
 and 100.
-Precision and recall were calculated for these tests (Figure 1).
+True positives, false positives and false negatives were used to calculate overall precision and recall
+at these three gene set sizes (Figure 1).
+We found that overall recall was positively associated with the size of the gene sets considered.
+At 20 genes per set, GSA had the highest overall performance, with F1 of 0.25, followed by LAW with 0.06.
+At 50 genes per set, GSA and LAM achieved similar F1 values of ~0.39.
+At 100 genes per set, LAM achieved the highest F1 performance of 0.70 followed by LAT with 0.61, while
+GSA scored only 0.42.
+This result suggests that FCS methods perform better than ORA with gene set sizes >50.
+Performance results for GSA and LAM methods with 100 genes per set shows that recall was higher for most
+simulations, particularly with lower group sizes.
+Taken together, this result suggests that LAM offers improved sensitivity over existing approaches when
+the gene sets are larger than 50 members.
+
+ in Figure 2.
+This 
 As expected, precision and recall values were better when group sizes and delta values were larger.

### 2023-11-04 figs/GSE158422_simulate_summary.Rmd
fig1

@@
 
-l <- list("GSA"=gres2,"LAT"=lacres2,"LTT"=lactopres2,"LAW"=nlacres2,
+l1 <- list("GSA"=gres2,"LAT"=lacres2,"LTT"=lactopres2,"LAW"=nlacres2,
   "LAM"=lacmres2,"ALT"=alcres2,"ALW"=nalcres2,"ALM"=almres2)
 
-x <- do.call(rbind,lapply(l,summarise))
+x1 <- do.call(rbind,lapply(l1,summarise))
 
-x %>% kbl(caption="sim 20 res") %>% kable_paper("hover", full_width = F)
+x1 %>% kbl(caption="sim 20 res") %>% kable_paper("hover", full_width = F)
 
@@
 
-barplot(t(x),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, cex.names=0.7, cex.axis=0.7, main="20 genes per set")
+barplot(t(x1),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, cex.names=0.7, cex.axis=0.7, main="20 genes per set")
 grid()
@@
 
-l <- list("GSA"=gres2,"LAT"=lacres2,"LTT"=lactopres2,"LAW"=nlacres2,
+l2 <- list("GSA"=gres2,"LAT"=lacres2,"LTT"=lactopres2,"LAW"=nlacres2,
   "LAM"=lacmres2,"ALT"=alcres2,"ALW"=nalcres2,"ALM"=almres2)
 
-x <- do.call(rbind,lapply(l,summarise))
+x2 <- do.call(rbind,lapply(l2,summarise))
 
-x %>% kbl(caption="sim 50 res") %>% kable_paper("hover", full_width = F)
+x2 %>% kbl(caption="sim 50 res") %>% kable_paper("hover", full_width = F)
 
-barplot(t(x),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, cex.names=0.7, cex.axis=0.7, main="50 genes per set")
+barplot(t(x2),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, cex.names=0.7, cex.axis=0.7, main="50 genes per set")
 grid()
@@
 
-l <- list("GSA"=gres2,"LAT"=lacres2,"LTT"=lactopres2,"LAW"=nlacres2,
+l3 <- list("GSA"=gres2,"LAT"=lacres2,"LTT"=lactopres2,"LAW"=nlacres2,
   "LAM"=lacmres2,"ALT"=alcres2,"ALW"=nalcres2,"ALM"=almres2)
 
-x <- do.call(rbind,lapply(l,summarise))
+x3 <- do.call(rbind,lapply(l3,summarise))
 
@@
 grid()
+
+```
+
+Make a PNG copy of the data.
+
+```{r,summary_png,fig.width=5,fig.height=4}
+
+png(filename="fig1.png",height=70,width=210,units="mm",res=200)
+
+par(mar = c(2.5, 4.1, 2.1, 0.5))
+
+par(mfrow=c(1,3))
+
+cols <- c("white","gray","black")
+
+barplot(t(x1),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, cex.names=0.75, cex.axis=0.75, main="20 genes per set")
+grid()
+
+legend("topleft", inset=.1, c("Precision","Recall","F1"), fill=cols, cex=0.75)
+
+barplot(t(x2),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, cex.names=0.75, cex.axis=0.75, main="50 genes per set")
+grid()
+
+barplot(t(x3),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, cex.names=0.75, cex.axis=0.75, main="100 genes per set")
+grid()
+
+dev.off()
+
+par(mfrow=c(1,1))
 

### 2023-11-04 manuscript/manuscript.Rmd
ms

@@
 We found that overall recall was positively associated with the size of the gene sets considered.
-At 20 genes per set, GSA had the highest overall performance, with F1 of 0.25, followed by LAW with 0.06.
-At 50 genes per set, GSA and LAM achieved similar F1 values of ~0.39.
+At 20 genes per set, GSA had the highest overall performance, with F1 of 0.25, followed by LAW with 0.06.At 50 genes per set, GSA and LAM achieved similar F1 values of ~0.39.
 At 100 genes per set, LAM achieved the highest F1 performance of 0.70 followed by LAT with 0.61, while
@@
 This result suggests that FCS methods perform better than ORA with gene set sizes >50.
-Performance results for GSA and LAM methods with 100 genes per set shows that recall was higher for most
-simulations, particularly with lower group sizes.
+
+![Figure 1. Precision, recall and F1 for enrichment tests with simulated data.]("path/to/image")
+
+Performance results for GSA and LAM methods with 100 genes per set shows that LAM recall was higher for
+most simulations, particularly with lower group sizes.
 Taken together, this result suggests that LAM offers improved sensitivity over existing approaches when

### 2023-11-04 manuscript/manuscript.Rmd
writing

@@
     fig_height: 5
-bibliography: gmea.bib
-csl: plos-computational-biology.csl
+bibliography: manuscript.bib
+csl: epigenetics.csl
 ---
@@
 Mark Ziemann<sup>1,2*</sup>,
-Mandhri Abeysooriya<sup>2,3*</sup>,
+Mandhri Abeysooriya<sup>2,3</sup>,
 Anusuiya Bora<sup>2</sup>,
@@
 True positives, false positives and false negatives were used to calculate overall precision and recall
-at these three gene set sizes (Figure 1).
+at these three gene set sizes (*Figure 1*).
 We found that overall recall was positively associated with the size of the gene sets considered.
@@
 
-![Figure 1. Precision, recall and F1 for enrichment tests with simulated data.]("path/to/image")
+![Figure 1. Precision, recall and F1 for enrichment tests with simulated data.](../figs/fig1.png)
 
-Performance results for GSA and LAM methods with 100 genes per set shows that LAM recall was higher for
-most simulations, particularly with lower group sizes.
+Performance results for GSA and LAM methods with 100 genes per set shows that precision was similar, with most simulations yielding
+>95% precision (*Figure 2*).
+LAM recall was superior for most simulations, particularly with lower group sizes.
 Taken together, this result suggests that LAM offers improved sensitivity over existing approaches when
@@
 
- in Figure 2.
-This 
-As expected, precision and recall values were better when group sizes and delta values were larger.
+![Figure 2. Precision, recall and F1 for GSA and LAM enrichment tests over a range of group sizes, delta values and gene set sizes.](../figs/fig2.png)
 
-These data showed a strong relationship between the size of the gene sets and the precision and recall,
-regardless of the enrichment test used.
-On the whole, precision and recall were better
+### Sensitivity of GSA and LAM methods with real cancer data
+
+Next, GSA and LAM methods were applied to compare paired normal and cancer methylation data.
+
+* Chart results with 37 pax.
+
+* Show sensitivity analysis with range of samples from 3 to 30.
 

### 2023-11-15 figs/GSE158422_sensitivity.Rmd
started working on the sensitivity analysis again

--- 2023-10-12 figs/GSE158422_sensitivity.Rmd
@@
 
-In the previous analysis, it was shown that the "AL" (aggregation-limma-enrichment) method
-yielded good performance with a cohort of cancer-normal tissue pairs.
+In the previous analysis, it was shown that the limma-average-mitch (LAM) method
+yielded good performance with simulated data.
 
 In this piece of work, we want to ascertain the sensitivity of this approach as compared to
-existing methods such as gsameth and ebGSEA.
+the existing method, GSAmeth.
 
@@
 the dataset down to n=3 patients.
-What we expect is that the "AL" method could have a higher degree of recall as compared to these
-other approaches.
-We expect that "AL" could identify gene sets with larger effect sizes (enrichment scores) using
+What we expect is that LAM might have a higher degree of recall as compared to GSAmeth.
+We expect LAM will identify gene sets with larger effect sizes (enrichment scores) using
 fewer samples.
 This means that for a given set of datasets, this approach could yield more findings, or
-alternatively if pathway analysis is the primary outcome of a study, it need not be as large which
+alternatively if pathway analysis is the primary outcome of a study, it need not be as large, which
 can save money and difficulty with establishing large sample collections.
@@
 RAM usage can be moderated by reducing the parallel core count.
+
+This analysis was run on an AMD Ryzen Threadripper PRO 5955WX 16-Cores with 3200 MT/s DDR4.
 
@@
 CORES=24
+
+TESTING=TRUE
 
@@
 
-```{r,functions}
+limma-average-mitch (LAM).
 
-# magg is a general function that aggregates probe methylation measurements by gene
-# in this variation it aggregates by MEAN, which performs better than median
-magg <- function(mval,myann,cores=1){
-  gn <- unique(unlist(strsplit( myann$UCSC_RefGene_Name ,";")))
-  gnl <- strsplit( myann$UCSC_RefGene_Name ,";")
-  gnl <- mclapply(gnl,unique,mc.cores=cores)
-  myann$gnl <- gnl
-  keep <- rownames(subset(myann,UCSC_RefGene_Name!=""))
-  mx <- mval[rownames(mval) %in% keep,]
-  mymed <- function(g) {
-    probes <- rownames(myann[grep(g,myann$gnl),])
-    rows <- which(rownames(mx) %in% probes)
-    if ( length(rows) > 1 ) {
-      b <- mx[rows,]
-      med <- apply(b,2,mean)
-      med <- matrix(med,nrow=1)
-      colnames(med) <- colnames(b)
-      rownames(med) <- g
-      return(med)
-    }
-  }
-  med <- mclapply(gn,mymed,mc.cores=cores)
-  med <- med[lapply(med,length)>0]
-  medf <- do.call(rbind,med)
-  return(medf)
-}
+```{r,limmafunc}
 
-# chragg is a specialised function that performs the magg routine on chromosomes in parallel
-chragg <- function(mval,myann,cores=1){
-  annodf <- as.data.frame(anno)
-  keep <- rownames(subset(myann,UCSC_RefGene_Name!=""))
-  mx <- mval[rownames(mval) %in% keep,]
-  chrs <- unique(anno$chr)
-  myorder <- unlist(lapply(chrs,function(mychr) { nrow( annodf[annodf$chr==mychr,] ) } ))
-  chrs <- chrs[order(-myorder)]
-  leadercores <- floor(sqrt(cores))
-  workercores <- ceiling(sqrt(cores))
-  chrmedf <- mclapply(chrs,function(chr) {
-    chrfrag <- annodf[annodf$chr==chr,]
-    chrprobes <-rownames(chrfrag)
-    chrmx <- mx[rownames(mx) %in% chrprobes,]
-    chranno <- myann[rownames(myann) %in% chrprobes,]
-    chrmedf <- magg(mval=chrmx,myann=chranno,cores=workercores)
-    return(chrmedf)
-  },mc.cores=leadercores)
-  medf <- do.call(rbind, chrmedf)
-  return(medf)
-}
-
-# agglimma is a helper function that runs limma analysis on the aggregated methylation data
-# and fixes any duplicated gene names - most significant one is retained
-agglimma <- function(medf,design) {
-  fit.reduced <- lmFit(medf,design)
+# limma
+runlimma <- function(mval,design,myann) {
+  fit.reduced <- lmFit(mval,design)
   fit.reduced <- eBayes(fit.reduced)
-  dmagg <- topTable(fit.reduced,coef=ncol(design), number = Inf)
-  nondup <- !duplicated(dmagg$ID)
-  dmagg <- dmagg[nondup,]
-  rownames(dmagg) <- dmagg$ID
-  dmagg$ID = NULL
-  return(dmagg)
-}
-
-# ttenrich is a function designed to conduct analysis of differential methylation.
-# It takes the t-statistics from limma for a geneset and determines whether this gene
-# set has a distribution different to the null (zero) using a one-sample two-sided t-test.
-ttenrich <- function(m,genesets,cores=1) {
-  res <- mclapply( 1:length(genesets), function(i) {
-    gs <- genesets[i]
-    name <- names(gs)
-    n_members <- length(which(rownames(m) %in% gs[[1]]))
-    if ( n_members > 4 ) {
-      tstats <- m[which(rownames(m) %in% gs[[1]]),]
-      myn <- length(tstats)
-      mymean <- mean(tstats)
-      mymedian <- median(tstats)
-      wt <- t.test(tstats)
-      res <- c(name,myn,mymean,mymedian,wt$p.value)
-    }
-  } , mc.cores = cores)
-  res_df <- do.call(rbind, res)
-  rownames(res_df) <- res_df[,1]
-  res_df <- res_df[,-1]
-  colnames(res_df) <- c("n_genes","t_mean","t_median","pval")
-  tmp <- apply(res_df,2,as.numeric)
-  rownames(tmp) <- rownames(res_df)
-  res_df <- tmp
-  res_df <- as.data.frame(res_df)
-  res_df <- res_df[order(res_df$pval),]
-  res_df$logp <- -log10(res_df$pval )
-  res_df$fdr <- p.adjust(res_df$pval,method="fdr")
-  res_df[order(abs(res_df$pval)),]
-  return(res_df)
+  dm <- topTable(fit.reduced,coef=ncol(design), number = Inf)
+  dm <- merge(myann,dm,by=0)
+  dm <- dm[order(dm$P.Value),]
+  rownames(dm) <- dm$Row.names
+  dm$Row.names=NULL
+  return(dm)
 }
@@
 
-Run AL method on methylation data.
+The following function is key for the aggregation of probe t-statistic values by gene using the arithmetic mean.
+In this step, ~800k probe values are summarised to ~28k genes.
+
+```{r,lamfunction}
+
+# LAM aggregate
+agg <- function(dm,cores=1) {
+  dm <- dm[which(dm$UCSC_RefGene_Name!=""),]
+  gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
+  gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
+  gnl <- mclapply(gnl,unique,mc.cores=cores)
+  dm$UCSC_RefGene_Name <- gnl
+  l <- mclapply(1:nrow(dm), function(i) {
+    a <- dm[i,]
+    len <- length(a[[1]][[1]])
+    tvals <- as.numeric(rep(a["t"],len))
+    genes <- a[[1]][[1]]
+    data.frame(genes,tvals)
+  },mc.cores=cores)
+  df <- do.call(rbind,l)
+  keep <- names(which(table(df$genes)>1))
+  df <- df[df$genes %in% keep,]
+  gn <- unique(df$genes)
+  gme_res <- lapply( 1:length(gn), function(i) {
+    g <- gn[i]
+    tstats <- df[which(df$genes==g),"tvals"]
+    myn <- length(tstats)
+    mymean <- mean(tstats)
+    mymedian <- median(tstats)
+    top <- tstats[order(-abs(tstats))][1]
+    if ( length(tstats) > 2 ) {
+      ttest <- t.test(tstats)
+      pval <- ttest$p.value
+    } else {
+      pval = 1
+    }
+    res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,
+      "median"=mymedian, "top"=top, "pval"=pval)
+  } )
+  gme_res_df <- do.call(rbind, gme_res)
+  rownames(gme_res_df) <- gme_res_df[,1]
+  gme_res_df <- gme_res_df[,-1]
+  tmp <- apply(gme_res_df,2,as.numeric)
+  rownames(tmp) <- rownames(gme_res_df)
+  gme_res_df <- as.data.frame(tmp)
+  gme_res_df$sig <- -log10(gme_res_df[,"pval"])
+  gme_res_df <- gme_res_df[order(-gme_res_df$sig),]
+  gme_res_df$fdr <- p.adjust(gme_res_df$pval)
+  out <- list("df"=df,"gme_res_df"=gme_res_df)
+  return(out)
+}
+
+```
+
+Function to run mitch analysis.
+
+```{r,mitchfunc}
+
+runmitch <- function(m,genesets,cores=1) {
+  suppressMessages({ mres <- mitch_calc(m,genesets,minsetsize=5,cores=cores) })
+  mres <- mres$enrichment_result
+  rownames(mres) <- mres$set
+  mres$set=NULL
+  return(mres)
+}
+
+```
+
+## Run LAM
+
+Run LAM method on tumour-normal methylation data.
 Record timings.
@@
 
-# aggregation step
+dim(mval)
+dim(design)
+
 tic()
-medf <- chragg(mval,myann,cores=CORES)
-toc() #51.5 sec
+dm <- runlimma(mval=mval,design=design,myann=myann)
+toc() ## mean(23.7 26.8 25.4)=23.7
+top <- head(dm,20)
+top <- top[order(top$logFC),]
 
-medf_means <- list("tumor"=matrix(colMeans(medf),ncol=2)[,1],"normal"=matrix(colMeans(medf),ncol=2)[,2])
-boxplot(medf_means, main="mean gene methylation mval",cex=0,col="white")
-beeswarm(medf_means,add=TRUE,pch=1,cex=1.5)
+tic()
+dmagg <- agg(dm,cores=32)
+m <- dmagg$gme_res_df[,"mean",drop=FALSE]
+lamres <- runmitch(m=m,genesets=gs_symbols,cores=32)
+toc()
 
-lapply(medf_means,mean)
+#tic()
+#dmagg <- agg(dm,cores=16)
+#m <- dmagg$gme_res_df[,"mean",drop=FALSE]
+#lamres <- runmitch(m=m,genesets=gs_symbols,cores=16)
+#toc()
+#c32<-c()
+#c16<-c(123.666,129.619)
+#c8<-c()
+#c4<-c()
+#c2<-c()
+#c1<-c()
+#duration <- list("c1"=c1, "c2"=c2, "c4"=c4, "c8"=c8, "c16"=c16,"c32"=c32)
 
-# differential analysis
-tic()
-dmagg <- agglimma(medf,design)
-toc() #0.46 sec
-
-hist(dmagg$t)
-
-head(dmagg, 30) %>%
-  kbl(caption="limma gene methylation results") %>%
+top %>%
+  kbl(caption="Top significant probes") %>%
   kable_paper("hover", full_width = F)
 
-# get t-stats
-m <- as.data.frame(dmagg$t)
-rownames(m) <- rownames(dmagg)
-colnames(m) <- "t"
+lamsig <- subset(lamres,p.adjustANOVA<0.05)
 
-# enrichment test
-tic()
-alres <- ttenrich(m=m,genesets=gs_symbols,cores=CORES)
-toc() #1.1 sec
+top <- head(lamsig,20)
+top <- top[order(top$s.dist),]
 
-hist(alres$t_mean)
-
-alsig <- subset(alres,fdr<0.05)
-
-nrow(alsig)
-
-nrow(subset(alsig,t_median>0))
-
-head(alsig[order(-alsig$t_median),],20) %>%
-  kbl(caption="hypermethylated pathways") %>%
+top %>%
+  kbl(caption="Top gene sets (biggest effect size with FDR<0.05)") %>%
   kable_paper("hover", full_width = F)
 
-nrow(subset(alsig,t_median<0))
+barnames <- gsub("_"," ",gsub("REACTOME_","",rownames(top)))
+cols <- gsub("TRUE","red",gsub("FALSE","blue",as.character(top$s.dist>0)))
 
-head(alsig[order(alsig$t_median),],20) %>%
-  kbl(caption="hypomethylated pathways") %>%
-  kable_paper("hover", full_width = F)
+par(mar = c(5.1, 29.1, 4.1, 2.1))
+barplot(abs(top$s.dist),horiz=TRUE,las=1,names.arg=barnames, cex.names=0.65,
+  cex.axis=0.8,col=cols, xlab="Enrichment score",main="Reactomes")
+grid()
+
+par( mar = c(5.1, 4.1, 4.1, 2.1) )
+
+lamsig$s.dist
+
+# distinguish up (TRUE) and down (FALSE)
+lamsigsets <- paste(rownames(lamsig),as.character(lamsig$s.dist > 0))
 
@@
 
-Try a competitive test.
+## Downsample LAM test
 
-```{r,compet}
-
-mres <- mitch_calc(x=m,minsetsize=5,genesets=gs_symbols,cores=32,priority="effect")
-
-# hypermethylated
-mres_up <- subset(mres$enrichment_result,p.adjustANOVA<0.05 & s.dist > 0)
-nrow(mres_up)
-head(mres_up,20) %>%
-  kbl(caption="hypermethylated pathways with competitive test") %>%
-  kable_paper("hover", full_width = F)
-
-# hypomethylated
-mres_dn <- subset(mres$enrichment_result,p.adjustANOVA<0.05 & s.dist < 0)
-nrow(mres_dn)
-head(mres_dn,20) %>%
-  kbl(caption="hypomethylated pathways with competitive test") %>%
-  kable_paper("hover", full_width = F)
-
-```
-
-These results look more in line with cancer pathogenesis.
-
-Downsample AL test
+This will run the LAM analysis at group sizes of 2, 3, 6, 10, 16, 22 and 30,
+50 times each, with pseudorandom seeds from 100 to 5000 jumping up by 100.
 
@@
 
-#medf already exists
-
-SAMPLESIZES=c(2,3,6,10,16,22,30)
-
+#testing
+SAMPLESIZES=c(22,10,3)
 ssres <- lapply(SAMPLESIZES,function(n) {
-  SEEDS <- seq(100,5000,100)
+  SEEDS <- seq(100,300,100)
   dsres <- mclapply(SEEDS, function(SEED) {
@@
     design2 <- design2[,colSums(design2)>0]
-    medf2 <- medf[,mysample]
-    dmagg2 <- agglimma(medf2,design2)
-    m2 <- as.data.frame(dmagg2$t)
-    rownames(m2) <- rownames(dmagg2)
-    colnames(m2) <- "t"
-    alres2 <- ttenrich(m=m2,genesets=gs_symbols,cores=CORES)
-    head(alres2)
-    alsig2 <- subset(alres2,fdr<0.05)
-    TOT=nrow(alsig)
-    TP=length(which(rownames(alsig2) %in% rownames(alsig)))
-    FP=length(which(!rownames(alsig2) %in% rownames(alsig)))
-    TPES=mean(abs(alsig[which(rownames(alsig) %in% rownames(alsig2)),"t_mean"]))
-    RES=c("TOT"=TOT,"TP"=TP,"FP"=FP,"TPES"=TPES)
+    mval2 <- mval[,mysample]
+    dm2 <- runlimma(mval=mval2,design=design2,myann=myann)
+    dmagg2 <- agg(dm2,cores=2)
+    m2 <- dmagg2$gme_res_df[,"mean",drop=FALSE]
+    lamres2 <- runmitch(m=m2,genesets=gs_symbols,cores=2)
+    lamsig2 <- subset(lamres2,p.adjustANOVA<0.05)
+    lamsigsets2 <- paste(rownames(lamsig2),as.character(lamsig2$s.dist > 0))
+    TOT=length(lamsigsets2)
+    TP=length(which(lamsigsets2 %in% lamsigsets))
+    FP=length(which(!lamsigsets2 %in% lamsigsets))
+    FN=length(which(!lamsigsets %in% lamsigsets2))
+    RES=c("TOT"=TOT,"TP"=TP,"FP"=FP,"FN"=FN)
     RES
-  },mc.cores=2)
+  },mc.cores=5)
   do.call(rbind,dsres)
 })
+blah()
+
+# full
+if (TESTING == FALSE) {
+
+SAMPLESIZES=c(2,3,6,10,16,22,30)
+ssres <- lapply(SAMPLESIZES,function(n) {
+  SEEDS <- seq(100,300,100)
+  dsres <- mclapply(SEEDS, function(SEED) {
+    set.seed(SEED)
+    mysample <- sample(1:37,n,replace=FALSE)*2
+    mysample <- c(mysample,(mysample-1))
+    mysample <- mysample[order(mysample)]
+    design2 <- design[mysample,]
+    design2 <- design2[,colSums(design2)>0]
+    mval2 <- mval[,mysample]
+    dm2 <- runlimma(mval=mval2,design=design2,myann=myann)
+    dmagg2 <- agg(dm2,cores=2)
+    m2 <- dmagg2$gme_res_df[,"mean",drop=FALSE]
+    lamres2 <- runmitch(m=m2,genesets=gs_symbols,cores=2)
+    lamsig2 <- subset(lamres2,p.adjustANOVA<0.05)
+    lamsigsets2 <- paste(rownames(lamsig2),as.character(lamsig2$s.dist > 0))
+    TOT=length(lamsigsets2)
+    TP=length(which(lamsigsets2 %in% lamsigsets))
+    FP=length(which(!lamsigsets2 %in% lamsigsets))
+    FN=length(which(!lamsigsets %in% lamsigsets2))
+    RES=c("TOT"=TOT,"TP"=TP,"FP"=FP,"FN"=FN)
+    RES
+  },mc.cores=5)
+  do.call(rbind,dsres)
+})
+
+}
 

### 2023-11-15 manuscript/manuscript.Rmd
started working on the sensitivity analysis again

@@
 ---
-title: "Finally, functional class scoring enrichment analysis of Infinium DNA methylation data"
+title: "Finally, a functional class scoring method for enrichment analysis of Infinium DNA methylation data"
 date: "`r Sys.Date()`"
@@
 
-![Figure 2. Precision, recall and F1 for GSA and LAM enrichment tests over a range of group sizes, delta values and gene set sizes.](../figs/fig2.png)
+![Figure 2. Precision, recall and F1 for GSA and LAM enrichment tests over a range of group sizes and delta values. One hundred genes per set.](../figs/fig2.png)
 

### 2023-11-15 figs/GSE158422_sensitivity.Rmd
progress on sensitivity analysis

@@
 CORES=24
-
-TESTING=TRUE
 
@@
 
-#testing
-SAMPLESIZES=c(22,10,3)
-ssres <- lapply(SAMPLESIZES,function(n) {
-  SEEDS <- seq(100,300,100)
-  dsres <- mclapply(SEEDS, function(SEED) {
-    set.seed(SEED)
-    mysample <- sample(1:37,n,replace=FALSE)*2
-    mysample <- c(mysample,(mysample-1))
-    mysample <- mysample[order(mysample)]
-    design2 <- design[mysample,]
-    design2 <- design2[,colSums(design2)>0]
-    mval2 <- mval[,mysample]
-    dm2 <- runlimma(mval=mval2,design=design2,myann=myann)
-    dmagg2 <- agg(dm2,cores=2)
-    m2 <- dmagg2$gme_res_df[,"mean",drop=FALSE]
-    lamres2 <- runmitch(m=m2,genesets=gs_symbols,cores=2)
-    lamsig2 <- subset(lamres2,p.adjustANOVA<0.05)
-    lamsigsets2 <- paste(rownames(lamsig2),as.character(lamsig2$s.dist > 0))
-    TOT=length(lamsigsets2)
-    TP=length(which(lamsigsets2 %in% lamsigsets))
-    FP=length(which(!lamsigsets2 %in% lamsigsets))
-    FN=length(which(!lamsigsets %in% lamsigsets2))
-    RES=c("TOT"=TOT,"TP"=TP,"FP"=FP,"FN"=FN)
-    RES
-  },mc.cores=5)
-  do.call(rbind,dsres)
-})
-blah()
-
-# full
-if (TESTING == FALSE) {
-
 SAMPLESIZES=c(2,3,6,10,16,22,30)
 ssres <- lapply(SAMPLESIZES,function(n) {
-  SEEDS <- seq(100,300,100)
+  SEEDS <- seq(100,5000,100)
   dsres <- mclapply(SEEDS, function(SEED) {
@@
 
-}
-
 par(mfrow=c(3,1))
@@
 dots <- lapply(ssres,function(x) { x[,"TP"] })
-boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. true positives",xlab="sample size",cex=0,main="GMEA") ; grid()
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. true positives",xlab="sample size",cex=0,main="LAM") ; grid()
 beeswarm(dots,add=TRUE,pch=1,cex=0.6)
@@
 
-# limma
-runlimma <- function(mval,design,myann) {
-  fit.reduced <- lmFit(mval,design)
-  fit.reduced <- eBayes(fit.reduced)
-  summary(decideTests(fit.reduced))
-  dm <- topTable(fit.reduced,coef=4, number = Inf)
-  dm <- merge(myann,dm,by=0)
-  dm <- dm[order(dm$P.Value),]
-  rownames(dm) <- dm$Row.names
-  dm$Row.names=NULL
-  return(dm)
-}
-
 dm <- runlimma(mval,design,myann)
@@
 
-head(dm, 10) %>% kbl() %>% kable_paper("hover", full_width = F)
+head(dm, 10) %>% kbl(caption="Top 10 probes by p-value") %>% kable_paper("hover", full_width = F)
 
 sigup <- rownames(subset(dm,logFC > 0 & adj.P.Val < 0.05))
+
 sigdn <- rownames(subset(dm,logFC < 0 & adj.P.Val < 0.05))
@@
 gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez)
+rownames(gsaup) <- paste(rownames(gsaup),"UP")
 gsadn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez)
-toc() # 12.9 sec
+rownames(gsadn) <- paste(rownames(gsadn),"DN")
+toc() # 12.9, 12.6, 11.5
 
@@
 gsig_up <-  subset(gsaup,FDR<0.05)
-gsig_up %>% kbl(caption="Hypermethylated pathways") %>% kable_paper("hover", full_width = F)
+head(gsig_up,20) %>% kbl(caption="Hypermethylated pathways") %>% kable_paper("hover", full_width = F)
+
 gsig_dn <-  subset(gsadn,FDR<0.05)
-gsig_dn %>% kbl(caption="Hypomethylated pathways") %>% kable_paper("hover", full_width = F)
+head(gsig_dn,20) %>% kbl(caption="Hypomethylated pathways") %>% kable_paper("hover", full_width = F)
+
 gsig <- rbind(gsig_up,gsig_dn)
@@
     dm2 <- runlimma(mval2,design2,myann)
-    sigup2 <- rownames(subset(dm2,adj.P.Val<0.05 & logFC>0))
-    sigdn2 <- rownames(subset(dm2,adj.P.Val<0.05 & logFC<0))
-    #gsamet
+    if ( is.na(dm2$logFC[1]) ) {
+      RES=c("TOT"=0,"TP"=0,"FP"=0,"FN"=length(rownames(gsig)) )
+    } else {
+      sigup2 <- rownames(subset(dm2,adj.P.Val<0.05 & UCSC_RefGene_Name != "" & logFC>0))
+      if ( length(sigup2) < 250 ) { sigup2 <- head(rownames(subset(dm2,logFC>0 & UCSC_RefGene_Name != "" )),250) }
+      sigdn2 <- rownames(subset(dm2,adj.P.Val<0.05 & UCSC_RefGene_Name != "" & logFC<0))
+      if ( length(sigdn2) < 250 ) { sigdn2 <- head(rownames(subset(dm2,logFC<0 & UCSC_RefGene_Name != "" )),250) }
       gsaup2 <- gsameth(sig.cpg=sigup2, all.cpg=rownames(dm2), collection=gs_entrez)
+      rownames(gsaup2) <- paste(rownames(gsaup2),"UP")
       gsadn2 <- gsameth(sig.cpg=sigdn2, all.cpg=rownames(dm2), collection=gs_entrez)
+      rownames(gsadn2) <- paste(rownames(gsadn2),"DN")
       gsig_up2 <- subset(gsaup2,FDR<0.05)
@@
       FP=length(which(!rownames(gsig2) %in% rownames(gsig)))
-    TPES=0
-    RES=c("TOT"=TOT,"TP"=TP,"FP"=FP,"TPES"=TPES)
+      FN=length(which(!rownames(gsig) %in% rownames(gsig2)))
+      RES=c("TOT"=TOT,"TP"=TP,"FP"=FP,"FN"=FN)
+    }
     RES
-  },mc.cores=8)
+  }, mc.cores=5)
   do.call(rbind,dsres)
@@
 mx <- matrix(rep(0,200),ncol=4)
-colnames(mx) <- c("TOT","TP","FP","TPES")
+colnames(mx) <- c("TOT","TP","FP","FN")
 
@@
 
+blah()
+
 alsig_up <- rownames(subset(alsig,t_median>0))

### 2023-11-15 manuscript/manuscript.Rmd
progress on sensitivity analysis

@@
 methods at differenent effect size levels.
-This process was repeated 20 times.
+This process was repeated 50 times.
 
@@
 
-We applied the three distinct approaches ("LA", "AL" and "AA") for pathway level differential
-methylation analysis to a dataset consisting of lung tumour and adjacent normal tissue from 37 patients.
-The dataset was split randomly into two groups of 18 patients and the overlap between subset was used to
-assess the consistency and validity of findings with 50 simulations.
-From the 1550 Reactome gene sets in the dataset, the AL method recorded the highest number of
-"valid" pathways with an average of 919.8 being identified in both subsets, which was much higher than
-LA and AA methods with 295.3 and 276.7 gene sets respectively (Figure 1A).
-The AL method also recorded the lowest number of uncommon sets (382.0) which was better than the LA and
-AA methods with 634.8 and 908.3 respectively (Figure 1B).
-We also noticed that the LA method sometimes yielded contradictory results, so we quantified the number
-of discordant observations, that is statistically significant but in the opposite direction.
-True to this observation, the LA method gave 461.6 discordant pathways which indicates a high rate of
-false-positives.
-Meanwhile the AL and AA methods gave a mean of 0.1 and 0.02 discordant pathways respectively.
-Taken together, these results indicate that the AL method gives the highest rate of common results
-suggesting good accuracy in terms of sensitivity and specificity in this context.
+* LAM has better performance than GSAmeth
+
+* Could be due to (1) filtering genes without annotations, and (2) FDR not applied for sets without members.
+
+* Can make changes to the GSAmeth codes to fix the problems.
 

### 2023-11-15 figs/GSE158422_sensitivity.Rmd
sensitivity done

@@
 ```
+## Run limma
 
-## Run LAM
+Run limma and look at the absolute effect sizes.
 
-Run LAM method on tumour-normal methylation data.
-Record timings.
-
-```{r,rungmea}
+```{r,limma1}
 
@@
 top <- top[order(top$logFC),]
+
+par(mfrow=c(3,1))
+
+sig <- head(subset(dm,adj.P.Val < 0.05 ),100)
+sigmval <- mval[rownames(mval) %in% rownames(sig),]
+es <- lapply(1:nrow(sigmval), function(i) {
+  x <- sigmval[i,]
+  mx <- matrix(x,ncol=2)
+  cancer <- median(mx[,1])
+  normal <- median(mx[,2])
+  abs(cancer-normal)
+} )
+hist(unlist(es),main="top 100 probes", xlab="probe effect size")
+
+sig <- head(subset(dm,adj.P.Val < 0.05 ),1000)
+sig <- sig[101:1000,]
+sigmval <- mval[rownames(mval) %in% rownames(sig),]
+es <- lapply(1:nrow(sigmval), function(i) {
+  x <- sigmval[i,]
+  mx <- matrix(x,ncol=2)
+  cancer <- median(mx[,1])
+  normal <- median(mx[,2])
+  abs(cancer-normal)
+} )
+hist(unlist(es),main="top 1000 probes ex top 100", xlab="probe effect size")
+
+sig <- head(subset(dm,adj.P.Val < 0.05 ),10000)
+sig <- sig[1001:10000,]
+sigmval <- mval[rownames(mval) %in% rownames(sig),]
+es <- lapply(1:nrow(sigmval), function(i) {
+  x <- sigmval[i,]
+  mx <- matrix(x,ncol=2)
+  cancer <- median(mx[,1])
+  normal <- median(mx[,2])
+  abs(cancer-normal)
+} )
+hist(unlist(es), main="top 10000 probes ex top 1000", xlab="probe effect size")
+
+par(mfrow=c(1,1))
+
+```
+
+## Run LAM
+
+Run LAM method on tumour-normal methylation data.
+Record timings.
+
+```{r,rungmea}
 
@@
 
-lamsig$s.dist
+hist(lamsig$s.dist)
 
@@
 lamsigsets <- paste(rownames(lamsig),as.character(lamsig$s.dist > 0))
+
+lamsigsets2 <- gsub(" TRUE"," UP",lamsigsets)
+lamsigsets2 <- gsub(" FALSE"," DN",lamsigsets2)
 
@@
 beeswarm(dots,add=TRUE,pch=1,cex=0.6)
-abline(h= ssres[[1]][1,1])
+abline(h=nrow(lamsig))
 
@@
 gsig <- rbind(gsig_up,gsig_dn)
+gsigsets <- rownames(gsig)
 
@@
 beeswarm(dots,add=TRUE,pch=1,cex=0.6)
-abline(h=gsres[[1]][1,1])
+abline(h=nrow(gsig))
 
@@
 
-blah()
+lamsigsets2
+gsigsets
 
-alsig_up <- rownames(subset(alsig,t_median>0))
-alsig_dn <- rownames(subset(alsig,t_median<0))
-
-msig_up <- subset(mres$enrichment_result,p.adjustANOVA<0.05 & s.dist > 0)$set
-msig_dn <- subset(mres$enrichment_result,p.adjustANOVA<0.05 & s.dist < 0)$set
-
-gsig_up <-  rownames(subset(gsaup,FDR<0.05))
-gsig_dn <-  rownames(subset(gsadn,FDR<0.05))
-
-v1 <- list("AL up"=alsig_up,"AL dn"=alsig_dn,
-  "mitch up"=msig_up,"mitch dn"=msig_dn,
-  "GSAmeth up"=gsig_up,"GSAmeth dn"=gsig_dn)
+v1 <- list("LAM"=lamsigsets2,"GSA"=gsigsets)
 

### 2023-11-16 figs/GSE158422_sensitivity.Rmd
sensitivity looking good!

@@
   library("missMethyl")
+  library("vioplot")
 })
@@
 
+Now look at the size of these genesets.
+Make sure these are ACTUALLY detected.
+Make a vector of genes.
+
+```{r,sizes}
+
+detgenes <- unique(unlist(strsplit(dm$UCSC_RefGene_Name,";")))
+
+gsigsets1 <- sapply(strsplit(gsigsets," "),'[[',1)
+gsigsets1set <- gs_symbols[names(gs_symbols) %in% gsigsets1]
+gsigsets1len <- unlist(lapply(gsigsets1set,function(set) {
+  length(set[which(set %in% detgenes)])
+}))
+
+lamsigsets1 <- sapply(strsplit(lamsigsets2," "),'[[',1)
+lamsigsets1set <- gs_symbols[names(gs_symbols) %in% lamsigsets1]
+lamsigsets1len <- unlist(lapply(lamsigsets1set,function(set) {
+  length(set[which(set %in% detgenes)])
+}))
+
+summary(gsigsets1len)
+summary(lamsigsets1len)
+
+GSANAME=paste("GSA (",length(gsigsets1len),")",sep="")
+LAMNAME=paste("LAM (",length(lamsigsets1len),")",sep="")
+mylist <- list(GSANAME=gsigsets1len,LAMNAME=lamsigsets1len)
+boxplot(mylist,cex=0, ylab="gene set size",names=c(GSANAME,LAMNAME))
+beeswarm(mylist,pch=19,add=TRUE,cex=0.7)
+
+```
+
+Interesting result that shows that LAM detected gene sets are smaller (59.5) than
+GSA sets (205).
+
+This could be due to one of the quirks in some ORA analyses were genes without an
+annotation are excluded.
+
 ## Session information

### 2023-11-16 figs/GSE158422_sensitivity.Rmd
finish

@@
   library("missMethyl")
-  library("vioplot")
 })

### 2023-11-16 figs/GSE158422_sensitivity.Rmd
sinsitivity figure

@@
 
+PNG output of boxplots.
+
+```{r,pngboxplot}
+
+png(filename="fig3.png",height=150,width=210,units="mm",res=200)
+par(mar = c(4.5, 4.1, 2.1, 0.5))
+par(mfrow=c(3,2))
+
+dots <- lapply(ssres,function(x) { x[,"TP"] })
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",cex=0,main="LAM") ; grid()
+beeswarm(dots,add=TRUE,pch=1,cex=0.6)
+abline(h=nrow(lamsig))
+
+dots <- lapply(gsres,function(x) { x[,"TP"] })
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",cex=0,main="GSA") ; grid()
+beeswarm(dots,add=TRUE,pch=1,cex=0.6)
+abline(h=nrow(gsig))
+
+dots <- lapply(ssres,function(x) { x[,"FP"] })
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. inconsistent",xlab="sample size",cex=0) ; grid()
+beeswarm(dots,add=TRUE,pch=1,cex=0.6)
+
+dots <- lapply(gsres,function(x) { x[,"FP"] })
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. inconsistent",xlab="sample size",cex=0) ; grid()
+beeswarm(dots,add=TRUE,pch=1,cex=0.6)
+
+dots <- lapply(ssres,function(x) { x[,"TP"] })
+TPdots <- do.call(rbind,dots)
+dots <- lapply(ssres,function(x) { x[,"FP"] })
+FPdots <- do.call(rbind,dots)
+ratio <- t(FPdots / (TPdots + FPdots))
+ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
+boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion inconsistent",xlab="sample size",cex=0) ; grid()
+beeswarm(ratio,add=TRUE,pch=1,cex=0.6)
+
+dots <- lapply(gsres,function(x) { x[,"TP"] })
+TPdots <- do.call(rbind,dots)
+dots <- lapply(gsres,function(x) { x[,"FP"] })
+FPdots <- do.call(rbind,dots)
+ratio <- t(FPdots / (TPdots + FPdots))
+ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
+boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion inconsistent",xlab="sample size",cex=0) ; grid()
+beeswarm(ratio,add=TRUE,pch=1,cex=0.6)
+
+dev.off()
+
+par(mfrow=c(1,1))
+
+```
+
 ## Overlap
@@
 plot(euler(v1),quantities = TRUE)
+
+message("common")
+intersect(lamsigsets2, gsigsets)
+
+message("specific to LAM")
+setdiff(lamsigsets2, gsigsets)
+
+message("specific to GSA")
+setdiff(gsigsets, lamsigsets2)
 

### 2023-11-16 figs/GSE158422_sensitivity.Rmd
next thing to calculate enrichment score of GSA pathways

@@
   library("missMethyl")
+  library("gridExtra")
 })
@@
 hist(unlist(es),main="top 100 probes", xlab="probe effect size")
+summary(unlist(es))
 
@@
 hist(unlist(es),main="top 1000 probes ex top 100", xlab="probe effect size")
+summary(unlist(es))
 
@@
 hist(unlist(es), main="top 10000 probes ex top 1000", xlab="probe effect size")
+summary(unlist(es))
 
@@
 
+Calc GSA enrichment scores.
+First calculate the background which is the detected genes with annotations.
+
+```{r,GSAes}
+
+detgenes <- unique(unlist(strsplit(dm$UCSC_RefGene_Name,";")))
+reactomegenes <- unique(unlist(gs_symbols))
+length(reactomegenes)
+BGSIZE=length(which(detgenes %in% reactomegenes))
+
+```
+
 ## GSAmeth Downsample
@@
 
-png(filename="fig3.png",height=150,width=210,units="mm",res=200)
+png(filename="fig4.png",height=150,width=210,units="mm",res=200)
 par(mar = c(4.5, 4.1, 2.1, 0.5))
@@
 message("specific to LAM")
-setdiff(lamsigsets2, gsigsets)
+LAMspec <- setdiff(lamsigsets2, gsigsets)
+LAMspec
 
 message("specific to GSA")
-setdiff(gsigsets, lamsigsets2)
+GSAspec <- setdiff(gsigsets, lamsigsets2)
+GSAspec
+
+message("discordant")
+intersect( sapply(strsplit(LAMspec," "),"[[",1) , sapply(strsplit(GSAspec," "),"[[",1) )
 
@@
 
+Make Fig3 an euler diagram, boxplot of sizes and barplot of top results.
+
+```{r,eulerbox}
+
+png(filename="euler.png",height=75,width=105,units="mm",res=200)
+plot(euler(v1),quantities = TRUE)
+dev.off()
+
+pic <- readPNG("euler.png")
+
+png(filename="fig3.png",height=150,width=210,units="mm",res=200)
+par(mar = c(4.5, 4.1, 2.1, 0.5))
+par(mfrow=c(2,2))
+
+#A overlap of LAM and GSA results
+par(mar = c(2,0,2,0))
+plot.new()
+plot.window(xlim = c(0 , 1), ylim = c( 0, 1))
+rasterImage(pic,0,0,1,1)
+par(mar = c(4.5, 4.1, 2.1, 0.5))
+
+#B
+boxplot(mylist,cex=0, ylab="gene set size",names=c(GSANAME,LAMNAME))
+beeswarm(mylist,pch=1,add=TRUE,cex=0.3)
+
+#C
+top <- head(lamsig,20)
+top <- top[order(top$s.dist),]
+barnames <- gsub("_"," ",gsub("REACTOME_","",rownames(top)))
+cols <- gsub("TRUE","red",gsub("FALSE","blue",as.character(top$s.dist>0)))
+par(mar = c(6.1, 18.1, 1.1, 0.5))
+barplot(abs(top$s.dist),horiz=TRUE,las=1,names.arg=barnames, cex.names=0.4,
+  cex.axis=0.5,col=cols)
+title(main="LAM",cex = 0.8)
+title(xlab="S dist",cex=0.6)
+grid()
+
+#D
+
+```
+
 Interesting result that shows that LAM detected gene sets are smaller (59.5) than

### 2023-11-16 figs/GSE158422_sensitivity.Rmd
looking good

@@
   library("gridExtra")
+  library("png")
 })
@@
 runmitch <- function(m,genesets,cores=1) {
-  suppressMessages({ mres <- mitch_calc(m,genesets,minsetsize=5,cores=cores) })
+  suppressMessages({ mres <- mitch_calc(m,genesets,minsetsize=5,cores=cores, priority=effect) })
   mres <- mres$enrichment_result
@@
 lamsigsets <- paste(rownames(lamsig),as.character(lamsig$s.dist > 0))
-
 lamsigsets2 <- gsub(" TRUE"," UP",lamsigsets)
@@
 
+sizes=c(1000,5000,10000,20000,50000,100000,200000,300000,350000,400000,450000,500000)
+nsigs <- lapply(sizes, function(z) {
+  dmsig <- head(dm,z)
+  dmsigup <- rownames(subset(dmsig,logFC>0))
+  dmsigdn <- rownames(subset(dmsig,logFC<0))
+  gsaup <- gsameth(sig.cpg=dmsigup, all.cpg=rownames(dm), collection=gs_entrez)
+  rownames(gsaup) <- paste(rownames(gsaup),"UP")
+  gsadn <- gsameth(sig.cpg=dmsigdn, all.cpg=rownames(dm), collection=gs_entrez)
+  rownames(gsadn) <- paste(rownames(gsadn),"DN")
+  nsig <- nrow(subset(rbind(gsaup,gsadn),FDR<0.05))
+  return(nsig)
+})
+nsigs <- unlist(nsigs)
+names(nsigs) <- sizes
+par( mar = c(5.1, 6.1, 4.1, 2.1) )
+barplot(nsigs,horiz=TRUE,las=1,xlab="no. significant sets")
+mtext("no. probes selected", 2, 4)
+
 sigup <- rownames(subset(dm,logFC > 0 & adj.P.Val < 0.05))
-
 sigdn <- rownames(subset(dm,logFC < 0 & adj.P.Val < 0.05))
@@
 length(sigdn)
+
+gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez)
+rownames(gsaup) <- paste(rownames(gsaup),"UP")
+gsadn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez)
+rownames(gsadn) <- paste(rownames(gsadn),"DN")
+ngsig <- nrow(subset(rbind(gsaup,gsadn),FDR<0.05))
 
@@
 length(reactomegenes)
-BGSIZE=length(which(detgenes %in% reactomegenes))
+BG=detgenes[(which(detgenes %in% reactomegenes))]
+BGSIZE=length(BG)
+FGUP <- unique(unlist(strsplit(subset(dm,logFC > 0 & adj.P.Val < 0.05)$UCSC_RefGene_Name,";")))
+FGUP <- FGUP[which(FGUP %in% reactomegenes)]
+FGDN <- unique(unlist(strsplit(subset(dm,logFC < 0 & adj.P.Val < 0.05)$UCSC_RefGene_Name,";")))
+FGDN <- FGDN[which(FGDN %in% reactomegenes)]
+
+myes <- lapply(1:nrow(gsig), function(i) {
+  set=rownames(gsig)[i]
+  set1=sapply(strsplit(set," "),"[[",1)
+  setg <- gs_symbols[[which(names(gs_symbols)==set1)]]
+  setg <- setg[which(setg %in% detgenes)]
+  SETSIZE=length(setg)
+  NUMUP = (length(intersect(FGUP,setg))/length(FGUP))
+  NUMDN = (length(intersect(FGDN,setg))/length(FGDN))
+  NUM=max(NUMUP,NUMDN)
+  DEN = (SETSIZE/BGSIZE)
+  ES=NUM/DEN
+  res=c("numerator"=NUM,"denominator"=DEN,"ES"=ES)
+  return(res)
+})
+
+myes <- data.frame(do.call(rbind,myes))
+gsiges <- data.frame(gsig,myes)
+gsiges <- head(gsiges[order(-gsiges$ES),],20)
+gsiges <- gsiges[order(gsiges$ES),]
+
+barnames <- gsub("_"," ",gsub("REACTOME_","",rownames(gsiges)))
+cols <- gsub("FALSE","blue",gsub("TRUE","red",grepl(" UP",barnames)))
+barnames <- gsub("DN","",gsub(" UP","",barnames))
+
+par(mar = c(5.1, 27.1, 4.1, 2.1))
+barplot(abs(gsiges$ES),horiz=TRUE,las=1,names.arg=barnames, cex.names=0.75,
+  cex.axis=0.85, col=cols, xlab="Enrichment score",main="GSA")
+grid()
 
@@
 #B
+par(mar = c(4.5, 4.1, 2.1, 0.5))
 boxplot(mylist,cex=0, ylab="gene set size",names=c(GSANAME,LAMNAME))
@@
 barnames <- gsub("_"," ",gsub("REACTOME_","",rownames(top)))
+barnames2 <- substring(barnames, 1, 60)
+barnames2 <- paste(barnames2,gsub("TRUE","...",gsub("FALSE","",sapply(barnames,nchar)>60)))
 cols <- gsub("TRUE","red",gsub("FALSE","blue",as.character(top$s.dist>0)))
-par(mar = c(6.1, 18.1, 1.1, 0.5))
-barplot(abs(top$s.dist),horiz=TRUE,las=1,names.arg=barnames, cex.names=0.4,
-  cex.axis=0.5,col=cols)
+par(mar = c(6.1, 20.1, 1.1, 0.5))
+barplot(abs(top$s.dist),horiz=TRUE,las=1,names.arg=barnames2, cex.names=0.55,
+  cex.axis=0.6,col=cols)
 title(main="LAM",cex = 0.8)
@@
 #D
+barnames <- gsub("_"," ",gsub("REACTOME_","",rownames(gsiges)))
+cols <- gsub("FALSE","blue",gsub("TRUE","red",grepl(" UP",barnames)))
+barnames <- gsub("DN","",gsub(" UP","",barnames))
+barnames2 <- substring(barnames, 1, 60)
+barnames2 <- paste(barnames2,gsub("TRUE","...",gsub("FALSE","",sapply(barnames,nchar)>60)))
+par(mar = c(6.1, 20.1, 1.1, 0.5))
+barplot(abs(gsiges$ES),horiz=TRUE,las=1,names.arg=barnames2, cex.names=0.55,
+  cex.axis=0.7, col=cols, xlab="Enrichment score",main="GSA")
+grid()
+
+dev.off()
 

### 2023-11-16 figs/GSE158422_sensitivity.Rmd
bugfix

@@
 runmitch <- function(m,genesets,cores=1) {
-  suppressMessages({ mres <- mitch_calc(m,genesets,minsetsize=5,cores=cores, priority=effect) })
+  suppressMessages({ mres <- mitch_calc(m,genesets,minsetsize=5,cores=cores, priority="effect") })
   mres <- mres$enrichment_result
@@
 barplot(abs(top$s.dist),horiz=TRUE,las=1,names.arg=barnames2, cex.names=0.55,
-  cex.axis=0.6,col=cols)
+  cex.axis=0.7,col=cols)
 title(main="LAM",cex = 0.8)

### 2023-11-17 figs/GSE158422_sensitivity.Rmd
sensitivity done

@@
 
+dim(dm)
+nrow(subset(dm,adj.P.Val<0.05 & logFC>0))
+nrow(subset(dm,adj.P.Val<0.05 & logFC<0))
+
+length(unique(unlist(strsplit(subset(dm,adj.P.Val<0.05 & logFC>0)$UCSC_RefGene_Name,";"))))
+length(unique(unlist(strsplit(subset(dm,adj.P.Val<0.05 & logFC<0)$UCSC_RefGene_Name,";"))))
+
+length(unique(unlist(strsplit(dm$UCSC_RefGene_Name,";"))))
+
 par(mfrow=c(3,1))
@@
   rownames(gsadn) <- paste(rownames(gsadn),"DN")
-  nsig <- nrow(subset(rbind(gsaup,gsadn),FDR<0.05))
+  nsig <- rownames(subset(rbind(gsaup,gsadn),FDR<0.05))
   return(nsig)
 })
-nsigs <- unlist(nsigs)
+
 names(nsigs) <- sizes
+nsigs
+
 par( mar = c(5.1, 6.1, 4.1, 2.1) )
-barplot(nsigs,horiz=TRUE,las=1,xlab="no. significant sets")
+barplot(unlist(lapply(nsigs,length)),horiz=TRUE,las=1,xlab="no. significant sets") ; grid()
 mtext("no. probes selected", 2, 4)
 
-sigup <- rownames(subset(dm,logFC > 0 & adj.P.Val < 0.05))
-sigdn <- rownames(subset(dm,logFC < 0 & adj.P.Val < 0.05))
-
-length(sigup)
-length(sigdn)
-
+#
+FCLIM <- c(0,0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1)
+ngsig <- lapply(FCLIM,function(F) {
+  sigup <- rownames(subset(dm,logFC > F & adj.P.Val < 0.05))
+  sigdn <- rownames(subset(dm,logFC < -F & adj.P.Val < 0.05))
   gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez)
@@
   rownames(gsadn) <- paste(rownames(gsadn),"DN")
-ngsig <- nrow(subset(rbind(gsaup,gsadn),FDR<0.05))
+  ngsig <- rownames(subset(rbind(gsaup,gsadn),FDR<0.05))
+  return(ngsig)
+})
 
+names(ngsig) <- FCLIM
+ngsig
+
+par( mar = c(5.1, 6.1, 4.1, 2.1) )
+barplot(unlist(lapply(ngsig,length)),horiz=TRUE,las=1,xlab="no. significant sets",xlim=c(0,60)) ; grid()
+mtext("fold change threshold", 2, 4)
+
+sigup <- rownames(subset(dm,logFC > 0 & adj.P.Val < 0.05))
+sigdn <- rownames(subset(dm,logFC < 0 & adj.P.Val < 0.05))
 tic()

### 2023-11-17 manuscript/manuscript.Rmd
sensitivity done

@@
 
-![Figure 1. Precision, recall and F1 for enrichment tests with simulated data.](../figs/fig1.png)
+![Figure 1. Precision, recall and F1 for enrichment tests with simulated data. Gene set sizes vary between 20 and 100 genes per set.](../figs/fig1.png)
 
@@
 
-![Figure 2. Precision, recall and F1 for GSA and LAM enrichment tests over a range of group sizes and delta values. One hundred genes per set.](../figs/fig2.png)
+![Figure 2. Precision, recall and F1 for GSA and LAM enrichment tests with simulated data over a range of group sizes and delta values. One hundred genes per set.](../figs/fig2.png)
 
@@
 
-Next, GSA and LAM methods were applied to compare paired normal and cancer methylation data.
+Next, GSA and LAM methods were applied to compare paired normal and cancer methylation data from 37 lung squamous cell carcinoma patients.
+From the 839,473 probes examined, 473,572 probes were considered significant at 5% FDR.
+Of these, 119,785 probes were significantly hypermethylated and 353,787 were significantly hypomethylated.
+These significantly hypermethylated probes mapped to 16,520 unique genes, while the hypomethylated probes mapped to 24,114 genes.
+These numbers are large, considering that there are 26,219 genes represented on this array.
 
-* Chart results with 37 pax.
+LAM resulted in 358 significant Reactome pathways at FDR<0.05 while there were just 21 with GSA (*Figure 3A*).
+The overlap of significant pathways was unexpectedly small, with just five common pathways.
 
-* Show sensitivity analysis with range of samples from 3 to 30.
+Given GSA found relatively few results
 
-* LAM has better performance than GSAmeth
+Considering the number of significant probes was very large we were concerned that
 
-* Could be due to (1) filtering genes without annotations, and (2) FDR not applied for sets without members.
+We were curious about whether LAM method had a bias towards larger gene sets as the recall of this
+method is higher with larger gene sets.
+Unexpectedly, pathways identified by LAM were collectively smaller than GSA (*Figure 3B*).
+Although algorithmic accuracy cannot be inferred from the types of pathways identified in real data,
+top ranked results from LAM appeared to be more related to cell differentiation, identity and development (*Figure 3C*),
+as compared to GSA (*Figure 3D*).
 
-* Can make changes to the GSAmeth codes to fix the problems.
+![Figure 3.
+Pathway enrichment analysis comparing normal and cancer samples using LAM and GSA methods.
+(A) Euler diagrams showing the overlap in statistically significant pathways (FDR<0.05) from LAM and GSA results.
+(B) Sizes of significant gene sets found with LAM and GSA methods.
+(C) Significant LAM pathways with largest S distance, a proxy for effect size.
+(D) Top ranked significant GSA pathways ranked by fold enrichment.](../figs/fig3.png)
 
-### Comparision of AL method to existing pathway enrichment approaches
+To compare the sensitivity of these methods, sample sizes were randomly downsampled, analysed
+as above and the significant pathways were compared to the results from the full
+group of 37 patients.
+This process was repeated 50 times and the results shown in (*Figure 4*).
+Pathways that were identified in the smaller and full group were termed "consistent",
+while pathays that were identified in the smaller but not in the full group were
+termed "inconsistent".
+LAM was able to detect 300 out of 358 consistent pathways with a sample size of 10,
+while GSA detected just 4 of 21 at this sample size.
+LAM identified more inconsistent pathways than GSA, but the proportion of
+inconsistent findings was lower across all sample sizes.
+At a sample size of 10, the proportion of inconsistent pathways was 10% for LAM and
+60% for GSA.
+
+![Figure 4. Sensitivity of LAM and GSA methods with real cancer array data.](../figs/fig4.png)
+
+This result suggests that LAM has superior sensitivity to detect differentially methylated
+pathways in real cancer data.
+
+## LAM and GSA integrative analysis with RNA expression
+
+* Separate promoter and gene body integration
+
+## Exploration of methylation pathways altered by ART
+
+## Exploration of methylation pathways altered by elevated homocysteine
+
+##
 

### 2023-11-17 figs/GSE158422_simulate020.Rmd
forece use of epic chip annotation

@@
   suppressWarnings(suppressMessages({
-    gsaup3 <- gsameth(sig.cpg=pup3, all.cpg=rownames(dm3), collection=gsets_entrez)
-    gsadn3 <- gsameth(sig.cpg=pdn3, all.cpg=rownames(dm3), collection=gsets_entrez)
+    gsaup3 <- gsameth(sig.cpg=pup3, all.cpg=rownames(dm3), collection=gsets_entrez, array.type="EPIC")
+    gsadn3 <- gsameth(sig.cpg=pdn3, all.cpg=rownames(dm3), collection=gsets_entrez, array.type="EPIC")
   }))

### 2023-11-17 figs/GSE158422_rna_int.Rmd
force use of EPIC annotation for GSAmeth

@@
 
+Let's take a look at how pathway methylation associates with gene expression.
+
+Partition methylation analysis between promoter and body.
+
+Compare to GSAmeth.
+
 ```{r,packages}
@@
   library("beeswarm")
+  library("missMethyl")
   library("HGNChelper")
@@
 
-## Load annotations
+## Load annotations and methylation data
 
@@
 
-load("GSE158422_gmea.Rdata")
+anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
+myann <- data.frame(anno[,c("UCSC_RefGene_Name","Regulatory_Feature_Group","Islands_Name","Relation_to_Island")])
+gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
+gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
+names(gp2) <- rownames(gp)
+sets <- split(rep(names(gp2), lengths(gp2)), unlist(gp2))
+summary(unlist(lapply(sets,length)))
+
+design <- readRDS("GSE158422_design.rds")
+mval <- readRDS("GSE158422_mx.rds")
+dm <- readRDS("GSE158422_dm.rds")
 
@@
 
-Pathway enrichment with mitch.
+RNA expression pathway enrichment with mitch.
 
@@
 
-y <- mitch_import(dge, DEtype="deseq2")
+rna <- mitch_import(dge, DEtype="deseq2")
 
-res <- mitch_calc(y, genesets=gs_symbols, priority="effect",cores=CORES)
+res <- mitch_calc(rna, genesets=gs_symbols, priority="effect",cores=32)
 
@@
 
-```{r,overlap1}
+```{r,lamfunc}
 
-load("GSE158422_gmea.Rdata")
-
-head(mres)
-
-head(alres)
-
-head(lares)
-
-msig_up <- subset(msig,s.dist>0)$set
-msig_dn <- subset(msig,s.dist<0)$set
-
-lasig_up <- rownames(subset(lasig,t_median>0))
-lasig_dn <- rownames(subset(lasig,t_median<0))
-
-alsig_up <- rownames(subset(alsig,t_median>0))
-alsig_dn <- rownames(subset(alsig,t_median<0))
-
-v1 <- list("LA up"=lasig_up, "LA dn"=lasig_dn,
-  "RNA up"=msig_up,"RNA dn"=msig_dn)
-plot(euler(v1),quantities = TRUE)
-
-v1 <- list("AL up"=alsig_up,"AL dn"=alsig_dn,
-  "RNA up"=msig_up,"RNA dn"=msig_dn)
-plot(euler(v1),quantities = TRUE)
+# LAM aggregate
+agg <- function(dm,cores=1) {
+  dm <- dm[which(dm$UCSC_RefGene_Name!=""),]
+  gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
+  gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
+  gnl <- mclapply(gnl,unique,mc.cores=cores)
+  dm$UCSC_RefGene_Name <- gnl
+  l <- mclapply(1:nrow(dm), function(i) {
+    a <- dm[i,]
+    len <- length(a[[1]][[1]])
+    tvals <- as.numeric(rep(a["t"],len))
+    genes <- a[[1]][[1]]
+    data.frame(genes,tvals)
+  },mc.cores=cores)
+  df <- do.call(rbind,l)
+  keep <- names(which(table(df$genes)>1))
+  df <- df[df$genes %in% keep,]
+  gn <- unique(df$genes)
+  gme_res <- lapply( 1:length(gn), function(i) {
+    g <- gn[i]
+    tstats <- df[which(df$genes==g),"tvals"]
+    myn <- length(tstats)
+    mymean <- mean(tstats)
+    mymedian <- median(tstats)
+    top <- tstats[order(-abs(tstats))][1]
+    if ( length(tstats) > 2 ) {
+      ttest <- t.test(tstats)
+      pval <- ttest$p.value
+    } else {
+      pval = 1
+    }
+    res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,
+      "median"=mymedian, "top"=top, "pval"=pval)
+  } )
+  gme_res_df <- do.call(rbind, gme_res)
+  rownames(gme_res_df) <- gme_res_df[,1]
+  gme_res_df <- gme_res_df[,-1]
+  tmp <- apply(gme_res_df,2,as.numeric)
+  rownames(tmp) <- rownames(gme_res_df)
+  gme_res_df <- as.data.frame(tmp)
+  gme_res_df$sig <- -log10(gme_res_df[,"pval"])
+  gme_res_df <- gme_res_df[order(-gme_res_df$sig),]
+  gme_res_df$fdr <- p.adjust(gme_res_df$pval)
+  out <- list("df"=df,"gme_res_df"=gme_res_df)
+  return(out)
+}
 
@@
 
-## Load promoter methylation enrichment data and calculate overlap
+Here we will look at parallel analysis of gene body and promoter.
 
-```{r,overlap2}
+```{r,promoterbody}
 
-head(alpres)
+table(dm$Regulatory_Feature_Group)
 
-head(lapres)
+# all gene
+dmagg <- agg(dm=dm,cores=16)
+str(dmagg)
+head(dmagg$gme_res_df)
+dim(dm)
+dim(dmagg$gme_res_df)
+mm <- dmagg$gme_res_df[,"mean",drop=FALSE]
+mmres <- mitch_calc(mm, genesets=gs_symbols, priority="effect",cores=32)
+mmres_up <- subset(mmres$enrichment_result,s.dist>0 & p.adjustANOVA < 0.05)
+mmres_dn <- subset(mmres$enrichment_result,s.dist<0 & p.adjustANOVA < 0.05)
+nrow(mmres_up)
+nrow(mmres_dn)
+head(mmres_up,20)
+head(mmres_dn,20)
 
-lapsig_up <- rownames(subset(lapsig,t_median>0))
-lapsig_dn <- rownames(subset(lapsig,t_median<0))
+# promoter
+dmpr <- dm[grep("Promoter",dm$Regulatory_Feature_Group),]
+dmpragg <- agg(dm=dmpr,cores=16)
+str(dmpragg)
+head(dmpragg$gme_res_df)
+dim(dmpr)
+dim(dmpragg$gme_res_df)
+mmpr <- dmpragg$gme_res_df[,"mean",drop=FALSE]
+mmprres <- mitch_calc(mmpr, genesets=gs_symbols, priority="effect",cores=32)
+mmprres_up <- subset(mmprres$enrichment_result,s.dist>0 & p.adjustANOVA < 0.05)
+mmprres_dn <- subset(mmprres$enrichment_result,s.dist<0 & p.adjustANOVA < 0.05)
+nrow(mmprres_up)
+nrow(mmprres_dn)
+head(mmprres_up,20)
+head(mmprres_dn,20)
 
-alpsig_up <- rownames(subset(alpsig,t_median>0))
-alpsig_dn <- rownames(subset(alpsig,t_median<0))
+# ex-promoter
+dmx <- dm[grep("Promoter",dm$Regulatory_Feature_Group,invert=TRUE),]
+dmxagg <- agg(dm=dmx,cores=16)
+str(dmxagg)
+head(dmxagg$gme_res_df)
+dim(dmx)
+dim(dmxagg$gme_res_df)
+mmx <- dmxagg$gme_res_df[,"mean",drop=FALSE]
+mmxres <- mitch_calc(mmx, genesets=gs_symbols, priority="effect",cores=32)
+mmxres_up <- subset(mmxres$enrichment_result,s.dist>0 & p.adjustANOVA < 0.05)
+mmxres_dn <- subset(mmxres$enrichment_result,s.dist<0 & p.adjustANOVA < 0.05)
+nrow(mmxres_up)
+nrow(mmxres_dn)
+head(mmxres_up,20)
+head(mmxres_dn,20)
 
-v1 <- list("LAP up"=lapsig_up, "LAP dn"=lapsig_dn,
-  "RNA up"=msig_up,"RNA dn"=msig_dn)
-plot(euler(v1),quantities = TRUE)
+# combined methylation
+mmm <- merge(mmpr,mmx,by=0)
+rownames(mmm) <- mmm$Row.names
+mmm$Row.names = NULL
+colnames(mmm) <- c("pr","x")
+mmmres <- mitch_calc(mmm, genesets=gs_symbols, priority="effect",cores=32)
+mmmres_sig <- subset(mmmres$enrichment_result, p.adjustMANOVA < 0.05)
+head(mmmres_sig,20)
+table(sign(mmmres_sig$s.pr) == sign(mmmres_sig$s.x))
 
-v1 <- list("ALP up"=alpsig_up,"ALP dn"=alpsig_dn,
-  "RNA up"=msig_up,"RNA dn"=msig_dn)
-plot(euler(v1),quantities = TRUE)
+```
+
+## RNA expression and methylation enrichment
+
+```{r,rnaoverlap2}
+
+# three way
+mmrna <- merge(mmm,rna,by=0)
+rownames(mmrna) <- mmrna$Row.names
+mmrna$Row.names = NULL
+colnames(mmrna) <- c("pr","x","rna")
+mmrnares <- mitch_calc(mmrna, genesets=gs_symbols, priority="effect",cores=32)
+mmrnares_sig <- subset(mmrnares$enrichment_result, p.adjustMANOVA < 0.05)
+head(mmrnares_sig,20)
+table(sign(mmrnares_sig$s.pr) == sign(mmrnares_sig$s.x))
+table(sign(mmrnares_sig$s.pr) == sign(mmrnares_sig$s.rna))
+table(sign(mmrnares_sig$s.x) == sign(mmrnares_sig$s.rna))
+mtop <- head(mmrnares_sig,30)
+mtop <- mtop[,c("set","s.pr","s.x","s.rna")]
+rownames(mtop) <- gsub("REACTOME_","",mtop$set)
+rownames(mtop) <- gsub("_"," ",rownames(mtop))
+mtop$set=NULL
+colnames(mtop) <- c("promoter","non-promoter","RNA")
+colfunc <- colorRampPalette(c("blue", "white", "red"))
+heatmap.2( as.matrix(mtop) , trace="none",margins = c(6,25), col=colfunc(25), cexRow=.6, cexCol=1)
+
+# rna vs promoter
+mmrna <- merge(mmpr,rna,by=0)
+rownames(mmrna) <- mmrna$Row.names
+mmrna$Row.names = NULL
+colnames(mmrna) <- c("pr","rna")
+mmrnares <- mitch_calc(mmrna, genesets=gs_symbols, priority="effect",cores=32)
+mmrnares_sig <- subset(mmrnares$enrichment_result, p.adjustMANOVA < 0.05)
+head(mmrnares_sig,20)
+table(sign(mmrnares_sig$s.pr) == sign(mmrnares_sig$s.rna))
+mmrnares_sig <- subset(mmrnares$enrichment_result, p.adjustMANOVA < 0.05)
+mylm <- lm(mmrnares$enrichment_result$s.rna ~ mmrnares$enrichment_result$s.pr)
+summary(mylm)
+mylm2 <- lm(mmrnares_sig$s.rna ~ mmrnares_sig$s.pr)
+summary(mylm2)
+plot(mmrnares$enrichment_result$s.pr,mmrnares$enrichment_result$s.rna,xlab="promoter",ylab="RNA",main="S dist",col="gray",pch=19)
+abline(v=0,h=0,lty=2,lwd=2)
+points(mmrnares_sig$s.pr,mmrnares_sig$s.rna,col="black",pch=19,cex=1.01)
+abline(mylm,col="blue",lty=3,lwd=3)
+abline(mylm2,col="red",lty=3,lwd=3)
+
+# rna vs expromoter
+mmrna <- merge(mmx,rna,by=0)
+rownames(mmrna) <- mmrna$Row.names
+mmrna$Row.names = NULL
+colnames(mmrna) <- c("x","rna")
+mmrnares <- mitch_calc(mmrna, genesets=gs_symbols, priority="effect",cores=32)
+mmrnares_sig <- subset(mmrnares$enrichment_result, p.adjustMANOVA < 0.05)
+head(mmrnares_sig,20)
+table(sign(mmrnares_sig$s.x) == sign(mmrnares_sig$s.rna))
+mmrnares_sig <- subset(mmrnares$enrichment_result, p.adjustMANOVA < 0.05)
+mylm <- lm(mmrnares$enrichment_result$s.rna ~ mmrnares$enrichment_result$s.x)
+summary(mylm)
+mylm2 <- lm(mmrnares_sig$s.rna ~ mmrnares_sig$s.x)
+summary(mylm2)
+plot(mmrnares$enrichment_result$s.x,mmrnares$enrichment_result$s.rna,xlab="ex-promoter",ylab="RNA",main="S dist",col="gray",pch=19)
+abline(v=0,h=0,lty=2,lwd=2)
+points(mmrnares_sig$s.x,mmrnares_sig$s.rna,col="black",pch=19,cex=1.01)
+abline(mylm,col="blue",lty=3,lwd=3)
+abline(mylm2,col="red",lty=3,lwd=3)
+
+```
+
+Now look at GSAmeth.
+
+```{r,rnagsa}
+
+# all meth
+sigup <- rownames(subset(dm,logFC > 0 & adj.P.Val < 0.05))
+sigdn <- rownames(subset(dm,logFC < 0 & adj.P.Val < 0.05))
+gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez, array.type="EPIC")
+rownames(gsaup) <- paste(rownames(gsaup),"UP")
+gsadn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez, array.type="EPIC")
+rownames(gsadn) <- paste(rownames(gsadn),"DN")
+gsig_up <- subset(gsaup,FDR<0.05)
+gsig_dn <- subset(gsadn,FDR<0.05)
+gsig <- rbind(gsig_up,gsig_dn)
+nrow(gsig)
+
+# all meth
+sigup <- rownames(subset(dm,logFC > 0 & adj.P.Val < 0.05))
+sigdn <- rownames(subset(dm,logFC < 0 & adj.P.Val < 0.05))
+gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez, array.type="EPIC")
+rownames(gsaup) <- paste(rownames(gsaup),"UP")
+gsadn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez, array.type="EPIC")
+rownames(gsadn) <- paste(rownames(gsadn),"DN")
+gsig_up <- subset(gsaup,FDR<0.05)
+gsig_dn <- subset(gsadn,FDR<0.05)
+gsig <- rbind(gsig_up,gsig_dn) 
+nrow(gsig)
 

### 2023-11-17 figs/GSE158422_sensitivity.Rmd
force use of EPIC annotation for GSAmeth

@@
 nrow(subset(dm,adj.P.Val<0.05 & logFC<0))
+saveRDS(dm,"GSE158422_dm.rds")
 
@@
   dmsigdn <- rownames(subset(dmsig,logFC<0))
-  gsaup <- gsameth(sig.cpg=dmsigup, all.cpg=rownames(dm), collection=gs_entrez)
+  gsaup <- gsameth(sig.cpg=dmsigup, all.cpg=rownames(dm), collection=gs_entrez, array.type="EPIC")
   rownames(gsaup) <- paste(rownames(gsaup),"UP")
-  gsadn <- gsameth(sig.cpg=dmsigdn, all.cpg=rownames(dm), collection=gs_entrez)
+  gsadn <- gsameth(sig.cpg=dmsigdn, all.cpg=rownames(dm), collection=gs_entrez, array.type="EPIC")
   rownames(gsadn) <- paste(rownames(gsadn),"DN")
@@
   sigdn <- rownames(subset(dm,logFC < -F & adj.P.Val < 0.05))
-  gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez)
+  gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez, array.type="EPIC")
   rownames(gsaup) <- paste(rownames(gsaup),"UP")
-  gsadn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez)
+  gsadn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez, array.type="EPIC")
   rownames(gsadn) <- paste(rownames(gsadn),"DN")
@@
 tic()
-gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez)
+gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez, array.type="EPIC")
 rownames(gsaup) <- paste(rownames(gsaup),"UP")
-gsadn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez)
+gsadn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez, array.type="EPIC")
 rownames(gsadn) <- paste(rownames(gsadn),"DN")
@@
       if ( length(sigdn2) < 250 ) { sigdn2 <- head(rownames(subset(dm2,logFC<0 & UCSC_RefGene_Name != "" )),250) }
-      gsaup2 <- gsameth(sig.cpg=sigup2, all.cpg=rownames(dm2), collection=gs_entrez)
+      gsaup2 <- gsameth(sig.cpg=sigup2, all.cpg=rownames(dm2), collection=gs_entrez, array.type="EPIC")
       rownames(gsaup2) <- paste(rownames(gsaup2),"UP")
-      gsadn2 <- gsameth(sig.cpg=sigdn2, all.cpg=rownames(dm2), collection=gs_entrez)
+      gsadn2 <- gsameth(sig.cpg=sigdn2, all.cpg=rownames(dm2), collection=gs_entrez, array.type="EPIC")
       rownames(gsadn2) <- paste(rownames(gsadn2),"DN")

### 2023-11-17 figs/GSE158422_rna_int.Rmd
use a better annotation set

@@
 Partition methylation analysis between promoter and body.
+
+Will need to do this at the annotation level.
 
@@
 anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
-myann <- data.frame(anno[,c("UCSC_RefGene_Name","Regulatory_Feature_Group","Islands_Name","Relation_to_Island")])
-gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
-gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
+myann <- data.frame(anno[,c("GencodeCompV12_NAME","GencodeCompV12_Group","Islands_Name","Relation_to_Island")])
+gp <- myann[,"GencodeCompV12_NAME",drop=FALSE]
+gp2 <- strsplit(gp$GencodeCompV12_NAME,";")
 names(gp2) <- rownames(gp)
@@
 agg <- function(dm,cores=1) {
-  dm <- dm[which(dm$UCSC_RefGene_Name!=""),]
-  gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
-  gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
+  dm <- dm[which(dm$GencodeCompV12_NAME!=""),]
+  gn <- unique(unlist(strsplit( dm$GencodeCompV12_NAME ,";")))
+  gnl <- strsplit( dm$GencodeCompV12_NAME ,";")
   gnl <- mclapply(gnl,unique,mc.cores=cores)
-  dm$UCSC_RefGene_Name <- gnl
+  dm$GencodeCompV12_NAME <- gnl
   l <- mclapply(1:nrow(dm), function(i) {

### 2023-11-17 figs/GSE158422_sensitivity.Rmd
use a better annotation set

@@
 anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
-myann <- data.frame(anno[,c("UCSC_RefGene_Name","Regulatory_Feature_Group","Islands_Name","Relation_to_Island")])
+myann <- data.frame(anno[,c("GencodeCompV12_NAME","Regulatory_Feature_Group","Islands_Name","Relation_to_Island")])
 
-gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
-gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
+gp <- myann[,"GencodeCompV12_NAME",drop=FALSE]
+gp2 <- strsplit(gp$GencodeCompV12_NAME,";")
 names(gp2) <- rownames(gp)
@@
 
-length(unique(unlist(strsplit(subset(dm,adj.P.Val<0.05 & logFC>0)$UCSC_RefGene_Name,";"))))
-length(unique(unlist(strsplit(subset(dm,adj.P.Val<0.05 & logFC<0)$UCSC_RefGene_Name,";"))))
+length(unique(unlist(strsplit(subset(dm,adj.P.Val<0.05 & logFC>0)$GencodeCompV12_NAME,";"))))
+length(unique(unlist(strsplit(subset(dm,adj.P.Val<0.05 & logFC<0)$GencodeCompV12_NAME,";"))))
 
-length(unique(unlist(strsplit(dm$UCSC_RefGene_Name,";"))))
+length(unique(unlist(strsplit(dm$GencodeCompV12_NAME,";"))))
 
@@
 
-detgenes <- unique(unlist(strsplit(dm$UCSC_RefGene_Name,";")))
+detgenes <- unique(unlist(strsplit(dm$GencodeCompV12_NAME,";")))
 reactomegenes <- unique(unlist(gs_symbols))
@@
 BGSIZE=length(BG)
-FGUP <- unique(unlist(strsplit(subset(dm,logFC > 0 & adj.P.Val < 0.05)$UCSC_RefGene_Name,";")))
+FGUP <- unique(unlist(strsplit(subset(dm,logFC > 0 & adj.P.Val < 0.05)$GencodeCompV12_NAME,";")))
 FGUP <- FGUP[which(FGUP %in% reactomegenes)]
-FGDN <- unique(unlist(strsplit(subset(dm,logFC < 0 & adj.P.Val < 0.05)$UCSC_RefGene_Name,";")))
+FGDN <- unique(unlist(strsplit(subset(dm,logFC < 0 & adj.P.Val < 0.05)$GencodeCompV12_NAME,";")))
 FGDN <- FGDN[which(FGDN %in% reactomegenes)]
@@
     } else {
-      sigup2 <- rownames(subset(dm2,adj.P.Val<0.05 & UCSC_RefGene_Name != "" & logFC>0))
-      if ( length(sigup2) < 250 ) { sigup2 <- head(rownames(subset(dm2,logFC>0 & UCSC_RefGene_Name != "" )),250) }
-      sigdn2 <- rownames(subset(dm2,adj.P.Val<0.05 & UCSC_RefGene_Name != "" & logFC<0))
-      if ( length(sigdn2) < 250 ) { sigdn2 <- head(rownames(subset(dm2,logFC<0 & UCSC_RefGene_Name != "" )),250) }
+      sigup2 <- rownames(subset(dm2,adj.P.Val<0.05 & GencodeCompV12_NAME != "" & logFC>0))
+      if ( length(sigup2) < 250 ) { sigup2 <- head(rownames(subset(dm2,logFC>0 & GencodeCompV12_NAME != "" )),250) }
+      sigdn2 <- rownames(subset(dm2,adj.P.Val<0.05 & GencodeCompV12_NAME != "" & logFC<0))
+      if ( length(sigdn2) < 250 ) { sigdn2 <- head(rownames(subset(dm2,logFC<0 & GencodeCompV12_NAME != "" )),250) }
       gsaup2 <- gsameth(sig.cpg=sigup2, all.cpg=rownames(dm2), collection=gs_entrez, array.type="EPIC")
@@
 
-detgenes <- unique(unlist(strsplit(dm$UCSC_RefGene_Name,";")))
+detgenes <- unique(unlist(strsplit(dm$GencodeCompV12_NAME,";")))
 

### 2023-11-17 figs/GSE158422_simulate020.Rmd
use a better annotation set

--- 2023-11-02 figs/GSE158422_simulate020.Rmd
@@
 anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
-myann <- data.frame(anno[,c("UCSC_RefGene_Name","Regulatory_Feature_Group","Islands_Name","Relation_to_Island")])
-
-gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
-gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
+myann <- data.frame(anno[,c("GencodeCompV12_NAME","Regulatory_Feature_Group","Islands_Name","Relation_to_Island")])
+
+gp <- myann[,"GencodeCompV12_NAME",drop=FALSE]
+gp2 <- strsplit(gp$GencodeCompV12_NAME,";")
 names(gp2) <- rownames(gp)
@@
 
-gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
+gene_catalog <- unique(unlist(strsplit(myann$GencodeCompV12_NAME,";")))
 
@@
   # make probe-gene vector
-  probe2gene <- strsplit(myann$UCSC_RefGene_Name,";")
+  probe2gene <- strsplit(myann$GencodeCompV12_NAME,";")
   names(probe2gene) <- rownames(myann)
@@
   # generate gene sets
-  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
+  gene_catalog <- unique(unlist(strsplit(myann$GencodeCompV12_NAME,";")))
   lengths <- unname(unlist(lapply(genesetdatabase,length)))
@@
 agg <- function(dm,cores=1) {
-  gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
-  gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
+  gn <- unique(unlist(strsplit( dm$GencodeCompV12_NAME ,";")))
+  gnl <- strsplit( dm$GencodeCompV12_NAME ,";")
   gnl <- mclapply(gnl,unique,mc.cores=cores)
-  dm$UCSC_RefGene_Name <- gnl
+  dm$GencodeCompV12_NAME <- gnl
   l <- mclapply(1:nrow(dm), function(i) {
@@
 geoagg <- function(dm,cores=1) {
-  gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
-  gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
+  gn <- unique(unlist(strsplit( dm$GencodeCompV12_NAME ,";")))
+  gnl <- strsplit( dm$GencodeCompV12_NAME ,";")
   gnl <- mclapply(gnl,unique,mc.cores=cores)
-  dm$UCSC_RefGene_Name <- gnl
+  dm$GencodeCompV12_NAME <- gnl
   l <- mclapply(1:nrow(dm), function(i) {
@@
 magg <- function(mval,myann,cores=1){
-  gn <- unique(unlist(strsplit( myann$UCSC_RefGene_Name ,";")))
-  gnl <- strsplit( myann$UCSC_RefGene_Name ,";")
+  gn <- unique(unlist(strsplit( myann$GencodeCompV12_NAME ,";")))
+  gnl <- strsplit( myann$GencodeCompV12_NAME ,";")
   gnl <- mclapply(gnl,unique,mc.cores=cores)
   myann$gnl <- gnl
-  keep <- rownames(subset(myann,UCSC_RefGene_Name!=""))
+  keep <- rownames(subset(myann,GencodeCompV12_NAME!=""))
   mx <- mval[rownames(mval) %in% keep,]
@@
   annodf <- as.data.frame(anno)
-  keep <- rownames(subset(myann,UCSC_RefGene_Name!=""))
+  keep <- rownames(subset(myann,GencodeCompV12_NAME!=""))
   mx <- mval[rownames(mval) %in% keep,]

### 2023-11-20 figs/GSE158422_pkgdev.Rmd
a walkthrough for infinium limma users

---
title: "GSE158422: Package Development"
author: "The GMEA team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
---

## Introduction

Here we are looking at putting together the easiest possible workflow for the package.

## Requirements

```{r,packages}

suppressPackageStartupMessages({
  library("limma")
  library("eulerr")
  library("IlluminaHumanMethylation450kmanifest")
  library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
  library("tictoc")
  library("mitch")
  library("kableExtra")
  library("beeswarm")
  library("missMethyl")
  library("gridExtra")
  library("png")
})

CORES=24

```

## Load data

Reactome pathways were downloaded on the 14th Sept 2023 from MsigDB.

```{r,load1}

if (!file.exists("GSE158422_design.rds")) {
  download.file("https://ziemann-lab.net/public/gmea_prototype/GSE158422_design.rds", "GSE158422_design.rds")
}
design <- readRDS("GSE158422_design.rds")

if (!file.exists("GSE158422_design.rds")) {
 download.file("https://ziemann-lab.net/public/gmea_prototype/GSE158422_mx.rds","GSE158422_mx.rds")
}
mval <- readRDS("GSE158422_mx.rds")

gs_symbols <- gmt_import("c2.cp.reactome.v2023.1.Hs.symbols.gmt")

```

## Curate the annotation

Use all probes on the chip.

```{r,anno1}

tic()
anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
myann <- data.frame(anno[,c("GencodeCompV12_NAME","GencodeCompV12_Group","Islands_Name","Relation_to_Island")])
gp <- myann[,"GencodeCompV12_NAME",drop=FALSE]
gp2 <- strsplit(gp$GencodeCompV12_NAME,";")
names(gp2) <- rownames(gp)
gp2 <- lapply(gp2,unique)
gt <- stack(gp2)
colnames(gt) <- c("gene","probe")
dim(gt)
str(gt)
toc() #9.0s

```

# AL GMEA: all probes

Start with curating the sample info and design matrix.

```{r,samplesheet}

sex <- as.data.frame(design)$sex
tumor <- as.data.frame(design)$tumor
patient <- as.character(unlist(lapply(1:ncol(mval),function(i) {c(i,i)})))
patient <- head(patient,ncol(mval))
design <- model.matrix(~ patient + tumor )

```

## Prep limma

```{r,limma1}

tic()
fit.reduced <- lmFit(mval,design)
fit.reduced <- eBayes(fit.reduced)
dm <- topTable(fit.reduced,coef=ncol(design), number = Inf)
head(dm)
toc() #25.53, 25.50, 25.51

```

## Merge gene name and aggregate

This only works properly when mitch's mapGeneIds function uses `mean` to aggregate (line 69 of mitch.R).

```{r,mg1}

# mitch aggregate function needs to be changes to `mean`
tic()
m <- mitch_import(x=dm,DEtype="limma",geneTable=gt)
toc() #9.58s, 9.71, 9.69
head(m)

```

This is the alternative aggregation approach which is faster but more complicated.

```{r,mg2,eval=FALSE}

tic()
anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
myann <- data.frame(anno[,c("GencodeCompV12_NAME","GencodeCompV12_Group","Islands_Name","Relation_to_Island")])
gp <- myann[,"GencodeCompV12_NAME",drop=FALSE]
gp2 <- strsplit(gp$GencodeCompV12_NAME,";")
names(gp2) <- rownames(gp)
gp2 <- lapply(gp2,unique)
gt <- stack(gp2)
colnames(gt) <- c("gene","probe")
dim(gt)
str(gt)
toc() #2.44, 2.41, 2.41s

```

##  Run mitch analysis.

```{r,mitch1}

tic()
mres <- mitch_calc(x=m,genesets=gs_symbols,minsetsize=5,cores=16, priority="effect")
toc() #10.92, 10.733, 10.91

mtable <- mres$enrichment_result
up <- subset(mtable,s.dist>0 & p.adjustANOVA<0.05)
dn <- subset(mtable,s.dist<0 & p.adjustANOVA<0.05)
nrow(up)
nrow(dn)
head(up,10)
head(dn,10)

top <- rbind(up[1:10,],dn[1:10,])
top <- top[order(top$s.dist),]

barnames <- gsub("_"," ",gsub("REACTOME_","",top$set))
cols <- as.character(((sign(top$s.dist)+1)/2)+1)
cols <- gsub("1","blue",gsub("2","red",cols))

par(mar = c(5.1, 29.1, 4.1, 2.1))
barplot(abs(top$s.dist),horiz=TRUE,las=1,names.arg=barnames, cex.names=0.8,
  cex.axis=0.8,col=cols, xlab="Enrichment score",main="Reactomes")
mtext("whole gene")
grid()

par( mar = c(5.1, 4.1, 4.1, 2.1) )

```

Make a html report and some charts.

```{r,report1}

mitch_report(res=mres,outfile="GSE158422_pkgdev_mitchreport.html",overwrite=TRUE)
mitch_plot2(res=mres,outfile="GSE158422_pkgdev_mitchcharts.pdf")

```

## Promoter only analysis

Now do the same but filter for promoters.
Using the GencodeCompV12_NAME field of the annotation and using TSS probes only.

```{r,promo}

head(myann)
table(anno$Regulatory_Feature_Group)
table(unlist(strsplit(myann$GencodeCompV12_Group,";")))

# promoter prep
tic()
keep=c("TSS200","TSS1500")
annos <- strsplit(myann$GencodeCompV12_Group,";")
genes <- strsplit(myann$GencodeCompV12_NAME,";")
keeps <- lapply(1:nrow(myann), function(i) {
  unique(genes[[i]][annos[[i]] %in% keep])
})
names(keeps) <- rownames(myann)
#gt <- as.data.frame((unlist(gp2)))
gt2 <- stack(gp2)
colnames(gt2) <- c("gene","probe")
toc() #20.37, 19.56, 21.23

tic()
m2 <- mitch_import(x=dm,DEtype="limma",geneTable=gt2)
toc() #9.68, 9.78, 9.62
head(m)
dim(m)

tic()
mres <- mitch_calc(x=m2,genesets=gs_symbols,minsetsize=5,cores=16, priority="effect")
toc() #11.12, 10.99, 11.01

mtable <- mres$enrichment_result
up <- subset(mtable,s.dist>0 & p.adjustANOVA<0.05)
dn <- subset(mtable,s.dist<0 & p.adjustANOVA<0.05)
nrow(up)
nrow(dn)
head(up,10)
head(dn,10)

top <- rbind(up[1:10,],dn[1:10,])
top<- top[order(top$s.dist),]

barnames <- gsub("_"," ",gsub("REACTOME_","",top$set))
cols <- as.character(((sign(top$s.dist)+1)/2)+1)
cols <- gsub("1","blue",gsub("2","red",cols))

par(mar = c(5.1, 29.1, 4.1, 2.1))
barplot(abs(top$s.dist),horiz=TRUE,las=1,names.arg=barnames, cex.names=0.8,
  cex.axis=0.8,col=cols, xlab="Enrichment score",main="Reactomes")
mtext("promoter only")
grid()

par( mar = c(5.1, 4.1, 4.1, 2.1) )

```

## Session information

```{r,save}

save.image("GSE158422_pkgdev.Rdata")

sessionInfo()

```

### 2023-11-21 figs/GSE158422_GSAfix.Rmd
Prefer UCSC annotations

--- 2023-11-17 figs/GSE158422_sensitivity.Rmd
@@
 ---
-title: "GSE158422: DNA methylation in Lung squamous cell carcinoma - gene methylation enrichment analysis - sensitivity analysis"
+title: "Can GSAmeth be fixed?"
 author: "The GMEA team"
@@
 
-In the previous analysis, it was shown that the limma-average-mitch (LAM) method
-yielded good performance with simulated data.
+In the previous analysis, it was shown that sensitivity of GSAmeth was poor compared to LAM with real data.
+This was surprising as GSAmeth performed well with simulated data.
 
-In this piece of work, we want to ascertain the sensitivity of this approach as compared to
-the existing method, GSAmeth.
+Possible explanations to this:
 
-To achieve this we will analyse with the full 37 patient sample pairs, and repeatedly downsample
-the dataset down to n=3 patients.
-What we expect is that LAM might have a higher degree of recall as compared to GSAmeth.
-We expect LAM will identify gene sets with larger effect sizes (enrichment scores) using
-fewer samples.
-This means that for a given set of datasets, this approach could yield more findings, or
-alternatively if pathway analysis is the primary outcome of a study, it need not be as large, which
-can save money and difficulty with establishing large sample collections.
+* GSAmeth filters out genes that do not have annotations.
+Ideally this filtering should be deactivated.
+
+* GSAmeth only performs FDR correction on gene sets where a minimum number of genes in the foreground.
+Ideally, FDR correction should be applied to genesets with a minimum number of genes in the background.
+
+Let's see if we can fix it.
 
 ## Requirements
-
-This analysis was run on a 16C/32T computer with 128GB RAM running at 4.2 GHz.
-RAM usage can be moderated by reducing the parallel core count.
-
-This analysis was run on an AMD Ryzen Threadripper PRO 5955WX 16-Cores with 3200 MT/s DDR4.
 
@@
   library("png")
+  source("gsameth_fix.R")
 })
@@
 anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
-myann <- data.frame(anno[,c("UCSC_RefGene_Name","Regulatory_Feature_Group","Islands_Name","Relation_to_Island")])
+myann <- data.frame(anno[,c("UCSC_RefGene_Name","UCSC_RefGene_Group","Islands_Name","Relation_to_Island")])
 
@@
 
-par(mfrow=c(3,1))
-
-sig <- head(subset(dm,adj.P.Val < 0.05 ),100)
-sigmval <- mval[rownames(mval) %in% rownames(sig),]
-es <- lapply(1:nrow(sigmval), function(i) {
-  x <- sigmval[i,]
-  mx <- matrix(x,ncol=2)
-  cancer <- median(mx[,1])
-  normal <- median(mx[,2])
-  abs(cancer-normal)
-} )
-hist(unlist(es),main="top 100 probes", xlab="probe effect size")
-summary(unlist(es))
-
-sig <- head(subset(dm,adj.P.Val < 0.05 ),1000)
-sig <- sig[101:1000,]
-sigmval <- mval[rownames(mval) %in% rownames(sig),]
-es <- lapply(1:nrow(sigmval), function(i) {
-  x <- sigmval[i,]
-  mx <- matrix(x,ncol=2)
-  cancer <- median(mx[,1])
-  normal <- median(mx[,2])
-  abs(cancer-normal)
-} )
-hist(unlist(es),main="top 1000 probes ex top 100", xlab="probe effect size")
-summary(unlist(es))
-
-sig <- head(subset(dm,adj.P.Val < 0.05 ),10000)
-sig <- sig[1001:10000,]
-sigmval <- mval[rownames(mval) %in% rownames(sig),]
-es <- lapply(1:nrow(sigmval), function(i) {
-  x <- sigmval[i,]
-  mx <- matrix(x,ncol=2)
-  cancer <- median(mx[,1])
-  normal <- median(mx[,2])
-  abs(cancer-normal)
-} )
-hist(unlist(es), main="top 10000 probes ex top 1000", xlab="probe effect size")
-summary(unlist(es))
-
-par(mfrow=c(1,1))
-
-```
-
-## Run LAM
-
-Run LAM method on tumour-normal methylation data.
-Record timings.
-
-```{r,rungmea}
-
-tic()
-dmagg <- agg(dm,cores=32)
-m <- dmagg$gme_res_df[,"mean",drop=FALSE]
-lamres <- runmitch(m=m,genesets=gs_symbols,cores=32)
-toc()
-
-#tic()
-#dmagg <- agg(dm,cores=16)
-#m <- dmagg$gme_res_df[,"mean",drop=FALSE]
-#lamres <- runmitch(m=m,genesets=gs_symbols,cores=16)
-#toc()
-#c32<-c()
-#c16<-c(123.666,129.619)
-#c8<-c()
-#c4<-c()
-#c2<-c()
-#c1<-c()
-#duration <- list("c1"=c1, "c2"=c2, "c4"=c4, "c8"=c8, "c16"=c16,"c32"=c32)
-
-top %>%
-  kbl(caption="Top significant probes") %>%
-  kable_paper("hover", full_width = F)
-
-lamsig <- subset(lamres,p.adjustANOVA<0.05)
-
-top <- head(lamsig,20)
-top <- top[order(top$s.dist),]
-
-top %>%
-  kbl(caption="Top gene sets (biggest effect size with FDR<0.05)") %>%
-  kable_paper("hover", full_width = F)
-
-barnames <- gsub("_"," ",gsub("REACTOME_","",rownames(top)))
-cols <- gsub("TRUE","red",gsub("FALSE","blue",as.character(top$s.dist>0)))
-
-par(mar = c(5.1, 29.1, 4.1, 2.1))
-barplot(abs(top$s.dist),horiz=TRUE,las=1,names.arg=barnames, cex.names=0.65,
-  cex.axis=0.8,col=cols, xlab="Enrichment score",main="Reactomes")
-grid()
-
-par( mar = c(5.1, 4.1, 4.1, 2.1) )
-
-hist(lamsig$s.dist)
-
-# distinguish up (TRUE) and down (FALSE)
-lamsigsets <- paste(rownames(lamsig),as.character(lamsig$s.dist > 0))
-lamsigsets2 <- gsub(" TRUE"," UP",lamsigsets)
-lamsigsets2 <- gsub(" FALSE"," DN",lamsigsets2)
-
-```
-
-## Downsample LAM test
-
-This will run the LAM analysis at group sizes of 2, 3, 6, 10, 16, 22 and 30,
-50 times each, with pseudorandom seeds from 100 to 5000 jumping up by 100.
-
-```{r,al_down1,fig.height=10}
-
-SAMPLESIZES=c(2,3,6,10,16,22,30)
-ssres <- lapply(SAMPLESIZES,function(n) {
-  SEEDS <- seq(100,5000,100)
-  dsres <- mclapply(SEEDS, function(SEED) {
-    set.seed(SEED)
-    mysample <- sample(1:37,n,replace=FALSE)*2
-    mysample <- c(mysample,(mysample-1))
-    mysample <- mysample[order(mysample)]
-    design2 <- design[mysample,]
-    design2 <- design2[,colSums(design2)>0]
-    mval2 <- mval[,mysample]
-    dm2 <- runlimma(mval=mval2,design=design2,myann=myann)
-    dmagg2 <- agg(dm2,cores=2)
-    m2 <- dmagg2$gme_res_df[,"mean",drop=FALSE]
-    lamres2 <- runmitch(m=m2,genesets=gs_symbols,cores=2)
-    lamsig2 <- subset(lamres2,p.adjustANOVA<0.05)
-    lamsigsets2 <- paste(rownames(lamsig2),as.character(lamsig2$s.dist > 0))
-    TOT=length(lamsigsets2)
-    TP=length(which(lamsigsets2 %in% lamsigsets))
-    FP=length(which(!lamsigsets2 %in% lamsigsets))
-    FN=length(which(!lamsigsets %in% lamsigsets2))
-    RES=c("TOT"=TOT,"TP"=TP,"FP"=FP,"FN"=FN)
-    RES
-  },mc.cores=5)
-  do.call(rbind,dsres)
-})
-
-par(mfrow=c(3,1))
-
-dots <- lapply(ssres,function(x) { x[,"TP"] })
-boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. true positives",xlab="sample size",cex=0,main="LAM") ; grid()
-beeswarm(dots,add=TRUE,pch=1,cex=0.6)
-abline(h=nrow(lamsig))
-
-dots <- lapply(ssres,function(x) { x[,"FP"] })
-boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. false positives",xlab="sample size",cex=0) ; grid()
-beeswarm(dots,add=TRUE,pch=1,cex=0.6)
-
-dots <- lapply(ssres,function(x) { x[,"TP"] })
-TPdots <- do.call(rbind,dots)
-dots <- lapply(ssres,function(x) { x[,"FP"] })
-FPdots <- do.call(rbind,dots)
-ratio <- t(FPdots / (TPdots + FPdots))
-ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
-boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion FP",xlab="sample size",cex=0) ; grid()
-beeswarm(ratio,add=TRUE,pch=1,cex=0.6)
-
-par(mfrow=c(1,1))
-
 ```
@@
 
+# the number of DM probes impact the results drastically
 sizes=c(1000,5000,10000,20000,50000,100000,200000,300000,350000,400000,450000,500000)
@@
 
-#
+# the fold change threshold also impacts the results
 FCLIM <- c(0,0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1)
@@
 sigdn <- rownames(subset(dm,logFC < 0 & adj.P.Val < 0.05))
+
 tic()
@@
 
-Only 4 hyper and 1 hypomethylated gene set were identified.
-This is surprising as there were 1299 hypermethylated probes corresponding to 808 genes, and
-1016 hypomethylated probes corresponding to 657 genes.
+Unload missmethyl install new one with fix.
 
-Calc GSA enrichment scores.
-First calculate the background which is the detected genes with annotations.
+```{r,gsafgix}
 
-```{r,GSAes}
+detach("package:missMethyl", unload=TRUE)
+install.packages("../pkgs/missMethyl", repos = NULL, type = "source")
+library("missMethyl")
 
-detgenes <- unique(unlist(strsplit(dm$UCSC_RefGene_Name,";")))
-reactomegenes <- unique(unlist(gs_symbols))
-length(reactomegenes)
-BG=detgenes[(which(detgenes %in% reactomegenes))]
-BGSIZE=length(BG)
-FGUP <- unique(unlist(strsplit(subset(dm,logFC > 0 & adj.P.Val < 0.05)$UCSC_RefGene_Name,";")))
-FGUP <- FGUP[which(FGUP %in% reactomegenes)]
-FGDN <- unique(unlist(strsplit(subset(dm,logFC < 0 & adj.P.Val < 0.05)$UCSC_RefGene_Name,";")))
-FGDN <- FGDN[which(FGDN %in% reactomegenes)]
+gsafixup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez)
+rownames(gsafixup) <- paste(rownames(gsafixup),"UP")
+gsafixdn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez)
+rownames(gsafixdn) <- paste(rownames(gsafixdn),"DN")
 
-myes <- lapply(1:nrow(gsig), function(i) {
-  set=rownames(gsig)[i]
-  set1=sapply(strsplit(set," "),"[[",1)
-  setg <- gs_symbols[[which(names(gs_symbols)==set1)]]
-  setg <- setg[which(setg %in% detgenes)]
-  SETSIZE=length(setg)
-  NUMUP = (length(intersect(FGUP,setg))/length(FGUP))
-  NUMDN = (length(intersect(FGDN,setg))/length(FGDN))
-  NUM=max(NUMUP,NUMDN)
-  DEN = (SETSIZE/BGSIZE)
-  ES=NUM/DEN
-  res=c("numerator"=NUM,"denominator"=DEN,"ES"=ES)
-  return(res)
+gsafixup <- gsafixup[order(gsafixup$P.DE),]
+gsafixdn <- gsafixdn[order(gsafixdn$P.DE),]
+
+gsigfix_up <- subset(gsafixup,FDR<0.05)
+head(gsigfix_up,20) %>% kbl(caption="Hypermethylated pathways (fix)") %>% kable_paper("hover", full_width = F)
+
+gsigfix_dn <- subset(gsafixdn,FDR<0.05)
+head(gsigfix_dn,20) %>% kbl(caption="Hypomethylated pathways (fix)") %>% kable_paper("hover", full_width = F)
+
+gsigfix <- rbind(gsigfix_up,gsigfix_dn)
+
+gsigfixsets <- rownames(gsigfix)
+
+# the number of DM probes impact the results drastically
+sizes=c(1000,5000,10000,20000,50000,100000,200000,300000,350000,400000,450000,500000)
+nsigs <- lapply(sizes, function(z) {
+  dmsig <- head(dm,z)
+  dmsigup <- rownames(subset(dmsig,logFC>0))
+  dmsigdn <- rownames(subset(dmsig,logFC<0))
+  gsaup <- gsameth(sig.cpg=dmsigup, all.cpg=rownames(dm), collection=gs_entrez)
+  rownames(gsaup) <- paste(rownames(gsaup),"UP")
+  gsadn <- gsameth(sig.cpg=dmsigdn, all.cpg=rownames(dm), collection=gs_entrez)
+  rownames(gsadn) <- paste(rownames(gsadn),"DN")
+  nsig <- rownames(subset(rbind(gsaup,gsadn),FDR<0.05))
+  return(nsig)
 })
 
-myes <- data.frame(do.call(rbind,myes))
-gsiges <- data.frame(gsig,myes)
-gsiges <- head(gsiges[order(-gsiges$ES),],20)
-gsiges <- gsiges[order(gsiges$ES),]
+names(nsigs) <- sizes
+nsigs
 
-barnames <- gsub("_"," ",gsub("REACTOME_","",rownames(gsiges)))
-cols <- gsub("FALSE","blue",gsub("TRUE","red",grepl(" UP",barnames)))
-barnames <- gsub("DN","",gsub(" UP","",barnames))
+par( mar = c(5.1, 6.1, 4.1, 2.1) )
+barplot(unlist(lapply(nsigs,length)),horiz=TRUE,las=1,xlab="no. significant sets") ; grid()
+mtext("no. probes selected", 2, 4)
 
-par(mar = c(5.1, 27.1, 4.1, 2.1))
-barplot(abs(gsiges$ES),horiz=TRUE,las=1,names.arg=barnames, cex.names=0.75,
-  cex.axis=0.85, col=cols, xlab="Enrichment score",main="GSA")
-grid()
+# the fold change threshold also impacts the results
+FCLIM <- c(0,0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1)
+ngsig <- lapply(FCLIM,function(F) {
+  sigup <- rownames(subset(dm,logFC > F & adj.P.Val < 0.05))
+  sigdn <- rownames(subset(dm,logFC < -F & adj.P.Val < 0.05))
+  gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez)
+  rownames(gsaup) <- paste(rownames(gsaup),"UP")
+  gsadn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez)
+  rownames(gsadn) <- paste(rownames(gsadn),"DN")
+  ngsig <- rownames(subset(rbind(gsaup,gsadn),FDR<0.05))
+  return(ngsig)
+})
+
+names(ngsig) <- FCLIM
+ngsig
+
+par( mar = c(5.1, 6.1, 4.1, 2.1) )
+barplot(unlist(lapply(ngsig,length)),horiz=TRUE,las=1,xlab="no. significant sets",xlim=c(0,60)) ; grid()
+mtext("fold change threshold", 2, 4)
 
 ```
-
-## GSAmeth Downsample
-
-```{r,gsamethdownsample,fig.height=10}
-
-SAMPLESIZES=c(2,3,6,10,16,22,30)
-
-gsres <- lapply(SAMPLESIZES,function(n) {
-  SEEDS <- seq(100,5000,100)
-  dsres <- mclapply(SEEDS, function(SEED) {
-    set.seed(SEED)
-    mysample <- sample(1:37,n,replace=FALSE)*2
-    mysample <- c(mysample,(mysample-1))
-    mysample <- mysample[order(mysample)]
-    design2 <- design[mysample,]
-    design2 <- design2[,colSums(design2)>0]
-    #mval downsample
-    mval2 <- mval[,mysample]
-    dm2 <- runlimma(mval2,design2,myann)
-    if ( is.na(dm2$logFC[1]) ) {
-      RES=c("TOT"=0,"TP"=0,"FP"=0,"FN"=length(rownames(gsig)) )
-    } else {
-      sigup2 <- rownames(subset(dm2,adj.P.Val<0.05 & UCSC_RefGene_Name != "" & logFC>0))
-      if ( length(sigup2) < 250 ) { sigup2 <- head(rownames(subset(dm2,logFC>0 & UCSC_RefGene_Name != "" )),250) }
-      sigdn2 <- rownames(subset(dm2,adj.P.Val<0.05 & UCSC_RefGene_Name != "" & logFC<0))
-      if ( length(sigdn2) < 250 ) { sigdn2 <- head(rownames(subset(dm2,logFC<0 & UCSC_RefGene_Name != "" )),250) }
-      gsaup2 <- gsameth(sig.cpg=sigup2, all.cpg=rownames(dm2), collection=gs_entrez)
-      rownames(gsaup2) <- paste(rownames(gsaup2),"UP")
-      gsadn2 <- gsameth(sig.cpg=sigdn2, all.cpg=rownames(dm2), collection=gs_entrez)
-      rownames(gsadn2) <- paste(rownames(gsadn2),"DN")
-      gsig_up2 <- subset(gsaup2,FDR<0.05)
-      gsig_dn2 <- subset(gsadn2,FDR<0.05)
-      gsig2 <- rbind(gsig_up2,gsig_dn2)
-      TOT=nrow(gsig2)
-      TP=length(which(rownames(gsig2) %in% rownames(gsig)))
-      FP=length(which(!rownames(gsig2) %in% rownames(gsig)))
-      FN=length(which(!rownames(gsig) %in% rownames(gsig2)))
-      RES=c("TOT"=TOT,"TP"=TP,"FP"=FP,"FN"=FN)
-    }
-    RES
-  }, mc.cores=5)
-  do.call(rbind,dsres)
-})
-
-mx <- matrix(rep(0,200),ncol=4)
-colnames(mx) <- c("TOT","TP","FP","FN")
-
-# fix empty matrix
-gsres <- lapply(gsres,function(x) {
-  if ( typeof(x) == "character" ) {
-    mx
-  } else {
-    x
-  }
-} )
-
-par(mfrow=c(3,1))
-
-dots <- lapply(gsres,function(x) { x[,"TP"] })
-boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. true positives",xlab="sample size",cex=0,main="GSAmeth")
-beeswarm(dots,add=TRUE,pch=1,cex=0.6)
-abline(h=nrow(gsig))
-
-dots <- lapply(gsres,function(x) { x[,"FP"] })
-boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. false positives",xlab="sample size",cex=0)
-beeswarm(dots,add=TRUE,pch=1,cex=0.6)
-
-dots <- lapply(gsres,function(x) { x[,"TP"] })
-TPdots <- do.call(rbind,dots)
-dots <- lapply(gsres,function(x) { x[,"FP"] })
-FPdots <- do.call(rbind,dots)
-ratio <- t(FPdots / (TPdots + FPdots))
-ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
-boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion FP",xlab="sample size",cex=0) ; grid()
-beeswarm(ratio,add=TRUE,pch=1,cex=0.6)
-
-par(mfrow=c(1,1))
-
-```
-
-PNG output of boxplots.
-
-```{r,pngboxplot}
-
-png(filename="fig4.png",height=150,width=210,units="mm",res=200)
-par(mar = c(4.5, 4.1, 2.1, 0.5))
-par(mfrow=c(3,2))
-
-dots <- lapply(ssres,function(x) { x[,"TP"] })
-boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",cex=0,main="LAM") ; grid()
-beeswarm(dots,add=TRUE,pch=1,cex=0.6)
-abline(h=nrow(lamsig))
-
-dots <- lapply(gsres,function(x) { x[,"TP"] })
-boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",cex=0,main="GSA") ; grid()
-beeswarm(dots,add=TRUE,pch=1,cex=0.6)
-abline(h=nrow(gsig))
-
-dots <- lapply(ssres,function(x) { x[,"FP"] })
-boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. inconsistent",xlab="sample size",cex=0) ; grid()
-beeswarm(dots,add=TRUE,pch=1,cex=0.6)
-
-dots <- lapply(gsres,function(x) { x[,"FP"] })
-boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. inconsistent",xlab="sample size",cex=0) ; grid()
-beeswarm(dots,add=TRUE,pch=1,cex=0.6)
-
-dots <- lapply(ssres,function(x) { x[,"TP"] })
-TPdots <- do.call(rbind,dots)
-dots <- lapply(ssres,function(x) { x[,"FP"] })
-FPdots <- do.call(rbind,dots)
-ratio <- t(FPdots / (TPdots + FPdots))
-ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
-boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion inconsistent",xlab="sample size",cex=0) ; grid()
-beeswarm(ratio,add=TRUE,pch=1,cex=0.6)
-
-dots <- lapply(gsres,function(x) { x[,"TP"] })
-TPdots <- do.call(rbind,dots)
-dots <- lapply(gsres,function(x) { x[,"FP"] })
-FPdots <- do.call(rbind,dots)
-ratio <- t(FPdots / (TPdots + FPdots))
-ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
-boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion inconsistent",xlab="sample size",cex=0) ; grid()
-beeswarm(ratio,add=TRUE,pch=1,cex=0.6)
-
-dev.off()
-
-par(mfrow=c(1,1))
-
-```
-
-## Overlap
-
-Use Euler diagrams to find the similarity.
-As medians are normally distributed about the zero, one would expect equal numbers of up and down-
-methylated gene sets.
-Therefore the t-test based approach could be considered better.
-
-```{r,ol1}
-
-lamsigsets2
-gsigsets
-
-v1 <- list("LAM"=lamsigsets2,"GSA"=gsigsets)
-
-plot(euler(v1),quantities = TRUE)
-
-message("common")
-intersect(lamsigsets2, gsigsets)
-
-message("specific to LAM")
-LAMspec <- setdiff(lamsigsets2, gsigsets)
-LAMspec
-
-message("specific to GSA")
-GSAspec <- setdiff(gsigsets, lamsigsets2)
-GSAspec
-
-message("discordant")
-intersect( sapply(strsplit(LAMspec," "),"[[",1) , sapply(strsplit(GSAspec," "),"[[",1) )
-
-```
-
-Now look at the size of these genesets.
-Make sure these are ACTUALLY detected.
-Make a vector of genes.
-
-```{r,sizes}
-
-detgenes <- unique(unlist(strsplit(dm$UCSC_RefGene_Name,";")))
-
-gsigsets1 <- sapply(strsplit(gsigsets," "),'[[',1)
-gsigsets1set <- gs_symbols[names(gs_symbols) %in% gsigsets1]
-gsigsets1len <- unlist(lapply(gsigsets1set,function(set) {
-  length(set[which(set %in% detgenes)])
-}))
-
-lamsigsets1 <- sapply(strsplit(lamsigsets2," "),'[[',1)
-lamsigsets1set <- gs_symbols[names(gs_symbols) %in% lamsigsets1]
-lamsigsets1len <- unlist(lapply(lamsigsets1set,function(set) {
-  length(set[which(set %in% detgenes)])
-}))
-
-summary(gsigsets1len)
-summary(lamsigsets1len)
-
-GSANAME=paste("GSA (",length(gsigsets1len),")",sep="")
-LAMNAME=paste("LAM (",length(lamsigsets1len),")",sep="")
-mylist <- list(GSANAME=gsigsets1len,LAMNAME=lamsigsets1len)
-boxplot(mylist,cex=0, ylab="gene set size",names=c(GSANAME,LAMNAME))
-beeswarm(mylist,pch=19,add=TRUE,cex=0.7)
-
-```
-
-Make Fig3 an euler diagram, boxplot of sizes and barplot of top results.
-
-```{r,eulerbox}
-
-png(filename="euler.png",height=75,width=105,units="mm",res=200)
-plot(euler(v1),quantities = TRUE)
-dev.off()
-
-pic <- readPNG("euler.png")
-
-png(filename="fig3.png",height=150,width=210,units="mm",res=200)
-par(mar = c(4.5, 4.1, 2.1, 0.5))
-par(mfrow=c(2,2))
-
-#A overlap of LAM and GSA results
-par(mar = c(2,0,2,0))
-plot.new()
-plot.window(xlim = c(0 , 1), ylim = c( 0, 1))
-rasterImage(pic,0,0,1,1)
-par(mar = c(4.5, 4.1, 2.1, 0.5))
-
-#B
-par(mar = c(4.5, 4.1, 2.1, 0.5))
-boxplot(mylist,cex=0, ylab="gene set size",names=c(GSANAME,LAMNAME))
-beeswarm(mylist,pch=1,add=TRUE,cex=0.3)
-
-#C
-top <- head(lamsig,20)
-top <- top[order(top$s.dist),]
-barnames <- gsub("_"," ",gsub("REACTOME_","",rownames(top)))
-barnames2 <- substring(barnames, 1, 60)
-barnames2 <- paste(barnames2,gsub("TRUE","...",gsub("FALSE","",sapply(barnames,nchar)>60)))
-cols <- gsub("TRUE","red",gsub("FALSE","blue",as.character(top$s.dist>0)))
-par(mar = c(6.1, 20.1, 1.1, 0.5))
-barplot(abs(top$s.dist),horiz=TRUE,las=1,names.arg=barnames2, cex.names=0.55,
-  cex.axis=0.7,col=cols)
-title(main="LAM",cex = 0.8)
-title(xlab="S dist",cex=0.6)
-grid()
-
-#D
-barnames <- gsub("_"," ",gsub("REACTOME_","",rownames(gsiges)))
-cols <- gsub("FALSE","blue",gsub("TRUE","red",grepl(" UP",barnames)))
-barnames <- gsub("DN","",gsub(" UP","",barnames))
-barnames2 <- substring(barnames, 1, 60)
-barnames2 <- paste(barnames2,gsub("TRUE","...",gsub("FALSE","",sapply(barnames,nchar)>60)))
-par(mar = c(6.1, 20.1, 1.1, 0.5))
-barplot(abs(gsiges$ES),horiz=TRUE,las=1,names.arg=barnames2, cex.names=0.55,
-  cex.axis=0.7, col=cols, xlab="Enrichment score",main="GSA")
-grid()
-
-dev.off()
-
-```
-
-Interesting result that shows that LAM detected gene sets are smaller (59.5) than
-GSA sets (205).
-
-This could be due to one of the quirks in some ORA analyses were genes without an
-annotation are excluded.
 

### 2023-11-21 figs/GSE158422_rna_int.Rmd
Prefer UCSC annotations

--- 2023-11-17 figs/GSE158422_rna_int.Rmd
@@
 anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
-myann <- data.frame(anno[,c("UCSC_RefGene_Name","Regulatory_Feature_Group","Islands_Name","Relation_to_Island")])
+myann <- data.frame(anno[,c("UCSC_RefGene_Name","GencodeCompV12_Group","Islands_Name","Relation_to_Island")])
 gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
@@
 
-table(dm$Regulatory_Feature_Group)
+table(dm$UCSC_RefGene_Group)
 
@@
 # promoter
-dmpr <- dm[grep("Promoter",dm$Regulatory_Feature_Group),]
+dmpr <- dm[grep("Promoter",dm$UCSC_RefGene_Group),]
 dmpragg <- agg(dm=dmpr,cores=16)
@@
 # ex-promoter
-dmx <- dm[grep("Promoter",dm$Regulatory_Feature_Group,invert=TRUE),]
+dmx <- dm[grep("Promoter",dm$UCSC_RefGene_Group,invert=TRUE),]
 dmxagg <- agg(dm=dmx,cores=16)

### 2023-11-21 figs/gmea.Rmd
Prefer UCSC annotations

@@
 ann450k <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
-myann <- data.frame(ann450k[,c("UCSC_RefGene_Name","Regulatory_Feature_Group")])
+myann <- data.frame(ann450k[,c("UCSC_RefGene_Name","UCSC_RefGene_Group")])
 gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
@@
 sets <- split(rep(names(gp2), lengths(gp2)), unlist(gp2))
-promoters <- grep("Prom",myann$Regulatory_Feature_Group)
+promoters <- grep("Prom",myann$UCSC_RefGene_Group)
 

### 2023-11-21 figs/step_01a_GSE74548_prep.Rmd
Prefer UCSC annotations

@@
 ann450k <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
-myann <- data.frame(ann450k[,c("UCSC_RefGene_Name","Regulatory_Feature_Group")])
-promoters <- grep("Prom",myann$Regulatory_Feature_Group)
+myann <- data.frame(ann450k[,c("UCSC_RefGene_Name","UCSC_RefGene_Group")])
+promoters <- grep("Prom",myann$UCSC_RefGene_Group)
 

### 2023-11-21 figs/GSE158422_pkgdev.Rmd
a walkthrough for infinium limma users
Prefer UCSC annotations

@@
 anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
-myann <- data.frame(anno[,c("GencodeCompV12_NAME","GencodeCompV12_Group","Islands_Name","Relation_to_Island")])
-gp <- myann[,"GencodeCompV12_NAME",drop=FALSE]
-gp2 <- strsplit(gp$GencodeCompV12_NAME,";")
+myann <- data.frame(anno[,c("UCSC_RefGene_Name","GencodeCompV12_Group","Islands_Name","Relation_to_Island")])
+gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
+gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
 names(gp2) <- rownames(gp)
@@
 mitch_report(res=mres,outfile="GSE158422_pkgdev_mitchreport.html",overwrite=TRUE)
-mitch_plot2(res=mres,outfile="GSE158422_pkgdev_mitchcharts.pdf")
+mitch_plots(res=mres,outfile="GSE158422_pkgdev_mitchcharts.pdf")
 
@@
 Now do the same but filter for promoters.
-Using the GencodeCompV12_NAME field of the annotation and using TSS probes only.
+Using the UCSC_RefGene_Name field of the annotation and using TSS probes only.
 
@@
 head(myann)
-table(anno$Regulatory_Feature_Group)
+table(anno$UCSC_RefGene_Group)
 table(unlist(strsplit(myann$GencodeCompV12_Group,";")))
@@
 annos <- strsplit(myann$GencodeCompV12_Group,";")
-genes <- strsplit(myann$GencodeCompV12_NAME,";")
+genes <- strsplit(myann$UCSC_RefGene_Name,";")
 keeps <- lapply(1:nrow(myann), function(i) {

### 2023-11-22 figs/GSE158422_simulate020.Rmd
use half of genes for sets

@@
 
-#genesets <- gmt_import("https://ziemann-lab.net/public/gmea_prototype/ReactomePathways.gmt")
-
 if (!file.exists("GSE158422_design.rds")) {
@@
 
-boxplot(list("normal"=matrix(colMeans(mval),ncol=2)[,2],"tumor"=matrix(colMeans(mval),ncol=2)[,1]),
-  main="mean probe methylation mval")
-
 ```
@@
 
+To reflect coverage of Reactome and other pathway databases, only half the genes will be included in sets.
+
 ```{r,gset_prep}
@@
 
+# set gene set size here
 lengths <- rep(20,1000)
@@
 randomGeneSets <- function(gene_catalog, lengths, seed){
+  set.seed(seed) ; gene_catalog_half <- sample(gene_catalog,length(gene_catalog)/2)
   num_gsets <- length(lengths)
@@
   gsets <- lapply(1:num_gsets,function(i) {
-    set.seed(seeds[i]) ; gs <- sample(gene_catalog,lengths[i])
+    set.seed(seeds[i]) ; gs <- sample(gene_catalog_half,lengths[i])
     return(gs)
@@
 gsets <- randomGeneSets(gene_catalog,lengths,seed=100)
-
-```
-
-Select nonoverlapping gene sets.
-
-Not in use now. Might delete later.
-
-```{r,gset_selection}
-
-gset_selection <- function(genesetdatabase,seed,gene_catalog) {
-  set.seed(seed)
-  gsets <- genesetdatabase[sample(1:length(genesetdatabase))]
-  genelist=NULL
-  genesets=NULL
-  for ( i in 1:length(gsets) ) {
-    gs <- gsets[i]
-    inx <- length(intersect(unlist(gs),genelist))
-    num <- length(intersect(unlist(gs),gene_catalog))
-    if ( inx == 0 & num >= 10 ) {
-      genesets <- c(genesets,gs)
-      genelist <- c(unname(unlist(gs)),genelist)
-    }
-  }
-  return(genesets)
-}
-
-gset_mod <- gset_selection(genesetdatabase=gsets,seed=100,gene_catalog=gene_catalog)
 
@@
 num_dm_sets=50
-sims=10
+sims=3
 
@@
 
-save.image("GSE158422_simulate020.Rdata")
+save.image("GSE158422_simulate050.Rdata")
 

### 2023-11-22 figs/GSE158422_simulate050.Rmd
use half of genes for sets

@@
 
+# set gene set size here
 lengths <- rep(50,1000)

### 2023-11-22 figs/GSE158422_simulate100.Rmd
use half of genes for sets

@@
 
+# set gene set size here
 lengths <- rep(100,1000)
@@
 
-save.image("GSE158422_simulate100.Rdata")
+save.image("GSE158422_simulate050.Rdata")
 

### 2023-11-22 figs/GSE158422_simulate050.Rmd
added extra 10% DM genes and probes

@@
 
-incorp_dm <- function(genesets,myann,mval,seed,
-  frac_genes,frac_probes,groupsize,delta=1) {
+incorp_dm <- function(genesets,myann,mval,seed, frac_genes,frac_probes,
+  groupsize,delta=1,gene_catalog) {
 
@@
   gdn <- unname(unlist(gtdn))
+  # add extra 10% DM genes
+  gnon <- setdiff(gene_catalog,c(gup,gdn))
+  gextra <- round(length(gnon)*0.1)
+  set.seed(seed) ; gup <- c(gup,sample(gnon,gextra))
+  gnon <- setdiff(gene_catalog,c(gup,gdn))
+  set.seed(seed) ; gdn <- c(gdn,sample(gnon,gextra))
   # make probe-gene vector
@@
   set.seed(seed) ; pdn2 <- sample(pdn,floor(length(pdn)*frac_probes))
+  # add 10% DM probes as well
+  probes <- rownames(myann)
+  pnon <- setdiff(probes,c(pup,pdn))
+  pextra <- round(length(pnon)*0.1)
+  set.seed(seed) ; pup <- c(pup,sample(pnon,pextra))
+  pnon <- setdiff(probes,c(pup,pdn))
+  set.seed(seed) ; pdn <- c(pdn,sample(pnon,pextra))
   # divide samples between ctrl and case
@@
   sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
-    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
+    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta,
+    gene_catalog=gene_catalog)
   # set up limma

### 2023-11-26 figs/GSE158422_simulate_summary.Rmd
fig2
bugfix

@@
 
-x %>% kbl(caption="sim 100 res") %>% kable_paper("hover", full_width = F)
+x3 %>% kbl(caption="sim 100 res") %>% kable_paper("hover", full_width = F)
 
-barplot(t(x),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, cex.names=0.7, cex.axis=0.7, main="100 genes per set")
+barplot(t(x3),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, cex.names=0.7, cex.axis=0.7, main="100 genes per set")
 grid()
@@
 
+# png
+png(filename="fig2.png",height=150,width=210,units="mm",res=200)
+
+par(mar = c(4.1, 4.1, 2.1, 0.5))
+
+par(mfrow=c(2,3))
+
+plot(rownames(gres3p),gres3p[,"0.5"],type="b",las=1,ylab="precision",xlab="",
+  main="",ylim=c(0,1))
+lines(rownames(gres3p),gres3p[,"0.4"],type="b",col="red")
+lines(rownames(gres3p),gres3p[,"0.3"],type="b",col="blue")
+lines(rownames(gres3p),gres3p[,"0.2"],type="b",col="darkgreen")
+grid()
+
+legend("bottomright", legend=c("0.5", "0.4", "0.3", "0.2"),title="delta",
+  lty=1, lwd=2, col=c("black", "red", "blue", "darkgreen"), cex=1)
+
+plot(rownames(gres3r),gres3r[,"0.5"],type="b",las=1,ylab="recall",xlab="group size",
+  main="GSA",ylim=c(0,1))
+lines(rownames(gres3r),gres3r[,"0.4"],type="b",col="red")
+lines(rownames(gres3r),gres3r[,"0.3"],type="b",col="blue")
+lines(rownames(gres3r),gres3r[,"0.2"],type="b",col="darkgreen")
+grid()
+
+plot(rownames(gres3r),gres3f[,"0.5"],type="b",las=1,ylab="F1",xlab="",
+  main="",ylim=c(0,1))
+lines(rownames(gres3r),gres3f[,"0.4"],type="b",col="red")
+lines(rownames(gres3r),gres3f[,"0.3"],type="b",col="blue")
+lines(rownames(gres3r),gres3f[,"0.2"],type="b",col="darkgreen")
+grid()
+
+plot(rownames(lacmres3p),lacmres3p[,"0.5"],type="b",las=1,ylab="precision",xlab="",
+  main="",ylim=c(0,1))
+lines(rownames(lacmres3p),lacmres3p[,"0.4"],type="b",col="red")
+lines(rownames(lacmres3p),lacmres3p[,"0.3"],type="b",col="blue")
+lines(rownames(lacmres3p),lacmres3p[,"0.2"],type="b",col="darkgreen")
+grid()
+
+legend("bottomright", legend=c("0.5", "0.4", "0.3", "0.2"),title="delta",
+  lty=1, lwd=2, col=c("black", "red", "blue", "darkgreen"), cex=1)
+
+plot(rownames(lacmres3r),lacmres3r[,"0.5"],type="b",las=1,ylab="recall",xlab="group size",
+  main="LAM",ylim=c(0,1))
+lines(rownames(lacmres3r),lacmres3r[,"0.4"],type="b",col="red")
+lines(rownames(lacmres3r),lacmres3r[,"0.3"],type="b",col="blue")
+lines(rownames(lacmres3r),lacmres3r[,"0.2"],type="b",,col="darkgreen")
+grid()
+
+plot(rownames(lacmres3r),lacmres3f[,"0.5"],type="b",las=1,ylab="F1",xlab="",
+  main="",ylim=c(0,1))
+lines(rownames(lacmres3r),lacmres3f[,"0.4"],type="b",col="red")
+lines(rownames(lacmres3r),lacmres3f[,"0.3"],type="b",col="blue")
+lines(rownames(lacmres3r),lacmres3f[,"0.2"],type="b",col="darkgreen")
+grid()
+
+dev.off()
+
 par(mfrow=c(1,1))

### 2023-11-28 figs/GSE158422_pkgdev.Rmd
slight update

@@
 anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
-myann <- data.frame(anno[,c("UCSC_RefGene_Name","GencodeCompV12_Group","Islands_Name","Relation_to_Island")])
+myann <- data.frame(anno[,c("UCSC_RefGene_Name","UCSC_RefGene_Group","Islands_Name","Relation_to_Island")])
 gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
@@
 colnames(gt) <- c("gene","probe")
+gt$probe <- as.character(gt$probe)
 dim(gt)
@@
 toc() #9.0s
+
+```
+
+Now try for efficient method to incorporate into the package.
+
+```{r,anno2}
+
+library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
 

### 2023-11-28 figs/GSE158422_rna_int.Rmd
progress on promoter versus body analysis

@@
   library("tictoc")
+  library("limma")
   library("mitch")
@@
 mval <- readRDS("GSE158422_mx.rds")
-dm <- readRDS("GSE158422_dm.rds")
+
+runlimma <- function(mval,design,myann) {
+  fit.reduced <- lmFit(mval,design)
+  fit.reduced <- eBayes(fit.reduced)
+  dm <- topTable(fit.reduced,coef=ncol(design), number = Inf)
+  dm <- merge(myann,dm,by=0)
+  dm <- dm[order(dm$P.Value),]
+  rownames(dm) <- dm$Row.names
+  dm$Row.names=NULL
+  return(dm)
+}
+
+dm <- runlimma(mval=mval,design=design,myann=myann)
 
@@
 
-```{r,lamfunc}
+## Curate the annotation
 
-# LAM aggregate
-agg <- function(dm,cores=1) {
-  dm <- dm[which(dm$UCSC_RefGene_Name!=""),]
-  gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
-  gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
-  gnl <- mclapply(gnl,unique,mc.cores=cores)
-  dm$UCSC_RefGene_Name <- gnl
-  l <- mclapply(1:nrow(dm), function(i) {
-    a <- dm[i,]
-    len <- length(a[[1]][[1]])
-    tvals <- as.numeric(rep(a["t"],len))
-    genes <- a[[1]][[1]]
-    data.frame(genes,tvals)
-  },mc.cores=cores)
-  df <- do.call(rbind,l)
-  keep <- names(which(table(df$genes)>1))
-  df <- df[df$genes %in% keep,]
-  gn <- unique(df$genes)
-  gme_res <- lapply( 1:length(gn), function(i) {
-    g <- gn[i]
-    tstats <- df[which(df$genes==g),"tvals"]
-    myn <- length(tstats)
-    mymean <- mean(tstats)
-    mymedian <- median(tstats)
-    top <- tstats[order(-abs(tstats))][1]
-    if ( length(tstats) > 2 ) {
-      ttest <- t.test(tstats)
-      pval <- ttest$p.value
-    } else {
-      pval = 1
-    }
-    res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,
-      "median"=mymedian, "top"=top, "pval"=pval)
-  } )
-  gme_res_df <- do.call(rbind, gme_res)
-  rownames(gme_res_df) <- gme_res_df[,1]
-  gme_res_df <- gme_res_df[,-1]
-  tmp <- apply(gme_res_df,2,as.numeric)
-  rownames(tmp) <- rownames(gme_res_df)
-  gme_res_df <- as.data.frame(tmp)
-  gme_res_df$sig <- -log10(gme_res_df[,"pval"])
-  gme_res_df <- gme_res_df[order(-gme_res_df$sig),]
-  gme_res_df$fdr <- p.adjust(gme_res_df$pval)
-  out <- list("df"=df,"gme_res_df"=gme_res_df)
-  return(out)
-}
+Use all probes on the chip.
+
+```{r,anno1}
+
+anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
+myann <- data.frame(anno[,c("UCSC_RefGene_Name","UCSC_RefGene_Group","Islands_Name","Relation_to_Island")])
+gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
+gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
+names(gp2) <- rownames(gp)
+gp2 <- lapply(gp2,unique)
+gt <- stack(gp2)
+colnames(gt) <- c("gene","probe")
+gt$probe <- as.character(gt$probe)
+dim(gt)
+str(gt)
 
@@
 
+The approach here is to generate two mapping tables of probes to genes based on the annotation which is split
+into gene body and promoter (TSS).
+
+gene body=gb
+
+gene promoter (TSS) = gp
+
 ```{r,promoterbody}
 
-table(dm$UCSC_RefGene_Group)
+# partition body and promoter data(probes)
+dmg <- strsplit(dm$UCSC_RefGene_Name,";")
+
+body <- lapply(1:nrow(dm),function(i) {
+  a <- dm[i,"UCSC_RefGene_Group"]
+  a <- unlist(strsplit(a,";"))
+  dmg[[i]][a=="Body"]
+})
+names(body) <- rownames(dm)
+body <- lapply(body,unique)
+gb <- stack(body)
+colnames(gb) <- c("gene","probe")
+gb$probe <- as.character(gb$probe)
+dim(gb)
+str(gb)
+
+tss <- lapply(1:nrow(dm),function(i) {
+  a <- dm[i,"UCSC_RefGene_Group"]
+  a <- unlist(strsplit(a,";"))
+  dmg[[i]][grep("TSS",a)]
+})
+names(tss) <- rownames(dm)
+tss <- lapply(tss,unique)
+gp <- stack(tss)
+colnames(gp) <- c("gene","probe")
+gp$probe <- as.character(gp$probe)
+dim(gp)
+str(gp)
+
+# HERENOW
 

### 2023-11-28 figs/GSE158422_rna_int.Rmd
RNA integration first draft

@@
 
-CORES=12
+CORES=16
 
@@
 
-res <- mitch_calc(rna, genesets=gs_symbols, priority="effect",cores=32)
+res <- mitch_calc(rna, genesets=gs_symbols, priority="effect",cores=16)
 
@@
 
-# HERENOW
+# import for promoters
+dmp <- dm[which(rownames(dm) %in%  gp$probe),]
+mp <- mitch_import(x=dmp,DEtype="limma",geneTable=gp)
+dim(dmp)
+dim(mp)
 
-# all gene
-dmagg <- agg(dm=dm,cores=16)
-str(dmagg)
-head(dmagg$gme_res_df)
-dim(dm)
-dim(dmagg$gme_res_df)
-mm <- dmagg$gme_res_df[,"mean",drop=FALSE]
-mmres <- mitch_calc(mm, genesets=gs_symbols, priority="effect",cores=32)
-mmres_up <- subset(mmres$enrichment_result,s.dist>0 & p.adjustANOVA < 0.05)
-mmres_dn <- subset(mmres$enrichment_result,s.dist<0 & p.adjustANOVA < 0.05)
-nrow(mmres_up)
-nrow(mmres_dn)
-head(mmres_up,20)
-head(mmres_dn,20)
+# import bodies
+dmb <- dm[which(rownames(dm) %in%  gb$probe),]
+mb <- mitch_import(x=dmb,DEtype="limma",geneTable=gb)
+dim(dmb)
+dim(mb)
 
 # promoter
-dmpr <- dm[grep("Promoter",dm$UCSC_RefGene_Group),]
-dmpragg <- agg(dm=dmpr,cores=16)
-str(dmpragg)
-head(dmpragg$gme_res_df)
-dim(dmpr)
-dim(dmpragg$gme_res_df)
-mmpr <- dmpragg$gme_res_df[,"mean",drop=FALSE]
-mmprres <- mitch_calc(mmpr, genesets=gs_symbols, priority="effect",cores=32)
-mmprres_up <- subset(mmprres$enrichment_result,s.dist>0 & p.adjustANOVA < 0.05)
-mmprres_dn <- subset(mmprres$enrichment_result,s.dist<0 & p.adjustANOVA < 0.05)
-nrow(mmprres_up)
-nrow(mmprres_dn)
-head(mmprres_up,20)
-head(mmprres_dn,20)
+dmp_res <- mitch_calc(mp, genesets=gs_symbols, priority="effect",cores=16)
+dmp_res_up <- subset(dmp_res$enrichment_result,s.dist>0 & p.adjustANOVA < 0.05)
+dmp_res_dn <- subset(dmp_res$enrichment_result,s.dist<0 & p.adjustANOVA < 0.05)
+nrow(dmp_res_up)
+nrow(dmp_res_dn)
+head(dmp_res_up,20)
+head(dmp_res_dn,20)
 
-# ex-promoter
-dmx <- dm[grep("Promoter",dm$UCSC_RefGene_Group,invert=TRUE),]
-dmxagg <- agg(dm=dmx,cores=16)
-str(dmxagg)
-head(dmxagg$gme_res_df)
-dim(dmx)
-dim(dmxagg$gme_res_df)
-mmx <- dmxagg$gme_res_df[,"mean",drop=FALSE]
-mmxres <- mitch_calc(mmx, genesets=gs_symbols, priority="effect",cores=32)
-mmxres_up <- subset(mmxres$enrichment_result,s.dist>0 & p.adjustANOVA < 0.05)
-mmxres_dn <- subset(mmxres$enrichment_result,s.dist<0 & p.adjustANOVA < 0.05)
-nrow(mmxres_up)
-nrow(mmxres_dn)
-head(mmxres_up,20)
-head(mmxres_dn,20)
+# body
+dmb_res <- mitch_calc(mb, genesets=gs_symbols, priority="effect",cores=16)
+dmb_res_up <- subset(dmb_res$enrichment_result,s.dist>0 & p.adjustANOVA < 0.05)
+dmb_res_dn <- subset(dmb_res$enrichment_result,s.dist<0 & p.adjustANOVA < 0.05)
+nrow(dmb_res_up)
+nrow(dmb_res_dn)
+head(dmb_res_up,20)
+head(dmb_res_dn,20)
 
 # combined methylation
-mmm <- merge(mmpr,mmx,by=0)
-rownames(mmm) <- mmm$Row.names
-mmm$Row.names = NULL
-colnames(mmm) <- c("pr","x")
-mmmres <- mitch_calc(mmm, genesets=gs_symbols, priority="effect",cores=32)
-mmmres_sig <- subset(mmmres$enrichment_result, p.adjustMANOVA < 0.05)
-head(mmmres_sig,20)
-table(sign(mmmres_sig$s.pr) == sign(mmmres_sig$s.x))
+mm <- merge(mp,mb,by=0)
+rownames(mm) <- mm$Row.names
+mm$Row.names=NULL
+colnames(mm) <- c("prom","body")
+mm_res <- mitch_calc(mm, genesets=gs_symbols, priority="effect",cores=16)
+mm_res_sig <- subset(mm_res$enrichment_result,p.adjustMANOVA < 0.01)
+nrow(mm_res_sig)
+head(mm_res_sig,20)
+table(sign(mm_res_sig$s.prom) == sign(mm_res_sig$s.body) )
+mm_res_sig[sign(mm_res_sig$s.prom) != sign(mm_res_sig$s.body),]
 
@@
 # three way
-mmrna <- merge(mmm,rna,by=0)
+mmrna <- merge(mm,rna,by=0)
 rownames(mmrna) <- mmrna$Row.names
 mmrna$Row.names = NULL
-colnames(mmrna) <- c("pr","x","rna")
-mmrnares <- mitch_calc(mmrna, genesets=gs_symbols, priority="effect",cores=32)
+colnames(mmrna) <- c("prom","body","rna")
+mmrnares <- mitch_calc(mmrna, genesets=gs_symbols, priority="effect",cores=16)
 mmrnares_sig <- subset(mmrnares$enrichment_result, p.adjustMANOVA < 0.05)
 head(mmrnares_sig,20)
-table(sign(mmrnares_sig$s.pr) == sign(mmrnares_sig$s.x))
-table(sign(mmrnares_sig$s.pr) == sign(mmrnares_sig$s.rna))
-table(sign(mmrnares_sig$s.x) == sign(mmrnares_sig$s.rna))
+table(sign(mmrnares_sig$s.prom) == sign(mmrnares_sig$s.body))
+table(sign(mmrnares_sig$s.prom) == sign(mmrnares_sig$s.rna))
+table(sign(mmrnares_sig$s.body) == sign(mmrnares_sig$s.rna))
 mtop <- head(mmrnares_sig,30)
-mtop <- mtop[,c("set","s.pr","s.x","s.rna")]
+mtop <- mtop[,c("set","s.prom","s.body","s.rna")]
 rownames(mtop) <- gsub("REACTOME_","",mtop$set)
@@
 mtop$set=NULL
-colnames(mtop) <- c("promoter","non-promoter","RNA")
+colnames(mtop) <- c("promoter","body","RNA")
 colfunc <- colorRampPalette(c("blue", "white", "red"))
@@
 # rna vs promoter
-mmrna <- merge(mmpr,rna,by=0)
-rownames(mmrna) <- mmrna$Row.names
-mmrna$Row.names = NULL
-colnames(mmrna) <- c("pr","rna")
-mmrnares <- mitch_calc(mmrna, genesets=gs_symbols, priority="effect",cores=32)
-mmrnares_sig <- subset(mmrnares$enrichment_result, p.adjustMANOVA < 0.05)
-head(mmrnares_sig,20)
-table(sign(mmrnares_sig$s.pr) == sign(mmrnares_sig$s.rna))
-mmrnares_sig <- subset(mmrnares$enrichment_result, p.adjustMANOVA < 0.05)
-mylm <- lm(mmrnares$enrichment_result$s.rna ~ mmrnares$enrichment_result$s.pr)
+mprna <- merge(mp,rna,by=0)
+rownames(mprna) <- mprna$Row.names
+mprna$Row.names = NULL
+colnames(mprna) <- c("prom","rna")
+mprnares <- mitch_calc(mprna, genesets=gs_symbols, priority="effect",cores=16)
+mprnares_sig <- subset(mprnares$enrichment_result, p.adjustMANOVA < 0.05)
+head(mprnares_sig,20)
+table(sign(mprnares_sig$s.prom) == sign(mmrnares_sig$s.rna))
+mprnares_sig <- subset(mprnares$enrichment_result, p.adjustMANOVA < 0.05)
+mylm <- lm(mprnares$enrichment_result$s.rna ~ mprnares$enrichment_result$s.prom)
 summary(mylm)
-mylm2 <- lm(mmrnares_sig$s.rna ~ mmrnares_sig$s.pr)
+mylm2 <- lm(mprnares_sig$s.rna ~ mprnares_sig$s.pr)
 summary(mylm2)
-plot(mmrnares$enrichment_result$s.pr,mmrnares$enrichment_result$s.rna,xlab="promoter",ylab="RNA",main="S dist",col="gray",pch=19)
+plot(mprnares$enrichment_result$s.prom,mprnares$enrichment_result$s.rna,xlab="promoter",
+  ylab="RNA", main="S dist",col="gray",pch=19)
 abline(v=0,h=0,lty=2,lwd=2)
-points(mmrnares_sig$s.pr,mmrnares_sig$s.rna,col="black",pch=19,cex=1.01)
+points(mprnares_sig$s.pr,mprnares_sig$s.rna,col="black",pch=19,cex=1.01)
 abline(mylm,col="blue",lty=3,lwd=3)
@@
 
-# rna vs expromoter
-mmrna <- merge(mmx,rna,by=0)
-rownames(mmrna) <- mmrna$Row.names
-mmrna$Row.names = NULL
-colnames(mmrna) <- c("x","rna")
-mmrnares <- mitch_calc(mmrna, genesets=gs_symbols, priority="effect",cores=32)
-mmrnares_sig <- subset(mmrnares$enrichment_result, p.adjustMANOVA < 0.05)
-head(mmrnares_sig,20)
-table(sign(mmrnares_sig$s.x) == sign(mmrnares_sig$s.rna))
-mmrnares_sig <- subset(mmrnares$enrichment_result, p.adjustMANOVA < 0.05)
-mylm <- lm(mmrnares$enrichment_result$s.rna ~ mmrnares$enrichment_result$s.x)
+# rna vs body
+mbrna <- merge(mb,rna,by=0)
+rownames(mbrna) <- mbrna$Row.names
+mbrna$Row.names = NULL
+colnames(mbrna) <- c("body","rna")
+mbrnares <- mitch_calc(mbrna, genesets=gs_symbols, priority="effect",cores=16)
+mbrnares_sig <- subset(mbrnares$enrichment_result, p.adjustMANOVA < 0.05)
+head(mbrnares_sig,20)
+table(sign(mbrnares_sig$s.body) == sign(mbrnares_sig$s.rna))
+mbrnares_sig <- subset(mbrnares$enrichment_result, p.adjustMANOVA < 0.05)
+mylm <- lm(mbrnares$enrichment_result$s.rna ~ mbrnares$enrichment_result$s.body)
 summary(mylm)
-mylm2 <- lm(mmrnares_sig$s.rna ~ mmrnares_sig$s.x)
+mylm2 <- lm(mbrnares_sig$s.rna ~ mbrnares_sig$s.body)
 summary(mylm2)
-plot(mmrnares$enrichment_result$s.x,mmrnares$enrichment_result$s.rna,xlab="ex-promoter",ylab="RNA",main="S dist",col="gray",pch=19)
+plot(mbrnares$enrichment_result$s.body,mbrnares$enrichment_result$s.rna,xlab="ex-promoter",
+  ylab="RNA",main="S dist",col="gray",pch=19)
 abline(v=0,h=0,lty=2,lwd=2)
-points(mmrnares_sig$s.x,mmrnares_sig$s.rna,col="black",pch=19,cex=1.01)
+points(mbrnares_sig$s.body,mbrnares_sig$s.rna,col="black",pch=19,cex=1.01)
 abline(mylm,col="blue",lty=3,lwd=3)
@@
 
-# all meth
+# body
 sigup <- rownames(subset(dm,logFC > 0 & adj.P.Val < 0.05))
 sigdn <- rownames(subset(dm,logFC < 0 & adj.P.Val < 0.05))
-gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez, array.type="EPIC")
+gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez,
+  array.type="EPIC", genomic.features = "Body")
 rownames(gsaup) <- paste(rownames(gsaup),"UP")
-gsadn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez, array.type="EPIC")
+gsadn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez,
+  array.type="EPIC", genomic.features = "Body")
 rownames(gsadn) <- paste(rownames(gsadn),"DN")
@@
 nrow(gsig)
+gsig_body <- gsig
+gsig_body_up <- gsig_up
+gsig_body_dn <- gsig_dn
 
-# all meth
-sigup <- rownames(subset(dm,logFC > 0 & adj.P.Val < 0.05))
-sigdn <- rownames(subset(dm,logFC < 0 & adj.P.Val < 0.05))
-gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez, array.type="EPIC")
+# promoter
+TSS = c("TSS200", "TSS1500")
+gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez,
+  array.type="EPIC", genomic.features = TSS)
 rownames(gsaup) <- paste(rownames(gsaup),"UP")
-gsadn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez, array.type="EPIC")
+gsadn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez,
+  array.type="EPIC", genomic.features = TSS)
 rownames(gsadn) <- paste(rownames(gsadn),"DN")
@@
 nrow(gsig)
+gsig_promoter <- gsig
+gsig_promoter_up <- gsig_up
+gsig_promoter_dn <- gsig_dn
+
+```
+
+Now make a Euler plot for GSA meth-RNA comparison.
+
+```{r,compare_rna_gsa}
+
+rna_up <- subset(msig,s.dist>0)$set
+rna_dn <- subset(msig,s.dist<0)$set
+
+gsig_body_up <- rownames(gsig_body_up)
+gsig_body_dn <- rownames(gsig_body_dn)
+
+gsig_promoter_up <- rownames(gsig_promoter_up)
+gsig_promoter_dn <- rownames(gsig_promoter_dn)
+
+vg <- list("RNA up"=rna_up,"RNA dn"=rna_dn,
+  "GSA body up"=gsig_body_up, "GSA body dn"=gsig_body_dn,
+  "GSA promoter up"=gsig_promoter_up, "GSA prom dn"=gsig_promoter_dn)
+
+plot(euler(vg),quantities=TRUE)
+
+```
+
+Now make a Euler plot for LAM-RNA comparison.
+
+```{r,compare_rna_mitch}
+
+dmp_res_up <- dmp_res_up$set
+dmp_res_dn <- dmp_res_dn$set
+
+dmb_res_up <- dmb_res_up$set
+dmb_res_dn <- dmb_res_dn$set
+
+vm <- list("RNA up"=rna_up,"RNA dn"=rna_dn,
+  "LAM body up"=dmb_res_up, "LAM body dn"=dmb_res_dn,
+  "LAM promoter up"=dmp_res_up, "LAM prom dn"=dmp_res_dn)
+
+plot(euler(vm),quantities=TRUE)
 

### 2023-12-09 figs/GSE158422_rna_int.Rmd
rna int

@@
   library("gplots")
+  library("ggplot2")
+
 })
@@
 
+p3 <- heatmap.2( as.matrix(mtop) , trace="none",margins = c(6,25), col=colfunc(25), cexRow=.6, cexCol=1)
+
 # rna vs promoter
@@
 
+p1 <- plot(euler(vg),quantities=TRUE)
+
 ```
@@
 
+p2 <- plot(euler(vm),quantities=TRUE)
+
 ```
@@
 they were not concordant with differential expression.
+
+## Assemble figure
+
+Grid arrange function.
+
+```{r,gridarrange}
+
+plot5 <- grid.arrange(arrangeGrob(p1, p2, ncol=1), p3,
+  ncol=2, widths=c(1,1))
+
+```
 

### 2023-12-09 manuscript/manuscript.Rmd
ms

@@
 EPIC Infinium array data derived from non-cancerous lung tissue of 37 cancer patients was used for our
-evaluations.
+evaluations (GSE158422).
+We downloaded raw intensity EPIC IDAT files from GEO for study GSE158422 which consists of
+lung tumour and normal adjacent tissues from 37 patients.
+Probe annotations were obtained from the "IlluminaHumanMethylationEPICanno.ilm10b4.hg19" Bioconductor
+package.
 We randomly sampled datasets to serve as control and cases (no replacement), with group sizes varying between
@@
 with decreased methylation.
-From these gene sets, half of the members were selected.
+From these gene sets, half of the member genes were selected.
 For those selected genes, half of the annotated probes were selected.
@@
 Following the incorporation of methylation changes to selected probes in the case group, the data underwent
-enrichment analysis using `GSAmeth` approach, or one of the seven FCS methods outlined above.
+limma differential methylation analysis, followed by enrichment analysis using `GSAmeth` approach, or one of
+the seven FCS methods outlined above.
 Enrichment results were then filtered for FDR<0.05 and correct direction of methylation change.
@@
 the number of true positives (TP), false positives (FP) and false negatives (FN) could be determined.
-At each setting for group size, delta and gene set size, 20 replications were conducted each with a set seed
+At each setting for group size, delta and gene set size, 50 replications were conducted each with a set seed
 to ensure reproducibility.
@@
 
-As there are no established methods for benchmarking Infinium array based pathway regulation using
-known ground truth synthetic data, we used a split-and-analyse approach with a real dataset to
-evaluate the three approaches outlined above.
-We downloaded raw intensity EPIC IDAT files from GEO for study GSE158422 which consists of
-lung tumour and normal adjacent tissues from 37 patients.
-Probe annotations were obtained from the "IlluminaHumanMethylationEPICanno.ilm10b4.hg19" Bioconductor
-package.
-Reactome gene sets were downloaded from the MSigDB website.
-From this set of samples, 18 patients were randomly selected into analysis A, where control and tumour
-samples were compared.
-From the remaining 19 samples, 18 were randomly selected into analysis B, where control and tumour
-samples were compared as above to act as an independent validation of analysis A.
-Analyses A and B were conducted using one of the three enrichment approaches outlined above, and the
-gene sets meeting a 5% FDR threshold in the hyper- and hypo-methylated directions were saved.
-Gene sets identified to be common in A and B analyses (FDR<0.05 and direction) were considered validated,
-gene sets identified in only one method were not considered validated, and gene sets that gave discordant
-direction (hyper- vs hypo-methylation) between A and B were classed as contradictory.
-This process was repeated 50 times to rigourously evaluate method performance.
-
 ### Sensitivity analysis
 
-The same dataset including all 37 patients underwent analysis using the LA method, gsameth and ebGSEA.
-The effect size (enrichment score) of significantly differentially methylated pathways (FDR<0.05) was
-recorded.
+The same dataset including all 37 patients underwent pathway enrichment analysis using the GSAmeth and LAM methods
+with Reactome pathways downloaded from MSigDB website (version 2023.1).
+Euler diagrams were created with the `eulerr` R package.
 Then, a random subset of n patients was selected followed by pathway enrichment analysis with the
-respective method.
+respective method, with n varied between 2 and 30.
 The subset results were compared to the full n=37 results to test the sensitivity of these enrichment
-methods at differenent effect size levels.
+methods at differenent effect size levels, and whether findings at a smaller sample size were consistent with the full group.
 This process was repeated 50 times.
@@
 
-As methylation at promoters and gene bodies is considered to have differing association with gene
-expression, we conducted LA analysis at gene promoters, gene bodies and whole gene separately and
-examined the relationship with RNA-seq based pathway regulation.
-RNA-seq counts were downloaded from the GEO database (Accession: GSE158420).
+RNA-seq based gene counts for the same set of lung cancer patients were downloaded from GEO [GSE158420](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE158420).
 HGNChelper was used to repair transmogrified gene names, followed by DESeq2 based differential analysis.
-Enrichment analysis of RNA-seq data was conducted using the mitch package with REACTOME gene sets.
+This underwent filtering to remove genes with expression below 10 reads per sample on average across the dataset.
+Then DESeq2 was used to compare normal and cancerous tissue gene expression taking into consideration sample pairing.
+The differential expression results underwent enrichment analysis using the mitch package with default settings for DESeq2 data tables.
+CpG sites annotated as promoters were considered separately from those located at gene bodies in mitch analysis.
+For comparison, GSAmeth was used to separately examine pathways using separate analyses of promoter and gene body CpGs.
+For this analysis, Reactome gene sets were used.
+Subsequently, to generate the mitch heatmap, we used mitch in multivariate mode by analysing differential RNA expression,
+promoter and gene body methylation together, prioritising the results by s-distance, a surrogate measure of effect size.
 
@@
 The overlap of significant pathways was unexpectedly small, with just five common pathways.
-
-Given GSA found relatively few results
-
-Considering the number of significant probes was very large we were concerned that
 

### 2023-12-09 figs/GSE158422_rna_int.Rmd
fig5

--- 2023-11-28 figs/GSE158422_rna_int.Rmd
@@
   library("gplots")
+  library("gridExtra")
+  library("png")
 })
@@
 
+pdf("fig5c.pdf")
+heatmap.2( as.matrix(mtop) , trace="none",margins = c(6,25), col=colfunc(25), cexRow=.6, cexCol=1)
+dev.off()
+
 # rna vs promoter
@@
 
+pdf("fig5a.pdf")
+plot(euler(vg),quantities=TRUE)
+dev.off()
+
 ```
@@
 
+pdf("fig5b.pdf")
+plot(euler(vm),quantities=TRUE)
+dev.off()
+
 ```

### 2023-12-09 manuscript/manuscript.Rmd
fig5

@@
 
+![Figure 5.
+Integration of gene expression data with methylation pathways.
+(A,B) Euler diagrams showing the overlap in statistically significant pathways (FDR<0.05) from gene expression and GSA results (A) and LAM (B) results.
+(C) Multi-dimensional enrichment analysis of promoter and gene body methylation with gene expression using mitch.
+Top 20 gene sets shown with largest s distance after FDR siltering at 0.05.](../figs/fig5.png)
+
 ## Exploration of methylation pathways altered by ART

### 2023-12-11 manuscript/manuscript.Rmd
writing

@@
 We found that overall recall was positively associated with the size of the gene sets considered.
-At 20 genes per set, GSA had the highest overall performance, with F1 of 0.25, followed by LAW with 0.06.At 50 genes per set, GSA and LAM achieved similar F1 values of ~0.39.
-At 100 genes per set, LAM achieved the highest F1 performance of 0.70 followed by LAT with 0.61, while
-GSA scored only 0.42.
+At 20 genes per set, GSA had the highest overall performance, with F1 of 0.097, followed by LAM with 0.035.
+At 50 genes per set, LAM achieved an F1 value of 0.29, while GSA scored 0.19.
+At 100 genes per set, LAM showed the highest F1 performance of 0.56 followed by ALM with 0.53, while
+GSA scored only 0.24.
 This result suggests that FCS methods perform better than ORA with gene set sizes >50.
@@
 
-To compare the sensitivity of these methods, sample sizes were randomly downsampled, analysed
+To compare the sensitivity of these methods with real data, sample sizes were randomly downsampled, analysed
 as above and the significant pathways were compared to the results from the full
@@
 60% for GSA.
+This result suggests that LAM has superior sensitivity to detect differentially methylated
+pathways in real cancer data.
 
@@
 
-This result suggests that LAM has superior sensitivity to detect differentially methylated
-pathways in real cancer data.
-
 ## LAM and GSA integrative analysis with RNA expression
 
-* Separate promoter and gene body integration
+Associations of epigenetic marks with gene expression are of great interest for understanding disease processes.
+Matching RNA-seq datasets were analysed and we conducted enrichment analysis of gene expression together with
+DNA methylation.
+There were 12,386 differentially expressed genes, with 7,418 and 4,968 up- and down-regulated in the tumour group respectively.
+At the pathway level there were 318 and 193 up- and down-regulated Reactome pathways.
+As the context of gene methylation is important in influencing gene expression, promoter and gene body methylation
+were considered separately.
+In promoters, GSA identified 15 and 8 pathways with higher and lower methylation respectively, while at gene bodies, there were were 17 and 67
+pathways with higher and lower methylation respectively.
+Using LAM, there were 133 and 12 pathways identified with higher and lower promoter methylation respectively, while there were 236 and 35 with higher and lower gene body methylation respectively.
+The relative overlap between promoter and gene body results was larger for LAM pathway sets as compared to GSA (*Figure 5A,B*).
+There was no observed overlap between GSA methylation-based pathways and gene expression pathways (*Figure 5A*), but there were some
+overlaps between pathways identified with LAM based promoter and gene body methylation and gene expression (*Figure 5B*).
+Interestingly, 123/318 upregulated gene expression pathways were associated with increasing gene body methylation with LAM.
+Likewise 19 pathways with downregulated expression were associated with increasing promoter methylation.
+Some of these pathways with large effect sizes across the three contrasts are depicted in heatmap form (*Figure 5C*).
+Of these selected pathways, there was a mixture of concordant and discordant signatures in the three contrasts,
+suggesting that promoter methylation differences do not strongly influence RNA expression at the level of pathways.
+However there were a small number of pathways that did exhibit what could be epigenetic influences, for example
+"DNA strand elongation", "Unwinding of DNA", "Mucopolysaccharidoses" and "Interaction with cumulus cells and the zona pellucida"
+demonstrated an inverse relationship between promoter methylation and gene expression.
+These results demonstrate that LAM can illuminate the complicated relationship between DNA methylation and gene expression at the pathway level in cancer.
 
@@
 (C) Multi-dimensional enrichment analysis of promoter and gene body methylation with gene expression using mitch.
-Top 20 gene sets shown with largest s distance after FDR siltering at 0.05.](../figs/fig5.png)
+Top 20 gene sets shown with largest s distance after FDR filtering at 0.05.](../figs/fig5.png)
 

### 2023-12-12 figs/GSE158422_simulate_summary2.Rmd

---
title: "Testing gene methylation enrichment analysis approaches using simulated data - summarise data"
author: "The GMEA team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
---

## Introduction

Here is an explanation of the different objects:

0. GSA: gres

1. LAT: lacres

2. LTT: lactopres

3. LAW: nlacres

4. LAM: lacmres

5. ALT: alcres

6. ALW: nalcres

7. ALM: almres

```{r,packages}

suppressPackageStartupMessages({
  library("stringi")
  library("psych")
  library("kableExtra")
  library("limma")
  library("missMethyl")
  library("IlluminaHumanMethylation450kmanifest")
  library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
  library('org.Hs.eg.db')
  library("mitch")
})

```

## Functions

```{r,functions}

F1 <- function(x,y) {
  ( 2 * x * y ) / ( x + y )
}

summarise <- function(x) {
  PREC <- sum(x[,"TP"]) / ( sum(x[,"TP"]) + sum(x[,"FP"]) )
  REC <- sum(x[,"TP"]) / ( sum(x[,"TP"]) + sum(x[,"FN"]) )
  F <- F1(PREC,REC)
  return(c("PRECISION"=PREC,"RECALL"=REC,"F1"=F))
}

```

## Overall summary

Make a set of three barplots which summarise method performance over the range of group size and
delta values, for gene set sizes of 20, 50 and 200.

```{r,summary,fig.width=5,fig.height=4}

mar = c(5.1, 4.1, 5.1, 2.1)

x <- read.table("GSE158422_simulate_summary2_data.tsv",header=TRUE)
cols <- c("white","gray","black")
NAMES=c("GSA","LAT","LTT","LAW","LAM","ALT","ALW","ALM")

xa <- aggregate(. ~ Method , x , mean)
xao <- t(sapply(NAMES, function(y) { subset(xa,Method==y)}))
xao <- as.matrix(t(xao[,7:9]))
xao <- matrix(unlist(xao),nrow=3)
barplot(xao,beside=TRUE,ylim=c(0,1.2),names.arg=NAMES, ylab="index", col=cols, cex.names=0.7, cex.axis=0.7, main="Overall performance")
abline(h=seq(0,1,0.2),lty=3, lwd=0.9, col="gray")
legend("topleft", inset=.01, c("Precision","Recall","F1"), fill=cols, cex=0.75)

x1 <- subset(x,GPS==20)
x1 %>% kbl(caption="sim 20 res") %>% kable_paper("hover", full_width = F)
x1a <- aggregate(. ~ Method , x1 , mean)
x1ao <- t(sapply(NAMES, function(y) { subset(x1a,Method==y)}))
x1ao <- as.matrix(t(x1ao[,7:9]))
x1ao <- matrix(unlist(x1ao),nrow=3)
barplot(x1ao,beside=TRUE,ylim=c(0,1.2),names.arg=NAMES, ylab="index", col=cols, cex.names=0.7, cex.axis=0.7, main="20 genes per set")
abline(h=seq(0,1,0.2),lty=3, lwd=0.9, col="gray")
legend("topleft", inset=.01, c("Precision","Recall","F1"), fill=cols, cex=0.75)

x2 <- subset(x,GPS==50)
x2 %>% kbl(caption="sim 50 res") %>% kable_paper("hover", full_width = F)
x2a <- aggregate(. ~ Method , x2 , mean)
x2ao <- t(sapply(NAMES, function(y) { subset(x2a,Method==y)}))
x2ao <- as.matrix(t(x2ao[,7:9]))
x2ao <- matrix(unlist(x2ao),nrow=3)
barplot(x2ao,beside=TRUE,ylim=c(0,1.2),names.arg=NAMES, ylab="index", col=cols, cex.names=0.7, cex.axis=0.7, main="50 genes per set")
abline(h=seq(0,1,0.2),lty=3, lwd=0.9, col="gray")

x3 <- subset(x,GPS==100)
x3 %>% kbl(caption="sim 100 res") %>% kable_paper("hover", full_width = F)
x3a <- aggregate(. ~ Method , x3 , mean)
x3ao <- t(sapply(NAMES, function(y) { subset(x3a,Method==y)}))
x3ao <- as.matrix(t(x3ao[,7:9]))
x3ao <- matrix(unlist(x3ao),nrow=3)
barplot(x3ao,beside=TRUE,ylim=c(0,1.1),names.arg=NAMES, ylab="index", col=cols, cex.names=0.7, cex.axis=0.7, main="100 genes per set")
abline(h=seq(0,1,0.2),lty=3, lwd=0.9, col="gray")

y <- data.frame(x1ao[3,],x2ao[3,],x3ao[3,])
colnames(y) <- c("g20","g50","g100")
rownames(y) <- NAMES

cols <- c("white","lightgray","darkgray")
barplot(t(y),beside=TRUE,ylim=c(0,0.8),ylab="F1 index",col=cols,cex.names=0.7, cex.axis=0.7,main="Effect of gene set size")
abline(h=seq(0,1,0.2),lty=3, lwd=0.9, col="gray")
legend("topleft", inset=.01, c("20","50","100"), fill=cols, cex=0.75, title="Genes per set")

```

Make a PNG copy of the figure.

Experimenting with an alternate scheme for figure 1.

```{r,summary_png1}

# ALT FIG1
png(filename="altfig1.png",height=100,width=210,units="mm",res=200)

par(mar = c(2.5, 4.1, 2.1, 0.5))

par(mfrow=c(1,2))

cols <- c("white","gray","black")

barplot(xao,beside=TRUE,ylim=c(0,1),names.arg=NAMES, ylab="index", col=cols, cex.names=0.75, cex.axis=0.75, main="Overall performance")
abline(h=seq(0,1,0.2),lty=3, lwd=0.9, col="gray")
legend("topleft", inset=.01, c("Precision","Recall","F1"), fill=cols, cex=0.75)

cols <- c("white","lightgray","darkgray")
barplot(t(y),beside=TRUE,ylim=c(0,0.65),ylab="F1 index",col=cols,cex.names=0.75, cex.axis=0.75,main="Effect of gene set size")
abline(h=seq(0,1,0.2),lty=3, lwd=0.9, col="gray")
legend("topleft", inset=.01, c("20","50","100"), fill=cols, cex=0.75, title="Genes per set")

dev.off()

```

```{r,summary_png2}

# CANONICAL FIG1
png(filename="fig1.png",height=70,width=210,units="mm",res=200)

par(mar = c(2.5, 4.1, 2.1, 0.5))

par(mfrow=c(1,3))

cols <- c("white","gray","black")

barplot(x1ao,beside=TRUE,ylim=c(0,1),names.arg=NAMES, ylab="index", col=cols, cex.names=0.75, cex.axis=0.75, main="20 genes per set")
abline(h=seq(0,1,0.2),lty=3, lwd=0.9, col="gray")
legend("topleft", inset=.01, c("Precision","Recall","F1"), fill=cols, cex=0.85)

barplot(x2ao,beside=TRUE,ylim=c(0,1),names.arg=NAMES, ylab="index", col=cols, cex.names=0.75, cex.axis=0.75, main="50 genes per set")
abline(h=seq(0,1,0.2),lty=3, lwd=0.9, col="gray")

barplot(x3ao,beside=TRUE,ylim=c(0,1),names.arg=NAMES, ylab="index", col=cols, cex.names=0.75, cex.axis=0.75, main="100 genes per set")
abline(h=seq(0,1,0.2),lty=3, lwd=0.9, col="gray")

dev.off()

par(mfrow=c(1,1))

```

## In depth analysis of GSA vs LAM

Abandoned.

## Session information

```{r,save}

sessionInfo()

```

## Notes

LA

* simla: parametric self-contained

* simlac: parametric competitive

* simnla: nonparametric self-contained

* simnlac: nonparametric competitive

AL

* simal: parametric self-contained

* simalc: parametric competitive

* simnal: nonparametric self-contained

* simnalc: nonparametric competitive

### 2023-12-18 manuscript/manuscript.Rmd
fix results

@@
 Associations of epigenetic marks with gene expression are of great interest for understanding disease processes.
-Matching RNA-seq datasets were analysed and we conducted enrichment analysis of gene expression together with
+Matching tumour-normal RNA-seq datasets were analysed and we conducted enrichment analysis of gene expression together with
 DNA methylation.
-There were 12,386 differentially expressed genes, with 7,418 and 4,968 up- and down-regulated in the tumour group respectively.
-At the pathway level there were 318 and 193 up- and down-regulated Reactome pathways.
+There were 12,386 differentially expressed genes (FDR<0.05), with 7,418 and 4,968 up- and down-regulated in the tumour group respectively.
+At the pathway level there were 318 and 193 up- and down-regulated Reactome pathways with altered gene expression (FDR<0.05).
 As the context of gene methylation is important in influencing gene expression, promoter and gene body methylation
@@
 In promoters, GSA identified 15 and 8 pathways with higher and lower methylation respectively, while at gene bodies, there were were 17 and 67
-pathways with higher and lower methylation respectively.
-Using LAM, there were 133 and 12 pathways identified with higher and lower promoter methylation respectively, while there were 236 and 35 with higher and lower gene body methylation respectively.
+pathways with higher and lower methylation respectively (FDR<0.05).
+Using LAM, there were 133 and 12 pathways identified with higher and lower promoter methylation respectively, while there were 236 and 35 with higher and lower gene body methylation respectively (FDR<0.05).
 The relative overlap between promoter and gene body results was larger for LAM pathway sets as compared to GSA (*Figure 5A,B*).
@@
 
-## Exploration of methylation pathways altered by elevated homocysteine
+Two way analysis.
 
-##
+## Exploration of methylation pathways altered by elevated homocysteine or smoking
+
+Two way analysis.
 

### 2023-12-18 figs/GSE158422_pkgdev.Rmd
pkgdev example

@@
 
-This is the alternative aggregation approach which is faster but more complicated.
-
-```{r,mg2,eval=FALSE}
-
-tic()
-anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
-myann <- data.frame(anno[,c("UCSC_RefGene_Name","GencodeCompV12_Group","Islands_Name","Relation_to_Island")])
-gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
-gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
-names(gp2) <- rownames(gp)
-gp2 <- lapply(gp2,unique)
-gt <- stack(gp2)
-colnames(gt) <- c("gene","probe")
-dim(gt)
-str(gt)
-toc() #2.44, 2.41, 2.41s
-
-```
-
-##  Run mitch analysis.
-
 ```{r,mitch1}
@@
 table(anno$UCSC_RefGene_Group)
-table(unlist(strsplit(myann$GencodeCompV12_Group,";")))
+table(unlist(strsplit(myann$UCSC_RefGene_Group,";")))
 
@@
 keep=c("TSS200","TSS1500")
-annos <- strsplit(myann$GencodeCompV12_Group,";")
+annos <- strsplit(myann$UCSC_RefGene_Group,";")
 genes <- strsplit(myann$UCSC_RefGene_Name,";")
@@
 names(keeps) <- rownames(myann)
-#gt <- as.data.frame((unlist(gp2)))
 gt2 <- stack(gp2)
 colnames(gt2) <- c("gene","probe")
-toc() #20.37, 19.56, 21.23
+toc() #15.4 16.2 20.3
 

### 2023-12-18 figs/aging_example.Rmd
adding aging example

---
title: "Aging and DNA meth (https://doi.org/10.1186/s13073-019-0693-z)"
author: "The GMEA team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
---

## Introduction

Here we are looking at putting together the easiest possible workflow for the package.

## Requirements

```{r,packages}

suppressPackageStartupMessages({
  library("limma")
  library("eulerr")
  library("IlluminaHumanMethylation450kmanifest")
  library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
  library("tictoc")
  library("mitch")
  library("gplots")
  library("kableExtra")
  library("beeswarm")
  library("missMethyl")
  library("gridExtra")
  library("png")
})

CORES=16

```

## Load data

Reactome pathways were downloaded on the 14th Sept 2023 from MsigDB.

We'll also score the probes by the signed -log pvalue

```{r,load1}

if(!file.exists("aging_TableS1.csv")){
  options(timeout=1000000)
  download.file("https://datashare.ed.ac.uk/bitstream/handle/10283/3507/Supplementary_Table1_Limma_Age_EWAS_w1w3_Autosomes.csv?sequence=10&isAllowed=y",
    destfile="aging_TableS1.csv")
}

x <- read.csv("aging_TableS1.csv",header=TRUE)
head(x)
y1 <- x[,"Discovery.t",drop=FALSE]
rownames(y1) <- x$ProbeID
y2 <- x[,"Replication.t",drop=FALSE]
rownames(y2) <- x$ProbeID
head(y1)
head(y2)
l <- list("Discovery"=y1,"Replication"=y2)
str(l)

gs_symbols <- gmt_import("c2.cp.reactome.v2023.1.Hs.symbols.gmt")

```

## Curate the annotation

Use all probes on the chip.

```{r,anno1}

tic()
anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
myann <- data.frame(anno[,c("UCSC_RefGene_Name","UCSC_RefGene_Group","Islands_Name","Relation_to_Island")])
gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
names(gp2) <- rownames(gp)
gp2 <- lapply(gp2,unique)
gt <- stack(gp2)
colnames(gt) <- c("gene","probe")
gt$probe <- as.character(gt$probe)
dim(gt)
str(gt)
toc() #9.0s

```

## Merge gene name and aggregate

This only works properly when mitch's mapGeneIds function uses `mean` to aggregate (line 69 of mitch.R).

```{r,mg1}

# mitch aggregate function needs to be changes to `mean`
tic()
m <- mitch_import(x=l,DEtype="prescored",geneTable=gt)
toc() #9.0s
head(m)

```

```{r,mitch1}

tic()
mres <- mitch_calc(x=m,genesets=gs_symbols,minsetsize=5,cores=16, priority="effect")
toc() #11.45s

mtable <- mres$enrichment_result
head(mtable,20)

```

Rank-rank

```{r,rankrank}

mrank <- mres$ranked_profile

palette <- colorRampPalette(c("white", "yellow", "orange", "red", "darkred", "black"))
k <- MASS::kde2d(mrank[,1], mrank[,2])
X_AXIS = "Discovery"
Y_AXIS = "Replication"

filled.contour(k, color.palette = palette, plot.title = {
        abline(v = 0, h = 0, lty = 2, lwd = 2, col = "blue")
        title(main = "rank-rank plot", xlab = X_AXIS,
        ylab = Y_AXIS)
    })

pdf("fig6a.pdf")
palette <- colorRampPalette(c("white", "yellow", "orange", "red", "darkred", "black"))
k <- MASS::kde2d(mrank[,1], mrank[,2])
X_AXIS = "Discovery"
Y_AXIS = "Replication"

filled.contour(k, color.palette = palette, plot.title = {
        abline(v = 0, h = 0, lty = 2, lwd = 2, col = "blue")
        title(main = "rank-rank plot", xlab = X_AXIS,
        ylab = Y_AXIS)
    })
dev.off()

```

PW Scatter

```{r,pwscatter}

head(mtable)

sig <- subset(mtable,p.adjustMANOVA<0.05)
plot(mtable$s.Discovery,mtable$s.Replication,pch=19,col="gray",
  xlab="Discovery",ylab="Replication",main="pathway enrichment score")
points(sig$s.Discovery,sig$s.Replication,pch=19,col="red")
abline(v = 0, h = 0, lty = 2, lwd = 2, col = "blue")

pdf("fig6b.pdf")
plot(mtable$s.Discovery,mtable$s.Replication,pch=19,col="gray",
  xlab="Discovery",ylab="Replication",main="pathway enrichment score")
points(sig$s.Discovery,sig$s.Replication,pch=19,col="red")
abline(v = 0, h = 0, lty = 2, lwd = 2, col = "blue")
dev.off()

```

Heatmap

```{r,mheat}

mt <- head(mtable,30)
rownames(mt) <- mt$set
mt <- mt[,c("s.Discovery","s.Replication")]
rownames(mt) <- gsub("REACTOME_","",rownames(mt))
rownames(mt) <- gsub("_"," ",rownames(mt))
colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2(as.matrix(mt),scale="none",trace="none",margins=c(6,25),col=colfunc(25),cexRow=0.6,cexCol=1)

pdf("fig6c.pdf")
heatmap.2(as.matrix(mt),scale="none",trace="none",margins=c(6,25),col=colfunc(25),cexRow=0.6,cexCol=1)
dev.off()

```

Example 2d

```{r,example}

det <- mres$detailed_sets$REACTOME_CREATION_OF_C4_AND_C2_ACTIVATORS

k <- MASS::kde2d(det[,1], det[,2])
X_AXIS = "Discovery"
Y_AXIS = "Replication"

XMAX=max(mrank[,1])
XMIN=min(mrank[,1])
YMAX=max(mrank[,2])
YMIN=min(mrank[,2])

filled.contour(k, color.palette = palette, xlim=c(XMIN,XMAX), ylim=c(YMIN,YMAX),
      plot.title = {
        abline(v = 0, h = 0, lty = 2, lwd = 2, col = "blue")
        title(main = "Creation of C4 and C2 Activators", xlab = X_AXIS,
        ylab = Y_AXIS)
    })

pdf("fig6d.pdf")
filled.contour(k, color.palette = palette, xlim=c(XMIN,XMAX), ylim=c(YMIN,YMAX),
      plot.title = {
        abline(v = 0, h = 0, lty = 2, lwd = 2, col = "blue")
        title(main = "Creation of C4 and C2 Activators", xlab = X_AXIS,
        ylab = Y_AXIS)
    })
dev.off()

```

Make a html report and some charts.

```{r,report1}

mitch_report(res=mres,outfile="aging_mitchreport.html",overwrite=TRUE)
mitch_plots(res=mres,outfile="aging_mitchcharts.pdf")

```

## Session information

```{r,save}

save.image("aging_example.Rdata")

sessionInfo()

```

### 2023-12-18 manuscript/manuscript.Rmd
work on aging

@@
 True positives, false positives and false negatives were used to calculate overall precision and recall
-at these three gene set sizes (*Figure 1*).
+at these three gene set sizes (**Figure 1**).
 We found that overall recall was positively associated with the size of the gene sets considered.
@@
 Performance results for GSA and LAM methods with 100 genes per set shows that precision was similar, with most simulations yielding
->95% precision (*Figure 2*).
+>95% precision (**Figure 2**).
 LAM recall was superior for most simulations, particularly with lower group sizes.
@@
 
-LAM resulted in 358 significant Reactome pathways at FDR<0.05 while there were just 21 with GSA (*Figure 3A*).
+LAM resulted in 358 significant Reactome pathways at FDR<0.05 while there were just 21 with GSA (**Figure 3A**).
 The overlap of significant pathways was unexpectedly small, with just five common pathways.
@@
 method is higher with larger gene sets.
-Unexpectedly, pathways identified by LAM were collectively smaller than GSA (*Figure 3B*).
+Unexpectedly, pathways identified by LAM were collectively smaller than GSA (**Figure 3B**).
 Although algorithmic accuracy cannot be inferred from the types of pathways identified in real data,
-top ranked results from LAM appeared to be more related to cell differentiation, identity and development (*Figure 3C*),
-as compared to GSA (*Figure 3D*).
+top ranked results from LAM appeared to be more related to cell differentiation, identity and development (**Figure 3C**),
+as compared to GSA (**Figure 3D**).
 
@@
 group of 37 patients.
-This process was repeated 50 times and the results shown in (*Figure 4*).
+This process was repeated 50 times and the results shown in (**Figure 4**).
 Pathways that were identified in the smaller and full group were termed "consistent",
@@
 Using LAM, there were 133 and 12 pathways identified with higher and lower promoter methylation respectively, while there were 236 and 35 with higher and lower gene body methylation respectively (FDR<0.05).
-The relative overlap between promoter and gene body results was larger for LAM pathway sets as compared to GSA (*Figure 5A,B*).
-There was no observed overlap between GSA methylation-based pathways and gene expression pathways (*Figure 5A*), but there were some
-overlaps between pathways identified with LAM based promoter and gene body methylation and gene expression (*Figure 5B*).
+The relative overlap between promoter and gene body results was larger for LAM pathway sets as compared to GSA (**Figure 5A,B**).
+There was no observed overlap between GSA methylation-based pathways and gene expression pathways (**Figure 5A**), but there were some
+overlaps between pathways identified with LAM based promoter and gene body methylation and gene expression (**Figure 5B**).
 Interestingly, 123/318 upregulated gene expression pathways were associated with increasing gene body methylation with LAM.
 Likewise 19 pathways with downregulated expression were associated with increasing promoter methylation.
-Some of these pathways with large effect sizes across the three contrasts are depicted in heatmap form (*Figure 5C*).
+Some of these pathways with large effect sizes across the three contrasts are depicted in heatmap form (**Figure 5C**).
 Of these selected pathways, there was a mixture of concordant and discordant signatures in the three contrasts,
@@
 
+## Exploration of methylation pathways in aging
+
+We sought to explore the utility of LAM for exploring large EWAS datasets.
+One of the best studied associations with DNA methylation is with chronological aging, although
+pathway level differential methylation is not well defined.
+Using summary statistics from independent discovery and replication studies using the EPIC array platform described
+by McCartney et al (2020), we applied mitch to explore the pathway level differential methylation.
+In the discovery cohort we observed XX differential pathways (FDR), while in the replication group we observed XX.
+There was a high degree of agreement, with XX pathways significant in both groups.
+This is not surprising, as the gene-level differential methylation scores show a strong positive association between discovery and replication studies (**Figure 6A**).
+Using the MANOVA test, we identified XX pathways with altered methylation, most of which were concordant between
+discovery and replication (**Figure 6B**).
+When prioritising the results by S distance, a surrogate measure of effect size, 29 of the top 30 pathways were concordant in their direction of regulation (**Figure 6C**).
+Some of the observed methylation changes make sense with reports in the literature.
+For example the complement pathway is associated with aging and in this analysis we observe a reduction in methylation of genes beloning to this set including *C1R*, *MBL2*, *C1S* and *C1QC* (**Figure 6D**).
+
+![Figure 6. Pathway level DNA methylation alterations with chronological age. (A) Contour heatmap showing the similarity in gene methylation score ranks in the discovery and replication studies.
+(B) Mitch pathway enrichment scores (S distances) in discovery and replication studies. Pathways with MANOVA FDR<0.05 are shown in red while others are shaded gray.
+(C) Heatmap of 30 pathways with most extreme S distances (a surrogate of effect size). Red indicates increasing methylation and blue shows lower methylation.
+(D) An example of a pathway identified with this method, "creation of C4 and C2 activators" shows lower methylation of member genes in both discovery and replication studies.
+](../figs/fig5.png)
+
 ## Exploration of methylation pathways altered by ART

### 2023-12-18 figs/ART_novakovic.Rmd
novakovic working

---
title: "Novakovic et al 2019 reanalysis"
author: "The methylation in ART meta-analysis group"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 7
    fig_height: 7
  toc: true
theme: cosmo
---

```{r,packages}

source("meth_functions.R")

```

# Obtaining array annotations

```{r,annotation}

anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
myann <- data.frame(anno[,c("UCSC_RefGene_Name","Regulatory_Feature_Group","Islands_Name","Relation_to_Island")])
promoters <- grep("Prom",myann$Regulatory_Feature_Group)

```

## Introduction

This report is the reanalysis of first analysed DNA methylation data by
[Novakovic et al (2019)] (https://doi.org/10.1038/s41467-019-11929-9)

In this study,  DNA methylation status was generated for 149 neonatal (84♀ 65♂) and 158 adult (87♀ 71♂) 
ART-conceived individuals and for 58 neonatal (37♀, 21♂) and 75 adult (51♀, 24♂) non-ART conceived individuals.

```{r,download}

WORKING_DIR="GSE131433"
ARRAY_DATA="GSE131433_RAW.tar"
DEST=paste(WORKING_DIR,"/",ARRAY_DATA,sep="")
IDAT="GSE131433/IDAT"
options(timeout=1000000)

if(!dir.exists(WORKING_DIR)){
  dir.create(WORKING_DIR)
}

if(!file.exists(DEST)){
  download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE131433&format=file",
  #  download.file("ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE131nnn/GSE131433/suppl/GSE131433_RAW.tar",
    destfile=DEST)
  untar(exdir = "IDAT", tarfile = DEST)
}

SERIES_MATRIX=paste(WORKING_DIR,"/","GSE131433_series_matrix.txt.gz",sep="")
if(!file.exists(SERIES_MATRIX)){
  download.file("ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE131nnn/GSE131433/matrix/GSE131433_series_matrix.txt.gz", 
    destfile=SERIES_MATRIX)
}
gse <- getGEO(filename=SERIES_MATRIX)

baseDir <- "."
sample_metadata <- pData(phenoData(gse))
targets <- sample_metadata

files <- list.files(WORKING_DIR,pattern = "GSM",recursive = TRUE)
mybase <- unique(gsub("_Red.idat.gz" ,"", gsub("_Grn.idat.gz", "" ,files)))
mybase <- paste(WORKING_DIR,"/",mybase,sep="")
targets$Basename <- mybase
head(targets)

# filter time of birth as were not interested in follow up samples
targets <- subset(targets,`time of collection:ch1`=="birth")
rgSet <- read.metharray.exp(targets = targets)
rgSet

```

```{r,norm,fig.cap="Figure 1. Normalisation of bead-array data with SWAN."}

mSet <- preprocessRaw(rgSet)
mSetSw <- SWAN(mSet,verbose=FALSE)
par(mfrow=c(1,2), cex=0.8)
densityByProbeType(mSet[,1], main = "Raw")
densityByProbeType(mSetSw[,1], main = "SWAN")

```

```{r,filterprobes}

# include sex chromosomes
detP <- detectionP(rgSet)
keep <- rowSums(detP < 0.01) == ncol(rgSet)
mSetSw <- mSetSw[keep,]

# exclude SNP probes
mSetSw <- mapToGenome(mSetSw)
mSetSw_nosnp <- dropLociWithSnps(mSetSw)

# exclude sex chromosomes
xyprobes <- anno$Name[anno$chr %in% c("chrX","chrY")]
mSetFlt <- mSetSw[which(!rownames(mSetSw) %in% xyprobes),]

```

```{r,beta_m_vals}

# include sex chromosomes
meth <- getMeth(mSetSw)
unmeth <- getUnmeth(mSetSw)
Mval <- log2((meth + 100)/(unmeth + 100))
beta <- getBeta(mSetSw)
dim(Mval)

# exclude sex chromosomes
meth <- getMeth(mSetFlt)
unmeth <- getUnmeth(mSetFlt)
Mval_flt <- log2((meth + 100)/(unmeth + 100))
beta_flt <- getBeta(mSetFlt)
dim(Mval_flt)

```

# MDS analysis

```{r,scree,fig.cap="Figure 2. Screeplot shows contribution of top principal components to the overall variation in the experiment."}

par(mfrow=c(2,1))
myscree(Mval,main="incl sex chr")
myscree(Mval_flt,main="excl sex chr")

```

#MDS plot 1

```{r,mds1,fig.width = 8 ,fig.height = 8,fig.cap="Figure 3. MDS plot coloured by ART classification."}

colnames(targets) <- gsub(" ","_",colnames(targets))
colnames(targets) <- gsub(":","_",colnames(targets))
targets$sex<-factor(targets$`gender_ch1`)
targets$art<-factor(targets$`art_subtype`)
sample_groups <- factor(targets$art)
colour_palette=brewer.pal(n = length(levels(sample_groups)), name = "Paired")
colours <- colour_palette[as.integer(factor(targets$art))]
plot(1,axes = FALSE,xlab="",ylab="",main="MDS by ART type")
legend("center",legend=levels(sample_groups),pch=16,cex=1.2,col=colour_palette)
mydist <- plotMDS(Mval, labels=targets$Sample_Name,col=colours,main="sex chromosomes included")
mydist_flt <- plotMDS(Mval_flt, labels=targets$Sample_Name,col=colours,main="sex chromosomes excluded")
colnames(mydist_flt@.Data[[5]])  <- sapply(strsplit(colnames(mydist_flt@.Data[[5]]),"_"),"[[",1)
plotMDS(mydist_flt, labels=targets$Sample_Name,col=colours,main="sex chromosomes excluded")

```

#save data object

```{r,save}

save.image("Novakovic.Rdata")

```

#Differential analysis

1. non-ART Vs Fresh embryo

2. non-ART Vs Frozen embryo

3. non-ART Vs GIFT

4. non-ART Vs FX

5. GIFT Vs FX

6. non-ART Vs ART

7. Fresh Vs Frozen

### non-ART Vs Fresh embryo

```{r,dm1}

# sex chromosomes excluded
beta <- beta_flt
Mval <- Mval_flt
colnames(beta) <- sapply(strsplit(colnames(beta),"_"),"[[",1)
colnames(Mval) <- sapply(strsplit(colnames(Mval),"_"),"[[",1)

#targets$Basename <- gsub("IDAT/","",targets$Basename)  
birth<-targets[which(targets$`time_of_collection`=="birth"),]
birth$Basename <- sapply(strsplit(birth$Basename,"/"),"[[",3)
birth$Basename <- sapply(strsplit(birth$Basename,"_"),"[[",1)

samplesheet <- subset(birth,art_subtype_ch1=="non-ART"|art_subtype_ch1=="Fresh embryo")
groups <- factor(samplesheet$art_subtype,levels=c("non-ART","Fresh embryo"))
sex <- factor(samplesheet$sex,levels=c("M","F"))
Mvals<-Mval[,colnames(Mval)%in% samplesheet$Basename]
betas<-beta[,colnames(beta)%in% samplesheet$Basename]
dim(Mvals)
dim(betas)

top_nat_vs_fh <- dm_analysis(samplesheet=samplesheet,
  sex=sex,groups=groups,mx=Mvals,name="top_nat_vs_fh",
  myann=myann ,beta= betas) 
head(top_nat_vs_fh$dma,10)
head(top_nat_vs_fh$dmr,10)
top_nat_vs_fh$comp
top_nat_vs_fh$cgi
names(top_nat_vs_fh)
head(top_nat_vs_fh$confects$table$name)

# Allele
top_nat_vs_fh$dma$unmeth <- "T"
top_nat_vs_fh$dma$meth <- "C"
top_nat_vs_fh$fit$SE <- sqrt(top_nat_vs_fh$fit$s2.post) * top_nat_vs_fh$fit$stdev.unscaled

# Extract required columns from dma
top_nat_vs_fh_metal <-top_nat_vs_fh$dma[,c("Row.names", "meth", "unmeth", "AveExpr", "P.Value")]
head(top_nat_vs_fh_metal)

# Convert fit outputs to dataframes
fitCE <- as.data.frame(top_nat_vs_fh$fit$coefficients)
fitCE$Row.names <- row.names(fitCE)
fitCE <- fitCE[,c(4,3)]
names(fitCE)[2]<- "coefficient"

fitSE <- as.data.frame(top_nat_vs_fh$fit$SE)
fitSE$Row.names <- row.names(fitSE)
fitSE <- fitSE[,c(4,3)]
names(fitSE)[2]<- "SE"

# Merge Datasets
top_nat_vs_fh_metal <- merge(top_nat_vs_fh_metal, fitCE)
top_nat_vs_fh_metal <- merge(top_nat_vs_fh_metal, fitSE)

# Number of effective participants
Neff = 4/sum(1/table(groups))
top_nat_vs_fh_metal$N <- Neff

# Output for Meta-analysis
write.table(top_nat_vs_fh_metal, file="novakovic_top_nat_vs_fh_metal.tsv",
  sep="\t",quote=FALSE, row.names = FALSE)

head(top_nat_vs_fh$dma)
head(top_nat_vs_fh$dmr)
head(top_nat_vs_fh_metal)

saveRDS(top_nat_vs_fh,file="novakovic_nat_vs_fh.rds")

```

### non-ART Vs Frozen embryo

```{r,dm2}

samplesheet<-subset(birth,art_subtype_ch1=="non-ART"|art_subtype_ch1=="Frozen embryo")
groups <- factor(samplesheet$art_subtype,levels=c("non-ART","Frozen embryo"))
sex <- factor(samplesheet$sex,levels=c("M","F"))
Mvals<-Mval[,colnames(Mval)%in% samplesheet$Basename]
betas<-beta[,colnames(beta)%in% samplesheet$Basename]

top_nat_vs_fz <- dm_analysis(samplesheet=samplesheet,
  sex=sex,groups=groups,mx=Mvals,name="top_nat_vs_fz",
  myann=myann ,beta= betas)

head(top_nat_vs_fz$dma,10)
head(top_nat_vs_fz$dmr,10)
top_nat_vs_fz$comp
top_nat_vs_fz$cgi
names(top_nat_vs_fz)
head(top_nat_vs_fz$confects$table$name)

# Allele
top_nat_vs_fz$dma$unmeth <- "T"
top_nat_vs_fz$dma$meth <- "C"
top_nat_vs_fz$fit$SE <- sqrt(top_nat_vs_fz$fit$s2.post) * top_nat_vs_fz$fit$stdev.unscaled

# Extract required columns from dma
top_nat_vs_fz_metal <-top_nat_vs_fz$dma[,c("Row.names", "meth", "unmeth", "AveExpr", "P.Value")]
head(top_nat_vs_fz_metal)

# Convert fit outputs to dataframes
fitCE <- as.data.frame(top_nat_vs_fz$fit$coefficients)
fitCE$Row.names <- row.names(fitCE)
fitCE <- fitCE[,c(4,3)]
names(fitCE)[2]<- "coefficient"

fitSE <- as.data.frame(top_nat_vs_fz$fit$SE)
fitSE$Row.names <- row.names(fitSE)
fitSE <- fitSE[,c(4,3)]
names(fitSE)[2]<- "SE"

# Merge Datasets
top_nat_vs_fz_metal <- merge(top_nat_vs_fz_metal, fitCE)
top_nat_vs_fz_metal <- merge(top_nat_vs_fz_metal, fitSE)

# Number of effective participants
Neff = 4/sum(1/table(groups))
top_nat_vs_fz_metal$N <- Neff

# Output for Meta-analysis
write.table(top_nat_vs_fz_metal, file="novakovic_top_nat_vs_fz_metal.tsv",
  sep="\t",quote=FALSE, row.names = FALSE)

head(top_nat_vs_fz$dma)
head(top_nat_vs_fz$dmr)
head(top_nat_vs_fz_metal)

saveRDS(top_nat_vs_fz,file="novakovic_nat_vs_fz.rds")

```

### non-ART Vs GIFT

```{r,dm3}

samplesheet<-subset(birth,art_subtype_ch1=="non-ART"|art_subtype_ch1=="GIFT")
groups <- factor(samplesheet$art_subtype,levels=c("non-ART","GIFT"))
sex <- factor(samplesheet$sex,levels=c("M","F"))
Mvals<-Mval[,colnames(Mval)%in% samplesheet$Basename]
betas<-beta[,colnames(beta)%in% samplesheet$Basename]

top_nat_vs_GIFT <- dm_analysis(samplesheet=samplesheet,
sex=sex,groups=groups,mx=Mvals,name="top_nat_vs_GIFT",
myann=myann ,beta= betas)

head(top_nat_vs_GIFT$dma,10)
head(top_nat_vs_GIFT$dmr,10)

top_nat_vs_GIFT$comp
top_nat_vs_GIFT$cgi
names(top_nat_vs_GIFT)
head(top_nat_vs_GIFT$confects$table$name)

# Allele
top_nat_vs_GIFT$dma$unmeth <- "T"
top_nat_vs_GIFT$dma$meth <- "C"
top_nat_vs_GIFT$fit$SE <- sqrt(top_nat_vs_GIFT$fit$s2.post) * top_nat_vs_GIFT$fit$stdev.unscaled

# Extract required columns from dma
top_nat_vs_GIFT_metal <-top_nat_vs_GIFT$dma[,c("Row.names", "meth", "unmeth", "AveExpr", "P.Value")]
head(top_nat_vs_GIFT_metal)

# Convert fit outputs to dataframes
fitCE <- as.data.frame(top_nat_vs_GIFT$fit$coefficients)
fitCE$Row.names <- row.names(fitCE)
fitCE <- fitCE[,c(4,3)]
names(fitCE)[2]<- "coefficient"

fitSE <- as.data.frame(top_nat_vs_GIFT$fit$SE)
fitSE$Row.names <- row.names(fitSE)
fitSE <- fitSE[,c(4,3)]
names(fitSE)[2]<- "SE"

# Merge Datasets
top_nat_vs_GIFT_metal <- merge(top_nat_vs_GIFT_metal, fitCE)
top_nat_vs_GIFT_metal <- merge(top_nat_vs_GIFT_metal, fitSE)

# Number of effective participants
Neff = 4/sum(1/table(groups))
top_nat_vs_GIFT_metal$N <- Neff

# Output for Meta-analysis
write.table(top_nat_vs_GIFT_metal, file="novakovic_top_nat_vs_GIFT_metal.tsv",
  sep="\t",quote=FALSE, row.names = FALSE)

head(top_nat_vs_GIFT$dma)
head(top_nat_vs_GIFT$dmr)
head(top_nat_vs_GIFT_metal)

saveRDS(top_nat_vs_GIFT,file="novakovic_nat_vs_GIFT.rds")

```

### non-ART Vs FX

```{r,dm4}

samplesheet <- subset(birth,art_subtype_ch1=="non-ART"|art_subtype_ch1=="Frozen embryo" | art_subtype_ch1=="Fresh embryo")
samplesheet$art_subtype <- gsub("Fresh embryo","FX",samplesheet$art_subtype)
samplesheet$art_subtype <- gsub("Frozen embryo","FX",samplesheet$art_subtype)
groups <- factor(samplesheet$art_subtype,levels=c("non-ART","FX"))

sex <- factor(samplesheet$sex,levels=c("M","F"))
Mvals<-Mval[,colnames(Mval)%in% samplesheet$Basename]
betas<-beta[,colnames(beta)%in% samplesheet$Basename]

top_nat_vs_FX <- dm_analysis(samplesheet=samplesheet,
  sex=sex,groups=groups,mx=Mvals,name="top_nat_vs_FX",
  myann=myann ,beta= betas)

head(top_nat_vs_FX$dma,10)
head(top_nat_vs_FX$dmr,10)

top_nat_vs_FX$comp
top_nat_vs_FX$cgi
names(top_nat_vs_FX)
head(top_nat_vs_FX$confects$table$name)

# Allele
top_nat_vs_FX$dma$unmeth <- "T"
top_nat_vs_FX$dma$meth <- "C"
top_nat_vs_FX$fit$SE <- sqrt(top_nat_vs_FX$fit$s2.post) * top_nat_vs_FX$fit$stdev.unscaled

# Extract required columns from dma
top_nat_vs_FX_metal <-top_nat_vs_FX$dma[,c("Row.names", "meth", "unmeth", "AveExpr", "P.Value")]
head(top_nat_vs_FX_metal)

# Convert fit outputs to dataframes
fitCE <- as.data.frame(top_nat_vs_FX$fit$coefficients)
fitCE$Row.names <- row.names(fitCE)
fitCE <- fitCE[,c(4,3)]
names(fitCE)[2]<- "coefficient"

fitSE <- as.data.frame(top_nat_vs_FX$fit$SE)
fitSE$Row.names <- row.names(fitSE)
fitSE <- fitSE[,c(4,3)]
names(fitSE)[2]<- "SE"

# Merge Datasets
top_nat_vs_FX_metal <- merge(top_nat_vs_FX_metal, fitCE)
top_nat_vs_FX_metal <- merge(top_nat_vs_FX_metal, fitSE)

# Number of effective participants
Neff = 4/sum(1/table(groups))
top_nat_vs_FX_metal$N <- Neff

# Output for Meta-analysis
write.table(top_nat_vs_FX_metal, file="novakovic_top_nat_vs_FX_metal.tsv",sep="\t",quote=FALSE, row.names = FALSE)

head(top_nat_vs_FX$dma)
head(top_nat_vs_FX$dmr)
head(top_nat_vs_FX_metal)

saveRDS(top_nat_vs_FX,file="novakovic_nat_vs_FX.rds")

```

### GIFT Vs FX

```{r,dm5}

samplesheet <- subset(birth,art_subtype_ch1=="GIFT"|art_subtype_ch1=="Frozen embryo" | art_subtype_ch1=="Fresh embryo")
samplesheet$art_subtype <- gsub("Fresh embryo","FX",samplesheet$art_subtype)
samplesheet$art_subtype <- gsub("Frozen embryo","FX",samplesheet$art_subtype)
groups <- factor(samplesheet$art_subtype,levels=c("GIFT","FX"))
sex <- factor(samplesheet$sex,levels=c("M","F"))
Mvals<-Mval[,colnames(Mval)%in% samplesheet$Basename]
betas<-beta[,colnames(beta)%in% samplesheet$Basename]

top_GIFT_vs_FX <- dm_analysis(samplesheet=samplesheet,
sex=sex,groups=groups,mx=Mvals,name="top_GIFT_vs_FX",
myann=myann ,beta= betas)
head(top_GIFT_vs_FX$dma,10)
head(top_GIFT_vs_FX$dmr,10)
top_GIFT_vs_FX$comp
top_GIFT_vs_FX$cgi
names(top_GIFT_vs_FX)
head(top_GIFT_vs_FX$confects$table$name)

write.table(top_GIFT_vs_FX$dma,file="novakovic_top_GIFT_vs_FX.tsv",sep="\t",quote=FALSE)

# Allele
top_GIFT_vs_FX$dma$unmeth <- "T"
top_GIFT_vs_FX$dma$meth <- "C"
top_GIFT_vs_FX$fit$SE <- sqrt(top_GIFT_vs_FX$fit$s2.post) * top_GIFT_vs_FX$fit$stdev.unscaled

# Extract required columns from dma
top_GIFT_vs_FX_metal <-top_GIFT_vs_FX$dma[,c("Row.names", "meth", "unmeth", "AveExpr", "P.Value")]
head(top_GIFT_vs_FX_metal)

# Convert fit outputs to dataframes
fitCE <- as.data.frame(top_GIFT_vs_FX$fit$coefficients)
fitCE$Row.names <- row.names(fitCE)
fitCE <- fitCE[,c(4,3)]
names(fitCE)[2]<- "coefficient"

fitSE <- as.data.frame(top_GIFT_vs_FX$fit$SE)
fitSE$Row.names <- row.names(fitSE)
fitSE <- fitSE[,c(4,3)]
names(fitSE)[2]<- "SE"

# Merge Datasets
top_GIFT_vs_FX_metal <- merge(top_GIFT_vs_FX_metal, fitCE)
top_GIFT_vs_FX_metal <- merge(top_GIFT_vs_FX_metal, fitSE)

# Number of effective participants
Neff = 4/sum(1/table(groups))
top_GIFT_vs_FX_metal$N <- Neff

# Output for Meta-analysis
write.table(top_GIFT_vs_FX_metal, file="novakovic_top_GIFT_vs_FX_metal.tsv",sep="\t",quote=FALSE, row.names = FALSE)

head(top_GIFT_vs_FX$dma)
head(top_GIFT_vs_FX$dmr)
head(top_GIFT_vs_FX_metal)

saveRDS(top_GIFT_vs_FX,file="novakovic_GIFT_vs_FX.rds")

```

### non-ART Vs ART

```{r,dm6}

samplesheet<-birth
groups <- factor(samplesheet$art_status,levels=c("non-ART","ART"))
sex <- factor(samplesheet$sex,levels=c("M","F"))
Mvals<-Mval[,colnames(Mval)%in% samplesheet$Basename]
betas<-beta[,colnames(beta)%in% samplesheet$Basename]

top_nat_vs_ART <- dm_analysis(samplesheet=samplesheet,
sex=sex,groups=groups,mx=Mvals,name="top_nat_vs_ART",
myann=myann ,beta= betas)
head(top_nat_vs_ART$dma,10)
head(top_nat_vs_ART$dmr,10)
top_nat_vs_ART$comp
top_nat_vs_ART$cgi
names(top_nat_vs_ART)
head(top_nat_vs_ART$confects$table$name)

write.table(top_nat_vs_ART$dma,file="novakovic_top_nat_vs_ART.tsv",sep="\t",quote=FALSE)

# Allele
top_nat_vs_ART$dma$unmeth <- "T"
top_nat_vs_ART$dma$meth <- "C"
top_nat_vs_ART$fit$SE <- sqrt(top_nat_vs_ART$fit$s2.post) * top_nat_vs_ART$fit$stdev.unscaled

# Extract required columns from dma
top_nat_vs_ART_metal <-top_nat_vs_ART$dma[,c("Row.names", "meth", "unmeth", "AveExpr", "P.Value")]
head(top_nat_vs_ART_metal)

# Convert fit outputs to dataframes
fitCE <- as.data.frame(top_nat_vs_ART$fit$coefficients)
fitCE$Row.names <- row.names(fitCE)
fitCE <- fitCE[,c(4,3)]
names(fitCE)[2]<- "coefficient"

fitSE <- as.data.frame(top_nat_vs_ART$fit$SE)
fitSE$Row.names <- row.names(fitSE)
fitSE <- fitSE[,c(4,3)]
names(fitSE)[2]<- "SE"

# Merge Datasets
top_nat_vs_ART_metal <- merge(top_nat_vs_ART_metal, fitCE)
top_nat_vs_ART_metal <- merge(top_nat_vs_ART_metal, fitSE)

# Number of effective participants
Neff = 4/sum(1/table(groups))
top_nat_vs_ART_metal$N <- Neff

# Output for Meta-analysis
write.table(top_nat_vs_ART_metal, file="novakovic_top_nat_vs_ART_metal.tsv",sep="\t",quote=FALSE, row.names = FALSE)

head(top_nat_vs_ART$dma)
head(top_nat_vs_ART$dmr)
head(top_nat_vs_ART_metal)

saveRDS(top_nat_vs_ART,file="novakovic_nat_vs_ART.rds")

```

### Fresh embryo Vs Frozen embryo

```{r, dm7}

samplesheet <- subset(birth,art_subtype_ch1=="Fresh embryo"|art_subtype_ch1=="Frozen embryo")
groups <- factor(samplesheet$art_subtype,levels=c("Fresh embryo","Frozen embryo"))
sex <- factor(samplesheet$sex,levels=c("M","F"))
Mvals<-Mval[,colnames(Mval)%in% samplesheet$Basename]
betas<-beta[,colnames(beta)%in% samplesheet$Basename]

top_fh_vs_fz <- dm_analysis(samplesheet=samplesheet,
sex=sex,groups=groups,mx=Mvals,name="top_fh_vs_fz",
myann=myann ,beta= betas)
head(top_fh_vs_fz$dma,10)
head(top_fh_vs_fz$dmr,10)
top_fh_vs_fz$comp
top_fh_vs_fz$cgi
names(top_fh_vs_fz)
head(top_fh_vs_fz$confects$table$name)

# Allele
top_fh_vs_fz$dma$unmeth <- "T"
top_fh_vs_fz$dma$meth <- "C"
top_fh_vs_fz$fit$SE <- sqrt(top_fh_vs_fz$fit$s2.post) * top_fh_vs_fz$fit$stdev.unscaled

# Extract required columns from dma
top_fh_vs_fz_metal <-top_fh_vs_fz$dma[,c("Row.names", "meth", "unmeth", "AveExpr", "P.Value")]
head(top_fh_vs_fz_metal)

# Convert fit outputs to dataframes
fitCE <- as.data.frame(top_fh_vs_fz$fit$coefficients)
fitCE$Row.names <- row.names(fitCE)
fitCE <- fitCE[,c(4,3)]
names(fitCE)[2]<- "coefficient"

fitSE <- as.data.frame(top_fh_vs_fz$fit$SE)
fitSE$Row.names <- row.names(fitSE)
fitSE <- fitSE[,c(4,3)]
names(fitSE)[2]<- "SE"

# Merge Datasets
top_fh_vs_fz_metal <- merge(top_fh_vs_fz_metal, fitCE)
top_fh_vs_fz_metal <- merge(top_fh_vs_fz_metal, fitSE)

# Number of effective participants
Neff = 4/sum(1/table(groups))
top_fh_vs_fz_metal$N <- Neff

# Output for Meta-analysis
write.table(top_fh_vs_fz_metal, file="novakovic_top_fh_vs_fz_metal.tsv",sep="\t",quote=FALSE, row.names = FALSE)

head(top_fh_vs_fz$dma)
head(top_fh_vs_fz$dmr)
head(top_fh_vs_fz_metal)

saveRDS(top_fh_vs_fz,file="novakovic_fh_vs_fz.rds")

```

## Venn diagrams of the differential methylated probes for each contrast

```{r,venn1,fig.cap="Comparison of probes altered by fresh and frozen IVF procedures"}

v1 <- list("FH up" = top_nat_vs_fh$dm_up , 
           "FZ up" = top_nat_vs_fz$dm_up ,
           "FH dn" = top_nat_vs_fh$dm_dn ,
           "FZ dn" = top_nat_vs_fz$dm_dn )
plot(euler(v1, shape = "ellipse"), quantities = TRUE)

```

```{r,venn2,fig.cap="Comparison of probes altered by GIFT, fresh and frozen IVF procedures"}

v2 <- list("FH up" = top_nat_vs_fh$dm_up , 
           "FH dn" = top_nat_vs_fh$dm_dn ,
           "GIFT up" = top_nat_vs_GIFT$dm_up ,
           "GIFT dn" = top_nat_vs_GIFT$dm_dn )
plot(euler(v2, shape = "ellipse"), quantities = TRUE)

v3 <- list("FZ up" = top_nat_vs_fz$dm_up , 
           "FZ dn" = top_nat_vs_fz$dm_dn ,
           "GIFT up" = top_nat_vs_GIFT$dm_up ,
           "GIFT dn" = top_nat_vs_GIFT$dm_dn )
plot(euler(v3, shape = "ellipse"), quantities = FALSE)

```

## Spearman correlations of each contrast

```{r,spearman}

mycontrasts <- list("top_nat_vs_fh"=top_nat_vs_fh, "top_nat_vs_fz"=top_nat_vs_fz, "top_nat_vs_FX"=top_nat_vs_FX ,
  "top_nat_vs_GIFT"=top_nat_vs_GIFT, "top_fh_vs_fz"=top_fh_vs_fz, "top_GIFT_vs_FX"=top_GIFT_vs_FX, "top_nat_vs_ART"=top_nat_vs_ART)
myrnks <- lapply(X = mycontrasts, FUN = myranks)
df <- join_all(myrnks,by="rn")
rownames(df) <- df$rn
df$rn=NULL
colnames(df) <- names(mycontrasts)
head(df)
mycors <- cor(df,method = "spearman")
my_palette <- colorRampPalette(c("darkred","red", "orange", "yellow","white"))(n = 25)
heatmap.2(mycors,scale="none",margin=c(10, 10),cexRow=0.8,trace="none",cexCol=0.8,
    col=my_palette,main="Spearman correlations")
mycors

```

## Mitch analysis

```{r,mitch1d}

library (mitch)

if (! file.exists("ReactomePathways.gmt") ) {
  download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", 
  destfile="ReactomePathways.gmt.zip")
  unzip("ReactomePathways.gmt.zip",overwrite = TRUE)
}

genesets <- gmt_import("ReactomePathways.gmt")

top_nat_vs_fh_mitch  <- run_mitch_1d(dma=top_nat_vs_fh$dma,name="nat_vs_fh") 
head(top_nat_vs_fh_mitch,20)

top_nat_vs_fz_mitch  <- run_mitch_1d(dma=top_nat_vs_fz$dma,name="nat_vs_fz")
head(top_nat_vs_fz_mitch,20)

top_nat_vs_GIFT_mitch  <- run_mitch_1d(dma=top_nat_vs_GIFT$dma,name="nat_vs_GIFT")
head(top_nat_vs_GIFT_mitch,20)

top_nat_vs_FX_mitch  <- run_mitch_1d(dma=top_nat_vs_FX$dma,name="nat_vs_FX")
head(top_nat_vs_FX_mitch,20)

top_GIFT_vs_FX_mitch  <- run_mitch_1d(dma=top_GIFT_vs_FX$dma,name="GIFT_vs_FX")
head(top_GIFT_vs_FX_mitch,20)

top_nat_vs_ART_mitch  <- run_mitch_1d(dma=top_nat_vs_ART$dma,name="nat_vs_ART")
head(top_nat_vs_ART_mitch,20)

top_fh_vs_fz_mitch  <- run_mitch_1d(dma=top_fh_vs_fz$dma,name="fh_vs_fz")
head(top_fh_vs_fz_mitch,20)

```

## SessionInfo

```{r,session}

sessionInfo()

```

### 2023-12-18 figs/ART_novakovic.Rmd
simplify

@@
 
-## Mitch analysis
-
-```{r,mitch1d}
-
-library (mitch)
-
-if (! file.exists("ReactomePathways.gmt") ) {
-  download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", 
-  destfile="ReactomePathways.gmt.zip")
-  unzip("ReactomePathways.gmt.zip",overwrite = TRUE)
-}
-
-genesets <- gmt_import("ReactomePathways.gmt")
-
-top_nat_vs_fh_mitch  <- run_mitch_1d(dma=top_nat_vs_fh$dma,name="nat_vs_fh") 
-head(top_nat_vs_fh_mitch,20)
-
-top_nat_vs_fz_mitch  <- run_mitch_1d(dma=top_nat_vs_fz$dma,name="nat_vs_fz")
-head(top_nat_vs_fz_mitch,20)
-
-top_nat_vs_GIFT_mitch  <- run_mitch_1d(dma=top_nat_vs_GIFT$dma,name="nat_vs_GIFT")
-head(top_nat_vs_GIFT_mitch,20)
-
-top_nat_vs_FX_mitch  <- run_mitch_1d(dma=top_nat_vs_FX$dma,name="nat_vs_FX")
-head(top_nat_vs_FX_mitch,20)
-
-top_GIFT_vs_FX_mitch  <- run_mitch_1d(dma=top_GIFT_vs_FX$dma,name="GIFT_vs_FX")
-head(top_GIFT_vs_FX_mitch,20)
-
-top_nat_vs_ART_mitch  <- run_mitch_1d(dma=top_nat_vs_ART$dma,name="nat_vs_ART")
-head(top_nat_vs_ART_mitch,20)
-
-top_fh_vs_fz_mitch  <- run_mitch_1d(dma=top_fh_vs_fz$dma,name="fh_vs_fz")
-head(top_fh_vs_fz_mitch,20)
-
-```
-
 ## SessionInfo

### 2023-12-18 figs/ART_estill2016.Rmd
added estill 450k dataset

---
title: "Estill 2016 reanalysis"
author: "The methylation in ART meta-analysis group"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    fig_width: 7
    fig_height: 7
theme: cosmo
---

## Introduction

In this report, I will take you through a re-analysis methylation data first described by 
[Estill et al (2016)](https://www.ncbi.nlm.nih.gov/pubmed/27288894).

In their study, they analysed the DNA methylation patterns of 137 neonatal blood spots
conceived naturally (NAT), through through insemination (IUI), or through ICSI using fresh or
cryopreserved (frozen) embryo transfer (FH and FZ respectively).

The platform used in the study is the Illumina Infinium HumanMethylation450k BeadChip assay.
The authors used a pipeline based on ChAMP 
([Tian et al, 2017](https://www.ncbi.nlm.nih.gov/pubmed/28961746)), together with Adjacent
Site Clustering ([Sofer et al, 2013](https://www.ncbi.nlm.nih.gov/pubmed/23990415)).

The methylation data have been deposited to NCBI GEO repository accession number 
[GSE79257](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE79257).

The main conclusions from the original study were:

* The methylation profiles of assisted reproductive technology and IUI newborns were dramatically
different from those of naturally (in vivo) conceived newborns.

* Profiles of ICSI-frozen (FET) and IUI infants were strikingly similar, suggesting that
cryopreservation may temper some of the epigenetic aberrations induced by IVF or ICSI.

*  The DNA methylation changes associated with IVF/ICSI culture conditions and/or parental
infertility were detected at metastable epialleles, suggesting a lasting impact on a child's
epigenome.

* Both infertility and ICSI alter DNA methylation at specific genomic loci, an effect that 
is mitigated to some extent by FET (freezing).

The aim of this work is to;

1. develop the analytical pipelines required for efficient re-analysis of 450K array data,

2. to confirm that we are able to obtain differential methylation results that are similar
to those obtained in the original study, and

3. to critically evaluate the conclusions made in the original study.

In this report I will be using the missMethyl vignette as a guide to analyse this dataset 
([Phipson et al, 2015](https://www.ncbi.nlm.nih.gov/pubmed/26424855)).

Previously I curated a list of good and bad probes using the script "filter_probes.Rmd".

## Loading packages and functions

These packackes will help us to perform vital steps such as normalisation, filtering, 
differential analysis, etc, and provide information about the array probe annotaions.

These functions provide shortcuts to help with charts and other analysis. They will
eventually be shoved into another Rscript or package but can stay here for now.

```{r,packages}

source("meth_functions.R")

# Annotation
anno <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
anno <- data.frame(anno[,c("UCSC_RefGene_Name","Regulatory_Feature_Group",
  "Islands_Name","Relation_to_Island")])
promoters <- grep("Prom",myann$Regulatory_Feature_Group)

```

## Probe sets

For each gene, extract out the probes.

```{r,probesets}

promoters <- grep("Prom",anno$Regulatory_Feature_Group)
promoters <- anno[promoters,]
gp <- anno[,"UCSC_RefGene_Name",drop=FALSE]
gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
names(gp2) <- rownames(gp)
sets <- split(rep(names(gp2), lengths(gp2)), unlist(gp2))
summary(unlist(lapply(sets,length)))

```

## Data import

TODO- add explanation

```{r,data}

dir.create("GSE79257")

ARRAY_SAMPLESHEET="GSE79257/GSE79257_Illumina_samplesheet.csv"
# only download it if it is not present on the system
if ( !file.exists(ARRAY_SAMPLESHEET ) ) {
    DLFILE=paste(ARRAY_SAMPLESHEET,".gz",sep="")
    download.file("https://ftp.ncbi.nlm.nih.gov/geo/series/GSE79nnn/GSE79257/suppl/GSE79257_Illumina_samplesheet.csv.gz",
        destfile = DLFILE)
    gunzip(DLFILE, overwrite = TRUE )
}

ARRAY_DATA="GSE79257/GSE79257_RAW.tar"
# only download it if it is not present on the system
if ( !dir.exists("GSE79257/IDAT") ) {
  dir.create("GSE79257/IDAT")
  if ( !file.exists(ARRAY_DATA)  ) {
    system('wget -O idats.tar.gz "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE79257&format=file"')
    file.rename("idats.tar.gz",ARRAY_DATA)
    untar(exdir = "GSE79257/IDAT", tarfile = ARRAY_DATA)
  }
}

baseDir <- "GSE79257"
targets <- read.metharray.sheet(baseDir)
head(targets)
rgSet <- read.metharray.exp(targets = targets)

```

## Normalisation

```{r,norm,fig.cap="Figure 1. Normalisation of bead-array data with SWAN."}

mSet <- preprocessRaw(rgSet)
mSetSw <- SWAN(mSet,verbose=FALSE)
par(mfrow=c(1,2), cex=0.8)
densityByProbeType(mSet[,1], main = "Raw")
densityByProbeType(mSetSw[,1], main = "SWAN")

```

## Filter probes

Here we are running parallel analyses, both including and excluding sex chromosomes.

```{r,filterprobes}

# include sex chromosomes
detP <- detectionP(rgSet)
keep <- rowSums(detP < 0.01) == ncol(rgSet)
mSetSw <- mSetSw[keep,]

# exclude SNP probes
mSetSw <- mapToGenome(mSetSw)
mSetSw_nosnp <- dropLociWithSnps(mSetSw)
dim(mSetSw)
dim(mSetSw_nosnp)
mSetSw <- mSetSw_nosnp

# exclude sex chromosomes
keep <- !(featureNames(mSetSw) %in% ann450k$Name[ann450k$chr %in% c("chrX","chrY")])
mSetFlt <- mSetSw[keep,]
head(mSetFlt)
dim(mSetFlt)

```

## Extracting Beta and M-values

```{r,beta_m_vals}

# include sex chromosomes
meth <- getMeth(mSetSw)
unmeth <- getUnmeth(mSetSw)
Mval <- log2((meth + 100)/(unmeth + 100))
beta <- getBeta(mSetSw)

# exclude sex chromosomes
meth <- getMeth(mSetFlt)
unmeth <- getUnmeth(mSetFlt)
Mval_flt <- log2((meth + 100)/(unmeth + 100))
beta_flt <- getBeta(mSetFlt)

```

## MDS analysis

[Multidimensional scaling(https://en.wikipedia.org/wiki/Multidimensional_scaling)  plot is a method used
to identify the major sources of variation in a dataset. In the MDS plots below, I will be plotting the 
first two dimensions (principal components [PCs]), with each sample label coloured either by ART 
classification, sex, ART and sex, and then array chip and then sample plate.

We wil begin with MDS analysis including the sex chromosomes and then exclude them.

First, let's quantify the contribution of the major principal components. with a scree plot, we can see 
whether most of the variation is captured in the first two PCs or whether it is spread over more PCs.
As we can see in Figure 2, the main source of variation is what is shown in PC1, and a much lesser
extent on the other dimensions. Interestingly, excluding sex chromosomes does not seem to change the 
relative contributions of PCs very much.

```{r,scree,fig.cap="Figure 2. Screeplot shows contribution of top principal components to the overall variation in the experiment."}

par(mfrow=c(2,1))
myscree(Mval,main="incl sex chr")
myscree(Mval_flt,main="excl sex chr")

```

Here is the MDS plot by ART classification (Figure 3). You can see that there are four clusters 
of samples when MDS is projected this way. Natural birth infants (NAT) are mostly seen in the 
two clusters at the bottom of the chart and the IVF frozen, IVF fresh and in intrauterine 
insemination (IUI) appear mostly as two clusters at the top of the chart. IUI seem to have a
few samples that appear as intermediate between the upper and lower clusters.

When excluding sex chromosomes, the samples tend to form two clusters with the NAT samples on
the left of the chart and others on the right of the chart. IUI and FH have some samples which
have intermediate methylation, which points to a slight effect of freezing.

```{r,mds1,fig.width = 8 ,fig.height = 8,fig.cap="Figure 3. MDS plot coloured by ART classification."}

targets$sex <- factor(sapply(strsplit(targets$Sample_Group,"_"),"[[",2))
targets$art <- factor(sapply(strsplit(targets$Sample_Group,"_"),"[[",1))
sample_groups <- factor(targets$art)
colour_palette=brewer.pal(n = length(levels(sample_groups)), name = "Paired")
colours <- colour_palette[as.integer(factor(targets$art))]
plot(1,axes = FALSE,xlab="",ylab="",main="MDS by ART type")
legend("center",legend=levels(sample_groups),pch=16,cex=1.2,col=colour_palette)
mydist <- plotMDS(Mval, labels=targets$Sample_Name,col=colours,main="sex chromosomes included")
mydist_flt <- plotMDS(Mval_flt, labels=targets$Sample_Name,col=colours,main="sex chromosomes excluded")

```

Next, we created an MDS plot by sex (Figure 4). The female samples appear in the two cluster on the 
right hand side of the chart and the male sample on the left. This could be simply due to the
chromosomal makeup of the male and female samples or could also be due to autosomes. To check this,
the MDS plot needs to be repeated with autosomes excluded.

When excluding the sex chromosomes, there is no clear clustering of the samples by sex. It appears 
that ART classification is the dominant source of variation.

```{r,mds2,fig.width = 8 ,fig.height = 8, fig.cap="Figure 4. MDS plot coloured by sex."}

sample_groups <- factor(targets$sex)
colour_palette=brewer.pal(n = length(levels(sample_groups)), name = "Paired")
colours <- colour_palette[as.integer(factor(targets$sex))]
plot(1,axes = FALSE,xlab="",ylab="",main="MDS by sex")
legend("center",legend=levels(sample_groups),pch=16,cex=1.2,col=colour_palette)
plotMDS(mydist, labels=targets$Sample_Name,col=colours,main="sex chromosomes included")
plotMDS(mydist_flt, labels=targets$Sample_Name,col=colours,main="sex chromosomes excluded")

```

For completeness, we show the MDS by ART and sex (Figure 5). It confirms the trends seen above where the 
samples are split on the x axis by sex and on the y axis by ART classification.

When sex chromosomes are removed, it is clear that ART classification is the dominant source of variance.

```{r,mds3,fig.width = 8 ,fig.height = 8, fig.cap="Figure 5. MDS plot coloured by ART classification and sex."}

sample_groups <- factor(targets$Sample_Group)
colour_palette=brewer.pal(n = length(levels(sample_groups)), name = "Paired")
colours <- colour_palette[as.integer(factor(targets$Sample_Group))]
plot(1,axes = FALSE,xlab="",ylab="",main="MDS by ART and sex")
legend("center",legend=levels(sample_groups),pch=16,cex=1.2,col=colour_palette)
plotMDS(mydist, labels=targets$Sample_Name,col=colours,main="sex chromosomes included")
plotMDS(mydist_flt, labels=targets$Sample_Name,col=colours,main="sex chromosomes excluded")

```

To acertain whether technical factors like batch effects account for variance on the two main PCs, we
create an MDS by array chip (Figure 6). There appears to be no relationship between array chip number
and the top two PCs in the two MDS plots shown below.

```{r,mds4,fig.width = 8 ,fig.height = 8, fig.cap="Figure 6. MDS plot coloured by array chip."}

sample_groups <- factor(targets$Array)
colour_palette=brewer.pal(n = length(levels(sample_groups)), name = "Paired")
colours <- colour_palette[as.integer(factor(targets$Array))]
plot(1,axes = FALSE,xlab="",ylab="",main="MDS by array chip")
legend("center",legend=levels(sample_groups),pch=16,cex=1.2,col=colour_palette)
plotMDS(mydist, labels=targets$Sample_Name,col=colours,main="sex chromosomes included")
plotMDS(mydist_flt, labels=targets$Sample_Name,col=colours,main="sex chromosomes excluded")

```

This was also performed for sample plate (Figure 7). It appears that the laboratory did not randomise
samples over plates 1 and 2. For example the samples on plate 2 appear to be mostly NAT. This does
not appear to be a major concern because NAT samples from plate 2 seem to cluster with NAT samples on
plate 1 in both MDS plots.

```{r,mds5,fig.width = 8 ,fig.height = 8, fig.cap="Figure 7. MDS plot coloured by sample plate."}

sample_groups <- factor(targets$Sample_Plate)
colour_palette=brewer.pal(n = length(levels(sample_groups)), name = "Paired")
colours <- colour_palette[as.integer(factor(targets$Sample_Plate))]
plot(1,axes = FALSE,xlab="",ylab="",main="MDS by sample plate")
legend("center",legend=levels(sample_groups),pch=16,cex=1.2,col=colour_palette)
plotMDS(mydist, labels=targets$Sample_Name,col=colours,main="sex chromosomes included")
plotMDS(mydist_flt, labels=targets$Sample_Name,col=colours,main="sex chromosomes excluded")

```

## Differential analysis

There are several differential contrasts that would be of interest to us in this study:

* __NAT (43) vs FH (38)__: to see what probes are affected by IVF where the embryos are not frozen.

* __NAT (43) vs FZ (38)__: to see what probes are affected by IVF where the embryos are frozen.

* __NAT (43) vs IUI (18)__: It will be interesting to see if the probes affected by IVF are also sensitive to
IUI, a process that involves in utero injection of sperm.

* __FH (38) vs FZ (38)__: to see whether freezing results in differential methylation.

* __NAT (43) vs FH+FZ (76)__: if freezing results in only a small effect, then these groups can be 
lumped together and provide more power to determine subtle effects of IVF.

* __IUI (18) vs FH+FZ (76)__: It will be interesting to see whether the process of extracting eggs, culturing
and fertilising in vitro has an effect that is separate to IUI.

The differential analysis is centred around limma to identify differentially methylated probes. 
TopConfects was also run to obtain the probes with the largest confident effect (topconfect).
There are five outputs below:

1. Volcano plot (limma result).

2. Beeswarm plot (top probes by limma p-value).

3. Heatmap (top probes by limma p-value).

4. Beeswarm plot (top probes by topconfect ranking).

5. Heatmap (top probes by topconfect ranking). (TODO)

### NAT vs FH

In this first analysis I will look at the effect of natural conception versus IVF (fresh).
I have conducted the differential analysis in parallel, both includng and excluding sex
chromosomes. Even when excluding autosomes, there are ~2000 probes on the autosomes that 
show differential methylation between sexes. Based on this data and MDS analysis, we will
now exclude sex chromosomes from downstream analysis.

In the comparison of NAT and FH samples there were ~150k probes with differential methylation
(FDR<0.05), with ~120k loss of methylation and ~30k with increased methylation.

```{r,nat_vs_fh_incl, fig.width = 8 ,fig.height = 8 }

# include sex chromosomes
samplesheet <- subset(targets,art=="NAT" | art=="FH")
samplesheet$Basename <- sapply(strsplit(samplesheet$Basename, "/"), "[[", 3)
groups <- factor(samplesheet$art,levels=c("NAT","FH"))
sex <- factor(samplesheet$sex,levels=c("Male","Female"))

top_nat_vs_fh_inc <- dm_analysis(samplesheet=samplesheet,
    sex=sex,groups=groups,mx=Mval,name="top_nat_vs_fh_inc",
    myann=myann ,beta= beta)

head(top_nat_vs_fh_inc$dma)
head(top_nat_vs_fh_inc$dmr)
top_nat_vs_fh_inc$comp
top_nat_vs_fh_inc$cgi
names(top_nat_vs_fh_inc)
head(top_nat_vs_fh_inc$confects$table$name)

```

```{r,nat_vs_fh_excl, fig.width = 8 ,fig.height = 8}

# exclude sex chromosomes
samplesheet <- subset(targets,art=="NAT" | art=="FH")
samplesheet$Basename <- sapply(strsplit(samplesheet$Basename, "/"), "[[", 3)
groups <- factor(samplesheet$art,levels=c("NAT","FH"))
sex <- factor(samplesheet$sex,levels=c("Male","Female"))

top_nat_vs_fh <- dm_analysis(samplesheet=samplesheet,
    sex=sex,groups=groups,mx=Mval_flt,name="top_nat_vs_fh",
    myann=myann, beta=beta_flt)

head(top_nat_vs_fh$dma)
 head(top_nat_vs_fh$dmr)
top_nat_vs_fh$comp
top_nat_vs_fh$cgi
write.table(top_nat_vs_fh$dma,file="estill_nat_vs_fh.tsv",sep="\t",quote=FALSE)

```

### NAT vs FZ

In the comparison of NAT and FZ samples, there were ~130k probes with lower methylation
and ~37k with higher methylation.

```{r,nat_vs_fz, fig.width = 8 ,fig.height = 8}

samplesheet <- subset(targets,art=="NAT" | art=="FZ")
samplesheet$Basename <- sapply(strsplit(samplesheet$Basename, "/"), "[[", 3)
groups <- factor(samplesheet$art,levels=c("NAT","FZ"))
sex <- factor(samplesheet$sex,levels=c("Male","Female"))

top_nat_vs_fz <- dm_analysis(samplesheet=samplesheet,
    sex=sex,groups=groups,mx=Mval_flt,name="top_nat_vs_fz",
    myann=myann, beta=beta_flt)

head(top_nat_vs_fz$dma)
head(top_nat_vs_fz$dmr)
top_nat_vs_fz$comp
write.table(top_nat_vs_fz$dma,file="estill_nat_vs_fz.tsv",sep="\t",quote=FALSE)

```

### NAT vs IUI

In the comparison of NAT and IUI samples there were ~114k probes with reduced methylation
and ~36k with higher methylation.

```{r,nat_vs_iui, fig.width = 8 ,fig.height = 8}

samplesheet <- subset(targets,art=="NAT" | art=="IUI")
groups <- factor(samplesheet$art,levels=c("NAT","IUI"))
samplesheet$Basename <- sapply(strsplit(samplesheet$Basename, "/"), "[[", 3)
sex <- factor(samplesheet$sex,levels=c("Male","Female"))
table(groups)
top_nat_vs_iui <- dm_analysis(samplesheet=samplesheet,
    sex=sex,groups=groups,mx=Mval_flt,name="top_nat_vs_iui",
    myann=myann, beta=beta_flt)

head(top_nat_vs_iui$dma)
head(top_nat_vs_iui$dmr)
top_nat_vs_iui$comp
write.table(top_nat_vs_iui$dma,file="estill_nat_vs_iui.tsv",sep="\t",quote=FALSE)

```

### FH vs FZ

In the comparison of FH and FZ, only ~13k probes were differentially methylated with roughly
half lower and higher. This suggests that freezing has only a small effect on methylation.

```{r,fh_vs_fz, fig.width = 8 ,fig.height = 8}

samplesheet <- subset(targets,art=="FH" | art=="FZ")
groups <- factor(samplesheet$art,levels=c("FH","FZ"))
samplesheet$Basename <- sapply(strsplit(samplesheet$Basename, "/"), "[[", 3)
sex <- factor(samplesheet$sex,levels=c("Male","Female"))
table(groups)
top_fh_vs_fz <- dm_analysis(samplesheet=samplesheet,
    sex=sex,groups=groups,mx=Mval_flt,name="top_fh_vs_fz",
    myann=myann, beta=beta_flt)

head(top_fh_vs_fz$dma)
head(top_fh_vs_fz$dmr)
top_fh_vs_fz$comp
write.table(top_fh_vs_fz$dma,file="estill_fh_vs_fz.tsv",sep="\t",quote=FALSE)

```

### NAT vs FH/FZ

Because the effect of embryo freezing was found to be very small, we compared the NAT samples
to all IVF samples (union of FH and FZ). There were ~200k probes with altered methylation
~157k with lower methylation and ~50k higher. This result points to a major effect of IVF on
infant genome methylation.

```{r,nat_vs_fhfz, fig.width = 8 ,fig.height = 8}

samplesheet <- subset(targets,art=="NAT" | art=="FH" | art=="FZ")
samplesheet$Basename <- sapply(strsplit(samplesheet$Basename, "/"), "[[", 3)
samplesheet$art <- gsub("FH","FX",samplesheet$art)
samplesheet$art <- gsub("FZ","FX",samplesheet$art)
groups <- factor(samplesheet$art,levels=c("NAT","FX"))
sex <- factor(samplesheet$sex,levels=c("Male","Female"))
table(groups)
top_nat_vs_fx <- dm_analysis(samplesheet=samplesheet,
    sex=sex,groups=groups,mx=Mval_flt,name="top_nat_vs_fx",
    myann=myann, beta=beta_flt)

head(top_nat_vs_fx$dma)
head(top_nat_vs_fx$dmr)
top_nat_vs_fx$comp
write.table(top_nat_vs_fx$dma,file="estill_nat_vs_fx.tsv",sep="\t",quote=FALSE)

```

### IUI vs FH/FZ

We know that IVF conceived infants have a major shift in genome methylation, but we don't know
exactly whether it is caused by the IVF process itself. To eliminate whether it might be potentially
caused by subfertility we looked at whether the differences between IUI and FH+FZ were large.

We observe only ~1500 differentiallymethylated probes which is drastically smaller than the NAT vs
FH/FZ comparison which suggests that the process of IVF accounts for only a small fraction (1%) of the 
variance in methylation. The difference in methylomes could be due to other factors such as subfertility
or even the effect of hormone therapy.

```{r,iui_vs_fhfz, fig.width = 8 ,fig.height = 8}

samplesheet <- subset(targets,art=="IUI" | art=="FH" | art=="FZ")
samplesheet$Basename <- sapply(strsplit(samplesheet$Basename, "/"), "[[", 3)
samplesheet$art <- gsub("FH","FX",samplesheet$art)
samplesheet$art <- gsub("FZ","FX",samplesheet$art)
groups <- factor(samplesheet$art,levels=c("IUI","FX"))
sex <- factor(samplesheet$sex,levels=c("Male","Female"))
table(groups)
top_iui_vs_fx <- dm_analysis(samplesheet=samplesheet,
    sex=sex,groups=groups,mx=Mval_flt,name="top_iui_vs_fx",
    myann=myann, beta=beta_flt)

head(top_iui_vs_fx$dma)
head(top_iui_vs_fx$dmr)
top_iui_vs_fx$comp
write.table(top_iui_vs_fx$dma,file="estill_iui_vs_fx.tsv",sep="\t",quote=FALSE)

```

## Venn diagrams of the differential methylated probes for each contrast

First, we look at the similarity of DMPs altered by fresh and frozen procedures. The overlap is very large.
This indicates a high degree of similarity between the profiles.

```{r,venn1,fig.cap="Comparison of probes altered by fresh and frozen IVF procedures", fig.width = 8 ,fig.height = 8}

v1 <- list("FH up" = top_nat_vs_fh$dm_up , 
           "FZ up" = top_nat_vs_fz$dm_up ,
           "FH dn" = top_nat_vs_fh$dm_dn ,
           "FZ dn" = top_nat_vs_fz$dm_dn )
plot(euler(v1, shape = "ellipse"), quantities = TRUE)

```

Now to take a look at whether IUI is similar to IVF fresh and frozen. The overlap is just a large as 
FH vs FZ.

```{r,venn2,fig.cap="Comparison of probes altered by IUI, fresh and frozen IVF procedures", fig.width = 8 ,fig.height = 8}

v2 <- list("FH up" = top_nat_vs_fh$dm_up , 
           "FH dn" = top_nat_vs_fh$dm_dn ,
           "IUI up" = top_nat_vs_iui$dm_up ,
           "IUI dn" = top_nat_vs_iui$dm_dn )
plot(euler(v2, shape = "ellipse"), quantities = TRUE)

v3 <- list("FZ up" = top_nat_vs_fz$dm_up , 
           "FZ dn" = top_nat_vs_fz$dm_dn ,
           "IUI up" = top_nat_vs_iui$dm_up ,
           "IUI dn" = top_nat_vs_iui$dm_dn )
plot(euler(v3, shape = "ellipse"), quantities = FALSE)

```

## Spearman correlations of each contrast

While the above Venn diagrams are suggestive of similarity between contrasts, the best 
way to assess this is with correlation analysis. I have chosen Spearman as based on
the directional p-value ranking metric.

```{r,spearman, fig.width = 8 ,fig.height = 8}

mycontrasts <- list("top_nat_vs_fh"=top_nat_vs_fh, "top_nat_vs_fz"=top_nat_vs_fz, "top_nat_vs_fx"=top_nat_vs_fx ,
  "top_nat_vs_iui"=top_nat_vs_iui, "top_fh_vs_fz"=top_fh_vs_fz, "top_iui_vs_fx"=top_iui_vs_fx)

myrnks <- lapply(X = mycontrasts, FUN = myranks)
df <- join_all(myrnks,by="rn")
rownames(df) <- df$rn
df$rn=NULL
colnames(df) <- names(mycontrasts)
head(df)

mycors <- cor(df,method = "spearman")

my_palette <- colorRampPalette(c("darkred","red", "orange", "yellow","white"))(n = 25)
heatmap.2(mycors,scale="none",margin=c(10, 10),cexRow=0.8,trace="none",cexCol=0.8,
    col=my_palette,main="Spearman correlations")

mycors

```

## References

Estill MS, Bolnick JM, Waterland RA, Bolnick AD, Diamond MP, Krawetz SA.
Assisted reproductive technology alters deoxyribonucleic acid methylation
profiles in bloodspots of newborn infants. Fertil Steril. 2016 Sep
1;106(3):629-639.e10. doi: 10.1016/j.fertnstert.2016.05.006. Epub 2016 Jun 8.
PubMed PMID: 27288894.

Phipson B, Maksimovic J, Oshlack A. missMethyl: an R package for analyzing
data from Illumina's HumanMethylation450 platform. Bioinformatics. 2016 Jan
15;32(2):286-8. doi: 10.1093/bioinformatics/btv560. Epub 2015 Sep 30. PubMed
PMID: 26424855.

Sofer T, Schifano ED, Hoppin JA, Hou L, Baccarelli AA. A-clustering: a novel
method for the detection of co-regulated methylation regions, and regions
associated with exposure. Bioinformatics. 2013 Nov 15;29(22):2884-91. doi:
10.1093/bioinformatics/btt498. Epub 2013 Aug 29. PubMed PMID: 23990415; PubMed
Central PMCID: PMC3810849.

Tian Y, Morris TJ, Webster AP, Yang Z, Beck S, Feber A, Teschendorff AE.
ChAMP: updated methylation analysis pipeline for Illumina BeadChips.
Bioinformatics. 2017 Dec 15;33(24):3982-3984. doi: 10.1093/bioinformatics/btx513.
PubMed PMID: 28961746; PubMed Central PMCID: PMC5860089.

## Session info

```{r,sessioninfo}

sessionInfo()

```
END OF REPORT

### 2023-12-18 figs/ART_novakovic.Rmd
simplify

@@
 
-# Allele
-top_nat_vs_fh$dma$unmeth <- "T"
-top_nat_vs_fh$dma$meth <- "C"
-top_nat_vs_fh$fit$SE <- sqrt(top_nat_vs_fh$fit$s2.post) * top_nat_vs_fh$fit$stdev.unscaled
-
-# Extract required columns from dma
-top_nat_vs_fh_metal <-top_nat_vs_fh$dma[,c("Row.names", "meth", "unmeth", "AveExpr", "P.Value")]
-head(top_nat_vs_fh_metal)
-
-# Convert fit outputs to dataframes
-fitCE <- as.data.frame(top_nat_vs_fh$fit$coefficients)
-fitCE$Row.names <- row.names(fitCE)
-fitCE <- fitCE[,c(4,3)]
-names(fitCE)[2]<- "coefficient"
-
-fitSE <- as.data.frame(top_nat_vs_fh$fit$SE)
-fitSE$Row.names <- row.names(fitSE)
-fitSE <- fitSE[,c(4,3)]
-names(fitSE)[2]<- "SE"
-
-# Merge Datasets
-top_nat_vs_fh_metal <- merge(top_nat_vs_fh_metal, fitCE)
-top_nat_vs_fh_metal <- merge(top_nat_vs_fh_metal, fitSE)
-
-# Number of effective participants
-Neff = 4/sum(1/table(groups))
-top_nat_vs_fh_metal$N <- Neff
-
-# Output for Meta-analysis
-write.table(top_nat_vs_fh_metal, file="novakovic_top_nat_vs_fh_metal.tsv",
-  sep="\t",quote=FALSE, row.names = FALSE)
-
-head(top_nat_vs_fh$dma)
-head(top_nat_vs_fh$dmr)
-head(top_nat_vs_fh_metal)
-
 saveRDS(top_nat_vs_fh,file="novakovic_nat_vs_fh.rds")
@@
 head(top_nat_vs_fz$confects$table$name)
-
-# Allele
-top_nat_vs_fz$dma$unmeth <- "T"
-top_nat_vs_fz$dma$meth <- "C"
-top_nat_vs_fz$fit$SE <- sqrt(top_nat_vs_fz$fit$s2.post) * top_nat_vs_fz$fit$stdev.unscaled
-
-# Extract required columns from dma
-top_nat_vs_fz_metal <-top_nat_vs_fz$dma[,c("Row.names", "meth", "unmeth", "AveExpr", "P.Value")]
-head(top_nat_vs_fz_metal)
-
-# Convert fit outputs to dataframes
-fitCE <- as.data.frame(top_nat_vs_fz$fit$coefficients)
-fitCE$Row.names <- row.names(fitCE)
-fitCE <- fitCE[,c(4,3)]
-names(fitCE)[2]<- "coefficient"
-
-fitSE <- as.data.frame(top_nat_vs_fz$fit$SE)
-fitSE$Row.names <- row.names(fitSE)
-fitSE <- fitSE[,c(4,3)]
-names(fitSE)[2]<- "SE"
-
-# Merge Datasets
-top_nat_vs_fz_metal <- merge(top_nat_vs_fz_metal, fitCE)
-top_nat_vs_fz_metal <- merge(top_nat_vs_fz_metal, fitSE)
-
-# Number of effective participants
-Neff = 4/sum(1/table(groups))
-top_nat_vs_fz_metal$N <- Neff
-
-# Output for Meta-analysis
-write.table(top_nat_vs_fz_metal, file="novakovic_top_nat_vs_fz_metal.tsv",
-  sep="\t",quote=FALSE, row.names = FALSE)
-
-head(top_nat_vs_fz$dma)
-head(top_nat_vs_fz$dmr)
-head(top_nat_vs_fz_metal)
 
@@
 head(top_nat_vs_GIFT$confects$table$name)
-
-# Allele
-top_nat_vs_GIFT$dma$unmeth <- "T"
-top_nat_vs_GIFT$dma$meth <- "C"
-top_nat_vs_GIFT$fit$SE <- sqrt(top_nat_vs_GIFT$fit$s2.post) * top_nat_vs_GIFT$fit$stdev.unscaled
-
-# Extract required columns from dma
-top_nat_vs_GIFT_metal <-top_nat_vs_GIFT$dma[,c("Row.names", "meth", "unmeth", "AveExpr", "P.Value")]
-head(top_nat_vs_GIFT_metal)
-
-# Convert fit outputs to dataframes
-fitCE <- as.data.frame(top_nat_vs_GIFT$fit$coefficients)
-fitCE$Row.names <- row.names(fitCE)
-fitCE <- fitCE[,c(4,3)]
-names(fitCE)[2]<- "coefficient"
-
-fitSE <- as.data.frame(top_nat_vs_GIFT$fit$SE)
-fitSE$Row.names <- row.names(fitSE)
-fitSE <- fitSE[,c(4,3)]
-names(fitSE)[2]<- "SE"
-
-# Merge Datasets
-top_nat_vs_GIFT_metal <- merge(top_nat_vs_GIFT_metal, fitCE)
-top_nat_vs_GIFT_metal <- merge(top_nat_vs_GIFT_metal, fitSE)
-
-# Number of effective participants
-Neff = 4/sum(1/table(groups))
-top_nat_vs_GIFT_metal$N <- Neff
-
-# Output for Meta-analysis
-write.table(top_nat_vs_GIFT_metal, file="novakovic_top_nat_vs_GIFT_metal.tsv",
-  sep="\t",quote=FALSE, row.names = FALSE)
-
-head(top_nat_vs_GIFT$dma)
-head(top_nat_vs_GIFT$dmr)
-head(top_nat_vs_GIFT_metal)
 
@@
 
-# Allele
-top_nat_vs_FX$dma$unmeth <- "T"
-top_nat_vs_FX$dma$meth <- "C"
-top_nat_vs_FX$fit$SE <- sqrt(top_nat_vs_FX$fit$s2.post) * top_nat_vs_FX$fit$stdev.unscaled
-
-# Extract required columns from dma
-top_nat_vs_FX_metal <-top_nat_vs_FX$dma[,c("Row.names", "meth", "unmeth", "AveExpr", "P.Value")]
-head(top_nat_vs_FX_metal)
-
-# Convert fit outputs to dataframes
-fitCE <- as.data.frame(top_nat_vs_FX$fit$coefficients)
-fitCE$Row.names <- row.names(fitCE)
-fitCE <- fitCE[,c(4,3)]
-names(fitCE)[2]<- "coefficient"
-
-fitSE <- as.data.frame(top_nat_vs_FX$fit$SE)
-fitSE$Row.names <- row.names(fitSE)
-fitSE <- fitSE[,c(4,3)]
-names(fitSE)[2]<- "SE"
-
-# Merge Datasets
-top_nat_vs_FX_metal <- merge(top_nat_vs_FX_metal, fitCE)
-top_nat_vs_FX_metal <- merge(top_nat_vs_FX_metal, fitSE)
-
-# Number of effective participants
-Neff = 4/sum(1/table(groups))
-top_nat_vs_FX_metal$N <- Neff
-
-# Output for Meta-analysis
-write.table(top_nat_vs_FX_metal, file="novakovic_top_nat_vs_FX_metal.tsv",sep="\t",quote=FALSE, row.names = FALSE)
-
-head(top_nat_vs_FX$dma)
-head(top_nat_vs_FX$dmr)
-head(top_nat_vs_FX_metal)
-
 saveRDS(top_nat_vs_FX,file="novakovic_nat_vs_FX.rds")
@@
 
-# Allele
-top_GIFT_vs_FX$dma$unmeth <- "T"
-top_GIFT_vs_FX$dma$meth <- "C"
-top_GIFT_vs_FX$fit$SE <- sqrt(top_GIFT_vs_FX$fit$s2.post) * top_GIFT_vs_FX$fit$stdev.unscaled
-
-# Extract required columns from dma
-top_GIFT_vs_FX_metal <-top_GIFT_vs_FX$dma[,c("Row.names", "meth", "unmeth", "AveExpr", "P.Value")]
-head(top_GIFT_vs_FX_metal)
-
-# Convert fit outputs to dataframes
-fitCE <- as.data.frame(top_GIFT_vs_FX$fit$coefficients)
-fitCE$Row.names <- row.names(fitCE)
-fitCE <- fitCE[,c(4,3)]
-names(fitCE)[2]<- "coefficient"
-
-fitSE <- as.data.frame(top_GIFT_vs_FX$fit$SE)
-fitSE$Row.names <- row.names(fitSE)
-fitSE <- fitSE[,c(4,3)]
-names(fitSE)[2]<- "SE"
-
-# Merge Datasets
-top_GIFT_vs_FX_metal <- merge(top_GIFT_vs_FX_metal, fitCE)
-top_GIFT_vs_FX_metal <- merge(top_GIFT_vs_FX_metal, fitSE)
-
-# Number of effective participants
-Neff = 4/sum(1/table(groups))
-top_GIFT_vs_FX_metal$N <- Neff
-
-# Output for Meta-analysis
-write.table(top_GIFT_vs_FX_metal, file="novakovic_top_GIFT_vs_FX_metal.tsv",sep="\t",quote=FALSE, row.names = FALSE)
-
-head(top_GIFT_vs_FX$dma)
-head(top_GIFT_vs_FX$dmr)
-head(top_GIFT_vs_FX_metal)
-
 saveRDS(top_GIFT_vs_FX,file="novakovic_GIFT_vs_FX.rds")
@@
 
-# Allele
-top_nat_vs_ART$dma$unmeth <- "T"
-top_nat_vs_ART$dma$meth <- "C"
-top_nat_vs_ART$fit$SE <- sqrt(top_nat_vs_ART$fit$s2.post) * top_nat_vs_ART$fit$stdev.unscaled
-
-# Extract required columns from dma
-top_nat_vs_ART_metal <-top_nat_vs_ART$dma[,c("Row.names", "meth", "unmeth", "AveExpr", "P.Value")]
-head(top_nat_vs_ART_metal)
-
-# Convert fit outputs to dataframes
-fitCE <- as.data.frame(top_nat_vs_ART$fit$coefficients)
-fitCE$Row.names <- row.names(fitCE)
-fitCE <- fitCE[,c(4,3)]
-names(fitCE)[2]<- "coefficient"
-
-fitSE <- as.data.frame(top_nat_vs_ART$fit$SE)
-fitSE$Row.names <- row.names(fitSE)
-fitSE <- fitSE[,c(4,3)]
-names(fitSE)[2]<- "SE"
-
-# Merge Datasets
-top_nat_vs_ART_metal <- merge(top_nat_vs_ART_metal, fitCE)
-top_nat_vs_ART_metal <- merge(top_nat_vs_ART_metal, fitSE)
-
-# Number of effective participants
-Neff = 4/sum(1/table(groups))
-top_nat_vs_ART_metal$N <- Neff
-
-# Output for Meta-analysis
-write.table(top_nat_vs_ART_metal, file="novakovic_top_nat_vs_ART_metal.tsv",sep="\t",quote=FALSE, row.names = FALSE)
-
-head(top_nat_vs_ART$dma)
-head(top_nat_vs_ART$dmr)
-head(top_nat_vs_ART_metal)
-
 saveRDS(top_nat_vs_ART,file="novakovic_nat_vs_ART.rds")
@@
 
-# Allele
-top_fh_vs_fz$dma$unmeth <- "T"
-top_fh_vs_fz$dma$meth <- "C"
-top_fh_vs_fz$fit$SE <- sqrt(top_fh_vs_fz$fit$s2.post) * top_fh_vs_fz$fit$stdev.unscaled
-
-# Extract required columns from dma
-top_fh_vs_fz_metal <-top_fh_vs_fz$dma[,c("Row.names", "meth", "unmeth", "AveExpr", "P.Value")]
-head(top_fh_vs_fz_metal)
-
-# Convert fit outputs to dataframes
-fitCE <- as.data.frame(top_fh_vs_fz$fit$coefficients)
-fitCE$Row.names <- row.names(fitCE)
-fitCE <- fitCE[,c(4,3)]
-names(fitCE)[2]<- "coefficient"
-
-fitSE <- as.data.frame(top_fh_vs_fz$fit$SE)
-fitSE$Row.names <- row.names(fitSE)
-fitSE <- fitSE[,c(4,3)]
-names(fitSE)[2]<- "SE"
-
-# Merge Datasets
-top_fh_vs_fz_metal <- merge(top_fh_vs_fz_metal, fitCE)
-top_fh_vs_fz_metal <- merge(top_fh_vs_fz_metal, fitSE)
-
-# Number of effective participants
-Neff = 4/sum(1/table(groups))
-top_fh_vs_fz_metal$N <- Neff
-
-# Output for Meta-analysis
-write.table(top_fh_vs_fz_metal, file="novakovic_top_fh_vs_fz_metal.tsv",sep="\t",quote=FALSE, row.names = FALSE)
-
-head(top_fh_vs_fz$dma)
-head(top_fh_vs_fz$dmr)
-head(top_fh_vs_fz_metal)
-
 saveRDS(top_fh_vs_fz,file="novakovic_fh_vs_fz.rds")
-
-```
-
-## Venn diagrams of the differential methylated probes for each contrast
-
-```{r,venn1,fig.cap="Comparison of probes altered by fresh and frozen IVF procedures"}
-
-v1 <- list("FH up" = top_nat_vs_fh$dm_up , 
-           "FZ up" = top_nat_vs_fz$dm_up ,
-           "FH dn" = top_nat_vs_fh$dm_dn ,
-           "FZ dn" = top_nat_vs_fz$dm_dn )
-plot(euler(v1, shape = "ellipse"), quantities = TRUE)
-
-```
-
-```{r,venn2,fig.cap="Comparison of probes altered by GIFT, fresh and frozen IVF procedures"}
-
-v2 <- list("FH up" = top_nat_vs_fh$dm_up , 
-           "FH dn" = top_nat_vs_fh$dm_dn ,
-           "GIFT up" = top_nat_vs_GIFT$dm_up ,
-           "GIFT dn" = top_nat_vs_GIFT$dm_dn )
-plot(euler(v2, shape = "ellipse"), quantities = TRUE)
-
-v3 <- list("FZ up" = top_nat_vs_fz$dm_up , 
-           "FZ dn" = top_nat_vs_fz$dm_dn ,
-           "GIFT up" = top_nat_vs_GIFT$dm_up ,
-           "GIFT dn" = top_nat_vs_GIFT$dm_dn )
-plot(euler(v3, shape = "ellipse"), quantities = FALSE)
-
-```
-
-## Spearman correlations of each contrast
-
-```{r,spearman}
-
-mycontrasts <- list("top_nat_vs_fh"=top_nat_vs_fh, "top_nat_vs_fz"=top_nat_vs_fz, "top_nat_vs_FX"=top_nat_vs_FX ,
-  "top_nat_vs_GIFT"=top_nat_vs_GIFT, "top_fh_vs_fz"=top_fh_vs_fz, "top_GIFT_vs_FX"=top_GIFT_vs_FX, "top_nat_vs_ART"=top_nat_vs_ART)
-myrnks <- lapply(X = mycontrasts, FUN = myranks)
-df <- join_all(myrnks,by="rn")
-rownames(df) <- df$rn
-df$rn=NULL
-colnames(df) <- names(mycontrasts)
-head(df)
-mycors <- cor(df,method = "spearman")
-my_palette <- colorRampPalette(c("darkred","red", "orange", "yellow","white"))(n = 25)
-heatmap.2(mycors,scale="none",margin=c(10, 10),cexRow=0.8,trace="none",cexCol=0.8,
-    col=my_palette,main="Spearman correlations")
-mycors
 

### 2023-12-18 figs/ART_estill2016.Rmd
fix anno

@@
 # Annotation
-anno <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
-anno <- data.frame(anno[,c("UCSC_RefGene_Name","Regulatory_Feature_Group",
+annofull <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
+anno <- data.frame(annofull[,c("UCSC_RefGene_Name","Regulatory_Feature_Group",
   "Islands_Name","Relation_to_Island")])
-promoters <- grep("Prom",myann$Regulatory_Feature_Group)
-
-```
-
-## Probe sets
-
-For each gene, extract out the probes.
-
-```{r,probesets}
-
-promoters <- grep("Prom",anno$Regulatory_Feature_Group)
-promoters <- anno[promoters,]
-gp <- anno[,"UCSC_RefGene_Name",drop=FALSE]
-gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
-names(gp2) <- rownames(gp)
-sets <- split(rep(names(gp2), lengths(gp2)), unlist(gp2))
-summary(unlist(lapply(sets,length)))
 
@@
 ## Data import
-
-TODO- add explanation
 
@@
     DLFILE=paste(ARRAY_SAMPLESHEET,".gz",sep="")
+    options(timeout=1000000)
     download.file("https://ftp.ncbi.nlm.nih.gov/geo/series/GSE79nnn/GSE79257/suppl/GSE79257_Illumina_samplesheet.csv.gz",
@@
 # exclude sex chromosomes
-keep <- !(featureNames(mSetSw) %in% ann450k$Name[ann450k$chr %in% c("chrX","chrY")])
+keep <- !(featureNames(mSetSw) %in% annofull$Name[annofull$chr %in% c("chrX","chrY")])
 mSetFlt <- mSetSw[keep,]
@@
 top_nat_vs_fh$cgi
-write.table(top_nat_vs_fh$dma,file="estill_nat_vs_fh.tsv",sep="\t",quote=FALSE)
+
+saveRDS(top_nat_vs_fh,file="estill_nat_vs_fh.rds")
 
@@
 top_nat_vs_fz$comp
-write.table(top_nat_vs_fz$dma,file="estill_nat_vs_fz.tsv",sep="\t",quote=FALSE)
+
+saveRDS(top_nat_vs_fz,file="estill_nat_vs_fz.rds")
 
@@
 
+saveRDS(top_nat_vs_iui,file="estill_nat_vs_iui.rds")
+
 ```
@@
 top_fh_vs_fz$comp
-write.table(top_fh_vs_fz$dma,file="estill_fh_vs_fz.tsv",sep="\t",quote=FALSE)
+
+saveRDS(top_fh_vs_fz,file="estill_fh_vs_fz.rds")
 
@@
 top_nat_vs_fx$comp
-write.table(top_nat_vs_fx$dma,file="estill_nat_vs_fx.tsv",sep="\t",quote=FALSE)
+
+saveRDS(top_nat_vs_fx,file="estill_nat_vs_fx.rds")
 
@@
 top_iui_vs_fx$comp
-write.table(top_iui_vs_fx$dma,file="estill_iui_vs_fx.tsv",sep="\t",quote=FALSE)
 
-```
-
-## Venn diagrams of the differential methylated probes for each contrast
-
-First, we look at the similarity of DMPs altered by fresh and frozen procedures. The overlap is very large.
-This indicates a high degree of similarity between the profiles.
-
-```{r,venn1,fig.cap="Comparison of probes altered by fresh and frozen IVF procedures", fig.width = 8 ,fig.height = 8}
-
-v1 <- list("FH up" = top_nat_vs_fh$dm_up , 
-           "FZ up" = top_nat_vs_fz$dm_up ,
-           "FH dn" = top_nat_vs_fh$dm_dn ,
-           "FZ dn" = top_nat_vs_fz$dm_dn )
-plot(euler(v1, shape = "ellipse"), quantities = TRUE)
-
-```
-
-Now to take a look at whether IUI is similar to IVF fresh and frozen. The overlap is just a large as 
-FH vs FZ.
-
-```{r,venn2,fig.cap="Comparison of probes altered by IUI, fresh and frozen IVF procedures", fig.width = 8 ,fig.height = 8}
-
-v2 <- list("FH up" = top_nat_vs_fh$dm_up , 
-           "FH dn" = top_nat_vs_fh$dm_dn ,
-           "IUI up" = top_nat_vs_iui$dm_up ,
-           "IUI dn" = top_nat_vs_iui$dm_dn )
-plot(euler(v2, shape = "ellipse"), quantities = TRUE)
-
-v3 <- list("FZ up" = top_nat_vs_fz$dm_up , 
-           "FZ dn" = top_nat_vs_fz$dm_dn ,
-           "IUI up" = top_nat_vs_iui$dm_up ,
-           "IUI dn" = top_nat_vs_iui$dm_dn )
-plot(euler(v3, shape = "ellipse"), quantities = FALSE)
-
-```
-
-## Spearman correlations of each contrast
-
-While the above Venn diagrams are suggestive of similarity between contrasts, the best 
-way to assess this is with correlation analysis. I have chosen Spearman as based on
-the directional p-value ranking metric.
-
-```{r,spearman, fig.width = 8 ,fig.height = 8}
-
-mycontrasts <- list("top_nat_vs_fh"=top_nat_vs_fh, "top_nat_vs_fz"=top_nat_vs_fz, "top_nat_vs_fx"=top_nat_vs_fx ,
-  "top_nat_vs_iui"=top_nat_vs_iui, "top_fh_vs_fz"=top_fh_vs_fz, "top_iui_vs_fx"=top_iui_vs_fx)
-
-myrnks <- lapply(X = mycontrasts, FUN = myranks)
-df <- join_all(myrnks,by="rn")
-rownames(df) <- df$rn
-df$rn=NULL
-colnames(df) <- names(mycontrasts)
-head(df)
-
-mycors <- cor(df,method = "spearman")
-
-my_palette <- colorRampPalette(c("darkred","red", "orange", "yellow","white"))(n = 25)
-heatmap.2(mycors,scale="none",margin=c(10, 10),cexRow=0.8,trace="none",cexCol=0.8,
-    col=my_palette,main="Spearman correlations")
-
-mycors
+saveRDS(top_iui_vs_fx,file="estill_iui_vs_fx.rds")
 

### 2023-12-18 figs/ART_estill2016.Rmd
fix anno

@@
     sex=sex,groups=groups,mx=Mval,name="top_nat_vs_fh_inc",
-    myann=myann ,beta= beta)
+    myann=anno ,beta= beta)
 
@@
     sex=sex,groups=groups,mx=Mval_flt,name="top_nat_vs_fh",
-    myann=myann, beta=beta_flt)
+    myann=anno, beta=beta_flt)
 
@@
     sex=sex,groups=groups,mx=Mval_flt,name="top_nat_vs_fz",
-    myann=myann, beta=beta_flt)
+    myann=anno, beta=beta_flt)
 
@@
     sex=sex,groups=groups,mx=Mval_flt,name="top_nat_vs_iui",
-    myann=myann, beta=beta_flt)
+    myann=anno, beta=beta_flt)
 
@@
     sex=sex,groups=groups,mx=Mval_flt,name="top_fh_vs_fz",
-    myann=myann, beta=beta_flt)
+    myann=anno, beta=beta_flt)
 
@@
     sex=sex,groups=groups,mx=Mval_flt,name="top_nat_vs_fx",
-    myann=myann, beta=beta_flt)
+    myann=anno, beta=beta_flt)
 
@@
     sex=sex,groups=groups,mx=Mval_flt,name="top_iui_vs_fx",
-    myann=myann, beta=beta_flt)
+    myann=anno, beta=beta_flt)
 

### 2023-12-19 figs/art_example.Rmd
integrative example

---
title: "DNA meth and IVF"
author: "The GMEA team and ART epigenetics team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
---

## Introduction

Here we are looking at putting together the easiest possible workflow for the package.

We will analyze methylation data for IVF conceived infants compared to naturally conceived.

Data from:

1. Estill MS, Bolnick JM, Waterland RA, Bolnick AD, Diamond MP, Krawetz SA. Assisted reproductive technology alters deoxyribonucleic acid methylation profiles in bloodspots of newborn infants. Fertil Steril. 2016 Sep 1;106(3):629-639.e10. doi: 10.1016/j.fertnstert.2016.05.006. Epub 2016 Jun 8. PMID: 27288894.

2. Novakovic B, Lewis S, Halliday J, Kennedy J, Burgner DP, Czajko A, Kim B, Sexton-Oates A, Juonala M, Hammarberg K, Amor DJ, Doyle LW, Ranganathan S, Welsh L, Cheung M, McBain J, McLachlan R, Saffery R. Assisted reproductive technologies are associated with limited epigenetic variation at birth that largely resolves by adulthood. Nat Commun. 2019 Sep 2;10(1):3922. doi: 10.1038/s41467-019-11929-9. PMID: 31477727; PMCID: PMC6718382.

## Requirements

```{r,packages}

suppressPackageStartupMessages({
  library("limma")
  library("eulerr")
  library("IlluminaHumanMethylation450kmanifest")
  library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
  library("tictoc")
  library("mitch")
  library("gplots")
  library("kableExtra")
  library("beeswarm")
  library("missMethyl")
  library("gridExtra")
  library("png")
})

CORES=16

```

## Load data

Reactome pathways were downloaded on the 14th Sept 2023 from MsigDB.

We'll also score the probes by the signed -log pvalue

```{r,load1}

# estill 2016 450k
art1 <- readRDS("estill_nat_vs_fh.rds")
art1 <- art1[[1]]
rownames(art1) <- art1$Row.names
str(art1)

# novakovic 2019 EPIC
art2 <- readRDS("novakovic_nat_vs_fh.rds")
art2 <-art2[[1]]
rownames(art2) <- art2$Row.names
str(art2)

gs_symbols <- gmt_import("c2.cp.reactome.v2023.1.Hs.symbols.gmt")

```

## Curate the annotation

Use all probes on the chip. Start with 450k chip.

```{r,anno1}

tic()
anno <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
myann <- data.frame(anno[,c("UCSC_RefGene_Name","UCSC_RefGene_Group","Islands_Name","Relation_to_Island")])
gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
names(gp2) <- rownames(gp)
gp2 <- lapply(gp2,unique)
gt1 <- stack(gp2)
colnames(gt1) <- c("gene","probe")
gt1$probe <- as.character(gt1$probe)
dim(gt1)
str(gt1)
length(unique(gt1$gene))
toc() #6.3s

```

Now EPIC chip.

```{r,anno2}

tic()
anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
myann <- data.frame(anno[,c("UCSC_RefGene_Name","UCSC_RefGene_Group","Islands_Name","Relation_to_Island")])
gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
names(gp2) <- rownames(gp)
gp2 <- lapply(gp2,unique)
gt2 <- stack(gp2)
colnames(gt2) <- c("gene","probe")
gt2$probe <- as.character(gt2$probe)
dim(gt2)
str(gt2)
length(unique(gt2$gene))
toc() #10.6s

```

## Import for mitch

```{r,import1}

m1 <- mitch_import(x=art1,DEtype="limma",geneTable=gt1)
dim(m1)
head(m1)

m2 <- mitch_import(x=art2,DEtype="limma",geneTable=gt2)
dim(m2)
head(m2)

m <- merge(m1,m2,by=0)
rownames(m) <- m[,1]
m[,1]=NULL
colnames(m) <- c("Estill","Novakovic")
dim(m)
head(m)

```

## Run mitch

```{r,mitch1}

tic()
mres <- mitch_calc(x=m,genesets=gs_symbols,minsetsize=5,cores=16, priority="effect")
toc() #11.45s

mtable <- mres$enrichment_result
sig <- subset(mtable,p.adjustMANOVA<0.05)
head(sig,20)

```

Rank-rank

```{r,rankrank}

mrank <- mres$ranked_profile

palette <- colorRampPalette(c("white", "yellow", "orange", "red", "darkred", "black"))
k <- MASS::kde2d(mrank[,1], mrank[,2])
X_AXIS = "Estill 450k"
Y_AXIS = "Novakovic EPIC"

filled.contour(k, color.palette = palette, plot.title = {
        abline(v = 0, h = 0, lty = 2, lwd = 2, col = "blue")
        title(main = "rank-rank plot", xlab = X_AXIS,
        ylab = Y_AXIS)
    })

pdf("fig7a.pdf")
palette <- colorRampPalette(c("white", "yellow", "orange", "red", "darkred", "black"))
k <- MASS::kde2d(mrank[,1], mrank[,2])
X_AXIS = "Estill 450k"
Y_AXIS = "Novakovic EPIC"

filled.contour(k, color.palette = palette, plot.title = {
        abline(v = 0, h = 0, lty = 2, lwd = 2, col = "blue")
        title(main = "rank-rank plot", xlab = X_AXIS,
        ylab = Y_AXIS)
    })
dev.off()

```

PW Scatter

```{r,pwscatter}

head(mtable)

sig <- subset(mtable,p.adjustMANOVA<0.05)
plot(mtable$s.Discovery,mtable$s.Replication,pch=19,col="gray",
  xlab="Estill 450k",ylab="Novakovic EPIC",main="pathway enrichment score")
points(sig$s.Discovery,sig$s.Replication,pch=19,col="red")
abline(v = 0, h = 0, lty = 2, lwd = 2, col = "blue")

pdf("fig7b.pdf")
plot(mtable$s.Discovery,mtable$s.Replication,pch=19,col="gray",
  xlab="Estill 450k",ylab="Novakovic EPIC",main="pathway enrichment score")
points(sig$s.Discovery,sig$s.Replication,pch=19,col="red")
abline(v = 0, h = 0, lty = 2, lwd = 2, col = "blue")
dev.off()

```

Heatmap

```{r,mheat}

mt <- head(mtable,30)
rownames(mt) <- mt$set
mt <- mt[,c("Estill 450k","Novakovic EPIC")]
rownames(mt) <- gsub("REACTOME_","",rownames(mt))
rownames(mt) <- gsub("_"," ",rownames(mt))
colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2(as.matrix(mt),scale="none",trace="none",margins=c(6,25),col=colfunc(25),cexRow=0.6,cexCol=1)

pdf("fig7c.pdf")
heatmap.2(as.matrix(mt),scale="none",trace="none",margins=c(6,25),col=colfunc(25),cexRow=0.6,cexCol=1)
dev.off()

```

Example 2d

```{r,example}

det <- mres$detailed_sets$REACTOME_SENSORY_PERCEPTION

k <- MASS::kde2d(det[,1], det[,2])
X_AXIS = "Discovery"
Y_AXIS = "Replication"

XMAX=max(mrank[,1])
XMIN=min(mrank[,1])
YMAX=max(mrank[,2])
YMIN=min(mrank[,2])

filled.contour(k, color.palette = palette, xlim=c(XMIN,XMAX), ylim=c(YMIN,YMAX),
      plot.title = {
        abline(v = 0, h = 0, lty = 2, lwd = 2, col = "blue")
        title(main = "Creation of C4 and C2 Activators", xlab = X_AXIS,
        ylab = Y_AXIS)
    })

pdf("fig7d.pdf")
filled.contour(k, color.palette = palette, xlim=c(XMIN,XMAX), ylim=c(YMIN,YMAX),
      plot.title = {
        abline(v = 0, h = 0, lty = 2, lwd = 2, col = "blue")
        title(main = "Creation of C4 and C2 Activators", xlab = X_AXIS,
        ylab = Y_AXIS)
    })
dev.off()

```

Make a html report and some charts.

```{r,report1}

mitch_report(res=mres,outfile="aging_mitchreport.html",overwrite=TRUE)
mitch_plots(res=mres,outfile="aging_mitchcharts.pdf")

```

## Session information

```{r,save}

save.image("aging_example.Rdata")

sessionInfo()

```

### 2023-12-19 figs/art_example.Rmd
art example

@@
 sig <- subset(mtable,p.adjustMANOVA<0.05)
-plot(mtable$s.Discovery,mtable$s.Replication,pch=19,col="gray",
+plot(mtable$s.Estill,mtable$s.Novakovic,pch=19,col="gray",
   xlab="Estill 450k",ylab="Novakovic EPIC",main="pathway enrichment score")
-points(sig$s.Discovery,sig$s.Replication,pch=19,col="red")
+points(sig$s.Estill,sig$s.Novakovic,pch=19,col="red")
 abline(v = 0, h = 0, lty = 2, lwd = 2, col = "blue")
@@
 pdf("fig7b.pdf")
-plot(mtable$s.Discovery,mtable$s.Replication,pch=19,col="gray",
+plot(mtable$s.Estill,mtable$s.Novakovic,pch=19,col="gray",
   xlab="Estill 450k",ylab="Novakovic EPIC",main="pathway enrichment score")
-points(sig$s.Discovery,sig$s.Replication,pch=19,col="red")
+points(sig$s.Estill,sig$s.Novakovic,pch=19,col="red")
 abline(v = 0, h = 0, lty = 2, lwd = 2, col = "blue")
@@
 k <- MASS::kde2d(det[,1], det[,2])
-X_AXIS = "Discovery"
-Y_AXIS = "Replication"
+X_AXIS = "Estill 450k"
+Y_AXIS = "Novakovic EPIC"
 
@@
         abline(v = 0, h = 0, lty = 2, lwd = 2, col = "blue")
-        title(main = "Creation of C4 and C2 Activators", xlab = X_AXIS,
+        title(main = "Reactome sensory perception", xlab = X_AXIS,
         ylab = Y_AXIS)
@@
 
-mitch_report(res=mres,outfile="aging_mitchreport.html",overwrite=TRUE)
-mitch_plots(res=mres,outfile="aging_mitchcharts.pdf")
+mitch_report(res=mres,outfile="art_mitchreport.html",overwrite=TRUE)
+mitch_plots(res=mres,outfile="art_mitchcharts.pdf")
 
@@
 
-save.image("aging_example.Rdata")
+save.image("art_example.Rdata")
 

### 2023-12-19 figs/art_example.Rmd
art example

@@
 rownames(mt) <- mt$set
-mt <- mt[,c("Estill 450k","Novakovic EPIC")]
+mt <- mt[,c("s.Estill","s.Novakovic")]
 rownames(mt) <- gsub("REACTOME_","",rownames(mt))

### 2023-12-19 figs/art_example.Rmd
better example

@@
 
-det <- mres$detailed_sets$REACTOME_SENSORY_PERCEPTION
+tic()
+mres <- mitch_calc(x=m,genesets=gs_symbols,minsetsize=5,cores=16, priority="effect")
+toc() #11.45s
+
+mtable <- mres$enrichment_result
+sig <- subset(mtable,p.adjustMANOVA<0.05)
+head(sig,20)
+
+det <- mres$detailed_sets$REACTOME_NGF_STIMULATED_TRANSCRIPTION
 
@@
         abline(v = 0, h = 0, lty = 2, lwd = 2, col = "blue")
-        title(main = "Reactome sensory perception", xlab = X_AXIS,
+        title(main = "Reactome NGF-stimulated transcription", xlab = X_AXIS,
         ylab = Y_AXIS)

### 2023-12-19 figs/art_example.Rmd
fix art example

--- 2023-12-19 figs/art_example.Rmd
@@
 
-det <- mres$detailed_sets$REACTOME_SENSORY_PERCEPTION
+tic()
+mres <- mitch_calc(x=m,genesets=gs_symbols,minsetsize=5,cores=16, priority="significance")
+toc() #11.45s
+
+mtable <- mres$enrichment_result
+sig <- subset(mtable,p.adjustMANOVA<0.05)
+head(sig,20)
+
+det <- mres$detailed_sets$REACTOME_NUCLEAR_EVENTS_KINASE_AND_TRANSCRIPTION_FACTOR_ACTIVATION
 
@@
         abline(v = 0, h = 0, lty = 2, lwd = 2, col = "blue")
-        title(main = "Reactome sensory perception", xlab = X_AXIS,
+        title(main = "Reactome nuclear kinase and transcription factor activation", xlab = X_AXIS,
         ylab = Y_AXIS)
@@
 dev.off()
+
+det <- cbind(det, apply(det,1,median))
+
+det <- det[order(det[,3]),]
+
+head(det,20) %>% kbl(caption="Top hypomethylated genes of the selectedpathway") %>% kable_paper("hover", full_width = F)
 

### 2023-12-19 manuscript/manuscript.Rmd
ms work
fig7

@@
 (D) An example of a pathway identified with this method, "creation of C4 and C2 activators" shows lower methylation of member genes in both discovery and replication studies.
-](../figs/fig5.png)
+](../figs/fig6.png)
 
-## Exploration of methylation pathways altered by ART
+## Exploration of methylation pathways altered by IVF
 
-Two way analysis.
+Another application for this method is the joint pathway enrichment analysis of studies conducted with different
+array systems.
+Epigenetic differences between infants conceived naturally and with assisted reproductive technologies have been the subject of studies since 20XX (eg: REFS).
+Two of the highest powered studies are Estill 2016 which used the 450K array and Novakovic 2019 study which used
+the EPIC array.
+Joint enrichment analysis can uncover pathways that are common or different between these studies and illuminate
+the biological differences between conception groups.
+For this example we focused on comparing naturally conceived infants to those conceived with in vitro fertilisation with fresh embryo transfer.
+The Estill dataset on 450k had 418,833 probes that met the quality criteria (Natural n=43; IVF n=38).
+Of these, 149,552 probes showed statistical significant differential methylation (FDR<0.05), of which 28,637 and 120,885 exhibited higher or lower methylation in the IVF group respectively.
+The Novakovic dataset on EPIC array had 810,518 probes that met the filtering criteria (n=133).
+Of these, 5,491 were statistically significant (FDR<0.05) with 1,326 and 4,165 exhibited higher or lower methylation in the IVF group respectively.
+After summarisation to the gene level, the Estill dataset on 450k array describes 19,951 genes, while the
+Novakovic dataset on EPIC array describes 26,150 genes.
+There were 19,942 genes common to both platforms.
+The rank-rank plot indicates a high degree of overlap overall, and interestingly shows a trend of more genes having lower methylation level in the IVF group (**Figure 7A**).
+A scatterplot of pathway enrichment scores indicates a high degree of agreement between these studies at the level of pathways despite they being conducted years apart and analysed with different array systems (**Figure 7B**).
+The heatmap shows a high degree of agreement with pathways with a higher level of methylation in the IVF groups,
+while pathways with overall lower methylation in the IVF group showed some variability in enrichment scores
+between studies (**Figure 7C**).
+We were interested in which pathways had consistently lower methylation in the IVF group and identified the Reactome pathway "Nuclear kinase and transcription factor activation events" which had enrichment scores of -0.26 in both contrasts and a MANOVA FDR value of 0.002 (**Figure 7D**).
+Top ranked genes from this pathway include *FOSL1*, *JUNB*, *ASCL1* and *ID1*.
+A loss of methylation of these genes could contribute to disorders of development.
+Taken together, this analysis shows that LAM method can be used to compare enrichment of independent studies
+conducted with 450k and EPIC arrays.
 
-## Exploration of methylation pathways altered by elevated homocysteine or smoking
-
-Two way analysis.
+![Figure 7.
+Pathway level DNA methylation differences in natural and IVF conceived infants.
+(A) Contour heatmap showing the similarity in gene methylation score ranks in the Estill (450k) and Novakovic (EPIC) studies.
+(B) Mitch pathway enrichment scores (S distances) in Estil and Novakovic studies. Pathways with MANOVA FDR<0.05 are shown in red while others are shaded gray.
+(C) Heatmap of 30 pathways with most extreme S distances (a surrogate of effect size). Red indicates higher methylation and blue shows lower methylation.
+(D) An example of a pathway identified with this joint enrichment analysis method, "Nuclear kinase and transcription factor activation" shows lower methylation of member genes in both Estill and Novakovic studies.
+](../figs/fig7.png)
 

### 2023-12-19 app/trial/app.R
start shiny

@@ -1,0 +1,94 @@
+library("shiny")
+library("vroom")
+library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
+library("mitch")
+
+options(shiny.maxRequestSize = 100 * 1024^2) #100MB
+
+arraytype <- c("450k", "EPIC")
+genesettype <- c("Reactome","KEGG","GO")
+
+ui <- fluidPage(
+  titlePanel(
+    "GMEA"
+  ),
+  sidebarPanel(
+    textInput("dataname", "Dataset name:"),
+    radioButtons("arraytype", "Array platform:", arraytype),
+    radioButtons("genesettype", "Gene set database:", genesettype),
+    fileInput("upload", multiple = FALSE,label="Upload limma data:"),
+    actionButton("analyse", "Analyse!",icon = icon("gears"))
+  ),
+  mainPanel(
+    textOutput("jobdata"),
+    tableOutput("fileinfo"),
+    tableOutput("head"),
+    dataTableOutput("mtable")
+  )
+)
+
+server <- function(input, output, session) {
+
+  mydataname <- eventReactive(input$analyse, {
+    as.character(input$dataname)
+  })
+    
+  myarraytype <- eventReactive(input$analyse, {
+    as.character(input$arraytype)
+  })
+
+  mygenesettype <- eventReactive(input$analyse, {
+    as.character(input$genesettype)
+  })
+
+  fileinfo <- eventReactive(input$analyse, {
+    input$upload
+  })
+ 
+  output$jobdata <- renderText(paste(mydataname(),myarraytype(),mygenesettype()))
+  output$fileinfo <- renderTable(fileinfo())
+
+  data <- reactive({
+    req(input$upload)
+    
+    ext <- tools::file_ext(input$upload$name)
+    switch(ext,
+           csv = vroom::vroom(input$upload$datapath, delim = ","),
+           tsv = vroom::vroom(input$upload$datapath, delim = "\t"),
+           validate("Invalid file; Please upload a .csv or .tsv file")
+    )
+  })
+  
+  output$head <- renderTable({
+    head(data(), 10)
+  })
+  
+  gt2 <- reactive({
+    req(input$upload)
+    anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
[…]

### 2023-12-20 manuscript/manuscript.Rmd
writing

@@
 * DNA methylation is the most widely studied epigenetic mark, for its role in development and
-disease.
+disease [@Robertson2005-tu].
 
 * Many hundreds of epigenome-wide association studies (EWASs) are conducted each year to understand DNA
-methylation patterns in disease and other contexts.
+methylation patterns in disease and other contexts [@Li2019-og].
 
 * Infinium arrays remain the preferred platform for EWASs due to low cost and analytical simplicity
-as compared to high throughput methylation sequencing.
+as compared to high throughput methylation sequencing [@Wei2021-qj].
 
 * Infinium arrays typically have multiple probes per gene in different locations including promoters,
-CpG islands, gene bodies and enhancers, which complicates downstream functional enrichment analysis,
+CpG islands, gene bodies and enhancers [@Pidsley2016-ix], which complicates downstream functional enrichment analysis,
 otherwise known as pathway analysis or gene set enrichment analysis.
@@
 * Enrichment analysis comes in two popular types; over-representation analysis (ORA) and functional
-class scoring (FCS).
+class scoring (FCS) [@Khatri2012-qy].
+A third type of enrichment analysis called pathway topology improves upon other methods with more sophisticated
+modelling of gene network activations, but these are yet to be adopted widely [@Xie2021-py].
 
@@
 The test seeks to identify sets of genes (eg: ontologies, pathways) that are over-represented in the
-gene list of interest relative to the background.
+gene list of interest relative to the background [@Tavazoie1999-ng].
 
-Functional class scoring takes a different approach by ranking all detected genes by a differential
+FCS takes a different approach by ranking all detected genes by a differential
 regulation score (eg: fold change, confident effect size) followed by a test to assess whether each
-set of genes has distribution of scores that is different to the null.
-Currently there are no specialised approaches for the application of FCS to Infinium array data.
+set of genes has distribution of scores that is different to the null [@Subramanian2005-ar].
+
+The `gsameth()` function of the `missMethyl` package is the state-of-the-art methods for ORA of infinium methylation array
+data as it addresses issues related to probes belonging to more than one gene and the fact that genes can have multiple
+probes [@Maksimovic2021-ur].
+As gsameth is an ORA method, results dependent strongly on the significance threshold used [@Ren2019-mr], and also
+on the proportion of probes that meet this threshold.
+
+FCS methods `ebGSEA` and `methylGSA` have been applied to EWAS data and are suggested to have better sensitivity to
+determine subtle associations between pathways and differential methylation [@Dong2019-np;@Ren2019-mr].
+The ebGSEA tool uses an empirical Bayes modeling approach followed by a one-tailed nonparametric test for gene set
+enrichment [@Dong2019-np].
+The `methylGSA` package implements two different approaches.
+Firstly, the `methylRRA` function which uses robust rank aggregation for detecting differentially methylated genes taking
+into consideration the variable number of probes followed by an FCS test based on z-scores [@Ren2019-mr].
+Secondly, `methylglm` uses a logistic regression approach to model differential methylation for genes inside and outside eachgene set [@Ren2019-mr].
+These methods are indeed more sensitive than ORA approaches, but as they are based on one-tailed tests irrespective of
+direction of methylation change, it limits the utility for downstream interpretation of the importance of methylation
+changes on genomic regulation.
+Neither gsameth, ebGSEA nor methylGSA report computed enrichment scores, which hinders downstream interpretation, as the
+enrichment score is a useful surrogate for effect size.
+Without an enrichment score, users are limited to using statistical significance values to prioritise results, which
+biases towards large gene sets that have non-specific biological functions.
+This can lead to end users over-looking smaller pathways with more specific biological functions that have larger effect
+sizes, as those p-values are relatively larger.
+
+Here we aim to develop and evaluate methods for two-tailed FCS of Infinium methylation array data that reports enrichment
+scores.
 
@@
 
+### Pathway-level differential methylation in aging
+In order to deminstrate the utility of the LAM method for 
+(2) the association of elevated plasma homocysteine with methylation patterns.
+
 ### Pathway-level differential methylation in assisted reproductive technology
 (i) the effect of assisted reproductive technology DNA methylation
-
-### Pathway-level differential methylation in hyperhomocysteinemia
-(2) the association of elevated plasma homocysteine with methylation patterns.
 
@@
 This process was repeated 50 times and the results shown in (**Figure 4**).
-Pathways that were identified in the smaller and full group were termed "consistent",
-while pathays that were identified in the smaller but not in the full group were
+Pathways that were identified as statistically significant (FDR<0.05) in the smaller and full group were
+termed "consistent", while pathways that were identified in the smaller but not in the full group were
 termed "inconsistent".
@@
 
-![Figure 4. Sensitivity of LAM and GSA methods with real cancer array data.](../figs/fig4.png)
+![Figure 4. Sensitivity of LAM and GSA methods with real cancer array data.
+Upper panels show the number of significant pathways identified in the smaller sample that were consistent
+with the full group for LAM (left) and GSA (right).
+Middle panels show the number of significant pathways identified in the smaller sample that were inconsistent
+with the full group.
+Lower panels show the proportion of significant pathways identified that were inconsistent with the full
+group.
+Left panels correspond to LAM, and right panels to GSA methods.
+](../figs/fig4.png)
 

### 2023-12-20 figs/GSE158422_sensitivity.Rmd
use half of genes for sets
what about probe mean rank - do it all again in docker

@@
 
-CORES=24
+CORES=16
 
@@
 summary(unlist(lapply(sets,length)))
-
-genesets <- gmt_import("https://ziemann-lab.net/public/gmea_prototype/ReactomePathways.gmt")
 
@@
 ```
+
 ## Run limma
@@
 
+## LRM - limma rank mean mitch
+
+```{r,lmrm1}
+
+tic()
+anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
+myann <- data.frame(anno[,c("UCSC_RefGene_Name","UCSC_RefGene_Group","Islands_Name","Relation_to_Island")])
+gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
+gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
+names(gp2) <- rownames(gp)
+gp2 <- lapply(gp2,unique)
+gt <- stack(gp2)
+colnames(gt) <- c("gene","probe")
+dim(gt)
+str(gt)
+toc() #8.735s
+
+dm$rank <-  rank(dm$t) - nrow(subset(dm,t<0))
+mm <- merge(dm,gt,by.x=0,by.y="probe")
+head(mm)
+mma <-aggregate(mm$rank ~ gene,mm,mean)
+rownames(mma) <- mma$gene
+mma$gene = NULL
+colnames(mma) <- "meanrank"
+lrmres <- runmitch(m=mma,genesets=gs_symbols,cores=16)
+
+lrmsig <- subset(lrmres,p.adjustANOVA<0.05)
+
+top <- head(lrmsig,20)
+top <- top[order(top$s.dist),]
+
+top %>%
+  kbl(caption="Top gene sets (biggest effect size with FDR<0.05)") %>%
+  kable_paper("hover", full_width = F)
+
+barnames <- gsub("_"," ",gsub("REACTOME_","",rownames(top)))
+cols <- gsub("TRUE","red",gsub("FALSE","blue",as.character(top$s.dist>0)))
+
+par(mar = c(5.1, 29.1, 4.1, 2.1))
+barplot(abs(top$s.dist),horiz=TRUE,las=1,names.arg=barnames, cex.names=0.65,
+  cex.axis=0.8,col=cols, xlab="Enrichment score",main="Reactomes")
+grid()
+
+par( mar = c(5.1, 4.1, 4.1, 2.1) )
+
+hist(lrmsig$s.dist)
+
+# distinguish up (TRUE) and down (FALSE)
+lrmsigsets <- paste(rownames(lrmsig),as.character(lrmsig$s.dist > 0))
+lrmsigsets2 <- gsub(" TRUE"," UP",lrmsigsets)
+lrmsigsets2 <- gsub(" FALSE"," DN",lrmsigsets2)
+
+```
+
 ## Run LAM
@@
 tic()
-dmagg <- agg(dm,cores=32)
+anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
+myann <- data.frame(anno[,c("UCSC_RefGene_Name","UCSC_RefGene_Group","Islands_Name","Relation_to_Island")])
+gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
+gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
+names(gp2) <- rownames(gp)
+gp2 <- lapply(gp2,unique)
+gt <- stack(gp2)
+colnames(gt) <- c("gene","probe")
+dim(gt)
+str(gt)
+toc() #8.735s
+
+tic()
+dmagg <- agg(dm,cores=16)
 m <- dmagg$gme_res_df[,"mean",drop=FALSE]
-lamres <- runmitch(m=m,genesets=gs_symbols,cores=32)
+lamres <- runmitch(m=m,genesets=gs_symbols,cores=16)
 toc()
@@
 
+## LMRM - limma mean rank mitch
+
+```{r,lmrm1}
+
+tic()
+anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
+myann <- data.frame(anno[,c("UCSC_RefGene_Name","UCSC_RefGene_Group","Islands_Name","Relation_to_Island")])
+gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
+gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
+names(gp2) <- rownames(gp)
+gp2 <- lapply(gp2,unique)
+gt <- stack(gp2)
+colnames(gt) <- c("gene","probe")
+dim(gt)
+str(gt)
+toc() #8.735s
+
+dm$rank <-  rank(dm$t) - nrow(subset(dm,t<0))
+mm <- merge(dm,gt,by.x=0,by.y="probe")
+head(mm)
+mma <-aggregate(mm$rank ~ gene,mm,mean)
+rownames(mma) <- mma$gene
+mma$gene = NULL
+colnames(mma) <- "meanrank"
+lmrmres <- runmitch(m=mma,genesets=gs_symbols,cores=16)
+
+lmrmsig <- subset(lmrmres,p.adjustANOVA<0.05)
+
+top <- head(lmrmsig,20)
+top <- top[order(top$s.dist),]
+
+top %>%
+  kbl(caption="Top gene sets (biggest effect size with FDR<0.05)") %>%
+  kable_paper("hover", full_width = F)
+
+barnames <- gsub("_"," ",gsub("REACTOME_","",rownames(top)))
+cols <- gsub("TRUE","red",gsub("FALSE","blue",as.character(top$s.dist>0)))
+
+par(mar = c(5.1, 29.1, 4.1, 2.1))
+barplot(abs(top$s.dist),horiz=TRUE,las=1,names.arg=barnames, cex.names=0.65,
+  cex.axis=0.8,col=cols, xlab="Enrichment score",main="Reactomes")
+grid()
+
+par( mar = c(5.1, 4.1, 4.1, 2.1) )
+
+hist(lmrmsig$s.dist)
+
+# distinguish up (TRUE) and down (FALSE)
+lmrmsigsets <- paste(rownames(lmrmsig),as.character(lmrmsig$s.dist > 0))
+lmrmsigsets2 <- gsub(" TRUE"," UP",lmrmsigsets)
+lmrmsigsets2 <- gsub(" FALSE"," DN",lmrmsigsets2)
+
+```
+
 ## Downsample LAM test
@@
 
+blah()
 SAMPLESIZES=c(2,3,6,10,16,22,30)

### 2023-12-20 figs/GSE158422_simulate050.Rmd
yet more bugfixes
adding the probe rank approach

@@
 
+# get probe-gene mapping
 anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
 myann <- data.frame(anno[,c("UCSC_RefGene_Name","UCSC_RefGene_Group","Islands_Name","Relation_to_Island")])
-
 gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
@@
 names(gp2) <- rownames(gp)
-sets <- split(rep(names(gp2), lengths(gp2)), unlist(gp2))
-
-summary(unlist(lapply(sets,length)))
+gp2 <- lapply(gp2,unique)
+gt <- stack(gp2)
+colnames(gt) <- c("gene","probe")
+dim(gt)
+str(gt)
+head(gt)
 
@@
 
-# aggregate
-agg <- function(dm,cores=1) {
-  gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
-  gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
-  gnl <- mclapply(gnl,unique,mc.cores=cores)
-  dm$UCSC_RefGene_Name <- gnl
-  l <- mclapply(1:nrow(dm), function(i) {
-    a <- dm[i,]
-    len <- length(a[[1]][[1]])
-    tvals <- as.numeric(rep(a["t"],len))
-    genes <- a[[1]][[1]]
-    data.frame(genes,tvals)
-  },mc.cores=cores)
-  df <- do.call(rbind,l)
-  keep <- names(which(table(df$genes)>1))
-  df <- df[df$genes %in% keep,]
-  gn <- unique(df$genes)
-  gme_res <- lapply( 1:length(gn), function(i) {
-    g <- gn[i]
-    tstats <- df[which(df$genes==g),"tvals"]
-    myn <- length(tstats)
-    mymean <- mean(tstats)
-    mymedian <- median(tstats)
-    top <- tstats[order(-abs(tstats))][1]
-    if ( length(tstats) > 2 ) {
-      ttest <- t.test(tstats)
-      pval <- ttest$p.value
-    } else {
-      pval = 1
-    }
-    res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,
-      "median"=mymedian, "top"=top, "pval"=pval)
-  } )
-  gme_res_df <- do.call(rbind, gme_res)
-  rownames(gme_res_df) <- gme_res_df[,1]
-  gme_res_df <- gme_res_df[,-1]
-  tmp <- apply(gme_res_df,2,as.numeric)
-  rownames(tmp) <- rownames(gme_res_df)
-  gme_res_df <- as.data.frame(tmp)
-  gme_res_df$sig <- -log10(gme_res_df[,"pval"])
-  gme_res_df <- gme_res_df[order(-gme_res_df$sig),]
-  gme_res_df$fdr <- p.adjust(gme_res_df$pval)
-  out <- list("df"=df,"gme_res_df"=gme_res_df)
-  return(out)
-}
-
-# geometric mean aggregate
-geoagg <- function(dm,cores=1) {
-  gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
-  gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
-  gnl <- mclapply(gnl,unique,mc.cores=cores)
-  dm$UCSC_RefGene_Name <- gnl
-  l <- mclapply(1:nrow(dm), function(i) {
-    a <- dm[i,]
-    len <- length(a[[1]][[1]])
-    tvals <- as.numeric(rep(a["t"],len))
-    genes <- a[[1]][[1]]
-    data.frame(genes,tvals)
-  },mc.cores=cores)
-  df <- do.call(rbind,l)
-  keep <- names(which(table(df$genes)>1))
-  df <- df[df$genes %in% keep,]
-  gn <- unique(df$genes)
-  gme_res <- lapply( 1:length(gn), function(i) {
-    g <- gn[i]
-    tstats <- df[which(df$genes==g),"tvals"]
-    myn <- length(tstats)
-    mymean <- jamGeomean(tstats)
-    mymedian <- median(tstats)
-    top <- tstats[order(-abs(tstats))][1]
-    if ( length(tstats) > 2 ) {
-      ttest <- t.test(tstats)
-      pval <- ttest$p.value
-    } else {
-      pval = 1
-    }
-    res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,
-      "median"=mymedian, "top"=top, "pval"=pval)
-  } )
-  gme_res_df <- do.call(rbind, gme_res)
-  rownames(gme_res_df) <- gme_res_df[,1]
-  gme_res_df <- gme_res_df[,-1]
-  tmp <- apply(gme_res_df,2,as.numeric)
-  rownames(tmp) <- rownames(gme_res_df)
-  gme_res_df <- as.data.frame(tmp)
-  gme_res_df$sig <- -log10(gme_res_df[,"pval"])
-  gme_res_df <- gme_res_df[order(-gme_res_df$sig),]
-  gme_res_df$fdr <- p.adjust(gme_res_df$pval)
-  out <- list("df"=df,"gme_res_df"=gme_res_df)
-  return(out)
-}
-
 # enrich parametric
@@
   dm3 <- runlimma(mval=mval2,design=d,myann=myann)
-  dmagg1 <- agg(dm3,cores=4)
-  m1 <- dmagg1$gme_res_df[,"mean",drop=FALSE]
+  dd <- merge(dm3,gt,by.x=0,by.y="probe")
+  m1 <- aggregate(t ~ gene,dd,mean)
+  rownames(m1) <- m1$gene
+  m1$gene=NULL
   lares1 <- ttenrich(m=m1,genesets=gsets,cores=4,testtype="competitive")
@@
   dm3 <- runlimma(mval=mval2,design=d,myann=myann)
-  dmagg1 <- agg(dm3,cores=4)
-  m1 <- dmagg1$gme_res_df[,"top",drop=FALSE]
-  lares1 <- ttenrich(m=m1,genesets=gsets,cores=4,testtype="competitive")
-  gsig_up3 <- rownames(subset(lares1, fdr < 0.05 & t_mean > 0))
-  gsig_dn3 <- rownames(subset(lares1, fdr < 0.05 & t_mean < 0))
-  gtup <- names(sim[[6]])
-  gtdn <- names(sim[[7]])
-  UPTP=length(intersect(gsig_up3 ,gtup))
-  UPFP=length(setdiff(gsig_up3 ,gtup))
-  UPFN=length(setdiff(gtup,gsig_up3))
-  DNTP=length(intersect(gsig_dn3 ,gtdn))
-  DNFP=length(setdiff(gsig_dn3 ,gtdn))
-  DNFN=length(setdiff(gtdn,gsig_dn3))
-  TP=UPTP+DNTP
-  FP=UPFP+DNFP
-  FN=UPFN+DNFN
-  TN=nrow(lares1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
-  PREC=TP/(TP+FP)
-  REC=TP/(TP+FN)
-  F1=TP/(TP+(0.5*(FP+FN)))
-  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
-  return(result)
-}
-
-# LA parametric competitive geometric
-simlacgeo <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
-  # generate gene sets
-  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
-  lengths <- unname(unlist(lapply(genesetdatabase,length)))
-  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
-  # select gene sets to alter
-  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
-  # incorporate select changes
-  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
-    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta,
-    gene_catalog=gene_catalog)
-  # set up limma
-  mval2 <- sim$mval
-  ncols <- ncol(mval2)
-  groupsize <- ncols/2
-  ss <- data.frame(colnames(mval2))
-  colnames(ss) <- "sample"
-  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
-  d <- model.matrix(~ ss$case )
-  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
-  dmagg1 <- geoagg(dm3,cores=4)
-  m1 <- dmagg1$gme_res_df[,"mean",drop=FALSE]
+  dd <- merge(dm3,gt,by.x=0,by.y="probe")
+  m1 <- aggregate(t ~ gene,dd, function(x) {
+    if (abs(max(x)) > abs(min(x))) { max(x) } else { min(x) }
+  })
+  rownames(m1) <- m1$gene
+  m1$gene=NULL
   lares1 <- ttenrich(m=m1,genesets=gsets,cores=4,testtype="competitive")
@@
   dm3 <- runlimma(mval=mval2,design=d,myann=myann)
-  dmagg1 <- agg(dm3,cores=4)
-  m1 <- dmagg1$gme_res_df[,"mean",drop=FALSE]
+  dd <- merge(dm3,gt,by.x=0,by.y="probe")
+  m1 <- aggregate(t ~ gene,dd,mean)
+  rownames(m1) <- m1$gene
+  m1$gene=NULL
   lares1 <- wtenrich(m=m1,genesets=gsets,cores=4,testtype="competitive")
@@
   dm3 <- runlimma(mval=mval2,design=d,myann=myann)
-  dmagg1 <- agg(dm3,cores=4)
-  m1 <- dmagg1$gme_res_df[,"mean",drop=FALSE]
+  dd <- merge(dm3,gt,by.x=0,by.y="probe")
+  m1 <- aggregate(t ~ gene,dd,mean)
+  rownames(m1) <- m1$gene
+  m1$gene=NULL
   lamres1 <- runmitch(m=m1,genesets=gsets,cores=4)
@@
 
-```
-
-## AL: Aggregate limma functions
+# LA rank competitive mitch
+simlrm <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
+  # generate gene sets
+  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
+  lengths <- unname(unlist(lapply(genesetdatabase,length)))
+  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
+  # select gene sets to alter
+  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
+  # incorporate select changes
+  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
+    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta,
+    gene_catalog=gene_catalog)
+  # set up limma
+  mval2 <- sim$mval
+  ncols <- ncol(mval2)
+  groupsize <- ncols/2
+  ss <- data.frame(colnames(mval2))
+  colnames(ss) <- "sample"
+  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
+  d <- model.matrix(~ ss$case )
+  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
+  # rank probes first
+  dm3$rank <-  rank(dm3$t) - nrow(subset(dm3,t<0))
+  mm <- merge(dm3,gt,by.x=0,by.y="probe")
+  head(mm)
+  mma <-aggregate(mm$rank ~ gene,mm,mean)
+  rownames(mma) <- mma$gene
+  mma$gene = NULL
+  colnames(mma) <- "meanrank"
+  lrmres <- runmitch(m=mma,genesets=gs_symbols,cores=16)
+  gsig_up3 <- rownames( subset( lrmres1, p.adjustANOVA < 0.05 & s.dist > 0 ) )
+  gsig_dn3 <- rownames( subset( lrmres1, p.adjustANOVA < 0.05 & s.dist < 0 ) )
+  gtup <- names(sim[[6]])
+  gtdn <- names(sim[[7]])
+  UPTP=length(intersect(gsig_up3 ,gtup))
+  UPFP=length(setdiff(gsig_up3 ,gtup))
+  UPFN=length(setdiff(gtup,gsig_up3))
+  DNTP=length(intersect(gsig_dn3 ,gtdn))
+  DNFP=length(setdiff(gsig_dn3 ,gtdn))
+  DNFN=length(setdiff(gtdn,gsig_dn3))
+  TP=UPTP+DNTP
+  FP=UPFP+DNFP
+  FN=UPFN+DNFN
+  TN=nrow(lamres1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
+  PREC=TP/(TP+FP)
+  REC=TP/(TP+FN)
+  F1=TP/(TP+(0.5*(FP+FN)))
+  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
+  return(result)
+}
+
+```
+
+## AL: Aggregate limma
 
@@
 
-```{r,alfunc}
-
-# chromosome by chromosome will be much faster
-magg <- function(mval,myann,cores=1){
-  gn <- unique(unlist(strsplit( myann$UCSC_RefGene_Name ,";")))
-  gnl <- strsplit( myann$UCSC_RefGene_Name ,";")
-  gnl <- mclapply(gnl,unique,mc.cores=cores)
-  myann$gnl <- gnl
-  keep <- rownames(subset(myann,UCSC_RefGene_Name!=""))
-  mx <- mval[rownames(mval) %in% keep,]
-  mymed <- function(g) {
-    probes <- rownames(myann[grep(g,myann$gnl),])
-    rows <- which(rownames(mx) %in% probes)
-    if ( length(rows) > 1 ) {
-      b <- mx[rows,]
-      med <- apply(b,2,mean)
-      med <- matrix(med,nrow=1)
-      colnames(med) <- colnames(b)
-      rownames(med) <- g
-      return(med)
-    }
-  }
-  med <- mclapply(gn,mymed,mc.cores=cores)
-  med <- med[lapply(med,length)>0]
-  medf <- do.call(rbind,med)
-  return(medf)
-}
-
-chragg <- function(mval,myann,cores=1){
-  annodf <- as.data.frame(anno)
-  keep <- rownames(subset(myann,UCSC_RefGene_Name!=""))
-  mx <- mval[rownames(mval) %in% keep,]
-  chrs <- unique(anno$chr)
-  myorder <- unlist(lapply(chrs,function(mychr) { nrow( annodf[annodf$chr==mychr,] ) } ))
-  chrs <- chrs[order(-myorder)]
-  leadercores <- floor(sqrt(cores))
-  workercores <- ceiling(sqrt(cores))
-  chrmedf <- mclapply(chrs,function(chr) {
-    chrfrag <- annodf[annodf$chr==chr,]
-    chrprobes <-rownames(chrfrag)
-    chrmx <- mx[rownames(mx) %in% chrprobes,]
-    chranno <- myann[rownames(myann) %in% chrprobes,]
-    chrmedf <- magg(mval=chrmx,myann=chranno,cores=workercores)
-    return(chrmedf)
-  },mc.cores=leadercores)
-  medf <- do.call(rbind, chrmedf)
-  return(medf)
-}
-
-agglimma <- function(medf,design) {
-  fit.reduced <- lmFit(medf,design)
-  fit.reduced <- eBayes(fit.reduced)
-  dmagg <- topTable(fit.reduced,coef=ncol(design), number = Inf)
-  nondup <- !duplicated(dmagg$ID)
-  dmagg <- dmagg[nondup,]
-  rownames(dmagg) <- dmagg$ID
-  dmagg$ID = NULL
-  return(dmagg)
-}
+```{r,al}
 
@@
   # al pipeline
-  medf1 <- chragg(mval=mval2,myann=myann,cores=4)
-  magg1 <- agglimma(medf1,d)
-  m1 <- as.data.frame(magg1$t)
-  rownames(m1) <- rownames(magg1)
+  mm <- merge(mval2,gt,by.x=0,by.y="probe")
+  mm$Row.names = NULL
+  a <- aggregate(. ~ gene,mm,mean)
+  rownames(a) <- a$gene
+  a$gene=NULL
+  fit.reduced <- lmFit(a,d)
+  fit.reduced <- eBayes(fit.reduced)
+  al <- topTable(fit.reduced,coef=ncol(d), number = Inf)
+  m1 <- as.data.frame(al$t)
+  rownames(m1) <- rownames(al)
   colnames(m1) <- "t"
@@
   # alm pipeline
-  medf1 <- chragg(mval2,myann,cores=4)
-  magg1 <- agglimma(medf1,d)
-  m1 <- as.data.frame(magg1$t)
-  rownames(m1) <- rownames(magg1)
+  mm <- merge(mval2,gt,by.x=0,by.y="probe")
+  mm$Row.names = NULL
+  a <- aggregate(. ~ gene,mm,mean)
+  rownames(a) <- a$gene
+  a$gene=NULL
+  fit.reduced <- lmFit(a,d)
+  fit.reduced <- eBayes(fit.reduced)
+  al <- topTable(fit.reduced,coef=ncol(d), number = Inf)
+  m1 <- as.data.frame(al$t)
+  rownames(m1) <- rownames(al)
   colnames(m1) <- "t"
@@
   d <- model.matrix(~ ss$case )
-  # aa pipeline
-  aares1 <- aal(mval=mval2, myann=myann, genesets=gsets, design=d, cores=4)
+  mm <- merge(mval2,gt,by.x=0,by.y="probe")
+  mm$Row.names = NULL
+  a <- aggregate(. ~ gene,mm,mean)
+  rownames(a) <- a$gene
+  a$gene=NULL
+  mystack <- stack(gsets)
+  mmm <- merge(a,mystack,by.x=0,by.y="values")
+  mmm$Row.names=NULL
+  aa <- aggregate(. ~ ind,mmm,mean)
+  rownames(aa) <- aa$ind
+  aa$ind=NULL
+  fit.reduced <- lmFit(aa,d)
+  fit.reduced <- eBayes(fit.reduced)
+  aares1 <- topTable(fit.reduced,coef=ncol(d), number = Inf)
   # summarise results
@@
 num_dm_sets=50
-sims=3
+sims=5
 
@@
 
-### parametric competitive aggregation with geometric mean
-
-Precision is same but recall was very low.
-
-Not included.
-
 ### nonparametric competitive
@@
 
+### LA rank probes
+
+```{r,runlrm}
+
+lrmres <- lapply(1:nrow(params) , function(j) {
+  groupsize = params[j,1]
+  delta = params[j,2]
+  res <- mclapply(1:sims,function(i) {
+    simlrm(genesetdatabase=gsets, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
+      groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
+  },mc.cores=4)
+  res <- do.call(rbind,res)
+  return(res)
+})
+
+lrmres2 <- do.call(rbind,lapply(lrmres,colMeans))
+lrmres2[,"PREC"] <- lrmres2[,"TP"] / ( lrmres2[,"TP"] + lrmres2[,"FP"] )
+
+lrmres2 %>% kbl(caption="LA rank mitch results") %>% kable_paper("hover", full_width = F)
+
+lrmres3p <- do.call(rbind,lapply(groupsizes, function (g) { lrmres2[params$groupsizes==g,"PREC"] }))
+colnames(lrmres3p) <- deltas
+rownames(lrmres3p) <- groupsizes
+lrmres3p %>% kbl(caption="LA rank mitch precision") %>% kable_paper("hover", full_width = F)
+
+lrmres3r <- do.call(rbind,lapply(groupsizes, function (g) { lrmres2[params$groupsizes==g,"REC"] }))
+colnames(lrmres3r) <- deltas
+rownames(lrmres3r) <- groupsizes
+lrmres3r %>% kbl(caption="LA rank mitch recall") %>% kable_paper("hover", full_width = F)
+
+F1(lrmres3p,lacmres3r) %>% kbl(caption="LA rank mitch F1") %>% kable_paper("hover", full_width = F)
+
+```
+
 ## AL sim

### 2023-12-20 figs/GSE158422_simulate050.Rmd
added lrm

@@
   colnames(mma) <- "meanrank"
-  lrmres <- runmitch(m=mma,genesets=gs_symbols,cores=16)
-  gsig_up3 <- rownames( subset( lrmres1, p.adjustANOVA < 0.05 & s.dist > 0 ) )
-  gsig_dn3 <- rownames( subset( lrmres1, p.adjustANOVA < 0.05 & s.dist < 0 ) )
+  lrmres <- runmitch(m=mma,genesets=gsets,cores=4)
+  gsig_up3 <- rownames( subset( lrmres, p.adjustANOVA < 0.05 & s.dist > 0 ) )
+  gsig_dn3 <- rownames( subset( lrmres, p.adjustANOVA < 0.05 & s.dist < 0 ) )
   gtup <- names(sim[[6]])
@@
   FN=UPFN+DNFN
-  TN=nrow(lamres1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
+  TN=nrow(lrmres)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
   PREC=TP/(TP+FP)
@@
       groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
-  },mc.cores=4)
+  },mc.cores=6)
   res <- do.call(rbind,res)

### 2023-12-20 figs/GSE158422_simulate020.Rmd
fig dumb bug
yet more bugfixes
final?

@@
   dm3 <- runlimma(mval=mval2,design=d,myann=myann)
-  dmagg1 <- agg(dm3,cores=4)
-  m1 <- dmagg1$gme_res_df[,"top",drop=FALSE]
+  dd <- merge(dm3,gt,by.x=0,by.y="probe")
+  m1 <- aggregate(t ~ gene,dd, function(x) {
+    if (abs(max(x)) > abs(min(x))) { max(x) } else { min(x) }
+  })
+  rownames(m1) <- m1$gene
+  m1$gene=NULL
   lares1 <- ttenrich(m=m1,genesets=gsets,cores=4,testtype="competitive")
@@
 
-# LA parametric competitive geometric
-simlacgeo <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
-  # generate gene sets
-  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
-  lengths <- unname(unlist(lapply(genesetdatabase,length)))
-  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
-  # select gene sets to alter
-  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
-  # incorporate select changes
-  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
-    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
-  # set up limma
-  mval2 <- sim$mval
-  ncols <- ncol(mval2)
-  groupsize <- ncols/2
-  ss <- data.frame(colnames(mval2))
-  colnames(ss) <- "sample"
-  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
-  d <- model.matrix(~ ss$case )
-  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
-  dmagg1 <- geoagg(dm3,cores=4)
-  m1 <- dmagg1$gme_res_df[,"mean",drop=FALSE]
-  lares1 <- ttenrich(m=m1,genesets=gsets,cores=4,testtype="competitive")
-  gsig_up3 <- rownames(subset(lares1, fdr < 0.05 & t_mean > 0))
-  gsig_dn3 <- rownames(subset(lares1, fdr < 0.05 & t_mean < 0))
-  gtup <- names(sim[[6]])
-  gtdn <- names(sim[[7]])
-  UPTP=length(intersect(gsig_up3 ,gtup))
-  UPFP=length(setdiff(gsig_up3 ,gtup))
-  UPFN=length(setdiff(gtup,gsig_up3))
-  DNTP=length(intersect(gsig_dn3 ,gtdn))
-  DNFP=length(setdiff(gsig_dn3 ,gtdn))
-  DNFN=length(setdiff(gtdn,gsig_dn3))
-  TP=UPTP+DNTP
-  FP=UPFP+DNFP
-  FN=UPFN+DNFN
-  TN=nrow(lares1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
-  PREC=TP/(TP+FP)
-  REC=TP/(TP+FN)
-  F1=TP/(TP+(0.5*(FP+FN)))
-  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
-  return(result)
-}
-
 # LA nonparametric competitive
@@
 
-```
-
-## AL: Aggregate limma functions
+# LA rank competitive mitch
+simlrm <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
+  # generate gene sets
+  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
+  lengths <- unname(unlist(lapply(genesetdatabase,length)))
+  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
+  # select gene sets to alter
+  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
+  # incorporate select changes
+  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
+    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta,
+    gene_catalog=gene_catalog)
+  # set up limma
+  mval2 <- sim$mval
+  ncols <- ncol(mval2)
+  groupsize <- ncols/2
+  ss <- data.frame(colnames(mval2))
+  colnames(ss) <- "sample"
+  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
+  d <- model.matrix(~ ss$case )
+  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
+  # rank probes first
+  dm3$rank <-  rank(dm3$t) - nrow(subset(dm3,t<0))
+  mm <- merge(dm3,gt,by.x=0,by.y="probe")
+  head(mm)
+  mma <-aggregate(mm$rank ~ gene,mm,mean)
+  rownames(mma) <- mma$gene
+  mma$gene = NULL
+  colnames(mma) <- "meanrank"
+  lrmres <- runmitch(m=mma,genesets=gsets,cores=4)
+  gsig_up3 <- rownames( subset( lrmres, p.adjustANOVA < 0.05 & s.dist > 0 ) )
+  gsig_dn3 <- rownames( subset( lrmres, p.adjustANOVA < 0.05 & s.dist < 0 ) )
+  gtup <- names(sim[[6]])
+  gtdn <- names(sim[[7]])
+  UPTP=length(intersect(gsig_up3 ,gtup))
+  UPFP=length(setdiff(gsig_up3 ,gtup))
+  UPFN=length(setdiff(gtup,gsig_up3))
+  DNTP=length(intersect(gsig_dn3 ,gtdn))
+  DNFP=length(setdiff(gsig_dn3 ,gtdn))
+  DNFN=length(setdiff(gtdn,gsig_dn3))
+  TP=UPTP+DNTP
+  FP=UPFP+DNFP
+  FN=UPFN+DNFN
+  TN=nrow(lrmres)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
+  PREC=TP/(TP+FP)
+  REC=TP/(TP+FN)
+  F1=TP/(TP+(0.5*(FP+FN)))
+  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
+  return(result)
+}
+
+```
+
+## AL: Aggregate limma
 
@@
 num_dm_sets=50
-sims=3
+sims=50
 
@@
 ```
-
-### parametric competitive aggregation with geometric mean
-
-Precision is same but recall was very low.
-
-Not included.
 
@@
 F1(lacmres3p,lacmres3r) %>% kbl(caption="LA mitch F1") %>% kable_paper("hover", full_width = F)
+
+```
+
+### LA rank probes
+
+```{r,runlrm}
+
+lrmres <- lapply(1:nrow(params) , function(j) {
+  groupsize = params[j,1]
+  delta = params[j,2]
+  res <- mclapply(1:sims,function(i) {
+    simlrm(genesetdatabase=gsets, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
+      groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
+  },mc.cores=6)
+  res <- do.call(rbind,res)
+  return(res)
+})
+
+lrmres2 <- do.call(rbind,lapply(lrmres,colMeans))
+lrmres2[,"PREC"] <- lrmres2[,"TP"] / ( lrmres2[,"TP"] + lrmres2[,"FP"] )
+
+lrmres2 %>% kbl(caption="LA rank mitch results") %>% kable_paper("hover", full_width = F)
+
+lrmres3p <- do.call(rbind,lapply(groupsizes, function (g) { lrmres2[params$groupsizes==g,"PREC"] }))
+colnames(lrmres3p) <- deltas
+rownames(lrmres3p) <- groupsizes
+lrmres3p %>% kbl(caption="LA rank mitch precision") %>% kable_paper("hover", full_width = F)
+
+lrmres3r <- do.call(rbind,lapply(groupsizes, function (g) { lrmres2[params$groupsizes==g,"REC"] }))
+colnames(lrmres3r) <- deltas
+rownames(lrmres3r) <- groupsizes
+lrmres3r %>% kbl(caption="LA rank mitch recall") %>% kable_paper("hover", full_width = F)
+
+F1(lrmres3p,lacmres3r) %>% kbl(caption="LA rank mitch F1") %>% kable_paper("hover", full_width = F)
 
@@
 
-save.image("GSE158422_simulate050.Rdata")
+save.image("GSE158422_simulate020.Rdata")
 

### 2023-12-20 figs/GSE158422_simulate100.Rmd
fig dumb bug
yet more bugfixes
final?

@@
 
-save.image("GSE158422_simulate050.Rdata")
+save.image("GSE158422_simulate100.Rdata")
 

### 2023-12-20 figs/GSE158422.Rmd
added main script

--- 2022-11-14 GSE158433/GSE158422.Rmd
@@
 sample_metadata <- pData(phenoData(gse))
+sample_metadata$patient <- factor(sapply(strsplit(sample_metadata$title,"_"),"[[",1))
+
 targets <- sample_metadata
@@
 sex_female <- as.numeric(colMeans(xdat)  > -0.1)
+names(sex_female) <- colnames(xdat)
 
@@
 
-#MDS plot 1
+MDS plot 1
 
@@
 
-samplesheet<-targets
-groups <- factor(samplesheet$case,levels=c("tumor: normal","tumor: tumor"))
-sex <- factor(samplesheet$sex,levels=c("0","1"))
+samplesheet <- targets
+
+patient <- factor(sapply(strsplit(samplesheet$title,"_"),"[[",1))
+
+tumor <- factor(samplesheet$`tumor:ch1`,levels=c("normal","tumor"))
+
+sex <- sex_female
+
 dim(Mval)
@@
 
-GSE158422 <- dm_analysis(samplesheet=samplesheet,
-  sex=sex,groups=groups,mx=Mval,name="GSE158422",
-  myann=myann ,beta= beta)
+design <- model.matrix(~ sex + patient + tumor )
+mx <- Mval_flt
+name="control vs tumor"
+
+mxs <- mx[,which( colnames(mx) %in% samplesheet$Basename )]
+fit.reduced <- lmFit(mxs,design)
+fit.reduced <- eBayes(fit.reduced)
+summary(decideTests(fit.reduced))
+dm <- topTable(fit.reduced,coef=ncol(design), number = Inf)
+dma <- merge(myann,dm,by=0)
+dma <- dma[order(dma$P.Value),]
+dm_up <- rownames(subset(dm,adj.P.Val<0.05 & logFC>0))
+dm_dn <- rownames(subset(dm,adj.P.Val<0.05 & logFC<0))
+sig <- min(length(dm_up),length(dm_dn))
+
+confects <- limma_confects(fit.reduced, coef=3, fdr=0.05)
+
+if (sig > 10) {
+  dmr <- run_dmrcate(mx=mxs,design=design)
+  head(dmr)
+  comp <- compartment_enrichment(dma)
+  comp
+  cgi <- cgi_enrichment(dma)
+  cgi
+} else {
+  dmr <- NULL
+  comp <- NULL
+  cgi <- NULL
+}
+#make_dm_plots(dm = dm ,name=name , mx=beta,mxs=mxs, groups = groups,
+  confects=confects,dmr = dmr, comp=comp, cgi=cgi)
+
+GSE158422 <- list("dma"=dma, "dm_up"=dm_up, "dm_dn"=dm_dn, "confects"=confects,
+  "dmr"= dmr, "comp"=comp, "cgi"=cgi, "fit"=fit.reduced)
 
@@
 saveRDS(GSE158422,file="GSE158422.rds")
+saveRDS(design,file="GSE158422_design.rds")
+saveRDS(mx,file="GSE158422_mx.rds")
+
 write.table(GSE158422$dma,file="GSE158422_limma.tsv",sep="\t",quote=FALSE)

### 2023-12-20 figs/main.Rmd
added main script

---
title: "Regenerate figures for the GMEA paper"
author: "The GMEA team"
date: "`r Sys.Date()`"
output:
  html_document:
  toc: true
theme: cosmo
---

## Introduction

Here we evaluate the performance of different FCS methods for Infinium methylation analysis.

# Workflow

## Simulation: Figure 1 and 2

```{r,fig12}

GSE158422.Rmd
GSE158422_simulate020.Rmd
GSE158422_simulate050.Rmd
GSE158422_simulate100.Rmd
GSE158422_simulate_summary.Rmd

```

## Sensitivity: Figure 3 and 4

```{r,fig34}

GSE158422_sensitivity.Rmd

```

## Integration with RNA: Figure 5

```{r,fig5}

GSE158422_rna_int.Rmd

```

## Aging study

```{r,fig6}

aging_example.Rmd

```

## IVF example

```{r,fig7}

ART_estill2016.Rmd
ART_novakovic.Rmd
art_example.Rmd

```

# Session information

```{r,session}

sessionInfo()

```

### 2023-12-20 figs/main.Rmd
added main script

@@
 
-GSE158422.Rmd
-GSE158422_simulate020.Rmd
-GSE158422_simulate050.Rmd
-GSE158422_simulate100.Rmd
-GSE158422_simulate_summary.Rmd
+rmarkdown::render("GSE158422.Rmd")
+rmarkdown::render("GSE158422_simulate020.Rmd")
+rmarkdown::render("GSE158422_simulate050.Rmd")
+rmarkdown::render("GSE158422_simulate100.Rmd")
+rmarkdown::render("GSE158422_simulate_summary.Rmd")
 
@@
 
-GSE158422_sensitivity.Rmd
+rmarkdown::render("GSE158422_sensitivity.Rmd")
 
@@
 
-GSE158422_rna_int.Rmd
+rmarkdown::render("GSE158422_rna_int.Rmd")
 
@@
 
-aging_example.Rmd
+rmarkdown::render("aging_example.Rmd")
 
@@
 
-ART_estill2016.Rmd
-ART_novakovic.Rmd
-art_example.Rmd
+rmarkdown::render("ART_estill2016.Rmd")
+rmarkdown::render("ART_novakovic.Rmd")
+rmarkdown::render("art_example.Rmd")
 

### 2023-12-21 figs/GSE158422.Rmd
fix source path
increase dl timeout
fix timeout

@@
 
-source("../meth_functions.R")
+source("meth_functions.R")
 
@@
 if(!file.exists(ARRAY_DATA)){
+  options(timeout=10000000000000)
   download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE158422&format=file",
@@
 if(!file.exists(SERIES_MATRIX)){
+  options(timeout=10000000000000)
   download.file("https://ftp.ncbi.nlm.nih.gov/geo/series/GSE158nnn/GSE158422/matrix/GSE158422_series_matrix.txt.gz",

### 2023-12-21 figs/GSE158422.Rmd
fix timeout

@@
     destfile=ARRAY_DATA)
-  untar(tarfile = ARRAY_DATA)
 }
+untar(tarfile = ARRAY_DATA,exdir="IDAT")
 

### 2023-12-21 figs/GSE158422.Rmd
fix plots

@@
 }
-#make_dm_plots(dm = dm ,name=name , mx=beta,mxs=mxs, groups = groups,
+
+make_dm_plots(dm = dm ,name=name , mx=beta,mxs=mxs, groups = groups,
   confects=confects,dmr = dmr, comp=comp, cgi=cgi)

### 2023-12-21 figs/GSE158422.Rmd
samplesheet
fixes

@@
 
-tumor <- factor(samplesheet$`tumor:ch1`,levels=c("normal","tumor"))
+tumor <- factor(samplesheet$tumor,levels=c("normal","tumor"))
 
@@
 
-make_dm_plots(dm = dm ,name=name , mx=beta,mxs=mxs, groups = groups,
+make_dm_plots(dm = dm ,name=name , mx=beta,mxs=mxs, groups = tumor,
   confects=confects,dmr = dmr, comp=comp, cgi=cgi)

### 2023-12-21 manuscript/manuscript.Rmd
mswork

@@
 
-* DNA methylation is the most widely studied epigenetic mark, for its role in development and
+DNA methylation is the most widely studied epigenetic mark, for its role in development and
 disease [@Robertson2005-tu].
 
-* Many hundreds of epigenome-wide association studies (EWASs) are conducted each year to understand DNA
+Many hundreds of epigenome-wide association studies (EWASs) are conducted each year to understand DNA
 methylation patterns in disease and other contexts [@Li2019-og].
 
-* Infinium arrays remain the preferred platform for EWASs due to low cost and analytical simplicity
+Infinium arrays remain the preferred platform for EWASs due to low cost and analytical simplicity
 as compared to high throughput methylation sequencing [@Wei2021-qj].
 
-* Infinium arrays typically have multiple probes per gene in different locations including promoters,
+Infinium arrays typically have multiple probes per gene in different locations including promoters,
 CpG islands, gene bodies and enhancers [@Pidsley2016-ix], which complicates downstream functional enrichment analysis,
@@
 
-* Enrichment analysis comes in two popular types; over-representation analysis (ORA) and functional
+Enrichment analysis comes in two popular types; over-representation analysis (ORA) and functional
 class scoring (FCS) [@Khatri2012-qy].
@@
 regulation score (eg: fold change, confident effect size) followed by a test to assess whether each
-set of genes has distribution of scores that is different to the null [@Subramanian2005-ar].
+set of genes has a distribution of scores that is different to the null [@Subramanian2005-ar].
 
-The `gsameth()` function of the `missMethyl` package is the state-of-the-art methods for ORA of infinium methylation array
+The `gsameth()` function of the `missMethyl` package is the state-of-the-art method for ORA of infinium methylation array
 data as it addresses issues related to probes belonging to more than one gene and the fact that genes can have multiple
@@
 
-FCS methods `ebGSEA` and `methylGSA` have been applied to EWAS data and are suggested to have better sensitivity to
+FCS methods `ebGSEA` and `methylGSA` have been developed for EWAS data and are suggested to have better sensitivity to
 determine subtle associations between pathways and differential methylation [@Dong2019-np;@Ren2019-mr].
@@
 
-Here we aim to develop and evaluate methods for two-tailed FCS of Infinium methylation array data that reports enrichment
-scores.
-
-* Here, we systematically evaluate different approaches to apply FCS to Infinium array.
-
-* Due to the fact that genes have many probes, there are many concevable ways that FCS could be applied
+Here we aim to develop and evaluate methods for two-tailed FCS of Infinium methylation array data that addresses these
+limitations.
+Due to the fact that genes have a variable number of probes, there are many concevable ways that FCS could be applied
 to Infinium Methylation data.
-
-* Method group 1: Limma is performed on the probe level and the probes beloning to each gene are scored
-and these scores are aggregated to the gene level.
-The gene scores undergo an enrichment test.
-
-* Method group 2: The probe methylation values beloning to each gene are aggregated before differential
+The first set of methods use differential probe methylation data as an input.
+The differential methylation of probes beloning to each gene are scored and these gene scores
+undergo an enrichment test.
+The second set of methods first aggregate probe methylation values for each gene before differential
 methylation analysis with limma.
-Then, differential methylation results for each gene are scored, followed by an enrichment test.
-
-With the above, there are options for how the aggregation steps are done, whether parametric or
-non-parametric tests are used and whether the enrichment tests are self-contained or competitive.
-
+Then, differential gene methylation results are used for a downstream enrichment test.
+With the above methods there are several methodological choices, such as the type of aggregation (eg: mean, median), and the type of downstream test (eg: parametric or nonparametric).
 To determine the best approach to applying FCS to Infinium methylation data, in this study we use
-simulated data.
-The proposed FCS methods are also compared against existing ORA methods.
+simulated data and compare results to existing ORA methods.
+We then examine the sensitivity of selected methods on real cancer data, investigate the association with gene expression and end
+with additional examples related to human aging and assisted reproductive technologies.
 
@@
 meaning the data needs to be summarised from probes to genes and then to gene sets.
-Here, we outline seven potential approaches for FCS of Infinium methylation data:
+Here, we outline eight potential approaches for FCS of Infinium methylation data:
 
@@
 Similar to LAW, however instead of the Wilcoxon test, the Mitch package is used to execute a two-way,
-two-sample ANOVA-on-ranks test (Kruskal Wallis).
+two-sample ANOVA-on-ranks test (Kruskal-Wallis).
 
-5. Aggregate Limma t-test (ALT).
+5. Limma Rank Mitch (LRM).
+Probes are ranked by t-statistic. The mean rank values for probes beloning to each gene are computed and used
+as input for a two-way, two-sample KruskalWallis test.
+
+6. Aggregate Limma t-test (ALT).
 In this method, all probe measurements beloning to each gene are averaged before conducting differential
@@
 
-6. Aggregate Limma Wilcox (ALW).
+7. Aggregate Limma Wilcox (ALW).
 Similar to ALT, however it uses the Wilcoxon non-parametric test for gene set enrichment.
 
-7. Aggregate Limma Mitch (ALM).
-Similar to ALW, however uses the Mitch package to execute a two-way, two-sample ANOVA-on-ranks test
-(Kruskal Wallis).
+8. Aggregate Limma Mitch (ALM).
+Similar to ALW, however uses the Mitch package to execute a two-way, two-sample Kruskal-Wallis test.
 
@@
 
-![Figure 1. Precision, recall and F1 for enrichment tests with simulated data. Gene set sizes vary between 20 and 100 genes per set.](../figs/fig1.png)
+![**Figure 1. Precision, recall and F1 for enrichment tests with simulated data.** Gene set sizes vary between 20 and 100 genes per set.](../figs/fig1.png)
 
@@
 
-![Figure 2. Precision, recall and F1 for GSA and LAM enrichment tests with simulated data over a range of group sizes and delta values. One hundred genes per set.](../figs/fig2.png)
+![**Figure 2. Precision, recall and F1 for GSA and LAM enrichment tests with simulated data over a range of group sizes and delta values.** One hundred genes per set.](../figs/fig2.png)
 
@@
 
-![Figure 3.
-Pathway enrichment analysis comparing normal and cancer samples using LAM and GSA methods.
+![**Figure 3. Pathway enrichment analysis comparing normal and cancer samples using LAM and GSA methods.**
 (A) Euler diagrams showing the overlap in statistically significant pathways (FDR<0.05) from LAM and GSA results.
@@
 (C) Significant LAM pathways with largest S distance, a proxy for effect size.
-(D) Top ranked significant GSA pathways ranked by fold enrichment.](../figs/fig3.png)
+(D) Top ranked significant GSA pathways ranked by fold enrichment.
+](../figs/fig3.png)
 
@@
 
-![Figure 4. Sensitivity of LAM and GSA methods with real cancer array data.
+![**Figure 4. Sensitivity of LAM and GSA methods with real cancer array data.**
 Upper panels show the number of significant pathways identified in the smaller sample that were consistent
@@
 
-![Figure 5.
-Integration of gene expression data with methylation pathways.
+![**Figure 5. Integration of gene expression data with methylation pathways.**
 (A,B) Euler diagrams showing the overlap in statistically significant pathways (FDR<0.05) from gene expression and GSA results (A) and LAM (B) results.
-(C) Multi-dimensional enrichment analysis of promoter and gene body methylation with gene expression using mitch.
-Top 20 gene sets shown with largest s distance after FDR filtering at 0.05.](../figs/fig5.png)
+(C) Multi-dimensional enrichment analysis of promoter and gene body methylation with gene expression using mitch. Top 20 gene sets shown with largest s distance after FDR filtering at 0.05.
+](../figs/fig5.png)
 
@@
 
-![Figure 6. Pathway level DNA methylation alterations with chronological age. (A) Contour heatmap showing the similarity in gene methylation score ranks in the discovery and replication studies.
+![**Figure 6. Pathway level DNA methylation alterations with chronological age.**
+(A) Contour heatmap showing the similarity in gene methylation score ranks in the discovery and replication studies.
 (B) Mitch pathway enrichment scores (S distances) in discovery and replication studies. Pathways with MANOVA FDR<0.05 are shown in red while others are shaded gray.
@@
 
-![Figure 7.
-Pathway level DNA methylation differences in natural and IVF conceived infants.
+![**Figure 7. Pathway level DNA methylation differences in natural and IVF conceived infants.**
 (A) Contour heatmap showing the similarity in gene methylation score ranks in the Estill (450k) and Novakovic (EPIC) studies.

### 2023-12-21 figs/GSE158422_pkgdev.Rmd
must update gene symbols

@@
   library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
+  library("HGNChelper")
   library("tictoc")
@@
 
-```
-
-Now try for efficient method to incorporate into the package.
-
-```{r,anno2}
-
-library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
+tic()
+new.hgnc.table <- getCurrentHumanMap()
+fix <- checkGeneSymbols(gt$gene,map=new.hgnc.table)
+fix2 <- fix[which(fix$x != fix$Suggested.Symbol),]
+length(unique(fix2$x))
+gt$gene <- fix$Suggested.Symbol
+toc()
 
@@
 
+# fix gene symbols
+tic()
+fix <- checkGeneSymbols(gt2$gene,map=new.hgnc.table)
+fix2 <- fix[which(fix$x != fix$Suggested.Symbol),]
+length(unique(fix2$x))
+gt2$gene <- fix$Suggested.Symbol
+toc()
+
 tic()

### 2023-12-21 figs/GSE158422_rna_int.Rmd
must update gene symbols

@@
 
+new.hgnc.table <- getCurrentHumanMap()
+fix <- checkGeneSymbols(gt$gene,map=new.hgnc.table)
+fix2 <- fix[which(fix$x != fix$Suggested.Symbol),]
+length(unique(fix2$x))
+gt$gene <- fix$Suggested.Symbol
+
 ```
@@
 
+fix <- checkGeneSymbols(gb$gene,map=new.hgnc.table)
+fix2 <- fix[which(fix$x != fix$Suggested.Symbol),]
+length(unique(fix2$x))
+gb$gene <- fix$Suggested.Symbol
+
 tss <- lapply(1:nrow(dm),function(i) {
@@
 str(gp)
+
+fix <- checkGeneSymbols(gp$gene,map=new.hgnc.table)
+fix2 <- fix[which(fix$x != fix$Suggested.Symbol),]
+length(unique(fix2$x))
+gp$gene <- fix$Suggested.Symbol
 

### 2023-12-21 figs/GSE158422_sensitivity.Rmd
must update gene symbols

@@
 myann <- data.frame(anno[,c("UCSC_RefGene_Name","UCSC_RefGene_Group","Islands_Name","Relation_to_Island")])
+gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
+gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
+names(gp2) <- rownames(gp)
+gp2 <- lapply(gp2,unique)
+gt <- stack(gp2)
+colnames(gt) <- c("gene","probe")
+dim(gt)
+str(gt)
+head(gt)
+
+new.hgnc.table <- getCurrentHumanMap()
+fix <- checkGeneSymbols(gt$gene,map=new.hgnc.table)
+fix2 <- fix[which(fix$x != fix$Suggested.Symbol),]
+length(unique(fix2$x))
+gt$gene <- fix$Suggested.Symbol
 
@@
 
-tic()
-anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
-myann <- data.frame(anno[,c("UCSC_RefGene_Name","UCSC_RefGene_Group","Islands_Name","Relation_to_Island")])
-gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
-gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
-names(gp2) <- rownames(gp)
-gp2 <- lapply(gp2,unique)
-gt <- stack(gp2)
-colnames(gt) <- c("gene","probe")
-dim(gt)
-str(gt)
-toc() #8.735s
-
 dm$rank <-  rank(dm$t) - nrow(subset(dm,t<0))
@@
 ```{r,rungmea}
-
-tic()
-anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
-myann <- data.frame(anno[,c("UCSC_RefGene_Name","UCSC_RefGene_Group","Islands_Name","Relation_to_Island")])
-gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
-gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
-names(gp2) <- rownames(gp)
-gp2 <- lapply(gp2,unique)
-gt <- stack(gp2)
-colnames(gt) <- c("gene","probe")
-dim(gt)
-str(gt)
-toc() #8.735s
 
@@
 lamsigsets2 <- gsub(" FALSE"," DN",lamsigsets2)
-
-```
-
-## LMRM - limma mean rank mitch
-
-```{r,lmrm1}
-
-tic()
-anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
-myann <- data.frame(anno[,c("UCSC_RefGene_Name","UCSC_RefGene_Group","Islands_Name","Relation_to_Island")])
-gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
-gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
-names(gp2) <- rownames(gp)
-gp2 <- lapply(gp2,unique)
-gt <- stack(gp2)
-colnames(gt) <- c("gene","probe")
-dim(gt)
-str(gt)
-toc() #8.735s
-
-dm$rank <-  rank(dm$t) - nrow(subset(dm,t<0))
-mm <- merge(dm,gt,by.x=0,by.y="probe")
-head(mm)
-mma <-aggregate(mm$rank ~ gene,mm,mean)
-rownames(mma) <- mma$gene
-mma$gene = NULL
-colnames(mma) <- "meanrank"
-lmrmres <- runmitch(m=mma,genesets=gs_symbols,cores=16)
-
-lmrmsig <- subset(lmrmres,p.adjustANOVA<0.05)
-
-top <- head(lmrmsig,20)
-top <- top[order(top$s.dist),]
-
-top %>%
-  kbl(caption="Top gene sets (biggest effect size with FDR<0.05)") %>%
-  kable_paper("hover", full_width = F)
-
-barnames <- gsub("_"," ",gsub("REACTOME_","",rownames(top)))
-cols <- gsub("TRUE","red",gsub("FALSE","blue",as.character(top$s.dist>0)))
-
-par(mar = c(5.1, 29.1, 4.1, 2.1))
-barplot(abs(top$s.dist),horiz=TRUE,las=1,names.arg=barnames, cex.names=0.65,
-  cex.axis=0.8,col=cols, xlab="Enrichment score",main="Reactomes")
-grid()
-
-par( mar = c(5.1, 4.1, 4.1, 2.1) )
-
-hist(lmrmsig$s.dist)
-
-# distinguish up (TRUE) and down (FALSE)
-lmrmsigsets <- paste(rownames(lmrmsig),as.character(lmrmsig$s.dist > 0))
-lmrmsigsets2 <- gsub(" TRUE"," UP",lmrmsigsets)
-lmrmsigsets2 <- gsub(" FALSE"," DN",lmrmsigsets2)
 

### 2023-12-21 figs/GSE158422_simulate020.Rmd
must update gene symbols

@@
   library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
+  library("HGNChelper")
   library('org.Hs.eg.db')
@@
 head(gt)
+
+new.hgnc.table <- getCurrentHumanMap()
+fix <- checkGeneSymbols(gt$gene,map=new.hgnc.table)
+fix2 <- fix[which(fix$x != fix$Suggested.Symbol),]
+length(unique(fix2$x))
+gt$gene <- fix$Suggested.Symbol
 

### 2023-12-21 figs/GSE158422_simulate050.Rmd
final?
must update gene symbols

@@
 num_dm_sets=50
-sims=5
+sims=50
 

### 2023-12-21 figs/aging_example.Rmd
must update gene symbols

@@
 
+Update old gene symbols.
+
 ```{r,anno1}
@@
 toc() #9.0s
+
+tic()
+new.hgnc.table <- getCurrentHumanMap()
+fix <- checkGeneSymbols(gt$gene,map=new.hgnc.table)
+fix2 <- fix[which(fix$x != fix$Suggested.Symbol),]
+length(unique(fix2$x))
+gt$gene <- fix$Suggested.Symbol
+toc()
 

### 2023-12-21 figs/art_example.Rmd
must update gene symbols

@@
 
+Update gene symbols.
+
 ```{r,anno1}
@@
 toc() #6.3s
+
+tic()
+new.hgnc.table <- getCurrentHumanMap()
+fix <- checkGeneSymbols(gt1$gene,map=new.hgnc.table)
+fix2 <- fix[which(fix$x != fix$Suggested.Symbol),]
+length(unique(fix2$x))
+gt1$gene <- fix$Suggested.Symbol
+toc()
 
@@
 toc() #10.6s
+
+tic()
+new.hgnc.table <- getCurrentHumanMap()
+fix <- checkGeneSymbols(gt2$gene,map=new.hgnc.table)
+fix2 <- fix[which(fix$x != fix$Suggested.Symbol),]
+length(unique(fix2$x))
+gt2$gene <- fix$Suggested.Symbol
+toc()
 

### 2023-12-21 figs/GSE158422_pkgdev.Rmd
put new HGNC table in the repo

@@
 tic()
-new.hgnc.table <- getCurrentHumanMap()
+#new.hgnc.table <- getCurrentHumanMap()
+new.hgnc.table <- readRDS("new.hgnc.table.rds")
 fix <- checkGeneSymbols(gt$gene,map=new.hgnc.table)

### 2023-12-21 figs/GSE158422_rna_int.Rmd
put new HGNC table in the repo

@@
 human=x$row.names
-fix <- checkGeneSymbols(human)
+new.hgnc.table <- readRDS("new.hgnc.table.rds")
+fix <- checkGeneSymbols(human,map=new.hgnc.table)
 x$gene <- fix$Suggested.Symbol
@@
 
-new.hgnc.table <- getCurrentHumanMap()
+#new.hgnc.table <- getCurrentHumanMap()
+new.hgnc.table <- readRDS("new.hgnc.table.rds")
 fix <- checkGeneSymbols(gt$gene,map=new.hgnc.table)

### 2023-12-21 figs/art_example.Rmd
put new HGNC table in the repo

@@
 tic()
-new.hgnc.table <- getCurrentHumanMap()
+#new.hgnc.table <- getCurrentHumanMap()
+new.hgnc.table <- readRDS("new.hgnc.table.rds")
 fix <- checkGeneSymbols(gt1$gene,map=new.hgnc.table)

### 2023-12-21 figs/GSE158422_simulate020.Rmd
fix gene catalog

@@
 
-gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
+gene_catalog <- unique(gt$gene)
 
@@
   # generate gene sets
-  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
+  gene_catalog <- unique(gt$gene)
   lengths <- unname(unlist(lapply(genesetdatabase,length)))

### 2023-12-21 manuscript/manuscript.Rmd
ms

@@
 into consideration the variable number of probes followed by an FCS test based on z-scores [@Ren2019-mr].
-Secondly, `methylglm` uses a logistic regression approach to model differential methylation for genes inside and outside eachgene set [@Ren2019-mr].
+Secondly, `methylglm` uses a logistic regression approach to model differential methylation for genes inside and outside each gene set [@Ren2019-mr].
 These methods are indeed more sensitive than ORA approaches, but as they are based on one-tailed tests irrespective of
@@
 changes on genomic regulation.
+Moreover, as pathways tend to be either upregulated or downregulated in omics assays, combining the two tends to dilute
+their signal and reduce sensitivity [@Hong2014-ab].
 Neither gsameth, ebGSEA nor methylGSA report computed enrichment scores, which hinders downstream interpretation, as the
-enrichment score is a useful surrogate for effect size.
+enrichment score is a useful surrogate for effect size (check this).
 Without an enrichment score, users are limited to using statistical significance values to prioritise results, which
@@
 to Infinium Methylation data.
-The first set of methods use differential probe methylation data as an input.
+The first set of methods use differential probe methylation data from `limma` [@Ritchie2015-zn] as an input.
 The differential methylation of probes beloning to each gene are scored and these gene scores
@@
 were selected.
+We conducted separate tests for increased and decreased probes, and we specified the background as all probes that
+passed quality control filtering.
 
-### Method validation
+### Method validation using simulations
 
-To assess these methods, we adopted a simulated data approach.
-EPIC Infinium array data derived from non-cancerous lung tissue of 37 cancer patients was used for our
-evaluations (GSE158422).
-We downloaded raw intensity EPIC IDAT files from GEO for study GSE158422 which consists of
-lung tumour and normal adjacent tissues from 37 patients.
+To assess these methods, we adopted a simulated data approach based on the selected modification of real methylation data.
+We downloaded raw intensity EPIC IDAT files from NCBI Gene Expression Omnibus (GEO) for study GSE158422 which consists of
+lung tumour and normal adjacent tissues from 37 patients [@Cho2022-yd].
+For the simulations, only non-cancerous datasets were used.
 Probe annotations were obtained from the "IlluminaHumanMethylationEPICanno.ilm10b4.hg19" Bioconductor
@@
 3 and 12.
-One thousand random gene sets were created with sizes varying between 20 and 100 members.
+One thousand random gene sets were created with sizes varying between 20 and 100 members, with member genes drawn from the
+annotation set.
 Random gene sets were used to avoid problems caused by the large overlap among real gene sets.
@@
 For those selected genes, half of the annotated probes were selected.
-We adjusted the m-values in the case group by a specified amount, which we call the "delta", which we varied
-between 0.1 and 0.4.
+We adjusted the M-values in the case group by a specified amount, which we call the "delta", which we varied
+between 0.1 and 0.5.
 Following the incorporation of methylation changes to selected probes in the case group, the data underwent
 limma differential methylation analysis, followed by enrichment analysis using `GSAmeth` approach, or one of
-the seven FCS methods outlined above.
+the eight FCS methods outlined above.
 Enrichment results were then filtered for FDR<0.05 and correct direction of methylation change.
@@
 
-The same dataset including all 37 patients underwent pathway enrichment analysis using the GSAmeth and LAM methods
+The full GSE158422 dataset including normal and cancer samples of all 37 patients underwent differential analysis with limma
+correcting for patient of origin effects.
+Downstream pathway enrichment analysis was conducted using the GSAmeth and LAM methods
 with Reactome pathways downloaded from MSigDB website (version 2023.1).
+LAM relies on probe-gene associations from the "UCSC_RefGene_Name" column of the "IlluminaHumanMethylationEPICanno.ilm10b4.hg19"
+annotation.
+As these annotations are several years old, we updated them using HGNChelper (database current as of 21-Dec-2023) [@Oh2022-ts].
+Circa 12% of defunct gene symbols were updated.
 Euler diagrams were created with the `eulerr` R package.
-Then, a random subset of n patients was selected followed by pathway enrichment analysis with the
+To test sensitivity, a random subset of n patients was selected followed by pathway enrichment analysis with the
 respective method, with n varied between 2 and 30.
-The subset results were compared to the full n=37 results to test the sensitivity of these enrichment
+The subset results were compared to the full group (n=37) results to test the sensitivity of these enrichment
 methods at differenent effect size levels, and whether findings at a smaller sample size were consistent with the full group.
-This process was repeated 50 times.
+This process was repeated 50 times at each sample size.
 
@@
 
-RNA-seq based gene counts for the same set of lung cancer patients were downloaded from GEO [GSE158420](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE158420).
+RNA-seq based gene counts for the same set of lung cancer patients were downloaded from GEO (accession: GSE158420).
 HGNChelper was used to repair transmogrified gene names, followed by DESeq2 based differential analysis.

### 2023-12-21 manuscript/manuscript.Rmd
add gmts

@@
 ---
-title: "Finally, a functional class scoring method for enrichment analysis of Infinium DNA methylation data"
+title: "Direction-aware functional class scoring enrichment analysis of Infinium DNA methylation data"
 date: "`r Sys.Date()`"

### 2023-12-21 figs/GSE158422_pkgdev.Rmd
update
remove stuff

@@
 
+Source code: https://github.com/markziemann/gmea/blob/main/figs/GSE158422_pkgdev.Rmd
+
 ## Requirements
+
+Load packages.
 
@@
 
+Updating gene names using the HGNC package because a lot of them have been updated over time.
+
 ```{r,anno1}
@@
 ## Prep limma
+
+Here we conduct a differential analysis using limma.
 
@@
 head(myann)
-table(anno$UCSC_RefGene_Group)
 table(unlist(strsplit(myann$UCSC_RefGene_Group,";")))

### 2023-12-21 figs/GSE158422_pkgdev.Rmd
fix promoter analysis

@@
 names(keeps) <- rownames(myann)
-gt2 <- stack(gp2)
+gt2 <- stack(keeps)
 colnames(gt2) <- c("gene","probe")

### 2023-12-21 figs/GSE158422_pkgdev.Rmd
cut down dm
pretty tables

@@
 dm <- topTable(fit.reduced,coef=ncol(design), number = Inf)
-head(dm)
+head(dm) %>%
+  kbl(caption = "Top significant probes") %>%
+  kable_paper("hover", full_width = F)
 toc() #25.53, 25.50, 25.51
@@
 toc() #9.58s, 9.71, 9.69
-head(m)
+head(m) %>%
+  kbl(caption = "Example of differential gene methylation scores used for pathway analysis") %>%
+  kable_paper("hover", full_width = F)
 
@@
 nrow(dn)
-head(up,10)
-head(dn,10)
+
+head(up,10) %>%
+  kbl(caption = "Top significant pathways with higher methylation") %>%
+  kable_paper("hover", full_width = F)
+
+head(dn,10) %>%
+  kbl(caption = "Top significant pathways with lower methylation") %>%
+  kable_paper("hover", full_width = F)
 
@@
 
-head(myann)
+head(myann) %>%
+  kbl(caption = "This is what the annotation looks like") %>%
+  kable_paper("hover", full_width = F)
+
 table(unlist(strsplit(myann$UCSC_RefGene_Group,";")))
@@
 tic()
-m2 <- mitch_import(x=dm,DEtype="limma",geneTable=gt2)
+# need to remove probes not associated to promoters
+dm2 <- dm[(which(rownames(dm) %in% gt2$probe)),]
+m2 <- mitch_import(x=dm2,DEtype="limma",geneTable=gt2)
 toc() #9.68, 9.78, 9.62
-head(m)
-dim(m)
+head(m2) %>%
+  kbl(caption = "This is what the differential methylation scores look like") %>%
+  kable_paper("hover", full_width = F)
+
+dim(m2)
 
@@
 nrow(dn)
-head(up,10)
-head(dn,10)
+
+head(up,10) %>%
+  kbl(caption = "Top significant pathways with increased promoter methylation") %>%
+  kable_paper("hover", full_width = F)
+
+head(dn,10) %>%
+  kbl(caption = "Top significant pathways with lower promoter methylation") %>%
+  kable_paper("hover", full_width = F)
 

### 2023-12-21 manuscript/manuscript.Rmd
mswork

@@
 disease [@Robertson2005-tu].
-
 Many hundreds of epigenome-wide association studies (EWASs) are conducted each year to understand DNA
 methylation patterns in disease and other contexts [@Li2019-og].
-
 Infinium arrays remain the preferred platform for EWASs due to low cost and analytical simplicity
 as compared to high throughput methylation sequencing [@Wei2021-qj].
-
 Infinium arrays typically have multiple probes per gene in different locations including promoters,
@@
 otherwise known as pathway analysis or gene set enrichment analysis.
-
 Enrichment analysis comes in two popular types; over-representation analysis (ORA) and functional
@@
 modelling of gene network activations, but these are yet to be adopted widely [@Xie2021-py].
-
 In ORA, genes with probes that meet an arbitrary significance threshold are selected and compared
@@
 gene list of interest relative to the background [@Tavazoie1999-ng].
-
 FCS takes a different approach by ranking all detected genes by a differential
@@
 set of genes has a distribution of scores that is different to the null [@Subramanian2005-ar].
-
 The `gsameth()` function of the `missMethyl` package is the state-of-the-art method for ORA of infinium methylation array
@@
 on the proportion of probes that meet this threshold.
-
 FCS methods `ebGSEA` and `methylGSA` have been developed for EWAS data and are suggested to have better sensitivity to
@@
 their signal and reduce sensitivity [@Hong2014-ab].
-Neither gsameth, ebGSEA nor methylGSA report computed enrichment scores, which hinders downstream interpretation, as the
-enrichment score is a useful surrogate for effect size (check this).
+Neither GSAmeth nor methylGSA report computed enrichment scores, which hinders downstream interpretation, as the
+enrichment score is a useful surrogate for effect size.
 Without an enrichment score, users are limited to using statistical significance values to prioritise results, which
@@
 
-The above were compared to a standard ORA based approach (GSA), which involves limma on probes, selection of
+The above were compared to a standard ORA-based approach (GSA), which involves limma on probes, selection of
 statistically significant probes (FDR<0.05) for ORA with the `GSAmeth()` function of missMethyl which conducts a
@@
 The full GSE158422 dataset including normal and cancer samples of all 37 patients underwent differential analysis with limma
-correcting for patient of origin effects.
+correcting for patient-of-origin effects.
 Downstream pathway enrichment analysis was conducted using the GSAmeth and LAM methods
@@
 As these annotations are several years old, we updated them using HGNChelper (database current as of 21-Dec-2023) [@Oh2022-ts].
-Circa 12% of defunct gene symbols were updated.
+Circa 12% of gene symbols were updated.
 Euler diagrams were created with the `eulerr` R package.
@@
 respective method, with n varied between 2 and 30.
-The subset results were compared to the full group (n=37) results to test the sensitivity of these enrichment
+Significantly enriched pathways were defined as those with ANOVA FDR values < 0.05.
+The subset significant results were compared to the full group (n=37) results to test the sensitivity of these enrichment
 methods at differenent effect size levels, and whether findings at a smaller sample size were consistent with the full group.
@@
 ### Pathway-level differential methylation in aging
-In order to deminstrate the utility of the LAM method for 
-(2) the association of elevated plasma homocysteine with methylation patterns.
+
+In order to deminstrate the utility of the LAM method for large EWAS studies, we examined
+methylation profile associations with age of two independent cohorts involving a total of 7036
+participants [@McCartney2020-ob].
+Limma summary statistics for chromological age were read into R and enrichment analysis was conducted as above.
+Briefly, gene-probe associations were first obtained from the
+"IlluminaHumanMethylationEPICanno.ilm10b4.hg19" dataset, followed by update of gene symbols with
+HGNChelper.
+Data were imported with mitch and enrichment analysis was conducted with gene sets from Reactome.
 
 ### Pathway-level differential methylation in assisted reproductive technology
-(i) the effect of assisted reproductive technology DNA methylation
+
+In order to demonstrate that methylation patterns across different array platforms can be compared using LAM,
+we examined methylation differences in infants associated with assisted reproduction as described by two
+independent studies [@Estill2016-jh;@Novakovic2019-jt].
+The Estill study was published in 2016 and is based on the HM450K array and the data is available from GEO Accession GSE79257.
+The Novakovic study in was published 2019, and used the EPIC methylation array and the data is available from GEO under accession
+GSE131433.
+Various assisted reproductive technology conception groups are described in these studies, but we examined
+just fresh in vitro fertilised conceived infants compared to naturally conceived infants.
+
+Raw intensity files (IDAT format) were read into R with minfi [aryee], normalistion was conducted with the SWAN method [maksimovic],
+and probe filtering was conducted to remove proves with a detection p-value > 0.01 in addition to probes located on X or Y
+chromosomes.
+M values were computed and underwent differential methylation analysis using limma [ritchie].
+The Estill study included 43 naturally conceived infants and 38 with fresh embryo transfer.
+The Novakovic study included XX naturally conceived infants and XX with fresh embryo transfer.
+Limma differential methylation tables were then imported with mitch.
+Probe gene associations for HM450K array were established using the "IlluminaHumanMethylation450kanno.ilmn12.hg19" annotation set
+and the gene symbols were updated with HGNChelper (database current as of 21-Dec-2023).
+As the Novakovic study involved EPIC array data, gene-probe associations were obtained from the
+"IlluminaHumanMethylationEPICanno.ilm10b4.hg19" annotation set, and gene symbols were updated with
+HGNChelper.
+Limma tables were read into R and imported with mitch.
+Gene level differential methylation score tables were merged and underwent mitch enrichment analysis using Reactome gene sets.
 

### 2023-12-22 figs/GSE158422_simulate020.Rmd
reduce cores

@@
   m1$gene=NULL
-  lares1 <- ttenrich(m=m1,genesets=gsets,cores=4,testtype="competitive")
+  lares1 <- ttenrich(m=m1,genesets=gsets,cores=2,testtype="competitive")
   gsig_up3 <- rownames(subset(lares1, fdr < 0.05 & t_mean > 0))
@@
   m1$gene=NULL
-  lares1 <- wtenrich(m=m1,genesets=gsets,cores=4,testtype="competitive")
+  lares1 <- wtenrich(m=m1,genesets=gsets,cores=2,testtype="competitive")
   gsig_up3 <- rownames(subset(lares1, fdr < 0.05 & t_mean > 0))
@@
   m1$gene=NULL
-  lamres1 <- runmitch(m=m1,genesets=gsets,cores=4)
+  lamres1 <- runmitch(m=m1,genesets=gsets,cores=2)
   gsig_up3 <- rownames( subset( lamres1, p.adjustANOVA < 0.05 & s.dist > 0 ) )
@@
   colnames(mma) <- "meanrank"
-  lrmres <- runmitch(m=mma,genesets=gsets,cores=4)
+  lrmres <- runmitch(m=mma,genesets=gsets,cores=2)
   gsig_up3 <- rownames( subset( lrmres, p.adjustANOVA < 0.05 & s.dist > 0 ) )
@@
   colnames(m1) <- "t"
-  alres1 <- ttenrich(m=m1,genesets=gsets,cores=4,testtype="competitive")
+  alres1 <- ttenrich(m=m1,genesets=gsets,cores=2,testtype="competitive")
   # summarise results
@@
   colnames(m1) <- "t"
-  alres1 <- wtenrich(m=m1,genesets=gsets,cores=4,testtype="competitive")
+  alres1 <- wtenrich(m=m1,genesets=gsets,cores=2,testtype="competitive")
   # summarise results
@@
   colnames(m1) <- "t"
-  almres1 <- runmitch(m=m1,genesets=gsets,cores=4)
+  almres1 <- runmitch(m=m1,genesets=gsets,cores=2)
   # summarise results

### 2023-12-22 figs/GSE158422_simulate050.Rmd
reduce cores

@@
 
-Cannot be run in multicore due to fragility of AnnotationDbi SQLite objects.
+CCannot be run in multicore due to fragility of AnnotationDbi SQLite objects.
 

### 2023-12-26 figs/GSE158422_sensitivity.Rmd
streamline

@@
 
-The following function is key for the aggregation of probe t-statistic values by gene using the arithmetic mean.
-In this step, ~800k probe values are summarised to ~28k genes.
-
-```{r,lamfunction}
-
-# LAM aggregate
-agg <- function(dm,cores=1) {
-  dm <- dm[which(dm$UCSC_RefGene_Name!=""),]
-  gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
-  gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
-  gnl <- mclapply(gnl,unique,mc.cores=cores)
-  dm$UCSC_RefGene_Name <- gnl
-  l <- mclapply(1:nrow(dm), function(i) {
-    a <- dm[i,]
-    len <- length(a[[1]][[1]])
-    tvals <- as.numeric(rep(a["t"],len))
-    genes <- a[[1]][[1]]
-    data.frame(genes,tvals)
-  },mc.cores=cores)
-  df <- do.call(rbind,l)
-  keep <- names(which(table(df$genes)>1))
-  df <- df[df$genes %in% keep,]
-  gn <- unique(df$genes)
-  gme_res <- lapply( 1:length(gn), function(i) {
-    g <- gn[i]
-    tstats <- df[which(df$genes==g),"tvals"]
-    myn <- length(tstats)
-    mymean <- mean(tstats)
-    mymedian <- median(tstats)
-    top <- tstats[order(-abs(tstats))][1]
-    if ( length(tstats) > 2 ) {
-      ttest <- t.test(tstats)
-      pval <- ttest$p.value
-    } else {
-      pval = 1
-    }
-    res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,
-      "median"=mymedian, "top"=top, "pval"=pval)
-  } )
-  gme_res_df <- do.call(rbind, gme_res)
-  rownames(gme_res_df) <- gme_res_df[,1]
-  gme_res_df <- gme_res_df[,-1]
-  tmp <- apply(gme_res_df,2,as.numeric)
-  rownames(tmp) <- rownames(gme_res_df)
-  gme_res_df <- as.data.frame(tmp)
-  gme_res_df$sig <- -log10(gme_res_df[,"pval"])
-  gme_res_df <- gme_res_df[order(-gme_res_df$sig),]
-  gme_res_df$fdr <- p.adjust(gme_res_df$pval)
-  out <- list("df"=df,"gme_res_df"=gme_res_df)
-  return(out)
-}
-
-```
-
 Function to run mitch analysis.
@@
 tic()
-dmagg <- agg(dm,cores=16)
-m <- dmagg$gme_res_df[,"mean",drop=FALSE]
-lamres <- runmitch(m=m,genesets=gs_symbols,cores=16)
+dd <- merge(dm,gt,by.x=0,by.y="probe")
+m <- aggregate(t ~ gene,dd,mean)
+rownames(m) <- m$gene
+m$gene=NULL
+lamres <- runmitch(m=m,genesets=gs_symbols,cores=8)
 toc()
-
-#tic()
-#dmagg <- agg(dm,cores=16)
-#m <- dmagg$gme_res_df[,"mean",drop=FALSE]
-#lamres <- runmitch(m=m,genesets=gs_symbols,cores=16)
-#toc()
-#c32<-c()
-#c16<-c(123.666,129.619)
-#c8<-c()
-#c4<-c()
-#c2<-c()
-#c1<-c()
-#duration <- list("c1"=c1, "c2"=c2, "c4"=c4, "c8"=c8, "c16"=c16,"c32"=c32)
 
@@
 
-blah()
 SAMPLESIZES=c(2,3,6,10,16,22,30)
@@
     dm2 <- runlimma(mval=mval2,design=design2,myann=myann)
-    dmagg2 <- agg(dm2,cores=2)
-    m2 <- dmagg2$gme_res_df[,"mean",drop=FALSE]
+    dd <- merge(dm2,gt,by.x=0,by.y="probe")
+    m2 <- aggregate(t ~ gene,dd,mean)
+    rownames(m2) <- m2$gene
+    m2$gene=NULL
     lamres2 <- runmitch(m=m2,genesets=gs_symbols,cores=2)

### 2023-12-27 figs/GSE158422_simulate020.Rmd
roll back gene name fix

@@
 #new.hgnc.table <- getCurrentHumanMap()
-new.hgnc.table <- readRDS("new.hgnc.table.rds")
-fix <- checkGeneSymbols(gt$gene,map=new.hgnc.table)
-fix2 <- fix[which(fix$x != fix$Suggested.Symbol),]
-length(unique(fix2$x))
-gt$gene <- fix$Suggested.Symbol
+#new.hgnc.table <- readRDS("new.hgnc.table.rds")
+#fix <- checkGeneSymbols(gt$gene,map=new.hgnc.table)
+#fix2 <- fix[which(fix$x != fix$Suggested.Symbol),]
+#length(unique(fix2$x))
+#gt$gene <- fix$Suggested.Symbol
 
@@
 num_dm_sets=50
-sims=50
+sims=20
 

### 2023-12-31 figs/GSE158422_simulate_summary.Rmd
lrm

@@
 
-5. ALT: alcres
+5. LRM: lrmres
 
-6. ALW: nalcres
+6. ALT: alcres
 
-7. ALM: almres
+7. ALW: nalcres
+
+8. ALM: almres
 
@@
 l1 <- list("GSA"=gres2,"LAT"=lacres2,"LTT"=lactopres2,"LAW"=nlacres2,
-  "LAM"=lacmres2,"ALT"=alcres2,"ALW"=nalcres2,"ALM"=almres2)
+  "LAM"=lacmres2, "LRM"=lrmres2, "ALT"=alcres2,"ALW"=nalcres2,"ALM"=almres2)
 
@@
 l2 <- list("GSA"=gres2,"LAT"=lacres2,"LTT"=lactopres2,"LAW"=nlacres2,
-  "LAM"=lacmres2,"ALT"=alcres2,"ALW"=nalcres2,"ALM"=almres2)
+  "LAM"=lacmres2, "LRM"=lrmres2, "ALT"=alcres2,"ALW"=nalcres2,"ALM"=almres2)
 
@@
 l3 <- list("GSA"=gres2,"LAT"=lacres2,"LTT"=lactopres2,"LAW"=nlacres2,
-  "LAM"=lacmres2,"ALT"=alcres2,"ALW"=nalcres2,"ALM"=almres2)
+  "LAM"=lacmres2, "LRM"=lrmres2, "ALT"=alcres2,"ALW"=nalcres2,"ALM"=almres2)
 

### 2023-12-31 figs/GSE158422_sensitivity.Rmd
updates

@@
 
-## LRM - limma rank mean mitch
-
-```{r,lmrm1}
-
-dm$rank <-  rank(dm$t) - nrow(subset(dm,t<0))
-mm <- merge(dm,gt,by.x=0,by.y="probe")
-head(mm)
-mma <-aggregate(mm$rank ~ gene,mm,mean)
-rownames(mma) <- mma$gene
-mma$gene = NULL
-colnames(mma) <- "meanrank"
-lrmres <- runmitch(m=mma,genesets=gs_symbols,cores=16)
-
-lrmsig <- subset(lrmres,p.adjustANOVA<0.05)
-
-top <- head(lrmsig,20)
-top <- top[order(top$s.dist),]
-
-top %>%
-  kbl(caption="Top gene sets (biggest effect size with FDR<0.05)") %>%
-  kable_paper("hover", full_width = F)
-
-barnames <- gsub("_"," ",gsub("REACTOME_","",rownames(top)))
-cols <- gsub("TRUE","red",gsub("FALSE","blue",as.character(top$s.dist>0)))
-
-par(mar = c(5.1, 29.1, 4.1, 2.1))
-barplot(abs(top$s.dist),horiz=TRUE,las=1,names.arg=barnames, cex.names=0.65,
-  cex.axis=0.8,col=cols, xlab="Enrichment score",main="Reactomes")
-grid()
-
-par( mar = c(5.1, 4.1, 4.1, 2.1) )
-
-hist(lrmsig$s.dist)
-
-# distinguish up (TRUE) and down (FALSE)
-lrmsigsets <- paste(rownames(lrmsig),as.character(lrmsig$s.dist > 0))
-lrmsigsets2 <- gsub(" TRUE"," UP",lrmsigsets)
-lrmsigsets2 <- gsub(" FALSE"," DN",lrmsigsets2)
-
-```
-
 ## Run LAM
@@
 
-top <- head(lamsig,20)
-top <- top[order(top$s.dist),]
+topup <- subset(lamres,p.adjustANOVA<0.05 & s.dist >0 )
+topup <- tail(topup[order(topup$s.dist),],10)
 
-top %>%
-  kbl(caption="Top gene sets (biggest effect size with FDR<0.05)") %>%
+topdn <- subset(lamres,p.adjustANOVA<0.05 & s.dist <0 )
+topdn <- head(topdn[order(topdn$s.dist),],10)
+
+topup %>%
+  kbl(caption="Top gene sets hypermethylated (biggest effect size with FDR<0.05)") %>%
   kable_paper("hover", full_width = F)
+
+topdn %>%
+  kbl(caption="Top gene sets hypomethylated (biggest effect size with FDR<0.05)") %>%
+  kable_paper("hover", full_width = F)
+
+top <- rbind(topdn,topup)
 
@@
 dots <- lapply(ssres,function(x) { x[,"TP"] })
+lapply(dots,summary)
 boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. true positives",xlab="sample size",cex=0,main="LAM") ; grid()
@@
 dots <- lapply(ssres,function(x) { x[,"FP"] })
+lapply(dots,summary)
 boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. false positives",xlab="sample size",cex=0) ; grid()
@@
 ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
+lapply(ratio,summary)
 boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion FP",xlab="sample size",cex=0) ; grid()
@@
 names(nsigs) <- sizes
-nsigs
+str(nsigs)
+lapply(nsigs,length)
 
@@
 names(ngsig) <- FCLIM
-ngsig
+str(ngsig)
+lapply(ngsig,length)
 
 par( mar = c(5.1, 6.1, 4.1, 2.1) )
-barplot(unlist(lapply(ngsig,length)),horiz=TRUE,las=1,xlab="no. significant sets",xlim=c(0,60)) ; grid()
+barplot(unlist(lapply(ngsig,length)),horiz=TRUE,las=1,xlab="no. significant sets") ; grid()
 mtext("fold change threshold", 2, 4)
@@
 
-Only 4 hyper and 1 hypomethylated gene set were identified.
-This is surprising as there were 1299 hypermethylated probes corresponding to 808 genes, and
-1016 hypomethylated probes corresponding to 657 genes.
+107 significant gene sets were identified.
 
@@
 gsiges <- data.frame(gsig,myes)
-gsiges <- head(gsiges[order(-gsiges$ES),],20)
-gsiges <- gsiges[order(gsiges$ES),]
+
+gsigesdn <- gsiges[grep(" DN",rownames(gsiges)),]
+gsigesdn <- tail(gsigesdn[order(gsigesdn$ES),],10)
+gsigesdn <- gsigesdn[nrow(gsigesdn):1,]
+gsigesup <- gsiges[grep(" UP",rownames(gsiges)),]
+gsigesup <- tail(gsigesup[order(gsigesup$ES),],10)
+
+gsiges <- rbind(gsigesdn,gsigesup)
 
@@
 dots <- lapply(gsres,function(x) { x[,"TP"] })
-boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. true positives",xlab="sample size",cex=0,main="GSAmeth")
+lapply(dots,summary)
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. true positives",xlab="sample size",
+  cex=0,main="GSAmeth",ylim=c(0,nrow(gsig)))
 beeswarm(dots,add=TRUE,pch=1,cex=0.6)
@@
 dots <- lapply(gsres,function(x) { x[,"FP"] })
+lapply(dots,summary)
 boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. false positives",xlab="sample size",cex=0)
@@
 dots <- lapply(ssres,function(x) { x[,"TP"] })
-boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",cex=0,main="LAM") ; grid()
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",
+  cex=0,main="LAM", ylim=c(0,nrow(lamsig))) ; grid()
 beeswarm(dots,add=TRUE,pch=1,cex=0.6)
@@
 dots <- lapply(gsres,function(x) { x[,"TP"] })
-boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",cex=0,main="GSA") ; grid()
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",
+  cex=0,main="GSA", ylim=c(0,nrow(gsig))) ; grid()
 beeswarm(dots,add=TRUE,pch=1,cex=0.6)
@@
 Use Euler diagrams to find the similarity.
-As medians are normally distributed about the zero, one would expect equal numbers of up and down-
-methylated gene sets.
-Therefore the t-test based approach could be considered better.
+First, distinguish based on direction.
+
+```{r,ol0}
+
+lamsigsets2up <- lamsigsets2[grep(" UP", lamsigsets2)]
+lamsigsets2dn <- lamsigsets2[grep(" DN", lamsigsets2)]
+
+gsigsetsup <- gsigsets[grep(" UP", gsigsets)]
+gsigsetsdn <- gsigsets[grep(" DN", gsigsets)]
+
+v1 <- list("LAM up"=lamsigsets2up,"LAM dn"=lamsigsets2dn,
+  "GSA up"=gsigsetsup, "GSA dn"=gsigsetsdn)
+
+plot(euler(v1),quantities = TRUE)
+
+message("common")
+intersect(lamsigsets2up, gsigsetsup)
+intersect(lamsigsets2dn, gsigsetsdn)
+
+message("specific to LAM")
+LAMspec <- setdiff(lamsigsets2up, gsigsetsup)
+LAMspec
+LAMspec <- setdiff(lamsigsets2dn, gsigsetsdn)
+LAMspec
+
+message("specific to GSA")
+GSAspec <- setdiff(gsigsetsup, lamsigsets2up)
+GSAspec
+GSAspec <- setdiff(gsigsetsdn, lamsigsets2dn)
+GSAspec
+
+```
+
+Euler without direction noted.
 

### 2024-01-01 figs/GSE158422_simulate_summary.Rmd
fig1

@@
 
+pdf("fig1a.pdf",width=5,height=4)
+barplot(t(x1),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, cex.names=0.7, cex.axis=0.7, main="20 genes per set")
+grid()
+legend("topleft", inset=.1, c("Precision","Recall","F1"), fill=cols, cex=0.7)
+dev.off()
+
 load("GSE158422_simulate050.Rdata")
@@
 
+pdf("fig1b.pdf",width=5,height=4)
+barplot(t(x2),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, cex.names=0.7, cex.axis=0.7, main="50 genes per set")
+grid()
+dev.off()
+
 load("GSE158422_simulate100.Rdata")
@@
 grid()
+
+pdf("fig1c.pdf",width=5,height=4)
+barplot(t(x3),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, cex.names=0.7, cex.axis=0.7, main="100 genes per set")
+grid()
+legend("topleft", inset=.1, c("Precision","Recall","F1"), fill=cols, cex=0.7)
+dev.off()
 

### 2024-01-02 manuscript/manuscript.Rmd
ms work

@@
 annotation.
-As these annotations are several years old, we updated them using HGNChelper (database current as of 21-Dec-2023) [@Oh2022-ts].
-Circa 12% of gene symbols were updated.
+As gene annotations are several years old, we used HGNChelper to update them (database current as of 21-Dec-2023) [@Oh2022-ts]
+(3253 gene symbols were updated).
 Euler diagrams were created with the `eulerr` R package.
@@
 Next, GSA and LAM methods were applied to compare paired normal and cancer methylation data from 37 lung squamous cell carcinoma patients.
-From the 839,473 probes examined, 473,572 probes were considered significant at 5% FDR.
+From the 839,473 probes passing quality control, 473,572 were considered significant at 5% FDR.
 Of these, 119,785 probes were significantly hypermethylated and 353,787 were significantly hypomethylated.
@@
 These numbers are large, considering that there are 26,219 genes represented on this array.
+LAM resulted in 406 significant Reactome pathways at FDR<0.05.
+Of these 370 involved higher methylation, while 36 involved lower methylation.
+LAM execution took 51 seconds using eight parallel threads.
+With GSA, there were 75 pathways with higher methylation and 32 with lower methylation.
+12 pathways were significant in both directions.
+There were 30 pathways that were common between LAM and GSA methods.
+This includes 26 and 4 with higher and lower methylation respectively (**Figure 3A**).
 
-LAM resulted in 358 significant Reactome pathways at FDR<0.05 while there were just 21 with GSA (**Figure 3A**).
-The overlap of significant pathways was unexpectedly small, with just five common pathways.
-
-We were curious about whether LAM method had a bias towards larger gene sets as the recall of this
-method is higher with larger gene sets.
-Unexpectedly, pathways identified by LAM were collectively smaller than GSA (**Figure 3B**).
+As the simulation results indicated LAM had better recall with larger gene sets, we were curious about whether there were differences in the sizes of
+gene sets found by LAM and GSA.
+GSA significant pathways had a median size of 135, while for LAM the median was 54.5, indicating pathways identified as significant by LAM were collectively smaller than GSA (**Figure 3B**).
 Although algorithmic accuracy cannot be inferred from the types of pathways identified in real data,
@@
 termed "inconsistent".
-LAM was able to detect 300 out of 358 consistent pathways with a sample size of 10,
-while GSA detected just 4 of 21 at this sample size.
+At a sample size of 10, LAM was able to detect 354 out of 406 (87%) consistent pathways with a sample size of 10,
+while GSA detected just 13 of 107 at this sample size (12%).
 LAM identified more inconsistent pathways than GSA, but the proportion of
 inconsistent findings was lower across all sample sizes.
-At a sample size of 10, the proportion of inconsistent pathways was 10% for LAM and
-60% for GSA.
+At a sample size of 10, the proportion of inconsistent pathways was 13% for LAM and 33% for GSA.
 This result suggests that LAM has superior sensitivity to detect differentially methylated
@@
 DNA methylation.
-There were 12,386 differentially expressed genes (FDR<0.05), with 7,418 and 4,968 up- and down-regulated in the tumour group respectively.
+From 18,704 detected genes, There were 12,380 differentially expressed genes (FDR<0.05), with 7,415 and 4,965 up- and down-regulated in the tumour group respectively.
 At the pathway level there were 318 and 193 up- and down-regulated Reactome pathways with altered gene expression (FDR<0.05).

### 2024-01-02 manuscript/manuscript.Rmd
ms work

@@
 From 18,704 detected genes, There were 12,380 differentially expressed genes (FDR<0.05), with 7,415 and 4,965 up- and down-regulated in the tumour group respectively.
-At the pathway level there were 318 and 193 up- and down-regulated Reactome pathways with altered gene expression (FDR<0.05).
+At the pathway level there were 304 and 135 up- and down-regulated Reactome pathways with altered gene expression (FDR<0.05).
 As the context of gene methylation is important in influencing gene expression, promoter and gene body methylation
 were considered separately.
-In promoters, GSA identified 15 and 8 pathways with higher and lower methylation respectively, while at gene bodies, there were were 17 and 67
+In promoters, GSA identified 15 and 7 pathways with higher and lower methylation respectively, while at gene bodies, there were were 18 and 75
 pathways with higher and lower methylation respectively (FDR<0.05).
-Using LAM, there were 133 and 12 pathways identified with higher and lower promoter methylation respectively, while there were 236 and 35 with higher and lower gene body methylation respectively (FDR<0.05).
-The relative overlap between promoter and gene body results was larger for LAM pathway sets as compared to GSA (**Figure 5A,B**).
+Using LAM at promoters, there were 190 and 20 pathways identified with higher and lower methylation respectively.
+LAM at gene bodies identified 251 and 31 pathways with higher and lower methylation respectively (FDR<0.05).
+The relative overlap between promoter and gene body results was larger for LAM pathway sets (22%) as compared to GSA (3%) (**Figure 5A,B**).
 There was no observed overlap between GSA methylation-based pathways and gene expression pathways (**Figure 5A**), but there were some
 overlaps between pathways identified with LAM based promoter and gene body methylation and gene expression (**Figure 5B**).
-Interestingly, 123/318 upregulated gene expression pathways were associated with increasing gene body methylation with LAM.
-Likewise 19 pathways with downregulated expression were associated with increasing promoter methylation.
+
+Interestingly, 136/304 upregulated gene expression pathways were associated with increasing gene body methylation with LAM.
+Likewise 21 pathways with downregulated expression were associated with increasing promoter methylation.
 Some of these pathways with large effect sizes across the three contrasts are depicted in heatmap form (**Figure 5C**).
@@
 However there were a small number of pathways that did exhibit what could be epigenetic influences, for example
-"DNA strand elongation", "Unwinding of DNA", "Mucopolysaccharidoses" and "Interaction with cumulus cells and the zona pellucida"
-demonstrated an inverse relationship between promoter methylation and gene expression.
+"Unwinding of DNA", "Mucopolysaccharidoses","Interaction with cumulus cells and the zona pellucida", "glucocorticoid biosynthesis" and
+"olfactory signaling" demonstrated an inverse relationship between promoter methylation and gene expression.
 These results demonstrate that LAM can illuminate the complicated relationship between DNA methylation and gene expression at the pathway level in cancer.

### 2024-01-02 figs/GSE158422_sensitivity.Rmd
figs

@@
 
+pdf("fig3c.pdf",height=6,width=7)
+par(mar = c(5.1, 29.1, 4.1, 2.1))
+barplot(abs(top$s.dist),horiz=TRUE,las=1,names.arg=barnames, cex.names=0.8,
+  cex.axis=0.8,col=cols, xlab="Enrichment score",main="Reactomes")
+grid()
+dev.off()
+
 par( mar = c(5.1, 4.1, 4.1, 2.1) )
@@
 
+pdf("fig3d.pdf",height=6,width=7)
+par(mar = c(5.1, 29.1, 4.1, 2.1))
+barplot(abs(gsiges$ES)-1,horiz=TRUE,las=1,names.arg=barnames, cex.names=0.8,
+  cex.axis=0.85, col=cols, xlab="Enrichment score",main="GSA")
+grid()
+dev.off()
+
 ```
@@
 
+pdf("fig3a.pdf",height=4,width=5)
+plot(euler(v1),quantities = TRUE)
+dev.off()
+
 message("common")
@@
 
+pdf("fig3b.pdf",height=4,width=5)
+boxplot(mylist,cex=0, ylab="gene set size",names=c(GSANAME,LAMNAME))
+beeswarm(mylist,pch=19,add=TRUE,cex=0.3)
+dev.off()
+
 ```

### 2024-01-02 figs/GSE158422_sensitivity.Rmd
fig outputs

@@
 lapply(dots,summary)
-boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. true positives",xlab="sample size",cex=0,main="LAM") ; grid()
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",cex=0,main="LAM") ; grid()
 beeswarm(dots,add=TRUE,pch=1,cex=0.6)
@@
 lapply(dots,summary)
-boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. false positives",xlab="sample size",cex=0) ; grid()
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. inconsistent",xlab="sample size",cex=0) ; grid()
 beeswarm(dots,add=TRUE,pch=1,cex=0.6)
@@
 lapply(ratio,summary)
-boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion FP",xlab="sample size",cex=0) ; grid()
+boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion inconsistent",xlab="sample size",cex=0) ; grid()
 beeswarm(ratio,add=TRUE,pch=1,cex=0.6)
 
+pdf("fig4a.pdf",height=7,width=5)
+par(mfrow=c(3,1))
+par( mar = c(4.1, 4.1, 2.1, 1.1) )
+dots <- lapply(ssres,function(x) { x[,"TP"] })
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",cex=0,main="LAM") ; grid()
+beeswarm(dots,add=TRUE,pch=1,cex=0.6)
+abline(h=nrow(lamsig))
+
+dots <- lapply(ssres,function(x) { x[,"FP"] })
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. inconsistent",xlab="sample size",cex=0) ; grid()
+beeswarm(dots,add=TRUE,pch=1,cex=0.6)
+
+dots <- lapply(ssres,function(x) { x[,"TP"] })
+TPdots <- do.call(rbind,dots)
+dots <- lapply(ssres,function(x) { x[,"FP"] })
+FPdots <- do.call(rbind,dots)
+ratio <- t(FPdots / (TPdots + FPdots))
+ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
+boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion inconsistent",xlab="sample size",cex=0) ; grid()
+beeswarm(ratio,add=TRUE,pch=1,cex=0.6)
+
+dev.off()
+
 par(mfrow=c(1,1))
+par( mar = c(5.1, 4.1, 4.1, 2.1) )
 
@@
 lapply(dots,summary)
-boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. true positives",xlab="sample size",
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",
   cex=0,main="GSAmeth",ylim=c(0,nrow(gsig)))
@@
 lapply(dots,summary)
-boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. false positives",xlab="sample size",cex=0)
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. inconsistent",xlab="sample size",cex=0)
 beeswarm(dots,add=TRUE,pch=1,cex=0.6)
@@
 lapply(ratio,summary)
-boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion FP",xlab="sample size",cex=0) ; grid()
+boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion inconsistent",xlab="sample size",cex=0) ; grid()
 beeswarm(ratio,add=TRUE,pch=1,cex=0.6)
 
+pdf("fig4b.pdf",height=7,width=5)
+par(mfrow=c(3,1))
+par( mar = c(4.1, 4.1, 2.1, 1.1) )
+
+dots <- lapply(gsres,function(x) { x[,"TP"] })
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",
+  cex=0,main="GSAmeth",ylim=c(0,nrow(gsig)))
+beeswarm(dots,add=TRUE,pch=1,cex=0.6)
+abline(h=nrow(gsig))
+
+dots <- lapply(gsres,function(x) { x[,"FP"] })
+boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. inconsistent",xlab="sample size",cex=0)
+beeswarm(dots,add=TRUE,pch=1,cex=0.6)
+
+dots <- lapply(gsres,function(x) { x[,"TP"] })
+TPdots <- do.call(rbind,dots)
+dots <- lapply(gsres,function(x) { x[,"FP"] })
+FPdots <- do.call(rbind,dots)
+ratio <- t(FPdots / (TPdots + FPdots))
+ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
+boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion inconsistent",xlab="sample size",cex=0) ; grid()
+beeswarm(ratio,add=TRUE,pch=1,cex=0.6)
+
+dev.off()
+
 par(mfrow=c(1,1))
+par( mar = c(5.1, 4.1, 4.1, 2.1) )
 

### 2024-01-02 figs/GSE158422_rna_int.Rmd
download counts
fix
save RNA integration results

@@
 ```{r,deseq1}
+
+if (!file.exists("GSE158420_counts.txt.gz")) {
+  download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE158420&format=file&file=GSE158420%5Fcounts%2Etxt%2Egz",
+    destfile="GSE158420_counts.txt.gz")
+}
 
@@
 pdf("fig5c.pdf")
-heatmap.2( as.matrix(mtop) , trace="none",margins = c(6,25), col=colfunc(25), cexRow=.6, cexCol=1)
+heatmap.2( as.matrix(mtop) , trace="none",margins = c(6,25), col=colfunc(25), cexRow=.7, cexCol=1)
 dev.off()
@@
 
+save.image("GSE158422_rna_int.Rdata")
+
 sessionInfo()

### 2024-01-03 figs/ART_novakovic.Rmd
art simplification
art fix

@@
 targets <- sample_metadata
-
-files <- list.files(WORKING_DIR,pattern = "GSM",recursive = TRUE)
-mybase <- unique(gsub("_Red.idat.gz" ,"", gsub("_Grn.idat.gz", "" ,files)))
-mybase <- paste(WORKING_DIR,"/",mybase,sep="")
-targets$Basename <- mybase
+targets <- subset(targets,`time of collection:ch1`=="birth")
+targets$Basename <- head( gsub("_Grn.idat.gz","",list.files("IDAT",full.names=TRUE,pattern="Grn")) , 207 )
 head(targets)
 
-# filter time of birth as were not interested in follow up samples
-targets <- subset(targets,`time of collection:ch1`=="birth")
 rgSet <- read.metharray.exp(targets = targets)
@@
 
-write.table(top_GIFT_vs_FX$dma,file="novakovic_top_GIFT_vs_FX.tsv",sep="\t",quote=FALSE)
-
 saveRDS(top_GIFT_vs_FX,file="novakovic_GIFT_vs_FX.rds")
@@
 head(top_nat_vs_ART$confects$table$name)
-
-write.table(top_nat_vs_ART$dma,file="novakovic_top_nat_vs_ART.tsv",sep="\t",quote=FALSE)
 

### 2024-01-03 figs/ART_novakovic.Rmd
fix path for idat

@@
 #targets$Basename <- gsub("IDAT/","",targets$Basename)  
-birth<-targets[which(targets$`time_of_collection`=="birth"),]
-birth$Basename <- sapply(strsplit(birth$Basename,"/"),"[[",3)
+birth <- targets[which(targets$`time_of_collection`=="birth"),]
+birth$Basename <- sapply(strsplit(birth$Basename,"/"),"[[",2)
 birth$Basename <- sapply(strsplit(birth$Basename,"_"),"[[",1)

### 2024-01-03 figs/art_example.Rmd
dont use ftp

@@
 tic()
-new.hgnc.table <- getCurrentHumanMap()
+#new.hgnc.table <- getCurrentHumanMap()
 fix <- checkGeneSymbols(gt2$gene,map=new.hgnc.table)

### 2024-01-03 figs/art_example.Rmd
art example pathway

@@
 
-det <- mres$detailed_sets$REACTOME_NUCLEAR_EVENTS_KINASE_AND_TRANSCRIPTION_FACTOR_ACTIVATION
+det <- mres$detailed_sets$REACTOME_CELLULAR_RESPONSES_TO_STIMULI
 
@@
         abline(v = 0, h = 0, lty = 2, lwd = 2, col = "blue")
-        title(main = "Reactome nuclear kinase and transcription factor activation", xlab = X_AXIS,
+        title(main = "Cellular responses to stimuli", xlab = X_AXIS,
         ylab = Y_AXIS)

### 2024-01-04 figs/aging_example.Rmd
developing the results section.

@@
 
+#numsig manova
+nrow(subset(mtable,p.adjustMANOVA<0.05))
+sig <- subset(mtable,p.adjustMANOVA<0.05)
+table(sign(sig$s.Discovery) == sign(sig$s.Replication))
+
+# num sig discovery
+length(which(p.adjust(mtable$p.Discovery,method="fdr")<0.05))
+sig_disc <- mtable[which(p.adjust(mtable$p.Discovery,method="fdr")<0.05),c("set","s.Discovery")]
+sig_disc_up <- sig_disc[which(sig_disc$s.Discovery>0),"set"]
+sig_disc_dn <- sig_disc[which(sig_disc$s.Discovery<0),"set"]
+
+# num sig replication
+length(which(p.adjust(mtable$p.Replication,method="fdr")<0.05))
+sig_repl <- mtable[which(p.adjust(mtable$p.Replication,method="fdr")<0.05),c("set","s.Replication")]
+sig_repl_up <- sig_disc[which(sig_repl$s.Replication>0),"set"]
+sig_repl_dn <- sig_disc[which(sig_repl$s.Replication<0),"set"]
+
+l <- list("Discovery up"=sig_disc_up,"Discovery dn"=sig_disc_dn,
+  "Replication up"=sig_repl_up,"Replication dn"=sig_repl_dn)
+
+str(l)
+
+plot(euler(l),quantities = TRUE)
+
 ```

### 2024-01-04 figs/art_example.Rmd
developing the results section.

@@
 tic()
-mres <- mitch_calc(x=m,genesets=gs_symbols,minsetsize=5,cores=16, priority="significance")
+mres <- mitch_calc(x=m,genesets=gs_symbols,minsetsize=5,cores=16, priority="effect")
 toc() #11.45s
@@
 
-det <- mres$detailed_sets$REACTOME_CELLULAR_RESPONSES_TO_STIMULI
+det <- mres$detailed_sets$REACTOME_ADRENORECEPTORS
 
@@
         abline(v = 0, h = 0, lty = 2, lwd = 2, col = "blue")
-        title(main = "Cellular responses to stimuli", xlab = X_AXIS,
+        title(main = "Adrenoreceptors", xlab = X_AXIS,
         ylab = Y_AXIS)

### 2024-01-04 manuscript/manuscript.Rmd
developing the results section.

@@
 GSA significant pathways had a median size of 135, while for LAM the median was 54.5, indicating pathways identified as significant by LAM were collectively smaller than GSA (**Figure 3B**).
-Although algorithmic accuracy cannot be inferred from the types of pathways identified in real data,
+Although algorithmic accuracy cannot be inferred from the types of pathways identified in real data [@Timmons2015-kr],
 top ranked results from LAM appeared to be more related to cell differentiation, identity and development (**Figure 3C**),
@@
 group of 37 patients.
-This process was repeated 50 times and the results shown in (**Figure 4**).
+This process was repeated 50 times at each sample size and the results shown in (**Figure 4**).
 Pathways that were identified as statistically significant (FDR<0.05) in the smaller and full group were
@@
 LAM at gene bodies identified 251 and 31 pathways with higher and lower methylation respectively (FDR<0.05).
-The relative overlap between promoter and gene body results was larger for LAM pathway sets (22%) as compared to GSA (3%) (**Figure 5A,B**).
+The relative overlap between significant promoter and gene body pathways was larger for LAM pathway sets (22%) as compared to GSA (3%) (**Figure 5A,B**).
 There was no observed overlap between GSA methylation-based pathways and gene expression pathways (**Figure 5A**), but there were some
 overlaps between pathways identified with LAM based promoter and gene body methylation and gene expression (**Figure 5B**).
-
 Interestingly, 136/304 upregulated gene expression pathways were associated with increasing gene body methylation with LAM.
@@
 However there were a small number of pathways that did exhibit what could be epigenetic influences, for example
-"Unwinding of DNA", "Mucopolysaccharidoses","Interaction with cumulus cells and the zona pellucida", "glucocorticoid biosynthesis" and
+"unwinding of DNA", "mucopolysaccharidoses","interaction with cumulus cells and the zona pellucida", "glucocorticoid biosynthesis" and
 "olfactory signaling" demonstrated an inverse relationship between promoter methylation and gene expression.
@@
 Using summary statistics from independent discovery and replication studies using the EPIC array platform described
-by McCartney et al (2020), we applied mitch to explore the pathway level differential methylation.
-In the discovery cohort we observed XX differential pathways (FDR), while in the replication group we observed XX.
-There was a high degree of agreement, with XX pathways significant in both groups.
+by McCartney et al (2020) [@McCartney2020-ob] irrespective of sex, we applied the LAM method to explore the pathway level differential methylation.
+In the discovery cohort we observed 304 differential pathways (FDR<0.05), while in the replication group we observed 107.
+There was a high degree of agreement, with 43 pathways significant in both groups.
 This is not surprising, as the gene-level differential methylation scores show a strong positive association between discovery and replication studies (**Figure 6A**).
-Using the MANOVA test, we identified XX pathways with altered methylation, most of which were concordant between
-discovery and replication (**Figure 6B**).
+Using the MANOVA test, we identified 390 pathways with altered methylation.
+Visualised as a scatterplot, the overall pattern is concordant, although a cluster of points in the lower right of the chart indicates a cluster of 268 pathways with higher methylation in the discovery sample that were lower in methylation in the replication sample (**Figure 6B**).
 When prioritising the results by S distance, a surrogate measure of effect size, 29 of the top 30 pathways were concordant in their direction of regulation (**Figure 6C**).
@@
 array systems.
-Epigenetic differences between infants conceived naturally and with assisted reproductive technologies have been the subject of studies since 20XX (eg: REFS).
-Two of the highest powered studies are Estill 2016 which used the 450K array and Novakovic 2019 study which used
-the EPIC array.
+Epigenetic differences between infants conceived naturally and with assisted reproductive technologies have been the subject of studies since 2009 [@Katari2009-ya].
+Two of the highest powered studies with publicly available data are Estill 2016 which used the 450K array and Novakovic 2019 study which used
+the EPIC array [@Estill2016-jh;@Novakovic2019-jt].
 Joint enrichment analysis can uncover pathways that are common or different between these studies and illuminate
@@
 The Estill dataset on 450k had 418,833 probes that met the quality criteria (Natural n=43; IVF n=38).
-Of these, 149,552 probes showed statistical significant differential methylation (FDR<0.05), of which 28,637 and 120,885 exhibited higher or lower methylation in the IVF group respectively.
-The Novakovic dataset on EPIC array had 810,518 probes that met the filtering criteria (n=133).
-Of these, 5,491 were statistically significant (FDR<0.05) with 1,326 and 4,165 exhibited higher or lower methylation in the IVF group respectively.
-After summarisation to the gene level, the Estill dataset on 450k array describes 19,951 genes, while the
-Novakovic dataset on EPIC array describes 26,150 genes.
-There were 19,942 genes common to both platforms.
+Of these, 149,589 probes showed statistical significant differential methylation (FDR<0.05), of which 28,767 and 120,822 exhibited higher or lower methylation in the IVF group respectively.
+The Novakovic dataset on EPIC array had 793,844 probes that met the filtering criteria (n=133).
+Of these, 5,562 were statistically significant (FDR<0.05) with 1,348 and 4,214 exhibited higher or lower methylation in the IVF group respectively.
+After summarisation to the gene level, the Estill dataset on 450k array describes 19,240 genes, while the
+Novakovic dataset on EPIC array describes 22,588 genes.
+There were 19,234 genes common to both platforms.
 The rank-rank plot indicates a high degree of overlap overall, and interestingly shows a trend of more genes having lower methylation level in the IVF group (**Figure 7A**).
-A scatterplot of pathway enrichment scores indicates a high degree of agreement between these studies at the level of pathways despite they being conducted years apart and analysed with different array systems (**Figure 7B**).
-The heatmap shows a high degree of agreement with pathways with a higher level of methylation in the IVF groups,
+A scatterplot of pathway enrichment scores indicates a moderate degree of agreement between these studies at the level of pathway enrichment scores (*r*=0.35,*p*=2.2e-16) despite these studies being conducted on independent cohorts years apart and analysed with different array systems (**Figure 7B**).
+A heatmap shows a high degree of agreement of pathways with higher methylation in the IVF groups,
 while pathways with overall lower methylation in the IVF group showed some variability in enrichment scores
 between studies (**Figure 7C**).
-We were interested in which pathways had consistently lower methylation in the IVF group and identified the Reactome pathway "Nuclear kinase and transcription factor activation events" which had enrichment scores of -0.26 in both contrasts and a MANOVA FDR value of 0.002 (**Figure 7D**).
+We were interested in which pathways had consistently lower methylation in the IVF group and identified the Reactome pathway "Adrenoreceptors" which had enrichment scores of -0.42 in the Estill study and -0.56 in the Novakovic study, recording a MANOVA FDR value of 0.005 (**Figure 7D**).
 Top ranked genes from this pathway include *FOSL1*, *JUNB*, *ASCL1* and *ID1*.

### 2024-01-04 manuscript/manuscript.Rmd
adrenoreceptors

@@
 between studies (**Figure 7C**).
-We were interested in which pathways had consistently lower methylation in the IVF group and identified the Reactome pathway "Adrenoreceptors" which had enrichment scores of -0.42 in the Estill study and -0.56 in the Novakovic study, recording a MANOVA FDR value of 0.005 (**Figure 7D**).
-Top ranked genes from this pathway include *FOSL1*, *JUNB*, *ASCL1* and *ID1*.
-A loss of methylation of these genes could contribute to disorders of development.
+We were interested in which pathways had consistently lower methylation in the IVF group and identified the Reactome pathway "Adrenoreceptors" which had enrichment scores of -0.42 in the Estill study and -0.56 in the Novakovic study, recording a MANOVA FDR value of 0.04 (**Figure 7D**).
+Top ranked genes from this pathway include *ADRA2C*, *ADRB2*, *ADRA2B* and *ADRA1B*.
+A loss of methylation of these genes could contribute to reported elevated blood pressure in people conceived by IVF [@Meister2018-cp].
 Taken together, this analysis shows that LAM method can be used to compare enrichment of independent studies
@@
 (C) Heatmap of 30 pathways with most extreme S distances (a surrogate of effect size). Red indicates higher methylation and blue shows lower methylation.
-(D) An example of a pathway identified with this joint enrichment analysis method, "Nuclear kinase and transcription factor activation" shows lower methylation of member genes in both Estill and Novakovic studies.
+(D) An example of a pathway identified with this joint enrichment analysis method, "Adrenoreceptors" shows lower methylation of member genes in both Estill and Novakovic studies.
 ](../figs/fig7.png)

### 2024-01-06 figs/GSE158422_simulate020_rslurm.Rmd
putting simulations on slurm

---
title: "Testing gene methylation enrichment analysis approaches using simulated data"
author: "The GMEA team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
---

## Introduction

In this report we establish a new method for generating simulated data with known ground truth.
This will be used to test different gene methylation enrichment approaches systematically.

The general steps are:

1. Import GSE158422 data corresponding to control (non-tumour tissue).

2. From the 37 samples, create two groups of 18 samples.
One of these will be considered "control" and the other "case".

3. Create random gene sets that have similar sized to Reactome pathways.

3. Some gene sets will be selected to be differentially methylated.
Half of these will be hypermethylated and the others will be hypomethylated.

4. The changes will be incorporated into the "case" samples.

5. The enrichment analysis will be conducted.

6. The accuracy will be calculated.

```{r,packages}

suppressPackageStartupMessages({
  library("stringi")
  library("limma")
  library("missMethyl")
  library("IlluminaHumanMethylation450kmanifest")
  library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
  library("HGNChelper")
  library('org.Hs.eg.db')
  library("psych")
  library("mitch")
  library("kableExtra")
  library("rslurm")
})

source("GSE158422_simulate_functions.R")

# optimised for 128 GB sever with 32 threads
CORES=6

```

## Load data

* annotations

* probe sets

* gene sets

* design matrix

* mval matrix

```{r,annotation}

# get probe-gene mapping
anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
myann <- data.frame(anno[,c("UCSC_RefGene_Name","UCSC_RefGene_Group","Islands_Name","Relation_to_Island")])
gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
names(gp2) <- rownames(gp)
gp2 <- lapply(gp2,unique)
gt <- stack(gp2)
colnames(gt) <- c("gene","probe")
dim(gt)
str(gt)
head(gt)

#new.hgnc.table <- getCurrentHumanMap()
#new.hgnc.table <- readRDS("new.hgnc.table.rds")
#fix <- checkGeneSymbols(gt$gene,map=new.hgnc.table)
#fix2 <- fix[which(fix$x != fix$Suggested.Symbol),]
#length(unique(fix2$x))
#gt$gene <- fix$Suggested.Symbol

if (!file.exists("GSE158422_design.rds")) {
  download.file("https://ziemann-lab.net/public/gmea_prototype/GSE158422_design.rds", "GSE158422_design.rds")
}
design <- readRDS("GSE158422_design.rds")

if (!file.exists("GSE158422_design.rds")) {
  options(timeout=10000000000000)
  download.file("https://ziemann-lab.net/public/gmea_prototype/GSE158422_mx.rds","GSE158422_mx.rds")
}
mval <- readRDS("GSE158422_mx.rds")

```

# Gene set database

We could use Reactome pathways, however these have a lot of overlapping sets, which could cause inflated
false positives.
After some testing, I found that the gene set size strongly impacts the accuracy.

For reference, Reactome sets have a mean of 48 and median of 15, while MSigDB has a median of 47 and mean of 116.

Therefore, a reasonable analysis would include small (20), medium (50) and large (100) sets.

To reflect coverage of Reactome and other pathway databases, only half the genes will be included in sets.

```{r,gset_prep}

gene_catalog <- unique(gt$gene)

# set gene set size here
lengths <- rep(20,1000)

gsets <- randomGeneSets(gene_catalog,lengths,seed=100)

```

# Incorporate changes

TODO: to incorporate changes to case samples.

Need to figure out what magnitude to change.
Will refer to the cancer/normal comparison.

Select genes and probes to alter.

```{r,incorp}

seed=100
frac_genes=0.5
frac_probes=0.5
delta=1
nsamples=10
normal_mval <- mval[,(1:(ncol(mval)/2)*2)]

```

# Run analyses

Set assumptions.

```{r,set}

num_dm_sets=50
sims=100
groupsizes=c(3,5,8,12)
deltas=c(0.1,0.2,0.3,0.4,0.5)

#groupsizes=5
#deltas=0.4

params <- expand.grid("groupsizes"=groupsizes,"deltas"=deltas,"seed"=seq(100,sims*100,100))

colnames(params) <- c("groupsize","delta","seed")

params %>% kbl(caption="Parameters to test") %>% kable_paper("hover", full_width = F)

```

## GSA

Using GSAmeth of significant probes.

```{r,rungsa,eval=FALSE}

sopt1 <- list(time = '1:00:00', mem="10G")

sjob <- slurm_apply(simgsa, params, jobname = 'gsa',
  nodes = 50, cpus_per_node = 2, submit = TRUE, slurm_options = sopt1,
  genesetdatabase=gsets, myann=myann, mval=normal_mval, frac_genes=0.5,
  frac_probes=0.5, num_dm_sets=num_dm_sets,
  pkgs = rev(.packages()),
  global_objects = c("randomGeneSets", "runlimma","incorp_dm", "F1", "gsets") )

gres <- get_slurm_out(sjob, outtype = 'raw', wait = TRUE)

gres <- do.call(rbind,gres)

gres <- data.frame(params,gres)

str(gres)

```

## LAT

Limma Average t-test.

```{r,runlat,eval=FALSE}

sopt1 <- list(time = '1:00:00', mem="16G")

sjob <- slurm_apply(simlac, params, jobname = 'lat',
  nodes = 50, cpus_per_node = 2, submit = TRUE, slurm_options = sopt1,
  genesetdatabase=gsets, myann=myann, mval=normal_mval, frac_genes=0.5,
  frac_probes=0.5, num_dm_sets=num_dm_sets,
  pkgs = rev(.packages()),
  global_objects = c("randomGeneSets", "runlimma","incorp_dm", "ttenrich","F1", "gsets","gt") )

latres <- get_slurm_out(sjob, outtype = 'raw', wait = TRUE)

latres <- do.call(rbind,latres)

latres <- data.frame(params,latres)

str(latres)

```

## LTT

Limma top t-test.

```{r,runltt,eval=TRUE}

sopt1 <- list(time = '1:00:00', mem="16G")

sjob <- slurm_apply(simlactop, params, jobname = 'ltt',
  nodes = 50, cpus_per_node = 2, submit = TRUE, slurm_options = sopt1,
  genesetdatabase=gsets, myann=myann, mval=normal_mval, frac_genes=0.5,
  frac_probes=0.5, num_dm_sets=num_dm_sets,
  pkgs = rev(.packages()),
  global_objects = c("randomGeneSets", "runlimma","incorp_dm", "ttenrich","F1", "gsets","gt") )

lttres <- get_slurm_out(sjob, outtype = 'raw', wait = TRUE)

lttres <- do.call(rbind,lttres)

lttres <- data.frame(params,lttres)

str(lttres)

```

## LAW

Limma average wilcox.

```{r,runlaw,eval=TRUE}

sopt1 <- list(time = '1:00:00', mem="16G")

sjob <- slurm_apply(simnlac, params, jobname = 'law',
  nodes = 50, cpus_per_node = 2, submit = TRUE, slurm_options = sopt1,
  genesetdatabase=gsets, myann=myann, mval=normal_mval, frac_genes=0.5,
  frac_probes=0.5, num_dm_sets=num_dm_sets,
  pkgs = rev(.packages()),
  global_objects = c("randomGeneSets", "runlimma","incorp_dm", "wtenrich","F1", "gsets","gt") )

lawres <- get_slurm_out(sjob, outtype = 'raw', wait = TRUE)

lawres <- do.call(rbind,lawres)

lawres <- data.frame(params,lawres)

str(lawres)

```

## LAM

Limma average mitch.

```{r,runlam,eval=TRUE}

sopt1 <- list(time = '1:00:00', mem="16G")

sjob <- slurm_apply(simlacm, params, jobname = 'lam',
  nodes = 50, cpus_per_node = 2, submit = TRUE, slurm_options = sopt1,
  genesetdatabase=gsets, myann=myann, mval=normal_mval, frac_genes=0.5,
  frac_probes=0.5, num_dm_sets=num_dm_sets,
  pkgs = rev(.packages()),
  global_objects = c("randomGeneSets", "runlimma","incorp_dm", "runmitch","F1", "gsets","gt") )

lamres <- get_slurm_out(sjob, outtype = 'raw', wait = TRUE)

lamres <- do.call(rbind,lamres)

lamres <- data.frame(params,lamres)

str(lamres)

```

## LRM

Limma rank mitch.

```{r,runlrm,eval=TRUE}

sopt1 <- list(time = '1:00:00', mem="16G")

sjob <- slurm_apply(simlrm, params, jobname = 'lrm',
  nodes = 50, cpus_per_node = 2, submit = TRUE, slurm_options = sopt1,
  genesetdatabase=gsets, myann=myann, mval=normal_mval, frac_genes=0.5,
  frac_probes=0.5, num_dm_sets=num_dm_sets,
  pkgs = rev(.packages()),
  global_objects = c("randomGeneSets", "runlimma","incorp_dm", "runmitch","F1", "gsets","gt") )

lrmres <- get_slurm_out(sjob, outtype = 'raw', wait = TRUE)

lrmres <- do.call(rbind,lrmres)

lrmres <- data.frame(params,lrmres)

str(lrmres)

```

## ALT

Aggregate limma t-test

```{r,runalt,eval=TRUE}

sopt1 <- list(time = '1:00:00', mem="16G")

sjob <- slurm_apply(simalc, params, jobname = 'alt',
  nodes = 50, cpus_per_node = 2, submit = TRUE, slurm_options = sopt1,
  genesetdatabase=gsets, myann=myann, mval=normal_mval, frac_genes=0.5,
  frac_probes=0.5, num_dm_sets=num_dm_sets,
  pkgs = rev(.packages()),
  global_objects = c("randomGeneSets", "runlimma","incorp_dm", "ttenrich","F1", "gsets","gt") )

altres <- get_slurm_out(sjob, outtype = 'raw', wait = TRUE)

altres <- do.call(rbind,altres)

altres <- data.frame(params,altres)

str(altres)

```

## ALW

Aggregate limma wilcox test

```{r,runalw,eval=TRUE}

sopt1 <- list(time = '1:00:00', mem="16G")

sjob <- slurm_apply(simnalc, params, jobname = 'alw',
  nodes = 50, cpus_per_node = 2, submit = TRUE, slurm_options = sopt1,
  genesetdatabase=gsets, myann=myann, mval=normal_mval, frac_genes=0.5,
  frac_probes=0.5, num_dm_sets=num_dm_sets,
  pkgs = rev(.packages()),
  global_objects = c("randomGeneSets", "runlimma","incorp_dm", "wtenrich","F1", "gsets","gt") )

alwres <- get_slurm_out(sjob, outtype = 'raw', wait = TRUE)

alwres <- do.call(rbind,alwres)

alwres <- data.frame(params,alwres)

str(alwres)

```

## ALM

Aggregate limma mitch

```{r,runalm,eval=TRUE}

sopt1 <- list(time = '1:00:00', mem="16G")

sjob <- slurm_apply(simalm, params, jobname = 'alm',
  nodes = 50, cpus_per_node = 2, submit = TRUE, slurm_options = sopt1,
  genesetdatabase=gsets, myann=myann, mval=normal_mval, frac_genes=0.5,
  frac_probes=0.5, num_dm_sets=num_dm_sets,
  pkgs = rev(.packages()),
  global_objects = c("randomGeneSets", "runlimma","incorp_dm", "runmitch","F1", "gsets","gt") )

almres <- get_slurm_out(sjob, outtype = 'raw', wait = TRUE)

almres <- do.call(rbind,almres)

almres <- data.frame(params,almres)

str(almres)

```

## Session information

```{r,save}

save.image("GSE158422_simulate020_rslurm.Rdata")

sessionInfo()

```

### 2024-01-06 figs/GSE158422_simulate_functions.R

@@ -1,0 +1,694 @@
+
+randomGeneSets <- function(gene_catalog, lengths, seed){
+  set.seed(seed) ; gene_catalog_half <- sample(gene_catalog,length(gene_catalog)/2)
+  num_gsets <- length(lengths)
+  set.seed(seed) ; seeds <- sample(1:1e6, num_gsets)
+  gsets <- lapply(1:num_gsets,function(i) {
+    set.seed(seeds[i]) ; gs <- sample(gene_catalog_half,lengths[i])
+    return(gs)
+  } )
+  names(gsets)<-stringi::stri_rand_strings(length(gsets), 15, pattern = "[A-Za-z]")
+  return(gsets)
+}
+
+incorp_dm <- function(genesets,myann,mval,seed, frac_genes,frac_probes,
+  groupsize,delta=1,gene_catalog) {
+
+  # divide gene sets between hyper and hypomethylated
+  nset <- floor(length(genesets)/2)
+  set.seed(seed) ; gtup <-sample(genesets,nset)
+  set.seed(seed) ; gtdn <- sample(setdiff(genesets,gtup),nset)
+  gup <- unname(unlist(gtup))
+  gdn <- unname(unlist(gtdn))
+  # add extra 10% DM genes
+  gnon <- setdiff(gene_catalog,c(gup,gdn))
+  gextra <- round(length(gnon)*0.1)
+  set.seed(seed) ; gup <- c(gup,sample(gnon,gextra))
+  gnon <- setdiff(gene_catalog,c(gup,gdn))
+  set.seed(seed) ; gdn <- c(gdn,sample(gnon,gextra))
+  # make probe-gene vector
+  probe2gene <- strsplit(myann$UCSC_RefGene_Name,";")
+  names(probe2gene) <- rownames(myann)
+  probe2gene <- unlist(probe2gene)
+  # select probes hypermethylated
+  set.seed(seed) ; gup2 <- sample(gup,floor(length(gup)*frac_genes))
+  pup <- names(probe2gene[which(probe2gene %in% gup2)])
+  set.seed(seed) ; pup2 <- sample(pup,floor(length(pup)*frac_probes))
+  # select probes hypomethylated
+  set.seed(seed) ; gdn2 <- sample(gdn,floor(length(gdn)*frac_genes))
[…]

### 2024-01-08 figs/GSE158422_simulate020.Rmd
offload functions to separate file

@@
 })
+
+source("GSE158422_simulate_functions.R")
 
@@
 
-randomGeneSets <- function(gene_catalog, lengths, seed){
-  set.seed(seed) ; gene_catalog_half <- sample(gene_catalog,length(gene_catalog)/2)
-  num_gsets <- length(lengths)
-  set.seed(seed) ; seeds <- sample(1:1e6, num_gsets)
-  gsets <- lapply(1:num_gsets,function(i) {
-    set.seed(seeds[i]) ; gs <- sample(gene_catalog_half,lengths[i])
-    return(gs)
-  } )
-  names(gsets)<-stri_rand_strings(length(gsets), 15, pattern = "[A-Za-z]")
-  return(gsets)
-}
-
 gsets <- randomGeneSets(gene_catalog,lengths,seed=100)
@@
 normal_mval <- mval[,(1:(ncol(mval)/2)*2)]
-
-incorp_dm <- function(genesets,myann,mval,seed, frac_genes,frac_probes,
-  groupsize,delta=1,gene_catalog) {
-
-  # divide gene sets between hyper and hypomethylated
-  nset <- floor(length(genesets)/2)
-  set.seed(seed) ; gtup <-sample(genesets,nset)
-  set.seed(seed) ; gtdn <- sample(setdiff(genesets,gtup),nset)
-  gup <- unname(unlist(gtup))
-  gdn <- unname(unlist(gtdn))
-  # add extra 10% DM genes
-  gnon <- setdiff(gene_catalog,c(gup,gdn))
-  gextra <- round(length(gnon)*0.1)
-  set.seed(seed) ; gup <- c(gup,sample(gnon,gextra))
-  gnon <- setdiff(gene_catalog,c(gup,gdn))
-  set.seed(seed) ; gdn <- c(gdn,sample(gnon,gextra))
-  # make probe-gene vector
-  probe2gene <- strsplit(myann$UCSC_RefGene_Name,";")
-  names(probe2gene) <- rownames(myann)
-  probe2gene <- unlist(probe2gene)
-  # select probes hypermethylated
-  set.seed(seed) ; gup2 <- sample(gup,floor(length(gup)*frac_genes))
-  pup <- names(probe2gene[which(probe2gene %in% gup2)])
-  set.seed(seed) ; pup2 <- sample(pup,floor(length(pup)*frac_probes))
-  # select probes hypomethylated
-  set.seed(seed) ; gdn2 <- sample(gdn,floor(length(gdn)*frac_genes))
-  pdn <- names(probe2gene[which(probe2gene %in% gdn2)])
-  set.seed(seed) ; pdn2 <- sample(pdn,floor(length(pdn)*frac_probes))
-  # add 10% DM probes as well
-  probes <- rownames(myann)
-  pnon <- setdiff(probes,c(pup,pdn))
-  pextra <- round(length(pnon)*0.1)
-  set.seed(seed) ; pup <- c(pup,sample(pnon,pextra))
-  pnon <- setdiff(probes,c(pup,pdn))
-  set.seed(seed) ; pdn <- c(pdn,sample(pnon,pextra))
-  # divide samples between ctrl and case
-  ncols <- ncol(mval)
-  maxgroupsize=floor(ncols/2)
-  if ( groupsize > maxgroupsize ) { stop("groupsize cannot be larger than half the ncols of mval") }
-  set.seed(seed) ; ctrl <- sample(1:ncols,groupsize)
-  set.seed(seed) ; case <- sample(setdiff(1:ncols,ctrl),groupsize)
-  mval_ctrl <- mval[,ctrl]
-  mval_case <- mval[,case]
-  # incorporate altered signals - change by +1 or -1
-  mval_case[rownames(mval_case) %in% pup2,] <-  mval_case[rownames(mval_case) %in% pup2,] + delta
-  mval_case[rownames(mval_case) %in% pdn2,] <-  mval_case[rownames(mval_case) %in% pdn2,] - delta
-  mval2 <- cbind(mval_ctrl,mval_case)
-  result <- list("mval"=mval2,"probes up"=pup2,"probes down"=pdn2,
-    "genes up"=gup2,"genes down"=gdn2,
-    "genesets up"=gtup,"genesets down"=gtdn)
-  return(result)
-}
-
-```
-
-## GSAMETH function
-
-```{r,gsameth_func}
-
-# limma
-runlimma <- function(mval,design,myann) {
-  fit.reduced <- lmFit(mval,design)
-  fit.reduced <- eBayes(fit.reduced)
-  dm <- topTable(fit.reduced,coef=ncol(design), number = Inf)
-  dm <- merge(myann,dm,by=0)
-  dm <- dm[order(dm$P.Value),]
-  rownames(dm) <- dm$Row.names
-  dm$Row.names=NULL
-  return(dm)
-}
-
-```
-
-This is how to use the function
-
-This could be complicated as it requires translation of symbols to entrez IDs, but
-could be simplified if the entrez translation is done at the gsameth step.
-
-```{r,rungsameth}
-
-simgsa <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
-  # generate gene sets
-  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
-  lengths <- unname(unlist(lapply(genesetdatabase,length)))
-  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
-  # select gene sets to alter
-  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
-  # incorporate select changes
-  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
-    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta,
-    gene_catalog=gene_catalog)
-  # set up limma
-  mval2 <- sim$mval
-  ncols <- ncol(mval2)
-  groupsize <- ncols/2
-  ss <- data.frame(colnames(mval2))
-  colnames(ss) <- "sample"
-  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
-  d <- model.matrix(~ ss$case )
-  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
-  pup3 <- rownames(subset(dm3,adj.P.Val<0.05 & logFC>0))
-  pdn3 <- rownames(subset(dm3,adj.P.Val<0.05 & logFC<0))
-  if ( length(pup3) < 250 ) { pup3 <- head(rownames(subset(dm3, logFC > 0)), 250) }
-  if ( length(pdn3) < 250 ) { pdn3 <- head(rownames(subset(dm3, logFC < 0)), 250) }
-  # convert gene sets to entrez
-  suppressWarnings(suppressMessages({ gene2entrez <- mapIds(org.Hs.eg.db, gene_catalog, 'ENTREZID', 'SYMBOL') }))
-  gsets_entrez <- lapply(gsets,function(gs) {
-    gs2 <- unique(gene2entrez[names(gene2entrez) %in% gs])
-    gs2 <- gs2[!is.na(gs2)]
-    return(gs2)
-  })
-  suppressWarnings(suppressMessages({
-    gsaup3 <- gsameth(sig.cpg=pup3, all.cpg=rownames(dm3), collection=gsets_entrez, array.type="EPIC")
-    gsadn3 <- gsameth(sig.cpg=pdn3, all.cpg=rownames(dm3), collection=gsets_entrez, array.type="EPIC")
-  }))
-  gsig_up3 <- rownames(subset(gsaup3,FDR<0.05))
-  gsig_dn3 <- rownames(subset(gsadn3,FDR<0.05))
-  gtup <- names(sim[[6]])
-  gtdn <- names(sim[[7]])
-  UPTP=length(intersect(gsig_up3 ,gtup))
-  UPFP=length(setdiff(gsig_up3 ,gtup))
-  UPFN=length(setdiff(gtup,gsig_up3))
-  DNTP=length(intersect(gsig_dn3 ,gtdn))
-  DNFP=length(setdiff(gsig_dn3 ,gtdn))
-  DNFN=length(setdiff(gtdn,gsig_dn3))
-  TP=UPTP+DNTP
-  FP=UPFP+DNFP
-  FN=UPFN+DNFN
-  TN=nrow(gsadn3)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
-  PREC=TP/(TP+FP)
-  REC=TP/(TP+FN)
-  F1=TP/(TP+(0.5*(FP+FN)))
-  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
-  return(result)
-}
-
-```
-
-## LA function
-
-This process runs limma first and then aggregates the results before doing an enrichment test.
-
-```{r,lafunc}
-
-# enrich parametric
-ttenrich <- function(m,genesets,cores=1,testtype="selfcontained") {
-  res <- mclapply( 1:length(genesets), function(i) {
-    scores <- m[,1]
-    gs <- genesets[i]
-    name <- names(gs)
-    n_members <- length(which(rownames(m) %in% gs[[1]]))
-    if ( n_members > 4 ) {
-      tstats <- m[which(rownames(m) %in% gs[[1]]),]
-      myn <- length(tstats)
-      mymean <- mean(tstats)
-      mymedian <- median(tstats)
-      if ( testtype == "selfcontained" ) { wt <- t.test(tstats) }
-      if ( testtype == "competitive" ) { wt <- t.test(tstats,scores) }
-      res <- c(name,myn,mymean,mymedian,wt$p.value)
-    }
-  } , mc.cores = cores)
-  res_df <- do.call(rbind, res)
-  rownames(res_df) <- res_df[,1]
-  res_df <- res_df[,-1]
-  colnames(res_df) <- c("n_genes","t_mean","t_median","pval")
-  tmp <- apply(res_df,2,as.numeric)
-  rownames(tmp) <- rownames(res_df)
-  res_df <- tmp
-  res_df <- as.data.frame(res_df)
-  res_df <- res_df[order(res_df$pval),]
-  res_df$logp <- -log10(res_df$pval )
-  res_df$fdr <- p.adjust(res_df$pval,method="fdr")
-  res_df[order(abs(res_df$pval)),]
-  return(res_df)
-}
-
-# enrich non-parametric
-wtenrich <- function(m,genesets,cores=1,testtype="selfcontained") {
-  res <- mclapply( 1:length(genesets), function(i) {
-    scores <- m[,1]
-    gs <- genesets[i]
-    name <- names(gs)
-    n_members <- length(which(rownames(m) %in% gs[[1]]))
-    if ( n_members > 4 ) {
-      tstats <- m[which(rownames(m) %in% gs[[1]]),]
-      myn <- length(tstats)
-      mymean <- mean(tstats)
-      mymedian <- median(tstats)
-      if ( testtype == "selfcontained" ) { wt <- wilcox.test(tstats) }
-      if ( testtype == "competitive" ) { wt <- wilcox.test(tstats,scores) }
-      res <- c(name,myn,mymean,mymedian,wt$p.value)
-    }
-  } , mc.cores = cores)
-  res_df <- do.call(rbind, res)
-  rownames(res_df) <- res_df[,1]
-  res_df <- res_df[,-1]
-  colnames(res_df) <- c("n_genes","t_mean","t_median","pval")
-  tmp <- apply(res_df,2,as.numeric)
-  rownames(tmp) <- rownames(res_df)
-  res_df <- tmp
-  res_df <- as.data.frame(res_df)
-  res_df <- res_df[order(res_df$pval),]
-  res_df$logp <- -log10(res_df$pval )
-  res_df$fdr <- p.adjust(res_df$pval,method="fdr")
-  res_df[order(abs(res_df$pval)),]
-  return(res_df)
-}
-
-# LA parametric competitive
-simlac <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
-  # generate gene sets
-  gene_catalog <- unique(gt$gene)
-  lengths <- unname(unlist(lapply(genesetdatabase,length)))
-  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
-  # select gene sets to alter
-  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
-  # incorporate select changes
-  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
-    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta,
-    gene_catalog=gene_catalog)
-  # set up limma
-  mval2 <- sim$mval
-  ncols <- ncol(mval2)
-  groupsize <- ncols/2
-  ss <- data.frame(colnames(mval2))
-  colnames(ss) <- "sample"
-  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
-  d <- model.matrix(~ ss$case )
-  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
-  dd <- merge(dm3,gt,by.x=0,by.y="probe")
-  m1 <- aggregate(t ~ gene,dd,mean)
-  rownames(m1) <- m1$gene
-  m1$gene=NULL
-  lares1 <- ttenrich(m=m1,genesets=gsets,cores=2,testtype="competitive")
-  gsig_up3 <- rownames(subset(lares1, fdr < 0.05 & t_mean > 0))
-  gsig_dn3 <- rownames(subset(lares1, fdr < 0.05 & t_mean < 0))
-  gtup <- names(sim[[6]])
-  gtdn <- names(sim[[7]])
-  UPTP=length(intersect(gsig_up3 ,gtup))
-  UPFP=length(setdiff(gsig_up3 ,gtup))
-  UPFN=length(setdiff(gtup,gsig_up3))
-  DNTP=length(intersect(gsig_dn3 ,gtdn))
-  DNFP=length(setdiff(gsig_dn3 ,gtdn))
-  DNFN=length(setdiff(gtdn,gsig_dn3))
-  TP=UPTP+DNTP
-  FP=UPFP+DNFP
-  FN=UPFN+DNFN
-  TN=nrow(lares1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
-  PREC=TP/(TP+FP)
-  REC=TP/(TP+FN)
-  F1=TP/(TP+(0.5*(FP+FN)))
-  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
-  return(result)
-}
-
-# LA parametric competitive top
-simlactop <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
-  # generate gene sets
-  gene_catalog <- unique(gt$gene)
-  lengths <- unname(unlist(lapply(genesetdatabase,length)))
-  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
-  # select gene sets to alter
-  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
-  # incorporate select changes
-  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
-    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta,
-    gene_catalog=gene_catalog)
-  # set up limma
-  mval2 <- sim$mval
-  ncols <- ncol(mval2)
-  groupsize <- ncols/2
-  ss <- data.frame(colnames(mval2))
-  colnames(ss) <- "sample"
-  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
-  d <- model.matrix(~ ss$case )
-  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
-  dd <- merge(dm3,gt,by.x=0,by.y="probe")
-  m1 <- aggregate(t ~ gene,dd, function(x) {
-    if (abs(max(x)) > abs(min(x))) { max(x) } else { min(x) }
-  })
-  rownames(m1) <- m1$gene
-  m1$gene=NULL
-  lares1 <- ttenrich(m=m1,genesets=gsets,cores=2,testtype="competitive")
-  gsig_up3 <- rownames(subset(lares1, fdr < 0.05 & t_mean > 0))
-  gsig_dn3 <- rownames(subset(lares1, fdr < 0.05 & t_mean < 0))
-  gtup <- names(sim[[6]])
-  gtdn <- names(sim[[7]])
-  UPTP=length(intersect(gsig_up3 ,gtup))
-  UPFP=length(setdiff(gsig_up3 ,gtup))
-  UPFN=length(setdiff(gtup,gsig_up3))
-  DNTP=length(intersect(gsig_dn3 ,gtdn))
-  DNFP=length(setdiff(gsig_dn3 ,gtdn))
-  DNFN=length(setdiff(gtdn,gsig_dn3))
-  TP=UPTP+DNTP
-  FP=UPFP+DNFP
-  FN=UPFN+DNFN
-  TN=nrow(lares1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
-  PREC=TP/(TP+FP)
-  REC=TP/(TP+FN)
-  F1=TP/(TP+(0.5*(FP+FN)))
-  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
-  return(result)
-}
-
-# LA nonparametric competitive
-simnlac <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
-  # generate gene sets
-  gene_catalog <- unique(gt$gene)
-  lengths <- unname(unlist(lapply(genesetdatabase,length)))
-  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
-  # select gene sets to alter
-  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
-  # incorporate select changes
-  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
-    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta,
-    gene_catalog=gene_catalog)
-  # set up limma
-  mval2 <- sim$mval
-  ncols <- ncol(mval2)
-  groupsize <- ncols/2
-  ss <- data.frame(colnames(mval2))
-  colnames(ss) <- "sample"
-  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
-  d <- model.matrix(~ ss$case )
-  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
-  dd <- merge(dm3,gt,by.x=0,by.y="probe")
-  m1 <- aggregate(t ~ gene,dd,mean)
-  rownames(m1) <- m1$gene
-  m1$gene=NULL
-  lares1 <- wtenrich(m=m1,genesets=gsets,cores=2,testtype="competitive")
-  gsig_up3 <- rownames(subset(lares1, fdr < 0.05 & t_mean > 0))
-  gsig_dn3 <- rownames(subset(lares1, fdr < 0.05 & t_mean < 0))
-  gtup <- names(sim[[6]])
-  gtdn <- names(sim[[7]])
-  UPTP=length(intersect(gsig_up3 ,gtup))
-  UPFP=length(setdiff(gsig_up3 ,gtup))
-  UPFN=length(setdiff(gtup,gsig_up3))
-  DNTP=length(intersect(gsig_dn3 ,gtdn))
-  DNFP=length(setdiff(gsig_dn3 ,gtdn))
-  DNFN=length(setdiff(gtdn,gsig_dn3))
-  TP=UPTP+DNTP
-  FP=UPFP+DNFP
-  FN=UPFN+DNFN
-  TN=nrow(lares1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
-  PREC=TP/(TP+FP)
-  REC=TP/(TP+FN)
-  F1=TP/(TP+(0.5*(FP+FN)))
-  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
-  return(result)
-}
-
-# LA competitive mitch
-simlacm <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
-  # generate gene sets
-  gene_catalog <- unique(gt$gene)
-  lengths <- unname(unlist(lapply(genesetdatabase,length)))
-  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
-  # select gene sets to alter
-  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
-  # incorporate select changes
-  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
-    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta,
-    gene_catalog=gene_catalog)
-  # set up limma
-  mval2 <- sim$mval
-  ncols <- ncol(mval2)
-  groupsize <- ncols/2
-  ss <- data.frame(colnames(mval2))
-  colnames(ss) <- "sample"
-  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
-  d <- model.matrix(~ ss$case )
-  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
-  dd <- merge(dm3,gt,by.x=0,by.y="probe")
-  m1 <- aggregate(t ~ gene,dd,mean)
-  rownames(m1) <- m1$gene
-  m1$gene=NULL
-  lamres1 <- runmitch(m=m1,genesets=gsets,cores=2)
-  gsig_up3 <- rownames( subset( lamres1, p.adjustANOVA < 0.05 & s.dist > 0 ) )
-  gsig_dn3 <- rownames( subset( lamres1, p.adjustANOVA < 0.05 & s.dist < 0 ) )
-  gtup <- names(sim[[6]])
-  gtdn <- names(sim[[7]])
-  UPTP=length(intersect(gsig_up3 ,gtup))
-  UPFP=length(setdiff(gsig_up3 ,gtup))
-  UPFN=length(setdiff(gtup,gsig_up3))
-  DNTP=length(intersect(gsig_dn3 ,gtdn))
-  DNFP=length(setdiff(gsig_dn3 ,gtdn))
-  DNFN=length(setdiff(gtdn,gsig_dn3))
-  TP=UPTP+DNTP
-  FP=UPFP+DNFP
-  FN=UPFN+DNFN
-  TN=nrow(lamres1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
-  PREC=TP/(TP+FP)
-  REC=TP/(TP+FN)
-  F1=TP/(TP+(0.5*(FP+FN)))
-  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
-  return(result)
-}
-
-# LA rank competitive mitch
-simlrm <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
-  # generate gene sets
-  gene_catalog <- unique(gt$gene)
-  lengths <- unname(unlist(lapply(genesetdatabase,length)))
-  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
-  # select gene sets to alter
-  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
-  # incorporate select changes
-  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
-    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta,
-    gene_catalog=gene_catalog)
-  # set up limma
-  mval2 <- sim$mval
-  ncols <- ncol(mval2)
-  groupsize <- ncols/2
-  ss <- data.frame(colnames(mval2))
-  colnames(ss) <- "sample"
-  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
-  d <- model.matrix(~ ss$case )
-  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
-  # rank probes first
-  dm3$rank <-  rank(dm3$t) - nrow(subset(dm3,t<0))
-  mm <- merge(dm3,gt,by.x=0,by.y="probe")
-  head(mm)
-  mma <-aggregate(mm$rank ~ gene,mm,mean)
-  rownames(mma) <- mma$gene
-  mma$gene = NULL
-  colnames(mma) <- "meanrank"
-  lrmres <- runmitch(m=mma,genesets=gsets,cores=2)
-  gsig_up3 <- rownames( subset( lrmres, p.adjustANOVA < 0.05 & s.dist > 0 ) )
-  gsig_dn3 <- rownames( subset( lrmres, p.adjustANOVA < 0.05 & s.dist < 0 ) )
-  gtup <- names(sim[[6]])
-  gtdn <- names(sim[[7]])
-  UPTP=length(intersect(gsig_up3 ,gtup))
-  UPFP=length(setdiff(gsig_up3 ,gtup))
-  UPFN=length(setdiff(gtup,gsig_up3))
-  DNTP=length(intersect(gsig_dn3 ,gtdn))
-  DNFP=length(setdiff(gsig_dn3 ,gtdn))
-  DNFN=length(setdiff(gtdn,gsig_dn3))
-  TP=UPTP+DNTP
-  FP=UPFP+DNFP
-  FN=UPFN+DNFN
-  TN=nrow(lrmres)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
-  PREC=TP/(TP+FP)
-  REC=TP/(TP+FN)
-  F1=TP/(TP+(0.5*(FP+FN)))
-  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
-  return(result)
-}
-
-```
-
-## AL: Aggregate limma
-
-Functions for aggregate-limma-enrich approach.
-
-```{r,al}
-
-# AL approach parametric competitive
-simalc <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
-  # generate gene sets
-  gene_catalog <- unique(gt$gene)
-  lengths <- unname(unlist(lapply(genesetdatabase,length)))
-  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
-  # select gene sets to alter
-  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
-  # incorporate select changes
-  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
-    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta,
-    gene_catalog=gene_catalog)
-  # set up limma
-  mval2 <- sim$mval
-  ncols <- ncol(mval2)
-  groupsize <- ncols/2
-  ss <- data.frame(colnames(mval2))
-  colnames(ss) <- "sample"
-  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
-  d <- model.matrix(~ ss$case )
-  # al pipeline
-  mm <- merge(mval2,gt,by.x=0,by.y="probe")
-  mm$Row.names = NULL
-  a <- aggregate(. ~ gene,mm,mean)
-  rownames(a) <- a$gene
-  a$gene=NULL
-  fit.reduced <- lmFit(a,d)
-  fit.reduced <- eBayes(fit.reduced)
-  al <- topTable(fit.reduced,coef=ncol(d), number = Inf)
-  m1 <- as.data.frame(al$t)
-  rownames(m1) <- rownames(al)
-  colnames(m1) <- "t"
-  alres1 <- ttenrich(m=m1,genesets=gsets,cores=2,testtype="competitive")
-  # summarise results
-  gsig_up3 <- rownames(subset(alres1, fdr < 0.05 & t_mean > 0))
-  gsig_dn3 <- rownames(subset(alres1, fdr < 0.05 & t_mean < 0))
-  gtup <- names(sim[[6]])
-  gtdn <- names(sim[[7]])
-  UPTP=length(intersect(gsig_up3 ,gtup))
-  UPFP=length(setdiff(gsig_up3 ,gtup))
-  UPFN=length(setdiff(gtup,gsig_up3))
-  DNTP=length(intersect(gsig_dn3 ,gtdn))
-  DNFP=length(setdiff(gsig_dn3 ,gtdn))
-  DNFN=length(setdiff(gtdn,gsig_dn3))
-  TP=UPTP+DNTP
-  FP=UPFP+DNFP
-  FN=UPFN+DNFN
-  TN=nrow(alres1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
-  PREC=TP/(TP+FP)
-  REC=TP/(TP+FN)
-  F1=TP/(TP+(0.5*(FP+FN)))
-  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
-  return(result)
-}
-
-# AL approach nonparametric competitive
-simnalc <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
-  # generate gene sets
-  gene_catalog <- unique(gt$gene)
-  lengths <- unname(unlist(lapply(genesetdatabase,length)))
-  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
-  # select gene sets to alter
-  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
-  # incorporate select changes
-  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
-    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta,
-    gene_catalog=gene_catalog)
-  # set up limma
-  mval2 <- sim$mval
-  ncols <- ncol(mval2)
-  groupsize <- ncols/2
-  ss <- data.frame(colnames(mval2))
-  colnames(ss) <- "sample"
-  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
-  d <- model.matrix(~ ss$case )
-  # al pipeline
-  mm <- merge(mval2,gt,by.x=0,by.y="probe")
-  mm$Row.names = NULL
-  a <- aggregate(. ~ gene,mm,mean)
-  rownames(a) <- a$gene
-  a$gene=NULL
-  fit.reduced <- lmFit(a,d)
-  fit.reduced <- eBayes(fit.reduced)
-  al <- topTable(fit.reduced,coef=ncol(d), number = Inf)
-  m1 <- as.data.frame(al$t)
-  rownames(m1) <- rownames(al)
-  colnames(m1) <- "t"
-  alres1 <- wtenrich(m=m1,genesets=gsets,cores=2,testtype="competitive")
-  # summarise results
-  gsig_up3 <- rownames(subset(alres1, fdr < 0.05 & t_mean > 0))
-  gsig_dn3 <- rownames(subset(alres1, fdr < 0.05 & t_mean < 0))
-  gtup <- names(sim[[6]])
-  gtdn <- names(sim[[7]])
-  UPTP=length(intersect(gsig_up3 ,gtup))
-  UPFP=length(setdiff(gsig_up3 ,gtup))
-  UPFN=length(setdiff(gtup,gsig_up3))
-  DNTP=length(intersect(gsig_dn3 ,gtdn))
-  DNFP=length(setdiff(gsig_dn3 ,gtdn))
-  DNFN=length(setdiff(gtdn,gsig_dn3))
-  TP=UPTP+DNTP
-  FP=UPFP+DNFP
-  FN=UPFN+DNFN
-  TN=nrow(alres1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
-  PREC=TP/(TP+FP)
-  REC=TP/(TP+FN)
-  F1=TP/(TP+(0.5*(FP+FN)))
-  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
-  return(result)
-}
-
-```
-
-## Agg-limma-mitch function
-
-This approach uses the aggregated mvals, limma and instead of a 1-sample t-test it uses mitch
-which is a competitive test and could give more interpretable results.
-
-```{r,alm}
-
-runmitch <- function(m,genesets,cores=1) {
-  suppressMessages({ mres <- mitch_calc(m,genesets,minsetsize=5,cores=cores) })
-  mres <- mres$enrichment_result
-  rownames(mres) <- mres$set
-  mres$set=NULL
-  return(mres)
-}
-
-simalm <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
-  # generate gene sets
-  gene_catalog <- unique(gt$gene)
-  lengths <- unname(unlist(lapply(genesetdatabase,length)))
-  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
-  # select gene sets to alter
-  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
-  # incorporate select changes
-  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
-    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta,
-    gene_catalog=gene_catalog)
-  # set up limma
-  mval2 <- sim$mval
-  ncols <- ncol(mval2)
-  groupsize <- ncols/2
-  ss <- data.frame(colnames(mval2))
-  colnames(ss) <- "sample"
-  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
-  d <- model.matrix(~ ss$case )
-  # alm pipeline
-  mm <- merge(mval2,gt,by.x=0,by.y="probe")
-  mm$Row.names = NULL
-  a <- aggregate(. ~ gene,mm,mean)
-  rownames(a) <- a$gene
-  a$gene=NULL
-  fit.reduced <- lmFit(a,d)
-  fit.reduced <- eBayes(fit.reduced)
-  al <- topTable(fit.reduced,coef=ncol(d), number = Inf)
-  m1 <- as.data.frame(al$t)
-  rownames(m1) <- rownames(al)
-  colnames(m1) <- "t"
-  almres1 <- runmitch(m=m1,genesets=gsets,cores=2)
-  # summarise results
-  gsig_up3 <- rownames( subset( almres1, p.adjustANOVA < 0.05 & s.dist > 0 ) )
-  gsig_dn3 <- rownames( subset( almres1, p.adjustANOVA < 0.05 & s.dist < 0 ) )
-  gtup <- names(sim[[6]])
-  gtdn <- names(sim[[7]])
-  UPTP=length(intersect(gsig_up3 ,gtup))
-  UPFP=length(setdiff(gsig_up3 ,gtup))
-  UPFN=length(setdiff(gtup,gsig_up3))
-  DNTP=length(intersect(gsig_dn3 ,gtdn))
-  DNFP=length(setdiff(gsig_dn3 ,gtdn))
-  DNFN=length(setdiff(gtdn,gsig_dn3))
-  TP=UPTP+DNTP
-  FP=UPFP+DNFP
-  FN=UPFN+DNFN
-  TN=nrow(almres1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
-  PREC=TP/(TP+FP)
-  REC=TP/(TP+FN)
-  F1=TP/(TP+(0.5*(FP+FN)))
-  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
-  return(result)
-}
-
-```
-
-## AA Aggregate-aggregate-limma functions
-
-Use mean value works well here.
-
-```{r,aafunc}
-
-gsagg <- function(x,genesets,cores=1) {
-  meds <- mclapply(1:length(genesets), function(i) {
-    gs = genesets[[i]]
-    xx <- x[rownames(x) %in% gs,]
-    med <- apply(xx,2,mean)
-  },mc.cores=cores)
-  mymed <- do.call(rbind,meds)
-  rownames(mymed) <- names(genesets)
-  as.data.frame(mymed)
-}
-
-aalimma <- function(agag,design) {
-  fit.reduced <- lmFit(agag,design)
-  fit.reduced <- eBayes(fit.reduced)
-  dmagg <- topTable(fit.reduced,coef=ncol(design), number = Inf)
-  return(dmagg)
-}
-
-aal <- function(mval,myann,genesets,design,cores=1) {
-  medf <- chragg(mval,myann,cores=cores)
-  agag <- gsagg(x=medf,genesets=genesets,cores=cores)
-  aalres <- aalimma(agag=agag,design=design)
-  return(aalres)
-}
-
-simaa <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
-  # generate gene sets
-  gene_catalog <- unique(gt$gene)
-  lengths <- unname(unlist(lapply(genesetdatabase,length)))
-  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
-  # select gene sets to alter
-  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
-  # incorporate select changes
-  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
-    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta,
-    gene_catalog=gene_catalog)
-  # set up limma
-  mval2 <- sim$mval
-  ncols <- ncol(mval2)
-  groupsize <- ncols/2
-  ss <- data.frame(colnames(mval2))
-  colnames(ss) <- "sample"
-  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
-  d <- model.matrix(~ ss$case )
-  mm <- merge(mval2,gt,by.x=0,by.y="probe")
-  mm$Row.names = NULL
-  a <- aggregate(. ~ gene,mm,mean)
-  rownames(a) <- a$gene
-  a$gene=NULL
-  mystack <- stack(gsets)
-  mmm <- merge(a,mystack,by.x=0,by.y="values")
-  mmm$Row.names=NULL
-  aa <- aggregate(. ~ ind,mmm,mean)
-  rownames(aa) <- aa$ind
-  aa$ind=NULL
-  fit.reduced <- lmFit(aa,d)
-  fit.reduced <- eBayes(fit.reduced)
-  aares1 <- topTable(fit.reduced,coef=ncol(d), number = Inf)
-  # summarise results
-  gsig_up3 <- rownames(subset(aares1, adj.P.Val < 0.05 & logFC > 0))
-  gsig_dn3 <- rownames(subset(aares1, adj.P.Val < 0.05 & logFC < 0))
-  gtup <- names(sim[[6]])
-  gtdn <- names(sim[[7]])
-  UPTP=length(intersect(gsig_up3 ,gtup))
-  UPFP=length(setdiff(gsig_up3 ,gtup))
-  UPFN=length(setdiff(gtup,gsig_up3))
-  DNTP=length(intersect(gsig_dn3 ,gtdn))
-  DNFP=length(setdiff(gsig_dn3 ,gtdn))
-  DNFN=length(setdiff(gtdn,gsig_dn3))
-  TP=UPTP+DNTP
-  FP=UPFP+DNFP
-  FN=UPFN+DNFN
-  TN=nrow(aares1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
-  PREC=TP/(TP+FP)
-  REC=TP/(TP+FN)
-  F1=TP/(TP+(0.5*(FP+FN)))
-  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
-  return(result)
-}
-
-F1 <- function(x,y) {
-  ( 2 * x * y ) / ( x + y )
-}
 
@@
 
-## GSA meth sim
+## GSA meth
 
@@
 
-## LA sim
+## LAT
 
-### parametric competitive
+Limma Average t-test.
 
-```{r,runlac}
+```{r,runlat}
 
@@
 
-### parametric competitive with "top" aggregation
+## LTT
 
-Too many false positives.
+Limma top t-test.
 
-```{r,runlactop}
+```{r,runltt}
 
@@
 
-### nonparametric competitive
+## LAW
 
-```{r,runnlac}
+Limma average wilcox.
+
+```{r,runlaw}
 
@@
 
-### LA competitive mitch
+### LAM
 
-```{r,runlacm}
+Limma average mitch.
+
+```{r,runlam}
 
@@
 
-### LA rank probes
+### LRM
+
+Limma rank mitch.
 
@@
 
-## AL sim
+## ALT
 
-### parametric competitive
+Aggregate limma t-test.
 
-```{r,runalc}
+```{r,runalt}
 
@@
 
-### nonparametric competitive
+## ALW
 
-```{r,runnalc}
+Aggregate limma wilcox.
+
+```{r,runalw}
 
@@
 
-## ALM competitive test
+## ALM
+
+Aggregate limma mitch.
 
@@
 ## AA sim
+
+Aggregate, aggregate limma.
 
@@
 ```
-
-## Notes
-
-LA
-
-* simla: parametric self-contained
-
-* simlac: parametric competitive
-
-* simnla: nonparametric self-contained
-
-* simnlac: nonparametric competitive
-
-AL
-
-* simal: parametric self-contained
-
-* simalc: parametric competitive
-
-* simnal: nonparametric self-contained
-
-* simnalc: nonparametric competitive

### 2024-01-08 figs/GSE158422_simulate050.Rmd
offload functions to separate file

@@
 
-## GSA meth sim
+## GSA meth
 
-CCannot be run in multicore due to fragility of AnnotationDbi SQLite objects.
+Cannot be run in multicore due to fragility of AnnotationDbi SQLite objects.
 

### 2024-01-08 figs/GSE158422_simulate100.Rmd
offload functions to separate file

@@
 
-incorp_dm <- function(genesets,myann,mval,seed, frac_genes,frac_probes,
-  groupsize,delta=1,gene_catalog) {
-
-  # divide gene sets between hyper and hypomethylated
-  nset <- floor(length(genesets)/2)
-  set.seed(seed) ; gtup <-sample(genesets,nset)
-  set.seed(seed) ; gtdn <- sample(setdiff(genesets,gtup),nset)
-  gup <- unname(unlist(gtup))
-  gdn <- unname(unlist(gtdn))
-  # add extra 10% DM genes
-  gnon <- setdiff(gene_catalog,c(gup,gdn))
-  gextra <- round(length(gnon)*0.1)
-  set.seed(seed) ; gup <- c(gup,sample(gnon,gextra))
-  gnon <- setdiff(gene_catalog,c(gup,gdn))
-  set.seed(seed) ; gdn <- c(gdn,sample(gnon,gextra))
-  # make probe-gene vector
-  probe2gene <- strsplit(myann$UCSC_RefGene_Name,";")
-  names(probe2gene) <- rownames(myann)
-  probe2gene <- unlist(probe2gene)
-  # select probes hypermethylated
-  set.seed(seed) ; gup2 <- sample(gup,floor(length(gup)*frac_genes))
-  pup <- names(probe2gene[which(probe2gene %in% gup2)])
-  set.seed(seed) ; pup2 <- sample(pup,floor(length(pup)*frac_probes))
-  # select probes hypomethylated
-  set.seed(seed) ; gdn2 <- sample(gdn,floor(length(gdn)*frac_genes))
-  pdn <- names(probe2gene[which(probe2gene %in% gdn2)])
-  set.seed(seed) ; pdn2 <- sample(pdn,floor(length(pdn)*frac_probes))
-  # add 10% DM probes as well
-  probes <- rownames(myann)
-  pnon <- setdiff(probes,c(pup,pdn))
-  pextra <- round(length(pnon)*0.1)
-  set.seed(seed) ; pup <- c(pup,sample(pnon,pextra))
-  pnon <- setdiff(probes,c(pup,pdn))
-  set.seed(seed) ; pdn <- c(pdn,sample(pnon,pextra))
-  # divide samples between ctrl and case
-  ncols <- ncol(mval)
-  maxgroupsize=floor(ncols/2)
-  if ( groupsize > maxgroupsize ) { stop("groupsize cannot be larger than half the ncols of mval") }
-  set.seed(seed) ; ctrl <- sample(1:ncols,groupsize)
-  set.seed(seed) ; case <- sample(setdiff(1:ncols,ctrl),groupsize)
-  mval_ctrl <- mval[,ctrl]
-  mval_case <- mval[,case]
-  # incorporate altered signals - change by +1 or -1
-  mval_case[rownames(mval_case) %in% pup2,] <-  mval_case[rownames(mval_case) %in% pup2,] + delta
-  mval_case[rownames(mval_case) %in% pdn2,] <-  mval_case[rownames(mval_case) %in% pdn2,] - delta
-  mval2 <- cbind(mval_ctrl,mval_case)
-  result <- list("mval"=mval2,"probes up"=pup2,"probes down"=pdn2,
-    "genes up"=gup2,"genes down"=gdn2,
-    "genesets up"=gtup,"genesets down"=gtdn)
-  return(result)
-}
-
-```
-
-## GSAMETH function
-
-```{r,gsameth_func}
-
-# limma
-runlimma <- function(mval,design,myann) {
-  fit.reduced <- lmFit(mval,design)
-  fit.reduced <- eBayes(fit.reduced)
-  dm <- topTable(fit.reduced,coef=ncol(design), number = Inf)
-  dm <- merge(myann,dm,by=0)
-  dm <- dm[order(dm$P.Value),]
-  rownames(dm) <- dm$Row.names
-  dm$Row.names=NULL
-  return(dm)
-}
-
-```
-
-This is how to use the function
-
-This could be complicated as it requires translation of symbols to entrez IDs, but
-could be simplified if the entrez translation is done at the gsameth step.
-
-```{r,rungsameth}
-
-simgsa <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
-  # generate gene sets
-  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
-  lengths <- unname(unlist(lapply(genesetdatabase,length)))
-  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
-  # select gene sets to alter
-  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
-  # incorporate select changes
-  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
-    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta,
-    gene_catalog=gene_catalog)
-  # set up limma
-  mval2 <- sim$mval
-  ncols <- ncol(mval2)
-  groupsize <- ncols/2
-  ss <- data.frame(colnames(mval2))
-  colnames(ss) <- "sample"
-  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
-  d <- model.matrix(~ ss$case )
-  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
-  pup3 <- rownames(subset(dm3,adj.P.Val<0.05 & logFC>0))
-  pdn3 <- rownames(subset(dm3,adj.P.Val<0.05 & logFC<0))
-  if ( length(pup3) < 250 ) { pup3 <- head(rownames(subset(dm3, logFC > 0)), 250) }
-  if ( length(pdn3) < 250 ) { pdn3 <- head(rownames(subset(dm3, logFC < 0)), 250) }
-  # convert gene sets to entrez
-  suppressWarnings(suppressMessages({ gene2entrez <- mapIds(org.Hs.eg.db, gene_catalog, 'ENTREZID', 'SYMBOL') }))
-  gsets_entrez <- lapply(gsets,function(gs) {
-    gs2 <- unique(gene2entrez[names(gene2entrez) %in% gs])
-    gs2 <- gs2[!is.na(gs2)]
-    return(gs2)
-  })
-  suppressWarnings(suppressMessages({
-    gsaup3 <- gsameth(sig.cpg=pup3, all.cpg=rownames(dm3), collection=gsets_entrez, array.type="EPIC")
-    gsadn3 <- gsameth(sig.cpg=pdn3, all.cpg=rownames(dm3), collection=gsets_entrez, array.type="EPIC")
-  }))
-  gsig_up3 <- rownames(subset(gsaup3,FDR<0.05))
-  gsig_dn3 <- rownames(subset(gsadn3,FDR<0.05))
-  gtup <- names(sim[[6]])
-  gtdn <- names(sim[[7]])
-  UPTP=length(intersect(gsig_up3 ,gtup))
-  UPFP=length(setdiff(gsig_up3 ,gtup))
-  UPFN=length(setdiff(gtup,gsig_up3))
-  DNTP=length(intersect(gsig_dn3 ,gtdn))
-  DNFP=length(setdiff(gsig_dn3 ,gtdn))
-  DNFN=length(setdiff(gtdn,gsig_dn3))
-  TP=UPTP+DNTP
-  FP=UPFP+DNFP
-  FN=UPFN+DNFN
-  TN=nrow(gsadn3)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
-  PREC=TP/(TP+FP)
-  REC=TP/(TP+FN)
-  F1=TP/(TP+(0.5*(FP+FN)))
-  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
-  return(result)
-}
-
-```
-
-## LA function
-
-This process runs limma first and then aggregates the results before doing an enrichment test.
-
-```{r,lafunc}
-
-# enrich parametric
-ttenrich <- function(m,genesets,cores=1,testtype="selfcontained") {
-  res <- mclapply( 1:length(genesets), function(i) {
-    scores <- m[,1]
-    gs <- genesets[i]
-    name <- names(gs)
-    n_members <- length(which(rownames(m) %in% gs[[1]]))
-    if ( n_members > 4 ) {
-      tstats <- m[which(rownames(m) %in% gs[[1]]),]
-      myn <- length(tstats)
-      mymean <- mean(tstats)
-      mymedian <- median(tstats)
-      if ( testtype == "selfcontained" ) { wt <- t.test(tstats) }
-      if ( testtype == "competitive" ) { wt <- t.test(tstats,scores) }
-      res <- c(name,myn,mymean,mymedian,wt$p.value)
-    }
-  } , mc.cores = cores)
-  res_df <- do.call(rbind, res)
-  rownames(res_df) <- res_df[,1]
-  res_df <- res_df[,-1]
-  colnames(res_df) <- c("n_genes","t_mean","t_median","pval")
-  tmp <- apply(res_df,2,as.numeric)
-  rownames(tmp) <- rownames(res_df)
-  res_df <- tmp
-  res_df <- as.data.frame(res_df)
-  res_df <- res_df[order(res_df$pval),]
-  res_df$logp <- -log10(res_df$pval )
-  res_df$fdr <- p.adjust(res_df$pval,method="fdr")
-  res_df[order(abs(res_df$pval)),]
-  return(res_df)
-}
-
-# enrich non-parametric
-wtenrich <- function(m,genesets,cores=1,testtype="selfcontained") {
-  res <- mclapply( 1:length(genesets), function(i) {
-    scores <- m[,1]
-    gs <- genesets[i]
-    name <- names(gs)
-    n_members <- length(which(rownames(m) %in% gs[[1]]))
-    if ( n_members > 4 ) {
-      tstats <- m[which(rownames(m) %in% gs[[1]]),]
-      myn <- length(tstats)
-      mymean <- mean(tstats)
-      mymedian <- median(tstats)
-      if ( testtype == "selfcontained" ) { wt <- wilcox.test(tstats) }
-      if ( testtype == "competitive" ) { wt <- wilcox.test(tstats,scores) }
-      res <- c(name,myn,mymean,mymedian,wt$p.value)
-    }
-  } , mc.cores = cores)
-  res_df <- do.call(rbind, res)
-  rownames(res_df) <- res_df[,1]
-  res_df <- res_df[,-1]
-  colnames(res_df) <- c("n_genes","t_mean","t_median","pval")
-  tmp <- apply(res_df,2,as.numeric)
-  rownames(tmp) <- rownames(res_df)
-  res_df <- tmp
-  res_df <- as.data.frame(res_df)
-  res_df <- res_df[order(res_df$pval),]
-  res_df$logp <- -log10(res_df$pval )
-  res_df$fdr <- p.adjust(res_df$pval,method="fdr")
-  res_df[order(abs(res_df$pval)),]
-  return(res_df)
-}
-
-# LA parametric competitive
-simlac <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
-  # generate gene sets
-  gene_catalog <- unique(gt$gene)
-  lengths <- unname(unlist(lapply(genesetdatabase,length)))
-  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
-  # select gene sets to alter
-  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
-  # incorporate select changes
-  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
-    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta,
-    gene_catalog=gene_catalog)
-  # set up limma
-  mval2 <- sim$mval
-  ncols <- ncol(mval2)
-  groupsize <- ncols/2
-  ss <- data.frame(colnames(mval2))
-  colnames(ss) <- "sample"
-  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
-  d <- model.matrix(~ ss$case )
-  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
-  dd <- merge(dm3,gt,by.x=0,by.y="probe")
-  m1 <- aggregate(t ~ gene,dd,mean)
-  rownames(m1) <- m1$gene
-  m1$gene=NULL
-  lares1 <- ttenrich(m=m1,genesets=gsets,cores=2,testtype="competitive")
-  gsig_up3 <- rownames(subset(lares1, fdr < 0.05 & t_mean > 0))
-  gsig_dn3 <- rownames(subset(lares1, fdr < 0.05 & t_mean < 0))
-  gtup <- names(sim[[6]])
-  gtdn <- names(sim[[7]])
-  UPTP=length(intersect(gsig_up3 ,gtup))
-  UPFP=length(setdiff(gsig_up3 ,gtup))
-  UPFN=length(setdiff(gtup,gsig_up3))
-  DNTP=length(intersect(gsig_dn3 ,gtdn))
-  DNFP=length(setdiff(gsig_dn3 ,gtdn))
-  DNFN=length(setdiff(gtdn,gsig_dn3))
-  TP=UPTP+DNTP
-  FP=UPFP+DNFP
-  FN=UPFN+DNFN
-  TN=nrow(lares1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
-  PREC=TP/(TP+FP)
-  REC=TP/(TP+FN)
-  F1=TP/(TP+(0.5*(FP+FN)))
-  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
-  return(result)
-}
-
-# LA parametric competitive top
-simlactop <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
-  # generate gene sets
-  gene_catalog <- unique(gt$gene)
-  lengths <- unname(unlist(lapply(genesetdatabase,length)))
-  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
-  # select gene sets to alter
-  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
-  # incorporate select changes
-  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
-    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta,
-    gene_catalog=gene_catalog)
-  # set up limma
-  mval2 <- sim$mval
-  ncols <- ncol(mval2)
-  groupsize <- ncols/2
-  ss <- data.frame(colnames(mval2))
-  colnames(ss) <- "sample"
-  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
-  d <- model.matrix(~ ss$case )
-  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
-  dd <- merge(dm3,gt,by.x=0,by.y="probe")
-  m1 <- aggregate(t ~ gene,dd, function(x) {
-    if (abs(max(x)) > abs(min(x))) { max(x) } else { min(x) }
-  })
-  rownames(m1) <- m1$gene
-  m1$gene=NULL
-  lares1 <- ttenrich(m=m1,genesets=gsets,cores=2,testtype="competitive")
-  gsig_up3 <- rownames(subset(lares1, fdr < 0.05 & t_mean > 0))
-  gsig_dn3 <- rownames(subset(lares1, fdr < 0.05 & t_mean < 0))
-  gtup <- names(sim[[6]])
-  gtdn <- names(sim[[7]])
-  UPTP=length(intersect(gsig_up3 ,gtup))
-  UPFP=length(setdiff(gsig_up3 ,gtup))
-  UPFN=length(setdiff(gtup,gsig_up3))
-  DNTP=length(intersect(gsig_dn3 ,gtdn))
-  DNFP=length(setdiff(gsig_dn3 ,gtdn))
-  DNFN=length(setdiff(gtdn,gsig_dn3))
-  TP=UPTP+DNTP
-  FP=UPFP+DNFP
-  FN=UPFN+DNFN
-  TN=nrow(lares1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
-  PREC=TP/(TP+FP)
-  REC=TP/(TP+FN)
-  F1=TP/(TP+(0.5*(FP+FN)))
-  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
-  return(result)
-}
-
-# LA nonparametric competitive
-simnlac <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
-  # generate gene sets
-  gene_catalog <- unique(gt$gene)
-  lengths <- unname(unlist(lapply(genesetdatabase,length)))
-  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
-  # select gene sets to alter
-  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
-  # incorporate select changes
-  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
-    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta,
-    gene_catalog=gene_catalog)
-  # set up limma
-  mval2 <- sim$mval
-  ncols <- ncol(mval2)
-  groupsize <- ncols/2
-  ss <- data.frame(colnames(mval2))
-  colnames(ss) <- "sample"
-  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
-  d <- model.matrix(~ ss$case )
-  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
-  dd <- merge(dm3,gt,by.x=0,by.y="probe")
-  m1 <- aggregate(t ~ gene,dd,mean)
-  rownames(m1) <- m1$gene
-  m1$gene=NULL
-  lares1 <- wtenrich(m=m1,genesets=gsets,cores=2,testtype="competitive")
-  gsig_up3 <- rownames(subset(lares1, fdr < 0.05 & t_mean > 0))
-  gsig_dn3 <- rownames(subset(lares1, fdr < 0.05 & t_mean < 0))
-  gtup <- names(sim[[6]])
-  gtdn <- names(sim[[7]])
-  UPTP=length(intersect(gsig_up3 ,gtup))
-  UPFP=length(setdiff(gsig_up3 ,gtup))
-  UPFN=length(setdiff(gtup,gsig_up3))
-  DNTP=length(intersect(gsig_dn3 ,gtdn))
-  DNFP=length(setdiff(gsig_dn3 ,gtdn))
-  DNFN=length(setdiff(gtdn,gsig_dn3))
-  TP=UPTP+DNTP
-  FP=UPFP+DNFP
-  FN=UPFN+DNFN
-  TN=nrow(lares1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
-  PREC=TP/(TP+FP)
-  REC=TP/(TP+FN)
-  F1=TP/(TP+(0.5*(FP+FN)))
-  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
-  return(result)
-}
-
-# LA competitive mitch
-simlacm <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
-  # generate gene sets
-  gene_catalog <- unique(gt$gene)
-  lengths <- unname(unlist(lapply(genesetdatabase,length)))
-  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
-  # select gene sets to alter
-  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
-  # incorporate select changes
-  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
-    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta,
-    gene_catalog=gene_catalog)
-  # set up limma
-  mval2 <- sim$mval
-  ncols <- ncol(mval2)
-  groupsize <- ncols/2
-  ss <- data.frame(colnames(mval2))
-  colnames(ss) <- "sample"
-  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
-  d <- model.matrix(~ ss$case )
-  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
-  dd <- merge(dm3,gt,by.x=0,by.y="probe")
-  m1 <- aggregate(t ~ gene,dd,mean)
-  rownames(m1) <- m1$gene
-  m1$gene=NULL
-  lamres1 <- runmitch(m=m1,genesets=gsets,cores=2)
-  gsig_up3 <- rownames( subset( lamres1, p.adjustANOVA < 0.05 & s.dist > 0 ) )
-  gsig_dn3 <- rownames( subset( lamres1, p.adjustANOVA < 0.05 & s.dist < 0 ) )
-  gtup <- names(sim[[6]])
-  gtdn <- names(sim[[7]])
-  UPTP=length(intersect(gsig_up3 ,gtup))
-  UPFP=length(setdiff(gsig_up3 ,gtup))
-  UPFN=length(setdiff(gtup,gsig_up3))
-  DNTP=length(intersect(gsig_dn3 ,gtdn))
-  DNFP=length(setdiff(gsig_dn3 ,gtdn))
-  DNFN=length(setdiff(gtdn,gsig_dn3))
-  TP=UPTP+DNTP
-  FP=UPFP+DNFP
-  FN=UPFN+DNFN
-  TN=nrow(lamres1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
-  PREC=TP/(TP+FP)
-  REC=TP/(TP+FN)
-  F1=TP/(TP+(0.5*(FP+FN)))
-  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
-  return(result)
-}
-
-# LA rank competitive mitch
-simlrm <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
-  # generate gene sets
-  gene_catalog <- unique(gt$gene)
-  lengths <- unname(unlist(lapply(genesetdatabase,length)))
-  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
-  # select gene sets to alter
-  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
-  # incorporate select changes
-  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
-    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta,
-    gene_catalog=gene_catalog)
-  # set up limma
-  mval2 <- sim$mval
-  ncols <- ncol(mval2)
-  groupsize <- ncols/2
-  ss <- data.frame(colnames(mval2))
-  colnames(ss) <- "sample"
-  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
-  d <- model.matrix(~ ss$case )
-  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
-  # rank probes first
-  dm3$rank <-  rank(dm3$t) - nrow(subset(dm3,t<0))
-  mm <- merge(dm3,gt,by.x=0,by.y="probe")
-  head(mm)
-  mma <-aggregate(mm$rank ~ gene,mm,mean)
-  rownames(mma) <- mma$gene
-  mma$gene = NULL
-  colnames(mma) <- "meanrank"
-  lrmres <- runmitch(m=mma,genesets=gsets,cores=2)
-  gsig_up3 <- rownames( subset( lrmres, p.adjustANOVA < 0.05 & s.dist > 0 ) )
-  gsig_dn3 <- rownames( subset( lrmres, p.adjustANOVA < 0.05 & s.dist < 0 ) )
-  gtup <- names(sim[[6]])
-  gtdn <- names(sim[[7]])
-  UPTP=length(intersect(gsig_up3 ,gtup))
-  UPFP=length(setdiff(gsig_up3 ,gtup))
-  UPFN=length(setdiff(gtup,gsig_up3))
-  DNTP=length(intersect(gsig_dn3 ,gtdn))
-  DNFP=length(setdiff(gsig_dn3 ,gtdn))
-  DNFN=length(setdiff(gtdn,gsig_dn3))
-  TP=UPTP+DNTP
-  FP=UPFP+DNFP
-  FN=UPFN+DNFN
-  TN=nrow(lrmres)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
-  PREC=TP/(TP+FP)
-  REC=TP/(TP+FN)
-  F1=TP/(TP+(0.5*(FP+FN)))
-  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
-  return(result)
-}
-
-```
-
-## AL: Aggregate limma
-
-Functions for aggregate-limma-enrich approach.
-
-```{r,al}
-
-# AL approach parametric competitive
-simalc <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
-  # generate gene sets
-  gene_catalog <- unique(gt$gene)
-  lengths <- unname(unlist(lapply(genesetdatabase,length)))
-  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
-  # select gene sets to alter
-  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
-  # incorporate select changes
-  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
-    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta,
-    gene_catalog=gene_catalog)
-  # set up limma
-  mval2 <- sim$mval
-  ncols <- ncol(mval2)
-  groupsize <- ncols/2
-  ss <- data.frame(colnames(mval2))
-  colnames(ss) <- "sample"
-  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
-  d <- model.matrix(~ ss$case )
-  # al pipeline
-  mm <- merge(mval2,gt,by.x=0,by.y="probe")
-  mm$Row.names = NULL
-  a <- aggregate(. ~ gene,mm,mean)
-  rownames(a) <- a$gene
-  a$gene=NULL
-  fit.reduced <- lmFit(a,d)
-  fit.reduced <- eBayes(fit.reduced)
-  al <- topTable(fit.reduced,coef=ncol(d), number = Inf)
-  m1 <- as.data.frame(al$t)
-  rownames(m1) <- rownames(al)
-  colnames(m1) <- "t"
-  alres1 <- ttenrich(m=m1,genesets=gsets,cores=2,testtype="competitive")
-  # summarise results
-  gsig_up3 <- rownames(subset(alres1, fdr < 0.05 & t_mean > 0))
-  gsig_dn3 <- rownames(subset(alres1, fdr < 0.05 & t_mean < 0))
-  gtup <- names(sim[[6]])
-  gtdn <- names(sim[[7]])
-  UPTP=length(intersect(gsig_up3 ,gtup))
-  UPFP=length(setdiff(gsig_up3 ,gtup))
-  UPFN=length(setdiff(gtup,gsig_up3))
-  DNTP=length(intersect(gsig_dn3 ,gtdn))
-  DNFP=length(setdiff(gsig_dn3 ,gtdn))
-  DNFN=length(setdiff(gtdn,gsig_dn3))
-  TP=UPTP+DNTP
-  FP=UPFP+DNFP
-  FN=UPFN+DNFN
-  TN=nrow(alres1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
-  PREC=TP/(TP+FP)
-  REC=TP/(TP+FN)
-  F1=TP/(TP+(0.5*(FP+FN)))
-  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
-  return(result)
-}
-
-# AL approach nonparametric competitive
-simnalc <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
-  # generate gene sets
-  gene_catalog <- unique(gt$gene)
-  lengths <- unname(unlist(lapply(genesetdatabase,length)))
-  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
-  # select gene sets to alter
-  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
-  # incorporate select changes
-  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
-    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta,
-    gene_catalog=gene_catalog)
-  # set up limma
-  mval2 <- sim$mval
-  ncols <- ncol(mval2)
-  groupsize <- ncols/2
-  ss <- data.frame(colnames(mval2))
-  colnames(ss) <- "sample"
-  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
-  d <- model.matrix(~ ss$case )
-  # al pipeline
-  mm <- merge(mval2,gt,by.x=0,by.y="probe")
-  mm$Row.names = NULL
-  a <- aggregate(. ~ gene,mm,mean)
-  rownames(a) <- a$gene
-  a$gene=NULL
-  fit.reduced <- lmFit(a,d)
-  fit.reduced <- eBayes(fit.reduced)
-  al <- topTable(fit.reduced,coef=ncol(d), number = Inf)
-  m1 <- as.data.frame(al$t)
-  rownames(m1) <- rownames(al)
-  colnames(m1) <- "t"
-  alres1 <- wtenrich(m=m1,genesets=gsets,cores=2,testtype="competitive")
-  # summarise results
-  gsig_up3 <- rownames(subset(alres1, fdr < 0.05 & t_mean > 0))
-  gsig_dn3 <- rownames(subset(alres1, fdr < 0.05 & t_mean < 0))
-  gtup <- names(sim[[6]])
-  gtdn <- names(sim[[7]])
-  UPTP=length(intersect(gsig_up3 ,gtup))
-  UPFP=length(setdiff(gsig_up3 ,gtup))
-  UPFN=length(setdiff(gtup,gsig_up3))
-  DNTP=length(intersect(gsig_dn3 ,gtdn))
-  DNFP=length(setdiff(gsig_dn3 ,gtdn))
-  DNFN=length(setdiff(gtdn,gsig_dn3))
-  TP=UPTP+DNTP
-  FP=UPFP+DNFP
-  FN=UPFN+DNFN
-  TN=nrow(alres1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
-  PREC=TP/(TP+FP)
-  REC=TP/(TP+FN)
-  F1=TP/(TP+(0.5*(FP+FN)))
-  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
-  return(result)
-}
-
-```
-
-## Agg-limma-mitch function
-
-This approach uses the aggregated mvals, limma and instead of a 1-sample t-test it uses mitch
-which is a competitive test and could give more interpretable results.
-
-```{r,alm}
-
-runmitch <- function(m,genesets,cores=1) {
-  suppressMessages({ mres <- mitch_calc(m,genesets,minsetsize=5,cores=cores) })
-  mres <- mres$enrichment_result
-  rownames(mres) <- mres$set
-  mres$set=NULL
-  return(mres)
-}
-
-simalm <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
-  # generate gene sets
-  gene_catalog <- unique(gt$gene)
-  lengths <- unname(unlist(lapply(genesetdatabase,length)))
-  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
-  # select gene sets to alter
-  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
-  # incorporate select changes
-  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
-    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta,
-    gene_catalog=gene_catalog)
-  # set up limma
-  mval2 <- sim$mval
-  ncols <- ncol(mval2)
-  groupsize <- ncols/2
-  ss <- data.frame(colnames(mval2))
-  colnames(ss) <- "sample"
-  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
-  d <- model.matrix(~ ss$case )
-  # alm pipeline
-  mm <- merge(mval2,gt,by.x=0,by.y="probe")
-  mm$Row.names = NULL
-  a <- aggregate(. ~ gene,mm,mean)
-  rownames(a) <- a$gene
-  a$gene=NULL
-  fit.reduced <- lmFit(a,d)
-  fit.reduced <- eBayes(fit.reduced)
-  al <- topTable(fit.reduced,coef=ncol(d), number = Inf)
-  m1 <- as.data.frame(al$t)
-  rownames(m1) <- rownames(al)
-  colnames(m1) <- "t"
-  almres1 <- runmitch(m=m1,genesets=gsets,cores=2)
-  # summarise results
-  gsig_up3 <- rownames( subset( almres1, p.adjustANOVA < 0.05 & s.dist > 0 ) )
-  gsig_dn3 <- rownames( subset( almres1, p.adjustANOVA < 0.05 & s.dist < 0 ) )
-  gtup <- names(sim[[6]])
-  gtdn <- names(sim[[7]])
-  UPTP=length(intersect(gsig_up3 ,gtup))
-  UPFP=length(setdiff(gsig_up3 ,gtup))
-  UPFN=length(setdiff(gtup,gsig_up3))
-  DNTP=length(intersect(gsig_dn3 ,gtdn))
-  DNFP=length(setdiff(gsig_dn3 ,gtdn))
-  DNFN=length(setdiff(gtdn,gsig_dn3))
-  TP=UPTP+DNTP
-  FP=UPFP+DNFP
-  FN=UPFN+DNFN
-  TN=nrow(almres1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
-  PREC=TP/(TP+FP)
-  REC=TP/(TP+FN)
-  F1=TP/(TP+(0.5*(FP+FN)))
-  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
-  return(result)
-}
-
-```
-
-## AA Aggregate-aggregate-limma functions
-
-Use mean value works well here.
-
-```{r,aafunc}
-
-gsagg <- function(x,genesets,cores=1) {
-  meds <- mclapply(1:length(genesets), function(i) {
-    gs = genesets[[i]]
-    xx <- x[rownames(x) %in% gs,]
-    med <- apply(xx,2,mean)
-  },mc.cores=cores)
-  mymed <- do.call(rbind,meds)
-  rownames(mymed) <- names(genesets)
-  as.data.frame(mymed)
-}
-
-aalimma <- function(agag,design) {
-  fit.reduced <- lmFit(agag,design)
-  fit.reduced <- eBayes(fit.reduced)
-  dmagg <- topTable(fit.reduced,coef=ncol(design), number = Inf)
-  return(dmagg)
-}
-
-aal <- function(mval,myann,genesets,design,cores=1) {
-  medf <- chragg(mval,myann,cores=cores)
-  agag <- gsagg(x=medf,genesets=genesets,cores=cores)
-  aalres <- aalimma(agag=agag,design=design)
-  return(aalres)
-}
-
-simaa <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
-  # generate gene sets
-  gene_catalog <- unique(gt$gene)
-  lengths <- unname(unlist(lapply(genesetdatabase,length)))
-  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
-  # select gene sets to alter
-  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
-  # incorporate select changes
-  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
-    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta,
-    gene_catalog=gene_catalog)
-  # set up limma
-  mval2 <- sim$mval
-  ncols <- ncol(mval2)
-  groupsize <- ncols/2
-  ss <- data.frame(colnames(mval2))
-  colnames(ss) <- "sample"
-  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
-  d <- model.matrix(~ ss$case )
-  mm <- merge(mval2,gt,by.x=0,by.y="probe")
-  mm$Row.names = NULL
-  a <- aggregate(. ~ gene,mm,mean)
-  rownames(a) <- a$gene
-  a$gene=NULL
-  mystack <- stack(gsets)
-  mmm <- merge(a,mystack,by.x=0,by.y="values")
-  mmm$Row.names=NULL
-  aa <- aggregate(. ~ ind,mmm,mean)
-  rownames(aa) <- aa$ind
-  aa$ind=NULL
-  fit.reduced <- lmFit(aa,d)
-  fit.reduced <- eBayes(fit.reduced)
-  aares1 <- topTable(fit.reduced,coef=ncol(d), number = Inf)
-  # summarise results
-  gsig_up3 <- rownames(subset(aares1, adj.P.Val < 0.05 & logFC > 0))
-  gsig_dn3 <- rownames(subset(aares1, adj.P.Val < 0.05 & logFC < 0))
-  gtup <- names(sim[[6]])
-  gtdn <- names(sim[[7]])
-  UPTP=length(intersect(gsig_up3 ,gtup))
-  UPFP=length(setdiff(gsig_up3 ,gtup))
-  UPFN=length(setdiff(gtup,gsig_up3))
-  DNTP=length(intersect(gsig_dn3 ,gtdn))
-  DNFP=length(setdiff(gsig_dn3 ,gtdn))
-  DNFN=length(setdiff(gtdn,gsig_dn3))
-  TP=UPTP+DNTP
-  FP=UPFP+DNFP
-  FN=UPFN+DNFN
-  TN=nrow(aares1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
-  PREC=TP/(TP+FP)
-  REC=TP/(TP+FN)
-  F1=TP/(TP+(0.5*(FP+FN)))
-  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
-  return(result)
-}
-
-F1 <- function(x,y) {
-  ( 2 * x * y ) / ( x + y )
-}
-
 ```
@@
 
-save.image("GSE158422_simulate100.Rdata")
+save.image("GSE158422_simulate020.Rdata")
 

### 2024-01-10 figs/GSE158422_simulate020_rslurm.Rmd
got rslurm working now

@@
 deltas=c(0.1,0.2,0.3,0.4,0.5)
-
-#groupsizes=5
-#deltas=0.4
+#groupsizes=12
+#deltas=0.5
 
@@
 
-params %>% kbl(caption="Parameters to test") %>% kable_paper("hover", full_width = F)
+str(params)
+
+params <- params[order(params$groupsize),]
 
@@
 
-```{r,rungsa,eval=FALSE}
+```{r,rungsa}
 
-sopt1 <- list(time = '1:00:00', mem="10G")
+sopt1 <- list(time = '6:00:00', mem="16G")
 
 sjob <- slurm_apply(simgsa, params, jobname = 'gsa',
-  nodes = 50, cpus_per_node = 2, submit = TRUE, slurm_options = sopt1,
+  nodes = 50, cpus_per_node = 1, submit = TRUE, slurm_options = sopt1,
   genesetdatabase=gsets, myann=myann, mval=normal_mval, frac_genes=0.5,
@@
 
-```{r,runlat,eval=FALSE}
+```{r,runlat}
 
-sopt1 <- list(time = '1:00:00', mem="16G")
+sopt1 <- list(time = '6:00:00', mem="32G")
 
@@
 
-```{r,runltt,eval=TRUE}
+```{r,runltt}
 
-sopt1 <- list(time = '1:00:00', mem="16G")
+sopt1 <- list(time = '6:00:00', mem="32G")
 
@@
 
-```{r,runlaw,eval=TRUE}
+```{r,runlaw}
 
-sopt1 <- list(time = '1:00:00', mem="16G")
+sopt1 <- list(time = '6:00:00', mem="32G")
 
@@
 
-```{r,runlam,eval=TRUE}
+```{r,runlam}
 
-sopt1 <- list(time = '1:00:00', mem="16G")
+sopt1 <- list(time = '6:00:00', mem="32G")
 
@@
 
-```{r,runlrm,eval=TRUE}
+```{r,runlrm}
 
-sopt1 <- list(time = '1:00:00', mem="16G")
+sopt1 <- list(time = '6:00:00', mem="32G")
 
@@
 
-```{r,runalt,eval=TRUE}
+```{r,runalt}
 
-sopt1 <- list(time = '1:00:00', mem="16G")
+sopt1 <- list(time = '6:00:00', mem="32G")
 
@@
 
-```{r,runalw,eval=TRUE}
+```{r,runalw}
 
-sopt1 <- list(time = '1:00:00', mem="16G")
+sopt1 <- list(time = '6:00:00', mem="32G")
 
@@
 
-```{r,runalm,eval=TRUE}
+```{r,runalm}
 
-sopt1 <- list(time = '1:00:00', mem="16G")
+sopt1 <- list(time = '6:00:00', mem="32G")
 

### 2024-01-12 figs/GSE158422_simulate_rslurm_summary.Rmd
added HPC results

---
title: "Testing gene methylation enrichment analysis approaches using simulated data - summarise data"
author: "The GMEA team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
---

## Introduction

Here is an explanation of the different objects:

0. GSA: gres

1. LAT: lacres

2. LTT: lactopres

3. LAW: nlacres

4. LAM: lacmres

5. LRM: lrmres

6. ALT: alcres

7. ALW: nalcres

8. ALM: almres

```{r,packages}

suppressPackageStartupMessages({
  library("kableExtra")
})

```

## Functions

```{r,functions}

F1 <- function(x,y) {
  ( 2 * x * y ) / ( x + y )
}

PRECREC <- function(x){
  y <- aggregate(.~groupsize+delta,x,mean)
  y$PREC <- y$TP / ( y$TP + y$FP )
  y$REC <- y$TP / ( y$TP + y$FN )
  y$F1 <- F1(y$PREC,y$REC)
  y
}
# usage: lapply(x20,PRECREC)

smrz <- function(x) {
  P <- mean(x$PREC)
  R <- mean(x$REC)
  F <- F1(P,R)
  c(P,R,F)
}
# usage: smrz(x20a)

```

## Load data

```{r,load}

load("GSE158422_simulate020_rslurm.Rdata")

x20 <- list("GSA"=gres,"LAT"=latres,"LTT"=lttres,"LAW"=lawres,
  "LAM"=lamres,"LRM"=lrmres,"ALT"=altres,"ALW"=alwres,"ALM"=almres)

load("GSE158422_simulate050_rslurm.Rdata")

x50 <- list("GSA"=gres,"LAT"=latres,"LTT"=lttres,"LAW"=lawres,
  "LAM"=lamres,"LRM"=lrmres,"ALT"=altres,"ALW"=alwres,"ALM"=almres)

load("GSE158422_simulate100_rslurm.Rdata")

x100 <- list("GSA"=gres,"LAT"=latres,"LTT"=lttres,"LAW"=lawres,
  "LAM"=lamres,"LRM"=lrmres,"ALT"=altres,"ALW"=alwres,"ALM"=almres)

rm(myann)
rm(gp2)
rm(normal_mval)
rm(mval,anno)

```

## Overall summary

Merge the data at gene set sizes of 20, 50 and 100 into a single result.

```{r,merge,fig.width=5,fig.height=4}

all <- lapply(1:length(x20),function(i) {
  mg <- rbind(x20[[i]],x50[[i]],x100[[i]])
  m <- apply(mg,2,mean)
  P <- m["TP"] / ( m["TP"] + m["FP"])
  R <- m["TP"] / ( m["TP"] + m["FN"] )
  F <- F1(P,R)
  res <- c(P,R,F)
  names(res) <- c("P","R","F")
  return(res)
} )
names(all) <- names(x20)
all <- do.call(rbind,all)
colnames(all) <- c("PREC","REC","F1")

all %>% kbl(caption="Overall simulation results") %>% kable_paper("hover", full_width = F)

par(mar = c(5.1, 4.1, 5.1, 2.1))
cols <- c("white","gray","black")

barplot(t(all),beside=TRUE,ylim=c(0,1),ylab="index",
  col=cols, cex.names=0.7, cex.axis=0.7, main="Overall simulation results")
grid()
legend("right", inset=.01, c("Precision","Recall","F1"), fill=cols, cex=0.7)

```

## Focus on differences caused byu gene set size

Make a set of three barplots which summarise method performance over the range of group size and
delta values, for gene set sizes of 20, 50 and 200.

```{r,x20summary,fig.width=5,fig.height=4}

groupsizes=unique(x20[[1]]$groupsize)
deltas=unique(x20[[1]]$delta)

x20a <- lapply(x20,PRECREC)
x20aa <- lapply(x20a,smrz)
x20aa <- do.call(rbind,x20aa)
colnames(x20aa) <- c("PREC","REC","F1")
x50a <- lapply(x50,PRECREC)
x50aa <- lapply(x50a,smrz)
x50aa <- do.call(rbind,x50aa)
colnames(x50aa) <- c("PREC","REC","F1")
x100a <- lapply(x100,PRECREC)
x100aa <- lapply(x100a,smrz)
x100aa <- do.call(rbind,x100aa)
colnames(x100aa) <- c("PREC","REC","F1")

x20aa %>% kbl(caption="sim 20 res") %>% kable_paper("hover", full_width = F)
x50aa %>% kbl(caption="sim 50 res") %>% kable_paper("hover", full_width = F)
x100aa %>% kbl(caption="sim 100 res") %>% kable_paper("hover", full_width = F)

barplot(t(x20aa),beside=TRUE,ylim=c(0,1),ylab="index",
  col=cols, cex.names=0.8, cex.axis=0.8, main="20 genes per set")
abline(h=seq(0.2,1,0.2),lty=2,col="gray")
legend("right", inset=.01, c("Precision","Recall","F1"), fill=cols, bg="white", cex=0.8)
barplot(t(x50aa),beside=TRUE,ylim=c(0,1),ylab="index",
  col=cols, cex.names=0.8, cex.axis=0.8, main="50 genes per set")
abline(h=seq(0.2,1,0.2),lty=2,col="gray")
legend("right", inset=.01, c("Precision","Recall","F1"), fill=cols, bg="white", cex=0.8)
barplot(t(x100aa),beside=TRUE,ylim=c(0,1),ylab="index",
  col=cols, cex.names=0.8, cex.axis=0.8, main="100 genes per set")
abline(h=seq(0.2,1,0.2),lty=2,col="gray")
legend("bottomright", inset=.01, c("Precision","Recall","F1"), fill=cols, bg="white", cex=0.8)
dev.off()

pdf("fig1_slurm.pdf",height=6,width=7)
par(mfrow=c(2,2))
par(mar = c(2.5, 4, 3, 1))
barplot(t(all),beside=TRUE,ylim=c(0,1),ylab="index",
  col=cols, cex.names=0.65, cex.axis=0.8, main="Overall simulation results")
abline(h=seq(0.2,1,0.2),lty=2,col="gray")
legend("right", inset=0, c("Precision","Recall","F1"), fill=cols, bg="white", cex=0.8)
mtext("A", 2, adj=2.5, las=1,padj=-4,cex=2)
barplot(t(x20aa),beside=TRUE,ylim=c(0,1),ylab="index",
  col=cols, cex.names=0.65, cex.axis=0.8, main="20 genes per set")
abline(h=seq(0.2,1,0.2),lty=2,col="gray")
legend("right", inset=0, c("Precision","Recall","F1"), fill=cols, bg="white", cex=0.8)
mtext("B", 2, adj=2.5, las=1,padj=-4,cex=2)
barplot(t(x50aa),beside=TRUE,ylim=c(0,1),ylab="index",
  col=cols, cex.names=0.65, cex.axis=0.8, main="50 genes per set")
abline(h=seq(0.2,1,0.2),lty=2,col="gray")
legend("right", inset=0, c("Precision","Recall","F1"), fill=cols, bg="white", cex=0.8)
mtext("C", 2, adj=2.5, las=1,padj=-4,cex=2)
barplot(t(x100aa),beside=TRUE,ylim=c(0,1),ylab="index",
  col=cols, cex.names=0.65, cex.axis=0.8, main="100 genes per set")
abline(h=seq(0.2,1,0.2),lty=2,col="gray")
legend("bottomright", inset=0, c("Precision","Recall","F1"), fill=cols, bg="white", cex=0.8)
mtext("D", 2, adj=2.5, las=1,padj=-4,cex=2)
dev.off()

```

## In depth analysis of GSA vs LAM

As GSA appears to be the best-performing existing method for enrichment analysis, it is a good idea
to compare it to LAM which is the best one out of the new methods evaluated here.
The following charts show the performance of GSA and LAM including precision and recall over the
range of group sizes and delta values when the sets contain 20, 50 and 100 genes each.

First with 20 genes per set.

```{r,020_gsa_vs_lam}

par(mfrow=c(2,3))
par(mar = c(4.1, 4.1, 3.5, 0.9))

gsa <- lapply(deltas,function(d) { subset(x20a$GSA,delta==d) } )
lam <- lapply(deltas,function(d) { subset(x20a$LAM,delta==d) } )

plot(gsa[[length(gsa)]][,"groupsize"],  gsa[[length(gsa)]][,"PREC"],type="b",las=1,ylab="precision",xlab="",
  main="",ylim=c(0,1))
lines(gsa[[length(gsa)-1]][,"groupsize"], gsa[[length(gsa)-1]][,"PREC"] ,type="b",col="red")
lines(gsa[[length(gsa)-2]][,"groupsize"], gsa[[length(gsa)-2]][,"PREC"] ,type="b",col="blue")
lines(gsa[[length(gsa)-3]][,"groupsize"], gsa[[length(gsa)-3]][,"PREC"] ,type="b",col="darkgreen")
lines(gsa[[length(gsa)-4]][,"groupsize"], gsa[[length(gsa)-4]][,"PREC"] ,type="b",col="darkorange")
grid()
legend("bottomright", legend=c("0.5", "0.4", "0.3", "0.2", "0.1"),title="delta",
  lty=1, lwd=2, col=c("black", "red", "blue", "darkgreen", "darkorange"), cex=1, bg="white")

plot(gsa[[length(gsa)]][,"groupsize"],  gsa[[length(gsa)]][,"REC"],type="b",las=1,ylab="recall",xlab="group size",
  main="GSA",ylim=c(0,1))
lines(gsa[[length(gsa)-1]][,"groupsize"], gsa[[length(gsa)-1]][,"REC"] ,type="b",col="red")
lines(gsa[[length(gsa)-2]][,"groupsize"], gsa[[length(gsa)-2]][,"REC"] ,type="b",col="blue")
lines(gsa[[length(gsa)-3]][,"groupsize"], gsa[[length(gsa)-3]][,"REC"] ,type="b",col="darkgreen")
lines(gsa[[length(gsa)-4]][,"groupsize"], gsa[[length(gsa)-4]][,"REC"] ,type="b",col="darkorange")
grid()

plot(gsa[[length(gsa)]][,"groupsize"],  gsa[[length(gsa)]][,"F1"],type="b",las=1,ylab="F1",xlab="",
  main="",ylim=c(0,1))
lines(gsa[[length(gsa)-1]][,"groupsize"], gsa[[length(gsa)-1]][,"F1"] ,type="b",col="red")
lines(gsa[[length(gsa)-2]][,"groupsize"], gsa[[length(gsa)-2]][,"F1"] ,type="b",col="blue")
lines(gsa[[length(gsa)-3]][,"groupsize"], gsa[[length(gsa)-3]][,"F1"] ,type="b",col="darkgreen")
lines(gsa[[length(gsa)-4]][,"groupsize"], gsa[[length(gsa)-4]][,"F1"] ,type="b",col="darkorange")
grid()

plot(lam[[length(lam)]][,"groupsize"],  lam[[length(lam)]][,"PREC"],type="b",las=1,ylab="precision",xlab="",
  main="",ylim=c(0,1))
lines(lam[[length(lam)-1]][,"groupsize"], lam[[length(lam)-1]][,"PREC"] ,type="b",col="red")
lines(lam[[length(lam)-2]][,"groupsize"], lam[[length(lam)-2]][,"PREC"] ,type="b",col="blue")
lines(lam[[length(lam)-3]][,"groupsize"], lam[[length(lam)-3]][,"PREC"] ,type="b",col="darkgreen")
lines(lam[[length(lam)-4]][,"groupsize"], lam[[length(lam)-4]][,"PREC"] ,type="b",col="darkorange")
grid()
legend("bottomright", legend=c("0.5", "0.4", "0.3", "0.2", "0.1"),title="delta",
  lty=1, lwd=2, col=c("black", "red", "blue", "darkgreen", "darkorange"), cex=1, bg="white")

plot(lam[[length(lam)]][,"groupsize"],  lam[[length(lam)]][,"REC"],type="b",las=1,ylab="recall",xlab="group size",
  main="LAM",ylim=c(0,1))
lines(lam[[length(lam)-1]][,"groupsize"], lam[[length(lam)-1]][,"REC"] ,type="b",col="red")
lines(lam[[length(lam)-2]][,"groupsize"], lam[[length(lam)-2]][,"REC"] ,type="b",col="blue")
lines(lam[[length(lam)-3]][,"groupsize"], lam[[length(lam)-3]][,"REC"] ,type="b",col="darkgreen")
lines(lam[[length(lam)-4]][,"groupsize"], lam[[length(lam)-4]][,"REC"] ,type="b",col="darkorange")
grid()

plot(lam[[length(lam)]][,"groupsize"],  lam[[length(lam)]][,"F1"],type="b",las=1,ylab="F1",xlab="",
  main="",ylim=c(0,1))
lines(lam[[length(lam)-1]][,"groupsize"], lam[[length(lam)-1]][,"F1"] ,type="b",col="red")
lines(lam[[length(lam)-2]][,"groupsize"], lam[[length(lam)-2]][,"F1"] ,type="b",col="blue")
lines(lam[[length(lam)-3]][,"groupsize"], lam[[length(lam)-3]][,"F1"] ,type="b",col="darkgreen")
lines(lam[[length(lam)-4]][,"groupsize"], lam[[length(lam)-4]][,"F1"] ,type="b",col="darkorange")
grid()

```

Now with 50 genes per set.

```{r,050_gsa_vs_lam}

par(mfrow=c(2,3))
par(mar = c(4.1, 4.1, 3.5, 0.9))

gsa <- lapply(deltas,function(d) { subset(x50a$GSA,delta==d) } )
lam <- lapply(deltas,function(d) { subset(x50a$LAM,delta==d) } )

plot(gsa[[length(gsa)]][,"groupsize"],  gsa[[length(gsa)]][,"PREC"],type="b",las=1,ylab="precision",xlab="",
  main="",ylim=c(0,1))
lines(gsa[[length(gsa)-1]][,"groupsize"], gsa[[length(gsa)-1]][,"PREC"] ,type="b",col="red")
lines(gsa[[length(gsa)-2]][,"groupsize"], gsa[[length(gsa)-2]][,"PREC"] ,type="b",col="blue")
lines(gsa[[length(gsa)-3]][,"groupsize"], gsa[[length(gsa)-3]][,"PREC"] ,type="b",col="darkgreen")
lines(gsa[[length(gsa)-4]][,"groupsize"], gsa[[length(gsa)-4]][,"PREC"] ,type="b",col="darkorange")
grid()
legend("bottomright", legend=c("0.5", "0.4", "0.3", "0.2", "0.1"),title="delta",
  lty=1, lwd=2, col=c("black", "red", "blue", "darkgreen", "darkorange"), cex=1, bg="white")

plot(gsa[[length(gsa)]][,"groupsize"],  gsa[[length(gsa)]][,"REC"],type="b",las=1,ylab="recall",xlab="group size",
  main="GSA",ylim=c(0,1))
lines(gsa[[length(gsa)-1]][,"groupsize"], gsa[[length(gsa)-1]][,"REC"] ,type="b",col="red")
lines(gsa[[length(gsa)-2]][,"groupsize"], gsa[[length(gsa)-2]][,"REC"] ,type="b",col="blue")
lines(gsa[[length(gsa)-3]][,"groupsize"], gsa[[length(gsa)-3]][,"REC"] ,type="b",col="darkgreen")
lines(gsa[[length(gsa)-4]][,"groupsize"], gsa[[length(gsa)-4]][,"REC"] ,type="b",col="darkorange")
grid()

plot(gsa[[length(gsa)]][,"groupsize"],  gsa[[length(gsa)]][,"F1"],type="b",las=1,ylab="F1",xlab="",
  main="",ylim=c(0,1))
lines(gsa[[length(gsa)-1]][,"groupsize"], gsa[[length(gsa)-1]][,"F1"] ,type="b",col="red")
lines(gsa[[length(gsa)-2]][,"groupsize"], gsa[[length(gsa)-2]][,"F1"] ,type="b",col="blue")
lines(gsa[[length(gsa)-3]][,"groupsize"], gsa[[length(gsa)-3]][,"F1"] ,type="b",col="darkgreen")
lines(gsa[[length(gsa)-4]][,"groupsize"], gsa[[length(gsa)-4]][,"F1"] ,type="b",col="darkorange")
grid()

plot(lam[[length(lam)]][,"groupsize"],  lam[[length(lam)]][,"PREC"],type="b",las=1,ylab="precision",xlab="",
  main="",ylim=c(0,1))
lines(lam[[length(lam)-1]][,"groupsize"], lam[[length(lam)-1]][,"PREC"] ,type="b",col="red")
lines(lam[[length(lam)-2]][,"groupsize"], lam[[length(lam)-2]][,"PREC"] ,type="b",col="blue")
lines(lam[[length(lam)-3]][,"groupsize"], lam[[length(lam)-3]][,"PREC"] ,type="b",col="darkgreen")
lines(lam[[length(lam)-4]][,"groupsize"], lam[[length(lam)-4]][,"PREC"] ,type="b",col="darkorange")
grid()
legend("bottomright", legend=c("0.5", "0.4", "0.3", "0.2", "0.1"),title="delta",
  lty=1, lwd=2, col=c("black", "red", "blue", "darkgreen", "darkorange"), cex=1, bg="white")

plot(lam[[length(lam)]][,"groupsize"],  lam[[length(lam)]][,"REC"],type="b",las=1,ylab="recall",xlab="group size",
  main="LAM",ylim=c(0,1))
lines(lam[[length(lam)-1]][,"groupsize"], lam[[length(lam)-1]][,"REC"] ,type="b",col="red")
lines(lam[[length(lam)-2]][,"groupsize"], lam[[length(lam)-2]][,"REC"] ,type="b",col="blue")
lines(lam[[length(lam)-3]][,"groupsize"], lam[[length(lam)-3]][,"REC"] ,type="b",col="darkgreen")
lines(lam[[length(lam)-4]][,"groupsize"], lam[[length(lam)-4]][,"REC"] ,type="b",col="darkorange")
grid()

plot(lam[[length(lam)]][,"groupsize"],  lam[[length(lam)]][,"F1"],type="b",las=1,ylab="F1",xlab="",
  main="",ylim=c(0,1))
lines(lam[[length(lam)-1]][,"groupsize"], lam[[length(lam)-1]][,"F1"] ,type="b",col="red")
lines(lam[[length(lam)-2]][,"groupsize"], lam[[length(lam)-2]][,"F1"] ,type="b",col="blue")
lines(lam[[length(lam)-3]][,"groupsize"], lam[[length(lam)-3]][,"F1"] ,type="b",col="darkgreen")
lines(lam[[length(lam)-4]][,"groupsize"], lam[[length(lam)-4]][,"F1"] ,type="b",col="darkorange")
grid()

```

Now 100 genes per set

```{r,100_gsa_vs_lam}

par(mfrow=c(2,3))
par(mar = c(4.1, 4.1, 3.5, 0.9))

gsa <- lapply(deltas,function(d) { subset(x100a$GSA,delta==d) } )
lam <- lapply(deltas,function(d) { subset(x100a$LAM,delta==d) } )

plot(gsa[[length(gsa)]][,"groupsize"],  gsa[[length(gsa)]][,"PREC"],type="b",las=1,ylab="precision",xlab="",
  main="",ylim=c(0,1))
lines(gsa[[length(gsa)-1]][,"groupsize"], gsa[[length(gsa)-1]][,"PREC"] ,type="b",col="red")
lines(gsa[[length(gsa)-2]][,"groupsize"], gsa[[length(gsa)-2]][,"PREC"] ,type="b",col="blue")
lines(gsa[[length(gsa)-3]][,"groupsize"], gsa[[length(gsa)-3]][,"PREC"] ,type="b",col="darkgreen")
lines(gsa[[length(gsa)-4]][,"groupsize"], gsa[[length(gsa)-4]][,"PREC"] ,type="b",col="darkorange")
grid()
legend("bottomright", legend=c("0.5", "0.4", "0.3", "0.2", "0.1"),title="delta",
  lty=1, lwd=2, col=c("black", "red", "blue", "darkgreen", "darkorange"), cex=1, bg="white")

plot(gsa[[length(gsa)]][,"groupsize"],  gsa[[length(gsa)]][,"REC"],type="b",las=1,ylab="recall",xlab="group size",
  main="GSA",ylim=c(0,1))
lines(gsa[[length(gsa)-1]][,"groupsize"], gsa[[length(gsa)-1]][,"REC"] ,type="b",col="red")
lines(gsa[[length(gsa)-2]][,"groupsize"], gsa[[length(gsa)-2]][,"REC"] ,type="b",col="blue")
lines(gsa[[length(gsa)-3]][,"groupsize"], gsa[[length(gsa)-3]][,"REC"] ,type="b",col="darkgreen")
lines(gsa[[length(gsa)-4]][,"groupsize"], gsa[[length(gsa)-4]][,"REC"] ,type="b",col="darkorange")
grid()

plot(gsa[[length(gsa)]][,"groupsize"],  gsa[[length(gsa)]][,"F1"],type="b",las=1,ylab="F1",xlab="",
  main="",ylim=c(0,1))
lines(gsa[[length(gsa)-1]][,"groupsize"], gsa[[length(gsa)-1]][,"F1"] ,type="b",col="red")
lines(gsa[[length(gsa)-2]][,"groupsize"], gsa[[length(gsa)-2]][,"F1"] ,type="b",col="blue")
lines(gsa[[length(gsa)-3]][,"groupsize"], gsa[[length(gsa)-3]][,"F1"] ,type="b",col="darkgreen")
lines(gsa[[length(gsa)-4]][,"groupsize"], gsa[[length(gsa)-4]][,"F1"] ,type="b",col="darkorange")
grid()

plot(lam[[length(lam)]][,"groupsize"],  lam[[length(lam)]][,"PREC"],type="b",las=1,ylab="precision",xlab="",
  main="",ylim=c(0,1))
lines(lam[[length(lam)-1]][,"groupsize"], lam[[length(lam)-1]][,"PREC"] ,type="b",col="red")
lines(lam[[length(lam)-2]][,"groupsize"], lam[[length(lam)-2]][,"PREC"] ,type="b",col="blue")
lines(lam[[length(lam)-3]][,"groupsize"], lam[[length(lam)-3]][,"PREC"] ,type="b",col="darkgreen")
lines(lam[[length(lam)-4]][,"groupsize"], lam[[length(lam)-4]][,"PREC"] ,type="b",col="darkorange")
grid()
legend("bottomright", legend=c("0.5", "0.4", "0.3", "0.2", "0.1"),title="delta",
  lty=1, lwd=2, col=c("black", "red", "blue", "darkgreen", "darkorange"), cex=1, bg="white")

plot(lam[[length(lam)]][,"groupsize"],  lam[[length(lam)]][,"REC"],type="b",las=1,ylab="recall",xlab="group size",
  main="LAM",ylim=c(0,1))
lines(lam[[length(lam)-1]][,"groupsize"], lam[[length(lam)-1]][,"REC"] ,type="b",col="red")
lines(lam[[length(lam)-2]][,"groupsize"], lam[[length(lam)-2]][,"REC"] ,type="b",col="blue")
lines(lam[[length(lam)-3]][,"groupsize"], lam[[length(lam)-3]][,"REC"] ,type="b",col="darkgreen")
lines(lam[[length(lam)-4]][,"groupsize"], lam[[length(lam)-4]][,"REC"] ,type="b",col="darkorange")
grid()

plot(lam[[length(lam)]][,"groupsize"],  lam[[length(lam)]][,"F1"],type="b",las=1,ylab="F1",xlab="",
  main="",ylim=c(0,1))
lines(lam[[length(lam)-1]][,"groupsize"], lam[[length(lam)-1]][,"F1"] ,type="b",col="red")
lines(lam[[length(lam)-2]][,"groupsize"], lam[[length(lam)-2]][,"F1"] ,type="b",col="blue")
lines(lam[[length(lam)-3]][,"groupsize"], lam[[length(lam)-3]][,"F1"] ,type="b",col="darkgreen")
lines(lam[[length(lam)-4]][,"groupsize"], lam[[length(lam)-4]][,"F1"] ,type="b",col="darkorange")
grid()

## PDF output

pdf("fig2_slurm.pdf",height=6,width=7)
par(mfrow=c(2,3))
par(mar = c(4.1, 4.1, 3.5, 0.9))

plot(gsa[[length(gsa)]][,"groupsize"],  gsa[[length(gsa)]][,"PREC"],type="b",las=1,ylab="precision",xlab="",
  main="",ylim=c(0,1))
lines(gsa[[length(gsa)-1]][,"groupsize"], gsa[[length(gsa)-1]][,"PREC"] ,type="b",col="red")
lines(gsa[[length(gsa)-2]][,"groupsize"], gsa[[length(gsa)-2]][,"PREC"] ,type="b",col="blue")
lines(gsa[[length(gsa)-3]][,"groupsize"], gsa[[length(gsa)-3]][,"PREC"] ,type="b",col="darkgreen")
lines(gsa[[length(gsa)-4]][,"groupsize"], gsa[[length(gsa)-4]][,"PREC"] ,type="b",col="darkorange")
grid()
legend("bottomright", legend=c("0.5", "0.4", "0.3", "0.2", "0.1"),title="delta",
  lty=1, lwd=2, col=c("black", "red", "blue", "darkgreen", "darkorange"), cex=1, bg="white")

plot(gsa[[length(gsa)]][,"groupsize"],  gsa[[length(gsa)]][,"REC"],type="b",las=1,ylab="recall",xlab="group size",
  main="GSA",ylim=c(0,1))
lines(gsa[[length(gsa)-1]][,"groupsize"], gsa[[length(gsa)-1]][,"REC"] ,type="b",col="red")
lines(gsa[[length(gsa)-2]][,"groupsize"], gsa[[length(gsa)-2]][,"REC"] ,type="b",col="blue")
lines(gsa[[length(gsa)-3]][,"groupsize"], gsa[[length(gsa)-3]][,"REC"] ,type="b",col="darkgreen")
lines(gsa[[length(gsa)-4]][,"groupsize"], gsa[[length(gsa)-4]][,"REC"] ,type="b",col="darkorange")
grid()

plot(gsa[[length(gsa)]][,"groupsize"],  gsa[[length(gsa)]][,"F1"],type="b",las=1,ylab="F1",xlab="",
  main="",ylim=c(0,1))
lines(gsa[[length(gsa)-1]][,"groupsize"], gsa[[length(gsa)-1]][,"F1"] ,type="b",col="red")
lines(gsa[[length(gsa)-2]][,"groupsize"], gsa[[length(gsa)-2]][,"F1"] ,type="b",col="blue")
lines(gsa[[length(gsa)-3]][,"groupsize"], gsa[[length(gsa)-3]][,"F1"] ,type="b",col="darkgreen")
lines(gsa[[length(gsa)-4]][,"groupsize"], gsa[[length(gsa)-4]][,"F1"] ,type="b",col="darkorange")
grid()

plot(lam[[length(lam)]][,"groupsize"],  lam[[length(lam)]][,"PREC"],type="b",las=1,ylab="precision",xlab="",
  main="",ylim=c(0,1))
lines(lam[[length(lam)-1]][,"groupsize"], lam[[length(lam)-1]][,"PREC"] ,type="b",col="red")
lines(lam[[length(lam)-2]][,"groupsize"], lam[[length(lam)-2]][,"PREC"] ,type="b",col="blue")
lines(lam[[length(lam)-3]][,"groupsize"], lam[[length(lam)-3]][,"PREC"] ,type="b",col="darkgreen")
lines(lam[[length(lam)-4]][,"groupsize"], lam[[length(lam)-4]][,"PREC"] ,type="b",col="darkorange")
grid()
legend("bottomright", legend=c("0.5", "0.4", "0.3", "0.2", "0.1"),title="delta",
  lty=1, lwd=2, col=c("black", "red", "blue", "darkgreen", "darkorange"), cex=1, bg="white")

plot(lam[[length(lam)]][,"groupsize"],  lam[[length(lam)]][,"REC"],type="b",las=1,ylab="recall",xlab="group size",
  main="LAM",ylim=c(0,1))
lines(lam[[length(lam)-1]][,"groupsize"], lam[[length(lam)-1]][,"REC"] ,type="b",col="red")
lines(lam[[length(lam)-2]][,"groupsize"], lam[[length(lam)-2]][,"REC"] ,type="b",col="blue")
lines(lam[[length(lam)-3]][,"groupsize"], lam[[length(lam)-3]][,"REC"] ,type="b",col="darkgreen")
lines(lam[[length(lam)-4]][,"groupsize"], lam[[length(lam)-4]][,"REC"] ,type="b",col="darkorange")
grid()

plot(lam[[length(lam)]][,"groupsize"],  lam[[length(lam)]][,"F1"],type="b",las=1,ylab="F1",xlab="",
  main="",ylim=c(0,1))
lines(lam[[length(lam)-1]][,"groupsize"], lam[[length(lam)-1]][,"F1"] ,type="b",col="red")
lines(lam[[length(lam)-2]][,"groupsize"], lam[[length(lam)-2]][,"F1"] ,type="b",col="blue")
lines(lam[[length(lam)-3]][,"groupsize"], lam[[length(lam)-3]][,"F1"] ,type="b",col="darkgreen")
lines(lam[[length(lam)-4]][,"groupsize"], lam[[length(lam)-4]][,"F1"] ,type="b",col="darkorange")
grid()

dev.off()

```

## Session information

```{r,save}

sessionInfo()

```

### 2024-01-12 manuscript/manuscript.Rmd
working on the ms

@@
 
-* Infinium methylation beadarrays remain one of the most popular techniques for studying genomic
-methylation, but tools for downstream pathways analysis remain limited.
+* Infinium methylation beadarrays remain one of the most popular platforms for studying genomic
+methylation, but tools for downstream pathways analysis have their limitations.
 
@@
 
-* Here we propose three different approaches to apply FCS to Infinium data, which vary upon when the
-differential methylation analysis is conducted, at the level of probes, genes or gene sets.
+* Here we evaluate several approaches to apply FCS to Infinium data retaining directional information.
+The best performing method involves the mean aggregation of probe limma t-statistics by gene followed by
+a rank-ANOVA enrichment test using the mitch package.
 
@@
 
-Simulations were conducted using the seven FCS methods (LAT, LTT, LAW, LAM, ALT, ALW, ALM) and GSAmeth,
+Simulations were conducted using the seven FCS methods (LAT, LTT, LAW, LAM, LRM, ALT, ALW, ALM) and GSAmeth,
 with group sizes of 3-12 and delta values between 0.1 and 0.5, and random gene sets with sizes 20, 50
@@
 True positives, false positives and false negatives were used to calculate overall precision and recall
-at these three gene set sizes (**Figure 1**).
-We found that overall recall was positively associated with the size of the gene sets considered.
-At 20 genes per set, GSA had the highest overall performance, with F1 of 0.097, followed by LAM with 0.035.
-At 50 genes per set, LAM achieved an F1 value of 0.29, while GSA scored 0.19.
-At 100 genes per set, LAM showed the highest F1 performance of 0.56 followed by ALM with 0.53, while
-GSA scored only 0.24.
-This result suggests that FCS methods perform better than ORA with gene set sizes >50.
+at these three gene set sizes (**Figure 1A**).
+The LAM and ALM methods recorded an overall precision of 0.94, while GSA scored lowest with 0.848.
+Recall was highest for LAM with 0.22, while ALM scored 0.21 and GSA scored 0.15
+F1 scores were highest for LAM with 0.36 followed by ALM with 0.35, while GSA scored 0.26.
 
-![**Figure 1. Precision, recall and F1 for enrichment tests with simulated data.** Gene set sizes vary between 20 and 100 genes per set.](../figs/fig1.png)
+We noticed that the size of gene sets used in the simulations strongly influenced the results, with overall recall increasing with
+gene set size (**Figure 1B-D**).
+The GSA method performed better than FCS methods with gene sets of 20 (**Figure 1B**), while at 50 genes per set GSA performance
+was on par with LAM (**Figure 1C**).
+At 100 genes per set however, LAM showed superior precision and recall compared to GSA (**Figure 1D**).
 
-Performance results for GSA and LAM methods with 100 genes per set shows that precision was similar, with most simulations yielding
->95% precision (**Figure 2**).
-LAM recall was superior for most simulations, particularly with lower group sizes.
-Taken together, this result suggests that LAM offers improved sensitivity over existing approaches when
-the gene sets are larger than 50 members.
+![**Figure 1. Precision, recall and F1 for enrichment tests with simulated data.**
+(A) Overall results including 20, 50 and 100 genes per set.
+(B-D) Results for simulations using 20, 50 and 100 genes per set respectively.
+](../figs/fig1_slurm.pdf){height="500px" width="500px"}
+
+Focusing on results from 100 genes per set, LAM precision was more consistent across the parameter ranges as compared to GSA which
+showed lower precision when group size and delta were lower (**Figure 2**).
+Recall was strongly dependent on group size and delta parameters.
+LAM recorded relatively higher recall when group size and delta were lower.
+F1 performance scores at 100 genes per set were better for LAM as compared to GSA.
 
 ![**Figure 2. Precision, recall and F1 for GSA and LAM enrichment tests with simulated data over a range of group sizes and delta values.** One hundred genes per set.](../figs/fig2.png)
+
+Overall, LAM was the superior FCS method across these tests and performed better than the existing GSA method.
 
@@
 
-![**Figure 5. Integration of gene expression data with methylation pathways.**
-(A,B) Euler diagrams showing the overlap in statistically significant pathways (FDR<0.05) from gene expression and GSA results (A) and LAM (B) results.
-(C) Multi-dimensional enrichment analysis of promoter and gene body methylation with gene expression using mitch. Top 20 gene sets shown with largest s distance after FDR filtering at 0.05.
-](../figs/fig5.png)
+#[**Figure 5. Integration of gene expression data with methylation pathways.**
+#(A,B) Euler diagrams showing the overlap in statistically significant pathways (FDR<0.05) from gene expression and GSA results (A) and LAM (B) results.
+#(C) Multi-dimensional enrichment analysis of promoter and gene body methylation with gene expression using mitch. Top 20 gene sets shown with largest s distance after FDR filtering at 0.05.
+#](../figs/fig5.png)
 
@@
 
-![**Figure 6. Pathway level DNA methylation alterations with chronological age.**
-(A) Contour heatmap showing the similarity in gene methylation score ranks in the discovery and replication studies.
-(B) Mitch pathway enrichment scores (S distances) in discovery and replication studies. Pathways with MANOVA FDR<0.05 are shown in red while others are shaded gray.
-(C) Heatmap of 30 pathways with most extreme S distances (a surrogate of effect size). Red indicates increasing methylation and blue shows lower methylation.
-(D) An example of a pathway identified with this method, "creation of C4 and C2 activators" shows lower methylation of member genes in both discovery and replication studies.
-](../figs/fig6.png)
+#![**Figure 6. Pathway level DNA methylation alterations with chronological age.**
+#(A) Contour heatmap showing the similarity in gene methylation score ranks in the discovery and replication studies.
+#(B) Mitch pathway enrichment scores (S distances) in discovery and replication studies. Pathways with MANOVA FDR<0.05 are shown in red while others are shaded gray.
+#(C) Heatmap of 30 pathways with most extreme S distances (a surrogate of effect size). Red indicates increasing methylation and blue shows lower methylation.
+#(D) An example of a pathway identified with this method, "creation of C4 and C2 activators" shows lower methylation of member genes in both discovery and replication studies.
+#](../figs/fig6.png)
 
@@
 
-![**Figure 7. Pathway level DNA methylation differences in natural and IVF conceived infants.**
-(A) Contour heatmap showing the similarity in gene methylation score ranks in the Estill (450k) and Novakovic (EPIC) studies.
-(B) Mitch pathway enrichment scores (S distances) in Estil and Novakovic studies. Pathways with MANOVA FDR<0.05 are shown in red while others are shaded gray.
-(C) Heatmap of 30 pathways with most extreme S distances (a surrogate of effect size). Red indicates higher methylation and blue shows lower methylation.
-(D) An example of a pathway identified with this joint enrichment analysis method, "Adrenoreceptors" shows lower methylation of member genes in both Estill and Novakovic studies.
-](../figs/fig7.png)
+#![**Figure 7. Pathway level DNA methylation differences in natural and IVF conceived infants.**
+#(A) Contour heatmap showing the similarity in gene methylation score ranks in the Estill (450k) and Novakovic (EPIC) studies.
+#(B) Mitch pathway enrichment scores (S distances) in Estil and Novakovic studies. Pathways with MANOVA FDR<0.05 are shown in red while others are shaded gray.
+#(C) Heatmap of 30 pathways with most extreme S distances (a surrogate of effect size). Red indicates higher methylation and blue shows lower methylation.
+#(D) An example of a pathway identified with this joint enrichment analysis method, "Adrenoreceptors" shows lower methylation of member genes in both Estill and Novakovic studies.
+#](../figs/fig7.png)
 

### 2024-01-12 figs/GSE158422_sensitivity.Rmd
work on paper

@@
 barplot(abs(top$s.dist),horiz=TRUE,las=1,names.arg=barnames, cex.names=0.65,
-  cex.axis=0.8,col=cols, xlab="Enrichment score",main="Reactomes")
+  cex.axis=0.8,col=cols, xlab="Enrichment score",main="LAM")
 grid()
@@
 barplot(abs(top$s.dist),horiz=TRUE,las=1,names.arg=barnames, cex.names=0.8,
-  cex.axis=0.8,col=cols, xlab="Enrichment score",main="Reactomes")
+  cex.axis=0.8,col=cols, xlab="Enrichment score",main="LAM")
 grid()

### 2024-01-12 manuscript/manuscript.Rmd
work on paper

@@
 
-* Infinium methylation beadarrays remain one of the most popular platforms for studying genomic
-methylation, but tools for downstream pathways analysis have their limitations.
-
-* Functional class scoring (FCS) is a group of pathway enrichment techniques that involve the ranking of
-genes and evaluation of their collective regulation in biological systems.
-
-* Here we evaluate several approaches to apply FCS to Infinium data retaining directional information.
-The best performing method involves the mean aggregation of probe limma t-statistics by gene followed by
-a rank-ANOVA enrichment test using the mitch package.
-
-* Using real EPIC array data, we demonstrate that aggregation of probe methylation to gene level before
-differential methylation analysis yields a high degree of true positives and low number of contradictory
-results.
-
-* Using matched RNA-seq data we evaluate the relationship of methylation differences at promoters and
-gene bodies with RNA expression at the level of pathways.
-
-* We apply this approach to two further contexts, where existing methods failed to identify any
-pathway-level effects in 450K and EPIC datasets: (i) the effect of assisted reproductive technology on
-DNA methylation, and (2) the association of elevated plasma homocysteine with methylation patterns.
-
-* The methods described here have been incorporated into a package called gmea, short for gene
-methylation enrichment analysis.
+Infinium methylation beadarrays remain one of the most popular platforms for studying genomic
+DNA methylation, but tools for downstream pathways analysis have their limitations.
+Functional class scoring (FCS) is a group of pathway enrichment techniques that involve the ranking of
+genes and evaluation of their collective regulation in biological systems, but the implementations described for
+Infinium methylation data do not retain direction information which is important for mechanistic understanding
+of genomic regulation.
+Here, we evaluate several candidate FCS methods that retain directional information.
+According to simulation results, the best performing method involves the mean aggregation of probe limma
+t-statistics by gene followed by a rank-ANOVA enrichment test using the mitch package.
+This method, which we call "LAM", outperformed an existing over-representation analysis method in
+simulations, and showed higher sensitivity and robustness in an analysis of real lung tumor-normal paired datasets.
+Using matched RNA-seq data we examine the relationship of methylation differences at promoters and
+gene bodies with RNA expression at the level of pathways in lung cancer.
+To further demonstrate the utility of our approach, we apply it to two other contexts; firstly an examination of
+methylation differences associated with chronological age; and secondly to examine methylation differences in
+infants conceived with assisted reproductive technology.
+These results indicate that the LAM method is more sensitive and precise as compared to existing methods.
+LAM has the potential to illuminate disease mechanisms and identify novel diagnostic signatures.
+A vignette is provided to illustrate how to implement this method.
 
@@
 
-DNA methylation is the most widely studied epigenetic mark, for its role in development and
-disease [@Robertson2005-tu].
+DNA methylation is the most widely studied epigenetic mark, for its role in
+development and disease [@Robertson2005-tu].
 Many hundreds of epigenome-wide association studies (EWASs) are conducted each year to understand DNA
@@
 Various assisted reproductive technology conception groups are described in these studies, but we examined
-just fresh in vitro fertilised conceived infants compared to naturally conceived infants.
-
+just fresh in vitro fertilisation (IVF) conceived infants compared to naturally conceived infants.
 Raw intensity files (IDAT format) were read into R with minfi [aryee], normalistion was conducted with the SWAN method [maksimovic],
@@
 and 100.
+At each parameter setting, the simulations were repeated 100 times with a different seed value.
 True positives, false positives and false negatives were used to calculate overall precision and recall
@@
 
-![**Figure 1. Precision, recall and F1 for enrichment tests with simulated data.**
-(A) Overall results including 20, 50 and 100 genes per set.
-(B-D) Results for simulations using 20, 50 and 100 genes per set respectively.
-](../figs/fig1_slurm.pdf){height="500px" width="500px"}
+***Figure 1. Precision, recall and F1 for enrichment tests with simulated data.***
+*(A) Overall results including 20, 50 and 100 genes per set.
+(B-D) Results for simulations using 20, 50 and 100 genes per set respectively.*
 
@@
 
-![**Figure 2. Precision, recall and F1 for GSA and LAM enrichment tests with simulated data over a range of group sizes and delta values.** One hundred genes per set.](../figs/fig2.png)
-
-Overall, LAM was the superior FCS method across these tests and performed better than the existing GSA method.
+***Figure 2. Precision, recall and F1 for GSA and LAM enrichment tests with simulated data over a range of group sizes and delta
+values.***
+*One hundred genes per set.*
 
@@
 LAM execution took 51 seconds using eight parallel threads.
-With GSA, there were 75 pathways with higher methylation and 32 with lower methylation.
+With GSA, there were 75 Reactome pathways with higher methylation and 32 with lower methylation.
 12 pathways were significant in both directions.
@@
 This includes 26 and 4 with higher and lower methylation respectively (**Figure 3A**).
+GSA execution for both directions in series took 22 seconds.
 
-As the simulation results indicated LAM had better recall with larger gene sets, we were curious about whether there were differences in the sizes of
-gene sets found by LAM and GSA.
-GSA significant pathways had a median size of 135, while for LAM the median was 54.5, indicating pathways identified as significant by LAM were collectively smaller than GSA (**Figure 3B**).
+As the simulation results indicated LAM had better recall with larger gene sets, we were curious about whether there were differences
+in the sizes of gene sets found by LAM and GSA.
+GSA significant pathways had a median size of 135, while for LAM the median was 54.5, indicating pathways identified as significant
+with LAM were collectively smaller than GSA (**Figure 3B**).
+This may seem counterintuitive, but may be due to LAM's better sensitivity in dectecting pathway enrichment when effect sizes are
+lower.
+
 Although algorithmic accuracy cannot be inferred from the types of pathways identified in real data [@Timmons2015-kr],
@@
 
-![**Figure 3. Pathway enrichment analysis comparing normal and cancer samples using LAM and GSA methods.**
-(A) Euler diagrams showing the overlap in statistically significant pathways (FDR<0.05) from LAM and GSA results.
+***Figure 3. Pathway enrichment analysis comparing normal and cancer samples using LAM and GSA methods.***
+*(A) Euler diagrams showing the overlap in statistically significant pathways (FDR<0.05) from LAM and GSA results.
 (B) Sizes of significant gene sets found with LAM and GSA methods.
 (C) Significant LAM pathways with largest S distance, a proxy for effect size.
-(D) Top ranked significant GSA pathways ranked by fold enrichment.
-](../figs/fig3.png)
+(D) Top ranked significant GSA pathways ranked by fold enrichment.*
 
@@
 termed "inconsistent".
-At a sample size of 10, LAM was able to detect 354 out of 406 (87%) consistent pathways with a sample size of 10,
+At a sample size of 10, LAM was able to detect 354 out of 406 (87%) consistent pathways,
 while GSA detected just 13 of 107 at this sample size (12%).
@@
 inconsistent findings was lower across all sample sizes.
-At a sample size of 10, the proportion of inconsistent pathways was 13% for LAM and 33% for GSA.
+At a sample size of 10, the proportion of inconsistent pathways was 10% for LAM and 28% for GSA.
 This result suggests that LAM has superior sensitivity to detect differentially methylated
@@
 
-![**Figure 4. Sensitivity of LAM and GSA methods with real cancer array data.**
-Upper panels show the number of significant pathways identified in the smaller sample that were consistent
+***Figure 4. Sensitivity of LAM and GSA methods with real cancer array data.***
+*Upper panels show the number of significant pathways identified in the smaller sample that were consistent
 with the full group for LAM (left) and GSA (right).
@@
 group.
-Left panels correspond to LAM, and right panels to GSA methods.
-](../figs/fig4.png)
+Left panels correspond to LAM, and right panels to GSA methods.*
 
-## LAM and GSA integrative analysis with RNA expression
+## Integrating methylation and RNA expression pathways
 
@@
 DNA methylation.
-From 18,704 detected genes, There were 12,380 differentially expressed genes (FDR<0.05), with 7,415 and 4,965 up- and down-regulated in the tumour group respectively.
+From 18,704 detected genes, There were 12,380 differentially expressed genes (FDR<0.05), with 7,415 and 4,965 up- and down-regulated
+in the tumour group respectively.
 At the pathway level there were 304 and 135 up- and down-regulated Reactome pathways with altered gene expression (FDR<0.05).
@@
 were considered separately.
-In promoters, GSA identified 15 and 7 pathways with higher and lower methylation respectively, while at gene bodies, there were were 18 and 75
-pathways with higher and lower methylation respectively (FDR<0.05).
+In promoters, GSA identified 15 and 7 pathways with higher and lower methylation respectively, while at gene bodies, there were were
+18 and 75 pathways with higher and lower methylation respectively (FDR<0.05).
 Using LAM at promoters, there were 190 and 20 pathways identified with higher and lower methylation respectively.
 LAM at gene bodies identified 251 and 31 pathways with higher and lower methylation respectively (FDR<0.05).
-The relative overlap between significant promoter and gene body pathways was larger for LAM pathway sets (22%) as compared to GSA (3%) (**Figure 5A,B**).
+The relative overlap between significant promoter and gene body pathways was larger for LAM pathway sets (22%) as compared to GSA
+(3%) (**Figure 5A,B**).
 There was no observed overlap between GSA methylation-based pathways and gene expression pathways (**Figure 5A**), but there were some
 overlaps between pathways identified with LAM based promoter and gene body methylation and gene expression (**Figure 5B**).
-Interestingly, 136/304 upregulated gene expression pathways were associated with increasing gene body methylation with LAM.
-Likewise 21 pathways with downregulated expression were associated with increasing promoter methylation.
-Some of these pathways with large effect sizes across the three contrasts are depicted in heatmap form (**Figure 5C**).
-Of these selected pathways, there was a mixture of concordant and discordant signatures in the three contrasts,
-suggesting that promoter methylation differences do not strongly influence RNA expression at the level of pathways.
-However there were a small number of pathways that did exhibit what could be epigenetic influences, for example
+
+Using LAM, 136/304 upregulated RNA expression pathways were associated with increasing gene body methylation.
+Likewise, 21 pathways with downregulated RNA expression were associated with increasing promoter methylation.
+Some of these pathways with large enrichment scores across the three contrasts are depicted in heatmap form (**Figure 5C**).
+There were a small number of pathways that exhibited an inverse relationship between promoter methylation and RNA expression such as
 "unwinding of DNA", "mucopolysaccharidoses","interaction with cumulus cells and the zona pellucida", "glucocorticoid biosynthesis" and
-"olfactory signaling" demonstrated an inverse relationship between promoter methylation and gene expression.
-These results demonstrate that LAM can illuminate the complicated relationship between DNA methylation and gene expression at the pathway level in cancer.
+"olfactory signaling".
+These were however less common than pathways that showed a positive relationship between RNA expression and promoter methylation.
+These results demonstrate that LAM can illuminate the complicated relationship between DNA methylation and gene expression at the
+pathway level in cancer.
 
-#[**Figure 5. Integration of gene expression data with methylation pathways.**
-#(A,B) Euler diagrams showing the overlap in statistically significant pathways (FDR<0.05) from gene expression and GSA results (A) and LAM (B) results.
-#(C) Multi-dimensional enrichment analysis of promoter and gene body methylation with gene expression using mitch. Top 20 gene sets shown with largest s distance after FDR filtering at 0.05.
-#](../figs/fig5.png)
+***Figure 5. Integration of gene expression data with methylation pathways.***
+*(A,B) Euler diagrams showing the overlap in statistically significant pathways (FDR<0.05) from gene expression and GSA results (A)
+and LAM (B) results.
+(C) Multi-dimensional enrichment analysis of promoter and gene body methylation with gene expression using mitch.
+Top 20 gene sets shown with largest s distance after FDR filtering at 0.05.*
 
@@
 pathway level differential methylation is not well defined.
-Using summary statistics from independent discovery and replication studies using the EPIC array platform described
-by McCartney et al (2020) [@McCartney2020-ob] irrespective of sex, we applied the LAM method to explore the pathway level differential methylation.
+Using limma summary statistics from independent discovery and replication studies using the EPIC array platform described
+by McCartney et al (2020) [@McCartney2020-ob], we applied the LAM method to explore the pathway level
+differential methylation.
 In the discovery cohort we observed 304 differential pathways (FDR<0.05), while in the replication group we observed 107.
 There was a high degree of agreement, with 43 pathways significant in both groups.
-This is not surprising, as the gene-level differential methylation scores show a strong positive association between discovery and replication studies (**Figure 6A**).
-Using the MANOVA test, we identified 390 pathways with altered methylation.
-Visualised as a scatterplot, the overall pattern is concordant, although a cluster of points in the lower right of the chart indicates a cluster of 268 pathways with higher methylation in the discovery sample that were lower in methylation in the replication sample (**Figure 6B**).
-When prioritising the results by S distance, a surrogate measure of effect size, 29 of the top 30 pathways were concordant in their direction of regulation (**Figure 6C**).
+This is not surprising, as the gene-level rank differential methylation scores show a strong positive association between discovery and
+replication studies (**Figure 6A**).
+Using the rank-MANOVA test of the mitch package, we identified 390 pathways with altered methylation.
+Visualised as a scatterplot, the overall pattern is concordant, although a cluster of points in the lower right of the chart
+indicates 268 pathways with higher methylation in the discovery sample and lower in methylation in the replication
+sample (**Figure 6B**).
+When prioritising the results by S distance, a surrogate measure of effect size, 29 of the top 30 pathways were concordant in their
+direction of regulation (**Figure 6C**).
 Some of the observed methylation changes make sense with reports in the literature.
-For example the complement pathway is associated with aging and in this analysis we observe a reduction in methylation of genes beloning to this set including *C1R*, *MBL2*, *C1S* and *C1QC* (**Figure 6D**).
+For example the complement pathway is associated with aging and in this analysis we observe a reduction in methylation of genes
+beloning to this set including *C1R*, *MBL2*, *C1S* and *C1QC* (**Figure 6D**).
 
-#![**Figure 6. Pathway level DNA methylation alterations with chronological age.**
-#(A) Contour heatmap showing the similarity in gene methylation score ranks in the discovery and replication studies.
-#(B) Mitch pathway enrichment scores (S distances) in discovery and replication studies. Pathways with MANOVA FDR<0.05 are shown in red while others are shaded gray.
-#(C) Heatmap of 30 pathways with most extreme S distances (a surrogate of effect size). Red indicates increasing methylation and blue shows lower methylation.
-#(D) An example of a pathway identified with this method, "creation of C4 and C2 activators" shows lower methylation of member genes in both discovery and replication studies.
-#](../figs/fig6.png)
+***Figure 6. Pathway level DNA methylation alterations with chronological age.***
+*(A) Contour heatmap showing the similarity in gene methylation score ranks in the discovery and replication studies.
+(B) Mitch pathway enrichment scores (S distances) in discovery and replication studies.
+Pathways with MANOVA FDR<0.05 are shown in red while others are shaded gray.
+(C) Heatmap of 30 pathways with most extreme S distances (a surrogate of effect size).
+Red indicates increasing methylation and blue shows lower methylation.
+(D) An example of a pathway identified with this method, "creation of C4 and C2 activators" shows lower methylation of member genes
+in both discovery and replication studies.*
 
-## Exploration of methylation pathways altered by IVF
+## Exploration of methylation pathway differences in infants conceived with in vitro fertilisation
 
@@
 array systems.
-Epigenetic differences between infants conceived naturally and with assisted reproductive technologies have been the subject of studies since 2009 [@Katari2009-ya].
-Two of the highest powered studies with publicly available data are Estill 2016 which used the 450K array and Novakovic 2019 study which used
-the EPIC array [@Estill2016-jh;@Novakovic2019-jt].
+Epigenetic differences between infants conceived naturally and with assisted reproductive technologies have been the subject of
+studies since 2009 [@Katari2009-ya].
+Two of the highest powered studies with publicly available data are Estill 2016 which used the 450K array and Novakovic 2019 study
+which used the EPIC array [@Estill2016-jh;@Novakovic2019-jt].
 Joint enrichment analysis can uncover pathways that are common or different between these studies and illuminate
 the biological differences between conception groups.
-For this example we focused on comparing naturally conceived infants to those conceived with in vitro fertilisation with fresh embryo transfer.
-The Estill dataset on 450k had 418,833 probes that met the quality criteria (Natural n=43; IVF n=38).
-Of these, 149,589 probes showed statistical significant differential methylation (FDR<0.05), of which 28,767 and 120,822 exhibited higher or lower methylation in the IVF group respectively.
+For this example we focused on comparing naturally conceived infants to those conceived with in vitro fertilisation (IVF) with fresh
+embryo transfer.
+The Estill dataset on the 450k platform had 418,833 probes that met the quality criteria (Natural n=43; IVF n=38).
+Of these, 149,589 probes showed statistical significant differential methylation (FDR<0.05), of which 28,767 and 120,822 exhibited
+higher or lower methylation in the IVF group respectively.
 The Novakovic dataset on EPIC array had 793,844 probes that met the filtering criteria (n=133).
-Of these, 5,562 were statistically significant (FDR<0.05) with 1,348 and 4,214 exhibited higher or lower methylation in the IVF group respectively.
-After summarisation to the gene level, the Estill dataset on 450k array describes 19,240 genes, while the
-Novakovic dataset on EPIC array describes 22,588 genes.
+Of these, 5,562 were statistically significant (FDR<0.05) with 1,348 and 4,214 exhibited higher or lower methylation in the IVF group
+respectively.
+After summarisation to the gene level, the Estill dataset on 450k array describes 19,240 genes, while the Novakovic dataset on EPIC
+array describes 22,588 genes.
 There were 19,234 genes common to both platforms.
-The rank-rank plot indicates a high degree of overlap overall, and interestingly shows a trend of more genes having lower methylation level in the IVF group (**Figure 7A**).
-A scatterplot of pathway enrichment scores indicates a moderate degree of agreement between these studies at the level of pathway enrichment scores (*r*=0.35,*p*=2.2e-16) despite these studies being conducted on independent cohorts years apart and analysed with different array systems (**Figure 7B**).
+The rank-rank plot of differential gene methylation scores indicates a high degree of similarity overall, and interestingly shows a
+trend of more genes having lower methylation level in the IVF group (**Figure 7A**).
+A scatterplot of pathway enrichment scores indicates a moderate degree of agreement between these studies
+(*r*=0.35,*p*=2.2e-16) despite these studies being conducted on independent cohorts years apart and analysed with different array
+systems (**Figure 7B**).
 A heatmap shows a high degree of agreement of pathways with higher methylation in the IVF groups,
@@
 between studies (**Figure 7C**).
-We were interested in which pathways had consistently lower methylation in the IVF group and identified the Reactome pathway "Adrenoreceptors" which had enrichment scores of -0.42 in the Estill study and -0.56 in the Novakovic study, recording a MANOVA FDR value of 0.04 (**Figure 7D**).
+We were interested in which pathways had consistently lower methylation in the IVF group and identified the Reactome pathway
+"Adrenoreceptors" which had enrichment scores of -0.42 in the Estill study and -0.56 in the Novakovic study, recording a MANOVA FDR
+value of 0.04 (**Figure 7D**).
 Top ranked genes from this pathway include *ADRA2C*, *ADRB2*, *ADRA2B* and *ADRA1B*.
@@
 
-#![**Figure 7. Pathway level DNA methylation differences in natural and IVF conceived infants.**
-#(A) Contour heatmap showing the similarity in gene methylation score ranks in the Estill (450k) and Novakovic (EPIC) studies.
-#(B) Mitch pathway enrichment scores (S distances) in Estil and Novakovic studies. Pathways with MANOVA FDR<0.05 are shown in red while others are shaded gray.
-#(C) Heatmap of 30 pathways with most extreme S distances (a surrogate of effect size). Red indicates higher methylation and blue shows lower methylation.
-#(D) An example of a pathway identified with this joint enrichment analysis method, "Adrenoreceptors" shows lower methylation of member genes in both Estill and Novakovic studies.
-#](../figs/fig7.png)
+***Figure 7. Pathway level DNA methylation differences in natural and IVF conceived infants.***
+*(A) Contour heatmap showing the similarity in gene methylation score ranks in the Estill (450k) and Novakovic (EPIC) studies.
+(B) Mitch pathway enrichment scores (S distances) in Estil and Novakovic studies.
+Pathways with MANOVA FDR<0.05 are shown in red while others are shaded gray.
+(C) Heatmap of 30 pathways with most extreme S distances (a surrogate of effect size).
+Red indicates higher methylation and blue shows lower methylation.
+(D) An example of a pathway identified with this joint enrichment analysis method, "Adrenoreceptors" shows lower methylation of
+member genes in both Estill and Novakovic studies.*
 

### 2024-01-12 manuscript/manuscript.Rmd
working on manuscript

@@
     fig_height: 5
-bibliography: manuscript.bib
+bibliography: references.bib
 csl: epigenetics.csl
@@
 Mandhri Abeysooriya<sup>2,3</sup>,
-Anusuiya Bora<sup>2</sup>,
+Anusuiya Bora<sup>1,2</sup>,
 Severin Lamon<sup>3</sup>,
-Yen Wong<sup>4</sup>,
-Jeffrey Craig<sup>4</sup>
+Mary Kasu<sup>1</sup>,
+Mitch Norris<sup>1</sup>,
+Yen Ting Wong<sup>4</sup>,
+Jeffrey Craig<sup>4,5</sup>
 </strong>
@@
 
-(*) Correspondence: m.ziemann@deakin.edu.au
+5. Murdoch Children's Research Institute, Melbourne, Australia.
+
+(*) Correspondence: mark.ziemann@burnet.edu.au
 
@@
 | Mark Ziemann | 0000-0002-7688-6974 |
+| Mandhri Abeysooriya | 0000-0003-2163-6203 |
 | Anusuiya Bora | 0009-0006-2908-1352 |
+| Severine Lamon | 0000-0002-3271-6551 |
 
@@
 into consideration the variable number of probes followed by an FCS test based on z-scores [@Ren2019-mr].
-Secondly, `methylglm` uses a logistic regression approach to model differential methylation for genes inside and outside each gene set [@Ren2019-mr].
+Secondly, `methylglm` uses a logistic regression approach to model differential methylation for genes inside and outside each gene set
+[@Ren2019-mr].
 These methods are indeed more sensitive than ORA approaches, but as they are based on one-tailed tests irrespective of
@@
 Due to the fact that genes have a variable number of probes, there are many concevable ways that FCS could be applied
-to Infinium Methylation data.
+to Infinium methylation array data.
 The first set of methods use differential probe methylation data from `limma` [@Ritchie2015-zn] as an input.
@@
 Then, differential gene methylation results are used for a downstream enrichment test.
-With the above methods there are several methodological choices, such as the type of aggregation (eg: mean, median), and the type of downstream test (eg: parametric or nonparametric).
+With the above methods there are several methodological choices, such as the type of aggregation (eg: mean, median), and the type of
+downstream test (eg: parametric or nonparametric).
 To determine the best approach to applying FCS to Infinium methylation data, in this study we use
@@
 Functional enrichment analysis is a process of data summarisation from genes to gene sets (pathways).
-This is made more complicated for Infinium array data due to the presence of multiple probes per gene,
+This is made more complicated for Infinium methylation array data due to the presence of multiple probes per gene,
 meaning the data needs to be summarised from probes to genes and then to gene sets.
@@
 Similar to LAT, except instead of the two-sided t-test of gene set enrichment, a non-parametric alternative,
-the Wilcoxon signed-rank test.
+the Wilcoxon signed-rank test is used.
 
 4. Limma Average Mitch (LAM).
-Similar to LAW, however instead of the Wilcoxon test, the Mitch package is used to execute a two-way,
+Similar to LAW, however instead of the Wilcoxon test, the mitch package[@Kaspi2020-yx] is used to execute a two-way,
 two-sample ANOVA-on-ranks test (Kruskal-Wallis).
@@
 the number of true positives (TP), false positives (FP) and false negatives (FN) could be determined.
-At each setting for group size, delta and gene set size, 50 replications were conducted each with a set seed
+At each setting for group size, delta and gene set size, 100 replications were conducted each with a set seed
 to ensure reproducibility.
+Simulations were conducted on a SLURM cluster [@Yoo2003-mc].
 Mean TP, FP and FN values were determined, and then precision, recall and F1 scores were calculated.
@@
 Downstream pathway enrichment analysis was conducted using the GSAmeth and LAM methods
-with Reactome pathways downloaded from MSigDB website (version 2023.1).
+with Reactome pathways downloaded from MSigDB website (version 2023.1)[@Milacic2024-js].
 LAM relies on probe-gene associations from the "UCSC_RefGene_Name" column of the "IlluminaHumanMethylationEPICanno.ilm10b4.hg19"
@@
 respective method, with n varied between 2 and 30.
-Significantly enriched pathways were defined as those with ANOVA FDR values < 0.05.
+Significantly enriched pathways were defined as those with FDR values < 0.05.
 The subset significant results were compared to the full group (n=37) results to test the sensitivity of these enrichment
-methods at differenent effect size levels, and whether findings at a smaller sample size were consistent with the full group.
+methods, and whether findings at a smaller sample size were consistent with the full group.
 This process was repeated 50 times at each sample size.
@@
 This underwent filtering to remove genes with expression below 10 reads per sample on average across the dataset.
-Then DESeq2 was used to compare normal and cancerous tissue gene expression taking into consideration sample pairing.
-The differential expression results underwent enrichment analysis using the mitch package with default settings for DESeq2 data tables.
+Then DESeq2[@Love2014-wt] was used to compare normal and cancerous tissue gene expression taking into consideration sample pairing.
+The differential expression results underwent enrichment analysis using the mitch package [@Kaspi2020-yx] with default settings for
+DESeq2 data tables.
 CpG sites annotated as promoters were considered separately from those located at gene bodies in mitch analysis.
@@
 participants [@McCartney2020-ob].
-Limma summary statistics for chromological age were read into R and enrichment analysis was conducted as above.
-Briefly, gene-probe associations were first obtained from the
-"IlluminaHumanMethylationEPICanno.ilm10b4.hg19" dataset, followed by update of gene symbols with
-HGNChelper.
+Limma summary statistics for chromological age were read into R and enrichment analysis was conducted with the LAM method.
+Briefly, gene-probe associations were first obtained from the "IlluminaHumanMethylationEPICanno.ilm10b4.hg19"
+dataset, followed by update of gene symbols with HGNChelper.
 Data were imported with mitch and enrichment analysis was conducted with gene sets from Reactome.
@@
 
-In order to demonstrate that methylation patterns across different array platforms can be compared using LAM,
+In order to demonstrate that methylation patterns across different array platforms can be compared using the LAM method,
 we examined methylation differences in infants associated with assisted reproduction as described by two
-independent studies [@Estill2016-jh;@Novakovic2019-jt].
-The Estill study was published in 2016 and is based on the HM450K array and the data is available from GEO Accession GSE79257.
+independent studies.
+The Estill study was published in 2016 and is based on the HM450K array and the data is available from GEO Accession GSE79257
+[@Estill2016-jh].
 The Novakovic study in was published 2019, and used the EPIC methylation array and the data is available from GEO under accession
-GSE131433.
+GSE131433[@Novakovic2019-jt].
 Various assisted reproductive technology conception groups are described in these studies, but we examined
 just fresh in vitro fertilisation (IVF) conceived infants compared to naturally conceived infants.
-Raw intensity files (IDAT format) were read into R with minfi [aryee], normalistion was conducted with the SWAN method [maksimovic],
-and probe filtering was conducted to remove proves with a detection p-value > 0.01 in addition to probes located on X or Y
-chromosomes.
-M values were computed and underwent differential methylation analysis using limma [ritchie].
+Raw intensity files (IDAT format) were read into R with minfi [@Aryee2014-su], normalistion was conducted with the SWAN method
+[@Maksimovic2012-hy], and probe filtering was conducted to remove proves with a detection p-value > 0.01 in addition to probes
+located on X or Y chromosomes.
+M values were computed and underwent differential methylation analysis using limma [@Ritchie2015-zn].
 The Estill study included 43 naturally conceived infants and 38 with fresh embryo transfer.
-The Novakovic study included XX naturally conceived infants and XX with fresh embryo transfer.
-Limma differential methylation tables were then imported with mitch.
+The Novakovic study included 58 naturally conceived infants and 75 with fresh embryo transfer.
+Limma differential analysis was conducted accounting for sex, then methylation tables underwent LAM analysis with Reactome gene sets.
 Probe gene associations for HM450K array were established using the "IlluminaHumanMethylation450kanno.ilmn12.hg19" annotation set
-and the gene symbols were updated with HGNChelper (database current as of 21-Dec-2023).
+and the gene symbols were updated with HGNChelper.
 As the Novakovic study involved EPIC array data, gene-probe associations were obtained from the
@@
 HGNChelper.
-Limma tables were read into R and imported with mitch.
-Gene level differential methylation score tables were merged and underwent mitch enrichment analysis using Reactome gene sets.
 
@@
 
-***Figure 1. Precision, recall and F1 for enrichment tests with simulated data.***
-*(A) Overall results including 20, 50 and 100 genes per set.
-(B-D) Results for simulations using 20, 50 and 100 genes per set respectively.*
+>***Figure 1. Precision, recall and F1 for enrichment tests with simulated data.***
+>*(A) Overall results including 20, 50 and 100 genes per set.
+>(B-D) Results for simulations using 20, 50 and 100 genes per set respectively.*
 
@@
 
-***Figure 2. Precision, recall and F1 for GSA and LAM enrichment tests with simulated data over a range of group sizes and delta
-values.***
-*One hundred genes per set.*
+>***Figure 2. Precision, recall and F1 for GSA and LAM enrichment tests with simulated data over a range of group sizes and delta
+>values.***
+>*One hundred genes per set.*
 
@@
 
-***Figure 3. Pathway enrichment analysis comparing normal and cancer samples using LAM and GSA methods.***
-*(A) Euler diagrams showing the overlap in statistically significant pathways (FDR<0.05) from LAM and GSA results.
-(B) Sizes of significant gene sets found with LAM and GSA methods.
-(C) Significant LAM pathways with largest S distance, a proxy for effect size.
-(D) Top ranked significant GSA pathways ranked by fold enrichment.*
+>***Figure 3. Pathway enrichment analysis comparing normal and cancer samples using LAM and GSA methods.***
+>*(A) Euler diagrams showing the overlap in statistically significant pathways (FDR<0.05) from LAM and GSA results.
+>(B) Sizes of significant gene sets found with LAM and GSA methods.
+>(C) Significant LAM pathways with largest S distance, a proxy for effect size.
+>(D) Top ranked significant GSA pathways ranked by fold enrichment.*
 
@@
 
-***Figure 4. Sensitivity of LAM and GSA methods with real cancer array data.***
-*Upper panels show the number of significant pathways identified in the smaller sample that were consistent
-with the full group for LAM (left) and GSA (right).
-Middle panels show the number of significant pathways identified in the smaller sample that were inconsistent
-with the full group.
-Lower panels show the proportion of significant pathways identified that were inconsistent with the full
-group.
-Left panels correspond to LAM, and right panels to GSA methods.*
+>***Figure 4. Sensitivity of LAM and GSA methods with real cancer array data.***
+>*Upper panels show the number of significant pathways identified in the smaller sample that were consistent
+>with the full group for LAM (left) and GSA (right).
+>Middle panels show the number of significant pathways identified in the smaller sample that were inconsistent
+>with the full group.
+>Lower panels show the proportion of significant pathways identified that were inconsistent with the full
+>group.
+>Left panels correspond to LAM, and right panels to GSA methods.*
 
@@
 
-***Figure 5. Integration of gene expression data with methylation pathways.***
-*(A,B) Euler diagrams showing the overlap in statistically significant pathways (FDR<0.05) from gene expression and GSA results (A)
-and LAM (B) results.
-(C) Multi-dimensional enrichment analysis of promoter and gene body methylation with gene expression using mitch.
-Top 20 gene sets shown with largest s distance after FDR filtering at 0.05.*
+>***Figure 5. Integration of gene expression data with methylation pathways.***
+>*(A,B) Euler diagrams showing the overlap in statistically significant pathways (FDR<0.05) from gene expression and GSA results (A)
+>and LAM (B) results.
+>(C) Multi-dimensional enrichment analysis of promoter and gene body methylation with gene expression using mitch.
+>Top 20 gene sets shown with largest s distance after FDR filtering at 0.05.*
 
@@
 
-***Figure 6. Pathway level DNA methylation alterations with chronological age.***
-*(A) Contour heatmap showing the similarity in gene methylation score ranks in the discovery and replication studies.
-(B) Mitch pathway enrichment scores (S distances) in discovery and replication studies.
-Pathways with MANOVA FDR<0.05 are shown in red while others are shaded gray.
-(C) Heatmap of 30 pathways with most extreme S distances (a surrogate of effect size).
-Red indicates increasing methylation and blue shows lower methylation.
-(D) An example of a pathway identified with this method, "creation of C4 and C2 activators" shows lower methylation of member genes
-in both discovery and replication studies.*
+>***Figure 6. Pathway level DNA methylation alterations with chronological age.***
+>*(A) Contour heatmap showing the similarity in gene methylation score ranks in the discovery and replication studies.
+>(B) Mitch pathway enrichment scores (S distances) in discovery and replication studies.
+>Pathways with MANOVA FDR<0.05 are shown in red while others are shaded gray.
+>(C) Heatmap of 30 pathways with most extreme S distances (a surrogate of effect size).
+>Red indicates increasing methylation and blue shows lower methylation.
+>(D) An example of a pathway identified with this method, "creation of C4 and C2 activators" shows lower methylation of member genes
+>in both discovery and replication studies.*
 
@@
 
-***Figure 7. Pathway level DNA methylation differences in natural and IVF conceived infants.***
-*(A) Contour heatmap showing the similarity in gene methylation score ranks in the Estill (450k) and Novakovic (EPIC) studies.
-(B) Mitch pathway enrichment scores (S distances) in Estil and Novakovic studies.
-Pathways with MANOVA FDR<0.05 are shown in red while others are shaded gray.
-(C) Heatmap of 30 pathways with most extreme S distances (a surrogate of effect size).
-Red indicates higher methylation and blue shows lower methylation.
-(D) An example of a pathway identified with this joint enrichment analysis method, "Adrenoreceptors" shows lower methylation of
-member genes in both Estill and Novakovic studies.*
+>***Figure 7. Pathway level DNA methylation differences in natural and IVF conceived infants.***
+>*(A) Contour heatmap showing the similarity in gene methylation score ranks in the Estill (450k) and Novakovic (EPIC) studies.
+>(B) Mitch pathway enrichment scores (S distances) in Estil and Novakovic studies.
+>Pathways with MANOVA FDR<0.05 are shown in red while others are shaded gray.
+>(C) Heatmap of 30 pathways with most extreme S distances (a surrogate of effect size).
+>Red indicates higher methylation and blue shows lower methylation.
+>(D) An example of a pathway identified with this joint enrichment analysis method, "Adrenoreceptors" shows lower methylation of
+>member genes in both Estill and Novakovic studies.*
 
@@
 
+## Code availability
+
+Code to generate the results shown here are included in a GitHub repository (https://github.com/markziemann/gmea/).
+To enable reproduction of the findings, a docker image is made available at DockerHub (https://hub.docker.com/repository/docker/mziemann/gmea/general).
+
 ## Bibliography

### 2024-01-13 manuscript/manuscript.Rmd
manuscript.Rmd

@@
 methylation differences associated with chronological age; and secondly to examine methylation differences in
-infants conceived with assisted reproductive technology.
+infants conceived with in vitro fertilisation.
 These results indicate that the LAM method is more sensitive and precise as compared to existing methods.
@@
 biases towards large gene sets that have non-specific biological functions.
-This can lead to end users over-looking smaller pathways with more specific biological functions that have larger effect
-sizes, as those p-values are relatively larger.
 
@@
 For this analysis, Reactome gene sets were used.
-Subsequently, to generate the mitch heatmap, we used mitch in multivariate mode by analysing differential RNA expression,
+Subsequently, to generate the pathway heatmap, we used mitch in multivariate mode by analysing differential RNA expression,
 promoter and gene body methylation together, prioritising the results by s-distance, a surrogate measure of effect size.
@@
 participants [@McCartney2020-ob].
-Limma summary statistics for chromological age were read into R and enrichment analysis was conducted with the LAM method.
+Limma summary statistics for chronological age including discovery and replication groups were read into R and enrichment analysis was conducted with the LAM method.
 Briefly, gene-probe associations were first obtained from the "IlluminaHumanMethylationEPICanno.ilm10b4.hg19"
 dataset, followed by update of gene symbols with HGNChelper.
-Data were imported with mitch and enrichment analysis was conducted with gene sets from Reactome.
+Discovery and replication data were imported separately with mitch and multivariate enrichment analysis was conducted with gene sets from Reactome.
 
@@
 
-Simulations were conducted using the seven FCS methods (LAT, LTT, LAW, LAM, LRM, ALT, ALW, ALM) and GSAmeth,
+Simulations were conducted using the eight FCS methods (LAT, LTT, LAW, LAM, LRM, ALT, ALW, ALM) and GSAmeth,
 with group sizes of 3-12 and delta values between 0.1 and 0.5, and random gene sets with sizes 20, 50
@@
 
->***Figure 1. Precision, recall and F1 for enrichment tests with simulated data.***
->*(A) Overall results including 20, 50 and 100 genes per set.
->(B-D) Results for simulations using 20, 50 and 100 genes per set respectively.*
+![**Figure 1. Precision, recall and F1 for enrichment tests with simulated data.**
+(A) Overall results including 20, 50 and 100 genes per set.
+(B-D) Results for simulations using 20, 50 and 100 genes per set respectively.
+](img/fig1.svg)
 
@@
 
->***Figure 2. Precision, recall and F1 for GSA and LAM enrichment tests with simulated data over a range of group sizes and delta
->values.***
->*One hundred genes per set.*
+![**Figure 2. Precision, recall and F1 for GSA and LAM enrichment tests with simulated data over a range of group sizes and delta
+values.**
+One hundred genes per set.](img/fig2.svg)
 
@@
 LAM resulted in 406 significant Reactome pathways at FDR<0.05.
-Of these 370 involved higher methylation, while 36 involved lower methylation.
+Of these, 370 involved higher methylation, while 36 involved lower methylation.
 LAM execution took 51 seconds using eight parallel threads.
@@
 This includes 26 and 4 with higher and lower methylation respectively (**Figure 3A**).
-GSA execution for both directions in series took 22 seconds.
+GSA execution for both directions in series took 22 seconds on one thread.
 
@@
 
->***Figure 3. Pathway enrichment analysis comparing normal and cancer samples using LAM and GSA methods.***
->*(A) Euler diagrams showing the overlap in statistically significant pathways (FDR<0.05) from LAM and GSA results.
->(B) Sizes of significant gene sets found with LAM and GSA methods.
->(C) Significant LAM pathways with largest S distance, a proxy for effect size.
->(D) Top ranked significant GSA pathways ranked by fold enrichment.*
+![**Figure 3. Pathway enrichment analysis comparing normal and cancer samples using LAM and GSA methods.**
+(A) Euler diagrams showing the overlap in statistically significant pathways (FDR<0.05) from LAM and GSA results.
+(B) Sizes of significant gene sets found with LAM and GSA methods.
+(C) Significant LAM pathways with largest S distance, a proxy for effect size.
+(D) Top ranked significant GSA pathways ranked by fold enrichment.](img/fig3.svg)
 
@@
 
->***Figure 4. Sensitivity of LAM and GSA methods with real cancer array data.***
->*Upper panels show the number of significant pathways identified in the smaller sample that were consistent
->with the full group for LAM (left) and GSA (right).
->Middle panels show the number of significant pathways identified in the smaller sample that were inconsistent
->with the full group.
->Lower panels show the proportion of significant pathways identified that were inconsistent with the full
->group.
->Left panels correspond to LAM, and right panels to GSA methods.*
+![**Figure 4. Sensitivity of LAM and GSA methods with real cancer array data.**
+Upper panels show the number of significant pathways identified in the smaller sample that were consistent
+with the full group for LAM (left) and GSA (right).
+Middle panels show the number of significant pathways identified in the smaller sample that were inconsistent
+with the full group.
+Lower panels show the proportion of significant pathways identified that were inconsistent with the full
+group.
+Left panels correspond to LAM, and right panels to GSA methods.](img/fig4.svg)
 
@@
 
->***Figure 5. Integration of gene expression data with methylation pathways.***
->*(A,B) Euler diagrams showing the overlap in statistically significant pathways (FDR<0.05) from gene expression and GSA results (A)
->and LAM (B) results.
->(C) Multi-dimensional enrichment analysis of promoter and gene body methylation with gene expression using mitch.
->Top 20 gene sets shown with largest s distance after FDR filtering at 0.05.*
+![**Figure 5. Integration of gene expression data with methylation pathways.**
+(A,B) Euler diagrams showing the overlap in statistically significant pathways (FDR<0.05) from gene expression and GSA results (A)
+and LAM (B) results.
+(C) Multi-dimensional enrichment analysis of promoter and gene body methylation with gene expression using mitch.
+Top 20 gene sets shown with largest s distance after FDR filtering at 0.05.](img/fig5.svg)
 
@@
 
->***Figure 6. Pathway level DNA methylation alterations with chronological age.***
->*(A) Contour heatmap showing the similarity in gene methylation score ranks in the discovery and replication studies.
->(B) Mitch pathway enrichment scores (S distances) in discovery and replication studies.
->Pathways with MANOVA FDR<0.05 are shown in red while others are shaded gray.
->(C) Heatmap of 30 pathways with most extreme S distances (a surrogate of effect size).
->Red indicates increasing methylation and blue shows lower methylation.
->(D) An example of a pathway identified with this method, "creation of C4 and C2 activators" shows lower methylation of member genes
->in both discovery and replication studies.*
+![**Figure 6. Pathway level DNA methylation alterations with chronological age.**
+(A) Contour heatmap showing the similarity in gene methylation score ranks in the discovery and replication studies.
+(B) Mitch pathway enrichment scores (S distances) in discovery and replication studies.
+Pathways with MANOVA FDR<0.05 are shown in red while others are shaded gray.
+(C) Heatmap of 30 pathways with most extreme S distances (a surrogate of effect size).
+Red indicates increasing methylation and blue shows lower methylation.
+(D) An example of a pathway identified with this method, "creation of C4 and C2 activators" shows lower methylation of member genes
+in both discovery and replication studies.](img/fig6.svg)
 
@@
 
->***Figure 7. Pathway level DNA methylation differences in natural and IVF conceived infants.***
->*(A) Contour heatmap showing the similarity in gene methylation score ranks in the Estill (450k) and Novakovic (EPIC) studies.
->(B) Mitch pathway enrichment scores (S distances) in Estil and Novakovic studies.
->Pathways with MANOVA FDR<0.05 are shown in red while others are shaded gray.
->(C) Heatmap of 30 pathways with most extreme S distances (a surrogate of effect size).
->Red indicates higher methylation and blue shows lower methylation.
->(D) An example of a pathway identified with this joint enrichment analysis method, "Adrenoreceptors" shows lower methylation of
->member genes in both Estill and Novakovic studies.*
+![**Figure 7. Pathway level DNA methylation differences in natural and IVF conceived infants.**
+(A) Contour heatmap showing the similarity in gene methylation score ranks in the Estill (450k) and Novakovic (EPIC) studies.
+(B) Mitch pathway enrichment scores (S distances) in Estil and Novakovic studies.
+Pathways with MANOVA FDR<0.05 are shown in red while others are shaded gray.
+(C) Heatmap of 30 pathways with most extreme S distances (a surrogate of effect size).
+Red indicates higher methylation and blue shows lower methylation.
+(D) An example of a pathway identified with this joint enrichment analysis method, "Adrenoreceptors" shows lower methylation of
+member genes in both Estill and Novakovic studies.](img/fig7.svg)
 

### 2024-01-14 app/trial/app.R
progress on shiny

@@ -3,9 +3,10 @@ library("vroom")
 library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
 library("mitch")
 
-options(shiny.maxRequestSize = 100 * 1024^2) #100MB
+options(shiny.maxRequestSize = 300 * 1024^2) #300MB
 
-arraytype <- c("450k", "EPIC")
+arraytype <- c("EPIC","450k")
+
 genesettype <- c("Reactome","KEGG","GO")
 
 ui <- fluidPage(
@@ -13,6 +14,7 @@ ui <- fluidPage(
     "GMEA"
   ),
   sidebarPanel(
+    "Welcome to the GMEA analysis server. Optionally give your analysis a name, select your array platform, your preferred gene set annotation set, upload your limma data set and hit the analyse button. Please be patient as analysis might take a few minutes. Keep in mind that only TSV is supported at this time. Please ensure that the first column is the probe ID, and that there is one column with the heading 't'. This app doesn't default R row names, so please check that the sample t values shown match your data.",
     textInput("dataname", "Dataset name:"),
     radioButtons("arraytype", "Array platform:", arraytype),
     radioButtons("genesettype", "Gene set database:", genesettype),
@@ -21,9 +23,16 @@ ui <- fluidPage(
   ),
   mainPanel(
     textOutput("jobdata"),
+    "file information",
     tableOutput("fileinfo"),
-    tableOutput("head"),
-    dataTableOutput("mtable")
+    "t values",
+    tableOutput("tvals"),
+    "Top differentially methylated pathways",
+    dataTableOutput("mtable"),
+    "Download PDF report",
+    downloadButton("reportpdf","Generate PDF report"),
+    "Download HTML report",
+    downloadButton("reporthtml","Generate HTML report")
   )
 )
 
@@ -45,9 +54,7 @@ server <- function(input, output, session) {
     input$upload
   })
  
-  output$jobdata <- renderText(paste(mydataname(),myarraytype(),mygenesettype()))
-  output$fileinfo <- renderTable(fileinfo())
-
+  
   data <- reactive({
     req(input$upload)
     
@@ -57,10 +64,6 @@ server <- function(input, output, session) {
            tsv = vroom::vroom(input$upload$datapath, delim = "\t"),
[…]

### 2024-01-15 example_workflow/example_workflow.Rmd
example workflow

---
title: "Example of using mitch package for Infinium methylation analysis"
author: "The GMEA team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
---

## Introduction

Here we are looking at putting together the easiest possible workflow for the package.

Source code: https://github.com/markziemann/gmea/blob/main/example_workflow/example_workflow.Rmd

## Requirements

Load packages.

```{r,packages}

suppressPackageStartupMessages({
  library("limma")
  library("eulerr")
  library("IlluminaHumanMethylation450kanno.ilmn12.hg19")
  library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
  library("HGNChelper")
  library("tictoc")
  library("mitch")
  library("kableExtra")
  library("beeswarm")
  library("missMethyl")
  library("gridExtra")
  library("png")
})

CORES=2

```

## Load methylation data

The limma data table is read in.
This dataset is available from GEO (GSE49667).
The limma data was obtained using the "A cross-package Bioconductor workflow for analysing methylation array data" workflow.

```{r,loaddata}

dm <- read.csv("DMPs.csv.gz",header=TRUE)

head(dm)

```
## Load pathways

Reactome pathways were downloaded on the 14th Sept 2023 from MsigDB.

```{r,loadpathways}

gene_sets <- gmt_import("c5.go.v2023.2.Hs.symbols.gmt")

```

## Curate the annotation

Use all probes on the HM450K chip.

Updating gene names using the HGNC package because a lot of them have been updated over time.

An example of making a gene table for EPIC data is included down below.

```{r,anno1}

tic()
anno <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
myann <- data.frame(anno[,c("UCSC_RefGene_Name","UCSC_RefGene_Group","Islands_Name","Relation_to_Island")])
gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
names(gp2) <- rownames(gp)
gp2 <- lapply(gp2,unique)
gt1 <- stack(gp2)
colnames(gt1) <- c("gene","probe")
gt1$probe <- as.character(gt1$probe)
dim(gt1)
str(gt1)
length(unique(gt1$gene))
toc() #6.0s

tic()
#new.hgnc.table <- getCurrentHumanMap()
new.hgnc.table <- readRDS("../figs/new.hgnc.table.rds")
fix <- checkGeneSymbols(gt1$gene,map=new.hgnc.table)
fix2 <- fix[which(fix$x != fix$Suggested.Symbol),]
length(unique(fix2$x))
gt1$gene <- fix$Suggested.Symbol
toc() #  36 sec (2788 gene names updated)

head(gt1)

```

## Mitch pipeline

The first part is to import the data into mitch.

This only works properly when mitch's mapGeneIds function uses `mean` to aggregate (line 69 of mitch.R).

```{r,mg1}

# mitch aggregate function needs to be changes to `mean`
tic()
m <- mitch_import(x=dm,DEtype="limma",geneTable=gt1,geneIDcol="Name")
toc() #2.5 sec

head(m) %>%
  kbl(caption = "Example of differential gene methylation scores used for pathway analysis") %>%
  kable_paper("hover", full_width = F)

```

Now run the enrichment analysis.

Note that the results are not sorted by p-value, rather S.distance, an enrichment score.

```{r,mitch1}

tic()
mres <- mitch_calc(x=m,genesets=gene_sets,minsetsize=5,cores=4, priority="effect")
toc() #7 sec on 4 threads

mtable <- mres$enrichment_result
up <- subset(mtable,s.dist>0 & p.adjustANOVA<0.05)
dn <- subset(mtable,s.dist<0 & p.adjustANOVA<0.05)
nrow(up)
nrow(dn)

head(up,10) %>%
  kbl(caption = "Top significant pathways with higher methylation") %>%
  kable_paper("hover", full_width = F)

head(dn,10) %>%
  kbl(caption = "Top significant pathways with lower methylation") %>%
  kable_paper("hover", full_width = F)

top <- rbind(up[1:10,],dn[1:10,])
top <- top[order(top$s.dist),]

barnames <- gsub("_"," ",gsub("REACTOME_","",top$set))
cols <- as.character(((sign(top$s.dist)+1)/2)+1)
cols <- gsub("1","blue",gsub("2","red",cols))

par(mar = c(5.1, 29.1, 4.1, 2.1))
barplot(abs(top$s.dist),horiz=TRUE,las=1,names.arg=barnames, cex.names=0.8,
  cex.axis=0.8,col=cols, xlab="Enrichment score",main="Reactomes")
mtext("whole gene")
grid()

par( mar = c(5.1, 4.1, 4.1, 2.1) )

```

Make a html report and some charts.

```{r,report1}

mitch_report(res=mres,outfile="example_mitchreport.html",overwrite=TRUE)
mitch_plots(res=mres,outfile="example_mitchcharts.pdf")

```

## Make gene table for EPIC array data

```{r,gt2}

tic()
anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
myann <- data.frame(anno[,c("UCSC_RefGene_Name","UCSC_RefGene_Group","Islands_Name","Relation_to_Island")])
gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
names(gp2) <- rownames(gp)
gp2 <- lapply(gp2,unique)
gt <- stack(gp2)
colnames(gt) <- c("gene","probe")
gt$probe <- as.character(gt$probe)
dim(gt)
str(gt)
toc() #9.0s

tic()
#new.hgnc.table <- getCurrentHumanMap()
new.hgnc.table <- readRDS("../figsnew.hgnc.table.rds")
fix <- checkGeneSymbols(gt$gene,map=new.hgnc.table)
fix2 <- fix[which(fix$x != fix$Suggested.Symbol),]
length(unique(fix2$x))
gt$gene <- fix$Suggested.Symbol
toc()

```

## Session information

```{r,save}

sessionInfo()

```

### 2024-01-15 example_workflow/example_workflow.Rmd
bugfix

@@
 #new.hgnc.table <- getCurrentHumanMap()
-new.hgnc.table <- readRDS("../figsnew.hgnc.table.rds")
+new.hgnc.table <- readRDS("../figs/new.hgnc.table.rds")
 fix <- checkGeneSymbols(gt$gene,map=new.hgnc.table)

### 2024-01-15 manuscript/manuscript.Rmd
update ms

@@
 Severin Lamon<sup>3</sup>,
-Mary Kasu<sup>1</sup>,
-Mitch Norris<sup>1</sup>,
+Mary Sravya Kasu<sup>1</sup>,
+Mitch W. Norris<sup>1</sup>,
 Yen Ting Wong<sup>4</sup>,
@@
 | Severine Lamon | 0000-0002-3271-6551 |
+| Mary Sravya Kasu | 0000-0002-7891-836X |
 
@@
 Probes are ranked by t-statistic. The mean rank values for probes beloning to each gene are computed and used
-as input for a two-way, two-sample KruskalWallis test.
+as input for a two-way, two-sample ANOVA-on-ranks test.
 
@@
 8. Aggregate Limma Mitch (ALM).
-Similar to ALW, however uses the Mitch package to execute a two-way, two-sample Kruskal-Wallis test.
+Similar to ALW, however uses the Mitch package to execute a two-way, two-sample ANOVA-on-ranks test.
 
@@
 
+* Overall, we found that the FCS methods performed better across all simulation conditions.
+Of the different FCS methods tested, LAM was superior in terms of precision and recall.
+LAM also fits relatively easily into existing workflows, as most EWAS studies seem to use
+limma in their analyses of differential probe methylation.
+These analyses also consistently showed that non-parametric methods showed better recall with only a small decrease in precision, in accordance with some previous reports.
+Summarisation to gene level only takes a few seconds with the efficient base R aggregation command,
+and once aggregated, signatures can be analysed with existing packages like mitch.
+Mitch does not require a large amount of system memory and due to its parallel architecture can
+make use of multi-threaded processors, so it is not much slower than conducting ORA with existing methods.
+
+* But it was interesting that ORA performed better when gene sets are smaller.
+Given that KEGG and Reactome pathways consist of both large (>100) and small (<20) gene sets,
+a hybrid FCS and ORA approach might increase sensitivity and recall.
+
+* At 100 genes per set, LAM had superior sensitivity compared to GSA, which agrees with mitch's performance with simulated RNA-seq gene
+expression data[@Kaspi2020-yx].
+The improved sensitivity of this method provides a new opportunity for researchers to reanalyse previously conducted EWASs with contemporary pathway databases to better understand
+the subtle signatures, as we have demonstrated with the chronological age and assisted reproduction examples.
+The analyses conducted here used Reactome pathways, but this method is general purpose, so different gene sets can be used to examine various hypotheses.
+Transcription factor target gene sets can be used to identify transcription factors associated with changes in pathway methylation.
+MicroRNA target gene sets can identify potential associations between microRNA targets and DNA methylation patterns.
+The MSigDB resource contains these and several other types of gene sets for exploration [@Liberzon2015-um].
+
+* The differences in pathway results obtained with LAM and GSA from lung cancer data were striking, not only in the numbers of pathways identified, but also their relevance to the disease.
+Interestingly the relationship between promoter methylation and gene expression was not always inverse as we had anticipated.
+Joint enrichment analysis showed some genes with increasing promoter methylation together with increasing RNA expression (eg: "Negative regulation of AP2 transcription factors").
+Likewise a few pathways showed decreases in methylation with decreases in RNA expression such as "Dectin family", "TLR regulation" and "glycan biosynthesis termination".
+These results indicate that there is a complicated relationship between DNA methylation and gene expression in lung cancer and further work is required to appreciate the mechanisms involved.
+
+* We have demonstrated that this method works with both HM450K and EPIC array data and can even facilitate the comparison of EWAS using different platforms.
+We have provided an example workflow that researchers can use as a template.
+It shows how to generate a probe-gene table from Bioconductor annotation datasets, how to update the defunct gene names and how to run the enrichment analysis.
+Using this example, generating similar gene tables for the new EPIC 2.0 array and any future arrays will be straightforward.
+Updating defunct gene symbols in important as 2788 and 3253 gene symbols were updated on the HM450K and EPIC array annotations, which is approximately 10-14% of genes on these platforms.
+Omitting this step could lead to a loss of these genes from the pathway enrichment and a loss of sensitivity.
+
 ## Code availability
@@
 Code to generate the results shown here are included in a GitHub repository (https://github.com/markziemann/gmea/).
-To enable reproduction of the findings, a docker image is made available at DockerHub (https://hub.docker.com/repository/docker/mziemann/gmea/general).
+To enable reproduction of the findings, a docker image is made available at DockerHub
+(https://hub.docker.com/repository/docker/mziemann/gmea/general), and an example workflow can be found
+in the GitHub repository (https://github.com/markziemann/gmea/tree/main/example_workflow).
 

### 2024-01-15 example_workflow/example_workflow.Rmd
instructions
refining the example workflow

@@
 theme: cosmo
+bibliography: references.bib
+csl: epigenetics.csl
 ---
@@
 
-CORES=2
-
 ```
@@
 
-The limma data table is read in.
-This dataset is available from GEO (GSE49667).
-The limma data was obtained using the "A cross-package Bioconductor workflow for analysing methylation array data" workflow.
+The limma[@Ritchie2015-zn] data table is read in.
+This dataset is available from GEO (GSE49667)[@Zhang2013-oy].
+The limma data was obtained using a previously prepared workflow [@Maksimovic2017-zv].
 
@@
 ```
+
+Note that this object has probes with their own column, not simply as row names.
+
 ## Load pathways
 
-Reactome pathways were downloaded on the 14th Sept 2023 from MsigDB.
+Gene ontologies were downloaded in GMT format from MSigDB on 15th Jan 2024[@Liberzon2015-um;@The_Gene_Ontology_Consortium2023-fm].
+The GMT file is read into R using the mitch function `gmt_import()`.
 
@@
 
-Use all probes on the HM450K chip.
+One of the critical parts of this workflow is the establishment of probe-gene relationships.
+This controls how the probe data is aggregated to make the gene level scores.
 
-Updating gene names using the HGNC package because a lot of them have been updated over time.
+In this demonstration we are using all probes on the HM450K chip.
+Alternative approaches could use only promoters CpGs for example.
 
-An example of making a gene table for EPIC data is included down below.
+As these annotations are several years old, many of the annotated gene names are no longer
+current.
+To remedy this, the gene names are screened with the HGNChelper package and any defunct
+symbols get updated to the newer gene name, so they will be recognised properly in the
+gene sets.
+
+The process for making a gene table for EPIC data is included down below.
 
@@
 
-# mitch aggregate function needs to be changes to `mean`
 tic()
@@
 Note that the results are not sorted by p-value, rather S.distance, an enrichment score.
+I think this works better for interpretation.
 
@@
 tic()
-mres <- mitch_calc(x=m,genesets=gene_sets,minsetsize=5,cores=4, priority="effect")
-toc() #7 sec on 4 threads
+mres <- mitch_calc(x=m,genesets=gene_sets,minsetsize=5, priority="effect")
+toc() #7 secs
 
@@
 
+```
+
+Now make a barplot of these top findings.
+
+```{r, barplot}
+
 top <- rbind(up[1:10,],dn[1:10,])
@@
 
-barnames <- gsub("_"," ",gsub("REACTOME_","",top$set))
+barnames <- gsub("_"," ",top$set)
 cols <- as.character(((sign(top$s.dist)+1)/2)+1)
@@
 barplot(abs(top$s.dist),horiz=TRUE,las=1,names.arg=barnames, cex.names=0.8,
-  cex.axis=0.8,col=cols, xlab="Enrichment score",main="Reactomes")
+  cex.axis=0.8,col=cols, xlab="Enrichment score",main="Gene Ontologies")
 mtext("whole gene")
@@
 ```
+
+There are some interesting results there that give hints to the differences between naive T and Treg cells.
 
@@
 ## Make gene table for EPIC array data
+
+In case you are working with EPIC array data, the probe-gene table can be make with this snippet of code.
 
@@
 ```
+
+## Bibliography

### 2024-01-15 manuscript/manuscript.Rmd
example workflow online

@@
 To enable reproduction of the findings, a docker image is made available at DockerHub
-(https://hub.docker.com/repository/docker/mziemann/gmea/general), and an example workflow can be found
-in the GitHub repository (https://github.com/markziemann/gmea/tree/main/example_workflow).
+(https://hub.docker.com/repository/docker/mziemann/gmea/general).
+A reproducible example workflow report is hosted on our website (https://ziemann-lab.net/public/gmea/example_workflow.html).
 

### 2024-01-16 figs/multi_ewas_example.Rmd
fig8?

---
title: "Multi-EWAS enrichment example"
author: "The GMEA team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
---

## Introduction

Here we are looking at applying LAM/mitch to a multi-EWAS dataset.

https://pubmed.ncbi.nlm.nih.gov/37410739/

https://zenodo.org/records/8021411

## Requirements

```{r,packages}

suppressPackageStartupMessages({
  library("limma")
  library("eulerr")
  library("IlluminaHumanMethylation450kmanifest")
  library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
  library("HGNChelper")
  library("tictoc")
  library("mitch")
  library("gplots")
  library("kableExtra")
  library("beeswarm")
  library("missMethyl")
  library("gridExtra")
  library("png")
})

CORES=8

```

## Load data

Reactome pathways were downloaded on the 14th Sept 2023 from MsigDB.

We'll also score the probes by the signed -log pvalue

```{r,download1}

if(!dir.exists("hillary")){
  dir.create("hillary")
}

if(!file.exists("hillary/compressed-disease-states-ewas.tar.gz")){
  setwd("hillary")
  download.file("https://zenodo.org/records/8021411/files/compressed-disease-states-ewas.tar.gz?download=1",destfile="compressed-disease-states-ewas.tar.gz")
  untar("compressed-disease-states-ewas.tar.gz")
  setwd("..")
}

```

```{r,load}

myfiles <- list.files("hillary",pattern="prevalent_full",full.names=TRUE)
l1 <- lapply(myfiles, read.csv)
l1 <- lapply(l1,function(x) { x[,c("CpG","Beta"),drop=FALSE] } )
names(l1) <- gsub("_gs.csv","",gsub("hillary/prevalent_full_","",myfiles))
str(l1)

gs_symbols <- gmt_import("c2.cp.reactome.v2023.1.Hs.symbols.gmt")

```

## Curate the annotation

Use all probes on the chip.

Update old gene symbols.

```{r,anno1}

tic()
anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
myann <- data.frame(anno[,c("UCSC_RefGene_Name","UCSC_RefGene_Group","Islands_Name","Relation_to_Island")])
gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
names(gp2) <- rownames(gp)
gp2 <- lapply(gp2,unique)
gt <- stack(gp2)
colnames(gt) <- c("gene","probe")
gt$probe <- as.character(gt$probe)
dim(gt)
str(gt)
toc() #9.0s

tic()
#new.hgnc.table <- getCurrentHumanMap()
new.hgnc.table <- readRDS("new.hgnc.table.rds")
fix <- checkGeneSymbols(gt$gene,map=new.hgnc.table)
fix2 <- fix[which(fix$x != fix$Suggested.Symbol),]
length(unique(fix2$x))
gt$gene <- fix$Suggested.Symbol
toc()

```

## Mitch import

As we don't have the test statistic, we will use the delta beta value,
which according to preliminary investigations works better than the directional
p-value.

```{r,mg1}

tic()
m <- mitch_import(x=l1,DEtype="prescored",geneTable=gt,geneIDcol="CpG")
toc() #117 sec
head(m)

```

## Mitch calc

Multi-comparison.

```{r,mitch1}

tic()
mres <- mitch_calc(x=m,genesets=gs_symbols,minsetsize=5,cores=8, priority="effect")
toc() #21.3s

mtable <- mres$enrichment_result
head(mtable,20)

#numsig manova
nrow(subset(mtable,p.adjustMANOVA<0.05))
sig <- subset(mtable,p.adjustMANOVA<0.05)

```

Individual comparisons.

```{r,mitch2,fig.height=5,fig.width=5}

r1 <- lapply(l1,function(m) {
  m2 <- mitch_import(x=m,DEtype="prescored",geneTable=gt,geneIDcol="CpG")
  mres <- mitch_calc(x=m2,genesets=gs_symbols,minsetsize=5,cores=8, priority="effect")
  mtable <- mres$enrichment_result
})

nup <- lapply(r1,function(r) {
  ups <- subset(r,p.adjustANOVA<0.05 & s.dist>0)$set
  length(ups)
} )

ndn <- lapply(r1,function(r) {
  dns <- subset(r,p.adjustANOVA<0.05 & s.dist<0)$set
  length(dns)
} )

nums <- cbind(unlist(ndn),unlist(nup))
colnames(nums) <- c("dn","up")
nums

rownames(nums) <- gsub("_"," ",rownames(nums))

par(mar = c(5.1, 10.1, 4.1, 2.1))
barplot(t(nums),beside=TRUE,horiz=TRUE,las=1,
  legend.text = c("down","up"),xlab="no. pathways")
abline(v=seq(0,200,50),lty=2,lwd=0.5,col="gray")

pdf("fig8a.pdf",height=6,width=4)
par(mar = c(5.1, 10.1, 4.1, 2.1))
barplot(t(nums),beside=TRUE,horiz=TRUE,las=1,
  legend.text = c("down","up"),xlab="no. pathways")
abline(v=seq(0,200,50),lty=2,lwd=0.5,col="gray")
dev.off()

r2 <- lapply(r1,function(r) {
  ups <- head(subset(r,p.adjustANOVA<0.05 & s.dist > 0.3)$set,5)
  dns <- head(subset(r,p.adjustANOVA<0.05 & s.dist < -0.3)$set,5)
  list("ups"=ups,"dns"=dns)
} )

r2

gsets <- unique(unname(unlist(r2)))
gsets

par(mar = c(5.1, 4.1, 4.1, 2.1))

```

Make a heatmap with these gene sets.

```{r,mitchheat,fig.height=9,fig.width=9}

top <- mtable[which(mtable$set %in% gsets),]

top <- top[,c(1,4:17)]

rownames(top) <- top$set
top$set=NULL
head(top,2)

rownames(top) <- gsub("REACTOME_","",rownames(top))
rownames(top) <- gsub("_"," ",rownames(top))

colnames(top) <- gsub("^s.","",colnames(top))

colnames(top) <- gsub("_"," ",colnames(top))

#colfunc <- colorRampPalette(c("#B80F0A", "white", "#1134A6"))
colfunc <- colorRampPalette(c("red", "white", "blue"))

heatmap.2(as.matrix(top),scale="none",trace="none",margins=c(6,25),
  col=colfunc(25),cexRow=0.6,cexCol=0.8)

pdf("fig8b.pdf",height=9,width=7)
heatmap.2(as.matrix(top),scale="none",trace="none",margins=c(6,22),
  col=colfunc(25),cexRow=0.5,cexCol=0.8)
dev.off()

```

## Session information

```{r,save}

save.image("multi_ewas_example.Rdata")

sessionInfo()

```

### 2024-01-16 figs/multi_ewas_example.Rmd
fix color pal
incidence

@@
 
-myfiles <- list.files("hillary",pattern="prevalent_full",full.names=TRUE)
-l1 <- lapply(myfiles, read.csv)
+myfiles1 <- list.files("hillary",pattern="prevalent_full",full.names=TRUE)
+l1 <- lapply(myfiles1, read.csv)
 l1 <- lapply(l1,function(x) { x[,c("CpG","Beta"),drop=FALSE] } )
@@
 str(l1)
+
+myfiles2 <- list.files("hillary",pattern="incident_full",full.names=TRUE)
+l2 <- lapply(myfiles2, read.csv)
+l2 <- lapply(l2,function(x) { x[,c("CpG","Beta"),drop=FALSE] } )
+names(l2) <- gsub("_gs.csv","",gsub("hillary/incident_full_","",myfiles2))
+str(l2)
 
@@
 
+l1: prevalent
+
+l2: incident
+
 ```{r,mg1}
@@
 tic()
-m <- mitch_import(x=l1,DEtype="prescored",geneTable=gt,geneIDcol="CpG")
+m1 <- mitch_import(x=l1,DEtype="prescored",geneTable=gt,geneIDcol="CpG")
 toc() #117 sec
-head(m)
+head(m1)
+
+tic()
+m2 <- mitch_import(x=l2,DEtype="prescored",geneTable=gt,geneIDcol="CpG")
+toc() #158 sec
+head(m2)
 
@@
 
-## Mitch calc
+## Mitch calc prevalence
 
-Multi-comparison.
+Multi-comparison for prevalence.
 
-```{r,mitch1}
+```{r,mitch1prev}
 
 tic()
-mres <- mitch_calc(x=m,genesets=gs_symbols,minsetsize=5,cores=8, priority="effect")
+mres1 <- mitch_calc(x=m1,genesets=gs_symbols,minsetsize=5,cores=8, priority="effect")
 toc() #21.3s
 
-mtable <- mres$enrichment_result
-head(mtable,20)
+mtable1 <- mres1$enrichment_result
+head(mtable1,20)
 
 #numsig manova
-nrow(subset(mtable,p.adjustMANOVA<0.05))
-sig <- subset(mtable,p.adjustMANOVA<0.05)
+nrow(subset(mtable1,p.adjustMANOVA<0.05))
+sig <- subset(mtable1,p.adjustMANOVA<0.05)
 
@@
 
-```{r,mitch2,fig.height=5,fig.width=5}
+```{r,mitch2prev,fig.height=5,fig.width=5}
 
@@
 
-r2 <- lapply(r1,function(r) {
+r2 <- lapply(r2,function(r) {
   ups <- head(subset(r,p.adjustANOVA<0.05 & s.dist > 0.3)$set,5)
@@
 
-```{r,mitchheat,fig.height=9,fig.width=9}
+```{r,mitchheatprev,fig.height=9,fig.width=9}
 
-top <- mtable[which(mtable$set %in% gsets),]
-
+top <- mtable1[which(mtable1$set %in% gsets),]
 top <- top[,c(1,4:17)]
-
 rownames(top) <- top$set
@@
 rownames(top) <- gsub("_"," ",rownames(top))
-
 colnames(top) <- gsub("^s.","",colnames(top))
-
 colnames(top) <- gsub("_"," ",colnames(top))
-
-#colfunc <- colorRampPalette(c("#B80F0A", "white", "#1134A6"))
-colfunc <- colorRampPalette(c("red", "white", "blue"))
+colfunc <- colorRampPalette(c("blue", "white", "red"))
 
@@
 heatmap.2(as.matrix(top),scale="none",trace="none",margins=c(6,22),
+  col=colfunc(25),cexRow=0.5,cexCol=0.8)
+dev.off()
+
+```
+
+## Mitch calc incidence
+
+Multi-comparison for incidence.
+
+```{r,mitch1incidence}
+
+tic()
+mres2 <- mitch_calc(x=m2,genesets=gs_symbols,minsetsize=5,cores=8, priority="effect")
+toc() #29.5s
+
+mtable2 <- mres2$enrichment_result
+head(mtable2,20)
+
+#numsig manova
+nrow(subset(mtable2,p.adjustMANOVA<0.05))
+sig <- subset(mtable2,p.adjustMANOVA<0.05)
+
+```
+
+Individual comparisons.
+
+```{r,mitch2incidence,fig.height=5,fig.width=5}
+
+r1 <- lapply(l2,function(m) {
+  m2 <- mitch_import(x=m,DEtype="prescored",geneTable=gt,geneIDcol="CpG")
+  mres <- mitch_calc(x=m2,genesets=gs_symbols,minsetsize=5,cores=8, priority="effect")
+  mtable <- mres$enrichment_result
+})
+
+nup <- lapply(r1,function(r) {
+  ups <- subset(r,p.adjustANOVA<0.05 & s.dist>0)$set
+  length(ups)
+} )
+
+ndn <- lapply(r1,function(r) {
+  dns <- subset(r,p.adjustANOVA<0.05 & s.dist<0)$set
+  length(dns)
+} )
+
+nums <- cbind(unlist(ndn),unlist(nup))
+colnames(nums) <- c("dn","up")
+nums
+
+rownames(nums) <- gsub("ibd","IBD",rownames(nums))
+rownames(nums) <- gsub("_"," ",rownames(nums))
+
+par(mar = c(5.1, 10.1, 4.1, 2.1))
+barplot(t(nums),beside=TRUE,horiz=TRUE,las=1,
+  legend.text = c("down","up"),xlab="no. pathways")
+abline(v=seq(0,200,50),lty=2,lwd=0.5,col="gray")
+
+pdf("fig8c.pdf",height=6,width=4)
+par(mar = c(5.1, 10.1, 4.1, 2.1))
+barplot(t(nums),beside=TRUE,horiz=TRUE,las=1,
+  legend.text = c("down","up"),xlab="no. pathways")
+abline(v=seq(0,200,50),lty=2,lwd=0.5,col="gray")
+dev.off()
+
+r2 <- lapply(r1,function(r) {
+  ups <- head(subset(r,p.adjustANOVA<0.05 & s.dist > 0.3)$set,3)
+  dns <- head(subset(r,p.adjustANOVA<0.05 & s.dist < -0.3)$set,3)
+  list("ups"=ups,"dns"=dns)
+} )
+
+r2
+
+gsets <- unique(unname(unlist(r2)))
+gsets
+
+par(mar = c(5.1, 4.1, 4.1, 2.1))
+
+```
+
+Make a heatmap with these gene sets.
+
+```{r,mitchheatincidence,fig.height=9,fig.width=9}
+
+top <- mtable2[which(mtable2$set %in% gsets),]
+top <- top[,c(1,4:17)]
+rownames(top) <- top$set
+top$set=NULL
+head(top,2)
+
+rownames(top) <- gsub("REACTOME_","",rownames(top))
+rownames(top) <- gsub("_"," ",rownames(top))
+colnames(top) <- gsub("^s.","",colnames(top))
+colnames(top) <- gsub("_"," ",colnames(top))
+colnames(top) <- gsub("ibd","IBD",colnames(top))
+
+colfunc <- colorRampPalette(c("blue", "white", "red"))
+
+heatmap.2(as.matrix(top),scale="none",trace="none",margins=c(6,22),
+  col=colfunc(25),cexRow=0.6,cexCol=0.7)
+
+pdf("fig8d.pdf",height=9,width=7)
+heatmap.2(as.matrix(top),scale="none",trace="none",margins=c(6,20),
   col=colfunc(25),cexRow=0.5,cexCol=0.8)

### 2024-01-16 figs/multi_ewas_example.Rmd
bugfix

@@
 
-r2 <- lapply(r2,function(r) {
+r2 <- lapply(r1,function(r) {
   ups <- head(subset(r,p.adjustANOVA<0.05 & s.dist > 0.3)$set,5)

### 2024-01-16 figs/multi_ewas_example.Rmd
fix bug with missing conds
resize text

@@
 top <- mtable2[which(mtable2$set %in% gsets),]
-top <- top[,c(1,4:17)]
+top <- top[,c(1,4:22)]
 rownames(top) <- top$set
@@
 heatmap.2(as.matrix(top),scale="none",trace="none",margins=c(6,20),
-  col=colfunc(25),cexRow=0.5,cexCol=0.8)
+  col=colfunc(25),cexRow=0.5,cexCol=0.7)
 dev.off()

### 2024-01-17 manuscript/manuscript.Rmd
comment
work on ms

@@
 
+### Differential pathway methylation in 19 common disease states
+
+To examine whether LAM can identify differentially methylated pathways associated with disease prevalence and incidence,
+we downloaded blood based EWAS summary statistics from a recent publication [@Hillary2023-mu], made available via
+Zenodo (https://zenodo.org/records/8021411).
+The study used a logitudinal design, with sample collection baseline and diagnosis status established via electronic health
+record linkage up to 14 years after baseline.
+The study included 18,413 volunteers of European ancestry.
+We used summary statistics from the "full model", which corrected for relevant potential confounders.
+Disease status at baseline was used to classify participants into prevalence groups.
+Disease status at follow-up was used to classify participants into incidence groups.
+The study used the EPIC Infinium array and 752,722 probes were described in the summary statistic dataset.
+As limma t-statistics were not available, we used probe delta-beta values as input for LAM.
+LAM was conducted individually for each disease condition with Reactome gene sets.
+
 ## Results
@@
 
+## Differential pathway methylation in 19 common disease states
+
+To examine the potential for LAM to illuminate differential pathway methylation associations
+with diseases, we obtained and analysed blood based EWAS summary statistics from a recent
+publication [@Hillary2023-mu].
+This study examined 14 prevalent disease states and incidence of 19 disease states in a group
+of 18,413 participants, using the EPIC array.
+Delta-beta values were used for the purpose of scoring probe differential methylation, and these
+values underwent LAM analysis with Reactome gene sets.
+
+There were 899 differentially methylated pathways in the prevalence arm of the study (FDR<0.05) (**Figure 8A**).
+On average, there were 64 pathways with differential methylation in each prevalent condition, with
+Chronic obstructive pulmonary disease (COPD) having the most (263) and Alzheimer's disease having
+the fewest (3).
+We selected up to five statistically significant pathways with an absolute S distance of >0.4 in either
+direction in each prevalent condition to include in a heatmap (**Figure 8B**).
+We observed reduced methylation in the angiotensinogen metabolism pathway in patients with diabetes.
+Patients with colorectal cancer had lower methylation of the interleukin 10 signaling pathway.
+Reduced methylation of the salty taste perception pathway was observed in patients with CKD.
+
+![**Figure 8. Differential pathway methylation associated with prevalence of 14 common disease states.**
+(A) A barplot showing the number of statistically significant pathways with higher and lower methylation identified in each prevalent condition (FDR<0.05).
+(B) A heatmap of s scores for selected pathways across 14 common prevalent disease states.
+Stars indicate that the pathway was identified as being among the top five differentially methylated pathways in each direction for each condition.
+](img/fig8.svg)
+
+We also wanted to know whether the LAM method could identify signatures that appear before patients are diagnosed
+with a condition.
+So we examined differential pathway methylation associated with incidence of 19 disease states.
+Over all 19 conditions, there were 1570 significant pathways (FDR<0.05) (**Figure 9A**).
+Liver cirrhosis has the most, with 464, followed by COPD with 400, while Alzheimer's has just one and COVID-19
+hospitalisation had none.
+A heatmap shows the enrichment scores for selected pathways across these incident conditions (**Figure 9B**).
+Breast cancer incidence was associated with reduced methylation to the metal sequestration pathway.
+Incident prostate cancer was associated with reduced methylation of the FGFR1 pathway.
+Incident Parkinson's disease was associated with reduced methylation of aquaporins.
+In COPD incidence, higher methylation of carbohydrate binding ficolins was observed together with reduced methylation
+of the GDP mannose synthesis pathway.
+Incident stroke was associated with higher methylation of the lipoxin synthesis pathway.
+These results indicate that LAM can identify differentially methylated pathways in disease groups, and
+detect altered methylation before a diagnosis has been made.
+
+![**Figure 9. Differential pathway methylation associated with incidence of 19 common disease states.**
+(A) A barplot showing the number of statistically significant pathways with higher and lower methylation identified in each incident condition (FDR<0.05).
+(B) A heatmap of s scores for selected pathways across 19 common incident disease states.
+Stars indicate that the pathway was identified as being among the top three differentially methylated pathways in each direction for each condition.
+](img/fig9.svg)
+
 ## Discussion
@@
 These results indicate that there is a complicated relationship between DNA methylation and gene expression in lung cancer and further work is required to appreciate the mechanisms involved.
-
 * We have demonstrated that this method works with both HM450K and EPIC array data and can even facilitate the comparison of EWAS using different platforms.
@@
 It shows how to generate a probe-gene table from Bioconductor annotation datasets, how to update the defunct gene names and how to run the enrichment analysis.
-Using this example, generating similar gene tables for the new EPIC 2.0 array and any future arrays will be straightforward.
-Updating defunct gene symbols in important as 2788 and 3253 gene symbols were updated on the HM450K and EPIC array annotations, which is approximately 10-14% of genes on these platforms.
-Omitting this step could lead to a loss of these genes from the pathway enrichment and a loss of sensitivity.
+Using this example, generating similar gene tables for the new EPIC 2.0 array and any future arrays will be straightforward, once their annotation sets are available through Bioconductor.
+Updating defunct gene symbols is an important step as 2788 and 3253 gene symbols were updated on the HM450K and EPIC array annotations, which is approximately 10-14% of genes on these platforms.
+We note that other packages for enrichment of Infinium array data to not by default address this, which could lead to a loss of these genes from pathway enrichment and a loss of sensitivity.
+* The analysis of the large scale disease EWAS shows that LAM can be readily applied to identify differential pathway methylation associated with prevalent and incident disease.
+The authors of the study used GOmeth for their pathway analysis and from the 33 models examined, significant findings were obtained for only two, prevalent diabetes and heart disease.
+On contrast, LAM identified
+Such work is important as it is known that methylation changes can be caused by disease states.
+For example metabolic changes in diabetes are probably the cause of the observed methylation signatures.
+It is perhaps unexpected that more significant pathways be identified from the incident disease arm of the study,
+
+* This work could be extended to include not just proximal CpG sites, but also enhancer based probes and their target genes based on resources such as GeneHancer.
 

### 2024-01-17 figs/multi_ewas_example.Rmd
randomisation test

@@
 
+## Randomise data
+
+To demonstrate that mitch is robust to false positives, we can perform
+randomisation of the data.
+
+```{r,randomise}
+
+fres <- lapply(1:100,function(i){
+  seed=i*1000
+  z <- lapply(l2,function(m) {
+    m=l2[[1]]
+    set.seed=sum(m$Beta)*seed ; m$Beta <- sample(x=m$Beta,replace=FALSE,size=752722)
+    m2 <- mitch_import(x=m,DEtype="prescored",geneTable=gt,geneIDcol="CpG")
+    mres <- mitch_calc(x=m2,genesets=gs_symbols,minsetsize=5,cores=1, priority="effect")
+    mtable <- mres$enrichment_result
+    nrow(subset(mtable,p.adjustANOVA<0.05))
+  })
+  sum(unlist(z))
+})
+
+fres <- unlist(fres)
+summary(fres)
+
+```
+
 ## Session information

### 2024-01-18 manuscript/manuscript.Rmd
ms progress

@@
 Incident stroke was associated with higher methylation of the lipoxin synthesis pathway.
-These results indicate that LAM can identify differentially methylated pathways in disease groups, and
-detect altered methylation before a diagnosis has been made.
 
@@
 
+To show that these findings are not the result of false positives from the LAM method, we randomised the
+incidence profiles for all 19 conditions prior to enrichment analysis.
+This was repeated 100 times, and these analyses of randimised data yielded few significant pathways.
+Specifically, from 100 runs 64 yielded no significant pathways across 19 conditions.
+Twenty eight repeats had fewer than five false positives and eight runs had more than five false positives.
+Across the 100 repeats, the mean number of false positives was 1.95 and median was 0.
+These results indicate that LAM can identify differentially methylated pathways in disease groups, and
+detect altered methylation before a diagnosis has been made, with few false positives.
+
 ## Discussion
@@
 limma in their analyses of differential probe methylation.
-These analyses also consistently showed that non-parametric methods showed better recall with only a small decrease in precision, in accordance with some previous reports.
-Summarisation to gene level only takes a few seconds with the efficient base R aggregation command,
+The ALM method had only slightly worse performance as compared to LAM, but ALM requires users to aggregate the probe methylation values
+for each sample for each gene prior to limma.
+This extra aggregation step is computationally intensive and not a standard part of existing methylation analysis workflows.
+
+Our analyses also consistently showed that non-parametric methods showed better recall with only a small decrease in precision, in accordance with some previous reports.
+Summarisation of probe level t-statistics to gene level only takes a few seconds with the efficient base R `aggregate` command,
 and once aggregated, signatures can be analysed with existing packages like mitch.
 Mitch does not require a large amount of system memory and due to its parallel architecture can
-make use of multi-threaded processors, so it is not much slower than conducting ORA with existing methods.
+make use of multi-threaded processors, so it is not much slower than conducting ORA with existing methods that are single threaded.
 
-* But it was interesting that ORA performed better when gene sets are smaller.
-Given that KEGG and Reactome pathways consist of both large (>100) and small (<20) gene sets,
+* An interesting observation is that ORA performed better when gene sets are smaller.
+Given that popular pathway databases like KEGG and Reactome pathways consist of both large (>100) and small (<20) gene sets,
 a hybrid FCS and ORA approach might increase sensitivity and recall.
 
-* At 100 genes per set, LAM had superior sensitivity compared to GSA, which agrees with mitch's performance with simulated RNA-seq gene
-expression data[@Kaspi2020-yx].
-The improved sensitivity of this method provides a new opportunity for researchers to reanalyse previously conducted EWASs with contemporary pathway databases to better understand
-the subtle signatures, as we have demonstrated with the chronological age and assisted reproduction examples.
-The analyses conducted here used Reactome pathways, but this method is general purpose, so different gene sets can be used to examine various hypotheses.
+* At 100 genes per set, LAM had superior sensitivity compared to GSA, which agrees with a prevous report of mitch's performance with
+simulated RNA-seq gene expression data[@Kaspi2020-yx].
+The improved sensitivity of the LAM method provides a new opportunity for researchers to reanalyse previously conducted EWASs with contemporary pathway databases to better understand
+the subtle signatures, as we have demonstrated with the chronological age, assisted reproduction examples and 19 disease EWAS analyses.
+The analyses conducted here used Reactome pathways, but LAM is general purpose, so different gene sets can be used to examine various hypotheses.
 Transcription factor target gene sets can be used to identify transcription factors associated with changes in pathway methylation.
@@
 Likewise a few pathways showed decreases in methylation with decreases in RNA expression such as "Dectin family", "TLR regulation" and "glycan biosynthesis termination".
-These results indicate that there is a complicated relationship between DNA methylation and gene expression in lung cancer and further work is required to appreciate the mechanisms involved.
+These results indicate that there is a complicated relationship between DNA methylation and gene expression in lung cancer and further work is required to appreciate the mechanisms involved and distinguish which CpGs are critical for influencing gene expression and which are not.
+
 * We have demonstrated that this method works with both HM450K and EPIC array data and can even facilitate the comparison of EWAS using different platforms.
@@
 It shows how to generate a probe-gene table from Bioconductor annotation datasets, how to update the defunct gene names and how to run the enrichment analysis.
-Using this example, generating similar gene tables for the new EPIC 2.0 array and any future arrays will be straightforward, once their annotation sets are available through Bioconductor.
-Updating defunct gene symbols is an important step as 2788 and 3253 gene symbols were updated on the HM450K and EPIC array annotations, which is approximately 10-14% of genes on these platforms.
-We note that other packages for enrichment of Infinium array data to not by default address this, which could lead to a loss of these genes from pathway enrichment and a loss of sensitivity.
+Using this example, generating similar gene tables for the new Infinium Mouse and EPIC 2.0 arrays and any future arrays will be straightforward, once their annotation sets are available through Bioconductor.
+
+* Updating defunct gene symbols is an important step as 2788 and 3253 gene symbols were updated on the HM450K and EPIC array annotations, which is approximately 10-14% of genes on these platforms.
+We note that other packages for enrichment of Infinium array data do not by default address this, which could lead to a loss of these genes from pathway enrichment and diminished sensitivity.
+
 * The analysis of the large scale disease EWAS shows that LAM can be readily applied to identify differential pathway methylation associated with prevalent and incident disease.
 The authors of the study used GOmeth for their pathway analysis and from the 33 models examined, significant findings were obtained for only two, prevalent diabetes and heart disease.
-On contrast, LAM identified
-Such work is important as it is known that methylation changes can be caused by disease states.
+On contrast, LAM identified 2,469 pathway associations (74.8 per model).
+To demonstrate these are not the results of false positives, permutated profiles yielded on average 0.10 significant findings per model,
+supporting the idea that LAM is identifying real pathway signatures.
+
+* Distinguishing prevalent and incident signatures is important as it is known that methylation changes can be caused by disease states.
 For example metabolic changes in diabetes are probably the cause of the observed methylation signatures.

### 2024-01-18 manuscript/manuscript.Rmd
ms work

@@
 Anusuiya Bora<sup>1,2</sup>,
-Severin Lamon<sup>3</sup>,
+Severine Lamon<sup>3</sup>,
 Mary Sravya Kasu<sup>1</sup>,
@@
 LAM was conducted individually for each disease condition with Reactome gene sets.
+Finally, as a test of type I errors, we randomised each of the 19 incident profiles using
+base R `sample` command with a set seed, followed by LAM enrichment and counting the number
+of significant pathways.
+This process was repeated 100 times with unique seeds.
 
@@
 incidence profiles for all 19 conditions prior to enrichment analysis.
-This was repeated 100 times, and these analyses of randimised data yielded few significant pathways.
+This was repeated 100 times, and these analyses of randomised data yielded few significant pathways.
 Specifically, from 100 runs 64 yielded no significant pathways across 19 conditions.
@@
 
-* Overall, we found that the FCS methods performed better across all simulation conditions.
+Overall, we found that the FCS methods performed better across all simulation conditions.
 Of the different FCS methods tested, LAM was superior in terms of precision and recall.
@@
 
-* An interesting observation is that ORA performed better when gene sets are smaller.
+An interesting observation is that ORA performed better when gene sets are smaller.
 Given that popular pathway databases like KEGG and Reactome pathways consist of both large (>100) and small (<20) gene sets,
@@
 
-* At 100 genes per set, LAM had superior sensitivity compared to GSA, which agrees with a prevous report of mitch's performance with
+At 100 genes per set, LAM had superior sensitivity compared to GSA, which agrees with a prevous report of mitch's performance with
 simulated RNA-seq gene expression data[@Kaspi2020-yx].
@@
 
-* The differences in pathway results obtained with LAM and GSA from lung cancer data were striking, not only in the numbers of pathways identified, but also their relevance to the disease.
+The differences in pathway results obtained with LAM and GSA from lung cancer data were striking, not only in the numbers of pathways identified, but also their relevance to the disease.
 Interestingly the relationship between promoter methylation and gene expression was not always inverse as we had anticipated.
@@
 
-* We have demonstrated that this method works with both HM450K and EPIC array data and can even facilitate the comparison of EWAS using different platforms.
+We have demonstrated that this method works with both HM450K and EPIC array data and can even facilitate the comparison of EWAS using different platforms.
 We have provided an example workflow that researchers can use as a template.
@@
 
-* Updating defunct gene symbols is an important step as 2788 and 3253 gene symbols were updated on the HM450K and EPIC array annotations, which is approximately 10-14% of genes on these platforms.
+Updating defunct gene symbols is an important step as 2,788 and 3,253 gene symbols were updated on the HM450K and EPIC array annotations, which is approximately 10-14% of genes on these platforms.
 We note that other packages for enrichment of Infinium array data do not by default address this, which could lead to a loss of these genes from pathway enrichment and diminished sensitivity.
 
-* The analysis of the large scale disease EWAS shows that LAM can be readily applied to identify differential pathway methylation associated with prevalent and incident disease.
+The analysis of the large scale disease EWAS shows that LAM can be readily applied to identify differential pathway methylation associated with prevalent and incident disease.
 The authors of the study used GOmeth for their pathway analysis and from the 33 models examined, significant findings were obtained for only two, prevalent diabetes and heart disease.
@@
 
-* Distinguishing prevalent and incident signatures is important as it is known that methylation changes can be caused by disease states.
+Distinguishing prevalent and incident signatures is important as it is known that methylation changes can be caused by disease states.
 For example metabolic changes in diabetes are probably the cause of the observed methylation signatures.
@@
 
-* This work could be extended to include not just proximal CpG sites, but also enhancer based probes and their target genes based on resources such as GeneHancer.
+This work could be extended to include not just proximal CpG sites, but also enhancer based probes and their target genes based on resources such as GeneHancer.
+
+There is also potential to generate new libraries of differentially methylated probe and gene sets, to contribute to the pool of molecular signatures in public resources like MSigDB,
+which will assist in understanding similarities between methylation profiles.
 
@@
 
+## Acknowledgements
+
+This research was supported by use of the Nectar Research Cloud, a collaborative Australian research platform supported by the NCRIS-funded Australian Research Data Commons (ARDC).
+Authors gratefully acknowledge the contribution to this work of the Victorian Operational Infrastructure Support Program received by the Burnet Institute.
+We thank the researchers who publicly shared data to enable this research.
+
 ## Bibliography

### 2024-01-18 figs/GSE158422_simulate100_rslurm.Rmd
try larger gene set size on slurm cluster
source

--- 2024-01-06 figs/GSE158422_simulate020_rslurm.Rmd
@@
 ---
+
+Source: https://github.com/markziemann/gmea/
 
@@
 # set gene set size here
-lengths <- rep(20,1000)
+lengths <- rep(100,1000)
 
@@
 
-save.image("GSE158422_simulate020_rslurm.Rdata")
+save.image("GSE158422_simulate100_rslurm.Rdata")
 

### 2024-01-18 figs/multi_ewas_example.Rmd
source
make reports

@@
 
-r1 <- lapply(l1,function(m) {
+r1 <- lapply(1:length(l1),function(i) {
+  name=names(l1[i])
+  m <- l1[[i]]
   m2 <- mitch_import(x=m,DEtype="prescored",geneTable=gt,geneIDcol="CpG")
-  mres <- mitch_calc(x=m2,genesets=gs_symbols,minsetsize=5,cores=8, priority="effect")
+  mres <- mitch_calc(x=m2,genesets=gs_symbols,minsetsize=5,cores=2, priority="effect")
   mtable <- mres$enrichment_result
+  #mitch_report(res=mres,outfile=paste("mitchreport_prev_",name,".html",sep=""))
 })
@@
 
-r1 <- lapply(l2,function(m) {
+r1 <- lapply(1:length(l2),function(m) {
+  name=names(l2[i])
+  m <- l2[[i]]
   m2 <- mitch_import(x=m,DEtype="prescored",geneTable=gt,geneIDcol="CpG")
-  mres <- mitch_calc(x=m2,genesets=gs_symbols,minsetsize=5,cores=8, priority="effect")
+  mres <- mitch_calc(x=m2,genesets=gs_symbols,minsetsize=5,cores=2, priority="effect")
   mtable <- mres$enrichment_result
+  #mitch_report(res=mres,outfile=paste("mitchreport_incid_",name,".html",sep=""))
 })

### 2024-01-18 figs/multi_ewas_example.Rmd
bugfix

@@
 
-r1 <- lapply(1:length(l2),function(m) {
+r1 <- lapply(1:length(l2),function(i) {
   name=names(l2[i])

### 2024-01-18 figs/multi_ewas_example.Rmd
changed prioritisaion

--- 2024-01-17 figs/multi_ewas_example.Rmd
@@
 
-r1 <- lapply(l1,function(m) {
+r1 <- lapply(1:length(l1),function(i) {
+  name=names(l1[i])
+  m <- l1[[i]]
   m2 <- mitch_import(x=m,DEtype="prescored",geneTable=gt,geneIDcol="CpG")
-  mres <- mitch_calc(x=m2,genesets=gs_symbols,minsetsize=5,cores=8, priority="effect")
+  mres <- mitch_calc(x=m2,genesets=gs_symbols,minsetsize=5,cores=2, priority="significance")
   mtable <- mres$enrichment_result
+  mitch_report(res=mres,outfile=paste("mitchreport_prev_",name,".html",sep="",overwrite=TRUE))
 })
@@
 
-r1 <- lapply(l2,function(m) {
+r1 <- lapply(1:length(l2),function(i) {
+  name=names(l2[i])
+  m <- l2[[i]]
   m2 <- mitch_import(x=m,DEtype="prescored",geneTable=gt,geneIDcol="CpG")
-  mres <- mitch_calc(x=m2,genesets=gs_symbols,minsetsize=5,cores=8, priority="effect")
+  mres <- mitch_calc(x=m2,genesets=gs_symbols,minsetsize=5,cores=2, priority="significance")
   mtable <- mres$enrichment_result
+  #mitch_report(res=mres,outfile=paste("mitchreport_incid_",name,".html",sep="",overwrite=TRUE))
 })

### 2024-01-18 figs/multi_ewas_example.Rmd
bugfix

@@
 l1 <- lapply(l1,function(x) { x[,c("CpG","Beta"),drop=FALSE] } )
-names(l1) <- gsub("_gs.csv","",gsub("hillary/prevalent_full_","",myfiles))
+names(l1) <- gsub("_gs.csv","",gsub("hillary/prevalent_full_","",myfiles1))
 str(l1)

### 2024-01-18 figs/multi_ewas_example.Rmd
reports not needed

@@
   mtable <- mres$enrichment_result
-  mitch_report(res=mres,outfile=paste("mitchreport_prev_",name,".html",sep="",overwrite=TRUE))
+  #mitch_report(res=mres,outfile=paste("mitchreport_prev_",name,".html",sep="",overwrite=TRUE))
 })

### 2024-01-22 app/trial/app.R
shiny

@@ -19,20 +19,18 @@ ui <- fluidPage(
     radioButtons("arraytype", "Array platform:", arraytype),
     radioButtons("genesettype", "Gene set database:", genesettype),
     fileInput("upload", multiple = FALSE,label="Upload limma data:"),
+    downloadLink("downloadData", "Download full table of results"),
     actionButton("analyse", "Analyse!",icon = icon("gears"))
   ),
   mainPanel(
+    "This area will populate with results once analysis is complete.",
     textOutput("jobdata"),
     "file information",
     tableOutput("fileinfo"),
     "t values",
     tableOutput("tvals"),
     "Top differentially methylated pathways",
-    dataTableOutput("mtable"),
-    "Download PDF report",
-    downloadButton("reportpdf","Generate PDF report"),
-    "Download HTML report",
-    downloadButton("reporthtml","Generate HTML report")
+    dataTableOutput("mtable")
   )
 )
 
@@ -100,20 +98,6 @@ server <- function(input, output, session) {
         myres$enrichment_result
   })
   
-  reportpdf <- reactive({
-    req(mres())
-    mitch_plots(mres(),outfile = "plots.pdf")
-    pl <- "plots.pdf"
-    readBin(pl, raw(), file.info(pl)$size)
-  })
-  
-  reporthtml <- reactive({
-    req(mres())
-    mitch_report(mres(),outfile = "report.html",overwrite = TRUE)
-    htm <- "report.html"
-    readBin(htm, raw(), file.info(htm)$size)
-  })
-  
   output$jobdata <- renderText(paste(mydataname(),myarraytype(),mygenesettype()))
   
   output$fileinfo <- renderTable(fileinfo())
@@ -127,18 +111,19 @@ server <- function(input, output, session) {
     head(m)
   })
   
-  output$mtable <- renderDataTable(mtable(),options = list(pageLength = 10, info = FALSE))
-
-  # report download not working
-  output$reportpdf <- downloadHandler(
-    # For PDF output, change this to "report.pdf"
-    filename = "report.pdf",
-    content = reportpdf()
+  output$mtable <- renderDataTable(
+    mtable(),options = list(pageLength = 10, info = FALSE)
   )
-  output$reporthtml <- downloadHandler(
-    # For PDF output, change this to "report.pdf"
-    filename = "report.html",
-    content = reporthtml()
+
+  output$downloadData <- downloadHandler(
+    filename = function() {
[…]

### 2024-01-22 app/trial/app.R
basic version working

@@ -121,7 +121,7 @@ server <- function(input, output, session) {
     },
     content = function(file) {
       req(mtable())
-      write.table(mtable(),file,sep="\t")
+      write.table(mtable(),file,sep="\t",row.names = FALSE)
       #write.csv(mtcars, file)
     }
   )

### 2024-01-22 app/trial/app.R
changes

@@ -14,12 +14,13 @@ ui <- fluidPage(
     "GMEA"
   ),
   sidebarPanel(
-    "Welcome to the GMEA analysis server. Optionally give your analysis a name, select your array platform, your preferred gene set annotation set, upload your limma data set and hit the analyse button. Please be patient as analysis might take a few minutes. Keep in mind that only TSV is supported at this time. Please ensure that the first column is the probe ID, and that there is one column with the heading 't'. This app doesn't default R row names, so please check that the sample t values shown match your data.",
+    "Welcome to the GMEA analysis server. Optionally give your analysis a name, select your array platform, your preferred gene set annotation set, upload your limma data set and hit the analyse button. Please be patient as analysis might take a few minutes. Keep in mind that only TSV is supported at this time. Please ensure that the first column is the probe ID, and that there is one column with the heading 't'. This app doesn't recognise default R row names, so please check that the sample t values shown match your data. After selecting the file to analyze, hit the 'Analyse!' button and once the top results are shown go ahead and download the full table and the HTML report.",
     textInput("dataname", "Dataset name:"),
     radioButtons("arraytype", "Array platform:", arraytype),
     radioButtons("genesettype", "Gene set database:", genesettype),
     fileInput("upload", multiple = FALSE,label="Upload limma data:"),
-    downloadLink("downloadData", "Download full table of results"),
+    downloadButton("downloadData", label = "Download full results table"),
+    downloadButton("downloadData2", label = "Download HTML report"),
     actionButton("analyse", "Analyse!",icon = icon("gears"))
   ),
   mainPanel(
@@ -30,7 +31,8 @@ ui <- fluidPage(
     "t values",
     tableOutput("tvals"),
     "Top differentially methylated pathways",
-    dataTableOutput("mtable")
+    dataTableOutput("mtable"),
+    
   )
 )
 
@@ -80,8 +82,17 @@ server <- function(input, output, session) {
   
   genesets <- reactive({
     if((mygenesettype())  == "Reactome") {
-      gmt_import("c2.cp.reactome.v2023.1.Hs.symbols.gmt")
+      gmt_import("~/gmea/app/trial/c2.cp.reactome.v2023.2.Hs.symbols.gmt")
     }
+    
[…]

### 2024-01-22 app/trial/app.R
updates

@@ -14,17 +14,14 @@ ui <- fluidPage(
     "GMEA"
   ),
   sidebarPanel(
-    "Welcome to the GMEA analysis server. Optionally give your analysis a name, select your array platform, your preferred gene set annotation set, upload your limma data set and hit the analyse button. Please be patient as analysis might take a few minutes. Keep in mind that only TSV is supported at this time. Please ensure that the first column is the probe ID, and that there is one column with the heading 't'. This app doesn't recognise default R row names, so please check that the sample t values shown match your data. After selecting the file to analyze, hit the 'Analyse!' button and once the top results are shown go ahead and download the full table and the HTML report.",
+    "Welcome to the GMEA analysis server. Optionally give your analysis a name, select your array platform, your preferred gene set annotation set, upload your limma data set and hit the analyse button. Please be patient as analysis might take a few minutes. Keep in mind that only TSV is supported at this time. Please ensure that the first column is the probe ID, and that there is one column with the heading 't'. This app doesn't default R row names, so please check that the sample t values shown match your data.",
     textInput("dataname", "Dataset name:"),
     radioButtons("arraytype", "Array platform:", arraytype),
     radioButtons("genesettype", "Gene set database:", genesettype),
     fileInput("upload", multiple = FALSE,label="Upload limma data:"),
-    downloadButton("downloadData", label = "Download full results table"),
-    downloadButton("downloadData2", label = "Download HTML report"),
     actionButton("analyse", "Analyse!",icon = icon("gears"))
   ),
   mainPanel(
-    "This area will populate with results once analysis is complete.",
     textOutput("jobdata"),
     "file information",
     tableOutput("fileinfo"),
@@ -32,7 +29,10 @@ ui <- fluidPage(
     tableOutput("tvals"),
     "Top differentially methylated pathways",
     dataTableOutput("mtable"),
-    
+    "Download PDF report",
+    downloadButton("reportpdf","Generate PDF report"),
+    "Download HTML report",
+    downloadButton("reporthtml","Generate HTML report")
   )
 )
 
[…]

### 2024-01-22 app/trial/app.R
rollback

@@ -14,14 +14,17 @@ ui <- fluidPage(
     "GMEA"
   ),
   sidebarPanel(
-    "Welcome to the GMEA analysis server. Optionally give your analysis a name, select your array platform, your preferred gene set annotation set, upload your limma data set and hit the analyse button. Please be patient as analysis might take a few minutes. Keep in mind that only TSV is supported at this time. Please ensure that the first column is the probe ID, and that there is one column with the heading 't'. This app doesn't default R row names, so please check that the sample t values shown match your data.",
+    "Welcome to the GMEA analysis server. Optionally give your analysis a name, select your array platform, your preferred gene set annotation set, upload your limma data set and hit the analyse button. Please be patient as analysis might take a few minutes. Keep in mind that only TSV is supported at this time. Please ensure that the first column is the probe ID, and that there is one column with the heading 't'. This app doesn't recognise default R row names, so please check that the sample t values shown match your data. After selecting the file to analyze, hit the 'Analyse!' button and once the top results are shown go ahead and download the full table and the HTML report.",
     textInput("dataname", "Dataset name:"),
     radioButtons("arraytype", "Array platform:", arraytype),
     radioButtons("genesettype", "Gene set database:", genesettype),
     fileInput("upload", multiple = FALSE,label="Upload limma data:"),
+    downloadButton("downloadData", label = "Download full results table"),
+    downloadButton("downloadData2", label = "Download HTML report"),
     actionButton("analyse", "Analyse!",icon = icon("gears"))
   ),
   mainPanel(
+    "This area will populate with results once analysis is complete.",
     textOutput("jobdata"),
     "file information",
     tableOutput("fileinfo"),
@@ -29,10 +32,7 @@ ui <- fluidPage(
     tableOutput("tvals"),
     "Top differentially methylated pathways",
     dataTableOutput("mtable"),
-    "Download PDF report",
-    downloadButton("reportpdf","Generate PDF report"),
-    "Download HTML report",
-    downloadButton("reporthtml","Generate HTML report")
+    
   )
 )
 
[…]

### 2024-01-22 app/trial/app.R
prioritisation and gene set library

@@ -7,7 +7,9 @@ options(shiny.maxRequestSize = 300 * 1024^2) #300MB
 
 arraytype <- c("EPIC","450k")
 
-genesettype <- c("Reactome","KEGG","GO")
+genesettype <- c("Reactome","KEGG MEDICUS","GO")
+
+prioritisationtype <- c("Significance","Enrichment score")
 
 ui <- fluidPage(
   titlePanel(
@@ -18,6 +20,7 @@ ui <- fluidPage(
     textInput("dataname", "Dataset name:"),
     radioButtons("arraytype", "Array platform:", arraytype),
     radioButtons("genesettype", "Gene set database:", genesettype),
+    radioButtons("prioritisationtype", "Prioritisation:", prioritisationtype),
     fileInput("upload", multiple = FALSE,label="Upload limma data:"),
     downloadButton("downloadData", label = "Download full results table"),
     downloadButton("downloadData2", label = "Download HTML report"),
@@ -49,12 +52,16 @@ server <- function(input, output, session) {
   mygenesettype <- eventReactive(input$analyse, {
     as.character(input$genesettype)
   })
+  
+  myprioritisationtype <- eventReactive(input$analyse, {
+    as.character(input$prioritisationtype)
+  })  
 
   fileinfo <- eventReactive(input$analyse, {
     input$upload
   })
  
-  
+
   data <- reactive({
     req(input$upload)
     
@@ -82,17 +89,26 @@ server <- function(input, output, session) {
   
   genesets <- reactive({
     if((mygenesettype())  == "Reactome") {
-      gmt_import("~/gmea/app/trial/c2.cp.reactome.v2023.2.Hs.symbols.gmt")
+      gs <- gmt_import("~/gmea/app/trial/c2.cp.reactome.v2023.2.Hs.symbols.gmt")
     }
     
     if((mygenesettype())  == "KEGG") {
-      gmt_import("~/gmea/app/trial/c2.cp.kegg_medicus.v2023.2.Hs.symbols.gmt")
+      gs <- gmt_import("~/gmea/app/trial/c2.cp.kegg_medicus.v2023.2.Hs.symbols.gmt")
     }
     
     if((mygenesettype())  == "GO") {
-      gmt_import("~/gmea/app/trial/c5.all.v2023.2.Hs.symbols.gmt")
+      gs <- gmt_import("~/gmea/app/trial/c5.all.v2023.2.Hs.symbols.gmt")
     }
-    
+    gs
+  })
+  
[…]

### 2024-01-22 app/make_genetable.R
gene table 450k and epic

@@ -1,0 +1,45 @@
+library("IlluminaHumanMethylation450kanno.ilmn12.hg19")
+library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
+library("HGNChelper")
+
+# HM450K
+anno <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
+myann <- data.frame(anno[,c("UCSC_RefGene_Name","UCSC_RefGene_Group","Islands_Name","Relation_to_Island")])
+gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
+gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
+names(gp2) <- rownames(gp)
+gp2 <- lapply(gp2,unique)
+gt1 <- stack(gp2)
+colnames(gt1) <- c("gene","probe")
+gt1$probe <- as.character(gt1$probe)
+dim(gt1)
+str(gt1)
+length(unique(gt1$gene))
+#new.hgnc.table <- getCurrentHumanMap()
+new.hgnc.table <- readRDS("~/gmea/figs/new.hgnc.table.rds")
+fix <- checkGeneSymbols(gt1$gene,map=new.hgnc.table)
+fix2 <- fix[which(fix$x != fix$Suggested.Symbol),]
+length(unique(fix2$x))
+gt1$gene <- fix$Suggested.Symbol
+head(gt1)
+saveRDS(gt1,file="~/gmea/app/hm450k.rds")
+
+## EPIC
+anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
+myann <- data.frame(anno[,c("UCSC_RefGene_Name","UCSC_RefGene_Group","Islands_Name","Relation_to_Island")])
+gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
+gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
+names(gp2) <- rownames(gp)
+gp2 <- lapply(gp2,unique)
+gt2 <- stack(gp2)
+colnames(gt2) <- c("gene","probe")
+gt2$probe <- as.character(gt2$probe)
+dim(gt2)
+str(gt2)
+length(unique(gt2$gene))
+fix <- checkGeneSymbols(gt2$gene,map=new.hgnc.table)
+fix2 <- fix[which(fix$x != fix$Suggested.Symbol),]
+length(unique(fix2$x))
[…]

### 2024-01-22 app/make_genetable.R
genetables

@@ -16,13 +16,13 @@ dim(gt1)
 str(gt1)
 length(unique(gt1$gene))
 #new.hgnc.table <- getCurrentHumanMap()
-new.hgnc.table <- readRDS("~/gmea/figs/new.hgnc.table.rds")
+new.hgnc.table <- readRDS("../figs/new.hgnc.table.rds")
 fix <- checkGeneSymbols(gt1$gene,map=new.hgnc.table)
 fix2 <- fix[which(fix$x != fix$Suggested.Symbol),]
 length(unique(fix2$x))
 gt1$gene <- fix$Suggested.Symbol
 head(gt1)
-saveRDS(gt1,file="~/gmea/app/hm450k.rds")
+saveRDS(gt1,file="hm450k.rds")
 
 ## EPIC
 anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
@@ -42,4 +42,4 @@ fix2 <- fix[which(fix$x != fix$Suggested.Symbol),]
 length(unique(fix2$x))
 gt2$gene <- fix$Suggested.Symbol
 head(gt2)
-saveRDS(gt2,file="~/gmea/app/epic.rds")
+saveRDS(gt2,file="epic.rds")

### 2024-01-22 app/app.R
array type epic and 450k

@@ -87,17 +87,26 @@ server <- function(input, output, session) {
     gt2
   })
   
+ genetable <- reactive({
+   if (myarraytype()=="EPIC") {
+     gt <- readRDS("~/gmea/app/epic.rds")
+   } else {
+     gt <- readRDS("~/gmea/app/hm450k.rds")
+   }
+   gt
+ })
+    
   genesets <- reactive({
     if((mygenesettype())  == "Reactome") {
-      gs <- gmt_import("~/gmea/app/trial/c2.cp.reactome.v2023.2.Hs.symbols.gmt")
+      gs <- gmt_import("~/gmea/app/c2.cp.reactome.v2023.2.Hs.symbols.gmt")
     }
     
     if((mygenesettype())  == "KEGG") {
-      gs <- gmt_import("~/gmea/app/trial/c2.cp.kegg_medicus.v2023.2.Hs.symbols.gmt")
+      gs <- gmt_import("~/gmea/app/c2.cp.kegg_medicus.v2023.2.Hs.symbols.gmt")
     }
     
     if((mygenesettype())  == "GO") {
-      gs <- gmt_import("~/gmea/app/trial/c5.all.v2023.2.Hs.symbols.gmt")
+      gs <- gmt_import("~/gmea/app/c5.all.v2023.2.Hs.symbols.gmt")
     }
     gs
   })
@@ -116,7 +125,7 @@ server <- function(input, output, session) {
     m <- cbind(m[,1],m[,"t"])
     rownames(m) <- m[,1]
     m[,1]=NULL
-    m2 <- mitch_import(x=m,DEtype="prescored",geneTable=gt2())
+    m2 <- mitch_import(x=m,DEtype="prescored",geneTable=genetable())
     mres <- mitch_calc(x=m2,genesets=genesets(),minsetsize=5,cores=1, priority=prioritisation())    
   })
   

### 2024-01-22 app/app.R
shiny port

@@ -4,6 +4,9 @@ library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
 library("mitch")
 
 options(shiny.maxRequestSize = 300 * 1024^2) #300MB
+# Specify the application port
+options(shiny.host = "0.0.0.0")
+options(shiny.port = 8180)
 
 arraytype <- c("EPIC","450k")
 

### 2024-01-22 app/app.R
port fix

@@ -6,7 +6,7 @@ library("mitch")
 options(shiny.maxRequestSize = 300 * 1024^2) #300MB
 # Specify the application port
 options(shiny.host = "0.0.0.0")
-options(shiny.port = 8180)
+options(shiny.port = 3838)
 
 arraytype <- c("EPIC","450k")
 

### 2024-01-22 app/app.R
paths

@@ -92,24 +92,24 @@ server <- function(input, output, session) {
   
  genetable <- reactive({
    if (myarraytype()=="EPIC") {
-     gt <- readRDS("~/gmea/app/epic.rds")
+     gt <- readRDS("epic.rds")
    } else {
-     gt <- readRDS("~/gmea/app/hm450k.rds")
+     gt <- readRDS("hm450k.rds")
    }
    gt
  })
     
   genesets <- reactive({
     if((mygenesettype())  == "Reactome") {
-      gs <- gmt_import("~/gmea/app/c2.cp.reactome.v2023.2.Hs.symbols.gmt")
+      gs <- gmt_import("c2.cp.reactome.v2023.2.Hs.symbols.gmt")
     }
     
     if((mygenesettype())  == "KEGG") {
-      gs <- gmt_import("~/gmea/app/c2.cp.kegg_medicus.v2023.2.Hs.symbols.gmt")
+      gs <- gmt_import("c2.cp.kegg_medicus.v2023.2.Hs.symbols.gmt")
     }
     
     if((mygenesettype())  == "GO") {
-      gs <- gmt_import("~/gmea/app/c5.all.v2023.2.Hs.symbols.gmt")
+      gs <- gmt_import("c5.all.v2023.2.Hs.symbols.gmt")
     }
     gs
   })

### 2024-01-22 app/app.R
fix kegg

@@ -104,7 +104,7 @@ server <- function(input, output, session) {
       gs <- gmt_import("c2.cp.reactome.v2023.2.Hs.symbols.gmt")
     }
     
-    if((mygenesettype())  == "KEGG") {
+    if((mygenesettype())  == "KEGG MEDICUS") {
       gs <- gmt_import("c2.cp.kegg_medicus.v2023.2.Hs.symbols.gmt")
     }
     

### 2024-01-22 app/app.R
path problems

@@ -92,24 +92,24 @@ server <- function(input, output, session) {
   
  genetable <- reactive({
    if (myarraytype()=="EPIC") {
-     gt <- readRDS("epic.rds")
+     gt <- readRDS("/home/app/epic.rds")
    } else {
-     gt <- readRDS("hm450k.rds")
+     gt <- readRDS("/home/app/hm450k.rds")
    }
    gt
  })
     
   genesets <- reactive({
     if((mygenesettype())  == "Reactome") {
-      gs <- gmt_import("c2.cp.reactome.v2023.2.Hs.symbols.gmt")
+      gs <- gmt_import("/home/app/c2.cp.reactome.v2023.2.Hs.symbols.gmt")
     }
     
     if((mygenesettype())  == "KEGG MEDICUS") {
-      gs <- gmt_import("c2.cp.kegg_medicus.v2023.2.Hs.symbols.gmt")
+      gs <- gmt_import("/home/app/c2.cp.kegg_medicus.v2023.2.Hs.symbols.gmt")
     }
     
     if((mygenesettype())  == "GO") {
-      gs <- gmt_import("c5.all.v2023.2.Hs.symbols.gmt")
+      gs <- gmt_import("/home/app/c5.all.v2023.2.Hs.symbols.gmt")
     }
     gs
   })
@@ -161,7 +161,6 @@ server <- function(input, output, session) {
     content = function(file) {
       req(mtable())
       write.table(mtable(),file,sep="\t",row.names = FALSE)
-      #write.csv(mtcars, file)
     }
   )
 

### 2024-01-23 app/app.R
don't need gt2 any more

@@ -76,20 +76,6 @@ server <- function(input, output, session) {
     )
   })
   
-  gt2 <- reactive({
-    req(input$upload)
-    anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
-    myann <- data.frame(anno[,c("UCSC_RefGene_Name","UCSC_RefGene_Group","Islands_Name","Relation_to_Island")])
-    gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
-    gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
-    names(gp2) <- rownames(gp)
-    gp2 <- lapply(gp2,unique)
-    gt2 <- stack(gp2)
-    colnames(gt2) <- c("gene","probe")
-    gt2$probe <- as.character(gt2$probe)
-    gt2
-  })
-  
  genetable <- reactive({
    if (myarraytype()=="EPIC") {
      gt <- readRDS("/home/app/epic.rds")

### 2024-01-24 manuscript/manuscript.Rmd
update

@@
 Anusuiya Bora<sup>1,2</sup>,
-Severine Lamon<sup>3</sup>,
+Séverine Lamon<sup>3</sup>,
 Mary Sravya Kasu<sup>1</sup>,
@@
 Yen Ting Wong<sup>4</sup>,
-Jeffrey Craig<sup>4,5</sup>
+Jeffrey M. Craig<sup>4,5</sup>
 </strong>
@@
 
-Infinium methylation beadarrays remain one of the most popular platforms for studying genomic
-DNA methylation, but tools for downstream pathways analysis have their limitations.
-Functional class scoring (FCS) is a group of pathway enrichment techniques that involve the ranking of
-genes and evaluation of their collective regulation in biological systems, but the implementations described for
+Infinium methylation bead arrays remain one of the most popular platforms for studying genomic DNA methylation,
+but tools for downstream pathways analysis have their limitations.
+Functional class scoring (FCS) is a group of pathway enrichment techniques that involve the ranking of genes
+and evaluation of their collective regulation in biological systems, but the implementations described for
 Infinium methylation data do not retain direction information which is important for mechanistic understanding
@@
 t-statistics by gene followed by a rank-ANOVA enrichment test using the mitch package.
-This method, which we call "LAM", outperformed an existing over-representation analysis method in
-simulations, and showed higher sensitivity and robustness in an analysis of real lung tumor-normal paired datasets.
-Using matched RNA-seq data we examine the relationship of methylation differences at promoters and
-gene bodies with RNA expression at the level of pathways in lung cancer.
-To further demonstrate the utility of our approach, we apply it to two other contexts; firstly an examination of
-methylation differences associated with chronological age; and secondly to examine methylation differences in
-infants conceived with in vitro fertilisation.
-These results indicate that the LAM method is more sensitive and precise as compared to existing methods.
-LAM has the potential to illuminate disease mechanisms and identify novel diagnostic signatures.
-A vignette is provided to illustrate how to implement this method.
+This method, which we call “LAM”, outperformed an existing over-representation analysis method in simulations,
+and showed higher sensitivity and robustness in an analysis of real lung tumour-normal paired datasets.
+Using matched RNA-seq data we examine the relationship of methylation differences at promoters and gene bodies
+with RNA expression at the level of pathways in lung cancer.
+To demonstrate the utility of our approach, we apply it to three other contexts where public data were available.
+Firstly we examine differential pathway methylation associated with chronological age.
+Secondly we investigate pathway methylation differences in infants conceived with in vitro fertilisation.
+Lastly, we analyse differential pathway methylation in 19 disease states, identifying hundreds of novel associations.
+These results show LAM is a powerful method for the detection of differential pathway methylation as compared to
+existing methods.
+A reproducible vignette is provided to illustrate how to implement this method.
 
@@
 development and disease [@Robertson2005-tu].
-Many hundreds of epigenome-wide association studies (EWASs) are conducted each year to understand DNA
+Hundreds of epigenome-wide association studies (EWASs) are conducted each year to understand DNA
 methylation patterns in disease and other contexts [@Li2019-og].
@@
 as compared to high throughput methylation sequencing [@Wei2021-qj].
-Infinium arrays typically have multiple probes per gene in different locations including promoters,
+Infinium arrays typically include multiple probes per gene in different locations including promoters,
 CpG islands, gene bodies and enhancers [@Pidsley2016-ix], which complicates downstream functional enrichment analysis,
@@
 In ORA, genes with probes that meet an arbitrary significance threshold are selected and compared
-to a background list consisting of genes measured in the assay.
+to a background list of all genes measured in the assay.
 The test seeks to identify sets of genes (eg: ontologies, pathways) that are over-represented in the
@@
 FCS takes a different approach by ranking all detected genes by a differential
-regulation score (eg: fold change, confident effect size) followed by a test to assess whether each
+regulation score (eg: fold change, confident effect size, t-statistic) followed by a test to assess whether each
 set of genes has a distribution of scores that is different to the null [@Subramanian2005-ar].
 The `gsameth()` function of the `missMethyl` package is the state-of-the-art method for ORA of infinium methylation array
-data as it addresses issues related to probes belonging to more than one gene and the fact that genes can have multiple
+data as it addresses issues related to probes belonging to more than one gene and the fact that one gene can have multiple
 probes [@Maksimovic2021-ur].
-As gsameth is an ORA method, results dependent strongly on the significance threshold used [@Ren2019-mr], and also
+As gsameth is an ORA method, results strongly depend on the significance threshold used [@Ren2019-mr], and also
 on the proportion of probes that meet this threshold.
@@
 The `methylGSA` package implements two different approaches.
-Firstly, the `methylRRA` function which uses robust rank aggregation for detecting differentially methylated genes taking
+Firstly, the `methylRRA` function, which uses robust rank aggregation for detecting differentially methylated genes taking
 into consideration the variable number of probes followed by an FCS test based on z-scores [@Ren2019-mr].
@@
 limitations.
-Due to the fact that genes have a variable number of probes, there are many concevable ways that FCS could be applied
+Since genes have a variable number of probes, there are many concevable ways that FCS could be applied
 to Infinium methylation array data.
-The first set of methods use differential probe methylation data from `limma` [@Ritchie2015-zn] as an input.
+The first set of methods (type I) use differential probe methylation data from `limma` [@Ritchie2015-zn] as an input.
 The differential methylation of probes beloning to each gene are scored and these gene scores
 undergo an enrichment test.
-The second set of methods first aggregate probe methylation values for each gene before differential
+The second set of methods (type II) first aggregate probe methylation values for each gene before differential
 methylation analysis with limma.
 Then, differential gene methylation results are used for a downstream enrichment test.
-With the above methods there are several methodological choices, such as the type of aggregation (eg: mean, median), and the type of
-downstream test (eg: parametric or nonparametric).
-To determine the best approach to applying FCS to Infinium methylation data, in this study we use
+With the above methods there are several methodological choices, such as the type of aggregation, and the type of
+downstream test.
+To determine the best approach to applying FCS to Infinium methylation data, in this study, we use
 simulated data and compare results to existing ORA methods.
-We then examine the sensitivity of selected methods on real cancer data, investigate the association with gene expression and end
-with additional examples related to human aging and assisted reproductive technologies.
+We then examine the sensitivity of selected methods on real cancer data, investigate the association with gene expression
+and end with additional examples related to human ageing, assisted reproductive technologies and a large-scale EWAS of
+19 diseases.
 
@@
 meaning the data needs to be summarised from probes to genes and then to gene sets.
-Here, we outline eight potential approaches for FCS of Infinium methylation data:
+Here, we outline eight potential approaches for FCS of Infinium methylation data, where methods 1-5 are type I
+and 6-8 are type II:
 
@@
 Differential methylation analysis is conducted at the probe level with limma, and the limma
-t-statistics for each gene are summarised (mean).
+t-statistics for each gene are summarised (arithmetic mean).
 The mean t-statistics are used in a downstream two-sample two-way t-test of gene set enrichment.
@@
 Similar to LAW, however instead of the Wilcoxon test, the mitch package[@Kaspi2020-yx] is used to execute a two-way,
-two-sample ANOVA-on-ranks test (Kruskal-Wallis).
+two-sample ANOVA-on-ranks test as described previously[@Cox2012-zu].
 
@@
 modified hypergeometric test that accounts for multiple probe biases.
-In the case that fewer than 250 significant probes are identified, then 250 probes with the smallest p-values
+In the case that fewer than 250 significant probes were identified, the 250 probes with the smallest p-values
 were selected.
@@
 To assess these methods, we adopted a simulated data approach based on the selected modification of real methylation data.
-We downloaded raw intensity EPIC IDAT files from NCBI Gene Expression Omnibus (GEO) for study GSE158422 which consists of
+We downloaded raw intensity EPIC IDAT files from NCBI Gene Expression Omnibus (GEO) for study GSE158422, which consists of
 lung tumour and normal adjacent tissues from 37 patients [@Cho2022-yd].
@@
 Following the incorporation of methylation changes to selected probes in the case group, the data underwent
-limma differential methylation analysis, followed by enrichment analysis using `GSAmeth` approach, or one of
+limma differential methylation analysis, followed by enrichment analysis using the `GSAmeth` approach, or one of
 the eight FCS methods outlined above.
@@
 the number of true positives (TP), false positives (FP) and false negatives (FN) could be determined.
-At each setting for group size, delta and gene set size, 100 replications were conducted each with a set seed
+At each setting for group size, delta and gene set size, 100 replications were conducted, each with a set seed
 to ensure reproducibility.
@@
 As gene annotations are several years old, we used HGNChelper to update them (database current as of 21-Dec-2023) [@Oh2022-ts]
-(3253 gene symbols were updated).
+(3,253 gene symbols were updated).
 Euler diagrams were created with the `eulerr` R package.
 To test sensitivity, a random subset of n patients was selected followed by pathway enrichment analysis with the
-respective method, with n varied between 2 and 30.
+respective method, with n varying between 2 and 30.
 Significantly enriched pathways were defined as those with FDR values < 0.05.
@@
 methods, and whether findings at a smaller sample size were consistent with the full group.
-This process was repeated 50 times at each sample size.
+This process was repeated 50 times for each sample size.
 
@@
 GSE131433[@Novakovic2019-jt].
-Various assisted reproductive technology conception groups are described in these studies, but we examined
-just fresh in vitro fertilisation (IVF) conceived infants compared to naturally conceived infants.
+Various assisted reproductive technology conception groups are described in these studies, but we only examined
+fresh in vitro fertilisation (IVF) conceived infants compared to naturally conceived infants.
 Raw intensity files (IDAT format) were read into R with minfi [@Aryee2014-su], normalistion was conducted with the SWAN method
@@
 with LAM were collectively smaller than GSA (**Figure 3B**).
-This may seem counterintuitive, but may be due to LAM's better sensitivity in dectecting pathway enrichment when effect sizes are
+This may seem counterintuitive but may be due to LAM's better sensitivity in dectecting pathway enrichment when effect sizes are
 lower.
@@
 DNA methylation.
-From 18,704 detected genes, There were 12,380 differentially expressed genes (FDR<0.05), with 7,415 and 4,965 up- and down-regulated
+From 18,704 detected genes, there were 12,380 differentially expressed genes (FDR<0.05), with 7,415 and 4,965 up- and down-regulated
 in the tumour group respectively.
@@
 Some of the observed methylation changes make sense with reports in the literature.
-For example the complement pathway is associated with aging and in this analysis we observe a reduction in methylation of genes
+For example, the complement pathway is associated with aging and in this analysis we observe a reduction in methylation of genes
 beloning to this set including *C1R*, *MBL2*, *C1S* and *C1QC* (**Figure 6D**).
@@
 studies since 2009 [@Katari2009-ya].
-Two of the highest powered studies with publicly available data are Estill 2016 which used the 450K array and Novakovic 2019 study
+Two of the highest powered studies with publicly available data are Estill 2016 which used the HM450K array and Novakovic 2019 study
 which used the EPIC array [@Estill2016-jh;@Novakovic2019-jt].
@@
 embryo transfer.
-The Estill dataset on the 450k platform had 418,833 probes that met the quality criteria (Natural n=43; IVF n=38).
+The Estill dataset on the HM450K platform had 418,833 probes that met the quality criteria (Natural n=43; IVF n=38).
 Of these, 149,589 probes showed statistical significant differential methylation (FDR<0.05), of which 28,767 and 120,822 exhibited
@@
 respectively.
-After summarisation to the gene level, the Estill dataset on 450k array describes 19,240 genes, while the Novakovic dataset on EPIC
+After summarisation to the gene level, the Estill dataset on HM450K array describes 19,240 genes, while the Novakovic dataset on EPIC
 array describes 22,588 genes.
@@
 A scatterplot of pathway enrichment scores indicates a moderate degree of agreement between these studies
-(*r*=0.35,*p*=2.2e-16) despite these studies being conducted on independent cohorts years apart and analysed with different array
+(*r*=0.35, *p*=2.2e-16) despite these studies being conducted on independent cohorts years apart and analysed with different array
 systems (**Figure 7B**).
@@
 Taken together, this analysis shows that LAM method can be used to compare enrichment of independent studies
-conducted with 450k and EPIC arrays.
+conducted with HM450K and EPIC arrays.
 
 ![**Figure 7. Pathway level DNA methylation differences in natural and IVF conceived infants.**
-(A) Contour heatmap showing the similarity in gene methylation score ranks in the Estill (450k) and Novakovic (EPIC) studies.
+(A) Contour heatmap showing the similarity in gene methylation score ranks in the Estill (HM450K) and Novakovic (EPIC) studies.
 (B) Mitch pathway enrichment scores (S distances) in Estil and Novakovic studies.
@@
 
-To examine the potential for LAM to illuminate differential pathway methylation associations
+To examine the potential for LAM to reveal differential pathway methylation associations
 with diseases, we obtained and analysed blood based EWAS summary statistics from a recent
@@
 (A) A barplot showing the number of statistically significant pathways with higher and lower methylation identified in each prevalent condition (FDR<0.05).
-(B) A heatmap of s scores for selected pathways across 14 common prevalent disease states.
+(B) A heatmap of S scores for selected pathways across 14 common prevalent disease states.
 Stars indicate that the pathway was identified as being among the top five differentially methylated pathways in each direction for each condition.
@@
 with a condition.
-So we examined differential pathway methylation associated with incidence of 19 disease states.
+To this end, we examined differential pathway methylation associated with incidence of 19 disease states.
 Over all 19 conditions, there were 1570 significant pathways (FDR<0.05) (**Figure 9A**).
@@
 (A) A barplot showing the number of statistically significant pathways with higher and lower methylation identified in each incident condition (FDR<0.05).
-(B) A heatmap of s scores for selected pathways across 19 common incident disease states.
+(B) A heatmap of S scores for selected pathways across 19 common incident disease states.
 Stars indicate that the pathway was identified as being among the top three differentially methylated pathways in each direction for each condition.
@@
 Specifically, from 100 runs 64 yielded no significant pathways across 19 conditions.
-Twenty eight repeats had fewer than five false positives and eight runs had more than five false positives.
+Twenty-eight repeats had fewer than five false positives and eight runs had more than five false positives.
 Across the 100 repeats, the mean number of false positives was 1.95 and median was 0.
-These results indicate that LAM can identify differentially methylated pathways in disease groups, and
+These results indicate that LAM can identify differentially methylated pathways in disease groups and
 detect altered methylation before a diagnosis has been made, with few false positives.
@@
 The improved sensitivity of the LAM method provides a new opportunity for researchers to reanalyse previously conducted EWASs with contemporary pathway databases to better understand
-the subtle signatures, as we have demonstrated with the chronological age, assisted reproduction examples and 19 disease EWAS analyses.
+subtle signatures, as we have demonstrated with the chronological age, assisted reproduction examples and 19 disease EWAS analyses.
 The analyses conducted here used Reactome pathways, but LAM is general purpose, so different gene sets can be used to examine various hypotheses.
@@
 Joint enrichment analysis showed some genes with increasing promoter methylation together with increasing RNA expression (eg: "Negative regulation of AP2 transcription factors").
-Likewise a few pathways showed decreases in methylation with decreases in RNA expression such as "Dectin family", "TLR regulation" and "glycan biosynthesis termination".
+Likewise, a few pathways showed decreases in methylation with decreases in RNA expression such as "Dectin family", "TLR regulation" and "glycan biosynthesis termination".
 These results indicate that there is a complicated relationship between DNA methylation and gene expression in lung cancer and further work is required to appreciate the mechanisms involved and distinguish which CpGs are critical for influencing gene expression and which are not.
@@
 
-Updating defunct gene symbols is an important step as 2,788 and 3,253 gene symbols were updated on the HM450K and EPIC array annotations, which is approximately 10-14% of genes on these platforms.
+Updating defunct gene symbols is an important step as 2,788 and 3,253 gene symbols were updated on the HM450K and EPIC array annotations, which represents approximately 10-14% of genes on these platforms.
 We note that other packages for enrichment of Infinium array data do not by default address this, which could lead to a loss of these genes from pathway enrichment and diminished sensitivity.
 
-The analysis of the large scale disease EWAS shows that LAM can be readily applied to identify differential pathway methylation associated with prevalent and incident disease.
+The analysis of the large-scale disease EWAS shows that LAM can be readily applied to identify differential pathway methylation associated with prevalent and incident disease.
 The authors of the study used GOmeth for their pathway analysis and from the 33 models examined, significant findings were obtained for only two, prevalent diabetes and heart disease.
@@
 Distinguishing prevalent and incident signatures is important as it is known that methylation changes can be caused by disease states.
-For example metabolic changes in diabetes are probably the cause of the observed methylation signatures.
-It is perhaps unexpected that more significant pathways be identified from the incident disease arm of the study,
+For example, metabolic changes in diabetes are probably the cause of the observed methylation signatures.
+Epigenetic differences may also contribute to disease risk.
 
-This work could be extended to include not just proximal CpG sites, but also enhancer based probes and their target genes based on resources such as GeneHancer.
+This work could be extended to include not just proximal CpG sites, but also enhancer-based probes and their target genes based on resources such as GeneHancer.
 

### 2024-01-25 manuscript/manuscript.Rmd
heading levels

@@
 
-## Abstract
+# Abstract
 
@@
 
-## Introduction
+# Introduction
 
@@
 
-## Methods
+# Methods
 
-### Implementation overview
+## Implementation overview
 
@@
 
-### Method validation using simulations
+## Method validation using simulations
 
@@
 
-### Sensitivity analysis
+## Sensitivity analysis
 
@@
 
-### Association of methylation with gene expression
+## Association of methylation with gene expression
 
@@
 
-### Pathway-level differential methylation in aging
+## Pathway-level differential methylation in aging
 
@@
 
-### Pathway-level differential methylation in assisted reproductive technology
+## Pathway-level differential methylation in assisted reproductive technology
 
@@
 
-### Differential pathway methylation in 19 common disease states
+## Differential pathway methylation in 19 common disease states
 
@@
 
-## Results
+# Results
 
-### Evaluation of FCS methods for Infinium methylation array data
+## Evaluation of FCS methods for Infinium methylation array data
 
@@
 
-### Sensitivity of GSA and LAM methods with real cancer data
+## Sensitivity of GSA and LAM methods with real cancer data
 
@@
 
-## Discussion
+# Discussion
 
@@
 
-## Bibliography
+# Bibliography

### 2024-02-12 manuscript/manuscript.Rmd
updates

@@
 
-Infinium methylation bead arrays remain one of the most popular platforms for studying genomic DNA methylation,
-but tools for downstream pathways analysis have their limitations.
-Functional class scoring (FCS) is a group of pathway enrichment techniques that involve the ranking of genes
-and evaluation of their collective regulation in biological systems, but the implementations described for
-Infinium methylation data do not retain direction information which is important for mechanistic understanding
-of genomic regulation.
+Infinium Methylation BeadChip arrays remain one of the most popular platforms for epigenome-wide association studies,
+but tools for downstream pathway analysis have their limitations.
+Functional class scoring (FCS) is a group of pathway enrichment techniques that involve the ranking of genes and
+evaluation of their collective regulation in biological systems, but the implementations described for Infinium
+methylation array data do not retain direction information, which is important for mechanistic understanding of
+genomic regulation.
 Here, we evaluate several candidate FCS methods that retain directional information.
-According to simulation results, the best performing method involves the mean aggregation of probe limma
-t-statistics by gene followed by a rank-ANOVA enrichment test using the mitch package.
-This method, which we call “LAM”, outperformed an existing over-representation analysis method in simulations,
-and showed higher sensitivity and robustness in an analysis of real lung tumour-normal paired datasets.
-Using matched RNA-seq data we examine the relationship of methylation differences at promoters and gene bodies
-with RNA expression at the level of pathways in lung cancer.
+According to simulation results, the best performing method involves the mean aggregation of probe limma t-statistics
+by gene followed by a rank-ANOVA enrichment test using the mitch package.
+This method, which we call “LAM”, outperformed an existing over-representation analysis method in simulations, and
+showed higher sensitivity and robustness in an analysis of real lung tumour-normal paired datasets.
+Using matched RNA-seq data we examine the relationship of methylation differences at promoters and gene bodies with
+RNA expression at the level of pathways in lung cancer.
 To demonstrate the utility of our approach, we apply it to three other contexts where public data were available.
-Firstly we examine differential pathway methylation associated with chronological age.
-Secondly we investigate pathway methylation differences in infants conceived with in vitro fertilisation.
+Firstly, we examine differential pathway methylation associated with chronological age.
+Secondly, we investigate pathway methylation differences in infants conceived with in vitro fertilisation.
 Lastly, we analyse differential pathway methylation in 19 disease states, identifying hundreds of novel associations.
 These results show LAM is a powerful method for the detection of differential pathway methylation as compared to
-existing methods.
-A reproducible vignette is provided to illustrate how to implement this method.
+existing methods. A reproducible vignette is provided to illustrate how to implement this method.
 
-**Keywords:** Pathway analysis, functional enrichment analysis, Infinium Array, DNA methylation,
-epigenetics.
+**Keywords:** Pathway analysis, functional enrichment analysis, Infinium Array, DNA methylation, epigenetics, epigenome-wide association study.
 
@@
 otherwise known as pathway analysis or gene set enrichment analysis.
-Enrichment analysis comes in two popular types; over-representation analysis (ORA) and functional
+Enrichment analysis comes in two popular types: over-representation analysis (ORA) and functional
 class scoring (FCS) [@Khatri2012-qy].
@@
 to a background list of all genes measured in the assay.
-The test seeks to identify sets of genes (eg: ontologies, pathways) that are over-represented in the
+The test seeks to identify sets of genes (e.g.: ontologies, pathways) that are over-represented in the
 gene list of interest relative to the background [@Tavazoie1999-ng].
 FCS takes a different approach by ranking all detected genes by a differential
-regulation score (eg: fold change, confident effect size, t-statistic) followed by a test to assess whether each
+regulation score (e.g.: fold change, confident effect size, t-statistic) followed by a test to assess whether each
 set of genes has a distribution of scores that is different to the null [@Subramanian2005-ar].
@@
 Following the incorporation of methylation changes to selected probes in the case group, the data underwent
-limma differential methylation analysis, followed by enrichment analysis using the `GSAmeth` approach, or one of
+limma differential methylation analysis, followed by enrichment analysis using `GSAmeth`, or one of
 the eight FCS methods outlined above.
@@
 RNA-seq based gene counts for the same set of lung cancer patients were downloaded from GEO (accession: GSE158420).
-HGNChelper was used to repair transmogrified gene names, followed by DESeq2 based differential analysis.
+HGNChelper was used to fix gene names converted to dates, followed by DESeq2 based differential analysis.
 This underwent filtering to remove genes with expression below 10 reads per sample on average across the dataset.
@@
 In order to deminstrate the utility of the LAM method for large EWAS studies, we examined
-methylation profile associations with age of two independent cohorts involving a total of 7036
+EPIC methylation profile associations with age of two independent cohorts involving a total of 7036
 participants [@McCartney2020-ob].
@@
 From the 839,473 probes passing quality control, 473,572 were considered significant at 5% FDR.
-Of these, 119,785 probes were significantly hypermethylated and 353,787 were significantly hypomethylated.
-These significantly hypermethylated probes mapped to 16,520 unique genes, while the hypomethylated probes mapped to 24,114 genes.
+Of these, 119,785 probes were significantly increased in cancer, while 353,787 were significantly decreased.
+These significantly increased probes mapped to 16,520 unique genes, while the decreased probes mapped to 24,114 genes.
 These numbers are large, considering that there are 26,219 genes represented on this array.
@@
 with LAM were collectively smaller than GSA (**Figure 3B**).
-This may seem counterintuitive but may be due to LAM's better sensitivity in dectecting pathway enrichment when effect sizes are
-lower.
 
@@
 "olfactory signaling".
-These were however less common than pathways that showed a positive relationship between RNA expression and promoter methylation.
+These were, however, less common than pathways that showed a positive relationship between RNA expression and promoter methylation.
 These results demonstrate that LAM can illuminate the complicated relationship between DNA methylation and gene expression at the
@@
 the biological differences between conception groups.
-For this example we focused on comparing naturally conceived infants to those conceived with in vitro fertilisation (IVF) with fresh
+For this example, we focused on comparing naturally conceived infants to those conceived with in vitro fertilisation (IVF) with fresh
 embryo transfer.
 The Estill dataset on the HM450K platform had 418,833 probes that met the quality criteria (Natural n=43; IVF n=38).
-Of these, 149,589 probes showed statistical significant differential methylation (FDR<0.05), of which 28,767 and 120,822 exhibited
+Of these, 149,589 probes showed statistically significant differential methylation (FDR<0.05), of which 28,767 and 120,822 exhibited
 higher or lower methylation in the IVF group respectively.
@@
 with a condition.
-To this end, we examined differential pathway methylation associated with incidence of 19 disease states.
+Therefore, we examined differential pathway methylation associated with incidence of 19 disease states.
 Over all 19 conditions, there were 1570 significant pathways (FDR<0.05) (**Figure 9A**).
@@
 
-Overall, we found that the FCS methods performed better across all simulation conditions.
+Overall, we found that the FCS methods performed better across most simulation conditions.
 Of the different FCS methods tested, LAM was superior in terms of precision and recall.
@@
 
-Our analyses also consistently showed that non-parametric methods showed better recall with only a small decrease in precision, in accordance with some previous reports.
+Our analyses also consistently showed that non-parametric methods showed better recall with only a small decrease in precision.
 Summarisation of probe level t-statistics to gene level only takes a few seconds with the efficient base R `aggregate` command,
@@
 
-An interesting observation is that ORA performed better when gene sets are smaller.
+An interesting observation is that ORA performed better when gene sets are smaller in our simulations.
 Given that popular pathway databases like KEGG and Reactome pathways consist of both large (>100) and small (<20) gene sets,
@@
 
-The differences in pathway results obtained with LAM and GSA from lung cancer data were striking, not only in the numbers of pathways identified, but also their relevance to the disease.
-Interestingly the relationship between promoter methylation and gene expression was not always inverse as we had anticipated.
-Joint enrichment analysis showed some genes with increasing promoter methylation together with increasing RNA expression (eg: "Negative regulation of AP2 transcription factors").
-Likewise, a few pathways showed decreases in methylation with decreases in RNA expression such as "Dectin family", "TLR regulation" and "glycan biosynthesis termination".
-These results indicate that there is a complicated relationship between DNA methylation and gene expression in lung cancer and further work is required to appreciate the mechanisms involved and distinguish which CpGs are critical for influencing gene expression and which are not.
+The differences in pathway results obtained with LAM and GSA from lung cancer data were striking, not only in the numbers of
+pathways identified, but also their relevance to the disease.
+Interestingly, the relationship between promoter methylation and gene expression was not always inverse as we had anticipated.
+Joint enrichment analysis showed some genes with increasing promoter methylation together with increasing RNA expression
+(eg: “Negative regulation of AP2 transcription factors”).
+Likewise, a few pathways showed decreases in methylation with decreases in RNA expression such as “Dectin family”, “TLR regulation”
+and “glycan biosynthesis termination”.
+These results are in line with previous reports of the widespread existence of both negative and positive relationships between DNA methylation and
+gene expression[@Wagner2014-gy; @Xu2019-nu; @Spainhour2019-ur; @Wang2022-mr],
+and support the idea that there is a complicated relationship between DNA methylation and gene expression in lung cancer.
 
@@
 The authors of the study used GOmeth for their pathway analysis and from the 33 models examined, significant findings were obtained for only two, prevalent diabetes and heart disease.
-On contrast, LAM identified 2,469 pathway associations (74.8 per model).
+In contrast, LAM identified 2,469 pathway associations (74.8 per model).
 To demonstrate these are not the results of false positives, permutated profiles yielded on average 0.10 significant findings per model,
@@
 
-Distinguishing prevalent and incident signatures is important as it is known that methylation changes can be caused by disease states.
-For example, metabolic changes in diabetes are probably the cause of the observed methylation signatures.
-Epigenetic differences may also contribute to disease risk.
-
-This work could be extended to include not just proximal CpG sites, but also enhancer-based probes and their target genes based on resources such as GeneHancer.
+This work could be extended to include not just proximal CpG sites, but also enhancer-based probes and their target genes based on resources such as GeneHancer [@Fishilevich2017-xx].
 

### 2024-02-14 manuscript/manuscript.Rmd
ms

@@
 Mitch W. Norris<sup>1</sup>,
-Yen Ting Wong<sup>4</sup>,
+Yen Ting Wong<sup>4,5</sup>,
 Jeffrey M. Craig<sup>4,5</sup>
@@
 | Mary Sravya Kasu | 0000-0002-7891-836X |
+| Yen Ting Wong | 0000-0003-3120-096 |
+| Jeffrey M Craig | 0000-0003-3979-7849 |
 
@@
 to a background list of all genes measured in the assay.
-The test seeks to identify sets of genes (e.g.: ontologies, pathways) that are over-represented in the
+The test seeks to identify sets of genes (e.g., ontologies, pathways) that are over-represented in the
 gene list of interest relative to the background [@Tavazoie1999-ng].
 FCS takes a different approach by ranking all detected genes by a differential
-regulation score (e.g.: fold change, confident effect size, t-statistic) followed by a test to assess whether each
+regulation score (e.g., fold change, confident effect size, t-statistic) followed by a test to assess whether each
 set of genes has a distribution of scores that is different to the null [@Subramanian2005-ar].
@@
 their signal and reduce sensitivity [@Hong2014-ab].
-Neither GSAmeth nor methylGSA report computed enrichment scores, which hinders downstream interpretation, as the
+Neither gsameth nor methylGSA report computed enrichment scores, which hinders downstream interpretation, as the
 enrichment score is a useful surrogate for effect size.
@@
 The above were compared to a standard ORA-based approach (GSA), which involves limma on probes, selection of
-statistically significant probes (FDR<0.05) for ORA with the `GSAmeth()` function of missMethyl which conducts a
+statistically significant probes (FDR<0.05) for ORA with the `gsameth()` function of missMethyl which conducts a
 modified hypergeometric test that accounts for multiple probe biases.
@@
 Following the incorporation of methylation changes to selected probes in the case group, the data underwent
-limma differential methylation analysis, followed by enrichment analysis using `GSAmeth`, or one of
+limma differential methylation analysis, followed by enrichment analysis using `gsameth`, or one of
 the eight FCS methods outlined above.
@@
 correcting for patient-of-origin effects.
-Downstream pathway enrichment analysis was conducted using the GSAmeth and LAM methods
+Downstream pathway enrichment analysis was conducted using the GSA and LAM methods
 with Reactome pathways downloaded from MSigDB website (version 2023.1)[@Milacic2024-js].
@@
 CpG sites annotated as promoters were considered separately from those located at gene bodies in mitch analysis.
-For comparison, GSAmeth was used to separately examine pathways using separate analyses of promoter and gene body CpGs.
+For comparison, GSA was used to separately examine pathways using separate analyses of promoter and gene body CpGs.
 For this analysis, Reactome gene sets were used.
@@
 
-Simulations were conducted using the eight FCS methods (LAT, LTT, LAW, LAM, LRM, ALT, ALW, ALM) and GSAmeth,
+Simulations were conducted using the eight FCS methods (LAT, LTT, LAW, LAM, LRM, ALT, ALW, ALM) and GSA,
 with group sizes of 3-12 and delta values between 0.1 and 0.5, and random gene sets with sizes 20, 50
@@
 was on par with LAM (**Figure 1C**).
-At 100 genes per set however, LAM showed superior precision and recall compared to GSA (**Figure 1D**).
+At 100 genes per set, however, LAM showed superior precision and recall compared to GSA (**Figure 1D**).
 
@@
 (B) Sizes of significant gene sets found with LAM and GSA methods.
-(C) Significant LAM pathways with largest S distance, a proxy for effect size.
+(C) Significant LAM pathways with the largest S distance, an enrichment score generated by mitch and effect size proxy.
 (D) Top ranked significant GSA pathways ranked by fold enrichment.](img/fig3.svg)
@@
 in the tumour group respectively.
-At the pathway level there were 304 and 135 up- and down-regulated Reactome pathways with altered gene expression (FDR<0.05).
+At the pathway level, there were 304 and 135 up- and down-regulated Reactome pathways with altered gene expression (FDR<0.05).
 As the context of gene methylation is important in influencing gene expression, promoter and gene body methylation
@@
 There were a small number of pathways that exhibited an inverse relationship between promoter methylation and RNA expression such as
-"unwinding of DNA", "mucopolysaccharidoses","interaction with cumulus cells and the zona pellucida", "glucocorticoid biosynthesis" and
+"unwinding of DNA", "mucopolysaccharidoses", "interaction with cumulus cells and the zona pellucida", "glucocorticoid biosynthesis" and
 "olfactory signaling".
@@
 (C) Multi-dimensional enrichment analysis of promoter and gene body methylation with gene expression using mitch.
-Top 20 gene sets shown with largest s distance after FDR filtering at 0.05.](img/fig5.svg)
+Top 20 gene sets shown with the largest S distance after FDR filtering at 0.05.](img/fig5.svg)
 
@@
 sample (**Figure 6B**).
-When prioritising the results by S distance, a surrogate measure of effect size, 29 of the top 30 pathways were concordant in their
-direction of regulation (**Figure 6C**).
+When prioritising results by S distance, the enrichment score reported by mitch and a surrogate measure of effect size,
+29 of the top 30 pathways were concordant in their direction of regulation (**Figure 6C**).
 Some of the observed methylation changes make sense with reports in the literature.
@@
 Joint enrichment analysis showed some genes with increasing promoter methylation together with increasing RNA expression
-(eg: “Negative regulation of AP2 transcription factors”).
+(e.g., “Negative regulation of AP2 transcription factors”).
 Likewise, a few pathways showed decreases in methylation with decreases in RNA expression such as “Dectin family”, “TLR regulation”

### 2024-02-14 manuscript/manuscript.Rmd
ms

@@
 set of genes has a distribution of scores that is different to the null [@Subramanian2005-ar].
-The `gsameth()` function of the `missMethyl` package is the state-of-the-art method for ORA of infinium methylation array
+The `gsameth()` function of the `missMethyl` package is the state-of-the-art method for ORA of Infinium methylation array
 data as it addresses issues related to probes belonging to more than one gene and the fact that one gene can have multiple
@@
 limitations.
-Since genes have a variable number of probes, there are many concevable ways that FCS could be applied
+Since genes have a variable number of probes, there are many conceivable ways that FCS could be applied
 to Infinium methylation array data.
 The first set of methods (type I) use differential probe methylation data from `limma` [@Ritchie2015-zn] as an input.
-The differential methylation of probes beloning to each gene are scored and these gene scores
+The differential methylation of probes belonging to each gene are scored and these gene scores
 undergo an enrichment test.
@@
 4. Limma Average Mitch (LAM).
-Similar to LAW, however instead of the Wilcoxon test, the mitch package[@Kaspi2020-yx] is used to execute a two-way,
+Similar to LAW, however, instead of the Wilcoxon test, the mitch package[@Kaspi2020-yx] is used to execute a two-way,
 two-sample ANOVA-on-ranks test as described previously[@Cox2012-zu].
@@
 5. Limma Rank Mitch (LRM).
-Probes are ranked by t-statistic. The mean rank values for probes beloning to each gene are computed and used
+Probes are ranked by t-statistic. The mean rank values for probes belonging to each gene are computed and used
 as input for a two-way, two-sample ANOVA-on-ranks test.
@@
 6. Aggregate Limma t-test (ALT).
-In this method, all probe measurements beloning to each gene are averaged before conducting differential
-methylation andlysis with limma.
+In this method, all probe measurements belonging to each gene are averaged before conducting differential
+methylation analysis with limma.
 The limma t-statistics are then used downstream for enrichment analysis with the two-sample two-way t-test.
@@
 7. Aggregate Limma Wilcox (ALW).
-Similar to ALT, however it uses the Wilcoxon non-parametric test for gene set enrichment.
+Similar to ALT, however, it uses the Wilcoxon non-parametric test for gene set enrichment.
 
 8. Aggregate Limma Mitch (ALM).
-Similar to ALW, however uses the Mitch package to execute a two-way, two-sample ANOVA-on-ranks test.
+Similar to ALW, however, uses the Mitch package to execute a two-way, two-sample ANOVA-on-ranks test.
 
@@
 to ensure reproducibility.
-Simulations were conducted on a SLURM cluster [@Yoo2003-mc].
+Simulations were conducted on a SLURM high performance computing cluster [@Yoo2003-mc].
 Mean TP, FP and FN values were determined, and then precision, recall and F1 scores were calculated.
@@
 Subsequently, to generate the pathway heatmap, we used mitch in multivariate mode by analysing differential RNA expression,
-promoter and gene body methylation together, prioritising the results by s-distance, a surrogate measure of effect size.
+promoter and gene body methylation together, prioritising the results by S distance, a surrogate measure of effect size.
 
@@
 
-In order to deminstrate the utility of the LAM method for large EWAS studies, we examined
+In order to demonstrate the utility of the LAM method for large EWAS studies, we examined
 EPIC methylation profile associations with age of two independent cohorts involving a total of 7036
@@
 fresh in vitro fertilisation (IVF) conceived infants compared to naturally conceived infants.
-Raw intensity files (IDAT format) were read into R with minfi [@Aryee2014-su], normalistion was conducted with the SWAN method
+Raw intensity files (IDAT format) were read into R with minfi [@Aryee2014-su], normalisation was conducted with the SWAN method
 [@Maksimovic2012-hy], and probe filtering was conducted to remove proves with a detection p-value > 0.01 in addition to probes
@@
 Zenodo (https://zenodo.org/records/8021411).
-The study used a logitudinal design, with sample collection baseline and diagnosis status established via electronic health
+The study used a longitudinal design, with sample collection baseline and diagnosis status established via electronic health
 record linkage up to 14 years after baseline.
@@
 were considered separately.
-In promoters, GSA identified 15 and 7 pathways with higher and lower methylation respectively, while at gene bodies, there were were
+In promoters, GSA identified 15 and 7 pathways with higher and lower methylation respectively, while at gene bodies, there were
 18 and 75 pathways with higher and lower methylation respectively (FDR<0.05).
@@
 Some of the observed methylation changes make sense with reports in the literature.
-For example, the complement pathway is associated with aging and in this analysis we observe a reduction in methylation of genes
-beloning to this set including *C1R*, *MBL2*, *C1S* and *C1QC* (**Figure 6D**).
+For example, the complement pathway is associated with aging [@Gaya_da_Costa2018-sy]
+and in this analysis we observe a reduction in methylation of genes
+belonging to this set including *C1R*, *MBL2*, *C1S* and *C1QC* (**Figure 6D**).
 
@@
 
-At 100 genes per set, LAM had superior sensitivity compared to GSA, which agrees with a prevous report of mitch's performance with
+At 100 genes per set, LAM had superior sensitivity compared to GSA, which agrees with a previous report of mitch's performance with
 simulated RNA-seq gene expression data[@Kaspi2020-yx].
@@
 In contrast, LAM identified 2,469 pathway associations (74.8 per model).
-To demonstrate these are not the results of false positives, permutated profiles yielded on average 0.10 significant findings per model,
+To demonstrate these are not the results of false positives, permuted profiles yielded on average 0.10 significant findings per model,
 supporting the idea that LAM is identifying real pathway signatures.
@@
 
-There is also potential to generate new libraries of differentially methylated probe and gene sets, to contribute to the pool of molecular signatures in public resources like MSigDB,
+There is also potential to generate new libraries of differentially methylated probes and gene sets, to contribute to the pool of molecular signatures in public resources like MSigDB,
 which will assist in understanding similarities between methylation profiles.

### 2024-02-14 manuscript/manuscript.Rmd
grammar and spelling

@@
 Here, we evaluate several candidate FCS methods that retain directional information.
-According to simulation results, the best performing method involves the mean aggregation of probe limma t-statistics
+According to simulation results, the best-performing method involves the mean aggregation of probe limma t-statistics
 by gene followed by a rank-ANOVA enrichment test using the mitch package.
@@
 regulation score (e.g., fold change, confident effect size, t-statistic) followed by a test to assess whether each
-set of genes has a distribution of scores that is different to the null [@Subramanian2005-ar].
+set of genes has a distribution of scores that is different from the null [@Subramanian2005-ar].
 The `gsameth()` function of the `missMethyl` package is the state-of-the-art method for ORA of Infinium methylation array
@@
 These methods are indeed more sensitive than ORA approaches, but as they are based on one-tailed tests irrespective of
-direction of methylation change, it limits the utility for downstream interpretation of the importance of methylation
+the direction of methylation change, it limits the utility for downstream interpretation of the importance of methylation
 changes on genomic regulation.
@@
 to Infinium methylation array data.
-The first set of methods (type I) use differential probe methylation data from `limma` [@Ritchie2015-zn] as an input.
-The differential methylation of probes belonging to each gene are scored and these gene scores
+The first set of methods (type I) uses differential probe methylation data from `limma` [@Ritchie2015-zn] as an input.
+The differential methylation values of probes belonging to each gene are scored and these gene scores
 undergo an enrichment test.
@@
 Then, differential gene methylation results are used for a downstream enrichment test.
-With the above methods there are several methodological choices, such as the type of aggregation, and the type of
+With the above methods, there are several methodological choices, such as the type of aggregation, and the type of
 downstream test.
@@
 We then examine the sensitivity of selected methods on real cancer data, investigate the association with gene expression
-and end with additional examples related to human ageing, assisted reproductive technologies and a large-scale EWAS of
+and end with additional examples related to human aging, assisted reproductive technologies and a large-scale EWAS of
 19 diseases.
@@
 the eight FCS methods outlined above.
-Enrichment results were then filtered for FDR<0.05 and correct direction of methylation change.
+Enrichment results were then filtered for FDR<0.05 and the correct direction of methylation change.
 Differentially methylated gene sets observed were compared to gene sets selected to be the ground truth, so that
@@
 to ensure reproducibility.
-Simulations were conducted on a SLURM high performance computing cluster [@Yoo2003-mc].
+Simulations were conducted on a SLURM high-performance computing cluster [@Yoo2003-mc].
 Mean TP, FP and FN values were determined, and then precision, recall and F1 scores were calculated.
@@
 
-RNA-seq based gene counts for the same set of lung cancer patients were downloaded from GEO (accession: GSE158420).
-HGNChelper was used to fix gene names converted to dates, followed by DESeq2 based differential analysis.
+RNA-seq gene counts for the same set of lung cancer patients were downloaded from GEO (accession: GSE158420).
+HGNChelper was used to fix gene names converted to dates, followed by DESeq2 differential analysis.
 This underwent filtering to remove genes with expression below 10 reads per sample on average across the dataset.
@@
 In order to demonstrate the utility of the LAM method for large EWAS studies, we examined
-EPIC methylation profile associations with age of two independent cohorts involving a total of 7036
+EPIC methylation profile associations with age in two independent cohorts involving a total of 7,036
 participants [@McCartney2020-ob].
@@
 Briefly, gene-probe associations were first obtained from the "IlluminaHumanMethylationEPICanno.ilm10b4.hg19"
-dataset, followed by update of gene symbols with HGNChelper.
+dataset, followed by an update of gene symbols with HGNChelper.
 Discovery and replication data were imported separately with mitch and multivariate enrichment analysis was conducted with gene sets from Reactome.
@@
 To examine whether LAM can identify differentially methylated pathways associated with disease prevalence and incidence,
-we downloaded blood based EWAS summary statistics from a recent publication [@Hillary2023-mu], made available via
+we downloaded blood-based EWAS summary statistics from a recent publication [@Hillary2023-mu], made available via
 Zenodo (https://zenodo.org/records/8021411).
@@
 LAM was conducted individually for each disease condition with Reactome gene sets.
-Finally, as a test of type I errors, we randomised each of the 19 incident profiles using
+Finally, as a test of type I errors, we randomised each of the 19 incident profiles using the
 base R `sample` command with a set seed, followed by LAM enrichment and counting the number
@@
 12 pathways were significant in both directions.
-There were 30 pathways that were common between LAM and GSA methods.
+Thirty pathways were common between LAM and GSA methods.
 This includes 26 and 4 with higher and lower methylation respectively (**Figure 3A**).
@@
 Although algorithmic accuracy cannot be inferred from the types of pathways identified in real data [@Timmons2015-kr],
-top ranked results from LAM appeared to be more related to cell differentiation, identity and development (**Figure 3C**),
+top-ranked results from LAM appeared to be more related to cell differentiation, identity and development (**Figure 3C**),
 as compared to GSA (**Figure 3D**).
@@
 (C) Significant LAM pathways with the largest S distance, an enrichment score generated by mitch and effect size proxy.
-(D) Top ranked significant GSA pathways ranked by fold enrichment.](img/fig3.svg)
+(D) Top-ranked significant GSA pathways ranked by fold enrichment.](img/fig3.svg)
 
@@
 group of 37 patients.
-This process was repeated 50 times at each sample size and the results shown in (**Figure 4**).
+This process was repeated 50 times at each sample size and the results are shown in (**Figure 4**).
 Pathways that were identified as statistically significant (FDR<0.05) in the smaller and full group were
@@
 There was no observed overlap between GSA methylation-based pathways and gene expression pathways (**Figure 5A**), but there were some
-overlaps between pathways identified with LAM based promoter and gene body methylation and gene expression (**Figure 5B**).
+overlaps between pathways identified with LAM-based promoter and gene body methylation and gene expression (**Figure 5B**).
 
@@
 We sought to explore the utility of LAM for exploring large EWAS datasets.
-One of the best studied associations with DNA methylation is with chronological aging, although
-pathway level differential methylation is not well defined.
+One of the best-studied associations with DNA methylation is with chronological aging, although
+pathway-level differential methylation is not well defined.
 Using limma summary statistics from independent discovery and replication studies using the EPIC array platform described
-by McCartney et al (2020) [@McCartney2020-ob], we applied the LAM method to explore the pathway level
+by McCartney et al (2020) [@McCartney2020-ob], we applied the LAM method to explore the pathway-level
 differential methylation.
-In the discovery cohort we observed 304 differential pathways (FDR<0.05), while in the replication group we observed 107.
+In the discovery cohort, we observed 304 differential pathways (FDR<0.05), while in the replication group, we observed 107.
 There was a high degree of agreement, with 43 pathways significant in both groups.
@@
 Visualised as a scatterplot, the overall pattern is concordant, although a cluster of points in the lower right of the chart
-indicates 268 pathways with higher methylation in the discovery sample and lower in methylation in the replication
+indicates 268 pathways with higher methylation in the discovery sample and lower methylation in the replication
 sample (**Figure 6B**).
@@
 
-![**Figure 6. Pathway level DNA methylation alterations with chronological age.**
+![**Figure 6. Pathway-level DNA methylation alterations with chronological age.**
 (A) Contour heatmap showing the similarity in gene methylation score ranks in the discovery and replication studies.
@@
 studies since 2009 [@Katari2009-ya].
-Two of the highest powered studies with publicly available data are Estill 2016 which used the HM450K array and Novakovic 2019 study
+Two of the highest-powered studies with publicly available data are Estill 2016 which used the HM450K array and Novakovic 2019 study
 which used the EPIC array [@Estill2016-jh;@Novakovic2019-jt].
@@
 The Novakovic dataset on EPIC array had 793,844 probes that met the filtering criteria (n=133).
-Of these, 5,562 were statistically significant (FDR<0.05) with 1,348 and 4,214 exhibited higher or lower methylation in the IVF group
+Of these, 5,562 were statistically significant (FDR<0.05) with 1,348 and 4,214 exhibiting higher or lower methylation in the IVF group
 respectively.
@@
 The rank-rank plot of differential gene methylation scores indicates a high degree of similarity overall, and interestingly shows a
-trend of more genes having lower methylation level in the IVF group (**Figure 7A**).
+trend of more genes having lower methylation levels in the IVF group (**Figure 7A**).
 A scatterplot of pathway enrichment scores indicates a moderate degree of agreement between these studies
@@
 value of 0.04 (**Figure 7D**).
-Top ranked genes from this pathway include *ADRA2C*, *ADRB2*, *ADRA2B* and *ADRA1B*.
+Top-ranked genes from this pathway include *ADRA2C*, *ADRB2*, *ADRA2B* and *ADRA1B*.
 A loss of methylation of these genes could contribute to reported elevated blood pressure in people conceived by IVF [@Meister2018-cp].
-Taken together, this analysis shows that LAM method can be used to compare enrichment of independent studies
+Taken together, this analysis shows that LAM method can be used to compare pathway enrichment across independent studies
 conducted with HM450K and EPIC arrays.
 
-![**Figure 7. Pathway level DNA methylation differences in natural and IVF conceived infants.**
+![**Figure 7. Pathway-level DNA methylation differences in natural and IVF conceived infants.**
 (A) Contour heatmap showing the similarity in gene methylation score ranks in the Estill (HM450K) and Novakovic (EPIC) studies.
@@
 To examine the potential for LAM to reveal differential pathway methylation associations
-with diseases, we obtained and analysed blood based EWAS summary statistics from a recent
+with diseases, we obtained and analysed blood-based EWAS summary statistics from a recent
 publication [@Hillary2023-mu].
-This study examined 14 prevalent disease states and incidence of 19 disease states in a group
+This study examined methylation in 14 prevalent disease states and the incidence of 19 disease states in a group
 of 18,413 participants, using the EPIC array.
@@
 with a condition.
-Therefore, we examined differential pathway methylation associated with incidence of 19 disease states.
-Over all 19 conditions, there were 1570 significant pathways (FDR<0.05) (**Figure 9A**).
-Liver cirrhosis has the most, with 464, followed by COPD with 400, while Alzheimer's has just one and COVID-19
+Therefore, we examined differential pathway methylation associated with the incidence of 19 disease states.
+Across all 19 conditions, there were 1570 significant pathways (FDR<0.05) (**Figure 9A**).
+Liver cirrhosis had the most, with 464, followed by COPD with 400, while Alzheimer's had just one and COVID-19
 hospitalisation had none.
@@
 Incident Parkinson's disease was associated with reduced methylation of aquaporins.
-In COPD incidence, higher methylation of carbohydrate binding ficolins was observed together with reduced methylation
+In COPD incidence, higher methylation of carbohydrate-binding ficolins was observed together with reduced methylation
 of the GDP mannose synthesis pathway.
@@
 Twenty-eight repeats had fewer than five false positives and eight runs had more than five false positives.
-Across the 100 repeats, the mean number of false positives was 1.95 and median was 0.
+Across the 100 repeats, the mean number of false positives was 1.95 and the median was 0.
 These results indicate that LAM can identify differentially methylated pathways in disease groups and
@@
 The ALM method had only slightly worse performance as compared to LAM, but ALM requires users to aggregate the probe methylation values
-for each sample for each gene prior to limma.
+for each sample for each gene before limma.
 This extra aggregation step is computationally intensive and not a standard part of existing methylation analysis workflows.
@@
 Mitch does not require a large amount of system memory and due to its parallel architecture can
-make use of multi-threaded processors, so it is not much slower than conducting ORA with existing methods that are single threaded.
+make use of multi-threaded processors, so it is not much slower than conducting ORA with existing methods that are single-threaded.
 
-An interesting observation is that ORA performed better when gene sets are smaller in our simulations.
+An interesting observation is that ORA performed relatively better when gene sets were smaller in our simulations.
 Given that popular pathway databases like KEGG and Reactome pathways consist of both large (>100) and small (<20) gene sets,
@@
 subtle signatures, as we have demonstrated with the chronological age, assisted reproduction examples and 19 disease EWAS analyses.
-The analyses conducted here used Reactome pathways, but LAM is general purpose, so different gene sets can be used to examine various hypotheses.
+The analyses conducted here used Reactome pathways, but LAM is general-purpose, so different gene sets can be used to examine various hypotheses.
 Transcription factor target gene sets can be used to identify transcription factors associated with changes in pathway methylation.
@@
 
-The differences in pathway results obtained with LAM and GSA from lung cancer data were striking, not only in the numbers of
-pathways identified, but also their relevance to the disease.
+The differences in pathway results obtained with LAM and GSA from lung cancer data were striking, not only in the number of
+pathways identified but also in their relevance to the disease.
 Interestingly, the relationship between promoter methylation and gene expression was not always inverse as we had anticipated.
@@
 
-Updating defunct gene symbols is an important step as 2,788 and 3,253 gene symbols were updated on the HM450K and EPIC array annotations, which represents approximately 10-14% of genes on these platforms.
+Updating defunct gene symbols is an important step as 2,788 and 3,253 gene symbols were updated on the HM450K and EPIC array annotations, which represent approximately 10-14% of genes on these platforms.
 We note that other packages for enrichment of Infinium array data do not by default address this, which could lead to a loss of these genes from pathway enrichment and diminished sensitivity.

### 2024-02-14 manuscript/manuscript.Rmd
msedits
ms

@@
 ![**Figure 8. Differential pathway methylation associated with prevalence of 14 common disease states.**
-(A) A barplot showing the number of statistically significant pathways with higher and lower methylation identified in each prevalent condition (FDR<0.05).
+(A) A bar plot showing the number of statistically significant pathways with higher and lower methylation identified in each prevalent condition (FDR<0.05).
 (B) A heatmap of S scores for selected pathways across 14 common prevalent disease states.
@@
 ![**Figure 9. Differential pathway methylation associated with incidence of 19 common disease states.**
-(A) A barplot showing the number of statistically significant pathways with higher and lower methylation identified in each incident condition (FDR<0.05).
+(A) A bar plot showing the number of statistically significant pathways with higher and lower methylation identified in each incident condition (FDR<0.05).
 (B) A heatmap of S scores for selected pathways across 19 common incident disease states.
@@
 
+## Data availability
+
+Data sets reanalysed here are publicly available from NCBI GEO, Zenodo or from the supplementary tables of journal articles as described in the Methods section.
+
 ## Code availability
@@
 Authors gratefully acknowledge the contribution to this work of the Victorian Operational Infrastructure Support Program received by the Burnet Institute.
-We thank the researchers who publicly shared data to enable this research.
+We thank the researchers who publicly shared data to enable this work.
 

### 2024-02-14 manuscript/manuscript.Rmd
mswork

@@
 Mary Sravya Kasu<sup>1</sup>,
-Mitch W. Norris<sup>1</sup>,
+Mitchell W. Norris<sup>1</sup>,
 Yen Ting Wong<sup>4,5</sup>,
@@
 | Anusuiya Bora | 0009-0006-2908-1352 |
-| Severine Lamon | 0000-0002-3271-6551 |
+| Séverine Lamon | 0000-0002-3271-6551 |
 | Mary Sravya Kasu | 0000-0002-7891-836X |
-| Yen Ting Wong | 0000-0003-3120-096 |
-| Jeffrey M Craig | 0000-0003-3979-7849 |
+| Mitchell W. Norris | 0009-0001-4338-6252 |
+| Yen Ting Wong | 0000-0003-3120-0964 |
+| Jeffrey M. Craig | 0000-0003-3979-7849 |
 
@@
 meaning the data needs to be summarised from probes to genes and then to gene sets.
-Here, we outline eight potential approaches for FCS of Infinium methylation data, where methods 1-5 are type I
+Here, we outline eight potential approaches for FCS of Infinium methylation data implemented in R (v4.3.2), where methods 1-5 are type I
 and 6-8 are type II:
@@
 1. Limma Average t-test (LAT).
-Differential methylation analysis is conducted at the probe level with limma, and the limma
+Differential methylation analysis is conducted at the probe level with limma (v3.58.1), and the limma
 t-statistics for each gene are summarised (arithmetic mean).
@@
 4. Limma Average Mitch (LAM).
-Similar to LAW, however, instead of the Wilcoxon test, the mitch package[@Kaspi2020-yx] is used to execute a two-way,
+Similar to LAW, however, instead of the Wilcoxon test, the mitch package[@Kaspi2020-yx] (v1.15.0) is used to execute a two-way,
 two-sample ANOVA-on-ranks test as described previously[@Cox2012-zu].
@@
 The above were compared to a standard ORA-based approach (GSA), which involves limma on probes, selection of
-statistically significant probes (FDR<0.05) for ORA with the `gsameth()` function of missMethyl which conducts a
+statistically significant probes (FDR<0.05) for ORA with the `gsameth()` function of missMethyl (v1.36.0) which conducts a
 modified hypergeometric test that accounts for multiple probe biases.
@@
 annotation.
-As gene annotations are several years old, we used HGNChelper to update them (database current as of 21-Dec-2023) [@Oh2022-ts]
+As gene annotations are several years old, we used HGNChelper (v0.8.1) to update them (database current as of 21-Dec-2023) [@Oh2022-ts]
 (3,253 gene symbols were updated).
-Euler diagrams were created with the `eulerr` R package.
+Euler diagrams were created with the `eulerr` R package (v7.0.0).
 To test sensitivity, a random subset of n patients was selected followed by pathway enrichment analysis with the
@@
 fresh in vitro fertilisation (IVF) conceived infants compared to naturally conceived infants.
-Raw intensity files (IDAT format) were read into R with minfi [@Aryee2014-su], normalisation was conducted with the SWAN method
+Raw intensity files (IDAT format) were read into R with minfi [@Aryee2014-su] (v1.48.0), normalisation was conducted with the SWAN method
 [@Maksimovic2012-hy], and probe filtering was conducted to remove proves with a detection p-value > 0.01 in addition to probes
@@
 
+## Funding
+
+This work was supported by a grant from the Australian National Health and Medical Research Council (grant number 1146333 to JMC).
+This research was supported by use of the Nectar Research Cloud, a collaborative Australian research platform supported by the NCRIS-funded Australian Research Data Commons (ARDC).
+Authors gratefully acknowledge the contribution to this work of the Victorian Operational Infrastructure Support Program received by the Burnet Institute.
+
 ## Acknowledgements
 
-This research was supported by use of the Nectar Research Cloud, a collaborative Australian research platform supported by the NCRIS-funded Australian Research Data Commons (ARDC).
-Authors gratefully acknowledge the contribution to this work of the Victorian Operational Infrastructure Support Program received by the Burnet Institute.
 We thank the researchers who publicly shared data to enable this work.

### 2024-02-14 manuscript/manuscript.Rmd
mswork

@@
 (B) Mitch pathway enrichment scores (S distances) in discovery and replication studies.
-Pathways with MANOVA FDR<0.05 are shown in red while others are shaded gray.
+Pathways with MANOVA FDR<0.05 are shown in red while others are shaded grey.
 (C) Heatmap of 30 pathways with most extreme S distances (a surrogate of effect size).
@@
 (B) Mitch pathway enrichment scores (S distances) in Estil and Novakovic studies.
-Pathways with MANOVA FDR<0.05 are shown in red while others are shaded gray.
+Pathways with MANOVA FDR<0.05 are shown in red while others are shaded grey.
 (C) Heatmap of 30 pathways with most extreme S distances (a surrogate of effect size).
@@
 Interestingly, the relationship between promoter methylation and gene expression was not always inverse as we had anticipated.
-Joint enrichment analysis showed some genes with increasing promoter methylation together with increasing RNA expression
+Joint enrichment analysis showed some pathways with increasing promoter methylation together with increasing RNA expression
 (e.g., “Negative regulation of AP2 transcription factors”).
@@
 
-We have demonstrated that this method works with both HM450K and EPIC array data and can even facilitate the comparison of EWAS using different platforms.
+We have demonstrated that this method works with both HM450K and EPIC array data and can even facilitate the comparison of EWAS across different platforms.
 We have provided an example workflow that researchers can use as a template.
@@
 This work was supported by a grant from the Australian National Health and Medical Research Council (grant number 1146333 to JMC).
+Severine Lamon is supported by an Australian Research Council (ARC) Future Fellowship (FT210100278).
 This research was supported by use of the Nectar Research Cloud, a collaborative Australian research platform supported by the NCRIS-funded Australian Research Data Commons (ARDC).

### 2024-02-14 manuscript/manuscript.Rmd
img

@@
 group.
-Left panels correspond to LAM, and right panels to GSA methods.](img/fig4.svg)
+Left panels correspond to LAM, and right panels to GSA methods.](img/fig4.png)
 
@@
 (C) Multi-dimensional enrichment analysis of promoter and gene body methylation with gene expression using mitch.
-Top 20 gene sets shown with the largest S distance after FDR filtering at 0.05.](img/fig5.svg)
+Top 20 gene sets shown with the largest S distance after FDR filtering at 0.05.](img/fig5.png)
 
@@
 (D) An example of a pathway identified with this method, "creation of C4 and C2 activators" shows lower methylation of member genes
-in both discovery and replication studies.](img/fig6.svg)
+in both discovery and replication studies.](img/fig6.png)
 
@@
 (D) An example of a pathway identified with this joint enrichment analysis method, "Adrenoreceptors" shows lower methylation of
-member genes in both Estill and Novakovic studies.](img/fig7.svg)
+member genes in both Estill and Novakovic studies.](img/fig7.png)
 
@@
 Stars indicate that the pathway was identified as being among the top five differentially methylated pathways in each direction for each condition.
-](img/fig8.svg)
+](img/fig8.png)
 

### 2024-02-17 manuscript/manuscript.Rmd
dockerhub link
affiliations

@@
 
-3. Deakin University, School of Exercise and Nutrition, Burwood, Australia.
+3. Deakin University, School of Exercise and Nutrition Sciences, Institute for Physical Activity and Nutrition, Geelong, Australia.
 
@@
 To enable reproduction of the findings, a docker image is made available at DockerHub
-(https://hub.docker.com/repository/docker/mziemann/gmea/general).
+(https://hub.docker.com/repository/docker/mziemann/gmea).
 A reproducible example workflow report is hosted on our website (https://ziemann-lab.net/public/gmea/example_workflow.html).

### 2024-02-22 manuscript/manuscript.Rmd
affiliation

@@
 Séverine Lamon<sup>3</sup>,
-Mary Sravya Kasu<sup>1</sup>,
-Mitchell W. Norris<sup>1</sup>,
+Mary Sravya Kasu<sup>2</sup>,
+Mitchell W. Norris<sup>2</sup>,
 Yen Ting Wong<sup>4,5</sup>,

### 2024-04-09 figs/GSE158422_sensitivity.Rmd
source
added analysis of number of probes

@@
 
+Do significant gene sets bias toward genes with more probes?
+
+```{r,lam_genelength}
+
+genes <- unique(gt$gene)
+probecount <- unlist(lapply(genes,function(gene) { length(which(gt$gene==gene)) } ) )
+names(probecount) <- genes
+head(probecount)
+
+sigsets <- rownames(lamsig)
+nonsigsets <- rownames(subset(lamres,p.adjustANOVA>0.05))
+
+sigmeancounts <- lapply(sigsets, function(set) {
+  genes <- gs_symbols[[which(names(gs_symbols) ==set)]]
+  counts <- probecount[which(names(probecount) %in% genes)]
+  return(mean(counts))
+} )
+
+names(sigmeancounts) <- sigsets
+sigmeancounts <- unlist(sigmeancounts)
+summary(sigmeancounts)
+
+nonsigmeancounts <- lapply(nonsigsets, function(set) {
+  genes <- gs_symbols[[which(names(gs_symbols) ==set)]]
+  counts <- probecount[which(names(probecount) %in% genes)]
+  return(mean(counts))
+} )
+
+names(nonsigmeancounts) <- nonsigsets
+nonsigmeancounts <- unlist(nonsigmeancounts)
+summary(nonsigmeancounts)
+
+mylist <- list("FDR>0.05"=nonsigmeancounts,"FDR<0.05"=sigmeancounts)
+
+boxplot(mylist,col="white",ylab="probes per gene",cex=0,main="LAM",ylim=c(0,100)) ; grid()
+beeswarm(mylist,add=TRUE,pch=1,cex=0.5)
+
+par(mar = c(5.1, 4.1, 4.1, 2.1) )
+pdf("fig3e.pdf",height=6,width=7)
+boxplot(mylist,col="white",ylab="probes per gene",cex=0,main="LAM",ylim=c(0,100)) ; grid()
+beeswarm(mylist,add=TRUE,pch=1,cex=0.5)
+dev.off()
+
+```
+
 ## Downsample LAM test
@@
 
-```{r,al_down1,fig.height=10}
+```{r,al_down1,fig.height=10,eval=FALSE}
 
@@
 
+Are gsameth significant pathways influenced by number of probes per gene?
+
+```{r,gsa_genelength}
+
+head(probecount)
+
+allsets <- sub(" UP$","",rownames(gsaup))
+
+gsaup_sets <- gsub(" UP$","",rownames(subset(gsaup,FDR<0.05)))
+gsadn_sets <- gsub(" DN$","",rownames(subset(gsadn,FDR<0.05)))
+sigsets <- unique(c(gsaup_sets,gsadn_sets))
+length(sigsets)
+
+nonsigsets <- setdiff(allsets,sigsets)
+
+sigmeancounts <- lapply(sigsets, function(set) {
+  genes <- gs_symbols[[which(names(gs_symbols) ==set)]]
+  counts <- probecount[which(names(probecount) %in% genes)]
+  return(mean(counts))
+} )
+
+names(sigmeancounts) <- sigsets
+sigmeancounts <- unlist(sigmeancounts)
+summary(sigmeancounts)
+
+nonsigmeancounts <- lapply(nonsigsets, function(set) {
+  genes <- gs_symbols[[which(names(gs_symbols) ==set)]]
+  counts <- probecount[which(names(probecount) %in% genes)]
+  return(mean(counts))
+} )
+
+names(nonsigmeancounts) <- nonsigsets
+nonsigmeancounts <- unlist(nonsigmeancounts)
+summary(nonsigmeancounts)
+
+mylist <- list("FDR>0.05"=nonsigmeancounts,"FDR<0.05"=sigmeancounts)
+
+boxplot(mylist,col="white",ylab="probes per gene",cex=0,main="GSA",ylim=c(0,100)) ; grid()
+beeswarm(mylist,add=TRUE,pch=1,cex=0.5)
+
+par(mar = c(5.1, 4.1, 4.1, 2.1) )
+pdf("fig3f.pdf",height=6,width=7)
+boxplot(mylist,col="white",ylab="probes per gene",cex=0,main="GSA",ylim=c(0,100)) ; grid()
+beeswarm(mylist,add=TRUE,pch=1,cex=0.5)
+dev.off()
+
+```
+
 ## GSAmeth Downsample
 
-```{r,gsamethdownsample,fig.height=10}
+```{r,gsamethdownsample,fig.height=10,eval=FALSE}
 

### 2024-04-11 manuscript/manuscript.Rmd
manuscript updates

@@
 enrichment score is a useful surrogate for effect size.
+In ORA, The fold enrichment score is a measure of the degree of over-representation of an annotated gene set in the differentially regulated set compared to a background set.
+In FCS, the enrichment score represents the degree to which a set is over-represented at the top or bottom of the ranked list.
 Without an enrichment score, users are limited to using statistical significance values to prioritise results, which
 biases towards large gene sets that have non-specific biological functions.
+In our experience, it is more fruitful to prioritise gene sets by how extreme the enrichment scores are after removing sets that do not meet a predetermined significance threshold.
 
@@
 t-statistics for each gene are summarised (arithmetic mean).
+If a gene has an equal measure of positive and negative t-statistics, the net result will be zero.
 The mean t-statistics are used in a downstream two-sample two-way t-test of gene set enrichment.
@@
 two-sample ANOVA-on-ranks test as described previously[@Cox2012-zu].
+Mitch was selected as it performs favourably in terms of accuracy, is extensible to multiple contrasts and has advanced visualisation options [@Kaspi2020-yx].
 
@@
 6. Aggregate Limma t-test (ALT).
-In this method, all probe measurements belonging to each gene are averaged before conducting differential
+In this method, all probe measurements (M-values) belonging to each gene are averaged before conducting differential
 methylation analysis with limma.
@@
 passed quality control filtering.
+
+We didn’t benchmark our methods against ebGSEA and methylGSA methods because they are non-directional tests, and so have a different null hypothesis as compared to LAM.
 
@@
 the eight FCS methods outlined above.
-Enrichment results were then filtered for FDR<0.05 and the correct direction of methylation change.
+Enrichment results were then filtered for FDR<0.05 and the same direction of methylation change as the ground truth.
 Differentially methylated gene sets observed were compared to gene sets selected to be the ground truth, so that
@@
 DESeq2 data tables.
-CpG sites annotated as promoters were considered separately from those located at gene bodies in mitch analysis.
+CpG sites annotated as promoters (±1.5 kb around transcription start sites) were considered separately from those located at gene bodies (transcription start site to transcription termination site) with LAM.
 For comparison, GSA was used to separately examine pathways using separate analyses of promoter and gene body CpGs.
@@
 Limma summary statistics for chronological age including discovery and replication groups were read into R and enrichment analysis was conducted with the LAM method.
-Briefly, gene-probe associations were first obtained from the "IlluminaHumanMethylationEPICanno.ilm10b4.hg19"
-dataset, followed by an update of gene symbols with HGNChelper.
+Gene-probe associations were obtained and gene symbols updated as above.
 Discovery and replication data were imported separately with mitch and multivariate enrichment analysis was conducted with gene sets from Reactome.
@@
 independent studies.
-The Estill study was published in 2016 and is based on the HM450K array and the data is available from GEO Accession GSE79257
+The Estill study was published in 2016 and is based on the HM450K array and the data is available from GEO accession number GSE79257
 [@Estill2016-jh].
 The Novakovic study in was published 2019, and used the EPIC methylation array and the data is available from GEO under accession
-GSE131433[@Novakovic2019-jt].
+number GSE131433[@Novakovic2019-jt].
 Various assisted reproductive technology conception groups are described in these studies, but we only examined
@@
 Limma differential analysis was conducted accounting for sex, then methylation tables underwent LAM analysis with Reactome gene sets.
-Probe gene associations for HM450K array were established using the "IlluminaHumanMethylation450kanno.ilmn12.hg19" annotation set
-and the gene symbols were updated with HGNChelper.
-As the Novakovic study involved EPIC array data, gene-probe associations were obtained from the
-"IlluminaHumanMethylationEPICanno.ilm10b4.hg19" annotation set, and gene symbols were updated with
-HGNChelper.
+Probe gene associations were obtained from "IlluminaHumanMethylation450kanno.ilmn12.hg19" and "IlluminaHumanMethylationEPICanno.ilm10b4.hg19" Bioconductor packages for HM450K and EPIC array data respectively and gene symbols updated as above.
 
@@
 
+As the number of probes per gene is thought to bias downstream enrichment analysis, we investigated the average probe number for gene sets that were identified as statistically significant or not.
+For LAM, the mean probe number for significant sets was 6% smaller than non-significant sets (32.9 versus 35.2) (**Figure 3E**).
+For GSA the probe number for significant sets was 24% larger than non-significant sets (45.9 versus 36.9) (**Figure 3F**).
+This result indicates that LAM is not adversely biased by the number of probes per gene.
+
 ![**Figure 3. Pathway enrichment analysis comparing normal and cancer samples using LAM and GSA methods.**
@@
 (B) Sizes of significant gene sets found with LAM and GSA methods.
-(C) Significant LAM pathways with the largest S distance, an enrichment score generated by mitch and effect size proxy.
-(D) Top-ranked significant GSA pathways ranked by fold enrichment.](img/fig3.svg)
+(C) Significant LAM pathways with the largest absolute enrichment score.
+(D) Top-ranked significant GSA pathways ranked by fold enrichment.
+(E) Mean number of probes per gene for statistically significant (FDR<0.05) and non-significant (FDR>0.05) gene sets identified with LAM.
+(F) Mean number of probes per gene for statistically significant (FDR<0.05) and non-significant (FDR>0.05) gene sets identified with GSA.
+](img/fig3.svg)
 
@@
 the fewest (3).
-We selected up to five statistically significant pathways with an absolute S distance of >0.4 in either
+We selected up to five statistically significant pathways with an absolute enrichment score of >0.4 in either
 direction in each prevalent condition to include in a heatmap (**Figure 8B**).
@@
 
+In the lung cancer dataset, significant gene sets identified by GSA have on average 24% more probes per gene suggesting GSA has a
+subtle bias towards gene sets where genes have a large number of probes.
+Using the same dataset, LAM does not appear to be biased by the number of probes per gene.
+
 An interesting observation is that ORA performed relatively better when gene sets were smaller in our simulations.
@@
 simulated RNA-seq gene expression data[@Kaspi2020-yx].
-The improved sensitivity of the LAM method provides a new opportunity for researchers to reanalyse previously conducted EWASs with contemporary pathway databases to better understand
+The increased sensitivity of FCS with larger gene sets is probably due to the contribution of probes with FDR>0.05, which confer
+some informative signal in FCS tests, but are disregarded in ORA tests.
+The improved sensitivity of the LAM method provides a new opportunity for researchers to reanalyse previously conducted EWASs with contemporary pathway
+databases to better understand
 subtle signatures, as we have demonstrated with the chronological age, assisted reproduction examples and 19 disease EWAS analyses.
@@
 We have provided an example workflow that researchers can use as a template.
-It shows how to generate a probe-gene table from Bioconductor annotation datasets, how to update the defunct gene names and how to run the enrichment analysis.
-Using this example, generating similar gene tables for the new Infinium Mouse and EPIC 2.0 arrays and any future arrays will be straightforward, once their annotation sets are available through Bioconductor.
+It shows how to generate a probe-gene table from Bioconductor annotation datasets, how to update the defunct gene names and how to run the enrichment
+analysis.
+Using this example, generating similar gene tables for the new Infinium Mouse and EPIC 2.0 arrays and any future arrays will be straightforward, once their
+annotation sets are available through Bioconductor.
 
-Updating defunct gene symbols is an important step as 2,788 and 3,253 gene symbols were updated on the HM450K and EPIC array annotations, which represent approximately 10-14% of genes on these platforms.
-We note that other packages for enrichment of Infinium array data do not by default address this, which could lead to a loss of these genes from pathway enrichment and diminished sensitivity.
+Updating defunct gene symbols is an important step as 2,788 and 3,253 gene symbols were updated on the HM450K and EPIC array annotations, which represent
+approximately 10-14% of genes on these platforms.
+We note that other packages for enrichment of Infinium array data do not by default address this, which could lead to a loss of these genes from pathway
+enrichment and diminished sensitivity.
 
-The analysis of the large-scale disease EWAS shows that LAM can be readily applied to identify differential pathway methylation associated with prevalent and incident disease.
-The authors of the study used GOmeth for their pathway analysis and from the 33 models examined, significant findings were obtained for only two, prevalent diabetes and heart disease.
+The analysis of the large-scale disease EWAS shows that LAM can be readily applied to identify differential pathway methylation associated with prevalent and
+incident disease.
+The authors of the study used GOmeth for their pathway analysis and from the 33 models examined, significant findings were obtained for only two, prevalent
+diabetes and heart disease.
 In contrast, LAM identified 2,469 pathway associations (74.8 per model).
@@
 
-This work could be extended to include not just proximal CpG sites, but also enhancer-based probes and their target genes based on resources such as GeneHancer [@Fishilevich2017-xx].
+As LAM is not locked-in to specific probe-gene annotation schemes, this offers researchers additional flexibility.
+For example, enhancer-target gene relationships from resources like GeneHancer[@Fishilevich2017-xx] could uncover coordinated methylation of distal
+regulatory elements important for regulation of key pathways.
+Moreover, the classification of promoters as regions ±1.5 kb around transcription start sites may be over simplistic and ignore the unique features of many
+genes.
 
-There is also potential to generate new libraries of differentially methylated probes and gene sets, to contribute to the pool of molecular signatures in public resources like MSigDB,
+There is also potential to generate new libraries of differentially methylated probes and gene sets, to contribute to the pool of molecular signatures in
+public resources like MSigDB,
 which will assist in understanding similarities between methylation profiles.
+
+There are some limitations to our method which are the focus of future development.
+Firstly, in ORA and LAM tests, there is a possibility that significance values are inflated due to high correlation between co-located probes; or between
+genes in a tightly regulated network [@Gatti2010-nn; @Wu2012-zr].
+Secondly, in these tests, the unit of sampling is the gene, which doesn’t take into consideration the number of replicate samples or biological variability.
+Alternative enrichment tests such as Globaltest or ssGSEA might overcome this [@Goeman2004-qy; @Barbie2009-vj; @Foroutan2018-ul].
+Lastly, workflows like LAM and GSA require substantial computing and coding skills, and it would be beneficial to provide these as web-based tools, similar
+to the excellent ShinyGO suite [@Ge2020-mw].
+
+In conclusion, LAM is a powerful method for identifying directionally coordinated methylation changes to gene sets that complements existing pathway
+enrichment tools for methylation array data.
 
@@
 
-Data sets reanalysed here are publicly available from NCBI GEO, Zenodo or from the supplementary tables of journal articles as described in the Methods section.
+Data sets reanalysed here are publicly available from NCBI GEO, Zenodo or from the supplementary tables of journal articles as described in the Methods
+section.
 
@@
 Severine Lamon is supported by an Australian Research Council (ARC) Future Fellowship (FT210100278).
-This research was supported by use of the Nectar Research Cloud, a collaborative Australian research platform supported by the NCRIS-funded Australian Research Data Commons (ARDC).
+This research was supported by use of the Nectar Research Cloud, a collaborative Australian research platform supported by the NCRIS-funded Australian
+Research Data Commons (ARDC).
 Authors gratefully acknowledge the contribution to this work of the Victorian Operational Infrastructure Support Program received by the Burnet Institute.

### 2024-04-11 manuscript/manuscript.Rmd
simplify manuscript
s dist --> enrichment score

@@
 Subsequently, to generate the pathway heatmap, we used mitch in multivariate mode by analysing differential RNA expression,
-promoter and gene body methylation together, prioritising the results by S distance, a surrogate measure of effect size.
+promoter and gene body methylation together, prioritising the results by S distance, the enrichment score provided by mitch.
 
@@
 (C) Multi-dimensional enrichment analysis of promoter and gene body methylation with gene expression using mitch.
-Top 20 gene sets shown with the largest S distance after FDR filtering at 0.05.](img/fig5.png)
+Top 20 gene sets shown with the largest absolute enrichment scores after FDR filtering at 0.05.](img/fig5.png)
 
@@
 sample (**Figure 6B**).
-When prioritising results by S distance, the enrichment score reported by mitch and a surrogate measure of effect size,
+When prioritising results by enrichment score,
 29 of the top 30 pathways were concordant in their direction of regulation (**Figure 6C**).
@@
 (A) Contour heatmap showing the similarity in gene methylation score ranks in the discovery and replication studies.
-(B) Mitch pathway enrichment scores (S distances) in discovery and replication studies.
+(B) Mitch pathway enrichment scores in discovery and replication studies.
 Pathways with MANOVA FDR<0.05 are shown in red while others are shaded grey.
-(C) Heatmap of 30 pathways with most extreme S distances (a surrogate of effect size).
+(C) Heatmap of 30 pathways with largest absolute enrichment scores.
 Red indicates increasing methylation and blue shows lower methylation.
@@
 (A) Contour heatmap showing the similarity in gene methylation score ranks in the Estill (HM450K) and Novakovic (EPIC) studies.
-(B) Mitch pathway enrichment scores (S distances) in Estil and Novakovic studies.
+(B) Mitch pathway enrichment scores in Estil and Novakovic studies.
 Pathways with MANOVA FDR<0.05 are shown in red while others are shaded grey.
-(C) Heatmap of 30 pathways with most extreme S distances (a surrogate of effect size).
+(C) Heatmap of 30 pathways with largest absolute enrichment score.
 Red indicates higher methylation and blue shows lower methylation.

### 2024-04-11 manuscript/manuscript.Rmd
s score --> enrichment score

@@
 (A) A bar plot showing the number of statistically significant pathways with higher and lower methylation identified in each prevalent condition (FDR<0.05).
-(B) A heatmap of S scores for selected pathways across 14 common prevalent disease states.
+(B) A heatmap of enrichment scores for selected pathways across 14 common prevalent disease states.
 Stars indicate that the pathway was identified as being among the top five differentially methylated pathways in each direction for each condition.
@@
 (A) A bar plot showing the number of statistically significant pathways with higher and lower methylation identified in each incident condition (FDR<0.05).
-(B) A heatmap of S scores for selected pathways across 19 common incident disease states.
+(B) A heatmap of enrichment scores for selected pathways across 19 common incident disease states.
 Stars indicate that the pathway was identified as being among the top three differentially methylated pathways in each direction for each condition.

### 2024-04-18 manuscript/manuscript.Rmd
explain enrichment score and effect size
fix typo

@@
 promoter and gene body methylation together, prioritising the results by S distance, the enrichment score provided by mitch.
+This score can be used as a surrogate measure of effect size, similar to how a fold change might be used as an effect size for individual genes.
 
@@
 [@Estill2016-jh].
-The Novakovic study in was published 2019, and used the EPIC methylation array and the data is available from GEO under accession
+The Novakovic study was published in 2019, and used the EPIC methylation array and the data is available from GEO under accession
 number GSE131433[@Novakovic2019-jt].

### 2024-06-22 figs/multi_ewas_example.Rmd
cancer bioinformatics symposium

@@
 
+Make a special heatmap for the cancer bioinformamtics symposium.
+
+```{r,cancerheat}
+
+names(r1) <- names(l2)
+
+r3 <- r1[grep("cancer",names(r1))]
+
+r4 <- lapply(r3,function(r) {
+  r <- subset(r,p.adjustANOVA<0.05)
+  ups <- head(r[order(-r$s.dist),],3)$set
+  dns <- head(r[order(r$s.dist),],3)$set
+  list("ups"=ups,"dns"=dns)
+} )
+
+r4
+
+gsets <- unique(unname(unlist(r4)))
+gsets
+
+top <- mtable2[which(mtable2$set %in% gsets),]
+top <- top[,c(1,4:22)]
+top <- top[,colnames(top) %in% c("set","s.breast_cancer","s.colorectal_can","s.lung_cancer","s.ovarian_cancer","s.prostate_cance")]
+
+rownames(top) <- top$set
+top$set=NULL
+head(top,3)
+
+rownames(top) <- gsub("REACTOME_","",rownames(top))
+rownames(top) <- gsub("_"," ",rownames(top))
+
+colnames(top) <- c("breast","colorectal","lung","ovarian","prostate")
+head(top)
+
+saveRDS(top,file="cancerheat.Rds")
+
+colfunc <- colorRampPalette(c("blue", "white", "red"))
+
+heatmap.2(as.matrix(top),scale="none",trace="none",margins=c(10,25),
+  col=colfunc(25),cexRow=0.75,cexCol=1.2)
+
+pdf("cancerheat.pdf",height=7,width=7)
+heatmap.2(as.matrix(top),scale="none",trace="none",margins=c(10,25),
+  col=colfunc(25),cexRow=0.7,cexCol=1.2)
+dev.off()
+
+```
+
 ## Randomise data

### 2024-06-28 example_workflow/example_workflow.Rmd
add mitch version warning
update for EPIC V2

@@
 
+Important: ensure that the mitch version used is 1.15.1 or higher.
+
 ```{r,packages}
@@
   library("eulerr")
+  library("AnnotationHub")
   library("IlluminaHumanMethylation450kanno.ilmn12.hg19")
@@
 })
+
+if( packageVersion("mitch") < "1.15.1") {
+  warning("This workflow requires mitch version 1.15.1 or higher")
+}
 
@@
 gt1$probe <- as.character(gt1$probe)
+toc() #6.0s
+
 dim(gt1)
@@
 length(unique(gt1$gene))
-toc() #6.0s
 
@@
 gp2 <- lapply(gp2,unique)
-gt <- stack(gp2)
-colnames(gt) <- c("gene","probe")
-gt$probe <- as.character(gt$probe)
-dim(gt)
-str(gt)
+gt2 <- stack(gp2)
+colnames(gt2) <- c("gene","probe")
+gt2$probe <- as.character(gt2$probe)
 toc() #9.0s
 
+dim(gt2)
+str(gt2)
+length(unique(gt2$gene))
+head(gt2)
+
 tic()
-#new.hgnc.table <- getCurrentHumanMap()
 new.hgnc.table <- readRDS("../figs/new.hgnc.table.rds")
-fix <- checkGeneSymbols(gt$gene,map=new.hgnc.table)
+fix <- checkGeneSymbols(gt2$gene,map=new.hgnc.table)
 fix2 <- fix[which(fix$x != fix$Suggested.Symbol),]
 length(unique(fix2$x))
-gt$gene <- fix$Suggested.Symbol
+gt2$gene <- fix$Suggested.Symbol
 toc()
+
+head(gt2)
+
+```
+
+## Make gene table for the new EPIC version 2 array
+
+For details, see the [Bioconductor package page](https://bioconductor.org/packages/EPICv2manifest/).
+
+```{r,gt3}
+
+tic()
+ah <- AnnotationHub()
+EPICv2manifest <- ah[["AH116484"]]
+myann <- data.frame(EPICv2manifest[,c("UCSC_RefGene_Name","UCSC_RefGene_Group","UCSC_CpG_Islands_Name","Relation_to_UCSC_CpG_Island","posrep_RepNum")])
+myann <- subset(myann,posrep_RepNum==1)
+gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
+gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
+names(gp2) <- rownames(gp)
+gp2 <- lapply(gp2,unique)
+gt3 <- stack(gp2)
+colnames(gt3) <- c("gene","probe")
+gt3$probe <- as.character(gt3$probe)
+dim(gt3)
+str(gt3)
+toc()
+
+tic()
+new.hgnc.table <- readRDS("../figs/new.hgnc.table.rds")
+fix <- checkGeneSymbols(gt3$gene,map=new.hgnc.table)
+fix2 <- fix[which(fix$x != fix$Suggested.Symbol),]
+length(unique(fix2$x))
+gt3$gene <- fix$Suggested.Symbol
+toc()
+
+head(gt3)
 

### 2024-07-19 figs/multi_ewas_example.Rmd
timeout

@@
   setwd("hillary")
+  options(timeout=180)
   download.file("https://zenodo.org/records/8021411/files/compressed-disease-states-ewas.tar.gz?download=1",destfile="compressed-disease-states-ewas.tar.gz")
